\chapter{Fields}
\label{chap:fields}

The minuscule amount of algebra we have developed so far allows us to
scratch the surface of the unfathomable subject of \emph{field theory}.
Fields are of basic importance in many subjects---number theory and
algebraic geometry come immediately to mind---and it is not surprising that
their study has been developed to a particularly high level of
sophistication. While my profoundly limited knowledge will prevent me from
even hinting at this sophistication, even a cursory overview of the basic
notions will allow us to deal with remarkable applications, such as the
famous problems of constructibility of geometric figures. We will also get a
glimpse of the beautiful subject of \emph{Galois theory}, an amazing
interplay of the theory of field extensions, the solvability of polynomials,
and group theory.

\section{Field extensions, I}
\label{sec:fields:field-extensions-i}

We first deal with several basic notions in field theory, mostly inspired
by easy linear algebra considerations. The keywords here are \emph{finite},
\emph{simple}, \emph{finitely generated}, \emph{algebraic}.

\subsection{Basic definitions}
\label{subsec:fields:basic-definitions}
The study of fields is (of course) the study of the category \textsf{Fld}
of fields (Definition~III.1.14), with ring homomorphisms as morphisms. The
task is to understand what fields are and above all what they are \emph{in
relation to one another}.

The place to start is a reminder from elementary ring theory\footnote{The
last time I have made this point was back in \S{}V.5.2.}: \emph{every ring
homomorphism from a field to a nonzero ring is injective}. Indeed, the
kernel of a ring homomorphism to a nonzero ring is a proper ideal (because
a homomorphism maps $1$ to $1$ by definition and $1 \neq 0$ in nonzero
rings), and the only proper ideal in a field is $(0)$. In particular, every
ring homomorphism of fields is injective (fields are nonzero rings by
definition!); every morphism in \textsf{Fld} is a monomorphism (cf.\
Proposition~III.2.4).
Thus, every morphism $k \to K$ between two fields identifies the first with
a subfield of the second. In other words, one can view $K$ as a particular
way to enlarge $k$, that is, as an \emph{extension} of $k$. Field theory
is first and foremost the study of \emph{field extensions}. I will denote
a field extension by $k \subseteq K$ (which is less than optimal, because
there may be many ways to embed a field into another); other popular choices
are $K/k$ (which I do not like, as it suggests a quotienting operation) and
$\frac{K}{k}$ (which we will avoid most of the time, as it is hard to
typeset).
The coarsest invariant of a field $k$ is its \emph{characteristic}; cf.\
Exercise~V.4.17. We have a unique ring homomorphism $i : \mathbb{Z} \to k$
($\mathbb{Z}$ is initial in \textsf{Ring}: $1$ must go to $1$ by definition
of homomorphism, and this fixes the value of $i(n)$ for all $n \in
\mathbb{Z}$); the \emph{characteristic} of $k$, $\operatorname{char} k$, is
defined to be the nonnegative generator of the ideal $\ker i$; that is,
$\operatorname{char} k = 0$ if $i$ is injective, and $\operatorname{char} k
= p > 0$ if $\ker i = (p) \neq (0)$.

Put otherwise, either $n \cdot 1$ is only $0$ in $k$ for $n = 0$, in which
case $\operatorname{char} k = 0$, or $n \cdot 1 = 0$ for some $n \neq 0$;
$\operatorname{char} k = p > 0$ is then the smallest positive integer for
which $p \cdot 1 = 0$ in $k$. Since fields are integral domains, the image
$i(\mathbb{Z})$ must be an integral domain; hence $\ker i$ must be a prime
ideal. Therefore, the characteristic of a field is either $0$ or a prime
number.
If $k \subseteq K$ is an extension, then $\operatorname{char} k =
\operatorname{char} K$ (Exercise~1.1). Thus, we could define and study
categories $\textsf{Fld}_0$, $\textsf{Fld}_p$ of fields of a given
characteristic, without losing any information: these categories live within
\textsf{Fld}, without interacting with each other.\footnote{They do interact
in other ways, however---for example, because $\mathbb{Z}$ is initial in
\textsf{Ring}, so $\mathbb{Z}$ maps to all fields.} Further, each of these
categories has an initial object: $\mathbb{Q}$ is initial in
$\textsf{Fld}_0$, and\footnote{I am going to denote the field
$\mathbb{Z}/p\mathbb{Z}$ by $\mathbb{F}_p$ in this chapter, to stress its
role as a field (rather than `just' as a group).} $\mathbb{F}_p :=
\mathbb{Z}/p\mathbb{Z}$ in $\textsf{Fld}_p$; the reader has checked this
easy fact in Exercise~V.4.17, and now is an excellent time to contemplate it
again. In each case, the initial object is called the \emph{prime subfield}
of the given field. Thus, every field is in a canonical way an extension of
its prime subfield: studying fields really means studying field extensions.
To a large extent, the `small' field $k$ will be fixed in what follows, and
our main object of study will be the category $\textsf{Fld}_k$ of extensions
of $k$,
with the evident (and by now familiar to the reader) notions of morphisms.

The first general remark concerning field extensions is that the larger
field is an algebra, and hence a \emph{vector space} over the smaller one
(by the very definition of algebra; cf.\ Example~III.5.6).

\begin{defn}{}{fields:1.1}
A field extension $k \subseteq F$ is \emph{finite}, of \emph{degree} $n$,
if $F$ has (finite) dimension $\dim F = n$ as a vector space over $k$. The
extension is \emph{infinite} otherwise.

The degree of a finite extension $k \subseteq F$ is denoted by $[F : k]$
(and we write $[F : k] = \infty$ if the extension is infinite).
\end{defn}
In \S{}V.5.2 we have encountered a prototypical example of finite field
extension: a procedure starting from an irreducible polynomial $f(x) \in
k[x]$ with coefficients in a field and producing an extension $K$ of $k$ in
which $f(t)$ has a root. Explicitly,
\[K = \frac{k[t]}{(f(t))}\]
is such an extension (Proposition~V.5.7). The quotient is a field because
$k[t]$ is a PID; hence irreducible elements generate maximal ideals---prime
because of the UFD property (cf.\ Theorem~V.2.5) and then maximal because
nonzero prime ideals are maximal in a PID (cf.\ Proposition~III.4.13). The
coset of $t$ is a root of $f(x)$ when this is viewed as an element of
$K[x]$. The degree of the extension $k \subseteq K$ equals the degree of
the polynomial $f(x)$ (this is hopefully clear at this point and was checked
carefully in Proposition~III.4.6).
The reader may wonder whether perhaps \emph{all} finite extensions are of
this type. This is unfortunately not the case, but as it happens, it is the
case in a large class of examples. As I often do, I promise that the
situation will clarify itself considerably once we have accumulated more
general knowledge; in this case, the relevant fact will be
Proposition~5.19.

Also recall that we proved that these extensions are almost universal with
respect to the problem of extending $k$ so that the polynomial $f(t)$
acquires a root. I pointed out, however, that the \emph{uni} in \emph{uni}versal
is missing; that is, if $k \subseteq F$ is an extension of $k$ in which
$f(t)$ has a root, there may be many different ways to put $K$ in between:
\[k \subseteq K \subseteq F.\]
Such questions are central to the study of extensions, so we begin by
giving a second look at this situation.

\subsection{Simple extensions}
\label{subsec:fields:simple-extensions}
Let $k \subseteq F$ be a field extension, and let $\alpha \in F$. The
smallest subfield of $F$ containing both $k$ and $\alpha$ is denoted
$k(\alpha)$; that is, $k(\alpha)$ is the intersection of all subfields of
$F$ containing $k$ and $\alpha$.

\begin{defn}{}{fields:1.2}
A field extension $k \subseteq F$ is \emph{simple} if there exists an
element $\alpha \in F$ such that $F = k(\alpha)$.
\end{defn}
The extensions recalled above are of this kind: if $K = k[t]/(f(t))$ and
$\alpha$ denotes the coset of $t$, then $K = k(\alpha)$: indeed, if a
subfield of $K$ contains the coset of $t$, then it must contain (the coset
of) every polynomial expression in $t$, and hence it must be the whole of
$K$.

I have always found the notation $k(\alpha)$ somewhat unfortunate, since it
suggests that all such extensions are in some way isomorphic and possibly
all isomorphic to the field $k(t)$ of rational functions in one
indeterminate $t$ (cf.\ Definition~V.4.13). This is not true, although it
is clear that every element of $k(\alpha)$ may be written as
a rational function in $\alpha$ with coefficients in $k$ (Exercise~1.3).
In any case, it is easy to classify simple extensions: they are either
isomorphic to $k(t)$ or they are of the prototypical kind recalled above.
Here is the precise statement.

\begin{prop}{}{fields:1.3}
Let $k \subseteq k(\alpha)$ be a simple extension. Consider the evaluation
map $\epsilon : k[t] \to k(\alpha)$, defined by $f(t) \mapsto f(\alpha)$.
Then we have the following:
\begin{itemize}
\item $\epsilon$ is injective if and only if $k \subseteq k(\alpha)$ is an
  infinite extension. In this case, $k(\alpha)$ is isomorphic to the field
  of rational functions $k(t)$.
\item $\epsilon$ is not injective if and only if $k \subseteq k(\alpha)$ is
  finite. In this case there exists a unique monic irreducible nonconstant
  polynomial $p(t) \in k[t]$ of degree $n = [k(\alpha) : k]$ such that
\[k(\alpha) \cong \frac{k[t]}{(p(t))}.\]
  Via this isomorphism, $\alpha$ corresponds to the coset of $t$. The
  polynomial $p(t)$ is the monic polynomial of smallest degree in $k[t]$
  such that $p(\alpha) = 0$ in $k(\alpha)$.
\end{itemize}
\end{prop}
The polynomial $p(t)$ appearing in this statement is called the
\emph{minimal polynomial} of $\alpha$ over $k$.

Of course the minimal polynomial of an element $\alpha$ of a (`large')
field depends on the base (`small') field $k$. For example, $\sqrt{2} \in
\mathbb{C}$ has minimal polynomial $t^2 - 2$ over $\mathbb{Q}$, but $t -
\sqrt{2}$ over $\mathbb{R}$.

\begin{proof}
Let $F = k(\alpha)$. By the `first isomorphism theorem', the image of
$\epsilon : k[t] \to F$ is isomorphic to $k[t]/\ker(\epsilon)$. Since $F$
is an integral domain, so is $k[t]/\ker(\epsilon)$; hence $\ker(\epsilon)$
is a prime ideal in $k[t]$.
---Assume $\ker(\epsilon) = 0$; that is, $\epsilon$ is an injective map
from the integral domain $k[t]$ to the field $F$. By the universal property
of fields of fractions (cf.\ \S{}V.4.2), $\epsilon$ extends to a unique
homomorphism
\[k(t) \to F.\]
The (isomorphic) image of $k(t)$ in $F$ is a field containing $k$ and
$\alpha$; hence it equals $F$ by definition of simple extension.

Since $\epsilon$ is injective, the powers $\alpha^0 = 1, \alpha, \alpha^2,
\alpha^3, \dots$ (that is, the images $\epsilon(t^i)$) are all distinct and
linearly independent over $k$ (because the powers $1, t, t^2, \dots$ are
linearly independent over $k$); therefore the extension $k \subseteq F$ is
infinite in this case.
---If $\ker(\epsilon) \neq 0$, then $\ker(\epsilon) = (p(t))$ for a unique
monic irreducible nonconstant polynomial $p(t)$, which has smallest degree
(cf.\ Exercise~III.4.4!) among all nonzero polynomials in $\ker(\epsilon)$.
As $(p(t))$ is then maximal in $k[t]$, the image of $\epsilon$ is a
subfield of $F$ containing $\alpha = \epsilon(t)$. By definition of simple
extension, $F = $ the image of $\epsilon$; that is, the induced
homomorphism
\[\frac{k[t]}{(p(t))} \to F\]
is an isomorphism. In this case $[F : k] = \deg p(t)$, as recalled in
\S{}\ref{subsec:fields:basic-definitions}, and in particular the extension
is finite, as claimed.
\end{proof}

The alert reader will have noticed that the proof of \cref{prop:fields:1.3}
is essentially a rehash of the argument proving the `versality' part of
Proposition~V.5.7.
\begin{exam}{}{fields:exam:1.4}
Consider the extension $\mathbb{Q} \subseteq \mathbb{R}$.

The polynomial $x^2 - 2 \in \mathbb{Q}[x]$ has roots in $\mathbb{R}$:
therefore, by Proposition~V.5.7 there exists a homomorphism (hence a field
extension)
\[\epsilon : \frac{\mathbb{Q}[t]}{(t^2 - 2)} \hookrightarrow \mathbb{R},\]
such that the image of (the coset of) $t$ is a root $\alpha$ of $x^2 - 2$.
\cref{prop:fields:1.3} simply identifies the image of this homomorphism
with $\mathbb{Q}(\alpha) \subseteq \mathbb{R}$.
This is hopefully crystal clear; however, note that even this simple
example shows that the induced morphism $\epsilon$ is not unique (hence the
`lack of \emph{uni}'): because there are more than one root of $x^2 - 2$
in $\mathbb{R}$. Concretely, there are two possible choices for $\alpha$:
$\alpha = +\sqrt{2}$ and $\alpha = -\sqrt{2}$. The choice of $\alpha$
determines the evaluation map $\epsilon$ and therefore the specific
realization of $\mathbb{Q}[t]/(t^2 - 2)$ as a subfield of $\mathbb{R}$.

The reader may be misled by one feature of this example: clearly
$\mathbb{Q}(\sqrt{2}) = \mathbb{Q}(-\sqrt{2})$, and this may seem to
compensate for the lack of uniqueness lamented above. The morphism may not
be unique, but the image of the morphism surely is? No. The reader will
check that there are three distinct subfields of $\mathbb{C}$ isomorphic to
$\mathbb{Q}[t]/(t^3 - 2)$ (Exercise~1.5).
\end{exam}

Ay, there's the rub. One of our main goals in this chapter will be to
single out a condition on an extension $k \subseteq F$ that guarantees that
no matter how we
embed $F$ in a larger extension, the images of these (possibly many
different) embeddings will all coincide. Up to further technicalities, this
is what makes $\mathbb{Q} \subseteq \mathbb{Q}(\sqrt{2})$ a Galois
extension, while $\mathbb{Q} \subseteq \mathbb{Q}(\sqrt[3]{2})$ will not be
a Galois extension. But Galois extensions will have to wait until \S{}6,
and the reader can put this issue aside for now.

One way to recover uniqueness is to incorporate the choice of the root in
the data. This leads to the following refinement of the (uni)versality
statement, adapted for future applications.
\begin{prop}{}{fields:1.5}
Let $k_1 \subseteq F_1 = k_1(\alpha_1)$, $k_2 \subseteq F_2 = k_2(\alpha_2)$
be two finite simple extensions. Let $p_1(t) \in k_1[t]$, resp., $p_2(t)
\in k_2[t]$, be the minimal polynomials of $\alpha_1$, resp., $\alpha_2$.
Let $i : k_1 \to k_2$ be an isomorphism, such that\footnote{Of course any
homomorphism of rings $f : R \to S$ induces a unique ring homomorphism $f :
R[t] \to S[t]$ sending $t$ to $t$, named in the same way by a harmless
abuse of language.}
\[i(p_1(t)) = p_2(t).\]
Then there exists a unique isomorphism $j : F_1 \to F_2$ agreeing with $i$
on $k_1$ and such that $j(\alpha_1) = \alpha_2$.
\end{prop}
\begin{proof}
Since every element of $k_1(\alpha_1)$ is a linear combination of powers of
$\alpha_1$ with coefficients in $k_1$, $j$ is determined by its action on
$k_1$ (which agrees with $i$) and by $j(\alpha_1)$, which is prescribed to
be $\alpha_2$. Thus an isomorphism $j$ as in the statement is uniquely
determined.

To see that the isomorphism $j$ exists, note that as $i$ maps $p_1(t)$ to
$p_2(t)$, it induces an isomorphism
\[\frac{k_1[t]}{(p_1(t))} \xrightarrow{\;\sim\;} \frac{k_2[t]}{(p_2(t))}.\]
Composing with the isomorphisms found in \cref{prop:fields:1.3} gives $j$:
\[j : k_1(\alpha_1) \xrightarrow{\;\sim\;} \frac{k_1[t]}{(p_1(t))}
  \xrightarrow{\;\sim\;} \frac{k_2[t]}{(p_2(t))} \xrightarrow{\;\sim\;}
  k_2(\alpha_2),\]
as needed.
\end{proof}
Thus, isomorphisms lift uniquely to simple extensions, so long as they
preserve minimal polynomials. We may say that $j$ \emph{extends} $i$, and
draw the following diagram of extensions:
\[\begin{tikzcd}
  k_1(\alpha_1) \arrow[r, "j", "\sim"'] & k_2(\alpha_2) \\
  k_1 \arrow[u, hook] \arrow[r, "i", "\sim"'] & k_2 \arrow[u, hook]
\end{tikzcd}\]
We will be particularly interested in considering isomorphisms of a field
to itself, subject to the condition of fixing a specified subfield $k$
(that is, extending the identity on $k$), that is, automorphisms in
$\textsf{Fld}_k$. Explicitly, for $k \subseteq F$ an extension, we will
analyze isomorphisms $j : F \xrightarrow{\sim} F$ such that $\forall c \in
k$, $j(c) = c$. I will denote the group of such automorphisms by
$\operatorname{Aut}_k(F)$. This is probably the most important object of
study in this chapter, so I should make its definition
official:

\begin{defn}{}{fields:1.6}
Let $k \subseteq F$ be a field extension. The \emph{group of automorphisms
of the extension}, denoted $\operatorname{Aut}_k(F)$, is the group of field
automorphisms $j : F \to F$ such that $j|_k = \mathrm{id}_k$.
\end{defn}

\begin{coro}{}{fields:1.7}
Let $k \subseteq F = k(\alpha)$ be a simple finite extension, and let
$p(x)$ be the minimal polynomial of $\alpha$ over $k$. Then
$|\operatorname{Aut}_k(F)|$ equals the number of distinct roots of $p(x)$
in $F$; in particular,
\[|\operatorname{Aut}_k(F)| \leq [F : k],\]
with equality if and only if $p(x)$ factors over $F$ as a product of
distinct linear polynomials.
\end{coro}

\begin{proof}
Let $j \in \operatorname{Aut}_k(F)$. Since every element of $F$ is a
polynomial expression in $\alpha$ with coefficients in $k$, and $j$ extends
the identity on $k$, $j$ is determined by $j(\alpha)$. Now
\[p(j(\alpha)) = j(p(\alpha)) = j(0) = 0:\]
therefore, $j(\alpha)$ is necessarily a root of $p(x)$. This shows that
$|\operatorname{Aut}_k(F)|$ is no larger than the number of roots of $p(x)$.

On the other hand, by \cref{prop:fields:1.5} every choice of a root of
$p(t)$ determines a lift of the identity to an element $j \in
\operatorname{Aut}_k(F)$, establishing the other inequality.
\end{proof}
\cref{coro:fields:1.7} is our first hint of the powerful interaction
between group theory and the theory of field extensions; this theme will
haunt us throughout the chapter.

The proof of \cref{coro:fields:1.7} really sets up a bijection between the
roots of an irreducible polynomial $p(t) \in k[t]$ in a larger field $F$
and the group $\operatorname{Aut}_k(F)$. A particularly optimistic reader
may then hope that the roots of $p(t)$ come endowed with a group structure,
but this is hoping for too much: the bijection depends on the choice of one
root $\alpha$, and no one root of $p(t)$ is `more beautiful' than the
others.

However, we have established that if $F = k(\alpha)$ is a simple extension
of $k$, then the group $\operatorname{Aut}_k(F)$ acts faithfully and
transitively on the set of roots of $p(t)$ in $F$. In other words,
$\operatorname{Aut}_k(F)$ can be identified with a certain subgroup of the
symmetric group acting on the set of roots of $p(t)$.

More generally (Exercise~1.6) the automorphisms of any extension act on
roots of polynomials. One conclusion we can draw from these preliminary
considerations is that the analysis of $\operatorname{Aut}_k(F)$ is
substantially simplified if $k \subseteq F$ is a simple extension $k(\alpha)$
and further if the minimal polynomial $p(x)$ of $\alpha$ factors into
$\deg p(x)$ distinct linear terms in $F$. It will not come as a surprise
that we will focus our attention on such extensions in later sections: an
extension is \emph{Galois} precisely if it is of this type. (The reader is
invited to remember this fact, but its import will not be fully appreciated
until we develop substantially more material.)

\subsection{Finite and algebraic extensions}
\label{subsec:fields:finite-and-algebraic-extensions}
The dichotomy encountered in \cref{prop:fields:1.3} provides us with one
of the most important definitions in field theory:

\begin{defn}{}{fields:1.8}
Let $k \subseteq F$ be a field extension, and let $\alpha \in F$. Then
$\alpha$ is \emph{algebraic} over $k$, of \emph{degree} $n$ if $n =
[k(\alpha) : k]$ is finite; $\alpha$ is \emph{transcendental} over $k$
otherwise.

The extension $k \subseteq F$ is \emph{algebraic} if every $\alpha \in F$
is algebraic over $k$.
\end{defn}
By \cref{prop:fields:1.3}, $\alpha \in F$ is algebraic over $k$ if and only
if there exists a nonzero polynomial $f(x) \in k[x]$ such that $f(\alpha) =
0$. The minimal polynomial of $\alpha$ is the monic polynomial of smallest
degree satisfying this condition; as we have seen, it is necessarily
irreducible.
Also note that if $\alpha$ is algebraic over $k$, then every element of
$k(\alpha)$ may in fact be written as a \emph{polynomial} with coefficients
in $k$.

\emph{Finite extensions are necessarily algebraic:}
\begin{lem}{}{fields:1.9}
Let $k \subseteq F$ be a finite extension. Then every $\alpha \in F$ is
algebraic over $k$, of degree $\leq [F : k]$.
\end{lem}
\begin{proof}
Since $k \subseteq k(\alpha) \subseteq F$, the dimension of $k(\alpha)$ as a
$k$-vector space is bounded by $\dim_k F = [F : k]$.
\end{proof}
Concretely, if $k \subseteq F$ is finite and $\alpha \in F$, then the powers
$1, \alpha, \alpha^2 \dots$ are necessarily linearly dependent; and any
nontrivial linear dependence relation among them provides us with a nonzero
polynomial $f(x) \in k[x]$ such that $f(\alpha) = 0$.
The literal converse of the claim that `finite extensions are algebraic' is
not true; we will see an example in a moment. However, something along these
lines holds; understanding the situation requires a more careful look at
finiteness conditions.
First of all, compositions of finite extensions are finite extensions, and
the degree behaves nicely with respect to this operation:
\begin{prop}{}{fields:1.10}
Let $k \subseteq E \subseteq F$ be field extensions. Then $k \subseteq F$ is
finite if and only if both $k \subseteq E$ and $E \subseteq F$ are finite.
In this case,
\[
  [F : k] = [F : E][E : k].
\]
\end{prop}
\begin{proof}
If $F$ is finite-dimensional as a vector space over $k$, then so is its
subspace $E$; and any linear dependence relation of elements of $F$ over the
field $k$ gives one over the larger field $E$. It follows that if $k
\subseteq F$ is finite, then so are $k \subseteq E$ and $E \subseteq F$.
Conversely, assume $k \subseteq E$ and $E \subseteq F$ are both finite. Let
$(e_1, \dots, e_m)$ be a basis for $E$ over $k$, and let $(f_1, \dots, f_n)$
be a basis for $F$ over $E$. It will suffice to show that the $mn$ products
\[
  (e_1 f_1, e_1 f_2, \dots, e_m f_n)
\]
form a basis for $F$ over $k$.
Let $g \in F$. Then $\exists d_1, \dots, d_n \in E$ such that
\[
  g = \sum_{j=1}^{n} d_j f_j,
\]
since the elements $f_j$ span $F$ over $E$. Since $E$ is spanned by the
elements $e_i$ over $k$, $\exists c_{1j}, \dots, c_{mj} \in k$ such that
$\forall j$
\[
  d_j = \sum_{i=1}^{m} c_{ij} e_i.
\]
It follows that
\[
  g = \sum_{i=1}^{m} \sum_{j=1}^{n} c_{ij} e_i f_j:
\]
therefore the products $e_i f_j$ span $F$ over $k$. (This suffices to prove
that $k \subseteq F$ is finite.)
To verify that these elements are linearly independent, assume
\[
  \sum_{i,j} \lambda_{ij} e_i f_j = 0
\]
for $\lambda_{ij} \in k$. Then we have
\[
  \sum_{j} \Bigl(\sum_{i} \lambda_{ij} e_i\Bigr) f_j = 0,
\]
implying $\sum_i \lambda_{ij} e_i = 0$ for all $j$, since the elements $f_j$
are linearly independent over $E$. As the elements $e_i$ are linearly
independent over $k$, this shows that all $\lambda_{ij}$ equal $0$, as
needed.
\end{proof}
The formula given in the statement should look familiar to the reader: glance
at the end of \S{}II.8.5. Of course this is not accidental; the reader
should take it as another hint that the world of groups and the world of
fields have deep interaction. As in the case of groups, we draw the
immediate (but powerful) consequence, reminiscent of Lagrange's theorem:
\begin{coro}{}{fields:1.11}
Let $k \subseteq F$ be a finite extension, and let $E$ be an intermediate
field (that is, $k \subseteq E \subseteq F$). Then both $[E : k]$ and $[F :
E]$ divide $[F : k]$.
\end{coro}
As with Lagrange's theorem, judicious use of this result will make otherwise
mysterious statements melt into trivialities.
\begin{exam}{}{fields:exam:1.12}
Let $k \subseteq F$ be a field extension, and let $\alpha \in F$ be an
algebraic element over $k$, of odd degree. Then I claim that $\alpha$ may be
written as a polynomial in $\alpha^2$, with coefficients in $k$.
Indeed, $k(\alpha^2)$ is intermediate between $k$ and $k(\alpha)$:
\[
  k \subseteq k(\alpha^2) \subseteq k(\alpha);
\]
what can we say about the degree $d$ of $k(\alpha)$ over $k(\alpha^2)$?
Since $\alpha$ satisfies the polynomial $t^2 - \alpha^2 \in k(\alpha^2)[t]$,
$d \leq 2$. On the other hand, $d$ divides $[k(\alpha) : k]$ by
\cref{coro:fields:1.11}, and $[k(\alpha) : k]$ is odd, so $d \neq 2$.
Therefore $d = 1$, proving $k(\alpha) = k(\alpha^2)$, and in particular
$\alpha \in k(\alpha^2)$, which is the claim.
\end{exam}
Here is something else that should evoke fond memories for the reader: back
in \S{}III.6.5 we had encountered an important distinction between
\emph{finite} algebras and algebras \emph{of finite type}. An algebra is
finite over the base ring if it is finitely generated as a module, that is,
if it admits an onto homomorphism (of modules) from a finitely generated free
module; a commutative algebra is of finite type if it admits an onto
homomorphism (of algebras) from a polynomial ring in finitely many
variables.
Something along the same lines is going to occur here. An extension $k
\subseteq F$ is finite if and only if $\dim_k F$ is finite, that is, if and
only if $F$ is a finite $k$-algebra. The other finiteness condition takes the
following form.
\begin{defn}{}{fields:1.13}
A field extension $k \subseteq F$ is \emph{finitely generated} if there exist
$\alpha_1, \dots, \alpha_n \in F$ such that
\[
  F = k(\alpha_1)(\alpha_2)\dots(\alpha_n).
\]
\end{defn}
\begin{rem}{}{fields:1.14}
This is not the same as saying that $F$ is generated by the $\alpha_i$'s as
a (finite-type) $k$-algebra: even in the case of simple extensions, if
$\alpha$ is transcendental, then the natural evaluation map $\epsilon : k[t]
\to k(\alpha)$ of \cref{prop:fields:1.3} is not onto. However, this is an
epimorphism of rings in the categorical sense, as the reader will check
(Exercise~1.17); thus, the `categorical spirit' behind the finite-type
condition is preserved in this context.
This subtlety raises an interesting question: what if a field extension $k
\subseteq F$ really is a finite-type $k$-algebra? This turns out to be an
important issue, and we will come back to it in \S{}2.2.
\end{rem}
We will use the notation
\[
  k(\alpha_1, \alpha_2, \dots, \alpha_n)
\]
for the field $k(\alpha_1)(\alpha_2)\dots(\alpha_n)$. For $k \subseteq F$
and $\alpha_1, \dots, \alpha_n \in F$, the elements of the field $F =
k(\alpha_1, \dots, \alpha_n)$ are all the elements of $F$ which may be
written as rational functions in the $\alpha_i$'s, with coefficients in $k$
(cf.\ Exercise~1.3). Put otherwise, $k(\alpha_1, \dots, \alpha_n)$ is the
smallest subfield of $F$ containing $k(\alpha_1), \dots, k(\alpha_n)$ (it is
the \emph{composite} of these subfields). It is clear that the order of the
elements $\alpha_i$ is irrelevant.
Coming back to the issue of finite vs.\ algebraic, these two notions do
coincide for finitely generated extensions.
\begin{prop}{}{fields:1.15}
Let $k \subseteq F = k(\alpha_1, \dots, \alpha_n)$ be a finitely generated
field extension. Then the following are equivalent:
\begin{enumerate}[label=(\roman*)]
  \item $k \subseteq F$ is a finite extension.
  \item $k \subseteq F$ is an algebraic extension.
  \item Each $\alpha_i$ is algebraic over $k$.
\end{enumerate}
If these conditions are satisfied, then $[F : k] \leq$ the product of the
degrees of $\alpha_i$ over $k$.
\end{prop}
\begin{proof}
\cref{lem:fields:1.9} shows that (i)$\implies$(ii); (ii)$\implies$(iii)
trivially. Thus, we only need to prove that (iii)$\implies$(i), and to bound
the degree of $F$ over $k$ in the process.
Assume that each $\alpha_i$ is algebraic over $k$, and let $d_i$ be the
degree of $\alpha_i$ over $k$. By definition, $k \subseteq k(\alpha_i)$ is
finite, of degree $d_i$. By Exercise~1.16, each extension $k(\alpha_1, \dots,
\alpha_{i-1}) \subseteq k(\alpha_1, \dots, \alpha_i)$ is finite, of degree
$\leq d_i$. Applying \cref{prop:fields:1.10} to the composition of
extensions
\[
  k \subseteq k(\alpha_1) \subseteq k(\alpha_1, \alpha_2) \subseteq \dots
  \subseteq k(\alpha_1, \dots, \alpha_n) = F
\]
proves that $k \subseteq F$ is finite and $[F : k] \leq d_1 \dots d_n$, as
needed.
\end{proof}
While rather straightforward, \cref{prop:fields:1.15} has always seemed
remarkable to us: it says (in particular) that if $\alpha$ and $\beta$ are
algebraic over a field $k$, then so are $\alpha \pm \beta$, $\alpha\beta$,
$\alpha\beta^{-1}$, and any other rational function of $\alpha$ and $\beta$.
If all we knew were the simple-minded definition of `algebraic' (that is,
`root of a polynomial'), this would seem rather mysterious: given a
polynomial $f(x)$ of which $\alpha$ is a root and a polynomial $g(x)$ of
which $\beta$ is a root, how do we construct a polynomial $h(x)$ such that
(for example) $h(\alpha + \beta) = 0$? The answer is that we do not need to
perform any such construction, by virtue of \cref{prop:fields:1.15}.
One immediate consequence of this observation is that the set of algebraic
elements of any extension forms a field:

\begin{coro}{}{fields:1.16}
Let $k \subseteq F$ be a field extension. Let
\[
  E = \{\alpha \in F \mid \alpha \text{ is algebraic over } k\}.
\]
Then $E$ is a field.
\end{coro}
\begin{exam}{}{fields:exam:1.17}
Let $\overline{\Q} \subseteq \C$ be the set of complex numbers that are
algebraic over $\Q$; then $\overline{\Q}$ is a field, by
\cref{coro:fields:1.16}, and the extension $\Q \subseteq \overline{\Q}$ is
(tautologically) algebraic. Note that $\Q \subseteq \overline{\Q}$ is not a
finite extension, because in it there are elements of arbitrarily high degree
over $\Q$: indeed, there exist irreducible polynomials in $\Q[x]$ of
arbitrarily high degree, as we observed in Corollary~V.5.16.
Elements of $\overline{\Q}$ are called \emph{algebraic numbers}; the complex
numbers in the complement of $\overline{\Q}$ are transcendental over $\Q$ (by
definition); they are simply called \emph{transcendental numbers}. Elementary
cardinality considerations show that $\overline{\Q}$ is countable; thus, the
set of transcendental numbers is uncountable. Surprisingly, it may be
substantially difficult to prove that a given number is transcendental: for
example, the fact that $e$ is transcendental was only proved around 1870, by
\name{Charles Hermite}. The number $\pi$ is transcendental (\name{Lindemann},
ca.\ 1880); $e^\pi$ is transcendental (\name{Gelfond}--\name{Schneider},
1934). No one knows whether $\pi^e$, $\pi + e$, or $\pi e$ are
transcendental.
\end{exam}
Another consequence of \cref{prop:fields:1.15} is the important fact that
compositions of algebraic extensions are algebraic (whether finitely
generated or not):
\begin{coro}{}{fields:1.18}
Let $k \subseteq E \subseteq F$ be field extensions. Then $k \subseteq F$
is algebraic if and only if both $k \subseteq E$ and $E \subseteq F$ are
algebraic.
\end{coro}
\begin{proof}
If $k \subseteq F$ is algebraic, then every element of $F$ is algebraic
over $k$, hence over $E$, and every element of $E$ is algebraic over $k$;
thus $E \subseteq F$ and $k \subseteq E$ are algebraic.

Conversely, assume $k \subseteq E$ and $E \subseteq F$ are both algebraic,
and let $\alpha \in F$. Since $\alpha$ is algebraic over $E$, there exists
a polynomial
\[
  f(x) = x^n + e_{n-1}x^{n-1} + \dots + e_0 \in E[x]
\]
such that $f(\alpha) = 0$. This implies that in fact $\alpha$ is `already'
algebraic over the subfield $k(e_0, \dots, e_{n-1}) \subseteq E$;
therefore,
\[
  k(e_0, \dots, e_{n-1}) \subseteq k(e_0, \dots, e_{n-1}, \alpha)
\]
is a finite extension. On the other hand,
\[
  k \subseteq k(e_0, \dots, e_{n-1})
\]
is a finite extension by \cref{prop:fields:1.15}, since each $e_i$ is in
$E$ and therefore algebraic over $k$. By \cref{prop:fields:1.10},
\[
  k \subseteq k(e_0, \dots, e_{n-1}, \alpha)
\]
is finite. This implies that $\alpha$ is algebraic over $k$, as needed,
by \cref{lem:fields:1.9}.
\end{proof}
We will essentially deal only with finitely generated extensions, and
\cref{prop:fields:1.15} will simplify our work considerably. Also, since
finitely generated extensions are compositions of simple extensions, the
reader should expect that we will give a careful look at automorphisms of
such extensions.

It may in fact come as a surprise that, in many cases, finitely generated
extensions turn out to be simple to begin with; we will prove a precise
statement to this effect in due time (Proposition~5.19). The reader already
has enough tools to contemplate easy (but interesting) examples, such as
the following. This should serve as an encouragement to look at many more.
\begin{exam}{}{fields:exam:1.19}
Consider the extension $\Q \subseteq \Q(\sqrt{2}, \sqrt{3})$.

By \cref{prop:fields:1.15} we know that this is a finite (hence algebraic)
extension, of degree at most $4$.

Thus any five elements in $\Q(\sqrt{2}, \sqrt{3})$ must be linearly
dependent over $\Q$. We consider powers of $\sqrt{2} + \sqrt{3}$:
\[
  1,\quad (\sqrt{2}+\sqrt{3}),\quad (\sqrt{2}+\sqrt{3})^2,\quad
  (\sqrt{2}+\sqrt{3})^3,\quad (\sqrt{2}+\sqrt{3})^4
\]
must be linearly dependent. Thus, there must be rational numbers
$q_0, \dots, q_3$ such that
\[
  (\sqrt{2}+\sqrt{3})^4 + q_3(\sqrt{2}+\sqrt{3})^3
  + q_2(\sqrt{2}+\sqrt{3})^2 + q_1(\sqrt{2}+\sqrt{3}) + q_0 = 0.
\]

Elementary arithmetic shows that this relation is satisfied for $q_0 = 1$,
$q_1 = 0$, $q_2 = -10$, $q_3 = 0$: that is, $\sqrt{2}+\sqrt{3}$ is a
root of the polynomial
\[
  f(t) = t^4 - 10t^2 + 1.
\]
In fact, it is easily checked that $f(t)$ vanishes at all four combinations
$\pm\sqrt{2} \pm \sqrt{3}$.

It follows that
\[
  f(t) = (t-(-\sqrt{2}-\sqrt{3}))(t-(-\sqrt{2}+\sqrt{3}))
         (t-(\sqrt{2}-\sqrt{3}))(t-(\sqrt{2}+\sqrt{3}))
\]
and that $f(t)$ is irreducible over $\Q$ (no proper factor of $f(t)$ has
rational coefficients). Thus $\sqrt{2}+\sqrt{3}$ has degree $4$ over $\Q$.

Consider the composition of extensions
\[
  \Q \subseteq \Q(\sqrt{2}+\sqrt{3}) \subseteq \Q(\sqrt{2},\sqrt{3}).
\]
By \cref{coro:fields:1.11},
\[
  4 = [\Q(\sqrt{2}+\sqrt{3}) : \Q]
    \leq [\Q(\sqrt{2},\sqrt{3}) : \Q].
\]
On the other hand we know that $[\Q(\sqrt{2},\sqrt{3}) : \Q] \leq 4$;
thus we can conclude that this degree is exactly $4$, and it follows that
\[
  \Q(\sqrt{2},\sqrt{3}) = \Q(\sqrt{2}+\sqrt{3}):
\]
the extension was simple to begin with. (In due time we will prove that
this is a typical example, not a contrived one.)

Staring now at the composition
\[
  \Q \subseteq \Q(\sqrt{2}) \subseteq \Q(\sqrt{2},\sqrt{3}),
\]
\cref{prop:fields:1.10} tells us that
$[\Q(\sqrt{2},\sqrt{3}) : \Q(\sqrt{2})] = 2$.
That is, a side-effect of the computation carried out above is that the
polynomial $t^2 - 3$ must be irreducible over $\Q(\sqrt{2})$; this is
not surprising, but note that this method is very different from anything
we encountered back in \S{}V.5.

The fact that the other roots of the minimal polynomial of
$\sqrt{2}+\sqrt{3}$ `looked so much like' $\sqrt{2}+\sqrt{3}$ is not
surprising. Indeed, since $\Q(\sqrt{2}) \subseteq \Q(\sqrt{2},\sqrt{3})$
is a simple extension, we know (cf.\ \cref{prop:fields:1.5} and following)
that there is an automorphism of $\Q(\sqrt{2},\sqrt{3})$ fixing
$\Q(\sqrt{2})$ (and hence $\Q$) and swapping $\sqrt{3}$ and $-\sqrt{3}$.
Similarly, there is an automorphism fixing $\Q(\sqrt{3})$ and swapping
$\sqrt{2}$ and $-\sqrt{2}$. These automorphisms must act on the set of
roots of $f(t)$ (Exercise~1.6): applying them and their composition to
$\sqrt{2}+\sqrt{3}$ produces the other three roots of $f(t)$.

In fact, at this point we know that
$G = \operatorname{Aut}_{\Q}(\Q(\sqrt{2},\sqrt{3}))$ must consist of $4$
elements (by \cref{coro:fields:1.7}) and has at least two elements of
order $2$ (both automorphisms found above have order $2$). This is enough
to conclude that $G$ is not a cyclic group, and hence it must be isomorphic
to the group $\Z/2\Z \times \Z/2\Z$.
\end{exam}

\subsection*{Exercises}

\begin{exer}{}{fields:exer:1.1}
Prove that if $k \subseteq K$ is a field extension, then
$\operatorname{char} k = \operatorname{char} K$. Prove that the category
$\mathbf{Fld}$ has no initial object.
\end{exer}

\begin{exer}{}{fields:exer:1.2}
Define carefully the category $\mathbf{Fld}_k$ of extensions of $k$.
\end{exer}

\begin{exer}{}{fields:exer:1.3}
Let $k \subseteq F$ be a field extension, and let $\alpha \in F$. Prove
that the field $k(\alpha)$ consists of all the elements of $F$ which may
be written as rational functions in $\alpha$, with coefficients in $k$.
Why does this not give (in general) an onto homomorphism $k(t) \to
k(\alpha)$?
\end{exer}

\begin{exer}{}{fields:exer:1.4}
Let $k \subseteq k(\alpha)$ be a simple extension, with $\alpha$
transcendental over $k$. Let $E$ be a subfield of $k(\alpha)$ properly
containing $k$. Prove that $k(\alpha)$ is a finite extension of $E$.
\end{exer}

\begin{exer}{}{fields:exer:1.5}
(Cf.\ \cref{exam:fields:exam:1.4}.)
\begin{itemize}
\item Prove that there is exactly one subfield of $\R$ isomorphic to
  $\Q[t]/(t^2 - 2)$.
\item Prove that there are exactly three subfields of $\C$ isomorphic to
  $\Q[t]/(t^3 - 2)$.
\end{itemize}
From a `topological' point of view, one of these copies of $\Q[t]/(t^3 -
2)$ looks very different from the other two: it is not dense in $\C$,
but the others are.
\end{exer}

\begin{exer}{}{fields:exer:1.6}
Let $k \subseteq F$ be a field extension, and let $f(x) \in k[x]$ be a
polynomial. Prove that $\operatorname{Aut}_k(F)$ acts on the set of
roots of $f(x)$ contained in $F$. Provide examples showing that this
action need not be transitive or faithful.
\end{exer}

\begin{exer}{}{fields:exer:1.7}
Let $k \subseteq F$ be a field extension, and let $\alpha \in F$ be
algebraic over $k$.
\begin{itemize}
\item Suppose $p(x) \in k[x]$ is an irreducible monic polynomial such
  that $p(\alpha) = 0$; prove that $p(x)$ is the minimal polynomial of
  $\alpha$ over $k$, in the sense of \cref{prop:fields:1.3}.
\item Let $f(x) \in k[x]$. Prove that $f(\alpha) = 0$ if and only if
  $p(x) \mid f(x)$.
\item Show that the minimal polynomial of $\alpha$ is the minimal
  polynomial of a certain $k$-linear transformation of $F$, in the
  sense of Definition~VI.6.12.
\end{itemize}
\end{exer}

\begin{exer}{}{fields:exer:1.8}
Let $f(x) \in k[x]$ be a polynomial over a field $k$ of degree $d$, and
let $\alpha_1, \dots, \alpha_d$ be the roots of $f(x)$ in an extension
of $k$ where the polynomial factors completely. For a subset $I \subseteq
\{1, \dots, d\}$, denote by $\alpha_I$ the sum $\sum_{i \in I} \alpha_i$.
Assume that $\alpha_I \in k$ only for $I = \emptyset$ and $I =
\{1, \dots, d\}$. Prove that $f(x)$ is irreducible over $k$.
\end{exer}

\begin{exer}{}{fields:exer:1.9}
Let $k$ be a finite field. Prove that the order $|k|$ is a power of a
prime integer.
\end{exer}

\begin{exer}{}{fields:exer:1.10}
Let $k$ be a field. Prove that the ring of square $n \times n$ matrices
$M_n(k)$ contains an isomorphic copy of every extension of $k$ of degree
$\leq n$. (Hint: If $k \subseteq F$ is an extension of degree $n$ and
$\alpha \in F$, then `multiplication by $\alpha$' is a $k$-linear
transformation of $F$.)
\end{exer}
\begin{exer}{}{fields:exer:1.11}
Let $k \subseteq F$ be a finite field extension, and let $p(x)$ be the
characteristic polynomial of the $k$-linear transformation of $F$ given
by multiplication by $\alpha$. Prove that $p(\alpha) = 0$.

This gives an effective way to find a polynomial satisfied by an element
of an extension. Use it to find a polynomial satisfied by $\sqrt{2} +
\sqrt{3}$ over $\Q$, and compare this method with the one used in
\cref{exam:fields:exam:1.19}.
\end{exer}

\begin{exer}{}{fields:exer:1.12}
Let $k \subseteq F$ be a finite field extension, and let $\alpha \in F$.
The \emph{norm} of $\alpha$, $N_{k \subseteq F}(\alpha)$, is the
determinant of the linear transformation of $F$ given by multiplication
by $\alpha$ (cf.\ Exercise~1.11, Definition~VI.6.4). Prove that the norm
is multiplicative: for $\alpha, \beta \in F$,
\[
  N_{k \subseteq F}(\alpha\beta)
  = N_{k \subseteq F}(\alpha)\, N_{k \subseteq F}(\beta).
\]
Compute the norm of a complex number viewed as an element of the
extension $\R \subseteq \C$ (and marvel at the excellent choice of
terminology). Do the same for elements of an extension $\Q(\sqrt{d})$ of
$\Q$, where $d$ is an integer that is not a square, and compare the
result with Exercise~III.4.10.
\end{exer}

\begin{exer}{}{fields:exer:1.13}
Define the \emph{trace} $\operatorname{tr}_{k \subseteq F}(\alpha)$ of an
element $\alpha$ of a finite extension $F$ of a field $k$ by following
the lead of Exercise~1.12. Prove that the trace is additive:
\[
  \operatorname{tr}_{k \subseteq F}(\alpha + \beta)
  = \operatorname{tr}_{k \subseteq F}(\alpha)
  + \operatorname{tr}_{k \subseteq F}(\beta)
\]
for $\alpha, \beta \in F$. Compute the trace of an element of an
extension $\Q \subseteq \Q(\sqrt{d})$, for $d$ an integer that is not
a square.
\end{exer}

\begin{exer}{}{fields:exer:1.14}
Let $k \subseteq k(\alpha)$ be a simple algebraic extension, and let
$x^d + a_{d-1}x^{d-1} + \dots + a_0$ be the minimal polynomial of
$\alpha$ over $k$. Prove that
\[
  \operatorname{tr}_{k \subseteq k(\alpha)}(\alpha) = -a_{d-1}
  \quad\text{and}\quad
  N_{k \subseteq k(\alpha)}(\alpha) = (-1)^d a_0.
\]
(Cf.\ Exercises~1.12 and~1.13.)
\end{exer}

\begin{exer}{}{fields:exer:1.15}
Let $k \subseteq F$ be a finite extension, and let $\alpha \in F$. Assume
$[F : k(\alpha)] = r$. Prove that
\[
  \operatorname{tr}_{k \subseteq F}(\alpha)
  = r\,\operatorname{tr}_{k \subseteq k(\alpha)}(\alpha)
  \quad\text{and}\quad
  N_{k \subseteq F}(\alpha)
  = N_{k \subseteq k(\alpha)}(\alpha)^r.
\]
(Cf.\ Exercises~1.12 and~1.13.) (Hint: If $f_1, \dots, f_r$ is a basis
of $F$ over $k(\alpha)$ and $\alpha$ has degree $d$ over $k$, then
$(f_i \alpha^j)_{\substack{i = 1, \dots, r \\ j = 0, \dots, d-1}}$ is
a basis of $F$ over $k$. The matrix corresponding to multiplication by
$\alpha$ with respect to this basis consists of $r$ identical square
blocks.)
\end{exer}

\begin{exer}{}{fields:exer:1.16}
Let $k \subseteq L \subseteq F$ be fields, and let $\alpha \in F$. If
$k \subseteq k(\alpha)$ is a finite extension, then $L \subseteq
L(\alpha)$ is finite and $[L(\alpha) : L] \leq [k(\alpha) : k]$.
\end{exer}

\begin{exer}{}{fields:exer:1.17}
Let $k \subseteq F = k(\alpha_1, \dots, \alpha_n)$ be a finitely
generated extension. Prove that the evaluation map
\[
  k[t_1, \dots, t_n] \to F, \quad t_i \mapsto \alpha_i
\]
is an epimorphism of rings (although it need not be onto).
\end{exer}

\begin{exer}{}{fields:exer:1.18}
Let $R$ be a ring sandwiched between a field $k$ and an algebraic
extension $F$ of $k$. Prove that $R$ is a field.

Is it necessary to assume that the extension is algebraic?
\end{exer}

\begin{exer}{}{fields:exer:1.19}
Let $k \subseteq F$ be a field extension of degree $p$, a prime integer.
Prove that there are no subrings of $F$ properly containing $k$ and
properly contained in $F$. (Use Exercise~1.18.)
\end{exer}

\begin{exer}{}{fields:exer:1.20}
Let $p$ be a prime integer, and let $\alpha = \sqrt[p]{2} \in \R$. Let
$g(x) \in \Q[x]$ be any nonconstant polynomial of degree $< p$. Prove
that $\alpha$ may be expressed as a polynomial in $g(\alpha)$ with
rational coefficients.

Prove that an analogous statement for $\sqrt[4]{2}$ is false.
\end{exer}
\begin{exer}{}{fields:exer:1.21}
Let $k \subseteq F$ be a field extension, and let $E$ be the intermediate
field consisting of the elements of $F$ which are algebraic over $k$. For
$\alpha \in F$, prove that $\alpha$ is algebraic over $E$ if and only if
$\alpha \in E$. Deduce that $\overline{\Q}$ is algebraically closed.
\end{exer}

\begin{exer}{}{fields:exer:1.22}
Let $k \subseteq F$ be a field extension, and let $\alpha, \beta \in F$
be algebraic, of degree $d$, $e$, respectively. Assume $d$, $e$ are
relatively prime, and let $p(x)$ be the minimal polynomial of $\beta$
over $k$. Prove $p(x)$ is irreducible over $k(\alpha)$.
\end{exer}

\begin{exer}{}{fields:exer:1.23}
Express $\sqrt{2}$ explicitly as a polynomial function in $\sqrt{2} +
\sqrt{3}$ with rational coefficients.
\end{exer}

\begin{exer}{}{fields:exer:1.24}
Generalize the situation examined in \cref{exam:fields:exam:1.19}: let
$k$ be a field of characteristic $\neq 2$, and let $a, b \in k$ be
elements that are not squares in $k$; prove that $k(\sqrt{a}, \sqrt{b})
= k(\sqrt{a} + \sqrt{b})$.

Prove that $k(\sqrt{a}, \sqrt{b})$ has degree $2$, resp., $4$, over $k$
according to whether $ab$ is, resp., is not, a square in $k$.
\end{exer}

\begin{exer}{}{fields:exer:1.25}
Let $\xi := \sqrt{2 + \sqrt{2}}$.
\begin{itemize}
\item Find the minimal polynomial of $\xi$ over $\Q$, and show that
  $\Q(\xi)$ has degree $4$ over $\Q$.
\item Prove that $\sqrt{2 - \sqrt{2}}$ is another root of the minimal
  polynomial of $\xi$.
\item Prove that $\sqrt{2 - \sqrt{2}} \in \Q(\xi)$. (Hint:
  $(a+b)(a-b) = a^2 - b^2$.)
\item By \cref{prop:fields:1.5}, sending $\xi$ to $\sqrt{2 - \sqrt{2}}$
  defines an automorphism of $\Q(\xi)$ over $\Q$. Find the matrix of
  this automorphism w.r.t.\ the basis $1, \xi, \xi^2, \xi^3$.
\item Prove that $\operatorname{Aut}_{\Q}(\Q(\xi))$ is cyclic of order
  $4$.
\end{itemize}
\end{exer}

\begin{exer}{}{fields:exer:1.26}
Let $k \subseteq F$ be a field extension, let $I$ be an indexing set,
and let $\{\alpha_i\}_{i \in I}$ be a choice of elements of $F$. This
choice determines a homomorphism $\phi$ of $k$-algebras from the
polynomial ring $k[I]$ on the set $I$ to $F$ (the polynomial ring is a
free commutative $k$-algebra; cf.\ Proposition~III.6.4). We say that
$\{\alpha_i\}_{i \in I}$ is \emph{algebraically independent} over $k$
if $\phi$ is injective. For example, distinct elements
$\alpha_1, \dots, \alpha_n$ of $F$ are algebraically independent over
$k$ if there is no nonzero polynomial $f(x_1, \dots, x_n) \in
k[x_1, \dots, x_n]$ such that $f(\alpha_1, \dots, \alpha_n) = 0$.

Prove that $\alpha_1, \dots, \alpha_n$ are algebraically independent if
and only if the assignment $t_1 \mapsto \alpha_1, \dots, t_n \mapsto
\alpha_n$ defines a homomorphism of $k$-algebras (and hence an
isomorphism) from the field of rational functions $k(t_1, \dots, t_n)$
to $k(\alpha_1, \dots, \alpha_n)$.
\end{exer}

\begin{exer}{}{fields:exer:1.27}
With notation and terminology as in Exercise~1.26, the indexed set
$\{\alpha_i\}_{i \in I}$ is a \emph{transcendence basis} for $F$ over
$k$ if it is a maximal algebraically independent set in $F$.
\begin{itemize}
\item Prove that $\{\alpha_i\}_{i \in I}$ is a transcendence basis for
  $F$ over $k$ if and only if it is algebraically independent and $F$
  is algebraic over $k(\{\alpha_i\}_{i \in I})$.
\item Prove that transcendence bases exist. (Zorn)
\item Prove that any two transcendence bases for $F$ over $k$ have the
  same cardinality. (Mimic the proof of Proposition~VI.1.9. Don't feel
  too bad if you prefer to deal only with the case of finite
  transcendence bases.)
\end{itemize}
The cardinality of a transcendence basis is called the \emph{transcendence
degree} of $F$ over $k$, denoted $\operatorname{tr.deg}_{k \subseteq F}$.
\end{exer}

\begin{exer}{}{fields:exer:1.28}
Let $k \subseteq E \subseteq F$ be field extensions. Prove that
$\operatorname{tr.deg}_{k \subseteq F}$ is finite if and only if both
$\operatorname{tr.deg}_{k \subseteq E}$ and
$\operatorname{tr.deg}_{E \subseteq F}$ (see Exercise~1.27) are finite,
and in this case
\[
  \operatorname{tr.deg}_{k \subseteq F}
  = \operatorname{tr.deg}_{k \subseteq E}
  + \operatorname{tr.deg}_{E \subseteq F}.
\]
\end{exer}

\begin{exer}{}{fields:exer:1.29}
An extension $k \subseteq F$ is \emph{purely transcendental} if it
admits a transcendence basis $\{\alpha_i\}_{i \in I}$ (see
Exercise~1.27) such that $F = k(\{\alpha_i\}_{i \in I})$.

Prove that any field extension $k \subseteq F$ may be decomposed as a
purely transcendental extension followed by an algebraic extension. (Not
all field extensions may be decomposed as an algebraic extension followed
by a purely transcendental extension.)
\end{exer}

\begin{exer}{}{fields:exer:1.30}
Let $k \subseteq k(\alpha)$ be a simple extension, with $\alpha$
transcendental over $k$. Let $E$ be a subfield of $k(\alpha)$ properly
containing $k$. Prove that $\operatorname{tr.deg}_{k \subseteq E} = 1$.

L\"uroth's theorem asserts that in this situation $k \subseteq E$ is
itself a simple transcendental extension of $k$; that is, it is purely
transcendental (Exercise~1.29).
\end{exer}

\section{Algebraic closure, Nullstellensatz, and a little algebraic
geometry}
\label{sec:fields:algebraic-closure-nullstellensatz-and-a-little-algebraic}

One of the most important extensions of a field $k$ is its algebraic
closure $k \subseteq \overline{k}$; this was mentioned already in
\S{}V.5.2, and we can now prove that $\overline{k}$ exists and is unique
(up to isomorphism). Once this circle of ideas is approached, the
temptation to say a few words about the important result known as
Hilbert's Nullstellensatz, at the cost of a small digression into
algebraic geometry, will simply be irresistible.

\subsection{Algebraic closure}
\label{subsec:fields:algebraic-closure}
Recall (Definition~V.5.9) that a field $K$ is algebraically closed if all
irreducible polynomials in $K[x]$ have degree $1$, that is, if every
polynomial in $K[x]$ factors completely as a product of linear terms.
Equivalently (Exercise~III.4.21) every maximal ideal in $K[x]$ is of the
form $(x - c)$, for $c \in K$. Now that we have a little more vocabulary,
we can recast this definition yet again, as follows.
\begin{lem}{}{fields:2.1}
For a field $K$, the following are equivalent:
\begin{enumerate}[label=(\roman*)]
  \item $K$ is algebraically closed.
  \item $K$ has no nontrivial algebraic extensions.
  \item If $K \subseteq L$ is any extension and $\alpha \in L$ is
    algebraic over $K$, then $\alpha \in K$.
\end{enumerate}
\end{lem}
The proof is a straightforward application of the definitions and a good
exercise (Exercise~2.1).
\begin{defn}{}{fields:2.2}
An \emph{algebraic closure} of a field $k$ is an algebraic extension
$k \subseteq \overline{k}$ such that $\overline{k}$ is algebraically
closed.
\end{defn}
Of course, every polynomial $f(x) \in k[x]$ will split into linear factors
over $\overline{k}$: so does every polynomial in the much larger
$\overline{k}[x]$, as $\overline{k}$ is algebraically closed. The
requirement that $k \subseteq \overline{k}$ be algebraic ensures that no
other intermediate field $L$,
\[
  k \subseteq L \subsetneq \overline{k},
\]
may be algebraically closed: indeed, $L \subseteq \overline{k}$ would be a
nontrivial algebraic extension, contradicting \cref{lem:fields:2.1}. In
fact, the only subfield of $\overline{k}$ containing all roots of all
nonconstant polynomials in $k[x]$ is $\overline{k}$ itself
(Exercise~2.2); thus, the algebraic closure of a field is `as small as
possible' subject to this requirement. Not surprisingly, $\overline{k}$
turns out to be unique up to isomorphism, so that we can speak of
\emph{the} algebraic closure of $k$.
\begin{thm}{}{fields:2.3}
Every field $k$ admits an algebraic closure $k \subseteq \overline{k}$;
this extension is unique up to isomorphism.
\end{thm}
Concerning existence, the idea is to construct `by hand' a huge extension
$K$ of $k$ where every polynomial $f(x) \in k[x]$ factors completely. The
elements of $K$ which are algebraic over $k$ will form an algebraic
closure of $k$.

The construction is done in steps, each step including `one more root' of
all nonconstant polynomials in $k[x]$. Here is a formalization of this
step.
\begin{lem}{}{fields:2.4}
Let $k$ be a field. Then there exists an extension $k \subseteq K$ such
that every nonconstant polynomial $f(x) \in k[x]$ has at least one root
in $K$.
\end{lem}
\begin{proof}
(This construction is apparently due to \name{Emil Artin}.) Consider a
set $T = \{t_f\}$ in bijection with the set of nonconstant monic
polynomials $f(x) \in k[x]$, and let $k[T]$ be the corresponding
polynomial ring\footnote{Here is another rare occasion where we need to
consider polynomial rings with possibly infinitely many indeterminates. An
element of $k[T]$ is simply an ordinary polynomial with coefficients in
$k$, involving a finite number of indeterminates from $T$.} in all the
indeterminates $t_f$. Let $I \subseteq k[T]$ be the ideal generated by
all polynomials $f(t_f)$.

Then $I$ is a proper ideal. Indeed, otherwise we could write
\begin{equation}
  \label{eq:fields:2.4-star}
  1 = \sum_{i=1}^{n} a_i \cdot f_i(t_{f_i}),
\end{equation}
where $a_i \in k[T]$. I claim that this cannot be done: indeed, we can
construct an extension $k \subseteq F$ where the polynomials
$f_1(x), \dots, f_n(x)$ have roots $\alpha_1, \dots, \alpha_n$,
respectively (apply Proposition~V.5.7 $n$ times); view
\eqref{eq:fields:2.4-star} as an identity in $F[T]$, and plug in
$t_{f_i} = \alpha_i$, obtaining
\[
  1 = \sum_{i=1}^{n} a_i \cdot f_i(\alpha_i)
    = \sum_{i=1}^{n} a_i \cdot 0 = 0,
\]
which is nonsense.

Since $I$ is proper, it is contained in a maximal ideal $\mathfrak{m}$
(Proposition~V.3.5). Thus, we obtain a field extension
\[
  k \subseteq K := \frac{k[T]}{\mathfrak{m}};
\]
by construction every nonconstant monic (and hence every nonconstant)
polynomial $f(x)$ has a root in $K$, namely the coset of $t_f$.
\end{proof}
The field constructed in \cref{lem:fields:2.4} contains at least one root
of each nonconstant polynomial $f(x) \in k[x]$, but this is not good
enough: we are seeking a field which contains all roots of such
polynomials. Equivalently, not only should $f(x)$ have (at least) one
linear factor $x - \alpha$ in $K[x]$, but we need the quotient polynomial
$f(x)/(x - \alpha) \in K[x]$ to have a linear factor and the quotient by
that new factor to have one, etc. This prompts us to consider a whole
chain of extensions
\[
  k \subseteq K_1 \subseteq K_2 \subseteq K_3 \subseteq \dots
\]
where $K_1$ is obtained from $k$ by applying \cref{lem:fields:2.4}, $K_2$
is similarly obtained from $K_1$, etc.

Now consider the union $L$ of this chain\footnote{This is an example of
direct limit, a notion which we will introduce more formally in
\S{}VIII.1.4.}. For every two $a, b \in L$, there is an $i$ such that
$a, b \in K_i$; we can define $a + b$, $a \cdot b$ in $L$ by adopting
their definition in $K_i$, and the result does not depend on the choice
of $i$. It follows easily that $L$ is a field.
\begin{fact}{}{fields:2.5}
The field $L$ is algebraically closed.
\end{fact}
\begin{proof}
If $f(x) \in L[x]$ is a nonconstant polynomial, then $f(x) \in K_i[x]$
for some $i$; hence $f(x)$ has a root in $K_{i+1} \subseteq L$. That is,
every nonconstant polynomial in $L[x]$ has a root in $L$, as needed.
\end{proof}
\paragraph{Proof of the existence of algebraic closures.}
The existence of algebraic closures is now completely transparent, because
of the following simple observation:
\begin{lem}{}{fields:2.6}
Let $k \subseteq L$ be a field extension, with $L$ algebraically closed.
Let
\[
  \overline{k} := \{\alpha \in L \mid \alpha \text{ is algebraic over }
  k\}.
\]
Then $\overline{k}$ is an algebraic closure of $k$.
\end{lem}
The construction reviewed above provides us with an algebraically closed
field $L$ containing any given field $k$, so the lemma is all we need to
prove.

By \cref{coro:fields:1.16}, $\overline{k}$ is a field, and the extension
$k \subseteq \overline{k}$ is tautologically algebraic. To verify that
$\overline{k}$ is algebraically closed, let $k \subseteq \overline{k}(\alpha)$
be a simple algebraic extension. The minimal polynomial of $\alpha$ has a
root in $L$ since $L$ is algebraically closed, so by versality
(Proposition~V.5.7) there exists an embedding $\overline{k}(\alpha)
\subseteq L$. We can then view $\alpha$ as an element of $L$;
$k \subseteq \overline{k} \subseteq \overline{k}(\alpha)$ is a
composition of algebraic extensions, so $k \subseteq \overline{k}(\alpha)$
is algebraic (\cref{coro:fields:1.18}), and in particular $\alpha$ is
algebraic over $k$. But then $\alpha \in \overline{k}$, by definition of
the latter. It follows that $\overline{k}$ is algebraically closed, by
\cref{lem:fields:2.1}. \qed

\begin{rem}{}{fields:2.7}
Zorn's lemma was sneakily used in this proof (we used the existence of
maximal ideals, which relies upon Zorn's lemma), and it is natural to
wonder whether the existence of algebraic closures may be another
statement equivalent to the axiom of choice. Apparently, this is not the
case.\footnote{Allegedly, the existence of algebraic closures is a
consequence of the compactness theorem for first-order logic, which is
known to be weaker than the axiom of choice.}
\end{rem}
Next, we deal with uniqueness. It may be tempting to try to set things up
in such a way that algebraic closures would end up being solutions to a
universal problem; this would guarantee uniqueness by abstract nonsense.
However, we run into the same obstacle encountered in Proposition~V.5.7:
morphisms between extensions depend on choices of roots of polynomials,
so that algebraic closures are not `universal' in the most straightforward
sense of the term. Therefore, a bit of independent work is needed.
\begin{lem}{}{fields:2.8}
Let $k \subseteq L$ be a field extension, with $L$ algebraically closed.
Let $k \subseteq F$ be any algebraic extension. Then there exists a
morphism of extensions $i : F \to L$.
\end{lem}
As pointed out above, $i$ is by no means unique!
\begin{proof}
This argument also relies on Zorn's lemma. Consider the set $\mathcal{Z}$
of homomorphisms
\[
  i_K : K \to L
\]
where $K$ is an intermediate field, $k \subseteq K \subseteq F$, and
$i_K$ restricts to the identity on $k$; $\mathcal{Z}$ is nonempty, since
the extension $i_k : k \hookrightarrow L$ defines an element of
$\mathcal{Z}$. We give a poset structure to $\mathcal{Z}$ by defining
$i_K \leq i_{K'}$ if $K \subseteq K' \subseteq F$ and $i_{K'}$ restricts
to $i_K$ on $K$. To verify that every chain $\mathcal{C}$ in $\mathcal{Z}$
has an upper bound in $\mathcal{Z}$, let $K_{\mathcal{C}}$ be the union
of the sources of all $i_K \in \mathcal{C}$ ($K_{\mathcal{C}}$ is clearly
a field); if $\alpha \in K_{\mathcal{C}}$, define $i_{K_{\mathcal{C}}}(\alpha)$
to be $i_K(\alpha)$, where $i_K$ is any element of $\mathcal{C}$ such
that $\alpha \in K$. This prescription is clearly independent of the
chosen $K$ and defines a homomorphism $K_{\mathcal{C}} \to L$ restricting
to the identity on $k$. This homomorphism is an upper bound for
$\mathcal{C}$.

By Zorn's lemma, $\mathcal{Z}$ admits a maximal element $i_G$,
corresponding to an intermediate field $k \subseteq G \subseteq F$. Let
$H = i_G(G)$ be the image of $G$ in $L$.

I claim that $G = F$: this will prove the statement, because it will imply
that there is a homomorphism $i_F : F \to L$ extending the identity on $k$.

Arguing by contradiction, assume that there exists an $\alpha \in F
\setminus G$, and consider the extension $G \subseteq G(\alpha)$. Since
$\alpha \in F$ is algebraic over $k$, it is algebraic over $G$; thus, it
is a root of an irreducible polynomial $g(x) \in G[x]$. Consider the
induced homomorphism
\[
  i_G : G[x] \to H[x],
\]
and let $h(x) = i_G(g(x))$. Then $h(x)$ is an irreducible polynomial
over $H$, and it has a root $\beta$ in $L$ (this is where we use the
hypothesis that $L$ is algebraically closed!). We are in the situation
considered in \cref{prop:fields:1.5}: we have an isomorphism of fields
$i_G : G \to H$; we have simple extensions $G(\alpha)$, $H(\beta)$; and
$\alpha$, $\beta$ are roots of irreducible polynomials $g(x)$,
$h(x) = i_G(g(x))$. By \cref{prop:fields:1.5}, $i_G$ lifts to an
isomorphism
\[
  i_{G(\alpha)} : G(\alpha) \to H(\beta) \subseteq L
\]
sending $\alpha$ to $\beta$. This contradicts the maximality of $i_G$;
hence $G = F$, concluding the argument.
\end{proof}
\paragraph{Proof of uniqueness of algebraic closures.}
Let $k \subseteq \overline{k}$, $k \subseteq \overline{k}_1$ be two
algebraic closures of $k$; we have to prove that there exists an
isomorphism $\overline{k}_1 \to \overline{k}$ extending the identity on
$k$.

Since $k \subseteq \overline{k}_1$ is algebraic and $\overline{k}$ is
algebraically closed, by \cref{lem:fields:2.8} there exists a
homomorphism $i : \overline{k}_1 \to \overline{k}$ extending the identity
on $k$; this homomorphism is trivially injective, since $\overline{k}_1$
is a field. It is also surjective, by \cref{lem:fields:2.1}: otherwise
$\overline{k}_1 \subseteq \overline{k}$ is a nontrivial algebraic
extension, contradicting the fact that $\overline{k}_1$ is algebraically
closed. Therefore $i$ is an isomorphism, as needed. \qed

\subsection{The Nullstellensatz}
\label{subsec:fields:the-nullstellensatz}
If $K$ is an algebraically closed field, then every maximal ideal in
$K[x]$ is of the form $(x - c)$, for $c \in K$ (Exercise~III.4.21).
This statement has a straightforward-looking generalization to polynomial
rings in more indeterminates: if $K$ is algebraically closed, then every
maximal ideal in $K[x_1, \dots, x_n]$ is of the form
$(x_1 - c_1, \dots, x_n - c_n)$.

Proving this statement is more challenging than it may seem from the looks
of it; it is one facet of the famous theorem known as \name{Hilbert}'s
Nullstellensatz (``theorem on the position of zeros''). The Nullstellensatz
has made a few cameo appearances earlier (see Example~III.4.15 and
\S{}III.6.5 in particular), and we will spend a little time on it now.
We will not prove the Nullstellensatz in its natural generality, that is,
for all fields; this would take us too far. Actually, there are reasonably
short proofs of the theorem in this generality, but the short arguments I
have run into end up replacing a wider (and useful) context with ingenious
cleverness; I do not find these arguments particularly insightful or
memorable.

By contrast, the theorem has a very simple and memorable proof if we make
the further assumption that the field is uncountable. Therefore, the
reader will have a complete proof of the theorem (in the form given below,
which does not assume the field to be algebraically closed) for fields
such as $\R$ and $\C$ but will have to trust on faith regarding other
extremely important fields such as $\Q$.

The most vivid applications of the Nullstellensatz involve basic
definitions in algebraic geometry, and we will get a small taste of this
in the next section; but the theorem itself is best understood in the
context of the considerations on `finiteness' mentioned in \S{}1.3; see
especially \cref{rem:fields:1.14}. I pointed out that if $k \subseteq F$
is finitely generated as a field extension, then $F$ need not be finitely
generated (that is, `finite-type') as a $k$-algebra, and I raised the
question of understanding extensions $F$ that are finite-type
$k$-algebras. The following result answers this question. It is my
favorite version of the Nullstellensatz; therefore I will name it so:
\begin{thm}{Nullstellensatz}{fields:2.9}
Let $k \subseteq F$ be a field extension, and assume that $F$ is a
finite-type $k$-algebra. Then $k \subseteq F$ is a finite (hence
algebraic) extension.
\end{thm}
The reader should pause and savor this statement for a moment. As pointed
out in \S{}III.6.5, an algebra $k \to S$ may very well be of finite type
(as an algebra), without being finite (i.e., finitely generated as a
module): $S = k[t]$ is an obvious example. The content of
\cref{thm:fields:2.9} is that the distinction between finite-type and
finite disappears if $S$ is a field.

As mentioned above, we will only prove this statement under an additional
(and somewhat unnatural) assumption, which reduces the argument to
elementary linear algebra.
\begin{proof}[Proof for uncountable fields]
Assume that $k$ is uncountable. Let $k \subseteq F$ be a field extension,
and assume that $F$ is finitely generated as an algebra over $k$; in
particular, it is finitely generated as a field extension. We have to
prove that $k \subseteq F$ is a finite extension, or equivalently (by
\cref{prop:fields:1.15}) that it is algebraic, that is, that every
$\alpha \in F \setminus k$ is the root of a nonzero $f(x) \in k[x]$.

Now, by hypothesis $F$ is the quotient of a polynomial ring over $k$:
\[
  k[x_1, \dots, x_n] \twoheadrightarrow F;
\]
this implies that there is a countable basis of $F$ as a vector space
over $k$, because this is the case for $k[x_1, \dots, x_n]$. Consider
then the set
\[
  \left\{ \frac{1}{\alpha - c} \right\}_{c \in k};
\]
this is an uncountable subset of $F$; therefore it is linearly dependent
over $k$ (Proposition~VI.1.9). That is, there exist distinct
$c_1, \dots, c_m \in k$ and nonzero coefficients
$\lambda_1, \dots, \lambda_m \in k$ such that
\[
  \frac{\lambda_1}{\alpha - c_1} + \cdots + \frac{\lambda_m}{\alpha - c_m}
  = 0.
\]
Expressing the left-hand side with a common denominator gives
\[
  \frac{f(\alpha)}{g(\alpha)} = 0,
\]
for $f(x) \neq 0$ in $k[x]$ and $g(\alpha) \neq 0$ (Exercise~2.4).
This implies $f(\alpha) = 0$ for a nonzero $f(x) \in k[x]$, and we are
done.
\end{proof}
Regardless of its proof, the form of the Nullstellensatz given in
\cref{thm:fields:2.9} does not look much like the other results I have
mentioned as associated with it. For one thing, it is not too clear why
\cref{thm:fields:2.9} should be called a theorem on the `position of
zeros'. The result deserves a little more attention.

The form stated at the beginning of this section is obtained as follows:
\begin{coro}{}{fields:2.10}
Let $K$ be an algebraically closed field, and let $I$ be an ideal of
$K[x_1, \dots, x_n]$. Then $I$ is maximal if and only if
\[
  I = (x_1 - c_1, \dots, x_n - c_n),
\]
for $c_1, \dots, c_n \in K$.
\end{coro}
\begin{proof}
For $c_1, \dots, c_n \in K$,
\[
  \frac{K[x_1, \dots, x_n]}{(x_1 - c_1, \dots, x_n - c_n)} \cong K
\]
(Exercise~III.4.12) is a field; therefore $(x_1 - c_1, \dots, x_n - c_n)$
is maximal. Conversely, let $\mathfrak{m}$ be a maximal ideal in
$K[x_1, \dots, x_n]$, so that the quotient is a field and the natural map
\[
  K \to L := \frac{K[x_1, \dots, x_n]}{\mathfrak{m}}
\]
is a field extension. The field $L$ is finitely generated as a
$K$-algebra; therefore $K \subseteq L$ is an algebraic extension by
\cref{thm:fields:2.9}. Since $K$ is algebraically closed, this implies
that $L = K$ (\cref{lem:fields:2.1}); therefore, there is a surjective
homomorphism
\[
  \phi : K[x_1, \dots, x_n] \to K
\]
such that $\mathfrak{m} = \ker\phi$. Let $c_i := \phi(x_i)$. Then
\[
  (x_1 - c_1, \dots, x_n - c_n) \subseteq \ker\phi = \mathfrak{m};
\]
but $(x_1 - c_1, \dots, x_n - c_n)$ is maximal, so this implies
$\mathfrak{m} = (x_1 - c_1, \dots, x_n - c_n)$, as needed.
\end{proof}
Note how very (trivially) false the statement of \cref{coro:fields:2.10}
is over fields which are not algebraically closed: $(x^2 + 1)$ is maximal
in $\R[x]$. No such silliness may occur over $\C$, for any number of
variables.

\subsection{A little affine algebraic geometry}
\label{subsec:fields:a-little-affine-algebraic-geometry}
\cref{coro:fields:2.10} is one of the main reasons why `classical'
algebraic geometry developed over $\C$ rather than fields such as $\R$ or
$\Q$: it may be interpreted as saying that if $K$ is algebraically closed,
then there is a natural bijection between the points of the product space
$K^n$ and the maximal ideals in the ring $K[x_1, \dots, x_n]$. This is
the beginning of a fruitful dictionary translating geometry into algebra
and conversely. It is worthwhile formalizing this statement and exploring
other `geometric' consequences of the Nullstellensatz.

For $K$ a field, $\mathbb{A}^n_K$ denotes the \emph{affine space} of
dimension $n$ over $K$, that is, the set of $n$-tuples of elements of
$K$:
\[
  \mathbb{A}^n_K = \{(c_1, \dots, c_n) \mid c_i \in K\}.
\]
Elements of $\mathbb{A}^n_K$ are called \emph{points}.

\emph{Objection}: Why don't we just use `$K^n$' for this object? Because
the latter already stands for the standard $n$-dimensional vector space
over $K$; so it carries with it a certain baggage: the tuple
$(0, \dots, 0)$ is special, so are the vector subspaces of $K^n$, etc.
By contrast, no point of $\mathbb{A}^n_K$ is `special'; we can carry any
point to any other point by a simple translation. Similarly, although it
is useful to consider affine subspaces, these (unlike vector subspaces)
are not required to go through the origin. Affine geometry and linear
algebra are different in many respects.

We have already established that, by the Nullstellensatz, if $K$ is
algebraically closed, then there is a natural bijection between the points
of $\mathbb{A}^n_K$ and the maximal ideals of the polynomial ring
$K[x_1, \dots, x_n]$. Concretely, the bijection works like this: the
point $p = (c_1, \dots, c_n)$ corresponds to the set of polynomials
\[
  \mathcal{I}(p)
  := \{f(x) \in K[x_1, \dots, x_n] \mid f(p) = 0\},
\]
where $f(p)$ is the result of applying the evaluation map:
$f(x_1, \dots, x_n) \mapsto f(c_1, \dots, c_n) \in K$. This of course
is nothing but the homomorphism $\phi : K[x_1, \dots, x_n] \to K$
defined by prescribing $x_i \mapsto c_i$; the set $\mathcal{I}(p)$ is
simply $\ker\phi$, which is the maximal ideal
$(x_1 - c_1, \dots, x_n - c_n)$.

This correspondence $p \mapsto \mathcal{I}(p)$ may be defined over every
field; the content of \cref{coro:fields:2.10} is that every maximal ideal
of $K[x_1, \dots, x_n]$ corresponds to a point of $\mathbb{A}^n_K$ in
this way, if $K$ is algebraically closed.

It is natural to upgrade the correspondence as follows (again, over
arbitrary fields): for every subset $S \subseteq \mathbb{A}^n_K$,
consider the ideal
\[
  \mathcal{I}(S)
  := \{f(x) \in K[x_1, \dots, x_n] \mid \forall p \in S,\, f(p) = 0\},
\]
that is, the set of polynomials `vanishing along $S$'; it is immediately
checked that this is indeed an ideal of $K[x_1, \dots, x_n]$. We can
also consider a correspondence in the reverse direction, from ideals of
$K[x_1, \dots, x_n]$ to subsets of $\mathbb{A}^n_K$, defined by setting
for every ideal $I$,
\[
  \mathcal{V}(I)
  := \{p = (c_1, \dots, c_n) \in \mathbb{A}^n_K
      \mid \forall f \in I,\, f(c_1, \dots, c_n) = 0\}.
\]
Thus, $\mathcal{V}(I)$ is the set of common solutions of all the
polynomial equations $f = 0$ as $f \in I$: the set of points `cut out in
$\mathbb{A}^n_K$' by the polynomials in $I$.

In its most elementary manifestation, the dictionary mentioned in the
beginning of the section consists precisely of the pair of correspondences
\[
  \left\{\text{subsets of }\mathbb{A}^n_K\right\}
  \begin{matrix}
    \xrightarrow{\;\mathcal{I}\;} \\[3pt]
    \xleftarrow{\;\mathcal{V}\;}
  \end{matrix}
  \left\{\text{ideals in }K[x_1,\dots,x_n]\right\}.
\]
The set $\mathcal{V}(I)$ is often (somewhat improperly) called the
\emph{variety} of $I$, while $\mathcal{I}(S)$ is (also improperly) the
\emph{ideal} of $S$.

The function $\mathcal{V}$ can be defined (using the same prescription)
for any set $A \subseteq K[x_1, \dots, x_n]$, whether $A$ is an ideal
or not; it is clear that $\mathcal{V}(A) = \mathcal{V}(I)$ where $I$ is
the ideal generated by $A$ (Exercise~2.5). \name{Hilbert}'s basis theorem
(Theorem~V.1.2) tells us something rather interesting: $K[x_1, \dots,
x_n]$ is Noetherian when $K$ is a field; hence every ideal $I$ is
generated by a finite number of elements. Thus, for every ideal $I$ there
exist polynomials $f_1, \dots, f_r$ such that (abusing notation a little)
\[
  \mathcal{V}(I) = \mathcal{V}(f_1, \dots, f_r).
\]
In other words, if a set may be defined by any set of polynomial
equations, it may in fact be defined by a finite set of polynomial
equations.\footnote{I distinctly remember being surprised by this fact
the first time I ran into it.}

I will pass in silence many simple-minded properties of the functions
$\mathcal{V}$, $\mathcal{I}$ (they reverse inclusions, they behave
reasonably well with respect to unions and intersections, etc.); the
reader should feel free to research the topic at will. But I want to
focus the reader's attention on the fact that (of course) $\mathcal{V}$,
$\mathcal{I}$ are not bijections; this is a problem, if we really want
to construct a dictionary between `geometry' and `algebra'.

For example, there are (in general) lots of subsets of $\mathbb{A}^n_K$
which are not cut out by a set of polynomial equations, and there are
lots of ideals in $K[x_1, \dots, x_n]$ which are not obtained as
$\mathcal{I}(S)$ for any subset $S \subseteq \mathbb{A}^n_K$.

I will leave to the reader the task of finding examples of the first
phenomenon (Exercise~2.6). The way around it is to restrict our attention
to subsets of $\mathbb{A}^n_K$ which are of the form $\mathcal{V}(I)$
for some ideal $I$.
\begin{defn}{}{fields:2.11}
An \emph{(affine) algebraic set} is a subset $S \subseteq \mathbb{A}^n_K$
such that there exists an ideal $I \subseteq K[x_1, \dots, x_n]$ for
which $S = \mathcal{V}(I)$.
\end{defn}
This definition may seem like a cheap patch: we do not really know what
the image of $\mathcal{V}$ is, so we label it in some way and proceed.
This is not entirely true --- the family of algebraic subsets of a given
$\mathbb{A}^n_K$ has a solid, recognizable structure: it is the family of
closed subsets in a topology on $\mathbb{A}^n_K$ (Exercise~2.7); this
topology is called the \emph{Zariski topology}.

Studying affine algebraic sets is the business of affine algebraic
geometry. The aim is to understand `geometric' properties of these sets
in terms of `algebraic' properties of corresponding ideals or other
algebraic entities associated with them.
\begin{exam}{}{fields:exam:2.12}
Here are pictures of two algebraic subsets of $\mathbb{A}^2_{\R}$: the
first is $\mathcal{V}((y - x^2))$; the second is $\mathcal{V}((y^2 -
x^3))$. What feature of the ideal $(y^2 - x^3)$ is responsible for the
`cusp' at $(0, 0)$ in the second picture? The reader will find out in any
course in elementary algebraic geometry.
\end{exam}
The fact that $\mathcal{I}$ is not surjective leads to interesting
considerations. The standard example of an ideal that is not the ideal of
any set is $(x^2)$ in $K[x]$: wherever $x^2$ vanishes, so does $x$;
hence if $x^2 \in \mathcal{I}(S)$ for a set $S$, then $x \in
\mathcal{I}(S)$ as well. This motivates the following definitions.
\begin{defn}{}{fields:2.13}
Let $I$ be an ideal in a commutative ring $R$. The \emph{radical} of $I$
is the ideal
\[
  \sqrt{I} := \{r \in R \mid \exists k \geq 0,\; r^k \in I\}.
\]
An ideal $I$ is a \emph{radical ideal} if $I = \sqrt{I}$.
\end{defn}
The reader should check that $\sqrt{I}$ is indeed an ideal; it may be
characterized as the intersection of all prime ideals containing $I$
(Exercise~2.8).
\begin{exam}{}{fields:exam:2.14}
An element of a commutative ring $R$ is nilpotent if and only if it is in
the radical of the ideal $(0)$ (cf.\ Exercise~V.4.19). The reader has
encountered this ideal, called the \emph{nilradical} of $R$, in the
exercises, beginning with Exercise~III.3.12.

Prime ideals $\mathfrak{p}$ are clearly radical: because $f^k \in
\mathfrak{p} \implies f \in \mathfrak{p}$, and therefore $\mathfrak{p}
\subseteq \sqrt{\mathfrak{p}}$ (the other inclusion holds trivially for
all ideals).
\end{exam}
\begin{lem}{}{fields:2.15}
Let $K$ be a field, and let $S$ be a subset of $\mathbb{A}^n_K$. Then
the ideal $\mathcal{I}(S)$ is a radical ideal of $K[x_1, \dots, x_n]$.
\end{lem}
\begin{proof}
The inclusion $\mathcal{I}(S) \subseteq \sqrt{\mathcal{I}(S)}$ holds for
every ideal, so it is trivially satisfied. To verify the inclusion
$\sqrt{\mathcal{I}(S)} \subseteq \mathcal{I}(S)$, let $f \in
\sqrt{\mathcal{I}(S)}$. Then there is an integer $k \geq 0$ such that
$f^k \in \mathcal{I}(S)$; that is, $(\forall p \in S),\; f(p)^k = 0$.
But then $(\forall p \in S),\; f(p) = 0$, proving $f \in \mathcal{I}(S)$,
as needed.
\end{proof}
In view of these considerations, for any field we can refine the
correspondence between subsets of $\mathbb{A}^n_K$ and ideals of
$K[x_1, \dots, x_n]$ as follows:
\[
  \left\{\text{algebraic subsets of }\mathbb{A}^n_K\right\}
  \begin{matrix}
    \xrightarrow{\;\mathcal{I}\;} \\[3pt]
    \xleftarrow{\;\mathcal{V}\;}
  \end{matrix}
  \left\{\text{radical ideals in }K[x_1,\dots,x_n]\right\};
\]
as I have argued, it is necessary to do so if we want to have a good
dictionary. In the rest of the section, $\mathcal{I}$ and $\mathcal{V}$
will be taken as acting between these two sets.

While we have tightened the correspondence substantially, the situation
is still less than idyllic. Over arbitrary fields, the function
$\mathcal{V}$ is now surjective by the very definition of an affine
algebraic subset (cf.\ Exercise~2.9); but it is not necessarily
injective. An immediate counterexample is offered over $K = \R$ by the
ideal $(x^2 + 1)$: this is maximal, hence prime, hence radical, and
\[
  \mathcal{V}(x^2 + 1) = \emptyset = \mathcal{V}(1).
\]
Here is where the Nullstellensatz comes to our aid, and here is where we
see what the Nullstellensatz has to do with the `zeros' of an ideal
($ = \mathcal{V}$ of that ideal).
\begin{prop}[Weak Nullstellensatz]{}{fields:2.16}
Let $K$ be an algebraically closed field, and let $I \subseteq
K[x_1, \dots, x_n]$ be an ideal. Then $\mathcal{V}(I) = \emptyset$ if
and only if $I = (1)$.
\end{prop}
\begin{proof}
If $I = (1)$, then $\mathcal{V}(I) = \emptyset$ by definition.
Conversely, assume that $I \neq (1)$. By Proposition~V.3.5, $I$ is then
contained in a maximal ideal $\mathfrak{m}$. Since $K$ is algebraically
closed, by \cref{coro:fields:2.10} we have
$\mathfrak{m} = (x_1 - c_1, \dots, x_n - c_n)$
for some $c_1, \dots, c_n \in K$; so for all $f(x_1, \dots, x_n) \in I$
there exist $g_1, \dots, g_n \in K[x_1, \dots, x_n]$ such that
\[
  f(x_1, \dots, x_n)
  = \sum_{i=1}^{n} g_i(x_1, \dots, x_n)(x_i - c_i).
\]
In particular,
\[
  f(c_1, \dots, c_n)
  = \sum_{i=1}^{n} g_i(c_1, \dots, c_n)(c_i - c_i) = 0:
\]
that is, $(c_1, \dots, c_n) \in \mathcal{V}(I)$. This proves
$\mathcal{V}(I) \neq \emptyset$ if $I \neq (1)$, and we are done.
\end{proof}
This is encouraging, but it does not seem to really prove that $\mathcal{V}$
is injective. As it happens, however, it does: we can derive from the
`weak' Nullstellensatz the following stronger result, which does imply
injectivity:
\begin{prop}[Strong Nullstellensatz]{}{fields:2.17}
Let $K$ be an algebraically closed field, and let $I \subseteq
K[x_1, \dots, x_n]$ be an ideal. Then
\[
  \mathcal{I}(\mathcal{V}(I)) = \sqrt{I}.
\]
This implies that the composition $\mathcal{I} \circ \mathcal{V}$ is the
identity on the set of radical ideals; that is, $\mathcal{V}$ has a
left-inverse, and therefore it is injective (cf.\ Proposition~I.2.1!).
\end{prop}
\begin{proof}
Note that $\mathcal{I}(\mathcal{V}(I))$ is a radical ideal (by
\cref{lem:fields:2.15}). It follows immediately from the definitions that
$I \subseteq \mathcal{I}(\mathcal{V}(I))$; therefore
\[
  \sqrt{I} \subseteq \mathcal{I}(\mathcal{V}(I))
  = \mathcal{I}(\mathcal{V}(I)).
\]
We have to verify the reverse inclusion: assume that $f \in
\mathcal{I}(\mathcal{V}(I))$, that is, $f(p) = 0$ for all $p \in
\mathbb{A}^n_K$ such that $\forall g \in I,\, g(p) = 0$; we have to show
that there exists an $m$ for which $f^m \in I$.
The argument is not difficult after the fact, but it is fiendishly
clever.\footnote{This is called the \name{Rabinowitsch} trick and dates
back to 1929. It was published in a one-page, eighteen-line article in the
\textit{Mathematische Annalen}.} Let $y$ be an extra variable, and
consider the ideal $J$ generated by $I$ and $1 - fy$ in
$K[x_1, \dots, x_n, y]$. More explicitly, assume
$I = (g_1, \dots, g_r)$
(since $K[x_1, \dots, x_n]$ is Noetherian, finitely many generators
suffice); we can view $g_i(x_1, \dots, x_n)$, resp.,
$f(x_1, \dots, x_n) \in K[x_1, \dots, x_n]$, as polynomials
$G_i(x_1, \dots, x_n, y)$, resp., $F(x_1, \dots, x_n, y) \in
K[x_1, \dots, x_n, y]$, and let
\[
  J = (G_1, \dots, G_r, 1 - Fy).
\]
As $K[x_1, \dots, x_n, y]$ has $n+1$ variables, the ideal $J$ defines a
subset $\mathcal{V}(J)$ in $\mathbb{A}^{n+1}_K$. I claim $\mathcal{V}(J)
= \emptyset$.

Indeed, assume on the contrary that $\mathcal{V}(J) \neq \emptyset$, and
let $p = (a_1, \dots, a_n, b) \in \mathbb{A}^{n+1}_K$ be a point of
$\mathcal{V}(J)$. Then for $i = 1, \dots, r$, $G_i(a_1, \dots, a_n, b)
= 0$; but this means $g_i(a_1, \dots, a_n) = 0$ for $i = 1, \dots, r$;
that is, $(a_1, \dots, a_n) \in \mathcal{V}(I)$. Since $f$ vanishes at
all points of $\mathcal{V}(I)$, this implies $f(a_1, \dots, a_n) = 0$,
and then
\[
  (1 - Fy)(a_1, \dots, a_n, b)
  = 1 - f(a_1, \dots, a_n)\,b = 1 - 0 = 1.
\]
But this contradicts the assumption that $p \in \mathcal{V}(J)$.
Therefore, $\mathcal{V}(J) = \emptyset$.
Now use the weak Nullstellensatz: since $K$ is algebraically closed, we
can conclude $J = (1)$. Therefore, there exist polynomials
$H_i(x_1, \dots, x_n, y)$, $i = 1, \dots, r$, and
$L(x_1, \dots, x_n, y)$, such that
\[
  \sum_{i=1}^{r} H_i(\mathbf{x}, y)\,G_i(\mathbf{x}, y)
  + L(\mathbf{x}, y)(1 - F(\mathbf{x}, y)\,y) = 1.
\]
We are not done being clever: the next step is to view this equality in
the field of rational functions over $K$ rather than in the polynomial
ring, which we can do, since the latter is a subring of the former. We
can then plug in $y = 1/F$ and still get an equality, with the advantage
of killing the last summand on the left-hand side:
\[
  \sum_{i=1}^{r} H_i\!\left(\mathbf{x}, \tfrac{1}{F}\right)
  G_i\!\left(\mathbf{x}, \tfrac{1}{F}\right) = 1.
\]
Further, $G_i(\mathbf{x}, 1/F) = g_i(x_1, \dots, x_n)$ ($G_i$ is just
the name of $g_i$ in the larger polynomial ring and only depends on the
first $n$ variables), while $H_i(\mathbf{x}, 1/F) =
h_i(x_1, \dots, x_n)/f(x_1, \dots, x_n)^m$, where $h_i \in
K[x_1, \dots, x_n]$ and $m$ is an integer large enough to work for all
$i = 1, \dots, r$. Expressing in terms of a common denominator, we can
rewrite the identity as
\[
  \frac{h_1 g_1 + \dots + h_r g_r}{f^m} = 1,
\]
or $f^m = h_1 g_1 + \dots + h_r g_r$. We have proved this identity in
the field of rational functions; as it only involves polynomials, it holds
in $K[x_1, \dots, x_n]$. The right-hand side is an element of $I$, so
this proves $f^m \in I$, and we are done.
\end{proof}
The addition of the variable $y$ in the proof looks more reasonable (and
less like a trick) if one views it geometrically. The variety
$\mathcal{V}(1 - xy)$ in $\mathbb{A}^2_K$ is an ordinary hyperbola. The
projection $(x, y) \mapsto x$ may be used to identify $\mathcal{V}(1 -
xy)$ with the complement of the origin (that is, $\mathcal{V}(x)$) in the
``$x$-axis'' (viewed as $\mathbb{A}^1_K$). Similarly, $\mathcal{V}(1 -
fy)$ is a way to realize the Zariski open set $\mathbb{A}^n_K \setminus
\mathcal{V}(f)$ as a Zariski closed subset in a space $\mathbb{A}^{n+1}_K$
of dimension one higher. The proof of \cref{prop:fields:2.17} hinges on
this fact, and this fact is also an important building block in the basic
set-up of algebraic geometry, since it can be used to show that the
Zariski topology has a basis consisting of (sets which are isomorphic in
a suitable sense to) affine algebraic sets.
I will officially record the conclusion we were able to establish
concerning the functions $\mathcal{V}$, $\mathcal{I}$:
\begin{coro}{}{fields:2.18}
Let $K$ be an algebraically closed field. Then for any $n \geq 0$ the
functions
\[
  \left\{\text{algebraic subsets of }\mathbb{A}^n_K\right\}
  \begin{matrix}
    \xrightarrow{\;\mathcal{I}\;} \\[3pt]
    \xleftarrow{\;\mathcal{V}\;}
  \end{matrix}
  \left\{\text{radical ideals in }K[x_1,\dots,x_n]\right\}
\]
are inverses of each other.
\end{coro}
\begin{proof}
\cref{prop:fields:2.17} shows that $\mathcal{I} \circ \mathcal{V}$ is the
identity on radical ideals, so $\mathcal{V}$ is injective, and
$\mathcal{V}$ is surjective by definition of affine algebraic set. It
follows that $\mathcal{V}$ is a bijection and $\mathcal{I}$ is its
inverse.
\end{proof}
Summarizing: If $K$ is algebraically closed, then studying sets defined by
polynomial equations in a space $K^n$ is `the same thing as' studying
radical ideals in a polynomial ring $K[x_1, \dots, x_n]$.
This correspondence is actually realized even more effectively at the
level of $K$-algebras. We say that a ring is \emph{reduced} if it has no
nonzero nilpotents; then $R/I$ is reduced if and only if $I$ is a radical
ideal (Exercise~2.8).
\begin{defn}{}{fields:2.19}
Let $S \subseteq \mathbb{A}^n_K$ be an algebraic set. The
\emph{coordinate ring} of $S$ is the quotient\footnote{The notation
`$K[S]$' is fairly standard, although it may lead to confusion with the
usual polynomial rings; hopefully the context takes care of this. It makes
sense to incorporate the name of the field in the notation, since one
could in principle change the base field while keeping the `same' equations
encoded in the ideal $\mathcal{I}(S)$.}
\[
  K[S] := \frac{K[x_1, \dots, x_n]}{\mathcal{I}(S)}.
\]
\end{defn}
Thus, the coordinate ring of an algebraic set is a reduced, commutative
$K$-algebra of finite type. The reader will establish a `concrete'
interpretation of this ring, as the ring of `polynomial functions' on $S$,
in Exercise~2.12. One way to think about the Nullstellensatz (in its
manifestation as \cref{coro:fields:2.18}) is that, if $K$ is algebraically
closed, then every reduced commutative $K$-algebra of finite type $R$ may
be realized as the coordinate ring of an affine algebraic set $S$; the
points of $S$ correspond in a natural way to the maximal ideals of $R$
(Exercise~2.14).
In fact, in basic algebraic geometry one defines the category of affine
algebraic sets over a field $K$ in terms of this correspondence: morphisms
of algebraic sets are defined so that they match homomorphisms of the
corresponding (reduced, finite-type) $K$-algebras. (The reader will
encounter a more precise definition in Example~VIII.1.9.)
This dictionary allows us to translate every `geometric' feature of an
algebraic set (such as dimension, smoothness, etc.) into corresponding
`algebraic' features of the corresponding coordinate rings. Once these key
algebraic features have been identified, then one can throw away the
restriction of working on reduced finite-type algebras over an algebraically
closed field and try to `do geometry' on any (say commutative, Noetherian)
ring. For example, it turns out that PIDs correspond to (certain) smooth
curves in the affine geometric world; and then one can try to think of $\Z$
as a `smooth curve' (although $\Z$ is not a reduced finite-type $K$-algebra
over any field $K$!) and try to use theorems inspired by the geometry of
curves to understand features of $\Z$---that is, use geometry to do number
theory.\footnote{This is of course a drastic oversimplification.}
Another direction in which these simple considerations may be generalized is
by `gluing' affine algebraic sets into manifold-like objects: sets which may
not be affine algebraic sets `globally' but may be covered by affine
algebraic sets. A concrete way to do this is by working in projective space
rather than affine space; the globalization process can then be carried out
in a straightforward way at the algebraic level, where it translates into
the study of `graded' rings and modules---more notions for which,
regretfully, I will find no room in this book (except for a glance, in
\S{}VIII.4.3). In the second half of the twentieth century a more abstract
viewpoint, championed by \name{Alexander Grothendieck} and others, has
proven to be extremely effective; it has led to the intense study of
schemes, the current language of choice in algebraic geometry. We have
encountered the simplest kind of schemes when we introduced the spectrum of
a ring $R$, $\operatorname{Spec} R$, back in \S{}III.4.3.

\subsection*{Exercises}

\begin{exer}{}{fields:exer:2.1}
  Prove \cref{lem:fields:2.1}.
\end{exer}
\begin{exer}{}{fields:exer:2.2}
  Let $k \subseteq \bar{k}$ be an algebraic closure, and let $L$ be an
  intermediate field. Assume that every polynomial $f(x) \in k[x] \subseteq
  L[x]$ factors as a product of linear terms in $L[x]$. Prove that
  $L = \bar{k}$.
\end{exer}
\begin{exer}{}{fields:exer:2.3}
  Prove that if $k$ is a countable field, then so is $\bar{k}$.
\end{exer}
\begin{exer}{}{fields:exer:2.4}
  Let $k$ be a field, let $c_1, \dots, c_m \in k$ be distinct elements,
  and let $\lambda_1, \dots, \lambda_m$ be nonzero elements of $k$. Prove
  that
  \[
    \frac{\lambda_1}{x - c_1} + \cdots + \frac{\lambda_m}{x - c_m} \neq 0
  \]
  in $k(x)$. (This fact is used in the proof of \cref{thm:fields:2.9}.)
\end{exer}
\begin{exer}{}{fields:exer:2.5}
  Let $K$ be a field, let $A$ be a subset of $K[x_1, \dots, x_n]$, and
  let $I$ be the ideal generated by $A$. Prove that $\mathcal{V}(A) =
  \mathcal{V}(I)$ in $\mathbb{A}^n_K$.
\end{exer}
\begin{exer}{}{fields:exer:2.6}
  Let $K$ be your favorite infinite field. Find examples of subsets $S
  \subseteq \mathbb{A}^n_K$ which cannot be realized as $\mathcal{V}(I)$
  for any ideal $I \subseteq K[x_1, \dots, x_n]$. Prove that if $K$ is a
  finite field, then every subset $S \subseteq \mathbb{A}^n_K$ equals
  $\mathcal{V}(I)$ for some ideal $I \subseteq K[x_1, \dots, x_n]$.
\end{exer}
\begin{exer}{}{fields:exer:2.7}
  Let $K$ be a field and $n$ a nonnegative integer. Prove that the set of
  algebraic subsets of $\mathbb{A}^n_K$ is the family of closed sets of a
  topology on $\mathbb{A}^n_K$.
\end{exer}
\begin{exer}{}{fields:exer:2.8}
  With notation as in \cref{defn:fields:2.13}:
  \begin{itemize}
    \item Prove that the set $\sqrt{I}$ is an ideal of $R$.
    \item Prove that $\sqrt{I}$ corresponds to the nilradical of $R/I$ via
      the correspondence between ideals of $R/I$ and ideals of $R$
      containing $I$.
    \item Prove that $\sqrt{I}$ is in fact the intersection of all prime
      ideals of $R$ containing $I$. (Cf.\ Exercise~V.3.13.)
    \item Prove that $I$ is radical if and only if $R/I$ is reduced.
      (Cf.\ Exercise~III.3.13.)
  \end{itemize}
\end{exer}
\begin{exer}{}{fields:exer:2.9}
  Prove that every affine algebraic set equals $\mathcal{V}(I)$ for a
  radical ideal $I$.
\end{exer}
\begin{exer}{}{fields:exer:2.10}
  Prove that every ideal in a Noetherian ring contains a power of its
  radical.
\end{exer}
\begin{exer}{}{fields:exer:2.11}
  Assume a field $K$ is not algebraically closed. Find a reduced
  finite-type $K$-algebra which is not the coordinate ring of any affine
  algebraic set.
\end{exer}
\begin{exer}{}{fields:exer:2.12}
  Let $K$ be an infinite field. A \emph{polynomial function} on an affine
  algebraic set $S \subseteq \mathbb{A}^n_K$ is the restriction to $S$ of
  (the evaluation function of) a polynomial $f(x_1, \dots, x_n) \in
  K[x_1, \dots, x_n]$. Polynomial functions on $S$ manifestly form a ring
  and in fact a $K$-algebra. Prove that this $K$-algebra is isomorphic to
  the coordinate ring of $S$.
\end{exer}
\begin{exer}{}{fields:exer:2.13}
  Let $K$ be an algebraically closed field. Prove that every reduced
  commutative $K$-algebra of finite type is the coordinate ring of an
  algebraic set $S$ in some affine space $\mathbb{A}^n_K$.
\end{exer}
\begin{exer}{}{fields:exer:2.14}
  Prove that, over an algebraically closed field $K$, the points of an
  algebraic set $S$ correspond to the maximal ideals of the coordinate ring
  $K[S]$ of $S$, in such a way that if $p$ corresponds to the maximal ideal
  $\mathfrak{m}_p$, then the value of the function $f \in K[S]$ at $p$
  equals the coset of $f$ in $K[S]/\mathfrak{m}_p \cong K$.
\end{exer}
\begin{exer}{}{fields:exer:2.15}
  Let $K$ be an algebraically closed field. An algebraic subset $S$ of
  $\mathbb{A}^n_K$ is \emph{irreducible} if it cannot be written as the
  union of two algebraic subsets properly contained in it. Prove that $S$
  is irreducible if and only if its ideal $\mathcal{I}(S)$ is prime, if and
  only if its coordinate ring $K[S]$ is an integral domain.

  An irreducible algebraic set is `all in one piece', like $\mathbb{A}^n_K$
  itself, and unlike (for example) $\mathcal{V}(xy)$ in the affine plane
  $\mathbb{A}^2_K$ with coordinates $x$, $y$. Irreducible affine algebraic
  sets are called \emph{(affine algebraic) varieties}.
\end{exer}
\begin{exer}{}{fields:exer:2.16}
  Let $K$ be an algebraically closed field. The field of rational functions
  $K(x_1, \dots, x_n)$ is the field of fractions of $K[\mathbb{A}^n_K] =
  K[x_1, \dots, x_n]$; every rational function $\alpha = F/G$ (with
  $G \neq 0$ and $F$, $G$ relatively prime) may be viewed as defining a
  function on the open set $\mathbb{A}^n_K \setminus \mathcal{V}(G)$; we
  say that $\alpha$ is `defined' for all points in the complement of
  $\mathcal{V}(G)$.

  Let $G \in K[x_1, \dots, x_n]$ be irreducible. The set of rational
  functions that are defined in the complement of $\mathcal{V}(G)$ is a
  subring of $K(x_1, \dots, x_n)$. Prove that this subring may be
  identified with the localization (Exercise~V.4.7) of $K[\mathbb{A}^n_K]$
  at the multiplicative set $\{1, G, G^2, G^3, \dots\}$. (Use the
  Nullstellensatz.)

  The same considerations may be carried out for any irreducible algebraic
  set $S$, adopting as field of `rational functions' $K(S)$ the field of
  fractions of the integral domain $K[S]$.
\end{exer}
\begin{exer}{}{fields:exer:2.17}
  Let $K$ be an algebraically closed field, and let $\mathfrak{m}$ be a
  maximal ideal of $K[x_1, \dots, x_n]$, corresponding to a point $p$ of
  $\mathbb{A}^n_K$. A \emph{germ of a function} at $p$ is determined by an
  open set containing $p$ and a function defined on that open set; in our
  context (dealing with rational functions and where the open set may be
  taken to be the complement of a function that does not vanish at $p$)
  this is the same information as a rational function defined at $p$, in
  the sense of Exercise~2.16. Show how to identify the ring of germs with
  the localization $K[\mathbb{A}^n_K]_{\mathfrak{m}}$ (defined in
  Exercise~V.4.11).

  As in Exercise~2.16, the same discussion can be carried out for any
  algebraic set. This is the origin of the name `localization': localizing
  the coordinate ring of a variety $V$ at the maximal ideal corresponding
  to a point $p$ amounts to considering only functions defined in a
  neighborhood of $p$, thus studying $V$ `locally', `near $p$'.
\end{exer}
\begin{exer}{}{fields:exer:2.18}
  Let $K$ be an algebraically closed field. Consider the two `curves'
  $C_1 : y = x^2$, $C_2 : y^2 = x^3$ in $\mathbb{A}^2_K$ (pictures of
  the real points of these algebraic sets are shown in
  \cref{exam:fields:exam:2.12}).
  \begin{itemize}
    \item Prove that $K[C_1] \cong K[t] = K[\mathbb{A}^1_K]$, while
      $K[C_2]$ may be identified with the subring $K[t^2, t^3]$ of $K[t]$
      consisting of polynomials $a_0 + a_2 t^2 + \dots + a_d t^d$ with
      zero $t$-coefficient. (Note that every polynomial in $K[x, y]$ may
      be written as $f(x) + g(x)y + h(x,y)(y^2 - x^3)$ for uniquely
      determined polynomials $f(x)$, $g(x)$, $h(x,y)$.)
    \item Show that $C_1$, $C_2$ are both irreducible
      (cf.\ Exercise~2.15).
    \item Prove that $K[C_1]$ is a UFD, while $K[C_2]$ is not.
    \item Show that the Krull dimension of both $K[C_1]$ and $K[C_2]$ is
      $1$. (This is why these sets would be called `curves'. You may use
      the fact that maximal chains of prime ideals in $K[x,y]$ have
      length~$2$.)
    \item The origin $(0,0)$ is in both $C_1$, $C_2$ and corresponds to
      the maximal ideals $\mathfrak{m}_1$, resp., $\mathfrak{m}_2$, in
      $K[C_1]$, resp., $K[C_2]$, generated by the classes of $x$ and $y$.
    \item Prove that the localization $K[C_1]_{\mathfrak{m}_1}$ is a DVR
      (Exercise~V.4.13). Prove that the localization
      $K[C_2]_{\mathfrak{m}_2}$ is not a DVR. (Note that the relation
      $y^2 = x^3$ still holds in this ring; prove that
      $K[C_2]_{\mathfrak{m}_2}$ is not a UFD.)
  \end{itemize}
  The fact that a DVR admits a local parameter, that is, a single
  generator for its maximal ideal (cf.\ Exercise~V.2.20), is a good
  algebraic translation of the fact that a curve such as $C_1$ has a
  single, smooth branch through $(0,0)$. The maximal ideal of
  $K[C_2]_{\mathfrak{m}_2}$ cannot be generated by just one element, as
  the reader may verify.
\end{exer}
\begin{exer}{}{fields:exer:2.19}
  Prove that the fields of rational functions (Exercise~2.16) of the
  curves $C_1$ and $C_2$ of Exercise~2.18 are isomorphic and both have
  transcendence degree $1$ over $k$ (cf.\ Exercise~1.27).

  This is another reason why we should think of $C_1$ and $C_2$ as
  `curves'. In fact, it can be proven that the Krull dimension of the
  coordinate ring of a variety equals the transcendence degree of its
  field of rational functions. This is a consequence of
  \name{Noether}'s normalization theorem, a cornerstone of commutative
  algebra.
\end{exer}
\begin{exer}{}{fields:exer:2.20}
  Recall from Exercise~VI.2.13 that $\mathbb{P}^n_K$ denotes the
  `projective space' parametrizing lines in the vector space $K^{n+1}$.
  Every such line consists of multiples of a nonzero vector $(c_0, \dots,
  c_n) \in K^{n+1}$, so that $\mathbb{P}^n_K$ may be identified with the
  quotient (in the set-theoretic sense of \S{}I.1.5) of
  $K^{n+1} \setminus \{(0, \dots, 0)\}$ by the equivalence relation $\sim$
  defined by
  \[
    (c_0, \dots, c_n) \sim (c_0', \dots, c_n')
    \iff (\exists\, \lambda \in K^*)\;
    (c_0', \dots, c_n') = (\lambda c_0, \dots, \lambda c_n).
  \]
  The `point' in $\mathbb{P}^n_K$ determined by the vector $(c_0, \dots,
  c_n)$ is denoted $(c_0 : \cdots : c_n)$; these are the `projective
  coordinates'\footnote{This is a convenient abuse of language. Keep in
  mind that the $c_i$'s are not determined by the point, so they are not
  `coordinates' in any strict sense. Their ratios are, however, determined
  by the point.} of the point. Note that there is no `point'
  $(0 : \cdots : 0)$.

  Prove that the function $\mathbb{A}^n_K \to \mathbb{P}^n_K$ defined by
  \[
    (c_1, \dots, c_n) \mapsto (1 : c_1 : \cdots : c_n)
  \]
  is a bijection. This function is used to realize $\mathbb{A}^n_K$ as a
  subset of $\mathbb{P}^n_K$. By using similar functions, prove that
  $\mathbb{P}^n_K$ can be covered with $n+1$ copies of $\mathbb{A}^n_K$,
  and relate this fact to the cell decomposition obtained in
  Exercise~VI.2.13. (Suggestion: Work out carefully the case $n = 2$.)
\end{exer}

\begin{exer}{}{fields:exer:2.21}
  Let $F(x_0, \dots, x_n) \in K[x_0, \dots, x_n]$ be a homogeneous
  polynomial. With notation as in Exercise~2.20, prove that the condition
  `$F(c_0, \dots, c_n) = 0$' for a point $(c_0 : \cdots : c_n) \in
  \mathbb{P}^n_K$ is well-defined: it does not depend on the
  representative $(c_0, \dots, c_n)$ chosen for the point $(c_0 : \cdots :
  c_n)$. We can then define the following subset of $\mathbb{P}^n_K$:
  \[
    \mathcal{V}(F) :=
    \{(c_0 : \cdots : c_n) \in \mathbb{P}^n_K \mid F(c_0,\dots,c_n)=0\}.
  \]
  Prove that this `projective algebraic set' can be covered with $n+1$
  affine algebraic sets.

  The basic definitions in `projective algebraic geometry' can be
  developed along essentially the same path taken in this section for
  affine algebraic geometry, using `homogeneous ideals' (that is, ideals
  generated by homogeneous polynomials; see \S{}VIII.4.3) rather than
  ordinary ideals. This problem shows one way to relate projective and
  affine algebraic sets, in one template example.
\end{exer}

\section{Geometric impossibilities}
\label{sec:fields:geometric-impossibilities}

Very simple considerations in field theory dispose easily of a whole class
of geometric problems which had seemed unapproachable for a very long
time---problems which preoccupied the Greeks and were only solved in the
relatively recent past. These problems have to do with the construction of
certain geometric figures with `straightedge and compass', that is, subject
to certain very strict rules. The practical utility of these constructions
would appear to be absolute zero, but the intellectual satisfaction of being
able to thoroughly understand matters which stumped extremely intelligent
people for a very long time is well worth the little side trip.

\subsection{Constructions by straightedge and compass}
\label{subsec:fields:constructions-by-straightedge-and-compass}
We begin with two points $O$, $P$ in the ordinary, real plane. You are
allowed to mark (`construct') more points and other geometric figures in the
plane, but only according to the following rules:
\begin{itemize}
  \item If you have constructed two points $A$, $B$, then you can draw the
    line joining them (using your straightedge).
  \item If you have constructed two points $A$, $B$, then you can draw the
    circle with center at $A$ and containing $B$ (using your compass).
  \item You can mark any number of points of intersection of any two
    distinct lines, line and circle, or circles that you have drawn
    already.
\end{itemize}
Performing these actions leads to complex (and beautiful) collections of
lines and circles; we say that we have `constructed' a geometric figure if
that figure appears as a subset of the whole picture.
For example, here is a recipe to construct an equilateral triangle with
$O$, $P$ as two of its vertices:
\begin{enumerate}
  \item draw the circle with center $O$, through $P$;
  \item draw the circle with center $P$, through $O$;
  \item let $Q$ be any of the two points of intersection of the two
    circles;
  \item draw the lines through $O$, $P$; $O$, $Q$; and $P$, $Q$.
\end{enumerate}
Of course $O$, $P$, $Q$ are vertices of an equilateral triangle.
It is a pleasant exercise to try to come up with a specific sequence of
basic moves constructing a given figure or geometric configuration. The
enterprising reader could try to produce the vertices of a pentagon by a
straightedge-and-compass construction, without looking it up and before we
learn enough to thwart pure geometric intuition.
The general question is to decide whether a figure can or cannot be
constructed this way. Three problems of this kind became famous in
antiquity:
\begin{itemize}
  \item trisecting angles;
  \item squaring circles;
  \item doubling cubes.
\end{itemize}
For example, one assumes one has already constructed two lines forming an
angle $\theta$ and asks whether one can construct two lines forming
$\theta/3$. This is trivially possible for some constructible $\theta$: we
just constructed an angle of $\pi/3$, so evidently we were able to trisect
the angle $\pi$; but can this be done for all constructible angles? The
answer is no. Similarly, it is not possible to construct a square whose
area equals the area of a (constructible) circle, and it is not possible
to construct the side of a cube whose volume is twice as large as a cube
with given (constructible) side.
Field theory will allow us to establish all this.\footnote{However, for
the impossibility of squaring circles we will use the transcendence of
$\pi$, which we are taking on faith.} Later we will return to these
constructions and study the question of the constructibility of regular
polygons with a given side: it is very easy to construct equilateral
triangles (we just did), squares, hexagons; it is not too hard to construct
pentagons; but the question of which polygons are constructible and which
are not connects beautifully with deeper questions in field theory.
The three problems listed above have an illustrious history; they
apparently date back to at least 414 B.C., if one is to take this fragment
from \name{Aristophanes}' ``The Birds'' as an indication:\footnote{I thank
\name{Matilde Marcolli} for the translation. For those who are less
ignorant than I am: ὀρθῷ μετρήσω κανόνι προστιθεὶς ἵνα, / ὁ κύκλος
γένηταί σοι τετράγωνος κἂν μέσωι ἀγορά.}
\begin{quote}
  \textsc{Meton}: By measuring with a straightedge there \\
  the circle becomes a square with a court within.
\end{quote}
The problems depend on the precise restrictions imposed on the
constructions. For example, one can trisect angles if one is allowed to
mark points on the straightedge (thus making it a ruler); in fact, this was
already known to \name{Archimedes} (Exercise~3.13).
Translating these geometric problems into algebra requires establishing the
feasibility of a few basic constructions.
First of all, one can construct a line containing a given point $A$ and
perpendicular to a given line $\ell$ (thus, $\ell$ contains at least
another constructible point $B$). For this, draw the circle with center $A$
and containing $B$; if this circle only intersects $\ell$ at $B$, then the
line through $A$ and $B$ is perpendicular to $\ell$; otherwise, the circle
intersects $\ell$ at a second point $C$ (whether $A$ is on $\ell$ or not),
and the line through $A$ perpendicular to $\ell$ may be found by joining
the two points of intersection $D$, $E$ of the two circles with centers at
$B$, resp., $C$, and containing $C$, resp., $B$. Applying this
construction twice gives the line through a point $A$ and parallel to a
given line, which is often handy.
Judicious application of these operations allows us to bring a Cartesian
reference system into the picture. Starting with the initial two points
$O$, $P$, we can construct perpendicular Cartesian axes centered at $O$,
with $P$ marking (say) the point $(1,0)$ on the `$x$-axis'; constructing a
point $A = (x, y)$ is equivalent to constructing its projections $X =
(x,0)$, $Y = (0,y)$ onto the axes, and in fact it is equivalent to
constructing the two points $X = (x,0)$ and $Y = (y,0)$. It follows that
determining which figures are constructible by straightedge and compass is
equivalent to determining which numbers may be realized as coordinates of a
constructible point.
\begin{defn}{}{fields:3.1}
  A real number $r$ is \emph{constructible} if the point $(r, 0)$ is
  constructible with straightedge and compass (assuming $O = (0,0)$ and
  $P = (1,0)$, as above). I will denote by $\mathcal{C}_\mathbb{R}
  \subseteq \mathbb{R}$ the set of constructible real numbers.

  Also, we can identify the real plane with $\mathbb{C}$, placing $O$ at
  $0$ and $P$ at $1$, and we say that $z = x + iy$ is constructible if the
  point $(x,y)$ is constructible by straightedge and compass. I will denote
  by $\mathcal{C}_\mathbb{C} \subseteq \mathbb{C}$ the set of constructible
  complex numbers.
\end{defn}
Summarizing the foregoing discussion, we have proved:
\begin{lem}{}{fields:3.2}
  A point $(x,y)$ is constructible by straightedge and compass if and only
  if $x + iy \in \mathcal{C}_\mathbb{C}$, if and only if $x, y \in
  \mathcal{C}_\mathbb{R}$.
\end{lem}
The next obvious question is what kind of structure these sets of
constructible numbers carry, and this prompts us to look at a few more
basic straightedge-and-compass constructions.
\begin{lem}{}{fields:3.3}
  The subset $\mathcal{C}_\mathbb{R} \subseteq \mathbb{R}$ of constructible
  numbers is a subfield of $\mathbb{R}$. Likewise, $\mathcal{C}_\mathbb{C}$
  is a subfield of $\mathbb{C}$, and in fact $\mathcal{C}_\mathbb{C} =
  \mathcal{C}_\mathbb{R}(i)$.
\end{lem}
\begin{proof}
  The set $\mathcal{C}_\mathbb{R} \subseteq \mathbb{R}$ is nonempty, so in
  order to show it is a field, we only need to show that it is closed with
  respect to subtraction and division by a nonzero constructible number
  (cf.\ Proposition~II.6.2).

  The reader will check that $\mathcal{C}_\mathbb{R}$ is closed under
  subtraction (Exercise~3.2). To see that $\mathcal{C}_\mathbb{R}$ is
  closed under division, let $a, b \in \mathcal{C}_\mathbb{R}$, with $b
  \neq 0$. Since $A = (0, a)$ and $B = (b, 0)$ are constructible by
  hypothesis, we can construct $C$ as the $y$-intercept of the line
  through $P = (1, 0)$ and parallel to the line through $A$ and
  $B$.\footnote{I am assuming $a > 0$ and $b > 1$; the reader will check
  that other possibilities lead to the same conclusion.} Since the
  triangles $AOB$ and $COP$ are similar, we see that $C = (0, a/b)$; it
  follows that $a/b$ is constructible.

  The fact that $\mathcal{C}_\mathbb{C}$ is a subfield of $\mathbb{C}$ is
  an immediate consequence of the fact that $\mathcal{C}_\mathbb{R}$ is a
  field, and the proof of this is left to the enjoyment of the reader
  (Exercise~3.7). The fact that $\mathcal{C}_\mathbb{C} =
  \mathcal{C}_\mathbb{R}(i)$ is a restatement of the fact that $x + iy \in
  \mathcal{C}_\mathbb{C}$ if and only if $x$ and $y$ are in
  $\mathcal{C}_\mathbb{R}$.
\end{proof}
We may view $\mathcal{C}_\mathbb{R}$ and $\mathcal{C}_\mathbb{C}$ as
extensions of $\Q$, drawing a bridge between constructibility by
straightedge and compass and field theory: we will be able to understand
constructibility of geometric figures if we can understand the field
extensions $\Q \subseteq \mathcal{C}_\mathbb{R} \subseteq
\mathcal{C}_\mathbb{C}$. Happily, we can understand these extensions!

\subsection{Constructible numbers and quadratic extensions}
\label{subsec:fields:constructible-numbers-and-quadratic-extensions}
Our goal is to prove the following amazingly explicit description of
$\mathcal{C}_\mathbb{R}$ (immediately implying one for
$\mathcal{C}_\mathbb{C}$).
\begin{thm}{}{fields:3.4}
  Let $\gamma \in \mathbb{R}$. Then $\gamma \in \mathcal{C}_\mathbb{R}$ if
  and only if there exist real numbers $\delta_1, \dots, \delta_k$ such
  that for all $j = 1, \dots, k$,
  \[
    [\Q(\delta_1, \dots, \delta_j) : \Q(\delta_1, \dots, \delta_{j-1})]
    = 2,
  \]
  and $\gamma \in \Q(\delta_1, \dots, \delta_k)$.

  In other words, $\gamma \in \mathbb{R}$ is constructible if and only if
  it can be placed in the top field of a sequence of (real) quadratic
  extensions over $\Q$. Since $\mathcal{C}_\mathbb{C} =
  \mathcal{C}_\mathbb{R}(i)$, the same statement holds for constructible
  complex numbers, including $i$ in the list of $\delta_j$ (or simply
  allowing the $\delta_j$ to be complex numbers; cf.\ Exercise~3.9).
\end{thm}
The proof of this theorem is as explicit as one can possibly hope. From a
given straightedge-and-compass construction of a point $(x,y)$ one can
obtain an explicit sequence $\delta_1, \dots, \delta_k$ as in the
statement, such that $x$ and $y$ belong to the extension $\Q(\delta_1,
\dots, \delta_k)$; conversely, from any element $\gamma$ of any such
extension one can obtain an explicit construction of a point $(\gamma, 0)$
by straightedge and compass.
\begin{proof}
  Let's first argue in the `geometry to algebra' direction. A
  configuration of points, lines, and circles obtained by a
  straightedge-and-compass construction may be described by the coordinates
  of the points and the equations of the lines and circles. Suppose that at
  one stage in a given construction all coordinates of all points and all
  coefficients in the equations of lines and circles belong to a field $F$;
  I will say that the configuration is \emph{defined over $F$}. Then I
  claim that for every object constructed at the next stage, there exists a
  number $\delta \in \mathbb{R}$, of degree at most $2$ over $F$, such that
  the new configuration is defined over $F(\delta)$. The `only if' part of
  the theorem follows by induction on the number of steps in the
  construction, since at the beginning the configuration (that is, the pair
  of points $O = (0,0)$, $P = (1,0)$) is defined over $\Q$.
  Verifying my claim amounts to verifying it for the basic operations
  defining straightedge-and-compass constructions. The reader will check
  (Exercise~3.5) that the point of intersection of two lines defined over
  $F$ has coordinates in $F$, and that lines and circles determined by
  points with coordinates in $F$ are defined over $F$; so $\delta = 1$
  works in all these cases.

  For the intersection of a line $\ell$ and a circle $C$, assume that
  $\ell$ is not parallel to the $y$-axis (the argument is entirely
  analogous otherwise) and that it does meet $C$; let $y = mx + r$ be the
  equation of $\ell$ and let $x^2 + y^2 + ax + by + c = 0$ be the equation
  of $C$. We are assuming that $a, b, c, m, r \in F$. Then the
  $x$-coordinates of the points of intersection of $\ell$ and $C$ are the
  solutions of the equation
  \[
    x^2 + (mx + r)^2 + ax + b(mx + r) + c = 0.
  \]
  The `quadratic formula' shows that these coordinates belong to the field
  $F(\sqrt{D})$, where $D$ is the discriminant of this polynomial:
  explicitly,
  \[
    D = (2mr + bm + a)^2 - 4(m^2 + 1)(r^2 + br + c),
  \]
  but this is unimportant. What is important is that $D \in F$; hence
  $\delta = \sqrt{D}$ satisfies our requirement.
  For the intersection of two (distinct) circles defined over $F$, nothing
  new is needed: if
  \[
    \begin{cases}
      x^2 + y^2 + a_1 x + b_1 y + c_1 = 0, \\
      x^2 + y^2 + a_2 x + b_2 y + c_2 = 0
    \end{cases}
  \]
  are two circles, subtracting the two equations shows that their points of
  intersection coincide with the points of intersection of a circle and a
  line:
  \[
    \begin{cases}
      x^2 + y^2 + a_1 x + b_1 y + c_1 = 0, \\
      (a_1 - a_2)x + (b_1 - b_2)y + (c_1 - c_2) = 0,
    \end{cases}
  \]
  with the same conclusion as in the previous case. This completes the
  verification of the `only if' part of the theorem.
  To prove that every element of an extension as stated is constructible,
  again argue by induction: it suffices to show that if (i) $\delta \in
  \mathbb{R}$, (ii) all elements of $F$ are constructible, and (iii)
  $r = \delta^2 \in F$, then $\delta$ is constructible (note that in order
  to construct an element of degree $2$ over $F$, it suffices to construct
  the square root of the discriminant of its minimal polynomial). Therefore,
  all we have to show is that we can `take square roots' by a
  straightedge-and-compass construction.

  If $A = (r, 0)$ is constructible, so is $B = (-r, 0)$; $C$ is the
  midpoint of the segment $BP$ (midpoints are constructible, Exercise~3.1);
  the circle with center $C$ and containing $P$ intersects the positive
  $y$-axis at a point $Q = (0, \delta)$, and elementary geometry shows that
  $\delta^2 = r$. Therefore $\delta$ is constructible, concluding the proof
  of the theorem.
\end{proof}
For example, one can construct an angle of $3^\circ$: allegedly
\[
  \cos 3^\circ
  = \tfrac{1}{8}\bigl[(\sqrt{3}+1)\sqrt{5+\sqrt{5}}
    + (\sqrt{6}-\sqrt{2})(\sqrt{5}-1)\bigr],
\]
and this expression shows that
\[
  \cos 3^\circ \in
  \Q(\sqrt{2},\sqrt{3},\sqrt{5})\bigl(\sqrt{5+\sqrt{5}}\bigr);
\]
that is (by \cref{thm:fields:3.4}), $\cos 3^\circ$ is constructible. Of
course once $A = (\cos\theta, 0)$ is constructed, so is the angle
$\theta$.
Here is another example of a `constructive' application of
\cref{thm:fields:3.4}:
\begin{exam}{}{fields:exam:3.5}
  Regular pentagons are constructible. Indeed, it suffices to construct the
  point $A = (\cos(2\pi/5), 0)$, and it so happens that $\gamma =
  \cos(2\pi/5)$ satisfies
  \[
    4\gamma^2 + 2\gamma - 1 = 0 \tag{$*$}
  \]
  (in fact, $\gamma$ is half of the inverse of the golden ratio: $\gamma =
  \frac{\sqrt{5}-1}{4}$). It follows that $4\gamma \in \Q(\sqrt{5})$, and
  hence it is constructible by \cref{thm:fields:3.4}. In fact, the proof
  shows how to construct $\sqrt{5}$, so the reader should now have no
  difficulty producing a straightedge-and-compass construction of a regular
  pentagon (Exercise~3.6).

  We will come back to constructions of regular polygons once we have
  studied field theory a little more thoroughly (cf.\ \S{}7.2). By the way,
  how does one figure out that $\gamma = \cos(2\pi/5)$ should satisfy the
  identity $(*)$? This is easy modulo some complex number arithmetic; see
  Exercise~3.11.
\end{exam}

\subsection{Famous impossibilities}
\label{subsec:fields:famous-impossibilities}
\cref{thm:fields:3.4} easily settles the three problems mentioned in
\S{}3.1, via the following immediate consequence:
\begin{coro}{}{fields:3.6}
  Let $\gamma \in \mathcal{C}_\mathbb{C}$ be a constructible number. Then
  $[\Q(\gamma) : \Q]$ is a power of $2$.
\end{coro}
\begin{proof}
  By \cref{lem:fields:3.3} and \cref{thm:fields:3.4}, there exist
  $\delta_1, \dots, \delta_k \in \mathbb{R}$ such that $\gamma \in
  \Q(\delta_1, \dots, \delta_k, i)$, and each $\delta_j$ has degree $\leq
  2$ over $\Q(\delta_1, \dots, \delta_{j-1})$. Repeated application of
  \cref{prop:fields:1.10} shows that $[\Q(\delta_1, \dots, \delta_k, i) :
  \Q]$ is a power of $2$, and since $\Q \subseteq \Q(\gamma) \subseteq
  \Q(\delta_1, \dots, \delta_k, i)$, the statement follows from
  \cref{coro:fields:1.11}.
\end{proof}
In particular, constructible real numbers must satisfy the same condition.
Consider the problem of trisecting an angle. We know we can construct an
angle of $60^\circ$ (this is a subproduct of the construction of an
equilateral triangle). Constructing an angle of $20^\circ$ is equivalent
(Exercise~3.10) to constructing the complex number $\gamma$ on the unit
circle and with argument $\pi/9$. Now $\gamma^9 = -1$; that is, $\gamma$
satisfies the polynomial
\[
  t^9 + 1 = (t^3 + 1)(t^6 - t^3 + 1).
\]
It does not satisfy $t^3 + 1$; therefore its minimal polynomial over $\Q$
is a factor of $t^6 - t^3 + 1$. However, this polynomial is irreducible
over $\Q$ (substitute $t \mapsto t-1$, and apply Eisenstein's criterion),
so $[\Q(\gamma) : \Q] = 6$. By \cref{coro:fields:3.6}, $\gamma$ is not
constructible. Therefore there exist constructible angles which cannot be
trisected.
Similarly, cubes cannot be doubled because that would imply the
constructibility of $\sqrt[3]{2}$, which has degree $3$ over $\Q$,
contradicting \cref{coro:fields:3.6}. Squaring circles amounts to
constructing $\pi$, and $\pi$ is not even algebraic (although, as I have
mentioned, the proof of this fact is not elementary), so this is also not
possible.
As another example, constructing a regular $7$-gon would amount to
constructing a $7$th (complex) root of $1$, $\zeta$. By definition,
$\zeta$ satisfies
\[
  t^7 - 1 = (t-1)(t^6 + t^5 + \dots + t + 1);
\]
as $\zeta \neq 1$, $\zeta$ must satisfy the cyclotomic polynomial $t^6 +
\dots + 1$. This is irreducible (Example~V.5.19); hence again we find that
$\zeta$ has degree $6$ over $\Q$, and \cref{coro:fields:3.6} implies that
the regular $7$-gon cannot be constructed with straightedge and compass.
Of course $7$ is not too special: if $p$ is a positive prime integer, the
cyclotomic polynomial of degree $p-1$ is irreducible (Example~V.5.19
again); hence $[\Q(\zeta_p) : \Q] = p - 1$, where $\zeta_p$ is the complex
$p$th root of $1$ with argument $2\pi/p$. Therefore, \cref{coro:fields:3.6}
reveals that if $p$ is prime, then the regular $p$-gon can be constructed
only if $p - 1$ is a power of $2$. This is even more restrictive than it
looks at first, since if $p = 2^k + 1$ is prime, then necessarily $k$ is
itself a power of $2$ (Exercise~3.15). Primes of the form $2^{2^n} + 1$
are called \emph{Fermat primes}:
$3, 5, 17, 257, 65537$
(the `next one', $2^{32} + 1 = 4294967297 = 641 \cdot 6700417$, is not
prime; in fact, no one knows if there are any other Fermat primes, and yet
some people conjecture that there are infinitely many).
These considerations alone do not tell us that if $p$ is a Fermat prime,
then the regular $p$-gon can be constructed with straightedge and compass;
but they do tell us that the next case to consider would be $p = 2^4 + 1 =
17$, and it just so happens that
\[
  \cos\frac{2\pi}{17} = \frac{-1 + \sqrt{17}
    + \sqrt{2(17-\sqrt{17})}
    + 2\sqrt{17 + 3\sqrt{17}
      - \sqrt{2(17-\sqrt{17})}
      - 2\sqrt{2(17+\sqrt{17})}}}{16},
\]
as noted by \name{Gauss} at age 19. Therefore (by \cref{thm:fields:3.4})
the $17$-gon is constructible by straightedge and compass. As I have
announced already, the situation will be further clarified soon, allowing
us to bypass heavy-duty trigonometry.
\subsection*{Exercises}

\begin{exer}{}{fields:exer:3.1}
  Prove that if $A$, $B$ are constructible, then the midpoint of the
  segment $AB$ is also constructible. Prove that if two lines $\ell_1$,
  $\ell_2$ are constructible and not parallel, then the two lines bisecting
  the angles formed by $\ell_1$ and $\ell_2$ are also constructible.
\end{exer}
\begin{exer}{}{fields:exer:3.2}
  Prove that if $a$, $b$ are constructible numbers, then so is $a - b$.
\end{exer}
\begin{exer}{}{fields:exer:3.3}
  Find an explicit straightedge-and-compass construction for the product of
  two real numbers.
\end{exer}
\begin{exer}{}{fields:exer:3.4}
  Show how to square a triangle by straightedge and compass.
\end{exer}
\begin{exer}{}{fields:exer:3.5}
  Let $F$ be a subfield of $\mathbb{R}$.
  \begin{itemize}
    \item Let $A = (x_A, y_A)$, $B = (x_B, y_B)$ be two points in
      $\mathbb{R}^2$, with $x_A, y_A, x_B, y_B \in F$. Prove that the
      line through $A$, $B$ is defined over $F$ (that is, it admits an
      equation with coefficients in $F$).
    \item Prove that the circle with center at $A$ and containing $B$ is
      defined over $F$.
    \item Let $\ell_1$, $\ell_2$ be two distinct, nonparallel lines in
      $\mathbb{R}^2$, defined over $F$, and let $(x, y) = \ell_1 \cap
      \ell_2$. Prove that $x, y \in F$.
  \end{itemize}
\end{exer}
\begin{exer}{}{fields:exer:3.6}
  Devise a way to construct a regular pentagon with straightedge and
  compass.
\end{exer}
\begin{exer}{}{fields:exer:3.7}
  Identify the (real) plane with $\mathbb{C}$, and place $O$, $P$ at $0,
  1 \in \mathbb{C}$. Let $\mathcal{C}_\mathbb{C}$ be the set of all
  constructible points, viewed as a subset of $\mathbb{C}$. Prove that
  $\mathcal{C}_\mathbb{C}$ is a subfield of $\mathbb{C}$.
\end{exer}
\begin{exer}{}{fields:exer:3.8}
  For $\delta \in \mathbb{C}$, $\delta \neq 0$, let $\theta_\delta$ be the
  argument of $\delta$ (that is, the angle formed by the line through $0$
  and $\delta$ with the real axis). Prove that $\delta \in
  \mathcal{C}_\mathbb{C}$ if and only if $|\delta|$, $\cos\theta_\delta$,
  $\sin\theta_\delta$ are all constructible real numbers.
\end{exer}
\begin{exer}{}{fields:exer:3.9}
  Let $\gamma_1, \dots, \gamma_k \in \mathcal{C}_\mathbb{C}$ be
  constructible complex numbers, and let $K$ be the field $\Q(\gamma_1,
  \dots, \gamma_k) \subseteq \mathcal{C}_\mathbb{C}$. Let $\delta$ be a
  complex number such that $[K(\delta) : K] = 2$. Prove that $\delta \in
  \mathcal{C}_\mathbb{C}$.

  Deduce that there are no irreducible polynomials of degree $2$ over
  $\mathcal{C}_\mathbb{C}$ and that $\mathcal{C}_\mathbb{C}$ is the
  smallest subfield of $\mathbb{C}$ with this property.
\end{exer}
\begin{exer}{}{fields:exer:3.10}
  Prove that if two lines in any constructible configuration form an angle
  of $\theta$, then the line through $O$ forming an angle of $\theta$
  clockwise with respect to the line $OP$ is constructible.

  Deduce that an angle $\theta$ is constructible anywhere in the plane if
  and only if $\cos\theta$ is constructible.
\end{exer}
\begin{exer}{}{fields:exer:3.11}
  Verify that $\gamma = \cos(2\pi/5)$ satisfies the relation $(*)$ given in
  \S{}3.2. (If $\sigma = \sin(2\pi/5)$, note that $z = \gamma + i\sigma$
  satisfies $z^5 = 1$.)
\end{exer}
\begin{exer}{}{fields:exer:3.12}
  Prove that the angles of $1^\circ$ and $2^\circ$ are not constructible.
  (Hint: Given what we know at this point, you only need to recall that
  there exist trigonometric formulas for the sum of two angles; the exact
  shape of these formulas is not important.) For what integers $n$ is the
  angle $n^\circ$ constructible?
\end{exer}
\begin{exer}{}{fields:exer:3.13}
  Prove that $\alpha = \theta$ in the following picture: [the angle
  trisection construction with a marked ruler, as in \S{}3.1]. This says
  that angles can be trisected if we allow the use of a `ruler', that is,
  a straightedge with markings (here, the construction works if we can mark
  a fixed distance of $1$ on the ruler). Apparently, this construction was
  known to \name{Archimedes}.
\end{exer}
\begin{exer}{}{fields:exer:3.14}
  Prove that the regular $9$-gon is not constructible.
\end{exer}

\begin{exer}{}{fields:exer:3.15}
  Prove that if $2^k + 1$ is prime, then $k$ is a power of $2$.
\end{exer}

\section{Field extensions, II}
\label{sec:fields:field-extensions-ii}

It is time to continue our survey of different flavors of field
extensions. The keywords here are splitting fields, normal, separable.

\subsection{Splitting fields and normal extensions}
\label{subsec:fields:splitting-fields-and-normal-extensions}
In \S{}2 we have constructed the algebraic closure $\bar{k}$ of any given
field $k$: every polynomial in $k[x]$ factors as a product of linear terms
(that is, `splits') in $\bar{k}[x]$, and $\bar{k}$ is the `smallest'
extension of $k$ satisfying this property.

Here is an analogous, but more modest, requirement: given a subset $\mathcal{F}
\subseteq k[x]$ of polynomials, construct an extension $k \subseteq F$ such
that every polynomial in $\mathcal{F}$ splits as a product of linear terms
over $F$, and require $F$ to be as small as possible with this property. We
then call $F$ the \emph{splitting field} for $\mathcal{F}$. In practice we
will only be interested in the case in which $\mathcal{F}$ is a finite
collection of polynomials $f_1(x), \dots, f_r(x)$; then requiring that
each $f_i(x)$ splits over $F$ is equivalent to requiring that the product
$f_1(x) \cdots f_r(x)$ splits over $F$. That is, for our purposes it is
not restrictive to assume that $\mathcal{F}$ consists of a single polynomial
$f(x) \in k[x]$.
\begin{defn}{}{fields:4.1}
  Let $k$ be a field, and let $f(x) \in k[x]$ be a polynomial of degree
  $d$. The \emph{splitting field} for $f(x)$ over $k$ is an extension $F$
  of $k$ such that
  \[
    f(x) = c \prod_{i=1}^d (x - \alpha_i)
  \]
  splits in $F[x]$, and further $F = k(\alpha_1, \dots, \alpha_d)$ is
  generated over $k$ by the roots of $f(x)$ in $F$.
\end{defn}
Note that it is clear that a splitting field exists: given $f(x) \in k[x]$,
the subfield $F \subseteq \bar{k}$ generated over $k$ by the roots of
$f(x)$ in $\bar{k}$ satisfies the requirements in \cref{defn:fields:4.1}.
But I have written \emph{the} splitting field, and this is justified by the
uniqueness part of the following basic observation, which also evaluates
`how big' this extension can be.
\begin{lem}{}{fields:4.2}
  Let $k$ be a field, and let $f(x) \in k[x]$. Then the splitting field
  $F$ for $f(x)$ over $k$ is unique up to isomorphism, and $[F:k] \leq
  (\deg f)!$.

  In fact, if $\iota : k' \to k$ is any isomorphism of fields and $g(x)
  \in k'[x]$ is such that $f(x) = \iota(g(x))$, then $\iota$ extends to
  an isomorphism of any splitting field of $g(x)$ over $k'$ to any
  splitting field of $f(x)$ over $k$.
\end{lem}
\begin{proof}
  We first construct explicitly a splitting field and obtain the bound on
  the degree mentioned in the statement. Then we prove the second part,
  which implies uniqueness up to isomorphism.

  The construction of a splitting field and the given bound on the degree
  are an easy application of our basic simple extension found in
  Proposition~V.5.7. Arguing inductively, assume the splitting field has
  been constructed and the bound has been proved for all fields and all
  polynomials of degree $(\deg f - 1)$. Let $q(t)$ be any irreducible
  factor of $f(t)$ over $k$; then
  \[
    k \subseteq F' := \frac{k[t]}{(q(t))}
  \]
  is an extension of degree $\deg q \leq \deg f$, in which $q(x)$ (and
  hence $f(x)$) has a root (the coset $\alpha$ of $t$) and therefore a
  linear factor $x - \alpha$. The polynomial $h(x) := f(x)/(x - \alpha)
  \in F'[x]$ has degree $(\deg f - 1)$; therefore a splitting field $F$
  exists for $h(x)$, and $[F : F'] \leq (\deg f - 1)!$. The factors of
  $f(x)$ over $F$ are $(x - \alpha)$ and the factors of $h(x)$, which are
  linear; it follows that $F$ is a splitting field for $f(x)$ and
  \[
    [F : k] = [F : F'][F' : k] \leq (\deg f)(\deg f - 1)! = (\deg f)!
  \]
  as stated.
  To prove that isomorphisms $\iota : k' \to k$ extend to isomorphisms of
  splitting fields as stated, let $G$ be a splitting field for $g(x)$ and
  consider the composition $k' \to k \subseteq \bar{k}$. Since the
  extension $k' \subseteq G$ is algebraic, by \cref{lem:fields:2.8} there
  exists a morphism of extensions $\tilde\iota : G \to \bar{k}$ over $k'$.
  Since $G$ is generated over $k'$ by the roots of $g(x)$ in $G$, we have
  \[
    \tilde\iota(G) = k(\alpha_1, \dots, \alpha_d) \subseteq \bar{k},
  \]
  where the $\alpha_i$'s are the roots of $f(x) = \iota(g(x))$ in
  $\bar{k}$. Therefore $L := \tilde\iota(G)$ is independent of the chosen
  isomorphism $\iota$ and splitting field $G$. Applying this observation to
  the given $\iota$ and to the identity $k \to k$ proves the statement (as
  both $G$ and $F$ are isomorphic to $L$).
\end{proof}
\begin{exam}{}{fields:exam:4.3}
  By definition, $\Q(i)$ is the splitting field of $x^2 + 1$ over $\Q$,
  and $\mathbb{C}$ is the splitting field for the same polynomial, over
  $\mathbb{R}$.
\end{exam}
\begin{exam}{}{fields:exam:4.4}
  The splitting field $F$ of $x^8 - 1$ over $\Q$ is generated by $\zeta :=
  e^{2\pi i/8}$: indeed, the roots of $x^8 - 1$ are all the $8$th roots of
  $1$, and all of them are powers of $\zeta$. In fact, $\zeta$ is a root
  of the polynomial $x^4 + 1$, which is irreducible over $\Q$; therefore
  $F = \Q(\zeta)$ is `already' the splitting field of $x^4 + 1$. The
  degree of $F$ over $\Q$ is $[\Q(\zeta) : \Q] = 4$, way less than the
  bounds $8!$, $4!$ obtained in \cref{lem:fields:4.2}.

  To understand this splitting field even better, note that $i = \zeta^2$
  is in $F$, and so is $\sqrt{2} = \zeta + \zeta^7$; thus $F$ contains
  $\Q(i, \sqrt{2})$. Conversely, $\zeta = \frac{\sqrt{2}}{2}(1 + i) \in
  \Q(i, \sqrt{2})$. Therefore, the splitting field of $x^4 + 1$ (a.k.a.\
  the splitting field of $x^8 - 1$) is $\Q(i, \sqrt{2})$. Analyzing $\Q
  \subseteq \Q(i, \sqrt{2})$ (as we did for the extension in
  \cref{exam:fields:exam:1.19}) shows that its group of automorphisms over
  $\Q$ is $\Z/2\Z \times \Z/2\Z$.
\end{exam}
\begin{exam}{}{fields:exam:4.5}
  Variation on the theme: $x^4 - 1$. The situation changes if instead of
  $x^4 + 1$ we consider $x^4 - 1$: this polynomial factors over $\Q$,
  \[
    x^4 - 1 = (x-1)(x+1)(x^2+1),
  \]
  and it follows that the splitting field is the same as for $x^2 + 1$,
  that is, just $\Q(i)$.
\end{exam}
\begin{exam}{}{fields:exam:4.6}
  Variation on the theme: $x^4 + 2$. The overoptimistic reader may now
  hope that the difference between the splitting fields of $x^4 + 1$ vs.\
  $x^4 - 1$ over $\Q$ is just due to the fact that the first polynomial is
  irreducible over $\Q$ and the second is not. This example will nip any
  such guess in the bud. With notation as in \cref{exam:fields:exam:4.4},
  the roots of $x^4 + 2$ are
  \[
    \sqrt[4]{2}\,\zeta,\quad
    \sqrt[4]{2}\,\zeta^3,\quad
    \sqrt[4]{2}\,\zeta^5,\quad
    \sqrt[4]{2}\,\zeta^7.
  \]
  Therefore, with $K = \Q(\sqrt[4]{2}\,\zeta, \sqrt[4]{2}\,\zeta^3,
  \sqrt[4]{2}\,\zeta^5, \sqrt[4]{2}\,\zeta^7)$ the splitting field of
  $x^4 + 2$,
  \[
    K \subseteq
    \Q\bigl(\zeta,\sqrt[4]{2}\bigr)
    = \Q\bigl(i,\sqrt{2},\sqrt[4]{2}\bigr)
    = \Q\bigl(i,\sqrt[4]{2}\bigr).
  \]
  On the other hand, $\sqrt{2} = (\sqrt[4]{2}\,\zeta)^3/(\sqrt[4]{2}\,
  \zeta^3) \in K$; hence $i = (\sqrt[4]{2}\,\zeta)^2/\sqrt{2} \in K$;
  hence $\zeta = \frac{\sqrt{2}}{2}(1+i) \in K$; hence $\sqrt[4]{2} =
  (\sqrt[4]{2}\,\zeta)/\zeta \in K$. Therefore, $\Q(i, \sqrt[4]{2})
  \subseteq K$, and the conclusion is that the splitting field of $x^4 + 2$
  equals $K = \Q(i, \sqrt[4]{2})$. A simple degree computation
  (Exercise~4.3) shows $[K : \Q] = 8$, and in particular the splitting
  field of $x^4 + 2$ is certainly not isomorphic to the splitting field of
  $x^4 + 1$.
\end{exam}
Examples such as these have always left me with the impression that field
theory must be rather mysterious: innocent-looking variations on the
parameters of a problem may cause dramatic changes in the corresponding
field extensions. On the plus side, this hints that field theory can indeed
be a very precise tool in (for example) the study of polynomials.

Splitting fields will play an important role in the rest of the story. They
are even more special than they may appear to be at first: it turns out
that not only do they split the given polynomial, but they also
automatically split any irreducible polynomial which dares touch them with
a root. This makes splitting fields normal extensions:
\begin{defn}{}{fields:4.7}
  A field extension $k \subseteq F$ is \emph{normal} if for every
  irreducible polynomial $f(x) \in k[x]$, $f(x)$ has a root in $F$ if and
  only if $f(x)$ splits as a product of linear factors over $F$.
\end{defn}
\begin{thm}{}{fields:4.8}
  A field extension $k \subseteq F$ is finite and normal if and only if
  $F$ is the splitting field of some polynomial $f(x) \in k[x]$.
\end{thm}
\begin{proof}
  Assume $k \subseteq F$ is finite and normal. Then $F$ is finitely
  generated: $F = k(\alpha_1, \dots, \alpha_r)$ with $\alpha_i$ algebraic
  over $k$. Let $p_i(t)$ be the minimal polynomial of $\alpha_i$ over $k$.
  As $F$ is normal over $k$, each $p_i(t)$ splits completely over $F$, and
  hence so does $f(t) = p_1(t) \cdots p_r(t)$. It follows that $F$ is the
  splitting field of $f(x)$.
  Conversely, assume that $F$ is a splitting field for a polynomial $f(x)
  \in k[x]$, and let $p(x) \in k[x]$ be an irreducible polynomial such
  that $F$ contains a root $\alpha$ of $p(x)$. Viewing $F$ as a subfield
  of the algebraic closure $\bar{k}$, let $\beta \in \bar{k}$ be any other
  root of $p(x)$; we are going to show that $\beta \in F$. This will prove
  that $F$ contains all roots of $p(x)$, implying that $k \subseteq F$ is
  normal, and $k \subseteq F$ is finite by \cref{lem:fields:4.2}.

  By \cref{prop:fields:1.5}, there exists an isomorphism $\iota : k(\alpha)
  \to k(\beta)$ extending the identity on $k$ and sending $\alpha \mapsto
  \beta$. We also consider the subfield $F(\beta) \subseteq \bar{k}$,
  viewed as an extension of $k(\beta)$. Now observe that $F$ may be viewed
  as the splitting field of $f(x)$ over $k(\alpha)$: indeed, it contains
  all the roots of $f(x)$ and is generated over $k$ (and hence over
  $k(\alpha)$) by these roots. By the same token, $F(\beta)$ is the
  splitting field for $f(x)$ over $k(\beta)$. By \cref{lem:fields:4.2},
  $\iota$ extends to an isomorphism $\iota' : F \to F(\beta)$. Since
  $\iota$ restricts to the identity on $k$, $\iota'$ is an isomorphism of
  $k$-vector spaces; in particular, $\dim_k F = \dim_k F(\beta)$ (keep in
  mind that splitting fields are finite extensions).

  Now consider the different $k$-linear map $i : F \to F(\beta)$ simply
  given by the inclusion $k \subseteq F \subseteq F(\beta) \subseteq
  \bar{k}$. Since $F$ and $F(\beta)$ are $k$-vector spaces of the same
  finite dimension, $i$ must also be an isomorphism. In other words,
  $[F(\beta) : F] = 1$; that is, $\beta \in F$ as needed.
\end{proof}
\begin{rem}{}{fields:4.9}
  This argument is somewhat delicate. With notation as in the proof, the
  diagram
  \[
    \begin{tikzcd}
      k(\alpha) \ar[r, hook] \ar[d, "\iota"'] & F \\
      k(\beta)
    \end{tikzcd}
  \]
  is commutative, while the diagram in which both $F$ and $F(\beta)$ appear
  is not commutative if $\beta \neq \alpha$ (since $\iota$ sends $\alpha$
  to $\beta$, while $i$ sends $\alpha$ to $\alpha$). The key point in the
  argument is the observation that if there exists one isomorphism between
  two finite-dimensional vector spaces $V$, $W$, then every injective
  linear map $V \to W$ must be an isomorphism. Finite dimensionality is
  necessary in order to draw this conclusion; cf.\ Exercise~VI.6.5.
\end{rem}
\begin{exam}{}{fields:exam:4.10}
  If a complex root of an irreducible polynomial $p(x) \in \Q[x]$ may be
  expressed as a polynomial in $i$ and $\sqrt[4]{2}$ with rational
  coefficients, then all roots of $p(x)$ may be expressed likewise in terms
  of $i$ and $\sqrt[4]{2}$. Indeed, we have checked
  (\cref{exam:fields:exam:4.6}) that $\Q(i, \sqrt[4]{2})$ is a splitting
  field over $\Q$; hence it is a normal extension of $\Q$.
\end{exam}

\subsection{Separable polynomials}
\label{subsec:fields:separable-polynomials}
Our intuition may lead us to think that if a polynomial factors as a
product of linear factors and we are not `purposely' repeating one of the
factors (as in $(x-1)^2(x-2)$), then these will be distinct. For example,
surely irreducible polynomials necessarily split as products of distinct
factors in an algebraic closure, right? Wrong.
\begin{exam}{}{fields:exam:4.11}
  Let $p$ be a prime, and consider the field $\mathbb{F}_p(t)$ of rational
  functions over $\mathbb{F}_p$. Then the polynomial
  \[
    x^p - t \in \mathbb{F}_p(t)[x]
  \]
  is irreducible: by Eisenstein's criterion it is irreducible in
  $\mathbb{F}_p[t][x]$ (since $(t)$ is prime in $\mathbb{F}_p[t]$), hence
  in $\mathbb{F}_p(t)[x]$ by Proposition~V.4.16. Let $u$ be a root of
  this polynomial in an extension $L$ of $\mathbb{F}_p(t)$ (for example
  $L$ could be the algebraic closure of $\mathbb{F}_p(t)$, or more
  modestly a splitting field for the polynomial). Then (Exercise~4.8)
  \[
    x^p - t = (x - u)^p
  \]
  in $L[x]$; that is, $u$ has multiplicity $p$ as a root of $f(x)$.

  In other words, the minimal polynomial over $\mathbb{F}_p(t)$ of $u \in
  L$ vanishes $p$ times at $u$, and there is nothing to do about this: no
  smaller power of $(x-u)$ than $(x-u)^p = x^p - t$ has coefficients in
  $\mathbb{F}_p(t)$ (a smaller power would give a nontrivial factor of
  $x^p - t$, and $x^p - t$ is irreducible).
\end{exam}
I have always found this example difficult to visualize, because of
intuition developed in characteristic $0$, and as we will see, no such
pathology can occur in characteristic $0$.
\begin{defn}{}{fields:4.12}
  Let $k$ be a field. A polynomial $f(x) \in k[x]$ is \emph{separable} if
  it has no multiple factors over its splitting field; $f(x)$ is
  \emph{inseparable} if it has multiple factors over its splitting field.
\end{defn}
Thus, the polynomial $x^p - t$ in \cref{exam:fields:exam:4.11} is
inseparable. I suppose the terminology reflects the fact that we cannot
`separate' its roots: they come clumped together like quarks in a proton,
and we cannot take them apart.

Of course \cref{defn:fields:4.12} does not depend on the chosen splitting
field, since these are unique up to isomorphism (\cref{lem:fields:4.2}).
In fact, to detect separability, we can use any field in which the
polynomial splits as a product of linear factors (by Exercise~4.1). The
first, somewhat surprising, observation about separability is that we can
in fact detect it without leaving the field of coefficients of $f(x)$. This
fact uses a notion borrowed from calculus: for a polynomial
\[
  f(x) = a_0 + a_1 x + a_2 x^2 + \dots + a_n x^n,
\]
we denote by $f'(x)$ the `derivative'
\[
  f'(x) = a_1 + 2a_2 x + \dots + na_n x^{n-1}.
\]
Of course this is a purely formal operation; no limiting process is at work
over arbitrary fields. Still, all expected properties of derivatives hold,
as the reader should check: for example, $(fg)' = f'g + fg'$ as
usual.
\begin{lem}{}{fields:4.13}
  Let $k$ be a field, and let $f(x) \in k[x]$. Then $f(x)$ is separable
  if and only if $f(x)$ and $f'(x)$ are relatively prime.
\end{lem}
By definition, $f(x)$ and $f'(x)$ are relatively prime precisely when
$\gcd(f(x), f'(x)) = 1$. Note that if $k \subseteq F$, the gcd of $f(x)$
and $f'(x)$ is the same whether it is considered in $k[x]$ or in $F[x]$:
for example because it can be computed by applying the Euclidean algorithm
(\S{}V.2.4), and this proceeds in exactly the same way whether it is
performed in $k[x]$ or in $F[x]$.
\begin{proof}
  First assume that $f(x)$ is not separable. Then $f(x)$ has a multiple
  root in a splitting field $F$; that is, $f(x) = (x - \alpha)^m g(x)$ for
  some $\alpha \in F$, $g(x) \in F[x]$, and $m \geq 2$. Therefore
  \[
    f'(x) = m(x-\alpha)^{m-1}g(x) + (x-\alpha)^m g'(x),
  \]
  and it follows that $(x - \alpha)$ is a common factor of $f(x)$ and
  $f'(x)$ in $F[x]$. Therefore $\gcd(f(x), f'(x)) \neq 1$ in $F[x]$;
  hence $\gcd(f(x), f'(x)) \neq 1$ in $k[x]$; that is, $f(x)$, $f'(x)$
  are not relatively prime.

  Conversely, assume $\gcd(f(x), f'(x)) \neq 1$, so $f(x)$ and $f'(x)$
  have a common irreducible factor $(x - \alpha)$ in the algebraic closure
  $\bar{k}$ of $k$. Write $f(x) = (x - \alpha)h(x)$; we have
  \[
    f'(x) = h(x) + (x-\alpha)h'(x),
  \]
  and it follows that $x - \alpha$ divides $h(x)$ since it divides
  $f'(x)$. But then $(x-\alpha)^2 \mid f(x)$; hence $f(x)$ is inseparable.
\end{proof}
For example, the polynomial $x^p - t$ of \cref{exam:fields:exam:4.11} may
be seen to be inseparable without invoking splitting fields: the derivative
of $x^p - t$ equals $px^{p-1} = 0$ in characteristic $p$, and $\gcd(x^p -
t, 0) = x^p - t \neq 1$. This example captures one of the key features of
inseparability:
\begin{lem}{}{fields:4.14}
  Let $k$ be a field, and let $f(x) \in k[x]$ be an inseparable
  irreducible polynomial. Then $f'(x) = 0$.
\end{lem}
\begin{proof}
  Since $f(x)$ is inseparable, $f(x)$ and $f'(x)$ have a common
  irreducible factor $q(x)$ by \cref{lem:fields:4.13}; but as $f(x)$ is
  itself irreducible, $q(x)$ must be an associate of $f(x)$, and in
  particular its degree is larger than the degree of $f'(x)$ if $f'(x)
  \neq 0$. As $q(x) \mid f'(x)$, the only option is that $f'(x) = 0$.
\end{proof}
This already tells us that irreducible polynomials are necessarily
separable in characteristic $0$: in characteristic $0$, the derivative of
a nonconstant polynomial clearly cannot vanish. In fact,
\cref{lem:fields:4.14} gives us a precise picture of what inseparable,
irreducible polynomials must look like. If $f(x) = \sum_{i=0}^n a_i x^i$
is irreducible and inseparable, then the characteristic of the field must
be a positive prime $p$, and by \cref{lem:fields:4.14} we must have
\[
  f'(x) = \sum_{i=0}^n i a_i x^{i-1} = 0;
\]
that is, $ia_i = 0$ for all $i$. Now $ia_i = 0$ is automatic if $i$ is a
multiple of $p$, and it implies $a_i = 0$ for all indices $i$ which are
not multiples of $p$. Therefore, the only nonvanishing coefficients in
$f(x)$ must be those corresponding to indices which are multiples of $p$:
\[
  f(x) = a_0 + a_p x^p + a_{2p} x^{2p} + \cdots;
\]
therefore, $f(x)$ must in fact be a polynomial in $x^p$. (The reader will
refine this picture further by working out Exercise~4.13 and following.)
This leads us to a precise result linking separability and a flavor of
fields that we have not yet run into.
\begin{defn}{}{fields:4.15}
  Let $k$ be a field of characteristic $p > 0$. The \emph{Frobenius
  homomorphism} is the map $k \to k$ defined by $x \mapsto x^p$.
\end{defn}
The Frobenius homomorphism may not look like a homomorphism of rings, but
it is (Exercise~4.8). It must be injective, as is every nontrivial ring
homomorphism from a field; but it is not necessarily surjective.
\begin{defn}{}{fields:4.16}
  A field $k$ is \emph{perfect} if $\operatorname{char} k = 0$ or if
  $\operatorname{char} k > 0$ and the Frobenius homomorphism is surjective.
\end{defn}
\begin{prop}{}{fields:4.17}
  Let $k$ be a field. Then $k$ is perfect if and only if all irreducible
  polynomials in $k[x]$ are separable.
\end{prop}
\begin{proof}
  I will prove that irreducible polynomials over a perfect field are
  separable, leaving the other implication to the reader (Exercise~4.12).

  We have already noted that irreducible polynomials are separable over
  fields of characteristic zero. In positive characteristic $p$, we have
  observed that an inseparable irreducible polynomial must be of the form
  \[
    f(x) = \sum_{i=0}^m a_i \cdot (x^p)^i.
  \]
  Since Frobenius is surjective, there exist $b_i$ such that $b_i^p = a_i$.
  Thus
  \[
    f(x)
    = \sum_{i=0}^m b_i^p (x^p)^i
    = \sum_{i=0}^m (b_i x^i)^p
    = \left(\sum_{i=0}^m b_i x^i\right)^p
    = g(x)^p,
  \]
  where $g(x) = \sum b_i x^i$, keeping in mind that the Frobenius map is a
  homomorphism. But this contradicts the irreducibility of $f(x)$, so there
  is no such polynomial.
\end{proof}
\begin{coro}{}{fields:4.18}
  Finite fields are perfect. Therefore, over finite fields, irreducible
  polynomials are separable.
\end{coro}
\begin{proof}
  The Frobenius map is injective (because it is a homomorphism of fields),
  so it is surjective over finite fields, by the pigeon-hole principle.
  Therefore finite fields are perfect, and the second part of the statement
  follows from \cref{prop:fields:4.17}.
\end{proof}
The reader now sees why I have chosen the somewhat unusual field
$\mathbb{F}_p(t)$ in Example~4.11: we need a field of positive
characteristic, but no example of irreducible inseparable polynomial can be
found over $\mathbb{F}_p$, by \cref{coro:fields:4.18}; to concoct an
example, we need to enlarge the field so that it is infinite and to throw
in an element (that is, $t$) for which there is no $p$-th root, that is,
which is not in the image of Frobenius.

\subsection{Separable extensions and embeddings in algebraic closures}
\label{subsec:fields:separable-extensions-and-embeddings-in-algebraic-closures}
The terminology examined in the previous section extends to the language of
field extensions. If $k \subseteq F$ is an extension and $\alpha \in F$ is
algebraic over $k$, we say that $\alpha$ is \emph{separable over $k$} if
the minimal polynomial of $\alpha$ over $k$ is separable; $\alpha$ is
\emph{inseparable} otherwise.
\begin{defn}{}{fields:4.19}
  An algebraic field extension $k \subseteq F$ is \emph{separable} if every
  $\alpha \in F$ is separable over $k$.
\end{defn}
With this terminology, \cref{prop:fields:4.17} may be restated in the
following more impressive form:
\begin{prop}{}{fields:4.20}
  A field $k$ is perfect if and only if every algebraic extension of $k$ is
  separable.
\end{prop}
In particular, algebraic extensions of $\Q$ (or any field of characteristic
zero) and of every finite field are necessarily separable.
The separability condition is exceedingly convenient, and we will
essentially adopt it henceforth: all extensions we will seriously consider
will be separable. One convenient feature of separable extensions is the
following alternative description of the separability condition.

We have seen (\cref{lem:fields:2.8}) that every algebraic extension
$k \subseteq F$ may be embedded in an algebraic closure $k \subseteq \bar{k}$,
and I pointed out that this can in general be done in many different ways.
If finite, the number of different homomorphisms $F \to \bar{k}$ extending
the identity on $k$ is denoted by $[F : k]_s$.
\begin{defn}{}{fields:4.21}
  The number $[F : k]_s$ is the \emph{separable degree} of $F$ over $k$.
\end{defn}
It is clear that this number is independent of the chosen algebraic closure
$k \subseteq \bar{k}$. What does it have to do with separability?
\begin{lem}{}{fields:4.22}
  Let $k \subseteq k(\alpha)$ be a simple algebraic extension. Then
  $[k(\alpha) : k]_s$ equals the number of distinct roots in $\bar{k}$ of
  the minimal polynomial of $\alpha$. In particular,
  $[k(\alpha) : k]_s \leq [k(\alpha) : k]$, with equality if and only if
  $\alpha$ is separable over $k$.
\end{lem}
\begin{proof}
  The proof is essentially (and not by coincidence) a rehash of the proof
  of \cref{coro:fields:1.7}. Associate with each $\iota : k(\alpha) \to
  \bar{k}$ extending $\mathrm{id}_k$ the image $\iota(\alpha)$, which must
  be a root of the minimal polynomial of $\alpha$. This correspondence is
  injective, since $\iota(\alpha)$ determines $\iota$ (as $\iota$ extends
  the identity on $k$). To see it is surjective, let $\beta \in \bar{k}$ be
  any other root, and consider the extension $k(\beta) \subseteq \bar{k}$;
  by \cref{prop:fields:1.5} there is an isomorphism $k(\alpha) \to k(\beta)$
  sending $\alpha$ to $\beta$, and composing with the embedding
  $k(\beta) \subseteq \bar{k}$ defines the corresponding $\iota$, as
  needed.
\end{proof}
Thus, the `separable degree' does detect separability, for simple
extensions. Further, it is multiplicative over successive extensions:
\begin{lem}{}{fields:4.23}
  Let $k \subseteq E \subseteq F$ be algebraic extensions. Then $[F : k]_s$
  is finite if and only if both $[F : E]_s$, $[E : k]_s$ are finite, and
  in this case
  \[
    [F : k]_s = [F : E]_s [E : k]_s.
  \]
\end{lem}
\begin{proof}
  Different embeddings of $E$ into $\bar{k}$ extend to different embeddings
  of $F$ into $\bar{k}$, by \cref{lem:fields:2.8}; and embeddings of $F$
  into $\bar{k}$ extending the identity on $E$ extend a fortiori the
  identity on $k$. Therefore, if any of $[F : E]_s$, $[E : k]_s$ is
  infinite, then so is $[F : k]_s$.

  For the converse implication, it suffices to prove the stated degree
  formula. But every embedding $F \to \bar{k}$ extending the identity on
  $k$ may be obtained in two steps: first extend the identity to an
  embedding $E \to \bar{k}$, which can be done in $[E : k]_s$ ways; and
  then extend the chosen embedding $E \to \bar{k}$ to an embedding
  $F \to \bar{k}$, which can be done in $[F : E]_s$ ways. There are
  precisely $[F : E]_s [E : k]_s$ ways to do this, as stated.
\end{proof}
\cref{lem:fields:4.22,lem:fields:4.23} allow us to recast separability of
finite extensions entirely in the light of counting embeddings in an
algebraic closure:
\begin{prop}{}{fields:4.24}
  Let $k \subseteq F$ be a finite extension. Then $[F : k]_s \leq [F : k]$,
  and the following are equivalent:
  \begin{enumerate}[label=(\roman*)]
    \item $F = k(\alpha_1, \dots, \alpha_r)$, where each $\alpha_i$ is
      separable over $k$;
    \item $k \subseteq F$ is separable;
    \item $[F : k]_s = [F : k]$.
  \end{enumerate}
\end{prop}
\begin{proof}
  Since $F$ is finite over $k$, it is finitely generated. Let
  $F = k(\alpha_1, \dots, \alpha_r)$. Then using \cref{lem:fields:4.23},
  \cref{lem:fields:4.22}, and \cref{prop:fields:1.10},
  \begin{align*}
    [F : k]_s
    &= [k(\alpha_1, \dots, \alpha_{r-1})(\alpha_r)
      : k(\alpha_1, \dots, \alpha_{r-1})]_s
      \cdots [k(\alpha_1) : k]_s \\
    &\leq [k(\alpha_1, \dots, \alpha_{r-1})(\alpha_r)
      : k(\alpha_1, \dots, \alpha_{r-1})]
      \cdots [k(\alpha_1) : k] \\
    &= [F : k].
  \end{align*}
  This proves the stated inequality.

  (i) $\Rightarrow$ (iii): If each $\alpha_i$ is separable over $k$, then
  it is separable over the field $k(\alpha_1, \dots, \alpha_{i-1})$
  (Exercise~4.15), so the inequality is an equality by \cref{lem:fields:4.22}.

  (iii) $\Rightarrow$ (ii): Assume $[F : k]_s = [F : k]$, and let
  $\alpha \in F$. We have $k \subseteq k(\alpha) \subseteq F$; hence by
  \cref{lem:fields:4.23}
  \[
    [F : k(\alpha)]_s [k(\alpha) : k]_s = [F : k]_s = [F : k]
    = [F : k(\alpha)][k(\alpha) : k].
  \]
  Since both separable degrees are less than or equal to their plain
  counterparts, the equality implies $[k(\alpha) : k]_s = [k(\alpha) : k]$,
  proving that $\alpha$ is separable, by \cref{lem:fields:4.22}. Therefore
  the extension $k \subseteq F$ is separable according to
  \cref{defn:fields:4.19}.

  (ii) $\Rightarrow$ (i) is immediate from \cref{defn:fields:4.19}, since
  finite extensions are finitely generated.
\end{proof}
For example, if $\alpha$ is separable over $k$, then every
$\beta \in k(\alpha)$ is separable. This would seem rather mysterious from
the definition alone: why should the fact that the minimal polynomial of
$\alpha$ has distinct roots in $\bar{k}$ imply the same for the minimal
polynomial of $\beta$? It does, by virtue of \cref{prop:fields:4.24}.
\begin{rem}{}{fields:4.25}
  While we have only proved $[F : k]_s \leq [F : k]$, one can in fact
  prove that $[F : k]_s$ divides $[F : k]$: in fact, $[F : k]_s$ is the
  degree of a certain intermediate field $F_{\mathrm{sep}}$ over $k$. The
  diligent reader will prove this in Exercise~4.18.
\end{rem}

\subsection*{Exercises}

\begin{exer}{}{fields:exer:4.1}
  Let $k$ be a field, $f(x) \in k[x]$, and let $F$ be the splitting field
  for $f(x)$ over $k$. Let $k \subseteq K$ be an extension such that
  $f(x)$ splits as a product of linear factors over $K$. Prove that there
  is a homomorphism $F \to K$ extending the identity on $k$. [$\S$4.2]
\end{exer}
\begin{exer}{}{fields:exer:4.2}
  Describe the splitting field of $x^6 + x^3 + 1$ over $\Q$. Do the same
  for $x^4 + 4$.
\end{exer}
\begin{exer}{}{fields:exer:4.3}
  Find the order of the automorphism group of the splitting field of
  $x^4 + 2$ over $\Q$ (cf.\ Example~4.6). [$\S$4.1]
\end{exer}
\begin{exer}{}{fields:exer:4.4}
  Prove that the field $\Q(\sqrt{2})$ is not the splitting field of any
  polynomial over $\Q$.
\end{exer}
\begin{exer}{}{fields:exer:4.5}
  Let $F$ be a splitting field for a polynomial $f(x) \in k[x]$, and let
  $g(x) \in k[x]$ be a factor of $f(x)$. Prove that $F$ contains a unique
  copy of the splitting field of $g(x)$. [$\S$5.1]
\end{exer}
\begin{exer}{}{fields:exer:4.6}
  Let $k \subseteq F_1$, $k \subseteq F_2$ be two finite extensions, viewed
  as embedded in the algebraic closure $\bar{k}$ of $k$. Assume that $F_1$
  and $F_2$ are splitting fields of polynomials in $k[x]$. Prove that the
  intersection $F_1 \cap F_2$ and the composite $F_1 F_2$ (the smallest
  subfield of $\bar{k}$ containing both $F_1$ and $F_2$) are both also
  splitting fields over $k$. (\cref{thm:fields:4.8} is likely going to be
  helpful.)
\end{exer}
\begin{exer}{}{fields:exer:4.7}
  Let $k \subseteq F = k(\alpha)$ be a simple algebraic extension. Prove
  that $F$ is normal over $k$ if and only if for every algebraic extension
  $F \subseteq K$ and every $\sigma \in \mathrm{Aut}_k(K)$, $\sigma(F) =
  F$. [$\S$6.1]
\end{exer}
\begin{exer}{}{fields:exer:4.8}
  Let $p$ be a prime, and let $k$ be a field of characteristic $p$. For
  $a, b \in k$, prove that\footnote{This is sometimes referred to as the
  \emph{freshman's dream}, for painful reasons that are likely all too
  familiar to the reader.} $(a + b)^p = a^p + b^p$. [$\S$4.2, $\S$5.1,
  $\S$5.2]
\end{exer}
\begin{exer}{}{fields:exer:4.9}
  Using the notion of `derivative' given in $\S$4.2, prove that
  $(fg)' = f'g + fg'$ for all polynomials $f$, $g$.
\end{exer}
\begin{exer}{}{fields:exer:4.10}
  Let $k \subseteq F$ be a finite extension in characteristic $p > 0$.
  Assume that $p$ does not divide $[F : k]$. Prove that $k \subseteq F$
  is separable.
\end{exer}
\begin{exer}{}{fields:exer:4.11}
  Let $p$ be a prime integer. Prove that the Frobenius homomorphism on
  $\mathbb{F}_p$ is the identity. (Hint: Fermat.) [$\S$5.1]
\end{exer}
\begin{exer}{}{fields:exer:4.12}
  Let $k$ be a field, and assume that $k$ is not perfect. Prove that there
  are inseparable irreducible polynomials in $k[x]$. (If $\operatorname{char}
  k = p$ and $u \in k$, how many roots does $x^p - u$ have in $\bar{k}$?)
  [$\S$4.2]
\end{exer}
\begin{exer}{}{fields:exer:4.13}
  Let $k$ be a field of positive characteristic $p$, and let $f(x)$ be an
  irreducible polynomial. Prove that there exist an integer $d$ and a
  separable irreducible polynomial $f_{\mathrm{sep}}(x)$ such that
  $f(x) = f_{\mathrm{sep}}(x^{p^d})$. The number $p^d$ is called the
  \emph{inseparable degree} of $f(x)$. If $f(x)$ is the minimal polynomial
  of an algebraic element $\alpha$, the inseparable degree of $\alpha$ is
  defined to be the inseparable degree of $f(x)$. Prove that $\alpha$ is
  inseparable if and only if its inseparable degree is $\geq p$.

  The picture to keep in mind is as follows: the roots of the minimal
  polynomial $f(x)$ of $\alpha$ are distributed into $\deg f_{\mathrm{sep}}$
  `clumps', each collecting a number of coincident roots equal to the
  inseparable degree of $\alpha$. We say that $\alpha$ is `purely
  inseparable' if there is only one clump, that is, if all roots of $f(x)$
  coincide (see Exercise~4.14). [$\S$4.2, 4.14, 4.18]
\end{exer}
\begin{exer}{}{fields:exer:4.14}
  Let $k \subseteq F$ be an algebraic extension, in positive characteristic
  $p$. An element $\alpha \in F$ is \emph{purely inseparable} over $k$ if
  $\alpha^{p^d} \in k$ for some\footnote{Note the slightly annoying clash
  of notation: elements of $k$ are not inseparable, yet they are purely
  inseparable according to this definition.} $d \geq 0$. The extension is
  defined to be \emph{purely inseparable} if every $\alpha \in F$ is purely
  inseparable over $k$.

  Prove that $\alpha$ is purely inseparable if and only if
  $[k(\alpha) : k]_s = 1$, if and only if its degree equals its
  inseparability degree (Exercise~4.13). [4.13, 4.17]
\end{exer}
\begin{exer}{}{fields:exer:4.15}
  Let $k \subseteq F$ be an algebraic extension, and let $\alpha \in F$ be
  separable over $k$. For every intermediate field $k \subseteq E \subseteq
  F$, prove that $\alpha$ is separable over $E$. [$\S$4.3]
\end{exer}
\begin{exer}{}{fields:exer:4.16}
  Let $k \subseteq E \subseteq F$ be algebraic field extensions, and assume
  that $k \subseteq E$ is separable. Prove that if $\alpha \in F$ is
  separable over $E$, then $k \subseteq E(\alpha)$ is a separable
  extension. (Reduce to the case of finite extensions.)

  Deduce that the set of elements of $F$ which are separable over $k$ form
  an intermediate field $F_{\mathrm{sep}}$, such that every element
  $\alpha \in F$, $\alpha \notin F_{\mathrm{sep}}$ is inseparable over
  $F_{\mathrm{sep}}$. For $F = \bar{k}$, $k_{\mathrm{sep}}$ is called the
  \emph{separable closure} of $k$. [4.17, 4.18]
\end{exer}
\begin{exer}{}{fields:exer:4.17}
  Let $k \subseteq F$ be an algebraic extension, in positive characteristic.
  With notation as in Exercises~4.14 and~4.16, prove that the extension
  $F_{\mathrm{sep}} \subseteq F$ is purely inseparable. Prove that an
  extension $k \subseteq F$ is purely inseparable if and only if
  $F_{\mathrm{sep}} = k$. [4.18]
\end{exer}
\begin{exer}{}{fields:exer:4.18}
  Let $k \subseteq F$ be a finite extension, in positive characteristic.
  Define the \emph{inseparable degree} $[F : k]_i$ to be the quotient
  $[F : k] / [F : k]_s$.
  \begin{itemize}
    \item Prove that $[k(\alpha) : k]_i$ equals the inseparable degree of
      $\alpha$, as defined in Exercise~4.13.
    \item Prove that the inseparable degree is multiplicative: if
      $k \subseteq E \subseteq F$ are finite extensions, then
      $[F : k]_i = [F : E]_i [E : k]_i$.
    \item Prove that a finite extension is purely inseparable if and only
      if its inseparable degree equals its degree.
    \item With notation as in Exercise~4.16, prove that
      $[F : k]_s = [F_{\mathrm{sep}} : k]$ and
      $[F : k]_i = [F : F_{\mathrm{sep}}]$. (Use Exercise~4.17.)
  \end{itemize}
  In particular, $[F : k]_s$ divides $[F : k]$; the inseparable degree
  $[F : k]_i$ is an integer. [$\S$4.3]
\end{exer}
\begin{exer}{}{fields:exer:4.19}
  Let $k \subseteq F$ be a finite separable extension, and let
  $\iota_1, \dots, \iota_d$ be the distinct embeddings of $F$ in $\bar{k}$
  extending $\mathrm{id}_k$. For $\alpha \in F$, prove that the norm
  $N_{k \subseteq F}(\alpha)$ (cf.\ Exercise~1.12) equals
  $\prod_{i=1}^d \iota_i(\alpha)$ and its trace $\mathrm{tr}_{k \subseteq F}
  (\alpha)$ (Exercise~1.13) equals $\sum_{i=1}^d \iota_i(\alpha)$. (Hint:
  Exercises~1.14 and~1.15.) [4.21, 4.22, 6.15]
\end{exer}
\begin{exer}{}{fields:exer:4.20}
  Let $k \subseteq F$ be a finite separable extension, and let $\alpha \in
  F$. Prove that for all $\sigma \in \mathrm{Aut}_k(F)$,
  $N_{k \subseteq F}(\alpha / \sigma(\alpha)) = 1$ and
  $\mathrm{tr}_{k \subseteq F}(\alpha - \sigma(\alpha)) = 0$. [6.16, 6.19]
\end{exer}
\begin{exer}{}{fields:exer:4.21}
  Let $k \subseteq E \subseteq F$ be finite separable extensions, and let
  $\alpha \in F$. Prove that
  \begin{align*}
    N_{k \subseteq F}(\alpha) &= N_{k \subseteq E}(N_{E \subseteq F}(\alpha)),
    \\
    \mathrm{tr}_{k \subseteq F}(\alpha)
      &= \mathrm{tr}_{k \subseteq E}(\mathrm{tr}_{E \subseteq F}(\alpha)).
  \end{align*}
  (Hint: Use Exercise~4.19: if $d = [E : k]$ and $e = [F : E]$, the $de$
  embeddings of $F$ into $\bar{k}$ lifting $\mathrm{id}_k$ must divide into
  $d$ groups of $e$ each, according to their restriction to $E$.)

  This `transitivity' of norm and trace extends the result of Exercise~1.15
  to separable extensions. The separability restriction is actually
  unnecessary; cf.\ Exercise~4.22. [4.22]
\end{exer}
\begin{exer}{}{fields:exer:4.22}
  Generalize Exercises~4.19--4.21 to all finite extensions $k \subseteq F$.
  (For the norm, raise to power $[F : k]_i$; for the trace, multiply by
  $[F : k]_i$.)
\end{exer}

\section{Field extensions, III}
\label{sec:fields:field-extensions-iii}

The material in $\S$4 provides us with the main tools needed to tackle
several key examples. In this section we study finite and cyclotomic
fields, and we return to the question of when a finite extension is in fact
simple (cf.\ Example~1.19).

\subsection{Finite fields}
\label{subsec:fields:finite-fields}
Let $F$ be a finite field, and let $p$ be its characteristic. We know
($\S$1.1) that $F$ may be viewed as an extension $\mathbb{F}_p \subseteq F$
of $\mathbb{F}_p = \Z/p\Z$; let $d = [F : \mathbb{F}_p]$. Since $F$ has
dimension $d$ as a vector space over $\mathbb{F}_p$, it is isomorphic to
$\mathbb{F}_p^d$ as a vector space, and in particular $|F| = p^d$ is a
power of $p$.

The general question we pose is whether there exist fields of cardinality
equal to every power of a prime, and we aim to classify all fields of a
given cardinality. The conclusion we will reach is as neat as it can be:
for every prime power $q$ there exists exactly one field $F$ with $q$
elements, up to isomorphism.

Also recall (Remark~III.1.16) that, by a theorem of Wedderburn, every
finite division ring is in fact a finite field. Thus, relaxing the
hypothesis of commutativity in this subsection would not lead to a
different classification.
\begin{thm}{}{fields:5.1}
  Let $q = p^d$ be a power of a prime integer $p$. Then the polynomial
  $x^q - x$ is separable over $\mathbb{F}_p$, and the splitting field of
  $x^q - x$ over $\mathbb{F}_p$ is a field with precisely $q$ elements.
  Conversely, let $F$ be a field with exactly $q$ elements; then $F$ is a
  splitting field for $x^q - x$ over $\mathbb{F}_p$.
\end{thm}
\begin{proof}
  Let $F$ be the splitting field of $x^q - x$ over $\mathbb{F}_p$. Let $E$
  be the set of roots of $f(x) = x^q - x$ in $F$. Since $f'(x) = qx^{q-1}
  - 1 = -1$ (as $q = 0$ in characteristic $p$), we have $(f(x), f'(x)) =
  1$; hence (\cref{lem:fields:4.13}) $f(x)$ is separable, and $E$ consists
  of precisely $q$ elements. I claim that $E$ is a field, and it follows
  that $E = F$. Indeed, $F$ is generated by the roots of $f(x)$; hence the
  smallest subfield of $F$ containing $E$ is $F$ itself.

  To see that $E$ is a field, let $a, b \in E$. Then $a^q = a$ and
  $b^q = b$; it follows that
  \[
    (a - b)^q = a^q + (-1)^q b^q = a - b
  \]
  (using Exercise~4.8; note that $(-1)^q = -1$ if $p$ is odd and
  $(-1)^q = +1 = -1$ if $p = 2$). If $b \neq 0$,
  \[
    (ab^{-1})^q = a^q (b^q)^{-1} = ab^{-1}.
  \]
  Thus $E$ is closed under subtraction and division by a nonzero element,
  proving that $E$ is a field and concluding the proof of the first
  statement.

  To prove the second statement, let $F$ be a field with exactly $q$
  elements. The nonzero elements of $F$ form a group under multiplication,
  consisting of $q - 1$ elements; therefore, the (multiplicative) order of
  every nonzero $a \in F$ divides $q - 1$ (Example~II.8.15). Therefore,
  \[
    a \neq 0 \implies a^{q-1} = 1 \implies a^q - a = 0;
  \]
  of course $0^q - 0 = 0$. In other words, the polynomial $x^q - x$ has
  $q$ roots in $F$ (that is, all elements of $F$!); it follows that $F$ is
  a splitting field for $x^q - x$, as stated.
\end{proof}
\begin{coro}{}{fields:5.2}
  For every prime power $q$ there exists one and only one finite field of
  order $q$, up to isomorphism.
\end{coro}
\begin{proof}
  This follows immediately from \cref{thm:fields:5.1} and the uniqueness of
  splitting fields (\cref{lem:fields:4.2}).
\end{proof}
Since there is exactly one isomorphism class of fields of order $q$ for
any given prime power $q$, we can devise a notation for a field of order
$q$; it makes sense to adopt\footnote{Keep in mind that $\mathbb{F}_q$ is
not the ring $\Z/q\Z$ unless $q$ is prime: if $q$ is composite, then
$\Z/q\Z$ is not an integral domain.}
$\mathbb{F}_q$. This is called the \emph{Galois field} of order $q$.
\begin{exam}{}{fields:exam:5.3}
  Let $p$ be a prime integer. Then I claim that the polynomial $x^4 + 1$ is
  reducible\footnote{Of course $x^4 + 1$ is irreducible over $\Z$. This
  example shows that Proposition~V.5.15 cannot be turned into an `if and
  only if' statement.} over $\mathbb{F}_p$ (and therefore over every finite
  field).

  Since $x^4 + 1 = (x+1)^4$ in $\mathbb{F}_2[x]$, the statement holds for
  $p = 2$. Thus, we may assume that $p$ is an odd prime. Then I claim that
  $x^4 + 1$ divides $x^{p^2} - x$. Indeed, the square of every odd number
  is congruent to $1 \pmod{8}$ (Exercise~II.2.11); hence $8 \mid (p^2-1)$;
  hence $x^8 - 1$ divides $x^{p^2-1} - 1$ (Exercise~V.2.13); hence
  \[
    (x^4+1) \mid (x^8-1) \mid (x^{p^2-1}-1) \mid (x^{p^2}-x).
  \]
  It follows that $x^4 + 1$ factors completely in the splitting field of
  $x^{p^2} - x$, that is, in $\mathbb{F}_{p^2}$. If $\alpha$ is a root of
  $x^4 + 1$ in $\mathbb{F}_{p^2}$, we have the extensions
  $\mathbb{F}_p \subseteq \mathbb{F}_p(\alpha) \subseteq \mathbb{F}_{p^2}$;
  therefore (\cref{coro:fields:1.11}) $[\mathbb{F}_p(\alpha) : \mathbb{F}_p]$
  divides $[\mathbb{F}_{p^2} : \mathbb{F}_p] = 2$. That is, $\alpha$ has
  degree $1$ or $2$ over $\mathbb{F}_p$. But then its minimal polynomial is
  a factor of degree $1$ or $2$ of $(x^4+1)$, showing that the latter is
  reducible.
\end{exam}
\cref{thm:fields:5.1} has many interesting consequences, and we sample a
few in the rest of this subsection.
\begin{coro}{}{fields:5.4}
  Let $p$ be a prime, and let $d \leq e$ be positive integers. Then there
  is an extension $\mathbb{F}_{p^d} \subseteq \mathbb{F}_{p^e}$ if and
  only if $d \mid e$. Further, if $d \mid e$, then there is exactly one
  such extension, in the sense that $\mathbb{F}_{p^e}$ contains a unique
  copy of $\mathbb{F}_{p^d}$. All extensions $\mathbb{F}_{p^d} \subseteq
  \mathbb{F}_{p^e}$ are simple.
\end{coro}
\begin{proof}
  If there is an extension as stated, then $\mathbb{F}_p \subseteq
  \mathbb{F}_{p^d} \subseteq \mathbb{F}_{p^e}$; hence $[\mathbb{F}_{p^d} :
  \mathbb{F}_p]$ divides $[\mathbb{F}_{p^e} : \mathbb{F}_p]$ by
  \cref{coro:fields:1.11}. This says precisely that $d \mid e$.

  Conversely, assume that $d \mid e$. As
  \[
    p^e - 1 = (p^d - 1)\bigl((p^d)^{e/d - 1} + \dots + 1\bigr),
  \]
  we see that $p^d - 1$ divides $p^e - 1$, and consequently $x^{p^d - 1}
  - 1$ divides $x^{p^e - 1} - 1$ (Exercise~V.2.13). Therefore
  $(x^{p^d} - x) \mid (x^{p^e} - x)$. By \cref{thm:fields:5.1},
  $\mathbb{F}_{p^e}$ is a splitting field for the second polynomial. It
  follows that it contains a unique copy of the splitting field for the
  first polynomial (Exercise~4.5), that is, of $\mathbb{F}_{p^d}$.

  For the last statement, recall that the multiplicative group of nonzero
  elements of a finite field is necessarily cyclic (Theorem~IV.6.10). If
  $\alpha \in \mathbb{F}_{p^e}$ is a generator of this group, then $\alpha$
  will generate $\mathbb{F}_{p^e}$ over any subfield; if $d \mid e$, this
  says $\mathbb{F}_{p^e} = \mathbb{F}_{p^d}(\alpha)$, so $\mathbb{F}_{p^d}
  \subseteq \mathbb{F}_{p^e}$ is simple.
\end{proof}
These results can be translated into rather precise information on the
structure of the polynomial ring over a finite field. For example,
\begin{coro}{}{fields:5.5}
  Let $F$ be a finite field. Then for all integers $n \geq 1$ there exist
  irreducible polynomials of degree $n$ in $F[x]$.
\end{coro}
\begin{proof}
  We know $F = \mathbb{F}_{p^d}$ for some prime $p$ and some $d \geq 1$.
  By \cref{coro:fields:5.4} there is an extension $\mathbb{F}_{p^d}
  \subseteq \mathbb{F}_{p^{dn}}$, generated by an element $\alpha$. Then
  $[\mathbb{F}_{p^{dn}} : \mathbb{F}_{p^d}] = n$, and it follows that the
  minimal polynomial of $\alpha$ over $F = \mathbb{F}_{p^d}$ is an
  irreducible polynomial of degree $n$ in $F[x]$.
\end{proof}
In fact, our analysis of extensions of finite fields tells us about
explicit factorizations, leading to an inductive algorithm to find all
irreducible polynomials in $\mathbb{F}_q[x]$:
\begin{coro}{}{fields:5.6}
  Let $F = \mathbb{F}_q$ be a finite field, and let $n$ be a positive
  integer. Then the factorization of $x^{q^n} - x$ in $F[x]$ consists of
  all irreducible monic polynomials of degree $d$, as $d$ ranges over the
  positive divisors of $n$. In particular, all these polynomials factor
  completely in $\mathbb{F}_{q^n}$.
\end{coro}
\begin{proof}
  By \cref{thm:fields:5.1}, $\mathbb{F}_{q^n}$ is the splitting field of
  $x^{q^n} - x$ over $\mathbb{F}_p$, and hence over $\mathbb{F}_q = F$.

  If $f(x)$ is a monic irreducible polynomial of degree $d$, then
  $F[x]/(f(x)) \cong F(\alpha)$ is an extension of degree $d$ of $F$, that
  is, an isomorphic copy of $\mathbb{F}_{q^d}$. By \cref{coro:fields:5.4},
  if $d \mid n$, then there is an embedding of $\mathbb{F}_{q^d}$ in
  $\mathbb{F}_{q^n}$. But then $\alpha$ must be a root of $x^{q^n} - x$,
  and hence $x^{q^n} - x$ is a multiple of $f(x)$, as this is the minimal
  polynomial of $\alpha$. This proves that every irreducible polynomial of
  degree $d \mid n$ is a factor of $x^{q^n} - x$.

  Conversely, if $f(x)$ is an irreducible factor of $x^{q^n} - x$, then
  $\mathbb{F}_{q^n}$ contains a root $\alpha$ of $f(x)$; we have the
  extensions $F = \mathbb{F}_q \subseteq \mathbb{F}_q(\alpha) \subseteq
  \mathbb{F}_{q^n}$, and $\mathbb{F}_q(\alpha) \cong \mathbb{F}_{q^d}$ for
  $d = \deg \alpha$. It follows that $d \mid n$, again by
  \cref{coro:fields:5.4}.
\end{proof}
The picture I am trying to convey is the following: the $q^n$ roots of
$x^{q^n} - x$ clump into disjoint subsets, with each subset collecting the
roots of each and every irreducible polynomial of degree $d \mid n$ in
$F[x]$.
\begin{exam}{}{fields:exam:5.7}
  Let's contemplate the case $q = 2$: $\mathbb{F}_2 = \Z/2\Z$.
  \begin{itemize}
    \item $n = 1$: the polynomial $x^2 - x$ factors as the product of $x$
      and $(x - 1)$ (which we could write as $(x + 1)$ just as well, since
      we are working over $\mathbb{F}_2$). These are all the irreducible
      polynomials of degree $1$ over $\mathbb{F}_2$.
    \item $n = 2$: the polynomial $x^4 - x$ must factor as the product of
      all irreducible polynomials of degree $1$ and $2$; in fact
      \[
        x^4 - x = x(x-1)(x^2+x+1),
      \]
      and the conclusion is that there is exactly one irreducible polynomial
      of degree $2$ over $\mathbb{F}_2$, namely $x^2 + x + 1$.
    \item $n = 3$: the quotient of $x^8 - x$ by $x(x-1)$ is a polynomial
      of degree $6$, which must therefore be the product of the two
      irreducible polynomials of degree $3$ over $\mathbb{F}_2$. It takes a
      moment to find them: $x^3 + x^2 + 1$ and $x^3 + x + 1$. It also
      follows that $\mathbb{F}_8$ may be realized in two ways as a quotient
      of $\mathbb{F}_2[x]$ modulo an irreducible polynomial:
      \[
        \frac{\mathbb{F}_2[x]}{(x^3+x^2+1)} \cong
        \frac{\mathbb{F}_2[x]}{(x^3+x+1)}.
      \]
      We know that these two fields must be isomorphic because of
      \cref{thm:fields:5.1}; it is instructive to find an explicit
      isomorphism between them (Exercise~5.3).
    \item $n = 4$ and $5$: these cases are left to the enjoyment of the
      reader (Exercise~5.4).
    \item $n = 6$: the factorization of $x^{64} - x$ must include $x$,
      $x - 1$, the one irreducible polynomial of degree $2$, and the two
      irreducible polynomials of degree $3$ found above, and that leaves
      room for $9$ polynomials of degree $6$, which must be all and only
      the irreducible polynomials of degree $6$ over $\mathbb{F}_2$. So the
      $64$ elements of $\mathbb{F}_{64}$ cluster into disjoint sets of
      roots of each irreducible polynomial. The unique copies of
      $\mathbb{F}_4$ and $\mathbb{F}_8$ contained in $\mathbb{F}_{64}$
      intersect in the unique copy of $\mathbb{F}_2$. Again,
      $\mathbb{F}_{64}$ may be realized by quotienting $\mathbb{F}_2[x]$
      by the ideal generated by any of the irreducible polynomials of
      degree $6$; this gives $9$ `different' realizations of this field.
  \end{itemize}
\end{exam}
The picture drawn above for $\mathbb{F}_{64}$ means next to nothing, but
it may help us focus on one last element of information we are going to
extract regarding finite fields. Note that the effect of any automorphism
of $\mathbb{F}_{64}$ must be to scramble elements in each sector
represented in the picture, without mixing elements in different sectors and
without interchanging sectors. Indeed, every automorphism of an extension
sends roots of an irreducible polynomial to roots of the same polynomial.

Since extensions of finite fields are simple extensions, our previous work
allows us to be much more precise. Restricting our attention to the
extensions $\mathbb{F}_p \subseteq \mathbb{F}_{p^d}$, for a prime $p$, we
know these can be realized as simple extensions by an element with minimal
polynomial of degree $d$. This polynomial is necessarily separable
($\mathbb{F}_p$ is perfect), so \cref{coro:fields:1.7} immediately gives
us the size of the automorphism group:
$|\mathrm{Aut}_{\mathbb{F}_p}(\mathbb{F}_{p^d})| = d$. But what is this
group?
\begin{prop}{}{fields:5.8}
  $\mathrm{Aut}_{\mathbb{F}_p}(\mathbb{F}_{p^d})$ is cyclic, generated by
  the Frobenius isomorphism.
\end{prop}
\begin{proof}
  Let $\phi$ be the Frobenius homomorphism $\mathbb{F}_{p^d} \to
  \mathbb{F}_{p^d}$: $\phi(x) = x^p$. The Frobenius homomorphism is an
  isomorphism on a finite field (\cref{coro:fields:4.18}) and restricts to
  the identity on $\mathbb{F}_p$ (Exercise~4.11), so $\phi \in
  \mathrm{Aut}_{\mathbb{F}_p}(\mathbb{F}_{p^d})$. Since we know a priori
  the size of this group, all we have to show is that the order of $\phi$
  is $d$.

  Let $e = |\phi|$. Then $\phi^e = \mathrm{id}$; hence $x^{p^e} = x$ for
  all $x \in \mathbb{F}_{p^d}$. In other words, the (nonzero) polynomial
  $x^{p^e} - x$ has $p^d$ roots in $\mathbb{F}_{p^d}$; this implies
  $p^d \leq p^e$ (Lemma~V.5.1). Equivalently, $d \leq e$, yielding
  $|\mathrm{Aut}_{\mathbb{F}_p}(\mathbb{F}_{p^d})| \leq |\phi|$. But then
  these two numbers must be equal, since the order of an element always
  divides the order of the group (Example~II.8.15).
\end{proof}
Two last comments on finite fields are in order:
\begin{itemize}
  \item For $n$ large enough, the stupendously large (but finite) field
    $\mathbb{F}_{p^{n!}}$ works as an approximation of the algebraic
    closure $\bar{\mathbb{F}}_p$: indeed, this field contains roots of all
    polynomials of degree $\leq n$ in $\mathbb{F}_p[x]$, by
    \cref{coro:fields:5.6}. This observation can be turned into the
    explicit construction of an algebraic closure of $\mathbb{F}_p$: by
    \cref{coro:fields:5.4} there are extensions
    \[
      \mathbb{F}_p \subseteq \mathbb{F}_{p^2} \subseteq \mathbb{F}_{p^6}
      \subseteq \mathbb{F}_{p^{4!}} \subseteq \cdots
    \]
    and the union of this chain of fields\footnote{A more formal
    construction requires the notion of direct limit; cf.\ $\S$VIII.1.4.}
    gives a copy of $\bar{\mathbb{F}}_p$.
  \item Finite fields form a category, which (by \cref{coro:fields:5.4})
    should strongly remind the reader of the `toy' category encountered in
    Exercise~I.5.6. There is an important difference, however: in that
    category every object has exactly one automorphism, while we have just
    checked that finite fields may have many automorphisms. Challenge: The
    reader has encountered a very similar situation in a different (but
    related) context. Where?
\end{itemize}

\subsection{Cyclotomic polynomials and fields}
\label{subsec:fields:cyclotomic-polynomials-and-fields}
In the last several sections we have often encountered roots of $1$ in
$\C$; the extensions they generate are very important. Let $n$ be a
positive integer. I will denote by $\zeta_n$ the complex number
$e^{2\pi i/n}$; thus, the $n$ roots of the polynomial $x^n - 1$ in $\C$
are the $n$ distinct powers of $\zeta_n$; they form a cyclic subgroup of
order $n$ of the multiplicative group of $\C$, which I will denote $\mu_n$.
Pictorially, the roots of $1$ are placed at the vertices of a regular
$n$-gon centered at $0$, with one vertex at $1$.

A primitive $n$-th root of $1$ is a generator of $\mu_n$. Thus, $\zeta_n$
is primitive; from our work on cyclic groups, we know
(\cref{coro:fields:II.2.5}) that $\zeta_n^m$ is a primitive $n$-th root of
$1$ if and only if $m$ is relatively prime to $n$. In particular, there are
$\varphi(n)$ primitive $n$-th roots of $1$, where $\varphi$ is Euler's
totient function (cf.\ Exercise~II.6.14). Thus,
\[
  \Phi_n(x) := \prod_{\substack{\zeta \text{ primitive} \\ n\text{-th root of }1}}
  (x - \zeta)
  = \prod_{\substack{1 \leq m \leq n \\ (m,n)=1}} (x - \zeta_n^m)
\]
is a polynomial of degree $\varphi(n)$.
\begin{defn}{}{fields:5.9}
  The polynomial $\Phi_n(x)$ is called the \emph{$n$-th cyclotomic
  polynomial}.
\end{defn}
It is clear that $\Phi_n(x)$ is a monic polynomial of degree $\varphi(n)$.
It is perhaps a little less immediate that $\Phi_n(x)$ has rational (in
fact, integer) coefficients and that it is irreducible over $\Q$. We will
prove these facts in short order.
\begin{exam}{}{fields:exam:5.10}
  If $n = p$ is prime, then every nonidentity element of $\mu_p \cong C_p$
  is a generator: every $p$-th root of $1$ is primitive except $1$ itself.
  Therefore
  \[
    \Phi_p(x) = \frac{x^p - 1}{x - 1} = x^{p-1} + \dots + 1
  \]
  is the particular case encountered in Example~V.5.19, where we proved
  that $\Phi_p(x)$ is indeed irreducible.
\end{exam}
What if $n$ is not prime?
\begin{lem}{}{fields:5.11}
  For all positive integers $n$,
  \[
    x^n - 1 = \prod_{1 \leq d \mid n} \Phi_d(x).
  \]
\end{lem}
\begin{proof}
  If $n = de$, then every $d$-th root $\zeta$ of $1$ is an $n$-th root of
  $1$, because $\zeta^n = \zeta^{de} = (\zeta^d)^e = 1$. In particular,
  every primitive $d$-th root $\zeta$ of $1$ is an $n$-th root of $1$.

  On the other hand, every $\zeta \in \mu_n$ generates a subgroup $H$ of
  $\mu_n$, and $H = \mu_d$ for $d$ equal to the order of $\zeta$, a
  divisor of $n$ (Proposition~II.6.11). Thus, every $\zeta \in \mu_n$ is a
  primitive $d$-th root of $1$ for some $d \mid n$.

  Thus the set of $n$-th roots of $1$ equals the union of the sets of
  primitive $d$-th roots of $1$, as $d$ ranges over all positive divisors
  of $n$. The statement follows immediately:
  \[
    x^n - 1
    = \prod_{\zeta \in \mu_n} (x - \zeta)
    = \prod_{1 \leq d \mid n} \prod_{\substack{\zeta \text{ primitive} \\ d\text{-th root of }1}}
      (x - \zeta)
    = \prod_{1 \leq d \mid n} \Phi_d(x),
  \]
  as claimed.
\end{proof}
This argument brings us back to simple group-theoretic considerations. The
reader can check that \cref{lem:fields:5.11} implies immediately the result
of Exercise~II.6.14, by comparing degrees of both sides of the stated
identity.

\cref{lem:fields:5.11} yields an inductive computation of cyclotomic
polynomials; the fact that $\Phi_n(x) \in \Z[x]$ follows from this fact.
Explicitly,
\begin{coro}{}{fields:5.12}
  The cyclotomic polynomials $\Phi_n(x)$ have integer coefficients.
\end{coro}
\begin{proof}
  Use induction on $n$. Note that $\Phi_1(x) = x - 1$, and assume we have
  shown that all $\Phi_m(x)$ have integer coefficients for $m < n$. In
  particular, $f(x) := \prod_{1 \leq d \mid n,\, d < n} \Phi_d(x)$ is a
  monic polynomial with integer coefficients. Since $f(x)$ is monic, we can
  divide it into $x^n - 1$ with remainder, within $\Z[x]$: there exist
  $q(x), r(x) \in \Z[x]$ such that
  \[
    x^n - 1 = f(x)q(x) + r(x),
  \]
  with $r(x) = 0$ or $\deg r(x) < \deg f(x)$. On the other hand, by
  \cref{lem:fields:5.11}, $x^n - 1 = f(x)\Phi_n(x)$ in $\C[x]$. Therefore
  \[
    f(x)(\Phi_n(x) - q(x)) = r(x)
  \]
  in $\C[x]$. But this forces $r(x) = 0$ (otherwise we would have
  $\deg r(x) \geq \deg f(x)$). Therefore $\Phi_n(x) = q(x) \in \Z[x]$.
\end{proof}
\begin{exam}{}{fields:exam:5.13}
  The reader can spend some quality time computing explicitly the cyclotomic
  polynomials $\Phi_n(x)$ for several nonprime numbers $n$, working
  inductively and capitalizing on the fact that we know explicitly
  $\Phi_p(x)$ for prime $p$. For example, $x^4 - 1 = \Phi_1(x)\Phi_2(x)
  \Phi_4(x)$; therefore
  \[
    \Phi_4(x) = \frac{x^4 - 1}{x^2 - 1} = x^2 + 1.
  \]
  Since $x^6 - 1 = \Phi_1(x)\Phi_2(x)\Phi_3(x)\Phi_6(x)$,
  \[
    \Phi_6(x) = \frac{x^6 - 1}{(x^2-1)(x^2+x+1)} = x^2 - x + 1.
  \]
  Since $x^{12} - 1 = \Phi_1(x)\Phi_2(x)\Phi_3(x)\Phi_4(x)\Phi_6(x)
  \Phi_{12}(x)$,
  \[
    \Phi_{12}(x) = \frac{x^{12}-1}{(x^6-1)(x^2+1)} = x^4 - x^2 + 1,
  \]
  and so on ad libitum.
\end{exam}
The optimistic reader may now start guessing that the coefficients of
$\Phi_n(x)$ are always $0$ or $\pm 1$ (so would I, if I didn't know
better). Apparently the first counterexample is found for $n = 105$
(Exercise~5.9), and the coefficients are known to get as large as one
pleases for $n \to \infty$.
The irreducibility of $\Phi_n(x)$ is a bit trickier, in particular because
separability sneaks into the standard argument (which is why I had to wait
until now to present it).
\begin{prop}{}{fields:5.14}
  For all positive $n$, $\Phi_n(x) \in \Z[x]$ is irreducible over $\Q$.
  (Since $\Phi_n(x)$ is monic, irreducibility in $\Q[x]$ is equivalent to
  irreducibility in $\Z[x]$; cf.\ Corollary~V.4.17.)
\end{prop}
\begin{proof}
  Arguing by contradiction, assume $\Phi_n(x)$ is reducible. Then its
  roots $\zeta_n^m$, with $(m,n) = 1$, are divided among the factors; we
  can choose a root $\zeta_n^m$ of one irreducible monic factor $f(x)$,
  such that another root $\zeta_n^{mp}$ (for some prime $p$ not dividing
  $n$) is not a root of $f(x)$. Write $\Phi_n(x) = f(x)g(x)$; since
  $\Phi_n(x) \in \Z[x]$ and $\Phi_n(x)$, $f(x)$ are monic, then $f(x)$
  and $g(x)$ have integer coefficients (cf.\ Exercise~V.4.23). By our
  choice, $f(x)$ is the minimal polynomial of $\zeta_n^m$ over $\Q$, and
  $g(\zeta_n^{mp}) = 0$.

  It follows that $\zeta_n^m$ is a root of $g(x^p)$, and hence
  $f(x) \mid g(x^p)$. Therefore we can write $g(x^p) = f(x)h(x)$ with
  $h(x) \in \Z[x]$. Reading the last equation modulo $p$, we get (again
  using Exercise~4.8, and denoting cosets by underlining)
  \[
    \underline{g}(x)^p = \underline{f}(x)\underline{h}(x)
    \quad\text{in } \mathbb{F}_p[x];
  \]
  in particular, $\underline{f}(x)$ and $\underline{g}(x)$ must have a
  nontrivial common factor in $\mathbb{F}_p[x]$. But then its square
  divides $\underline{f}(x)\underline{g}(x)$: the reduction of $\Phi_n(x)$
  modulo $p$ must have a multiple factor.

  This implies that $x^n - 1 \in \mathbb{F}_p[x]$ has a multiple factor;
  that is, it is inseparable. However, its derivative $nx^{n-1} \in
  \mathbb{F}_p[x]$ is nonzero (because $p$ does not divide $n$ by
  assumption), and \cref{lem:fields:4.13} implies that $x^n - 1$ is
  separable in $\mathbb{F}_p[x]$. This contradiction shows that our
  assumption that $\Phi_n(x)$ is reducible must be nonsense, proving the
  statement.
\end{proof}
\begin{defn}{}{fields:5.15}
  The splitting field $\Q(\zeta_n)$ for the polynomial $x^n - 1$ over $\Q$
  is the \emph{$n$-th cyclotomic field}.
\end{defn}
By \cref{prop:fields:5.14}, $\Q(\zeta_n)$ is an extension of $\Q$ of
degree $\varphi(n)$; $\Phi_n(x)$ is the minimal polynomial of $\zeta_n$.
As usual, we now preoccupy ourselves with the automorphism group of this
extension, and as in previous examples the situation is very neat.
\begin{prop}{}{fields:5.16}
  $\mathrm{Aut}_\Q(\Q(\zeta_n))$ is isomorphic to the group of units in
  $\Z/n\Z$.
\end{prop}
\begin{proof}
  We know that $\mathrm{Aut}_\Q(\Q(\zeta_n))$ has cardinality $\varphi(n)$
  (\cref{coro:fields:1.7}; the roots are distinct since $\Phi_n(x)$ is
  separable), so all we need to do is exhibit an injective homomorphism
  $j : (\Z/n\Z)^* \to \mathrm{Aut}_\Q(\Q(\zeta_n))$. For $[m]_n \in
  (\Z/n\Z)^*$, that is, for $m$ an integer relatively prime to $n$, let
  $j([m]_n)$ be the unique automorphism $\Q(\zeta_n) \to \Q(\zeta_n)$
  sending $\zeta_n$ to $\zeta_n^m$ (cf.\ \cref{prop:fields:1.5}); this is
  evidently independent of the representative $m$. Then $j$ is clearly
  injective, and
  \[
    j([m_1]_n) \circ j([m_2]_n)(\zeta_n)
    = j([m_1]_n)(\zeta_n^{m_2})
    = \zeta_n^{m_2 m_1}
  \]
  agrees with the action of $j([m_1]_n [m_2]_n)$, showing that $j$ is a
  homomorphism and concluding the proof.
\end{proof}
\begin{exam}{}{fields:exam:5.17}
  The reader should consider Example~4.4 again: by what we have just
  proved, the automorphism group of the splitting field of $x^8 - 1$ is
  isomorphic to the group of units in $\Z/8\Z$; this group is immediately
  seen to be isomorphic to $(\Z/2\Z) \times (\Z/2\Z)$, confirming the
  claim made in Example~4.4.
\end{exam}
The constructibility of the regular $n$-gon by straightedge and compass is
of course equivalent to the constructibility of the complex number
$\zeta_n$. Since we now know that $[\Q(\zeta_n) : \Q] = \varphi(n)$,
\cref{coro:fields:3.6} tells us that the regular $n$-gon is constructible
by straightedge and compass only if $\varphi(n)$ is a power of $2$. This
condition can be chased a little further (Exercise~5.19). In any case, we
shall return to this theme one more time, after we absorb a little Galois
theory.

\subsection{Separability and simple extensions}
\label{subsec:fields:separability-and-simple-extensions}
All the examples of field extensions we have encountered so far were simple
extensions, although sometimes this may not be completely apparent at first
(cf.\ Example~1.19). There is a good reason for this: most of the examples
we have seen were separable extensions, and we can now prove that a finite
separable extension is necessarily simple. Here is a nice criterion for
simplicity (which does not involve any separability condition):
\begin{prop}{}{fields:5.18}
  An algebraic extension $k \subseteq F$ is simple if and only if the
  number of distinct intermediate fields $k \subseteq E \subseteq F$ is
  finite.
\end{prop}
\begin{proof}
  Assume $F = k(\alpha)$ is simple and algebraic, and let $q_k(x)$ be the
  minimal polynomial of $\alpha$ over $k$. Embed $F$ in an algebraic
  closure $\bar{k}$ of $k$. If $E$ is an intermediate field, then
  $F = E(\alpha)$ is also a simple, algebraic extension; denote by $q_E(x)$
  the minimal polynomial of $\alpha$ over $E$. Since $q_k(x) \in E[x]$ for
  all $E$ and $q_k(\alpha) = 0$, we know that each $q_E(x)$ is a factor of
  $q_k(x)$.

  I claim that $E$ is in fact determined by $q_E(x)$. Since $q_k(x)$ has
  finitely many factors in $\bar{k}$, this proves that there are only
  finitely many intermediate fields, that is, the `only if' part of the
  statement.

  To verify my claim, I will show that $E$ is generated (over $k$) by the
  coefficients of $q_E(x)$. To see this, let $E'$ be the subfield of $E$
  generated by $k$ and the coefficients of $q_E(x)$. Then $q_E(x) \in
  E'[x]$, and since $q_E(x)$ is irreducible over $E$, it must be
  irreducible over $E'$. Note that $E'(\alpha) = F = E(\alpha)$. We have
  $E' \subseteq E \subseteq F$; therefore
  \[
    \deg(q_E(x)) = [F : E'] = [F : E][E : E'] = \deg(q_E(x))[E : E'],
  \]
  from which $[E : E'] = 1$; that is, $E = E'$ as needed. This concludes
  the proof of the `only if' part of the statement.

  To prove the `if' part, assume that there are only finitely many
  intermediate fields $k \subseteq E \subseteq F$. The extension
  $k \subseteq F$ must be finitely generated (otherwise we could construct
  an infinite sequence of subextensions $k \subseteq k(\alpha_1) \subseteq
  k(\alpha_1, \alpha_2) \subseteq \dots \subseteq F$, and we would have
  infinitely many intermediate fields), hence finite since it is algebraic.
  If $k$ is a finite field, then so is $F$, and the extension is
  automatically simple by \cref{coro:fields:5.4}; thus we may assume that
  $k$ is infinite, and we have to show that every finitely generated
  algebraic extension $F = k(\alpha_1, \dots, \alpha_r)$ is simple.

  Arguing inductively, we may assume without loss of generality that
  $F = k(\alpha, \beta)$. For every $c \in k$, we have the intermediate
  field $k \subseteq k(c\alpha + \beta) \subseteq k(\alpha, \beta)$. But
  there are only finitely many intermediate fields, while $k$ is infinite,
  so for some $c \neq c'$ in $k$ we must have
  $k(c'\alpha + \beta) = k(c\alpha + \beta)$. It follows that
  \[
    \alpha = \frac{(c'\alpha + \beta) - (c\alpha + \beta)}{c' - c}
    \in k(c\alpha + \beta)
    \quad\text{and}\quad
    \beta = (c\alpha + \beta) - c\alpha \in k(c\alpha + \beta),
  \]
  and therefore $k(\alpha, \beta) \subseteq k(c\alpha + \beta)$. The other
  inclusion holds trivially, so $k(\alpha, \beta) = k(c\alpha + \beta)$ is
  simple.
\end{proof}
Here is the connection with separability:
\begin{prop}{}{fields:5.19}
  Every finite separable extension is simple.
\end{prop}
\begin{proof}
  Arguing inductively as in the proof of \cref{prop:fields:5.18}, we may
  assume $F = k(\alpha, \beta)$, with $\alpha$ and $\beta$ separable (and
  in particular algebraic) over $k$, and we may assume $k$ is an infinite
  field.

  Consider the set $I$ of embeddings $\iota : F \to \bar{k}$ of $F$ in an
  algebraic closure of $k$, extending the identity of $k$. If $\iota \neq
  \iota'$ in $I$ and $x$ is an indeterminate, then the polynomials
  $\iota(\alpha) x + \iota(\beta)$ and $\iota'(\alpha) x + \iota'(\beta)$
  are different: otherwise $\iota'(\alpha) = \iota(\alpha)$ and
  $\iota'(\beta) = \iota(\beta)$, so $\iota$, $\iota'$ would act in the
  same way on the whole of $k(\alpha, \beta)$, forcing $\iota = \iota'$.
  Therefore, the polynomial
  \[
    f(x) = \prod_{\iota \neq \iota'}
    \bigl((\iota(\alpha)x + \iota(\beta)) - (\iota'(\alpha)x +
    \iota'(\beta))\bigr) \in \bar{k}[x]
  \]
  is not identically $0$. Since $k$ is infinite, it follows that there
  exists $c \in k$ such that $f(c) \neq 0$; that is, distinct $\iota \in I$
  map
  \[
    \gamma = c\alpha + \beta
  \]
  to distinct elements $\iota(\gamma) = \iota(\alpha)c + \iota(\beta)$
  (each $\iota$ extends $\mathrm{id}_k$, so $\iota(c) = c$). Since the
  cardinality of $I$ is $[F : k]_s$ (\cref{defn:fields:4.21}) and each
  $\iota(\gamma)$ is a root of the minimal polynomial of $\gamma$ over $k$,
  we have
  \[
    [F : k]_s \leq [k(\gamma) : k] \leq [F : k].
  \]
  By assumption the extension is separable, so $[F : k]_s = [F : k]$ by
  \cref{prop:fields:4.24}, implying $[k(\gamma) : k] = [F : k]$ and finally
  $F = k(\gamma)$, concluding the proof.
\end{proof}
\cref{prop:fields:5.19} is a good illustration of why separability is a
technically desirable condition. To stress the point even further, note
that we are now in a position to extend the convenient inequality found in
\cref{coro:fields:1.7} to all finite separable extensions:
\begin{coro}{}{fields:5.20}
  Let $k \subseteq F$ be a finite, separable extension. Then
  $|\mathrm{Aut}_k(F)| \leq [F : k]$, with equality if and only if
  $k \subseteq F$ is a normal extension.
\end{coro}
\begin{proof}
  Since $k \subseteq F$ is finite and separable, \cref{prop:fields:5.19}
  implies it is simple: $F = k(\alpha)$ for some $\alpha \in F$. The
  inequality follows immediately from \cref{coro:fields:1.7}, and equality
  holds if and only if the minimal polynomial $f(x)$ of $\alpha$ factors
  into distinct linear factors in $F$. In this case $F$ is the splitting
  field of $f(x)$, so it is normal over $k$ by \cref{thm:fields:4.8}.
  Conversely, if $F$ is normal over $k$, then $f(x)$ splits completely in
  $F$ and has distinct roots since $\alpha$ is separable over $k$;
  therefore $|\mathrm{Aut}_k(F)| = [F : k]$, again by
  \cref{coro:fields:1.7}.
\end{proof}
\begin{exam}{}{fields:exam:5.21}
  By \cref{prop:fields:5.19}, if we want to construct a nonsimple finite
  extension, we have to use inseparable elements; to produce an example, we
  jazz up Example~4.11 a little. Consider the field of rational functions
  $F = \mathbb{F}_p(u, v)$ in two variables over $\mathbb{F}_p$ and the
  subfield $k = \mathbb{F}_p(s, t)$, where $s = u^p$ and $t = v^p$. Then
  $k \subseteq F$ is an algebraic extension; the minimal polynomial of $u$
  over $k$ is $x^p - s$; the minimal polynomial of $v$ over $k(u)$ is
  $y^p - t$; it follows that $[F : k] = p^2$.

  As $c$ ranges in $k$, we obtain intermediate fields
  $k \subseteq k(cu + v) \subseteq F$. If any two choices $c$, $c'$ led to
  the same intermediate field, then we would deduce that $k(u, v) =
  k(cu+v)$ by arguing as in the proof of \cref{prop:fields:5.18}. In
  particular, we would have $[k(cu + v) : k] = p^2$. But this is not the
  case, since $(cu + v)^p = c^p s + t \in k(s, t) = k$; hence the minimal
  polynomial of $cu + v$ over $k$ has degree at most $p$. Since
  $k = \mathbb{F}_p(s, t)$ is infinite, there are infinitely many
  intermediate fields, and it follows (by \cref{prop:fields:5.18}) that
  $F$ is not a simple extension of $k$.
\end{exam}

\subsection*{Exercises}

\begin{exer}{}{fields:exer:5.1}
  Let $p > 0$ be a prime integer. Prove that the additive group of a finite
  field with $p^d$ elements is isomorphic to $(\Z/p\Z)^d$.
\end{exer}
\begin{exer}{}{fields:exer:5.2}
  Prove that every element of a finite field $F$ is a sum of two squares in
  $F$. (Hint: Keep in mind that the multiplicative group of $F$ is cyclic.)
\end{exer}
\begin{exer}{}{fields:exer:5.3}
  Find an explicit isomorphism
  \[
    \frac{\mathbb{F}_2[x]}{(x^3+x^2+1)} \cong \frac{\mathbb{F}_2[x]}{(x^3+x+1)}.
  \]
  [$\S$5.1]
\end{exer}
\begin{exer}{}{fields:exer:5.4}
  Find all irreducible polynomials of degree $4$ over $\mathbb{F}_2$, and
  count those of degree $5$. [$\S$5.1]
\end{exer}
\begin{exer}{}{fields:exer:5.5}
  Find the number of irreducible polynomials of degree $12$ over
  $\mathbb{F}_9$.
\end{exer}
\begin{exer}{}{fields:exer:5.6}
  Write out explicitly the action of the cyclic group $C_4$ on
  $\mathbb{F}_{16}$, in terms of any realization of this field as a
  quotient of $\mathbb{F}_2[x]$.
\end{exer}
\begin{exer}{}{fields:exer:5.7}
  Let $p$ be a prime integer. View the Frobenius automorphism
  $\phi : \mathbb{F}_{p^d} \to \mathbb{F}_{p^d}$ as a linear
  transformation of the $\mathbb{F}_p$-vector space $\mathbb{F}_{p^d}$.
  Find the rational canonical form of $\phi$. (Adapt the proof of
  \cref{prop:fields:5.8} to show that the minimal polynomial of $\phi$ is
  $x^d - 1$.)
\end{exer}
\begin{exer}{}{fields:exer:5.8}
  For a prime $p$, find the factorization of $\Phi_p(x)$ over
  $\mathbb{F}_p$.
\end{exer}
\begin{exer}{}{fields:exer:5.9}
  Find all cyclotomic polynomials $\Phi_n(x)$ for $1 \leq n \leq 15$.
  Compute $\Phi_{105}(x)$. [$\S$5.2]
\end{exer}
\begin{exer}{}{fields:exer:5.10}
  Find the cyclotomic polynomials $\Phi_{2^m}(x)$ for all $m \geq 0$.
\end{exer}
\begin{exer}{}{fields:exer:5.11}
  Prove that if $n > 1$ is odd, then $\Phi_{2n}(x) = \Phi_n(-x)$. (Hint:
  Draw the primitive $14$-th roots of $1$ side-by-side to the primitive
  $7$-th roots of $1$; then go back to Exercise~II.2.15 to justify the
  fact you observe.)
\end{exer}
\begin{exer}{}{fields:exer:5.12}
  Let $a$, $n$ be positive integers, with $a > 1$. Prove that if
  $\Phi_n(a)$ divides $a - 1$, then $n = 1$. (Remember that $\Phi_n(a)$
  is a product of complex numbers $a - \zeta$, where $\zeta$ is a
  primitive $n$-th root of $1$. What does this tell you about the size of
  $\Phi_n(a)$?) [5.14]
\end{exer}
\begin{exer}{}{fields:exer:5.13}
  Let $a$, $d$, $n$ be positive integers, with $d < n$ and $a > 1$. Assume
  that $a^d - 1$ divides $a^n - 1$. Prove that $\Phi_n(a)$ divides the
  quotient $(a^n - 1)/(a^d - 1)$. (Hint: Exercise~V.2.13.) [5.14]
\end{exer}
\begin{exer}{}{fields:exer:5.14}
  Let $R$ be a finite division ring.
  \begin{itemize}
    \item Prove that the center of $R$ (Exercise~III.2.9) is isomorphic to
      $\mathbb{F}_q$, for $q$ a prime power. Prove that $|R| = q^n$ for
      some $n$.
    \item For every $r \in R$, prove that the centralizer of $r$ in the
      multiplicative group $(R^*, \cdot)$ has order $q^d - 1$ for some
      $d \leq n$.
    \item Prove that there are integers $d_1, \dots, d_r < n$ such that
      \[
        \tag{$*$} q^n - 1 = q - 1 + \sum_{i=1}^r \frac{q^n - 1}{q^{d_i} - 1}.
      \]
      (Hint: Class equation.)
    \item Deduce that $\Phi_n(q)$ divides $q - 1$ and hence $n = 1$. (Use
      Exercises~5.12 and~5.13.)
    \item Conclude that $R$ equals its center, showing that $R$ is
      commutative.
  \end{itemize}
  Thus, every finite division ring is a field: this is \emph{Wedderburn's
  little theorem}. The argument given here is due to \name{Ernst Witt}.
  [$\S$III.1.2]
\end{exer}
\begin{exer}{}{fields:exer:5.15}
  Let $a$, $p$, $n$ be integers, with $p$, $n$ positive and $p$ prime,
  $p \nmid n$.
  \begin{itemize}
    \item Show that $x^n - 1$ has no multiple roots modulo $p$.
    \item Show that if $p$ divides $\Phi_n(a)$, then $a^n \equiv 1 \pmod{p}$.
      (In particular, $p \nmid a$, so $[a]_p \in (\Z/p\Z)^*$.)
    \item Show that if $p$ divides $\Phi_n(a)$, then $a^d \equiv 1 \pmod{p}$
      for every $d < n$.
    \item Deduce that $p \mid \Phi_n(a)$ if and only if the order of $[a]_p$
      in $(\Z/p\Z)^*$ is $n$.
    \item Compute $\Phi_{15}(9)$, and show it is divisible by $31$. Then
      look back at the first part of Exercise~II.4.12.
  \end{itemize}
  [$\S$II.4.3, 5.16, 5.17]
\end{exer}
\begin{exer}{}{fields:exer:5.16}
  If $q$ is a prime that divides $a^n - 1$ and does not divide $a^d - 1$
  for any $d < n$, then $q$ is said to be a \emph{primitive prime divisor}
  of $a^n - 1$. By Exercise~5.15, if $q \nmid n$ and $q \mid \Phi_n(a)$,
  then $q$ is a primitive prime divisor of $a^n - 1$. The
  \emph{Birkhoff--Vandiver theorem} asserts that $a^n - 1$ has primitive
  prime divisors for all but a very short list of exceptions: $n = 1$,
  $a = 2$; $n = 2$, $a + 1$ a power of $2$; and $n = 6$, $a = 2$.
  Assuming this statement, we can give another proof of Wedderburn's
  theorem (published in 2003 by \name{Nicolas Lichiardopol}). Let $R$ be a
  finite division ring.
  \begin{itemize}
    \item Prove that there is a prime $p$ such that $|R| = p^n$ for some
      integer $n$.
    \item If $n = 1$, then $R \cong \mathbb{F}_p$. If $n = 2$ or if
      $p = 2$ and $n = 6$, then $R$ is commutative: the reader has proved
      this in Exercises~III.2.11 and~IV.2.17.
    \item Therefore, by Birkhoff--Vandiver we may assume that $p^n - 1$ has
      a primitive prime divisor $p'$.
    \item Prove that there is an element $a$ of order $p'$ in the
      multiplicative group $(R^*, \cdot)$ of $R$.
    \item Prove that $R$ is the only sub-division ring of $R$ containing
      $a$.
    \item Prove that the centralizer of $a$ in $R$ (Exercise~III.2.10) is
      $R$ itself; deduce that $a$ is in the center of $R$
      (Exercise~III.2.9).
    \item Prove that the center of $R$ is $R$: this shows that $R$ is
      commutative. Thus, $R$ is a finite field.
  \end{itemize}
  Note that the Birkhoff--Vandiver theorem gives another rapid way to
  conclude Witt's proof: a primitive prime divisor of $q^n - 1$ would
  divide $q - 1$ by $(*)$, showing that $n = 1$.
\end{exer}
\begin{exer}{}{fields:exer:5.17}
  Let $a$, $p$, $n$ be integers, with $p$, $n$ positive and $p$ prime,
  $p \nmid n$. Assume that $p$ divides $\Phi_n(a)$. Prove that
  $p \equiv 1 \pmod{n}$. (Use Exercise~5.15.) [5.18]
\end{exer}
\begin{exer}{}{fields:exer:5.18}
  Let $n > 0$ be any integer. Prove that there are infinitely many prime
  integers $\equiv 1 \pmod{n}$. (Use Exercise~5.17 together with
  Exercise~V.2.25.) The result of this exercise is a particular case of
  Dirichlet's theorem on primes in arithmetic progressions; see $\S$7.6.
  [$\S$7.6]
\end{exer}
\begin{exer}{}{fields:exer:5.19}
  Prove that the regular $n$-gon can be constructed by straightedge and
  compass only if $n = 2^m p_1 \cdots p_r$, where $m \geq 0$ and the
  factors $p_i$ are distinct Fermat primes. (Hint: Use Exercise~V.6.8.)
  [$\S$5.2, $\S$7.2]
\end{exer}
\begin{exer}{}{fields:exer:5.20}
  Recall from Exercise~1.10 that every field extension of degree $\leq n$
  over a field $k$ is a sub-$k$-algebra of the ring of matrices $M_n(k)$.
  Prove that if $k$ is finite or has characteristic $0$, then every
  extension of $k$ contained in $M_n(k)$ has degree $\leq n$.
\end{exer}
\begin{exer}{}{fields:exer:5.21}
  Prove that if $k \subseteq F$ is the splitting field of a separable
  polynomial, then it is the splitting field of an irreducible separable
  polynomial.
\end{exer}
\begin{exer}{}{fields:exer:5.22}
  Let $k$ be an infinite field. If $F = k(\alpha_1, \dots, \alpha_r)$,
  with each $\alpha_i$ separable over $k$, prove that there exist
  $c_1, \dots, c_r \in k$ such that $F = k(c_1\alpha_1 + \dots +
  c_r\alpha_r)$.
\end{exer}
\begin{exer}{}{fields:exer:5.23}
  Let $k$ be a field, and let $n > 0$ be an integer. Assume that there are
  no irreducible polynomials of degree $n$ in $k[x]$. Prove that there are
  no separable extensions of $k$ of degree $n$. [$\S$7.1]
\end{exer}

\section{A little Galois theory}
\label{sec:fields:a-little-galois-theory}

Galois theory is a beautiful interplay of field theory and group theory,
originally motivated by the problem of determining `symmetries' among roots
of a polynomial. \name{Galois} was interested in concrete relations which
must necessarily hold among the roots of a polynomial, the most trivial
example of which being the fact that if $\alpha$, $\beta$ are the two roots
of $x^2 + px + q$, then $\alpha + \beta = -p$ and $\alpha\beta = q$.
Interchanging the roots $\alpha$, $\beta$ has no effect on the quantities
$\alpha + \beta$, $\alpha\beta$. More generally, for a higher degree
polynomial there may be several quantities which are invariant under certain
permutations of the roots. These quantities and the corresponding groups of
permutations may be viewed as invariants determined by the polynomial,
yielding a sophisticated tool to study the polynomial.

\subsection{The Galois correspondence and Galois extensions}
\label{subsec:fields:the-galois-correspondence-and-galois-extensions}
In the language of field theory, subsets of the roots of a polynomial
$f(x)$ determine intermediate fields of the splitting field of $f(x)$, and
groups of permutations of the roots give automorphisms of these
intermediate fields. With this in mind, it is not surprising that splitting
fields should come to the fore; the fact that one can characterize such
fields in terms of groups of automorphisms (Theorem~6.9 below) will be key
to the whole discussion. Before wading into these waters, we should
formalize the precise relation between groups of automorphisms of an
extension and intermediate fields.
\begin{defn}{}{fields:6.1}
  Let $k \subseteq F$ be a field extension, and let $G \subseteq
  \mathrm{Aut}_k(F)$ be a group of automorphisms of the extension. The
  \emph{fixed field} of $G$ is the intermediate field
  \[
    F^G := \{\alpha \in F \mid \forall g \in G,\; g\alpha = \alpha\}.
  \]
\end{defn}
The fact that $F^G$ is indeed a subfield of $F$ containing $k$ is
immediate. The notion of fixed field allows us to set up a correspondence
\[
  \{\text{intermediate fields } E : k \subseteq E \subseteq F\}
  \rightleftharpoons
  \{\text{subgroups of } \mathrm{Aut}_k(F)\},
\]
sending the intermediate field $E$ to the subgroup $\mathrm{Aut}_E(F)$ of
$\mathrm{Aut}_k(F)$ and the subgroup $G \subseteq \mathrm{Aut}_k(F)$ to
the fixed field $F^G$.
\begin{defn}{}{fields:6.2}
  This is known as the \emph{Galois correspondence}.
\end{defn}
\begin{lem}{}{fields:6.3}
  The Galois correspondence is inclusion-reversing: for all subgroups $G$
  of $\mathrm{Aut}_k(F)$ and all intermediate fields $k \subseteq E
  \subseteq F$:
  \begin{itemize}
    \item $E \subseteq F^{\mathrm{Aut}_E(F)}$;
    \item $G \subseteq \mathrm{Aut}_{F^G}(F)$.
  \end{itemize}
  Further, denote by $E_1 E_2$ the smallest subfield of $F$ containing two
  intermediate fields $E_1$, $E_2$, and denote by $\langle G_1, G_2
  \rangle$ the smallest subgroup of $\mathrm{Aut}_k(F)$ containing two
  subgroups $G_1$, $G_2$. Then
  \begin{itemize}
    \item $\mathrm{Aut}_{E_1 E_2}(F) = \mathrm{Aut}_{E_1}(F) \cap
      \mathrm{Aut}_{E_2}(F)$;
    \item $F^{\langle G_1, G_2 \rangle} = F^{G_1} \cap F^{G_2}$.
  \end{itemize}
\end{lem}
\begin{proof}
  Exercise~6.1.
\end{proof}
Of course the reader should start wondering whether there are situations
guaranteeing that the inclusions appearing in \cref{lem:fields:6.3} are
equalities, making the Galois correspondence a bijection. This is precisely
where we are heading. The first observation is that this is not always the
case:
\begin{exam}{}{fields:exam:6.4}
  Consider the extension $\Q \subseteq \Q(\sqrt[3]{2})$. Since
  $[\Q(\sqrt[3]{2}) : \Q] = 3$ is prime, the only intermediate fields are
  $\Q$ and $\Q(\sqrt[3]{2})$ (by \cref{coro:fields:1.11}). Concerning
  $\mathrm{Aut}_\Q(\Q(\sqrt[3]{2}))$: since $\sqrt[3]{2} \in \R$, we have
  an extension $\Q(\sqrt[3]{2}) \subseteq \R$; since $\sqrt[3]{2}$ is the
  only cube root of $2$ in $\R$, the minimal polynomial $t^3 - 2$ of
  $\sqrt[3]{2}$ has a single root in $\Q(\sqrt[3]{2})$. By
  \cref{coro:fields:1.7}, $\mathrm{Aut}_\Q(\Q(\sqrt[3]{2}))$ consists of a
  single element: it is trivial.

  Thus, in this example the Galois correspondence acts between a set with
  two elements and a singleton:
  \[
    \{\Q,\; \Q(\sqrt[3]{2})\}
    \rightleftharpoons
    \{\mathrm{Aut}_\Q(\Q(\sqrt[3]{2}))\} = \{e\}.
  \]
  In particular, the function associating with each intermediate field the
  corresponding automorphism group is not injective in general.
\end{exam}
This example shows that the inclusion $E \subseteq F^{\mathrm{Aut}_E(F)}$
in \cref{lem:fields:6.3} may be proper: $\Q$ is properly contained in
$\Q(\sqrt[3]{2})$, which is the fixed field of the (only) subgroup of
$\mathrm{Aut}_\Q(\Q(\sqrt[3]{2}))$. Fortunately, the situation with the
other inclusion is more constrained:
\begin{prop}{}{fields:6.5}
  Let $k \subseteq F$ be a finite extension, and let $G$ be a subgroup of
  $\mathrm{Aut}_k(F)$. Then $|G| = [F : F^G]$, and $G = \mathrm{Aut}_{F^G}(F)$.
  In particular, the Galois correspondence (from intermediate fields to
  automorphism groups) is surjective for a finite extension.
\end{prop}
Proving this fact leads us to examining very carefully finite extensions of
the type $F^G \subseteq F$. Surprisingly, these turn out to satisfy just
about every property we have encountered so far:
\begin{lem}{}{fields:6.6}
  Let $k \subseteq F$ be a finite extension, and let $G$ be a subgroup of
  $\mathrm{Aut}_k(F)$. Then $F^G \subseteq F$ is a finite, simple, normal,
  separable extension.
\end{lem}
\begin{rem}{}{fields:6.7}
  The key to this lemma is the following observation, which is worth
  highlighting. If $\alpha \in F$ and $g \in G$, note that $g\alpha$ must
  be a root of the minimal polynomial of $\alpha$ over $k$; there are only
  finitely many roots, so the $G$-orbit of $\alpha$ consists of finitely
  many elements $\alpha = \alpha_1, \dots, \alpha_n$. The group $G$ acts on
  the orbit by permuting its elements; therefore, every element of $G$
  leaves the polynomial
  \[
    q_\alpha(t) = (t - \alpha_1) \cdots (t - \alpha_n)
  \]
  fixed. In other words, the coefficients of this polynomial must be in the
  fixed field $F^G$. Further, $q_\alpha(t)$ is separable since it has
  distinct roots. Finally, note that $\deg q_\alpha(t) \leq |G|$.
\end{rem}
\begin{proof}[Proof of \cref{lem:fields:6.6}]
  The extension $F^G \subseteq F$ is finite because $k \subseteq F$ is
  finite. Let $\alpha \in F$; by \cref{rem:fields:6.7}, $\alpha$ is a root
  of a separable polynomial $q_\alpha(t)$ with coefficients in $F^G$. It
  follows that $\alpha$ is separable over $F^G$; hence the extension is
  separable according to \cref{defn:fields:4.12}.

  Since $F^G \subseteq F$ is finite and separable, it is simple by
  \cref{prop:fields:5.19}; let $\alpha$ be a generator. The polynomial
  $q_\alpha(t)$ splits in $F$, and $F$ is generated over $F^G$ by the
  roots of $q_\alpha(t)$ (indeed, $\alpha_1 = \alpha$ suffices to generate
  $F$ over $F^G$); therefore, $F$ is a splitting field for $q_\alpha(t)$
  over $F^G$ according to \cref{defn:fields:4.1}. Therefore $F^G \subseteq
  F$ is normal by \cref{thm:fields:4.8}, and we are done.
\end{proof}
\begin{proof}[Proof of \cref{prop:fields:6.5}]
  By \cref{lem:fields:6.3}, $G$ is a subgroup of $\mathrm{Aut}_{F^G}(F)$,
  and in particular $|G| \leq |\mathrm{Aut}_{F^G}(F)|$. In order to
  verify that $G = \mathrm{Aut}_{F^G}(F)$, it suffices to prove the
  converse inequality. By \cref{lem:fields:6.6}, $F = F^G(\alpha)$ for
  some $\alpha \in F$; therefore $|\mathrm{Aut}_{F^G}(F)|$ equals the
  number of distinct roots in $F$ of the minimal polynomial of $\alpha$
  over $F^G$ (by \cref{coro:fields:1.7}). With notation as in
  \cref{rem:fields:6.7}, $\alpha$ is a root of $q_\alpha(x) \in F^G[x]$;
  hence the minimal polynomial of $\alpha$ is a factor of $q_\alpha(x)$.
  Since by construction the number of roots of $q_\alpha(x)$ is $\leq
  |G|$, we obtain $|\mathrm{Aut}_{F^G}(F)| \leq |G|$, as needed.

  To verify that $[F : F^G] = |G|$, just note that $[F : F^G] =
  |\mathrm{Aut}_{F^G}(F)|$ by \cref{coro:fields:5.20}, since $F^G
  \subseteq F$ is normal and separable by \cref{lem:fields:6.6}.
\end{proof}
\begin{rem}{}{fields:6.8}
  The hypothesis that the extension is finite in \cref{prop:fields:6.5} is
  necessary: the Galois correspondence is not necessarily surjective in the
  infinite case. Not all is lost, though: one can give a suitable topology
  to the group of automorphisms and limit the Galois correspondence to
  closed subgroups of the automorphism group, recovering results such as
  \cref{prop:fields:6.5}. The reader is welcome to explore this
  further---we will not be able to take this more general point of view
  here.
\end{rem}
\begin{thm}{}{fields:6.9}
  Let $k \subseteq F$ be a finite field extension. Then the following are
  equivalent:
  \begin{enumerate}[(1)]
    \item $F$ is the splitting field of a separable polynomial
      $f(t) \in k[t]$ over $k$;
    \item $k \subseteq F$ is normal and separable;
    \item $|\mathrm{Aut}_k(F)| = [F : k]$;
    \item $k = F^{\mathrm{Aut}_k(F)}$ is the fixed field of
      $\mathrm{Aut}_k(F)$;
    \item the Galois correspondence for $k \subseteq F$ is a bijection;
    \item $k \subseteq F$ is separable, and if $F \subseteq K$ is an
      algebraic extension and $\sigma \in \mathrm{Aut}_k(K)$, then
      $\sigma(F) = F$.
  \end{enumerate}
\end{thm}
\begin{proof}
  Most of the needed implications have been proven along the way.
  $(1) \iff (2)$ by \cref{thm:fields:4.8}; $(2) \Rightarrow (3)$ by
  \cref{coro:fields:5.20}. $(3) \iff (4)$ follows from
  \cref{prop:fields:6.5}, applied to the extension $F^{\mathrm{Aut}_k(F)}
  \subseteq F$: by \cref{prop:fields:6.5}, we have
  $[F : F^{\mathrm{Aut}_k(F)}] = |\mathrm{Aut}_k(F)|$; since
  $k \subseteq F^{\mathrm{Aut}_k(F)} \subseteq F$, it follows that
  $k = F^{\mathrm{Aut}_k(F)}$ if and only if $|\mathrm{Aut}_k(F)| =
  [F : k]$. $(4) \Rightarrow (2)$ by \cref{lem:fields:6.6}.

  $(2) \iff (6)$ holds by Exercise~4.7, since finite separable extensions
  are simple (\cref{prop:fields:5.19}).

  To prove $(5) \Rightarrow (4)$, let $E = F^{\mathrm{Aut}_k(F)}$; by
  \cref{prop:fields:6.5}, $\mathrm{Aut}_E(F) = \mathrm{Aut}_k(F)$.
  Assuming that the Galois correspondence is bijective, it follows that
  $k = E = F^{\mathrm{Aut}_k(F)}$, giving (4).

  Finally, we prove that $(1) \Rightarrow (5)$. Since $k \subseteq F$ is a
  finite extension, we already know that the Galois correspondence from
  intermediate fields to subgroups of $\mathrm{Aut}_k(F)$ has a right
  inverse (\cref{prop:fields:6.5}), so it suffices to show it has a
  left-inverse. Therefore, it suffices to verify that every intermediate
  field $E$ equals the fixed field of the corresponding subgroup
  $\mathrm{Aut}_E(F)$. Let $E$ be an intermediate field. By (1), $F$ is
  the splitting field of a separable polynomial $f(t) \in k[t] \subseteq
  E[t]$; therefore, condition (1) holds for the extension $E \subseteq F$.
  As we have already proved that $(1) \iff (4)$, this implies that
  $E = F^{\mathrm{Aut}_E(F)}$, and we are done.
\end{proof}
\begin{defn}{}{fields:6.10}
  A finite extension $k \subseteq F$ is \emph{Galois} if it satisfies any
  of the conditions listed in \cref{thm:fields:6.9}.
\end{defn}
\emph{Warning:} We will not study the Galois condition for infinite
extensions (cf.\ \cref{rem:fields:6.8}). Thus Galois extensions will
implicitly be finite in what follows.

Condition (6) is technically useful and helps with visualizing the Galois
condition. If a finite separable extension $k \subseteq F$ is not Galois,
then $F$ can be embedded in some larger extension $k \subseteq K$ (for
example, in the algebraic closure $k \subseteq \bar{k}$) in many possible
ways, with different images $\sigma(F)$. If $k \subseteq F$ is Galois, all
these images must coincide: $F = \sigma(F)$. Of course there are possibly
still many ways\footnote{In fact, precisely $[F : k]_s = [F : k] =
|\mathrm{Aut}_k(F)|$ if $K = \bar{k}$ and hence for all $K \supseteq F$.}
to embed $F$ in $K$, but they all have the same image $F$.

The mysterious comments at the end of Example~1.4 are now hopefully clear.
Galois extensions of a field $k$ are well-defined subfields of $\bar{k}$,
preserved by automorphisms of $\bar{k}$ over $k$. Contrast this situation
with the non-Galois extension $\Q \subseteq \Q(\sqrt[3]{2})$, which admits
three different embeddings in $\bar{\Q}$. For example, it does not make
sense to talk about `the' composite of $\Q(\sqrt[3]{2})$ and
$\Q(\sqrt[3]{2})$, as a subfield of (for instance) $\C$: according to the
chosen embedding of $\Q(\sqrt[3]{2})$, this may or may not be a subfield
of $\R$. But it does make sense to talk about the composite of two Galois
extensions $k \subseteq F_1$, $k \subseteq F_2$, since both $F_1$ and
$F_2$ are well-defined as subfields of $\bar{k}$, so their composite
$F_1 F_2$ is independent of the choice of embeddings $F_1 \subseteq
\bar{k}$, $F_2 \subseteq \bar{k}$.
\begin{defn}{}{fields:6.11}
  If $k \subseteq F$ is a Galois extension, the corresponding automorphism
  group $\mathrm{Aut}_k(F)$ is called the \emph{Galois group} of the
  extension.
\end{defn}
To reiterate, the extension $\Q \subseteq \Q(\sqrt[3]{2})$ is not a Galois
extension: as we have observed in Example~6.4, $\mathrm{Aut}_\Q(\Q(
\sqrt[3]{2}))$ is trivial (hence, for example, condition (3) of
\cref{thm:fields:6.9} does not hold).

On the other hand, we have already run into several examples of Galois
extensions: the examples of splitting fields we have seen in $\S$4.1 are
all Galois extensions. The `Galois fields' of $\S$5.1 are (surprise,
surprise) Galois extensions of their prime subfields, by
\cref{thm:fields:5.1}. The cyclotomic fields ($\S$5.2) are Galois
extensions of $\Q$.

A Galois extension $k \subseteq F$ is \emph{cyclic}, \emph{abelian},
etc., if the corresponding group of automorphisms $\mathrm{Aut}_k(F)$ is
cyclic, abelian, etc. For example, $\Q \subseteq \Q(\zeta_8)$ is an
abelian, but not cyclic, Galois extension (cf.\ \cref{prop:fields:5.16}
and Example~5.17).

\subsection{The fundamental theorem of Galois theory, I}
\label{subsec:fields:the-fundamental-theorem-of-galois-theory-i}
The fundamental theorem of Galois theory amounts to a more complete
description of the Galois correspondence for Galois extensions. At this
stage, this is what we know:\footnote{`Separability' does not play a role
in these considerations; thus they best convey `Galoisness' in, say,
characteristic $0$.}
\begin{itemize}
  \item If $k \subseteq F$ is a (finite) Galois extension, then there is
    an inclusion-reversing bijection
    \[
      \{\text{intermediate fields } E : k \subseteq E \subseteq F\}
      \leftrightarrow
      \{\text{subgroups of } \mathrm{Aut}_k(F)\}:
    \]
    through this bijection, the intermediate field $E$ corresponds to the
    subgroup $\mathrm{Aut}_E(F)$ of $\mathrm{Aut}_k(F)$, and the subgroup
    $G \subseteq \mathrm{Aut}_k(F)$ corresponds to the fixed field $F^G$.
  \item For every intermediate field $E$,
    $[F : E] = |\mathrm{Aut}_E(F)|$. The extension $E \subseteq F$ is
    then also a Galois extension (Exercise~6.3), and
    \[
      [E : k] = [\mathrm{Aut}_k(F) : \mathrm{Aut}_E(F)]:
    \]
    this follows from Lagrange's theorem (\cref{coro:fields:II.8.14}) and
    its field theory counterpart (\cref{prop:fields:1.10}).
  \item The extension $k \subseteq E$ is not necessarily Galois. (Once
    more $\Q(\sqrt[3]{2})$ gives a counterexample, since it can be viewed
    as intermediate in the splitting field of $t^3 - 2$ over $\Q$.)
\end{itemize}
The fact that the Galois correspondence is a bijection may be upgraded as
follows:
\begin{thm}{}{fields:6.12}
  Let $k \subseteq F$ be a Galois extension. The Galois correspondence is
  an inclusion-reversing isomorphism of the lattice of intermediate
  subfields of $k \subseteq F$ with the lattice of subgroups of
  $\mathrm{Aut}_k(F)$. That is (with notation as in \cref{lem:fields:6.3}),
  if $E_1$, $E_2$ are intermediate fields and $G_1$, $G_2$ are the
  corresponding subgroups of $\mathrm{Aut}_k(F)$, then $E_1 \cap E_2$
  corresponds to $\langle G_1, G_2 \rangle$ and $E_1 E_2$ corresponds to
  $G_1 \cap G_2$.
\end{thm}
\begin{proof}
  This follows immediately from \cref{thm:fields:6.9} and
  \cref{lem:fields:6.3}, which gives
  \[
    \mathrm{Aut}_{E_1 E_2}(F) = G_1 \cap G_2,
    \qquad
    F^{\langle G_1, G_2 \rangle} = E_1 \cap E_2,
  \]
  as needed.
\end{proof}
In this context, it is common to use the `vertical' depiction of field
extensions, with a parallel notation for subgroups: the field $F$ at the
top corresponds to $\{e\}$, an intermediate field $E$ to $\mathrm{Aut}_E(F)$,
and $k$ at the bottom to $\mathrm{Aut}_k(F)$.

The content of \cref{thm:fields:6.12} is that the lattice of intermediate
fields of a Galois extension $k \subseteq F$ and the lattice of subgroups
of the corresponding Galois group are identical.
\begin{exam}{}{fields:exam:6.13}
  For finite fields, this coincidence of lattices was essentially proven
  `by hand' in \cref{coro:fields:5.4} and \cref{prop:fields:5.8}. For
  example, the extension $\mathbb{F}_2 \subseteq \mathbb{F}_{64}$ is
  Galois, with cyclic Galois group $C_6$, generated by the Frobenius
  automorphism $\phi$. The divisors of $6$ are $1, 2, 3, 6$, corresponding
  to subgroups $\{e\}$, $\{e, \phi^3\}$, $\{e, \phi^2, \phi^4\}$, and
  $C_6$; and to intermediate fields $\mathbb{F}_{64}$, $\mathbb{F}_8$,
  $\mathbb{F}_4$, and $\mathbb{F}_2$, respectively.
\end{exam}
\cref{thm:fields:6.12} is often used in the `groups to fields' direction:
finding the lattice of subgroups of a finite group is essentially a
combinatorial problem, and through the Galois correspondence it determines
the lattice of intermediate fields of a corresponding Galois extension.
\begin{exam}{}{fields:exam:6.14}
  The extension $\Q(\sqrt{2}, \sqrt{3}) = \Q(\sqrt{2} + \sqrt{3})$ studied
  in Example~1.19 is the splitting field of the polynomial $t^4 - 10t^2 +
  1$, so it is Galois. We found that its Galois group is $\Z/2\Z \times
  \Z/2\Z$; the lattice of subgroups of this group is: $\{(0,0)\}$ at the
  top, with three index-2 subgroups $\langle(1,0)\rangle$,
  $\langle(0,1)\rangle$, $\langle(1,1)\rangle$ in the middle, and
  $\Z/2\Z \times \Z/2\Z$ at the bottom. Therefore the lattice of
  intermediate fields is: $\Q(\sqrt{2}, \sqrt{3})$ at the top, with three
  degree-2 extensions $\Q(\sqrt{2})$, $\Q(\sqrt{3})$, $\Q(\sqrt{6})$ in
  the middle, and $\Q$ at the bottom. (The intermediate fields are
  determined by recalling the generators of the corresponding subgroups, as
  found in Example~1.19: the flip $\sqrt{3} \mapsto -\sqrt{3}$ leaves
  $\Q(\sqrt{2})$ fixed, etc.) Galois theory tells us that there is no other
  intermediate field.
\end{exam}

\subsection{The fundamental theorem of Galois theory, II}
\label{subsec:fields:the-fundamental-theorem-of-galois-theory-ii}
Another popular notational device is to mark an extension by the
corresponding Galois group, if the extension is Galois: one writes the
group label $\mathrm{Aut}_k(F)$ on the segment from $k$ to $F$, and
$\mathrm{Aut}_E(F)$ on the segment from $E$ to $F$.

Example~6.4 shows that the extension $k \subseteq E$ need not be a Galois
extension: $\Q(\sqrt[3]{2})$ is an intermediate field for the extension of
$\Q$ into the splitting field for the polynomial $t^3 - 2$ (which is
Galois, by part (1) of \cref{thm:fields:6.9}), but as we have seen, it is
not Galois. The splitting field has degree $6$ over $\Q$ and is a quadratic
extension of $\Q(\sqrt[3]{2})$. This is in fact the typical situation:
every finite separable extension may be enlarged to a Galois extension
(Exercise~6.4).

In any case, the question of whether the extension of the base field $k$
in an intermediate field $E$ of a Galois extension $k \subseteq F$ is
Galois is central to the theory. The outcome of the discussion will be very
neat, and this is possibly the most striking part of the fundamental theorem
of Galois theory:
\begin{thm}{}{fields:6.15}
  Let $k \subseteq F$ be a Galois extension, and let $E$ be an intermediate
  field. Then $k \subseteq E$ is Galois if and only if $\mathrm{Aut}_E(F)$
  is normal in $\mathrm{Aut}_k(F)$; in this case, there is an isomorphism
  \[
    \mathrm{Aut}_k(E) \cong \frac{\mathrm{Aut}_k(F)}{\mathrm{Aut}_E(F)}.
  \]
\end{thm}
The `problem' with $\Q \subseteq \Q(\sqrt[3]{2})$ is that the automorphism
group of the splitting field of $t^3 - 2$ is $S_3$ (as we will soon be
able to verify: cf.\ Example~7.19). The subgroup corresponding to
$\Q(\sqrt[3]{2})$ has index $3$; hence (as we know, since we are well
acquainted with $S_3$) it is not normal.

The fact that the Galois group of $k \subseteq E$ will turn out to be
isomorphic to the quotient $\mathrm{Aut}_k(F)/\mathrm{Aut}_E(F)$ in the
Galois case should not be too surprising, since we know already that in
any case $[E : k]$ equals the index of $\mathrm{Aut}_E(F)$ in
$\mathrm{Aut}_k(F)$, and the index equals the order of the quotient if the
subgroup is normal. In general, $[E : k]$ equals the number of cosets of
$\mathrm{Aut}_E(F)$ in $\mathrm{Aut}_k(F)$, and it is worth shooting for
a concrete interpretation of these cosets, even when $k \subseteq E$ does
not turn out to be a Galois extension.
We already know that the number of left-cosets of $\mathrm{Aut}_E(F)$
equals the index of $\mathrm{Aut}_E(F)$ in $\mathrm{Aut}_k(F)$, hence
$[E : k]$, hence $[E : k]_s$ (because $k \subseteq E$ is separable, as a
subextension of a separable extension), and hence there must exist a
bijection between the set of left-cosets $g\,\mathrm{Aut}_E(F)$ as $g$
ranges in $\mathrm{Aut}_k(F)$ and the set of different embeddings
$\iota : E \to \bar{k}$ in an algebraic closure of $k$, extending
$\mathrm{id}_k$. Not surprisingly, there is a reason for this---that is,
a `meaningful' bijection between these two sets.
This is an excellent moment to review the material on group actions we
covered in \S{}II.9, especially \S{}II.9.3. Recall in particular that we
were able to classify all transitive actions of a group $G$ on a set $S$
(Proposition~II.9.9): these $G$-sets are all isomorphic to the set of
left-cosets of a subgroup $H$ of $G$ under the natural action of $G$; and
$H$ may be taken to be the stabilizer of any element of $S$.
We are going to define a natural action of $\mathrm{Aut}_k(F)$ on the set
$I$ of embeddings of $E$ in $\bar{k}$ extending the identity, and prove
that, as an $\mathrm{Aut}_k(F)$-set, $I$ is isomorphic to the set of
left-cosets of $\mathrm{Aut}_E(F)$ in $\mathrm{Aut}_k(F)$. This will be
our `meaningful bijection', and after this is done Theorem~6.15 will look
perfectly reasonable.
Fix an embedding $F \subseteq \bar{k}$ of $F$ in an algebraic closure of
$k$, and let $I$ be the set of embeddings $\iota : E \to \bar{k}$
extending the identity on $k$. It is a good idea to view $k \subseteq E$
as an abstract extension, rather than identifying $E$ with a specific
intermediate field in $F$; the fact that we can view $E$ as an
intermediate field of the extension $k \subseteq F$ simply means that
$\iota(E) \subseteq F$ for some $\iota \in I$.
\begin{itemize}
  \item For all $\iota \in I$, $\iota(E) \subseteq F$.

    Indeed, $k \subseteq E$ is simple (finite and separable $\Rightarrow$
    simple, Proposition~5.19). If $E = k(\alpha)$, then $\iota$ is
    determined by $\iota(\alpha)$, which is necessarily a root of the
    minimal polynomial of $\alpha$ over $k$. But $k \subseteq F$ is normal
    and contains some root $\iota(\alpha)$ of this polynomial; hence $F$
    contains all of them (Definition~4.7). That is, $\iota(\alpha) \in F$
    for all $\iota \in I$, implying $\iota(k(\alpha)) \subseteq F$ for all
    $\iota \in I$.
  \item Let $\iota \in I$, and let $g \in \mathrm{Aut}_k(F)$; then
    $g \circ \iota \in I$. Therefore, $\mathrm{Aut}_k(F)$ acts on $I$.

    Here $g \circ \iota$ is interpreted in the following way:
    $\iota(E) \subseteq F$, as we have seen; hence $g$ restricts to a
    homomorphism $\iota(E) \to F$, and $g \circ \iota$ is the composition
    of this homomorphism with $\iota$.
  \item The action of $\mathrm{Aut}_k(F)$ on $I$ is transitive.

    This follows from the `uniqueness of splitting fields', Lemma~4.2.
    Indeed, let $\iota_1, \iota_2 \in I$. Both $\iota_1(E)$, $\iota_2(E)$
    contain $k$, and $F$ is a splitting field over $k$ (part~(1) of
    Theorem~6.9); hence $F$ is a splitting field (for the same polynomial)
    over both $\iota_1(E)$ and $\iota_2(E)$. By Lemma~4.2, there exists
    an automorphism $g : F \to F$ extending the isomorphism
    $\iota_2 \circ \iota_1^{-1} : \iota_1(E) \to \iota_2(E)$.
    Then $g \in \mathrm{Aut}_k(F)$ (since
    $\iota_2 \circ \iota_1^{-1}$ restricts to the identity on $k$), and
    the fact that $g$ restricts to $\iota_2 \circ \iota_1^{-1}$ means
    precisely that $g \circ \iota_1 = \iota_2$.
  \item As an $\mathrm{Aut}_k(F)$-set, $I$ is isomorphic to the set of
    left-cosets of $\mathrm{Aut}_E(F)$ in $\mathrm{Aut}_k(F)$.
\end{itemize}

Now choose one $\iota \in I$, and use it to identify $E$ with an
intermediate field of $k \subseteq F$. Then
$\mathrm{Aut}_E(F) = \mathrm{Aut}_{\iota(E)}(F)$ is the subgroup of
$\mathrm{Aut}_k(F)$ consisting of those elements $g$ which restrict to the
identity on $E = \iota(E)$, that is, the stabilizer of $\iota$. At this
point, Proposition~II.9.9 is all that is needed to conclude the promised
result in the last bullet above.

I view this observation as an integral part of the fundamental theorem of
Galois theory. The conventional statement (Theorem~6.15) is an essentially
immediate consequence of this fact.
\begin{proof}[Proof of \cref{thm:fields:6.15}]
  As observed above, the $\mathrm{Aut}_k(F)$-stabilizer of $\iota \in I$
  equals $\mathrm{Aut}_{\iota(E)}(F)$. By Proposition~II.9.12,
  stabilizers of different $\iota$ are conjugates of each other; if
  $\mathrm{Aut}_E(F)$ is normal, it follows that for all $\iota \in I$,
  $\mathrm{Aut}_{\iota(E)}(F) = \mathrm{Aut}_E(F)$.
  This implies that $\iota(E) = E$ for all $\iota \in I$, since the Galois
  correspondence is bijective for the Galois extension $k \subseteq F$. It
  follows that $k \subseteq E$ satisfies condition~(6) of Theorem~6.9;
  hence it is Galois.

  Conversely, assume that $k \subseteq E$ is Galois; by the same
  condition~(6), $\iota(E) = E$ for all $\iota \in I$. Restriction to
  $E = \iota(E)$ then defines a homomorphism
  $\rho : \mathrm{Aut}_k(F) \to \mathrm{Aut}_k(E)$.
  This homomorphism is surjective (because the $\mathrm{Aut}_k(F)$-action
  on $I$ is transitive), and its kernel consists of those
  $g \in \mathrm{Aut}_k(F)$ which restrict to the identity on $E$, that
  is, precisely of $\mathrm{Aut}_E(F)$. This shows that $\mathrm{Aut}_E(F)$
  is normal and establishes the stated isomorphism by virtue of the first
  isomorphism theorem for groups, Corollary~II.8.2.
\end{proof}
\begin{rem}{}{fields:6.16}
  There is a parallel between Galois theory and the theory of covering
  spaces in topology. In this analogy, Galois extensions correspond to
  regular covers; the Galois group of an extension corresponds to the group
  of deck transformations; and Theorem~6.15 corresponds to the fact that
  the quotient of a regular cover by a normal subgroup of the group of deck
  transformations is again a regular cover. More general (connected) covers
  correspond to more general algebraic extensions. A space is simply
  connected if and only if it admits no nontrivial connected covers, so
  this notion corresponds in field theory to the condition that a field $K$
  admits no nontrivial algebraic extensions, that is, that $K$ is
  algebraically closed. Viewing the fundamental group of a space as the
  group of deck transformations of its fundamental cover suggests that we
  should think of the Galois group of the algebraic closure
  $k \subseteq \bar{k}$ as `the fundamental group' of a field
  $k$.\footnote{In algebraic geometry this analogy is carried out to its
  natural consequences. A covering map of algebraic varieties $X \to Y$
  determines a field extension $K(Y) \subseteq K(X)$, where $K(X)$,
  $K(Y)$ are the `fields of rational functions' (in the affine case, these
  are just the fields of fractions of the corresponding coordinate rings;
  cf.\ Exercise~2.16). One can then use Galois theory to transfer to the
  algebro-geometric environment notions such as the fundamental group,
  without appealing to topological notions (such as `continuous maps from
  $S^1$'), which would be problematic in, e.g., positive characteristic.}
\end{rem}

\subsection{Further remarks and examples}
\label{subsec:fields:further-remarks-and-examples}
Suppose $k \subseteq F$ and $k \subseteq K$ are two finite extensions
contained in a larger extension of $k$ (for example, we could choose
specific embeddings $F \subseteq \bar{k}$, $K \subseteq \bar{k}$ in an
algebraic closure of $k$). Then we can consider the composite $KF$ (cf.\
Lemma~6.3) of $F$, $K$ in the larger extension. It is natural to question
how the Galois condition behaves under this operation.
\begin{prop}{}{fields:6.17}
  Suppose $k \subseteq F$ is a Galois extension and $k \subseteq K$ is any
  finite extension. Then $K \subseteq KF$ is a Galois extension, and
  $\mathrm{Aut}_K(KF) \cong \mathrm{Aut}_{F \cap K}(F)$.
\end{prop}

The situation is depicted in the lattice:
\[
  \begin{array}{ccc}
    KF & & \\
    \big\uparrow & \nwarrow & \\
    F & & K \\
    & \nwarrow & \big\uparrow \\
    & & k
  \end{array}
  \qquad
  \begin{array}{c}
    \mathrm{Aut}_K(KF) \cong \mathrm{Aut}_{F \cap K}(F)
  \end{array}
\]
\begin{proof}
  As $k \subseteq F$ is Galois, it is the splitting field of a separable
  polynomial $f(x) \in k[x] \subseteq K[x]$. The roots of $f(x)$ generate
  $F$ over $k$, so they generate $KF$ over $K$; in other words, $KF$ is
  the splitting field of $f(x)$ over $K$, so $K \subseteq KF$ is Galois.

  If $\sigma : KF \to KF$ is an automorphism extending the identity on $K$
  (and hence on $k$), then $\sigma$ restricts to a homomorphism
  $F \to \sigma(F)$ extending the identity on $k$. Since $k \subseteq F$
  is Galois, $\sigma(F) = F$ (condition~(6) in Theorem~6.9). Thus,
  `restriction' defines a natural group homomorphism
  $\rho : \mathrm{Aut}_K(KF) \to \mathrm{Aut}_k(F)$.

  I claim that $\rho$ is injective. Indeed, by definition every
  $\sigma \in \mathrm{Aut}_K(KF)$ is the identity on $K$; if $\sigma$
  restricts to the identity on $F$, then $\sigma$ is the identity on the
  composite $KF$; that is, $\sigma = \mathrm{id}$ in
  $\mathrm{Aut}_K(KF)$.

  Therefore, $\mathrm{Aut}_K(KF)$ may be identified with a subgroup $G$
  of $\mathrm{Aut}_k(F)$. Since every $\sigma \in \mathrm{Aut}_K(KF)$
  fixes $K$, the restriction $\rho(\sigma)$ fixes $F \cap K$; that is,
  $F^G \supseteq F \cap K$. Conversely, suppose $\alpha \in F^G$; then
  $K(\alpha)$ is in the fixed field of $\mathrm{Aut}_K(KF)$, which is $K$
  itself since $K \subseteq KF$ is Galois (condition~(4) of
  Theorem~6.9). That is, $\alpha \in K$, and it follows that
  $F^G = F \cap K$. By Proposition~6.5, $G = \mathrm{Aut}_{F \cap K}(F)$,
  concluding the proof.
\end{proof}
I will leave to the reader the pleasure of verifying that if both
$k \subseteq F$ and $k \subseteq K$ are Galois, then $KF$ is also Galois
over $k$, and so is $F \cap K$ (Exercise~6.13).
\begin{exam}{}{fields:exam:6.18}
  We have studied cyclotomic fields $\Q(\zeta_n)$ as extensions of $\Q$;
  $\Q(\zeta_n)$ is the splitting field of $x^n - 1$, so these extensions
  are Galois; we have proved (Proposition~5.16) that
  $\mathrm{Aut}_\Q(\Q(\zeta_n))$ is isomorphic to the group of units in
  $\Z/n\Z$.

  Now let $k$ be any field of characteristic zero. The splitting field of
  $x^n - 1$ over $k$ is the composite $k(\zeta)$ of $k$ and $\Q(\zeta)$.
  By Proposition~6.17 the extension $k \subseteq k(\zeta)$ is Galois, and
  $\mathrm{Aut}_k(k(\zeta))$ is isomorphic to a subgroup of the group of
  units of $\Z/n\Z$. In particular, it is abelian.
\end{exam}
The previous example gives an important class of abelian Galois
extensions. Remarkably, we can classify all cyclic Galois extensions, if
we impose some restriction on the base field $k$.

\begin{prop}{}{fields:6.19}
  Let $k \subseteq F$ be an extension of degree $m$. Assume that $k$
  contains a primitive $m$-th root of $1$ and $\mathrm{char}\,k$ does not
  divide\footnote{We will apply this result in \S{}7.4; for convenience,
  we will take $\mathrm{char}\,k = 0$ in that application.} $m$. Then
  $k \subseteq F$ is Galois and cyclic if and only if $F = k(\delta)$,
  with $\delta^m \in k$.
\end{prop}

That is, under the given hypotheses on $k$, the cyclic Galois extensions
of $k$ are precisely those of the form $k \subseteq k(\sqrt[m]{c})$, with
$c \in k$. This is part of Kummer theory, a basic tool in the study of
abelian extensions. Proposition~6.19 will be a key ingredient in our
application of Galois theory to the solvability of polynomial equations.
\begin{proof}
  Let $\zeta \in k$ be a primitive $m$-th root of $1$.

  First assume that $F = k(\delta)$, with $\delta^m = c \in k$. Then all
  $m$ roots of the polynomial $x^m - c$,
  \[
    \delta,\; \zeta\delta,\; \zeta^2\delta,\; \dots,\; \zeta^{m-1}\delta,
  \]
  are in $F$, and $F$ is generated by them; therefore $F$ is the splitting
  field of the separable (since $\mathrm{char}\,k$ does not divide $m$)
  polynomial $x^m - c$ over $k$; hence (Theorem~6.9, part~(1))
  $k \subseteq F$ is Galois. To see that $\mathrm{Aut}_k(F)$ is cyclic,
  note that every $\phi \in \mathrm{Aut}_k(F)$ is determined by
  $\phi(\delta)$, which is necessarily a root of $x^m - c$; therefore
  $\phi(\delta) = \zeta^i \delta$ for some $i$, determined up to a
  multiple of $m$. This defines an isomorphism of $\mathrm{Aut}_k(F)$
  with $\Z/m\Z$, as is immediately verified.
  Conversely, assume that $k \subseteq F$ is Galois and
  $\mathrm{Aut}_k(F)$ is cyclic, generated by $\phi$. The automorphisms
  $\phi^i$, $i = 0, \dots, m-1$, are pairwise distinct; thus they are
  linearly independent over $F$ (Exercise~6.14). Therefore, there exists
  an $\alpha \in F$ such that\footnote{This is called a Lagrange
  resolvent.}
  \[
    \delta := \alpha + \zeta^{-1}\phi(\alpha) + \dots
    + \zeta^{-(m-2)}\phi^{m-2}(\alpha)
    + \zeta^{-(m-1)}\phi^{m-1}(\alpha) \neq 0.
  \]
  Note that
  \[
    \phi(\delta) = \phi(\alpha) + \zeta^{-1}\phi^2(\alpha) + \dots
    + \zeta^{-(m-2)}\phi^{m-1}(\alpha) + \zeta^{-(m-1)}\alpha
    = \zeta\delta
  \]
  (since $\zeta \in k$, so $\phi(\zeta) = \zeta$). This implies that
  $\delta$ is not fixed by any nonidentity element of $\mathrm{Aut}_k(F)$;
  that is, $\mathrm{Aut}_{k(\delta)}(F) = \{e\}$; that is,
  $k(\delta) = F$. Further,
  \[
    \phi(\delta^m) = (\phi(\delta))^m = \zeta^m \delta^m = \delta^m;
  \]
  that is, $\delta^m$ is fixed by $\phi$, and hence by the whole of
  $\mathrm{Aut}_k(F)$. Since $k \subseteq F$ is Galois, this implies
  $\delta^m \in k$ (Theorem~6.9, part~(4)). That is, $\delta$ satisfies
  the stated conditions.
\end{proof}

\subsection*{Exercises}

\begin{exer}{}{fields:exer:6.1}
  Prove Lemma~6.3. [\S{}6.1]
\end{exer}
\begin{exer}{}{fields:exer:6.2}
  Prove that quadratic extensions in characteristic $\neq 2$ are Galois.
\end{exer}
\begin{exer}{}{fields:exer:6.3}
  Let $k \subseteq F$ be a Galois extension, and let $E$ be an
  intermediate field. Prove that $E \subseteq F$ is a Galois extension.
  [\S{}6.2]
\end{exer}
\begin{exer}{}{fields:exer:6.4}
  Let $k \subseteq E$ be a finite separable extension. Prove that $E$ may
  be identified with an intermediate field of a Galois extension
  $k \subseteq F$ of $k$. In fact, prove that there is a smallest such
  extension $k \subseteq F$, in the sense that if $k \subseteq E
  \subseteq K$, with $k \subseteq K$ Galois, then there exists an
  embedding of $F$ in $K$ which is the identity on $E$. (The extension
  $k \subseteq F$ is the \emph{Galois closure} of the extension
  $k \subseteq E$. It is clearly uniquely determined up to isomorphism.)
  [\S{}6.3, 6.5]
\end{exer}
\begin{exer}{}{fields:exer:6.5}
  We have proved (Proposition~5.19) that all finite separable extensions
  $k \subseteq E$ are simple. Let $k \subseteq F$ be a Galois closure of
  $k \subseteq E$ (Exercise~6.4). For $\alpha \in E$, prove that
  $E = k(\alpha)$ if and only if $\alpha$ is moved by all
  $\sigma \in \mathrm{Aut}_k(F) \setminus \mathrm{Aut}_E(F)$.
\end{exer}
\begin{exer}{}{fields:exer:6.6}
  \begin{enumerate}[label=(\roman*)]
    \item Prove that $\Q \subseteq \Q(\sqrt{2 + \sqrt{2}})$ is Galois,
      with cyclic Galois group. (Cf.\ Exercise~1.25.)
    \item Prove that $\Q \subseteq \Q(\sqrt{3 + \sqrt{5}})$ is Galois and
      its Galois group is isomorphic to $(\Z/2\Z) \times (\Z/2\Z)$.
    \item Prove that $\Q \subseteq \Q(\sqrt{1 + \sqrt{2}})$ is not Galois,
      and compute its Galois closure $\Q \subseteq F$. Prove that
      $\mathrm{Aut}_\Q(F) \cong D_8$. (Use Exercise~IV.2.16.)
  \end{enumerate}
\end{exer}
\begin{exer}{}{fields:exer:6.7}
  Let $p > 0$ be prime, and let $d \mid e$ be positive integers, so that
  there is an extension $\F_{p^d} \subseteq \F_{p^e}$. Prove that
  $\mathrm{Aut}_{\F_{p^d}}(\F_{p^e})$ is cyclic, and describe a generator
  of this group. (This generalizes Proposition~5.8; use Galois theory.)
\end{exer}
\begin{exer}{}{fields:exer:6.8}
  Let $k \subseteq F$ be a Galois extension of degree $n$, and let $E$ be
  an intermediate field. Assume that $[E : k]$ is the smallest prime
  dividing $n$. Prove $k \subseteq E$ is Galois.
\end{exer}
\begin{exer}{}{fields:exer:6.9}
  Let $k \subseteq F$ be a Galois extension of degree $75$. Prove that
  there exists an intermediate field $E$, with $k \subsetneq E \subsetneq
  F$, such that the extension $k \subseteq E$ is Galois.
\end{exer}
\begin{exer}{}{fields:exer:6.10}
  Let $k \subseteq F$ be a Galois extension and $E$ an intermediate field.
  Prove that the normalizer of $\mathrm{Aut}_E(F)$ in $\mathrm{Aut}_k(F)$
  is the set of $\sigma \in \mathrm{Aut}_k(F)$ such that
  $\sigma(E) \subseteq E$. Use this to give an alternative proof of the
  fact that $E$ is Galois over $k$ if and only if $\mathrm{Aut}_E(F)$ is
  normal in $\mathrm{Aut}_k(F)$.
\end{exer}
\begin{exer}{}{fields:exer:6.11}
  Let $k \subseteq E$ and $E \subseteq F$ be Galois extensions.
  \begin{enumerate}[label=(\roman*)]
    \item Find an example showing that $k \subseteq F$ is not necessarily
      Galois.
    \item Prove that if every $\sigma \in \mathrm{Aut}_k(E)$ is the
      restriction of an element of $\mathrm{Aut}_k(F)$, then
      $k \subseteq F$ is Galois.
  \end{enumerate}
\end{exer}
\begin{exer}{}{fields:exer:6.12}
  Find two algebraic extensions $k \subseteq F$, $k \subseteq K$ and
  embeddings $F \subseteq \bar{k}$, $\sigma_1 : K \to \bar{k}$,
  $\sigma_2 : K \to \bar{k}$ extending $k \subseteq \bar{k}$, such that
  the composites $F\sigma_1(K)$, $F\sigma_2(K)$ are not isomorphic.
  Prove that no such example exists if $F$ and $K$ are Galois over $k$.
\end{exer}
\begin{exer}{}{fields:exer:6.13}
  Let $k \subseteq F$ and $k \subseteq K$ be Galois extensions, and assume
  $F$ and $K$ are subfields of a larger field. Prove that $k \subseteq FK$
  and $k \subseteq F \cap K$ are both Galois extensions. [\S{}6.4,
  \S{}7.4]
\end{exer}
\begin{exer}{}{fields:exer:6.14}
  Let $k \subseteq F$ be a field extension, and let
  $\phi_1, \dots, \phi_m \in \mathrm{Aut}_k(F)$ be pairwise distinct
  automorphisms. Prove that $\phi_1, \dots, \phi_m$ are linearly
  independent over $F$. (Recycle/adapt the hint for
  Exercise~VI.6.15.) [\S{}6.4, 6.16, \S{}IX.7.6]
\end{exer}
\begin{exer}{}{fields:exer:6.15}
  Let $k \subseteq F$ be a Galois extension, and let $\alpha \in F$. Prove
  that
  \[
    N_{k \subseteq F}(\alpha) = \prod_{\sigma \in \mathrm{Aut}_k(F)}
    \sigma(\alpha), \qquad
    \mathrm{tr}_{k \subseteq F}(\alpha) = \sum_{\sigma \in \mathrm{Aut}_k(F)}
    \sigma(\alpha).
  \]
  (Exercise~4.19.)
\end{exer}
\begin{exer}{}{fields:exer:6.16}
  Let $k \subseteq F$ be a cyclic Galois extension of degree $d$, and let
  $\phi$ be a generator of $\mathrm{Aut}_k(F)$. Let $\alpha \in F$ be an
  element such that $N_{k \subseteq F}(\alpha) = 1$.
  \begin{enumerate}[label=(\roman*)]
    \item Prove that the automorphisms $\mathrm{id}_F, \phi, \dots,
      \phi^{d-1}$ are linearly independent over $F$. (Exercise~6.14.)
    \item Prove that there exists a $\gamma \in F$ such that
      \[
        \beta := \gamma + \alpha\phi(\gamma)
        + \alpha\phi(\alpha)\phi^2(\gamma) + \dots
        + \alpha\phi(\alpha)\cdots\phi^{d-2}(\alpha)\phi^{d-1}(\gamma)
        \neq 0.
      \]
    \item Prove that
      $\alpha\phi(\alpha)\phi^2(\alpha)\cdots\phi^{d-1}(\alpha)\phi^d(\gamma)
      = \gamma$, and deduce that $\alpha = \beta/\phi(\beta)$.
  \end{enumerate}
  Together with the result of Exercise~4.20, the conclusion is that an
  element $\alpha$ of a cyclic Galois extension as above has norm $1$ if
  and only if there exists a $\beta$ such that $\alpha = \beta/\phi(\beta)$.
  This is \emph{\name{Hilbert}'s theorem 90} (the 90th theorem in
  \name{Hilbert}'s \emph{Zahlbericht}, a report on the state of number
  theory at the end of the nineteenth century commissioned by the German
  Mathematical Society). [6.17, \S{}IX.7.6, IX.7.18]
\end{exer}
\begin{exer}{}{fields:exer:6.17}
  Exercise~6.16 should remind the reader of the proof of
  Proposition~6.19, and for good reasons. Assume that $k \subseteq F$ is
  a cyclic Galois extension of degree $m$ and that $k$ contains a
  primitive $m$-th root $\zeta$ of $1$. Use Hilbert's theorem 90 to prove
  that $F = k(\delta)$ for a $\delta$ such that $\delta^m \in k$, thereby
  recovering one direction of the statement of Proposition~6.19. (Hint:
  What is $N_{k \subseteq F}(\zeta^{-1})$?) The case of Hilbert's theorem
  90 needed here is due to \name{Kummer}.
\end{exer}
\begin{exer}{}{fields:exer:6.18}
  Use Hilbert's theorem 90 to find all rational solutions $a, b$ of the
  equation $a^2 + b^2 d = 1$, where $d$ is a positive integer that is not
  a perfect square. (Hint: $a^2 + b^2 d$ equals
  $N_{\Q \subseteq \Q(\sqrt{-d})}(a + b\sqrt{-d})$; cf.\
  Exercise~1.12.)
\end{exer}
\begin{exer}{}{fields:exer:6.19}
  Let $k \subseteq F$ be a cyclic Galois extension of degree $d$, and let
  $\phi$ be a generator of $\mathrm{Aut}_k(F)$. Let $\alpha \in F$ be an
  element such that $\mathrm{tr}_{k \subseteq F}(\alpha) = 0$. Prove an
  `additive version' of Hilbert's theorem 90:
  \begin{enumerate}[label=(\roman*)]
    \item Prove that there exists a $\gamma \in F$ such that
      $\mathrm{tr}_{k \subseteq F}(\gamma) \neq 0$.
    \item Consider the expression
      \[
        \alpha\phi(\gamma) + (\alpha + \phi(\alpha))\phi^2(\gamma)
        + \dots
        + (\alpha + \phi(\alpha) + \dots + \phi^{d-2}(\alpha))
          \phi^{d-1}(\gamma).
      \]
    \item Prove that there exists a $\beta \in F$ such that
      $\alpha = \beta - \phi(\beta)$.
  \end{enumerate}
  Together with Exercise~4.20, this says that an element $\alpha$ of a
  cyclic Galois extension as above has trace $0$ if and only if there
  exists a $\beta$ such that $\alpha = \beta - \phi(\beta)$. For an
  application, see Exercise~7.15. [7.15]
\end{exer}

\section{Short march through applications of Galois theory}
\label{sec:fields:short-march-through-applications-of-galois-theory}

Galois theory is a pervasive tool, with wide-ranging applications. I have
neither the competence nor room in this book for an adequate survey of
applications of the theory, but even the few examples reviewed in this
section should suffice to convey its considerable power.

\subsection{Fundamental theorem of algebra}
\label{subsec:fields:fundamental-theorem-of-algebra}
Theorem~6.15 and a little group theory are essentially all that is
needed\footnote{I am cheating a little, actually, since the intermediate
value theorem will also be used behind the scenes. This is not too
surprising: the completeness of $\R$ must enter the proof somewhere.}
to give an algebraic proof of the fundamental theorem of algebra,
promised back in \S{}V.5.3.
\begin{thm}{}{fields:7.1}
  $\C$ is algebraically closed.
\end{thm}

\begin{proof}
  Let $f(x) \in \C[x]$ be a nonconstant polynomial; we have to prove
  that $f(x)$ has roots in $\C$. Note that if $f(x)$ has no roots in
  $\C$, then neither does $f(x)\overline{f(x)} \in \R[x]$; that is, we
  may assume that $f(x)$ has real coefficients. Let $F$ be a splitting
  field for $f(x)$ over $\R$; embed $F$ in an algebraic closure of $\R$
  (we `don't know yet' that $\C$ is algebraically closed!), and consider
  the extension $\R \subseteq F(i)$.

  This extension is Galois: it is the splitting field of the
  square-free part\footnote{That is, take each irreducible factor of the
  polynomial with a power of exactly $1$. Over a perfect field,
  square-free polynomials are clearly separable.} of $f(x)(x^2 + 1)$.
  Let $G = \mathrm{Aut}_\R(F(i))$.
  By Sylow I (Theorem~IV.2.5), $G$ has a $2$-Sylow subgroup $H$. The
  index $[G : H]$ is odd, so it corresponds to a finite, separable
  extension $\R \subseteq E$ with $[E : \R]$ odd. However, every
  polynomial of odd degree over $\R$ has a real root (Exercise~V.5.17,
  which only uses elementary calculus), and it follows that
  $\R \subseteq E$ is trivial (Exercise~5.23). Therefore $[G : H] = 1$,
  proving that $G$ is a $2$-group.

  Since $\C = \R(i) \subseteq F(i)$, the Galois group of the (Galois)
  extension $\C \subseteq F(i)$ is a subgroup of $G$; therefore it is a
  $2$-group: $|\mathrm{Aut}_\C(F(i))| = 2^n$ for some $n \geq 0$. Now
  recall (Proposition~IV.2.6) that every group of order $p^n$, with $p$
  prime, contains subgroups of order $p^m$ for all $0 \leq m \leq n$. In
  particular, if $n \geq 1$, then $\mathrm{Aut}_\C(F(i))$ contains a
  subgroup of order $2^{n-1}$, which corresponds to a quadratic extension
  of $\C$ via the Galois correspondence. However, there are no quadratic
  extensions of $\C$, because there are no irreducible quadratic
  polynomials in $\C[x]$ (Exercise~5.23 again).

  This contradiction shows that $n = 0$; that is,
  $\mathrm{Aut}_\C(F(i))$ is necessarily trivial. This proves that
  $F(i) = \C$ and in particular that $\C$ contains the roots of $f(x)$.
\end{proof}
The argument can be simplified further, bypassing the use of Sylow's
theorem at the price of a slight increase in length. But why should we
attempt to bypass Sylow's theorem, after the effort put into proving it
in Chapter~IV?

\subsection{Constructibility of regular n-gons}
\label{subsec:fields:constructibility-of-regular-n-gons}
We return one last time to the issue of constructing regular $n$-gons.
Simpler considerations have given us constraints on $n$ that must
necessarily be fulfilled for the regular $n$-gon to be constructible: if
$n = p$ is prime, we have found that $p - 1$ must be a power of $2$
(\S{}3.3); this was upgraded for any $n$ to the condition that $\varphi(n)$
is a power of $2$, in \S{}5.2 (and an even more explicit condition is
given in Exercise~5.19). But these are all `negative' results: we have
not as yet proved that if $\varphi(n)$ is a power of $2$, then the
regular $n$-gon is constructible. This is where Galois theory comes into
the picture.
\begin{prop}{}{fields:7.2}
  Let $k \subseteq F$ be a Galois extension, and assume $[F : k] = p^r$
  for some prime $p$ and $r \geq 0$. Then there exist intermediate fields
  \[
    k = E_0 \subseteq E_1 \subseteq E_2 \subseteq \dots \subseteq E_r = F
  \]
  such that $[E_i : E_{i-1}] = p$ for $i = 1, \dots, r$.
\end{prop}

\begin{proof}
  As the Galois correspondence is bijective for Galois extensions
  (Theorem~6.9, part~(5)), this statement follows immediately from the
  fact that a group of order $p^r$, with $p$ prime, has a complete series
  of $p$-subgroups; cf.\ for example the discussion following the
  statement of Theorem~IV.2.8.
\end{proof}
\begin{thm}{}{fields:7.3}
  The regular $n$-gon is constructible by straightedge and compass if and
  only if $\varphi(n)$ is a power of $2$.
\end{thm}

\begin{proof}
  As recalled above, we have already established the $(\Rightarrow)$
  direction. For the converse, assume $\varphi(n) = 2^r$ for some $r$.
  The extension $\Q \subseteq \Q(\zeta_n)$ is Galois (it is the splitting
  field of $\Phi_n(x)$), of order $[\Q(\zeta_n) : \Q] = \varphi(n) = 2^r$
  (Proposition~5.14). Proposition~7.2 shows that the condition given in
  Theorem~3.4 is satisfied and therefore that $\zeta_n$ is constructible,
  as needed.
\end{proof}
For example, $\varphi(17) = 16 = 2^4$; the Galois group of $\Q(\zeta_{17})$
over $\Q$ is isomorphic to $(\Z/17\Z)^\ast$ (by Proposition~5.16), and
therefore (Theorem~IV.6.10) it is cyclic, of order $16$. A generator of
$(\Z/17\Z)^\ast$ is $[6]_{17}$; it follows that $\mathrm{Aut}_\Q(\Q(\zeta_{17}))$
is generated by the automorphism $\sigma$ defined by
$\zeta_{17} \mapsto \zeta_{17}^6$. The sequence of subgroups
\[
  \mathrm{Aut}_\Q(\Q(\zeta_{17})) = \langle\sigma\rangle \supseteq
  \langle\sigma^2\rangle \supseteq \langle\sigma^4\rangle \supseteq
  \langle\sigma^8\rangle \supseteq \{e\}
\]
corresponds via the Galois correspondence to a sequence
\[
  \Q \subseteq \Q(\delta_1) \subseteq \Q(\delta_2) \subseteq
  \Q(\delta_3) \subseteq \Q(\zeta_{17}).
\]
It is instructive to apply what we now know to see how one may determine
a generator $\delta_1$ for the first extension, that is, the fixed field
of $\sigma^2$. Let $\zeta = \zeta_{17}$. As a $\Q$-vector space,
$\Q(\zeta)$ is generated by $\zeta, \zeta^2, \dots, \zeta^{16}$. To find
the fixed field of $\sigma^2$, write out a general element of $\Q(\zeta)$:
\begin{align*}
  &a_1\zeta + a_2\zeta^2 + a_3\zeta^3 + a_4\zeta^4 + a_5\zeta^5
  + a_6\zeta^6 + a_7\zeta^7 + a_8\zeta^8 \\
  &\quad + a_9\zeta^9 + a_{10}\zeta^{10} + a_{11}\zeta^{11}
  + a_{12}\zeta^{12} + a_{13}\zeta^{13} + a_{14}\zeta^{14}
  + a_{15}\zeta^{15} + a_{16}\zeta^{16}
\end{align*}
and apply $\sigma^2 : \zeta \mapsto (\zeta^6)^6 = \zeta^{36} = \zeta^2$:
\begin{align*}
  &a_1\zeta^2 + a_2\zeta^4 + a_3\zeta^6 + a_4\zeta^8 + a_5\zeta^{10}
  + a_6\zeta^{12} + a_7\zeta^{14} + a_8\zeta^{16} + a_9\zeta \\
  &\quad + a_{10}\zeta^3 + a_{11}\zeta^5 + a_{12}\zeta^7
  + a_{13}\zeta^9 + a_{14}\zeta^{11} + a_{15}\zeta^{13}
  + a_{16}\zeta^{15}.
\end{align*}
The element is fixed if and only if
\[
  a_1 = a_2 = a_4 = a_8 = a_{16} = a_{15} = a_{13} = a_9,\quad
  a_3 = a_6 = a_{12} = a_7 = a_{14} = a_{11} = a_5 = a_{10},
\]
and it follows that the fixed field of $\sigma^2$ is generated by
\[
  \delta_1 = \zeta + \zeta^2 + \zeta^4 + \zeta^8 + \zeta^{16}
  + \zeta^{15} + \zeta^{13} + \zeta^9
\]
and by $\zeta^3 + \zeta^6 + \zeta^{12} + \zeta^7 + \zeta^{14}
+ \zeta^{11} + \zeta^5 + \zeta^{10} = -1 - \delta_1$ (remember that
$\zeta$ is a root of $\Phi_{17}(x) = x^{16} + \dots + x + 1$).
The theory tells us that $\delta_1$ has degree $2$ over $\Q$, so
$\delta_1^2$ must be a rational combination of $1$ and $\delta_1$. Indeed,
pencil and paper and a bit of patience (and the relation $\zeta^{17} = 1$)
give
\begin{align*}
  \delta_1^2 &= (\zeta + \zeta^2 + \zeta^4 + \zeta^8 + \zeta^{16}
    + \zeta^{15} + \zeta^{13} + \zeta^9)^2 \\
  &= 3(\zeta + \zeta^2 + \zeta^4 + \zeta^8 + \zeta^{16} + \zeta^{15}
    + \zeta^{13} + \zeta^9) \\
  &\quad + 4(\zeta^3 + \zeta^6 + \zeta^{12} + \zeta^7 + \zeta^{14}
    + \zeta^{11} + \zeta^5 + \zeta^{10}) + 8 \\
  &= 3\delta_1 + 4(-1 - \delta_1) + 8 = -\delta_1 + 4.
\end{align*}
Thus, we find that $\delta_1^2 + \delta_1 - 4 = 0$. Solving for $\delta_1$
(note that $\delta_1 > 0$),
\[
  \delta_1 = \frac{-1 + \sqrt{17}}{2}.
\]
So we could easily construct $\delta_1$. The same procedure applied to
the other extensions may be used to produce $\delta_2$, $\delta_3$, and
finally $\zeta_{17}$, whose real part is the complicated expression given
at the end of \S{}3.3.

\subsection{Fundamental theorem on symmetric functions}
\label{subsec:fields:fundamental-theorem-on-symmetric-functions}
Let $t_1, \dots, t_n$ be indeterminates. The polynomial
\[
  P_n(x) := (x - t_1)\cdots(x - t_n) \in \Z[t_1, \dots, t_n][x]
\]
is universal in the sense that every polynomial of degree $n$ with
coefficients in (say) an integral domain may be obtained from $P_n(x)$
by suitably specifying $t_1, \dots, t_n$, possibly in a larger ring
(Exercise~7.2). The coefficients of the expansion
\[
  P_n(x) = x^n - s_1(t_1, \dots, t_n)x^{n-1} + \dots
  + (-1)^n s_n(t_1, \dots, t_n)
\]
are (up to sign) the \emph{elementary symmetric functions} of
$t_1, \dots, t_n$. For example, for $n = 3$,
\begin{align*}
  s_1(t_1,t_2,t_3) &= t_1 + t_2 + t_3, \\
  s_2(t_1,t_2,t_3) &= t_1 t_2 + t_1 t_3 + t_2 t_3, \\
  s_3(t_1,t_2,t_3) &= t_1 t_2 t_3.
\end{align*}
The functions $s_i(t_1, \dots, t_n)$ are \emph{symmetric} in the sense
that they are invariant under permutations of $t_1, \dots, t_n$ (because
$P_n(x)$ is invariant); they are elementary by virtue of the following
important result, which has a particularly simple proof thanks to Galois
theory.
\begin{thm}{Fundamental theorem on symmetric functions}{fields:7.4}
  Let $K$ be a field, and let $\phi \in K(t_1, \dots, t_n)$. Then $\phi$
  is symmetric if and only if it is a rational function (with coefficients
  in $K$) of the elementary symmetric functions $s_1, \dots, s_n$.
\end{thm}

Here we are viewing the $s_i$'s as elements of $K[t_1, \dots, t_n]$,
which we can do unambiguously since $\Z$ is initial in $\mathsf{Ring}$.

\begin{proof}
  Let $F = K(t_1, \dots, t_n)$, and let $k = K(s_1, \dots, s_n)$ be the
  subfield generated by the elementary symmetric functions over $K$. Then
  $F$ is a splitting field of the separable polynomial $P_n(x)$ over $k$.
  In particular $k \subseteq F$ is a Galois extension, and
  $|\mathrm{Aut}_k(F)| = [F : k] \leq n!$ by Lemma~4.2. The symmetric
  group $S_n$ acts (faithfully) on $F$ by permuting $t_1, \dots, t_n$,
  and this action is the identity on $k = K(s_1, \dots, s_n)$ since each
  $s_i$ is symmetric. Thus $S_n$ may be identified with a subgroup of
  $\mathrm{Aut}_k(F)$; but $|S_n| = n!$, so it follows that
  $\mathrm{Aut}_k(F) = S_n$.

  Since $k \subseteq F$ is Galois, we obtain $k = F^{S_n}$. This says
  precisely that if $\phi \in K(t_1, \dots, t_n)$, then $\phi$ is
  invariant under the $S_n$ action on $t_1, \dots, t_n$ if and only if
  $\phi \in k = K(s_1, \dots, s_n)$, which is the statement.
\end{proof}
It is not difficult to strengthen the result of Theorem~7.4 and prove
that every symmetric polynomial is in fact a polynomial in the elementary
symmetric functions. Note that the argument given in the proof amounts to
the following result, which should be recorded for individual attention:
\begin{lem}{}{fields:7.5}
  Let $K$ be a field, $t_1, \dots, t_n$ indeterminates, and let
  $s_1, \dots, s_n$ be the elementary symmetric functions on
  $t_1, \dots, t_n$. Then the extension
  $K(s_1, \dots, s_n) \subseteq K(t_1, \dots, t_n)$ is Galois, with
  Galois group $S_n$.
\end{lem}

This result and Cayley's theorem (Theorem~II.9.5) immediately yield the
following:

\begin{coro}{}{fields:7.6}
  Let $G$ be a finite group. Then there exists a Galois extension
  $k \subseteq F$ such that $\mathrm{Aut}_k(F) \cong G$.
\end{coro}

\begin{proof}
  Exercise~7.4.
\end{proof}
It is not known whether $k$ may be chosen to be $\Q$ in
Corollary~7.6. The proof hinted at above realizes $k$ as a rather large
field, and it is not at all clear how one may `descend' from this field
to $\Q$. This is the \emph{inverse Galois problem}, a subject of current
research. The reader can appreciate the difficulty of this problem by
noting that even realizing $S_n$ as a Galois group over $\Q$ (for all
$n$) is not so easy, although it is true that for every $n$ there are
infinitely many polynomials of degree $n$ in $\Z[x]$ whose splitting
fields have Galois group $S_n$ over $\Q$ (see Example~7.20 and
Exercise~7.11 for the case $n = p$ prime). If you want a memorable
example to carry with you, \name{Schur} showed (1931) that the splitting
field of the `truncated exponential'
\[
  1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots + \frac{x^n}{n!}
\]
has Galois group $S_n$ over $\Q$ for $n \neq 4$ (and $A_n$ for $n = 4$).
It is known (\name{Shafarevich}, 1954) that every finite solvable group
is the Galois group of a Galois extension over $\Q$. The inverse Galois
problem may be rephrased as asking whether every finite group is a
quotient of the Galois group of $\Q$ over $\Q$. I have already mentioned
that this object (which may be interpreted as the `fundamental group of
$\Q$'; cf.\ Remark~6.16) is mysterious and extremely
interesting.\footnote{According to J.~S.\ \name{Milne}, the `most
interesting object in mathematics' is the \'etale fundamental group of
the projective line over $\Q$ with the three points $0, 1, \infty$
removed. This is an extension of the Galois group of $\Q$ over $\Q$.}
Going back to Lemma~7.5, the alternating group $A_n$ has index $2$ in
$S_n$, so it corresponds to a quadratic extension of $K(s_1, \dots, s_n)$
via the Galois correspondence. It is easy to determine a generator of
this extension: it must be a function of $t_1, \dots, t_n$ which is
invariant under the action of all even permutations and not invariant
under the action of odd permutations. We have used such a function
precisely to define even/odd permutations, in \S{}IV.4.3:
\[
  \Delta = \prod_{1 \leq i < j \leq n} (t_i - t_j).
\]
To see that $\Delta$ has degree $2$ over $K(s_1, \dots, s_n)$, simply
note that the discriminant
\[
  D = \Delta^2 = \prod_{1 \leq i < j \leq n} (t_i - t_j)^2
\]
is invariant under the action of all permutations, and therefore it
belongs to the field $K(s_1, \dots, s_n)$ by Theorem~7.4. This proves

\begin{coro}{}{fields:7.7}
  With notation as above, the extension
  $K(s_1, \dots, s_n)(\sqrt{D}) \subseteq K(t_1, \dots, t_n)$ is Galois,
  with Galois group $A_n$.
\end{coro}

\subsection{Solvability of polynomial equations by radicals}
\label{subsec:fields:solvability-of-polynomial-equations-by-radicals}
This is possibly the most famous elementary application of Galois theory
and one that is very close to \name{Galois}' own original motivation.
Quadratic polynomial equations
\[
  x^2 + bx + c = 0
\]
are famously solved (in\footnote{The formula requires extracting a square
root; in characteristic $2$ this may not be possible.} characteristic
$\neq 2$) by the quadratic formula:
\[
  x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2}.
\]
Analogous, but more complicated, formulas exist for equations of degree
$3$ and $4$ (\name{Tartaglia}/\name{Cardano}/\name{Ferrari}, around 1540;
this generated a priority dispute which degenerated into base personal
attacks). Equations of degree $5$ and higher resisted the best efforts of
mathematicians for centuries. The most ambitious goal, in line with what
occurs for degrees $2$, $3$, and $4$, would be to produce a formula for
the solutions to the `general' polynomial equation
\[
  x^n + a_{n-1}x^{n-1} + \dots + a_0 = 0
\]
in terms of the coefficients $a_i$ and basic operations such as taking
roots. This would be a solution `by radicals' of the equation. Well, such
a formula simply does not exist:
\begin{thm}{}{fields:7.8}
  The general polynomial equation of degree $5$ or higher admits no
  solution by radicals.
\end{thm}

I am being intentionally vague about the ground field in which the
problem is posed; a `formula' such as the quadratic formula ought to hold
over any field in which it makes sense; for example, the field should not
have characteristic $2$ in the case of the quadratic formula. An easy way
to avoid any such snag is to work in characteristic zero: this has the
main technical advantage of ensuring automatic separability of polynomials
$x^m - 1$, whose roots define the `radicals'; cf.\ Proposition~6.19.
Therefore, we will assume in what follows that the ground field has
characteristic $0$.
The `general polynomial' is the `universal' polynomial $P_n(x)$
introduced in \S{}7.3, taken over a chosen ground field $K$: the general
polynomial
\[
  x^n - s_1 x^{n-1} + \dots + (-1)^n s_n \in k[x],
\]
with $k = K(s_1, \dots, s_n)$ and $s_i$ elementary symmetric functions
in (`indeterminate') roots $t_1, \dots, t_n$. It is in fact no harder to
deal with the more general case of arbitrary irreducible polynomials
$f(x) \in k[x]$, over any field $k$ (of characteristic zero); this is
what we will do.

A formula by radicals expresses a root $t_i$ of $f(x)$ as an element of
an extension obtained from $k$ in the following fashion:
\[
  k \subseteq k(\delta_1) \subseteq k(\delta_1, \delta_2) \subseteq
  \dots \subseteq k(\delta_1, \dots, \delta_r)
\]
where for all $i = 1, \dots, r$ we have $\delta_i^{m_i}
\in k(\delta_1, \dots, \delta_{i-1})$ for some exponent $m_i$.

\begin{defn}{}{fields:7.9}
  Extensions of this type are called \emph{radical extensions}.
\end{defn}
Some facts about radical extensions are immediate from the definition:
for example, if $k \subseteq F$ and $k \subseteq K$ are two radical
extensions and $F$, $K$ are subfields of a larger field, then
$k \subseteq FK$ is radical (Exercise~7.5). Further, radical extensions
are composites of cyclic ones provided the appropriate roots of $1$ are
in $k$, by Proposition~6.19. Also,

\begin{lem}{}{fields:7.10}
  Every separable radical extension is contained in a Galois radical
  extension.
\end{lem}

\begin{proof}
  Let $k \subseteq F$ be a separable radical extension. In particular
  $k \subseteq F$ is finite and separable, so $F = k(\alpha)$ for some
  $\alpha \in F$ (Proposition~5.19). Let $p(x)$ be the minimal polynomial
  of $\alpha$ over $k$. The splitting field $L$ of $p(x)$ over $k$ may be
  obtained as the composite of $\sigma(F)$ as $\sigma$ ranges over the
  embeddings of $F$ in $\bar{k}$, extending $\mathrm{id}_k$: indeed, this
  composite is generated over $k$ by all $\sigma(\alpha)$, which range
  over all roots of $p(x)$. As a composite of radical extensions, $L$ is
  radical over $k$. As a splitting field of a separable polynomial, $L$
  is Galois over $k$ (Theorem~6.9, part~(1)), and we are done.
\end{proof}
Galois theory will allow us to relate solvability by radicals of a
polynomial $f(x)$ with a precise statement on the splitting field of
$f(x)$. Here is a first formalization of this connection:

\begin{lem}{}{fields:7.11}
  Let $k$ be a field of characteristic $0$, and let $f(x) \in k[x]$ be
  an irreducible polynomial. Then there exists a formula solving $f(x)$
  by radicals if and only if the splitting field of $f(x)$ is contained
  in a Galois radical extension.
\end{lem}

\begin{proof}
  If the splitting field of $f(x)$ is contained in a radical extension
  (Galois or not), then the roots may be written as combinations of field
  operations and radicals, as needed.

  For the converse, assume $f(x)$ is solvable by radicals. A formula for
  a root $t_1$ can be turned into a formula for any other root $t_i$ by
  applying an element $\sigma_i$ of (for example) $\mathrm{Aut}_k(\bar{k})$
  sending $t_1$ to $t_i$ (such an automorphism exists since $f(x)$ is
  irreducible). Thus, if a formula exists for one root, then a formula
  exists for all roots. The composite of all the corresponding radical
  extensions gives one, possibly larger, radical extension of $k$
  containing all roots of $f(x)$. Therefore, the splitting field of
  $f(x)$ is contained in a radical extension of $k$. By Lemma~7.10, this
  extension may be assumed to be Galois, as needed.
\end{proof}
In a nutshell, we will understand solvability of polynomial equations by
radicals if we understand radical Galois extensions. As demanded by the
general philosophy underlying Galois theory, we should aim to
characterize these extensions in terms of their Galois groups. It turns
out that the right condition (modulo technical considerations) is that the
Galois group should be solvable.

\begin{defn}{}{fields:7.12}
  A Galois extension $k \subseteq F$ is \emph{solvable} if
  $\mathrm{Aut}_k(F)$ is a solvable group.
\end{defn}

The reader may want to revisit \S{}IV.3.3 now, for a reminder on
solvability in group theory. (We are approaching an explanation for the
`solvable' terminology!) Just as radical extensions decompose as a
sequence of cyclic extensions, solvable groups have cyclic composition
factors (Proposition~IV.3.11). I will say\footnote{This is a harmless
simplification; the precise condition is a little more careful about
which roots of unity are needed.} that a radical, resp., solvable,
extension $k \subseteq F$ `has enough roots of $1$' if $k$ contains a
primitive $M$-th root of $1$ for a common multiple $M$ of the order of
the corresponding cyclic factors.
\begin{lem}{}{fields:7.13}
  Let $k \subseteq F$ be a Galois extension, with $\mathrm{char}\,k = 0$.
  Provided it has enough roots of $1$, $k \subseteq F$ is radical if and
  only if it is solvable.
\end{lem}
\begin{proof}
  Modulo the fundamental theorem of Galois theory, this is a
  straightforward generalization of Proposition~6.19.

  Indeed, assume that $k \subseteq F$ is radical:
  \[
    k \subseteq k(\delta_1) \subseteq \dots \subseteq
    k(\delta_1, \dots, \delta_r) = F
  \]
  with $\delta_i^{m_i} \in k(\delta_1, \dots, \delta_{i-1})$. By
  hypothesis, $k$ contains primitive $m_i$-th roots of $1$ for all $i$.
  Consider the corresponding series of subgroups of
  $G = \mathrm{Aut}_k(F)$:
  \[
    G = G_0 \supseteq G_1 \supseteq \dots \supseteq G_r = \{e\},
  \]
  with $G_i = \mathrm{Aut}_{k(\delta_1,\dots,\delta_i)}(F)$. Note that
  $F$ is Galois over each intermediate field. Each extension
  $k(\delta_1, \dots, \delta_{i-1}) \subseteq k(\delta_1, \dots,
  \delta_i)$ satisfies the hypothesis of Proposition~6.19, and it follows
  that it is Galois and cyclic. By the fundamental theorem
  (Theorem~6.15) each $G_i$ is normal in the previous $G_{i-1}$, with
  cyclic quotient, and it follows that $G$ is solvable
  (Proposition~IV.3.11~(ii)). The same argument in reverse, using the
  converse implication in Proposition~6.19, proves that solvable
  extensions with enough roots of $1$ are radical.
\end{proof}
How do we account for the possible absence of the needed roots of $1$?
Surprisingly, this affects the statement in a rather minor way.
\begin{prop}{}{fields:7.14}
  Let $k$ be a field of characteristic $0$, and let $k \subseteq F$ be a
  Galois extension. Then $k \subseteq F$ is solvable if and only if it is
  contained in a Galois radical extension.
\end{prop}

The characteristic $0$ in Proposition~7.14 is an overshoot; as in
Proposition~6.19, it would suffice to require that $\mathrm{char}\,k$
does not divide the relevant degrees.
\begin{proof}
  Assume $\mathrm{Aut}_k(F)$ is solvable. Let $M$ be a common multiple
  of the orders of the cyclic quotients, and let $\zeta$ be a primitive
  $M$-th root of $1$ in an algebraic closure $\bar{k}$ of $k$. The
  splitting field $k(\zeta)$ of $x^M - 1$ over $k$ is Galois over $k$
  ($x^M - 1$ is separable since $k$ has characteristic $0$). By
  Proposition~6.17, the extension $k(\zeta) \subseteq F(\zeta)$ is also
  Galois, with Galois group isomorphic to $\mathrm{Aut}_{k(\zeta) \cap F}(F)$.
  It follows that $\mathrm{Aut}_{k(\zeta)}(F(\zeta))$ is solvable, since
  $\mathrm{Aut}_{k(\zeta) \cap F}(F)$ is a subgroup of $\mathrm{Aut}_k(F)$
  and the latter is solvable by assumption. (See the comment following
  Corollary~IV.3.13.) Since $k(\zeta) \subseteq F(\zeta)$ has enough
  roots of $1$, Lemma~7.13 implies that it is radical. So is
  $k \subseteq F(\zeta)$, since $k \subseteq k(\zeta)$ is itself
  trivially radical. This proves that $k \subseteq F$ is contained in a
  radical extension, hence in a Galois radical extension by Lemma~7.10.

  Conversely, assume that $k \subseteq F$ is Galois and contained in a
  radical Galois extension $k \subseteq L$. Let $M$ be a common multiple
  of the orders of cyclic factors in this extension, and let $\zeta$ be a
  primitive $M$-th root of $1$. The composite $L(\zeta)$ is Galois and
  radical over $k(\zeta)$ (Proposition~6.17 and Exercise~7.5); further,
  it has enough roots of $1$. By Lemma~7.13, $k(\zeta) \subseteq L(\zeta)$
  is solvable. It follows that $k \subseteq L(\zeta)$ is solvable. Indeed,
  this extension is Galois (Exercise~6.13); apply the fundamental theorem
  to $k \subseteq k(\zeta) \subseteq L(\zeta)$ to establish that
  $\mathrm{Aut}_k(k(\zeta))$ is isomorphic to the quotient of
  $\mathrm{Aut}_k(L(\zeta))$ by its normal subgroup
  $\mathrm{Aut}_{k(\zeta)}(L(\zeta))$. Both $\mathrm{Aut}_{k(\zeta)}(L(\zeta))$
  (as shown above) and $\mathrm{Aut}_k(k(\zeta))$ (an abelian group;
  cf.\ Example~6.18) are solvable, and it follows that
  $\mathrm{Aut}_k(L(\zeta))$ is solvable as promised, by
  Corollary~IV.3.13. Since $F$ is an intermediate field and Galois over
  $k$, by the fundamental theorem $\mathrm{Aut}_k(F)$ is a homomorphic
  image of $\mathrm{Aut}_k(L(\zeta))$. This shows that $\mathrm{Aut}_k(F)$
  is a homomorphic image of a solvable group; hence it is itself solvable
  (Corollary~IV.3.13 again), and we are done.
\end{proof}
We may have lost sight of where we were heading; but we are ready to
reconnect with the study of solutions to polynomial equations.
\begin{defn}{}{fields:7.15}
  The \emph{Galois group} $\mathrm{Gal}_k(f(x))$ of a separable
  polynomial $f(x) \in k[x]$ is the Galois group of the splitting field
  of $f(x)$ over $k$.
\end{defn}
\begin{coro}{}{fields:7.16}
  Let $k$ be a field of characteristic $0$, and let $f(x) \in k[x]$ be
  an irreducible polynomial. Then $f(x)$ is solvable by radicals if and
  only if its Galois group is solvable.
\end{coro}

\begin{proof}
  This is an immediate consequence of Lemma~7.11 and
  Proposition~7.14.
\end{proof}
Corollary~7.16 is called \emph{\name{Galois}' criterion}.
\name{Ruffini} (1799) and \name{Abel} (1824) had previously established
that general formulas in radicals for the solutions of equations of
degree $\geq 5$ do not exist (that is, Theorem~7.8); but it was
\name{Galois} who identified the precise condition given in
Corollary~7.16. Of course, Theorem~7.8 follows immediately from
\name{Galois}' criterion:

\begin{proof}[Proof of \cref{thm:fields:7.8}]
  By Lemma~7.5, the Galois group of the general polynomial of degree $n$
  is $S_n$. The group $S_n$ is not solvable for $n \geq 5$ (Corollary
  IV.4.21); hence the statement follows from Corollary~7.16.
\end{proof}
In fact, we could now do more: we know that $S_3$ and $S_4$ are solvable
(cf.\ Exercise~IV.3.16); from a composition series with cyclic quotients
we could in principle decompose explicitly the splitting field of general
polynomials of degree $3$ and $4$ as radical extensions and as a
consequence recover the \name{Tartaglia}/\name{Cardano}/\name{Ferrari}
formulas for their solutions.

\subsection{Galois groups of polynomials}
\label{subsec:fields:galois-groups-of-polynomials}
Impressive as \name{Galois}' criterion is, it does not in itself produce
a single polynomial of degree (say) $5$ over $\Q$ that is not solvable
by radicals. Finding such polynomials requires a certain amount of extra
work; we will be able to do this by the end of this subsection.
Computing Galois groups of polynomials is in fact a popular sport among
algebraists, excelling at which would require much more information than
the reader will glean here. I will just list a few straightforward
observations.
To begin with, recall that an element of $\mathrm{Aut}_k(F)$ must send
roots of a polynomial $f(x) \in k[x]$ to roots of the same polynomial,
and if $f(x)$ is irreducible and $F$ is its splitting field, then there
are automorphisms of $F$ sending any root of $f(x)$ to any other root
(Proposition~1.5, Lemma~4.2). This observation may be rephrased as
follows:
\begin{lem}{}{fields:7.17}
  Let $f(x) \in k[x]$ be a separable irreducible polynomial of degree
  $n$. Then $\mathrm{Gal}_k(f(x))$ acts transitively on the set of roots
  of $f(x)$ in $\bar{k}$. In particular, $\mathrm{Gal}_k(f(x))$ may be
  identified with a transitive subgroup of the symmetric group $S_n$.
\end{lem}
Of course a subgroup of $S_n$ is transitive if the corresponding action
on $\{1, \dots, n\}$ is transitive; the diligent reader has run across
this terminology in Exercise~IV.4.12.
The reader will easily produce a statement analogous to Lemma~7.17 in
case $f(x)$ is reducible. Of course the action is not transitive in this
case, and if the factors of $f(x)$ have degrees $n_1, \dots, n_r$, then
$\mathrm{Gal}_k(f(x))$ is contained in a subgroup of $S_n$ isomorphic
to $S_{n_1} \times \dots \times S_{n_r}$.
Lemma~7.17 (or its `reducible' variations) already give some
information. For example, we see that the Galois group of a separable
irreducible cubic can only be $A_3$ or $S_3$, since these are the only
transitive subgroups of $S_3$. Similarly, the range of possibilities for
irreducible polynomials of degree $4$ is rather restricted: $S_4$, $A_4$,
and isomorphic copies of the dihedral group $D_8$, of
$\Z/2\Z \times \Z/2\Z$, and of $\Z/4\Z$ (Exercise~7.8). One approach to
the computation of Galois groups of polynomials amounts to defining
invariants imposing further restrictions, leading to algorithms deciding
which of such possibilities occurs for a given polynomial.
We have already encountered the most important of these invariants: if
the roots of a separable polynomial $f(x) \in k[x]$ in its splitting
field are $\alpha_1, \dots, \alpha_n$, the \emph{discriminant} of $f(x)$
is the element $D = \Delta^2$, where
\[
  \Delta = \prod_{1 \leq i < j \leq n} (\alpha_i - \alpha_j).
\]
Every permutation of the roots fixes $D$, so $D$ must be fixed by the
whole Galois group; therefore, $D \in k$. Odd permutations move $\Delta$
(if $\mathrm{char}\,k \neq 2$), and even permutations fix it; therefore
$\Delta$ is fixed by the Galois group $G$ (in other words, $\Delta \in k$)
if and only if $G \subseteq A_n$. This proves
\begin{lem}{}{fields:7.18}
  Let $k$ be a field of characteristic $\neq 2$, and let $f(x) \in k[x]$
  be a separable polynomial with discriminant $D$. Then the Galois group
  of $f(x)$ is contained in the alternating group $A_n$ if and only if
  $D$ is a square in $k$.
\end{lem}
\begin{exam}{}{fields:exam:7.19}
  Lemma~7.18 and a discriminant computation are all that is needed to
  compute the Galois group of an irreducible cubic polynomial
  $f(x) = x^3 + ax^2 + bx + c$. It would be futile to try to remember
  the discriminant
  \[
    D = a^2 b^2 - 4a^3 c - 4b^3 + 18abc - 27c^2;
  \]
  but one may remember the trick of shifting $x$ by $a/3$ (in
  characteristic $\neq 3$), with the effect of killing the coefficient of
  $x^2$:
  \[
    f\!\left(x - \tfrac{a}{3}\right) = x^3 + px + q
  \]
  for suitable $p$ and $q$. This does not change $D$ (shifting all roots
  $\alpha_i$ by the same amount has no effect on the differences
  $\alpha_i - \alpha_j$), yet $D = -4p^3 - 27q^2$ is a little more
  memorable. By Lemma~7.18 and the preceding considerations, the Galois
  group is $A_3$ if $D$ is a square and $S_3$ otherwise. For example,
  $x^3 - 2$ has Galois group $S_3$ over $\Q$, since $D = -108$ is not a
  square in $\Q$; $x^3 - 3x + 1$ has discriminant $81 = 9^2$, and
  therefore it has Galois group $A_3 \cong \Z/3\Z$.
\end{exam}
The reader will have no difficulty locating a discussion of the different
possibilities for polynomials of degree $4$ and detailed information for
higher degree polynomials. I will just highlight the following simple
observation.
\begin{exam}{}{fields:exam:7.20}
  Let $f(x) \in \Q[x]$ be an irreducible polynomial of degree $p$, where
  $p$ is prime. Assume that $f(x)$ has $p - 2$ real roots and $2$
  nonreal, complex roots. Then the Galois group of $f(x)$ is $S_p$.

  Indeed, complex conjugation induces an automorphism of the splitting
  field and acts by interchanging the two nonreal roots, so the Galois
  group $G$, as a subgroup of $S_p$, contains a transposition. On the
  other hand, the degree of the splitting field (and hence $|G|$) is
  divisible by $p$, because it contains a simple extension of order $p$,
  obtained by adjoining any one root to $\Q$. Since $p$ is prime, $G$
  contains an element of order $p$ by \name{Cauchy}'s theorem
  (Theorem~IV.2.1); the only elements of order $p$ in $S_p$ are
  $p$-cycles, so $G$ contains a $p$-cycle. It follows that $G = S_p$,
  by (a simple variation of) Exercise~IV.4.7. For example, the Galois
  group of $f(x) = x^5 - 5x - 1$ over $\Q$ is $S_5$, giving a concrete
  example of a quintic that cannot be solved by radicals
  (Exercise~7.10). The reader will use this technique to produce
  polynomials of every prime degree $p$ in $\Z[x]$ whose Galois group is
  $S_p$ (Exercise~7.11).
\end{exam}

\subsection{Abelian groups as Galois groups over Q}
\label{subsec:fields:abelian-groups-as-galois-groups-over-q}
Having mentioned the inverse Galois problem in \S{}7.3, I should point
out that the reader is now in the position of proving that every finite
abelian group may be realized as the Galois group of an extension over
$\Q$.
In fact, the reader can prove a much more precise result: every finite
abelian group may be realized as the Galois group of some intermediate
field of the extension $\Q \subseteq \Q(\zeta_n)$ of $\Q$ in a
cyclotomic field. This uses a number-theoretic fact: for every integer
$N$, there are infinitely many primes $p$ such that $p \equiv 1 \pmod{N}$.
This is a particular case of \name{Dirichlet}'s theorem (1837), which
states that if $a$, $b$ are positive integers and $\gcd(a, b) = 1$, then
there are infinitely many primes of
the form $a + nb$ with $n > 0$. The particular case $a = 1$, $b = N$
needed here was apparently already known to \name{Euler} and can be
proven by elementary means (in fact, there is a proof using cyclotomic
polynomials: see Exercise~5.18). Assuming this fact, argue as follows:
\begin{itemize}
  \item By the classification theorem (Theorem~IV.6.6), every finite
    abelian group $G$ is isomorphic to a product of cyclic groups
    $(\Z/n_1\Z) \times \dots \times (\Z/n_r\Z)$.
  \item By \name{Dirichlet}'s theorem, we can then choose distinct primes
    $p_i$ such that $p_i \equiv 1 \pmod{n_i}$.
  \item Let $n = p_1 \cdots p_r$; by the Chinese Remainder Theorem
    (Theorem~V.6.1),
    $(\Z/n\Z)^\ast \cong (\Z/p_1\Z)^\ast \times \dots \times
    (\Z/p_r\Z)^\ast$, and the $i$-th factor on the right-hand side is
    cyclic of order $p_i - 1$, a multiple of $n_i$.
  \item It follows that $(\Z/n\Z)^\ast$ has a subgroup $H$ such that
    $G \cong (\Z/n\Z)^\ast / H$.
  \item Since $(\Z/n\Z)^\ast \cong \mathrm{Aut}_\Q(\Q(\zeta_n))$
    (Proposition~5.16) and cyclotomic fields are Galois over $\Q$, $H$
    corresponds to an intermediate field $F$,
    $\Q \subseteq F \subseteq \Q(\zeta_n)$.
  \item Since $\mathrm{Aut}_\Q(\Q(\zeta_n))$ is abelian, $H$ is
    automatically normal; hence $\Q \subseteq F$ is Galois, and
    $\mathrm{Aut}_\Q(F) \cong G$, as needed.
\end{itemize}
Thus, the `inverse Galois problem' has an easy solution for abelian
groups. A much stronger result is true: it can be proven that every
finite abelian extension of $\Q$ is contained in some cyclotomic field.
This is the \emph{\name{Kronecker}--\name{Weber} theorem}, dating back
to the second half of the nineteenth century. The natural context for
this result is class field theory and is well beyond the scope of this
book. The reader will be able to verify (Exercise~7.19) that every
quadratic extension of $\Q$ is contained in some cyclotomic field.

\subsection*{Exercises}

\begin{exer}{}{fields:exer:7.1}
  Find explicitly a generator of a quadratic intermediate field of the
  extension $\Q \subseteq \Q(\zeta_{10})$.
\end{exer}
\begin{exer}{}{fields:exer:7.2}
  Let $R$ be an integral domain, and let $f(x) \in R[x]$ be a polynomial
  of degree $n$. Show how to obtain $f(x)$ by specializing the
  `universal' polynomial $P_n(x)$ defined in \S{}7.3. [\S{}7.3]
\end{exer}
\begin{exer}{}{fields:exer:7.3}
  Prove that the elementary symmetric functions $s_1, \dots, s_n$ (see
  \S{}7.3) are algebraically independent. (Hint: Use Exercise~1.28.)
\end{exer}
\begin{exer}{}{fields:exer:7.4}
  Prove that every finite group is isomorphic to the group of
  automorphisms of some Galois extension. [\S{}7.3]
\end{exer}
\begin{exer}{}{fields:exer:7.5}
  Let $k \subseteq F$ be a radical extension, and let $k \subseteq K$ be
  any extension; assume $F$ and $K$ are contained in a larger field.
  Prove that $K \subseteq FK$ is radical. [\S{}7.4]
\end{exer}
\begin{exer}{}{fields:exer:7.6}
  Let $f(x) \in \Q[x]$ be an irreducible cubic, and let $\rho$ be a real
  root of $f(x)$. Prove that the splitting field of $f(x)$ over $\Q$ is
  $\Q(\rho, \sqrt{D})$, where $D$ is the discriminant of $f(x)$.
\end{exer}
\begin{exer}{}{fields:exer:7.7}
  Let $f(x) \in \Q[x]$ be an irreducible cubic with exactly one real
  root. Prove that the discriminant of $f(x)$ is not a square in $\Q$.
\end{exer}
\begin{exer}{}{fields:exer:7.8}
  (Cf.\ Exercise~IV.4.12.) Find (mathematically or by library search)
  the lattice of subgroups of $S_4$, and verify that the transitive
  subgroups of $S_4$ are (isomorphic to) $S_4$, $A_4$, $D_8$,
  $\Z/2\Z \times \Z/2\Z$, and $\Z/4\Z$. [\S{}7.5]
\end{exer}
\begin{exer}{}{fields:exer:7.9}
  Compute the Galois group of the polynomial $x^4 - 2$.
\end{exer}
\begin{exer}{}{fields:exer:7.10}
  Prove that the polynomial $x^5 - 5x - 1$ has exactly $3$ real roots
  (this is a calculus exercise!) and is irreducible over $\Q$. Prove
  that its Galois group is $S_5$. [\S{}7.5]
\end{exer}
\begin{exer}{}{fields:exer:7.11}
  Let $n > 0$ be an integer. Note that
  \[
    f_n(x) := (x^2 + 2) \cdot x \cdot (x-2) \cdots (x - 2(n-4))
    \cdot (x - 2(n-3))
  \]
  has $n - 2$ rational roots and $2$ nonreal, complex roots. Prove that
  for infinitely many integers $q$, the polynomial $qf_n(x) + 2 \in \Z[x]$
  has $(n-2)$ real roots, $2$ nonreal complex roots, and is irreducible
  over $\Q$. Conclude that for each prime $p$ there are infinitely many
  polynomials of degree $p$ in $\Z[x]$ whose Galois group is $S_p$.
  [\S{}7.3, \S{}7.5]
\end{exer}
\begin{exer}{}{fields:exer:7.12}
  Let $f(x) \in k[x]$ be a separable irreducible polynomial of prime
  degree $p$ over a field $k$, and let $\alpha_1, \dots, \alpha_p$ be
  the roots of $f(x)$ in its splitting field $F$. Prove that the Galois
  group of $f(x)$ contains an element $\sigma$ of order $p$, `cycling'
  through the roots. [7.13]
\end{exer}
\begin{exer}{}{fields:exer:7.13}
  Let $f(x) \in k[x]$ be a separable irreducible polynomial of prime
  degree $p$ over a field $k$. Let $\alpha$ be a root of $f(x)$ in
  $\bar{k}$, and suppose you can express another root of $f(x)$ as a
  polynomial in $\alpha$, with coefficients in $k$. Prove that you can
  express all roots as polynomials in $\alpha$ and that the Galois group
  of $f(x)$ is $\Z/p\Z$. (Use Exercise~7.12.) [7.14]
\end{exer}
\begin{exer}{}{fields:exer:7.14}
  Let $k$ be a field of characteristic $p > 0$, and let
  $f(x) = x^p - x - a \in k[x]$. Assume that $f(x)$ has no roots in
  $k$. Prove that $f(x)$ is irreducible and that its Galois group is
  $\Z/p\Z$. (Note that if $\alpha$ is a root of $f(x)$, so is
  $\alpha + 1$; use Exercises~1.8 and~7.13.) The splitting field of
  $f(x)$ is an \emph{\name{Artin}--\name{Schreier} extension} of
  $\F_p$. [7.16]
\end{exer}
\begin{exer}{}{fields:exer:7.15}
  Let $k \subseteq F$ be a cyclic Galois extension of degree $p$, where
  $\mathrm{char}\,k = p > 0$. Prove that it is an
  \name{Artin}--\name{Schreier} extension. (Hint: What is
  $\mathrm{tr}_{k \subseteq F}(1)$? Use the additive version of
  Hilbert's theorem~90, Exercise~6.19.) [6.19]
\end{exer}
\begin{exer}{}{fields:exer:7.16}
  The \emph{\name{Artin}--\name{Schreier} map} on a field $k$ of
  positive characteristic $p$ is the function $x \mapsto x^p - x$.
  Denote this function by $\mathrm{AS}$, and let $\sqrt[\mathrm{AS}]{\cdot}$
  denote its inverse (which is only defined up to the addition of an
  element of $\F_p$: see Exercise~7.14). Express the solutions of a
  quadratic equation $x^2 + bx + c = 0$ in characteristic $2$ in terms
  of $\sqrt[\mathrm{AS}]{\cdot}$, for $b \neq 0$. (For $b \neq 0$, the
  roots must live in an \name{Artin}--\name{Schreier} extension of $k$,
  since the polynomial is then separable. It is inseparable for $b = 0$;
  solving the equation in this case amounts to extending $k$ by the
  unique square root of $c$.) [\S{}7.4]
\end{exer}
\begin{exer}{}{fields:exer:7.17}
  Let $f(x) \in k[x]$ be a separable irreducible polynomial of degree
  $n$ over a field $k$, and let $F$ be its splitting field. Assume
  $\mathrm{Aut}_k(F) \cong S_n$, and let $\alpha$ be a root of $f(x)$
  in $F$.
  \begin{enumerate}[label=(\roman*)]
    \item Prove that $\mathrm{Aut}_{k(\alpha)}(F) \cong S_{n-1}$.
    \item Prove that there are no proper subfields of $k(\alpha)$
      properly containing $k$. (Hint: Exercise~IV.4.8.)
  \end{enumerate}
  [7.18]
\end{exer}
\begin{exer}{}{fields:exer:7.18}
  Let $f(x) \in k[x]$ be a separable irreducible polynomial of degree
  $n$ over a field $k$, with Galois group $S_n$. Let $\alpha$ be a root
  of $f(x)$ in $\bar{k}$, and let $g(x) \in k[x]$ be any nonconstant
  polynomial of degree $< n$. Prove that $\alpha$ may be expressed as a
  polynomial in $g(\alpha)$, with coefficients in $k$. (Use
  Exercise~7.17.)
\end{exer}
\begin{exer}{}{fields:exer:7.19}
  Let $d$ be a positive integer.
  \begin{enumerate}[label=(\roman*)]
    \item Prove that the discriminant of a polynomial
      $f(x) = \prod_{i=1}^{d}(x - \alpha_i)$ equals the product
      $\pm \prod_{i=1}^{d} f'(\alpha_i)$.
    \item Prove that the discriminant of $x^d - 1$ is $\pm d^d$, and
      note that this is a square in $\Q(\zeta_d)$.
    \item Prove that $\Q(\zeta_{8d})$ contains square roots of $d$ and
      $-d$.
    \item Conclude that every quadratic extension of $\Q$ may be
      embedded in a cyclotomic field $\Q(\zeta_n)$, for a suitable $n$.
  \end{enumerate}
  (This is a very special case of the
  \name{Kronecker}--\name{Weber} theorem.) [\S{}7.6]
\end{exer}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-engine: xetex
%%% TeX-master: "../master"
%%% End:
