\chapter{Groups, second encounter}

In this chapter we return to Grp and study several topics of a less `general'
nature
than those considered in Chapter II. Most of what we do here will apply
exclusively
to finite groups; this is an important example in its own right, as it has
spectacular
applications (for example, in Galois theory; cf. \S{}VII.7), and it is a good
subject
from the expository point of view, since it gives us the opportunity to see
several
general concepts at work in a context that is complex enough to carry substance,
but simple enough (in this tiny selection of elementary topics) to be
appreciated
easily.

\section{The conjugation action}


\subsection{Actions of groups on sets, reminder}
Groups really shine when you let

them act on something. This section will make this point very effectively, since
we will get surprisingly precise results on finite groups by extremely
simple-minded
applications of the elementary facts concerning group actions that we
established
back in \S{}II.9.
Recall that we proved (Proposition II.9.9) that every transitive (left-) action
of
a group G on a set S is, up to a natural notion of isomorphism,
`left-multiplication
on the set of left-cosets G/H'. Here, H may be taken to be the stabilizer Stab
G(a)
of any element a \in  S, that is (Definition II.9.8) the subgroup of G fixing a.
This
fact applies to the orbits of every left-action of G on a set; in particular,
the number
of elements in a finite orbit O equals the index of the stabilizer of any a \in
O; in
particular (Corollary II.9.10) the number of elements |O| of an orbit must
divide
the order |G| of G, if G is finite.
These considerations may be packaged into a useful `counting' formula, which
we could call the class formula for that action; this name is usually reserved
to the
particular case of the action of G onto itself by conjugation, which we will
explore
more carefully below.
In order to state the formula, assume G acts on a set S; for a \in  S, let Ga
denote the stabilizer StabG (a). Also, let Z be the set of fixed points of the
action:
Z = {a \in  S | (\forall g \in  G) : ga = a}.
Note that a \in  Z \iff  Ga = G; we could say that a \in  Z if and only if the
orbit
of a is `trivial', in the sense that it consists of a alone.
Proposition 1.1. Let S be a finite set, and let G be a group acting on S. With
notation as above,
?
|S| = |Z| +
 [G : Ga],
a\in A
where A \subseteq  S has exactly one element for each nontrivial orbit of the
action.
Proof. The orbits form a partition of S, and Z collects the trivial orbits;
hence
?
|S| = |Z| +
 |Oa |,
a\in A
where Oa denotes the orbit of a. By Proposition II.9.9, the order |Oa | equals
the
index of the stabilizer of a, yielding the statement.
 ?
The main strength of Proposition 1.1 rests in the fact that, if G is finite,
each
summand [G : Ga ] divides the order of G (and is > 1). This can be a strong
constraint, when some information is known about |G|. For example, let's see
what
this says when G is a p-group:
Definition 1.2. A p-group is a finite group whose order is a power of a prime
integer p.
 ?
Corollary 1.3. Let G be a p-group acting on a finite set S, and let Z be the
fixed
point set of the action. Then
|Z| \equiv  |S| mod p.
Proof. Indeed, each summand [G : Ga ] in Proposition 1.1 is a number larger
than 1, and a power of p; hence it is 0 mod p.
 ?
For instance, in certain situations this can be used to establish1 that Z =
\emptyset : see
Exercise 1.1. Such immediate consequences of Proposition 1.1 will assist us
below,
in the proof of Sylow's theorems.
1 In this sense, Proposition 1.1 is an instance of a class of results known as
`fixed point
theorems'. The reader will likely encounter a few such theorems in topology
courses, where the
role of the `size' of a set may be played by (for example) the Euler
characteristic of a topological
space.

\subsection{Center, centralizer, conjugacy classes}
Recall (Example II.9.3) that ev-

ery group G acts on itself in at least two interesting ways: by (left-)
multiplication
and by conjugation. The latter action is defined by the following \rho  : G
\times  G \to  G:
\rho (g, a) = gag -1.
As we know (\S{}II.9.2), this datum is equivalent to the datum of a certain
group
homomorphism:
\sigma  : G \to  SG
from G to the permutation group on G.
This action highlights several interesting objects:
Definition 1.4. The center of G, denoted Z(G), is the subgroup ker \sigma  of G.
 ?
Concretely, the center2 of G is
Z(G) = {g \in  G | (\forall a \in  G) : ga = ag}.
Indeed, \sigma (g) is the identity in S G if and only if \sigma (g) acts as the
identity on G; that
is, if and only if gag-1 = a for all a \in  G; that is, if and only if g
commutes with all
elements of G. In other words, the center is the set of fixed points in G under
the
conjugation action.
Note that the center of a group G is automatically normal in G: this is nearly
immediate to check `by hand', but there is no need to do so since it is a kernel
by
definition and kernels are normal.
A group G is commutative if and only if Z(G) = G, that is, if and only if the
conjugation action is trivial on G.
In general, feel happy when you discover that the center of a group is not
trivial:
this will often allow you to set up proofs by induction on the number of
elements
of the group, by mod-ing out by the center (this is, roughly, how we will prove
the
first Sylow theorem). Or note the following useful fact, which comes in handy
when
trying to prove that a group is commutative:
Lemma 1.5. Let G be a finite group, and assume G/Z(G) is cyclic. Then G is
commutative (and hence G/Z(G) is in fact trivial).
Proof. (Cf. Exercise 1.5.) As G/Z(G) is cyclic, there exists an element g \in  G
such
that the class gZ(G) generates G/Z(G). Then \forall a \in  G
aZ(G) = (gZ(G))r
for some r \in  Z; that is, there is an element z \in  Z(G) of the center such
that
a = g r z.
If now a, b are in G, use this fact to write
a = g r z,
 b = g s w
for some s \in  Z and w \in  Z(G); but then
ab = (g r z)(g sw) = g r+szw = (g s w)(g r z) = ba,
2 Why `Z' ? `Center'
 i\'C Zentrum auf Deuts\v{E}.
where I have used the fact that z and w commute with every element of G. As a
and b were arbitrary, this proves that G is commutative.
 ?
Next, the stabilizer of a \in  G under conjugation has a special name:
Definition 1.6. The centralizer (or normalizer ) ZG (a) of a \in  G is its
stabilizer
under conjugation.
 ?
Thus,
ZG (a) = {g \in  G | gag-1 = a} = {g \in  G | ga = ag}
consists of those elements in G which commute with a. In particular, Z(G)
\subseteq  ZG(a)
?
for all a \in  G; in fact, Z(G) = a\in G ZG (a). Clearly a \in  Z(G) \iff  ZG(a)
= G.
If there is no ambiguity concerning the group G containing a, the index G may
be dropped.
Definition 1.7. The conjugacy class of a \in  G is the orbit [a] of a under the
conjugation action. Two elements a, b of G are conjugate if they belong to the
same conjugacy class.
 ?
The notation [a] is not standard; C(a) is used more frequently, but I am not
fond of it. Using [a] reminds us that these are nothing but the equivalence
classes
of elements of G under a certain interesting equivalence relation.
Note that [a] = {a} if and only if gag -1 = a for all g \in  G; that is, if and
only
if ga = ag for all g \in  G; that is, if and only if a \in  Z(G).

\subsection{The Class Formula}
The `official' Class Formula for a finite group G is the

particular case of Proposition 1.1 for the conjugation action.
Proposition 1.8 (Class formula). Let G be a finite group. Then
?
|G| = |Z(G)| +
 [G : Z(a)],
a\in A
where A \subseteq  G is a set containing one representative for each nontrivial
conjugacy
class in G.
Proof. The set of fixed points is Z(G), and the stabilizer of a is the
centralizer Z(a);
apply Proposition 1.1.
 ?
The class formula is surprisingly useful. In applying it, keep in mind that
every
summand on the right (that is, both |Z(G)| and each [G : Z(a)]) is a divisor of
|G|;
this fact alone often suffices to draw striking conclusions about G.
Possibly the most famous such application is to p-groups, via Corollary 1.3:
Corollary 1.9. Let G be a nontrivial p-group. Then G has a nontrivial center.
Proof. Since |Z(G)| \equiv  |G| mod p and |G| > 1 is a power of p, necessarily
|Z(G)| is
a multiple of p. As Z(G) = \emptyset  (since eG \in  Z(G)), this implies |Z(G)|
\geq  p.
 ?
For example, it follows immediately (from Corollary 1.9 and Lemma 1.5; cf. Ex-
ercise 1.6) that if p is prime, then every group of order p2 is commutative.
In general, the class formula poses a strong constraint on what can go on in a
group.
Example 1.10. Consider a group G of order 6; what are the possibilities for its
class formula?
If G is commutative, then the class formula will tell us very little:
6 = 6.
If G is not commutative, then its center must be trivial (as a consequence of
La-
grange's theorem and Lemma 1.5); so the class formula is 6 = 1 + \dots  , where
\dots
collects the sizes of the nontrivial conjugacy classes. But each of these
summands
must be larger than 1, smaller than 6, and must divide 6; that is, there are no
choices:
6=1+2+3
is the only possibility. The reader should check that this is indeed the class
formula
for S3 ; in fact, S3 is the only noncommutative group of order 6 up to
isomorphism
(Exercise 1.13).
 ?
Another useful observation is that normal subgroups must be unions of conju-
gacy classes: because if H is a normal subgroup, a \in  H, and b = gag-1 is
conjugate
to a, then
b \in  gHg-1 = H.
To stick with the |G| = 6 example, note that every subgroup of a group must
contain the identity and its size must divide the order of the group; it follows
that
a normal subgroup of a noncommutative group of order 6 cannot have order 2,
since
2 cannot be written as sums of orders of conjugacy classes (including the class
of
the identity).

\subsection{Conjugation of subsets and subgroups}
We may also act by conjugation

on the power set of G: if A \subseteq  G is a subset and g \in  G, the conjugate
of A is
the subset gAg -1 . By cancellation, the conjugation map a ?\to  gag-1 is a
bijection
between A and gAg -1.
This leads to terminology analogous to the one introduced in \S{}1.2.
Definition 1.11. The normalizer NG (A) of A is its stabilizer under conjugation.
The centralizer of A is the subgroup ZG (A) \subseteq  NG (A) fixing each
element of A. ?
Thus, g \in  NG (A) if and only if3 gAg-1 = A, and g \in  ZG(A) if and only if
\forall a \in  A, gag -1 = a.
For A = {a} a singleton, we have NG ({a}) = ZG ({a}) = ZG (a). In general,
ZG (A) ? NG (A).
If H is a subgroup of G, every conjugate gHg-1 of H is also a subgroup of G;
conjugate subgroups have the same order.
3 If A is finite (but not in general), this condition is equivalent to gAg-1
\subseteq  A.
Remark 1.12. The definition implies immediately that H \subseteq  NG (H) and
that H
is normal in G if and only if NG (H) = G. More generally, the normalizer NG (H)
of H in G is (clearly) the largest subgroup of G in which H is normal.
 ?
One could apply Proposition 1.1 to the conjugation action on subsets or sub-
group; however, there are too many subsets, and one has little control over the
number of subgroups. Other numerical considerations involving the number of
conjugates of a given subset or subgroups may be very useful.
Lemma 1.13. Let H \subseteq  G be a subgroup. Then (if finite) the number of
subgroups
conjugate to H equals the index [G : NG (H)] of the normalizer of H in G.
Proof. This is again an immediate consequence of Proposition II.9.9.
 ?
Corollary 1.14. If [G : H] is finite, then the number of subgroups conjugate to
H
is finite and divides [G : H].
Proof.
(cf. \S{}II.8.5).
[G : H] = [G : NG (H)] \cdot  [NG (H) : H]
?
One of the celebrated Sylow theorems will strengthen this statement substan-
tially in the case in which H is a maximal p-group contained in a finite group
G.
For a statement concerning the size of the normalizer of an arbitrary p-subgroup
of
a group, see Lemma 2.9.
Another useful numerical tool is the observation that if H and K are subgroups
of a group G and H \subseteq  NG (K)---so that gKg -1 = K for all g \in
H---then conjuga-
tion by g \in  H gives an automorphism of K. Indeed, I have already observed
that
conjugation is a bijection, and it is immediate to see that it is a
homomorphism:
\forall k 1 , k2 \in  K
(gk1 g -1 )(gk2 g-1 ) = gk 1 (g -1 g)k2g -1 = g(k1k2 )g -1 .
Thus, conjugation gives a set-function
\gamma  : H \to  AutGrp (K).
The reader will check that this is a group homomorphism and will determine ker
\gamma
(Exercise 1.21).
This is especially useful if H is finite and some information is available con-
cerning AutGrp (K) (for an example, see Exercise 4.14). A classic application is
presented in Exercise 1.22.

\subsection*{Exercises}

1.1. ? Let p be a prime integer, let G be a p-group, and let S be a set such
that
|S| \equiv  0 mod p. If G acts on S, prove that the action must have fixed
points. [\S{}1.1,
\S{}2.3]
1.2. Find the center of D2n . (The answer depends on the parity of n. You have
actually done this already: Exercise II.2.7. This time, use a presentation.)
1.3. Prove that the center of Sn is trivial for n \geq  3. (Suppose that \sigma
\in  Sn sends
a to b = a, and let c = a, b. Let \tau  be the permutation that acts solely by
swapping
b and c. Then compare the action of \sigma \tau  and \tau  \sigma  on a.)
1.4. ? Let G be a group, and let N be a subgroup of Z(G). Prove that N is normal
in G. [\S{}2.2]
1.5. ? Let G be a group. Prove that G/Z(G) is isomorphic to the group Inn(G) of
inner automorphisms of G. (Cf. Exercise II.4.8.) Then prove Lemma 1.5 again by
using the result of Exercise II.6.7. [\S{}1.2]
1.6. ? Let p, q be prime integers, and let G be a group of order pq. Prove that
either G is commutative or the center of G is trivial. Conclude (using Corollary
1.9)
that every group of order p2 , for a prime p, is commutative. [\S{}1.3]
1.7. Prove or disprove that if p is prime, then every group of order p3 is
commu-
tative.
1.8. ? Let p be a prime number, and let G be a p-group: |G| = pr . Prove that G
contains a normal subgroup of order p k for every nonnegative k \leq  r.
[\S{}2.2]
1.9. \lnot  Let p be a prime number, G a p-group, and H a nontrivial normal
subgroup
of G. Prove that H \cap  Z(G) = {e}. (Hint: Use the class formula.) [3.11]
1.10. Prove that if G is a group of odd order and g \in  G is conjugate to g -1
, then
g = eG .
1.11. Let G be a finite group, and suppose there exist representatives g1, . . .
, gr of
the r distinct conjugacy classes in G, such that \forall i, j, gi gj = gj gi .
Prove that G is
commutative. (Hint: What can you say about the sizes of the conjugacy classes?)
1.12. Verify that the class formula for both D8 and Q8 (cf. Exercise III.1.12)
is
8 = 2 + 2 + 2 + 2. (Also note that D8 = \sim 
 Q8 .)
1.13. ? Let G be a noncommutative group of order 6. As observed in Example 1.10,
G must have trivial center and exactly two conjugacy classes, of order 2 and 3.
\bullet  Prove that if every element of a group has order \leq  2, then the group is com-
mutative. Conclude that G has an element y of order 3.
\bullet  Prove that ?y? is normal in G.
\bullet  Prove that [y] is the conjugacy class of order 2 and [y] = {y, y 2 }.
\bullet  Prove that there is an x \in  G such that yx = xy2 .
\bullet  Prove that x has order 2.
\bullet  Prove that x and y generate G.
\bullet  Prove that G \sim 
 = S3 .
[\S{}1.3, \S{}2.5]
1.14. Let G be a group, and assume [G : Z(G)] = n is finite. Let A \subseteq  G
be any
subset. Prove that the number of conjugates of A is at most n.
1.15. Suppose that the class formula for a group G is 60 = 1 + 15 + 20 + 12 +
12.
Prove that the only normal subgroups of G are {e} and G.
1.16. ? Let G be a finite group, and let H \subseteq  G be a subgroup of index
2. For
a \in  H, denote by [a]H , resp., [a]G , the conjugacy class of a in H, resp.,
G. Prove
that either [a]H = [a]G or [a]H is half the size of [a]G , according to whether
the
centralizer ZG(a) is not or is contained in H. (Hint: Note that H is normal in
G,
by Exercise II.8.2; apply Proposition II.8.11.) [\S{}4.4]
1.17. \lnot  Let H be a proper subgroup of a finite group G. Prove that G is not
the
union of the conjugates of H. (Hint: You know the number of conjugates of H;
keep in mind that any two subgroups overlap, at least at the identity.) [1.18,
1.20]
1.18. Let S be a set endowed with a transitive action of a finite group G, and
assume |S| \geq  2. Prove that there exists a g \in  G without fixed points in
S, that
is, such that gs = s for all s \in  S. (Hint: By Proposition II.9.9, you may
assume
S = G/H, with H proper in G. Use Exercise 1.17.)
1.19. Let H be a proper subgroup of a finite group G. Prove that there exists a
g \in  G whose conjugacy class is disjoint from H.
1.20. Let G = GL2 (C), and let H be the subgroup consisting of upper triangular
matrices (Exercise II.6.2). Prove that G is the union of the conjugates of H.
Thus,
the finiteness hypothesis in Exercise 1.17 is necessary. (Hint: Equivalently,
prove
that every 2 \times  2 matrix is conjugate to a matrix in H. You will use the
fact that
C is algebraically closed; see Example III.4.14.)
1.21. ? Let H, K be subgroups of a group G, with H \subseteq  NG(K). Verify that
the
function \gamma  : H \to  AutGrp(K) defined by conjugation is a homomorphism of
groups
and that ker \gamma  = H \cap  ZG (K), where ZG (K) is the centralizer of K.
[\S{}1.4, 1.22]
1.22. ? Let G be a finite group, and let H be a cyclic subgroup of G of order p.
Assume that p is the smallest prime dividing the order of G and that H is normal
in G. Prove that H is contained in the center of G.
(Hint: By Exercise 1.21 there is a homomorphism \gamma  : G \to  AutGrp (H); by
Exercise II.4.14, AutGrp (H) has order p - 1. What can you say about \gamma ?)
[\S{}1.4]

\section{The Sylow theorems}


\subsection{Cauchy's theorem}
The `Sylow theorems' consist of three statements con-

cerning p-subgroups (cf. Definition 1.2) of a given finite group G. The form I
will
give for the first of these statements will tell us that G contains p-groups of
all sizes
allowed by Lagrange's theorem: if p is a prime and pk divides |G|, then G
contains
a subgroup of order pk . The proof of this statement is an easy induction,
provided
the statement for k = 1 is known: that is, provided that one has established
Theorem 2.1 (Cauchy's theorem). Let G be a finite group, and let p be a prime
divisor of |G|. Then G contains an element of order p.
As it happens, only the abelian version of this statement is needed for the
proof
of the first Sylow theorem; then the full statement of Cauchy's theorem follows
from the first Sylow theorem itself. Since the (diligent) reader has already
proved
Cauchy's theorem for abelian groups (in Exercise II.8.17), we could directly
move
on to Sylow theorems.
However, there is a quick proof4 of the full statement of Cauchy's theorem
which does not rely on Sylow and is a good illustration of the power of the
general
`class formula for arbitrary actions' (Proposition 1.1). I will present this
proof,
while also encouraging the reader to go back and (re)do Exercise II.8.17 now.
Proof of Theorem 2.1. Consider the set S of ordered p-tuples of elements of G:
(a1 , . . . , ap )
such that a1 \dots  ap = e. I claim that |S| = |G|p-1 : indeed, once a1, . . . ,
ap-1 are
chosen (arbitrarily), then ap is determined as it is the inverse of a1 \dots  a
p-1 .
Therefore, p divides the order of S as it divides the order of G.
Also note that if a1 \dots  ap = e, then
a2 \dots  apa1 = e
(even if G is not commutative): because if a1 is a left-inverse to a2 \dots  ap,
then it is
also a right-inverse to it.
Therefore, we may act with the group Z/pZ on S: given [m] in Z/pZ, with
0 \leq  m < p, act by [m] on
(a1 , . . . , ap )
by sending it to
(am+1 , . . . , ap , a1 , . . . , am) :
as we just observed, this is still an element of S.
Now Corollary 1.3 implies
|Z| \equiv  |S| \equiv  0 mod p,
where Z is the set of fixed points of this action. Fixed points are p-tuples of
the
form
(*)
 (a, . . . , a);
and note that Z = \emptyset , since {e, . . . , e} \in  Z. Since p \geq  2 and p
divides |Z|, we
conclude that |Z| > 1; therefore there exists some element in Z of the form (*),
with a = e.
4This argument is apparently due to James McKay.
This says that there exists an a \in  G, a = e, such that ap = e, proving the
statement.
 ?
We should remark that the proof given here proves a more precise result than
the raw statement of Theorem 2.1: every element of order p in G generates a
cyclic
subgroup of G of order p, and we are able to say something about the number of
such subgroups.
Claim 2.2. Let G be a finite group, let p be a prime divisor of |G|, and let N
be
the number of cyclic subgroups of G of order p. Then N \equiv  1 mod p.
The proof of this fact is left to the reader (as an incentive to really
understand
the proof of Theorem 2.1).
Claim 2.2, coupled with the simple observation that if there is only 1 cyclic
subgroup H of order p, then that subgroup must be normal (Exercise 2.2),
suffices
for interesting applications.
Definition 2.3. A group G is simple if it is nontrivial and its only normal sub-
groups are {e} and G itself.
 ?
Simple groups occupy a special place in the theory of groups: one can `break
up' any finite group into basic constituents which are simple groups; we will
see
how this is done in \S{}3.1. Thus, it is important to be able to tell whether a
group is
simple or not5.
Example 2.4. Let p be a positive prime integer. If |G| = mp, with 1 < m < p,
then G is not simple.
Indeed, consider the subgroups of G with p elements. By Claim 2.2, the number
of such subgroups is \equiv  1 mod p. Thus, if there is more than one such
subgroup,
then there must be at least p + 1. Any two distinct subgroups of prime order can
only meet at the identity (why?); therefore this would account for at least
1 + (p + 1)(p - 1) = p2
elements in G. Since |G| = mp < p 2 , this is impossible. Therefore there is
only
one cyclic subgroup of order p in G, which must be normal as mentioned above,
proving that G is not simple.
 ?

\subsection{Sylow I. Let p be a prime integer}
A p-Sylow subgroup of a finite group G

is a subgroup of order pr , where |G| = pr m and gcd(p, m) = 1. That is, P
\subseteq  G is
a p-Sylow subgroup if it is a p-group and p does not divide [G : P ].
If p does not divide the order of G, then G contains a p-Sylow subgroup: namely,
{e}. This is not very interesting; what is interesting is that G contains a
p-Sylow
subgroup even when p does divide the order of G:
Theorem 2.5 (First Sylow theorem). Every finite group contains a p-Sylow sub-
group, for all primes p.
5 In fact, a complete list of all finite simple groups is known: this is the
classification result
mentioned at the end of \S{}II.6.3, arguably one of the deepest and hardest
results in mathematics.
The first Sylow theorem follows from the seemingly stronger statement:
Proposition 2.6. If pk divides the order of G, then G has a subgroup of order pk
.
The statements are actually easily seen to be equivalent, by Exercise 1.8; in
any case, the standard argument proving Theorem 2.5 proves Proposition 2.6, and
I see no reason to hide this fact. Here is the argument:
Proof of Proposition 2.6. If k = 0, there is nothing to prove, so we may assume
k \geq  1 and in particular that |G| is a multiple of p.
Argue by induction on |G|: if |G| = p, again there is nothing to prove; if |G| >
p
and G contains a proper subgroup H such that [G : H] is relatively prime to p,
then pk divides the order of H, and hence H contains a subgroup of order pk by
the induction hypothesis, and thus so does G.
Therefore, we may assume that all proper subgroups of G have index divisible
by p. By the class formula (Proposition 1.8), p divides the order of the center
Z(G).
By Cauchy's theorem6 , \exists a \in  Z(G) such that a has order p. The cyclic
subgroup
N = ?a? is contained in Z(G), and hence it is normal in G (Exercise 1.4).
Therefore
we can consider the quotient G/N .
Since |G/N | = |G|/p and pk divides |G| by hypothesis, we have that pk-1
divides the order of G/N . By the induction hypothesis, we may conclude that
G/N contains a subgroup of order pk-1 . By the structure of the subgroups of
a quotient (\S{}II.8.3, especially Proposition II.8.9), this subgroup must be of
the
form P/N , for P a subgroup of G.
But then |P | = |P/N | \cdot  |N | = pk-1 \cdot  p = pk , as needed.
 ?
There are slicker ways to prove Theorem 2.5. We will see a pretty (and in-
sightful) alternative in \S{}2.3; but the proof given above is easy to remember
and is
a good template for similar arguments.
Remark 2.7. The diligent reader worked out in Exercise II.8.20 a stronger state-
ment than Proposition 2.6, for abelian groups. The arguments are similar; the
advantage in the abelian case is that any cyclic subgroup produced by Cauchy's
theorem is automatically normal, while ensuring normality requires a few twists
and turns in the general case (and, as a result, yields a weaker statement).
 ?

\subsection{Sylow II}
Theorem 2.5 tells us that some maximal p-group in G attains

the largest size allowed by Lagrange's theorem, that is, the maximal power of
the
prime p dividing |G|.
One can be more precise: the second Sylow theorem tells us that every maximal
p-group in |G| is in fact a p-Sylow subgroup. It is as large as is allowed by
Lagrange's
theorem.
The situation is in fact even better: all p-Sylow subgroups are conjugates of
each other7 . Moreover, even better than this, every p-group inside G must be
contained in a conjugate of any fixed p-Sylow subgroup.
6 Note that, as mentioned in \S{}2.1, we only need the abelian case of this
theorem.
7 Of course if P is a p-Sylow subgroup of G, then so are all conjugates gP g-1
of P .
The proof of this very precise result is very easy!
Theorem 2.8 (Second Sylow theorem). Let G be a finite group, let P be a p-Sylow
subgroup, and let H \subseteq  G be a p-group. Then H is contained in a
conjugate of P :
there exists g \in  G such that H \subseteq  gP g-1 .
Proof. Act with H on the set of left-cosets of P , by left-multiplication. Since
there
are [G : P ] cosets and p does not divide [G : P ], we know this action must
have
fixed points (Exercise 1.1): let gP be one of them. This means that \forall h
\in  H:
hgP = gP ;
that is, g -1 hgP = P for all h in H; that is, g-1 Hg \subseteq  P ; that is, H
\subseteq  gP g -1, as
needed.
 ?
We can obtain an even more complete picture of the situation. Suppose we
have constructed a chain
H0 = {e} \subseteq  H1 \subseteq  \dots  \subseteq  Hk
of p-subgroups of a group G, where |Hi | = pi . By Theorem 2.8 we know that Hk
is contained in some p-Sylow subgroup, of order pr = the maximum power of p
dividing the order of G. But I claim that the chain can in fact be continued one
step at a time all the way up to the Sylow subgroup:
H0 = {e} \subseteq  H1 \subseteq  \dots  \subseteq  Hk \subseteq  Hk+1 \subseteq
\dots  \subseteq  Hr ;
and, further, Hk may be assumed to be normal in Hk+1. The following lemma will
simplify the proof of this fact considerably and will also help us prove the
third
Sylow theorem.
Lemma 2.9. Let H be a p-group contained in a finite group G. Then
[NG (H) : H] \equiv  [G : H]
 mod p.
Proof. If H is trivial, then NG (H) = G and the two numbers are equal.
Assume then that H is nontrivial, and act with H on the set of left-cosets of H
in G, by left-multiplication. The fixed points of this action are the cosets gH
such
that \forall h \in  H
hgH = gH,
that is, such that g-1 hg \in  H for all h \in  H; in other words, H \subseteq
gHg -1, and hence
(by order considerations) gHg -1 = H. This means precisely that g \in  NG (H).
Therefore, the set of fixed points of the action consists of the set of cosets
of H
in NG (H).
The statement then follows immediately from Corollary 1.3.
 ?
As a consequence, if Hk is not a p-Sylow subgroup `already', in the sense that p
`still' divides [G : Hk ], then p must also divide [NG(Hk ) : Hk ]. Another
application
of Cauchy's theorem tells us how to obtain the next subgroup Hk+1 in the chain.
More precisely, we have the following result.
Proposition 2.10. Let H be a p-subgroup of a finite group G, and assume that H
\setminus is not a p-Sylow subgroup. Then there exists a p-subgroup H of G containing H,
\setminus  \setminus such that [H : H] = p and H is normal in H .
Proof. Since H is not a p-Sylow subgroup of G, p divides [NG (H) : H], by
Lemma 2.9. Since H is normal in NG (H), we may consider the quotient group
NG (H)/H, and p divides the order of this group. By Theorem 2.1, NG (H)/H has
an element of order p; this generates a subgroup of order p of NG (H)/H, which
must be (cf. \S{}II.8.3) of the form H \setminus  /H for a subgroup H \setminus
of NG (H).
It is straightforward to verify that H \setminus  satisfies the stated
requirements.
 ?
The statement about `chains of p-subgroups' follows immediately from this
result.
Note that Cauchy's theorem and Proposition 2.10 provide a new proof of Propo-
sition 2.6 and hence of the first Sylow theorem.

\subsection{Sylow III}
The third (and last) Sylow theorem gives a good handle on the

number of p-Sylow subgroups of a given finite group G. This is especially useful
in
establishing the existence of normal subgroups of G: since all p-Sylow subgroups
of a group are conjugates of each other (by the second Sylow theorem), if there
is
only one p-Sylow subgroup, then that subgroup must be normal8 .
Theorem 2.11 (Third Sylow theorem). Let p be a prime integer, and let G be
a finite group of order |G| = pr m. Assume that p does not divide m. Then the
number of p-Sylow subgroups of G divides m and is congruent to 1 modulo p.
Proof. Let Np denote the number of p-Sylow subgroups of G.
By Theorem 2.8, the p-Sylow subgroups of G are the conjugates of any given
p-Sylow subgroup P . By Lemma 1.13, Np is the index of the normalizer NG (P )
of P ; thus (Corollary 1.14) it divides the index m of P . In fact,
m = [G : P ] = [G : NG(P )] \cdot  [NG (P ) : P ] = Np \cdot  [NG (P ) : P ].
Now, by Lemma 2.9 we have
m = [G : P ] \equiv  [NG (P ) : P ]multiplying by Np , we get
mNp \equiv  m
 mod p.
Since m \equiv  0 mod p and p is prime, this implies
Np \equiv  1 mod p,
as needed.
mod p;
?
Of course there are other ways to prove Theorem 2.11: see for example Exer-
cise 2.11.
8For an alternative viewpoint, see Exercise 2.2.

\subsection{Applications}
Consequences stemming from the group actions we have en-

countered, and especially the Sylow theorems, may be applied to establish
exquisitely
precise facts about individual groups as well as whole classes of groups; this
is often
based on some simple but clever numerology.
The following examples are exceedingly simple-minded but will hopefully con-
vey the flavor of what can be done with the tools we have built in the previous
two sections. More examples may be found among the exercises at the end of this
section.
2.5.1. More nonsimple groups.
Claim 2.12. Let G be a group of order mpr , where p is a prime integer and
1 < m < p. Then G is not simple.
(Cf. Example 2.4.)
Proof. By the third Sylow theorem, the number Np of p-Sylow subgroups divides m
and is of the form 1 + kp. Since m < p, this forces k = 0, Np = 1. Therefore G
has
a normal subgroup of order pr ; hence it is not simple.
 ?
Of course the same argument gives the same conclusion for every group of order
mpr , where (m, p) = 1 and the only divisor d of m such that d \equiv  1 mod p
is d = 1.
Example 2.13. There are no simple groups of order 2002.
Indeed9 ,
2002 = 2 \cdot  7 \cdot  11 \cdot  13;
the divisors of 2 \cdot  7 \cdot  13 are
1 , 2 , 7 , 13 , 14 , 26 , 91 , 182 :
of these, only 1 is congruent to 1 mod 11.order 11 in every group of order 2002.
Thus there is a normal subgroup of
?
The reader should not expect the third Sylow theorem to always yield its fruits
so readily, however.
Example 2.14. There are no simple groups of order 12.
Note that 3 \equiv  1 mod 2 and 4 \equiv  1 mod 3: thus the argument used above
does not
guarantee the existence of either a normal 2-Sylow subgroup or a normal 3-Sylow
subgroup.
However, suppose that there is more than one 3-Sylow subgroup. Then there
must be 4, by the third Sylow theorem. Since any two such subgroups must
intersect
in the identity, this accounts for exactly 8 elements of order 3. Excluding
these
leaves us with the identity and 3 elements of order 2 or 4; that is just enough
room
to fit one 2-Sylow subgroup. This subgroup will then have to be normal.
Thus, either there is a 3-Sylow normal subgroup or there is a 2-Sylow normal
subgroup---either way, the group is not simple.
 ?
9 It is safe to guess that this statement has been assigned on hundreds of
algebra tests across
the world in the year 2002.
Even this more refined counting will often fail, and one has to dig deeper.
Example 2.15. There are no simple groups of order 24.
Indeed, let G be a group of order 24, and consider its 2-Sylow subgroups; by
the third Sylow theorem, there are either 1 or 3 such subgroups. If there is 1,
the
2-Sylow subgroup is normal and G is not simple. Otherwise, G acts (nontrivially)
by conjugation on this set of three 2-Sylow subgroups; this action gives a
nontrivial
homomorphism G \to  S3 , whose kernel is a proper, nontrivial normal subgroup
of G---thus again G is not simple.
 ?
The reader should practice by selecting a random number n and trying to say
as much as he/she can, in general, about groups of order n. Beware: such
problems
are a common feature of qualifying exams.
2.5.2. Groups of order pq, p < q prime.
Claim 2.16. Assume p < q are prime integers and q \equiv  1 mod p. Let G be a
group
of order pq. Then G is cyclic.
Proof. By the third Sylow theorem, G has a unique (hence normal) subgroup H
of order p. Indeed, the number Np of p-Sylow subgroups must divide q, and q is
prime, so Np = 1 or q. Necessarily Np \equiv  1 mod p, and q \equiv  1 mod p by
hypothesis;
therefore Np = 1.
Since H is normal, conjugation gives an action of G on H, hence (by Exer-
cise 1.21) a homomorphism \gamma  : G \to  Aut(H). Now H is cyclic of order p,
so
| Aut(H)| = p - 1 (Exercise II.4.14); the order of \gamma (G) must divide both
pq and
p - 1, and it follows that \gamma  is the trivial map.
Therefore, conjugation is trivial on H: that is, H \subseteq  Z(G). Lemma 1.5
implies
that G is abelian.
Finally, an abelian group of order pq, with p < q primes, is necessarily cyclic:
indeed it must contain elements g, h of order p, q, respectively (for example by
Cauchy's theorem), and then |gh| = pq by Exercise II.1.14.
 ?
For example, this statement `classifies' all groups of order 15, 33, 35, 51, . .
. :
such groups are necessarily cyclic.
The argument given in the proof is rather `high-brow', as it involves the auto-
morphism group of H; that is precisely why I gave it. For low-brow alternatives,
see Exercise 2.18 or Remark 5.4.
The condition q \equiv  1 mod p in Claim 2.16 is clearly necessary: indeed, |S3|
= 2\cdot 3
is the product of two distinct primes, and yet S3 is not cyclic. The argument
given
in the proof shows that if |G| = pq, with p < q prime, and G has a normal
subgroup
of order p, then G is cyclic. If q \equiv  1 mod p, it can be shown that there
is in fact
a unique noncommutative group of order pq up to isomorphism: the reader will
work this out after learning about semidirect products (Exercise 5.12). But we
are
in fact already in the position of obtaining rather sophisticated information
about
this group, even without knowing its construction in general (Exercise 2.19).
For fun, let's tackle the case in which p = 2.
Claim 2.17. Let q be an odd prime, and let G be a noncommutative group of
order 2q. Then G \sim 
 = D2q , the dihedral group.
Proof. By Cauchy's theorem, \exists y \in  G such that y has order q. By the
third Sylow
theorem, ?y? is the unique subgroup of order q in G (and is therefore normal).
Since G is not commutative and in particular it is not cyclic, it has no
elements of
order 2q; therefore, every element in the complement of ?y? has order 2; let x
be
any such element.
The conjugate xyx-1 of y by x is an element of order q, so xyx-1 \in  ?y?. Thus,
xyx-1
 = y r for some r between 0 and q - 1.
Now observe that
(yr )r = (xyx-1 )r = xy r x-1 = x2 y(x-1 )2 = y
since |x| = 2. Therefore, y r2
-1 = e, which implies
q | (r 2 - 1) = (r - 1)(r + 1)
by Corollary II.1.11. Since q is prime, this says that q | (r - 1) or q | (r +
1); since
0 \leq  r \leq  q - 1, it follows that r = 1 or r = q - 1.
If r = 1, then xyx-1 = y; that is, xy = yx. But then the order of xy is 2q (by
Exercise II.1.14), and G is cyclic, a contradiction.
Therefore r = q - 1, and we have established the relations
\{ 2
|
 | x = e,
yq = e,
|
 \}
 q-1yx = xy .
These are the relations satisfied by generators x, y of D2q , as the reader
hopefully
verified in Exercise II.2.5; the statement follows.
 ?
Claim 2.17 yields a classification of groups of order 2q, for q an odd prime:
such
a group must be either abelian (and hence cyclic, by the usual considerations)
or
isomorphic to a dihedral group. For q = 3, we recover the result of Exercise
1.13:
every noncommutative group of order 6 is isomorphic to D6 \sim 
 = S3 .

\subsection*{Exercises}

2.1. ? Prove Claim 2.2. [\S{}2.1]
2.2. ? Let G be a group. A subgroup H of G is characteristic if \phi (H)
\subseteq  H for
every automorphism \phi  of G.
\bullet  Prove that characteristic subgroups are normal.
\bullet  Let H \subseteq  K \subseteq  G, with H characteristic in K and K normal in G. Prove that
H is normal in G.
\bullet  Let G, K be groups, and assume that G contains a single subgroup H isomor-
phic to K. Prove that H is normal in G.
\bullet  Let K be a normal subgroup of a finite group G, and assume that |K| and
|G/K| are relatively prime. Prove that K is characteristic in G.
[\S{}2.1, \S{}2.4, 2.13, \S{}3.3]
2.3. Prove that a nonzero abelian group G is simple if and only if G = \sim 
 Z/pZ for
some positive prime integer p.
2.4. ? Prove that a nontrivial group G is simple if and only if its only
homomorphic
images (i.e., groups G\setminus  such that there is an onto homomorphism G \to
G\setminus ) are the
trivial group and G itself (up to isomorphism). [\S{}3.2]
2.5. Let G be a simple group, and assume \phi  : G \to  G\setminus  is a
nontrivial group
homomorphism. Prove that \phi  is injective.
2.6. Prove that there are no simple groups of order 4, 8, 9, 16, 25, 27, 32, or
49.
In fact, prove that no p-group of order \geq  p2 is simple.
2.7. Prove that there are no simple groups of order 6, 10, 14, 15, 20, 21, 22,
26,
28, 33, 34, 35, 38, 39, 42, 44, 46, 51, 52, 55, 57, or 58. (Hint: Example 2.4.)
2.8. Let G be a finite group, p a prime integer, and let N be the intersection
of
the p-Sylow subgroups of G. Prove that N is a normal p-subgroup of G and that
every normal p-subgroup of G is contained in N . (In other words, G/N is final
with respect to the property of being a homomorphic image of G of order
|G|/p\alpha
for some \alpha .)
2.9. \lnot  Let P be a p-Sylow subgroup of a finite group G, and let H \subseteq
G be a p-
subgroup. Assume H \subseteq  NG (P ). Prove that H \subseteq  P . (Hint: P is
normal in NG (P ),
so P H is a subgroup of NG (P ) by Proposition II.8.11, and |P H/P | = |H/(P
\cap  H)|.
Show that this implies that P H is a p-group, and hence P H = P since P is a
maximal p-subgroup of G. Deduce that H \subseteq  P .) [2.10]
2.10. \lnot  Let P be a p-Sylow subgroup of a finite group G, and act with P by
conjugation on the set of p-Sylow subgroups of G. Show that P is the unique
fixed
point of this action. (Hint: Use Exercise 2.9.) [2.11]
2.11. ? Use the second Sylow theorem, Corollary 1.14, and Exercise 2.10 to paste
together an alternative proof of the third Sylow theorem. [\S{}2.4]
2.12. Let P be a p-Sylow subgroup of a finite group G, and let H \subseteq  G be
a
subgroup containing the normalizer NG (P ). Prove that [G : H] \equiv  1 mod p.
2.13. \lnot  Let P be a p-Sylow subgroup of a finite group G.
\bullet  Prove that if P is normal in G, then it is in fact characteristic in G (cf. Exer-
cise 2.2).
\bullet  Let H \subseteq  G be a subgroup containing the Sylow subgroup P . Assume P is
normal in H and H is normal in G. Prove that P is normal in G.
\bullet  Prove that NG (N G(P )) = NG (P ).
[3.12]
2.14. Prove that there are no simple groups of order 18, 40, 45, 50, or 54.
2.15. Classify all groups of order n \leq  15, n = 8, 12: that is, produce a
list of
nonisomorphic groups such that every group of order n = 8, 12, n \leq  15 is
isomorphic
to one group in the list.
2.16. \lnot  Let G be a noncommutative group of order 8.
\bullet  Prove that G contains elements of order 4 and no elements of order 8.
\bullet  Let y be an element of order 4. Prove that G is generated by y and by an
element x \in  ?y?, such that x2 = e or x2 = y2 .
\bullet  In either case, G = {e, y, y 2 , y 3 , x, yx, y2 x, y 3 x}. Prove that the multiplication
table of G is determined by whether x2 = e or x2 = y2 , and by the value of xy.
\bullet  Prove that necessarily xy = y 3 x. (Hint: To eliminate xy = y 2 x, multiply on
the right by y.)
\bullet  Prove that G \sim 
 = D8 or G \sim 
 = Q8.
[6.2, VII.6.6]
2.17. \lnot  Let R be a division ring (Definition III.1.13), and assume |R| =
64. Prove
that R is necessarily commutative (hence, a field), as follows:
\bullet  The group of units of R has order 63. Prove it has a commutative subgroup G
of order 9. (Sylow.)
\bullet  Prove that R is the only sub-division ring of R containing G.
\bullet  Prove that the set of elements of R commuting with every element of G is a
sub-division ring of R containing G. (Cf. Exercise III.2.10.)
\bullet  Conclude that G is contained in the center of R. Recall that the center of
R is a sub-division ring of R (cf. Exercise III.2.9), and conclude that R is
commutative.
Like Exercise III.2.11, this is a particular case of a theorem of Wedderburn,
accord-
ing to which every finite division ring is a field. [VII.5.16]
2.18. ? Give an alternative proof of Claim 2.16 as follows: use the third Sylow
theorem to count the number of elements of order p and q in G; use this to show
that there are elements in G of order neither 1 nor p nor q; deduce that G is
cyclic.
[\S{}2.5]
2.19. ? Let G be a noncommutative group of order pq, where p < q are primes.
\bullet  Show that q \equiv  1 mod p.
\bullet  Show that the center of G is trivial.
\bullet  Draw the lattice of subgroups of G.
\bullet  Find the number of elements of each possible order in G.
\bullet  Find the number and size of the conjugacy classes in G.
[\S{}2.5]
2.20. How many elements of order 7 are there in a simple group of order 168?
2.21. Let p < q < r be prime integers, and let G be a group of order pqr. Prove
that G is not simple.

\section{Composition series and solvability}


\subsection{Let G be a finite group, n = |G|, and p be a prime divisor of n}
Assume

that the only divisor of n that is congruent to 1 modulo p is 1. Prove that G is
not
simple.

\subsection{\lnot  Let Np denote the number of p-Sylow subgroups of a group G}
Prove

that if G is simple, then |G| divides Np ! for all primes p in the factorization
of |G|.
More generally, prove that if G is simple and H is a subgroup of G of index N >
1,
then |G| divides N !. (Hint: Exercise II.9.12.) This problem capitalizes on the
idea
behind Example 2.15. [2.25]

\subsection{? Prove that there are no noncommutative simple groups of order less than 60}

If you have sufficient stamina, prove that the next possible order for a
noncommu-
tative simple group is 168. (Don't feel too bad if you have to cheat and look up
a
few particularly troublesome orders > 60.) [\S{}4.4]

\subsection{\lnot  Assume that G is a simple group of order 60}

\bullet  Use Sylow's theorems and simple numerology to prove that G has either five
or fifteen 2-Sylow subgroups, accounting for fifteen elements of order 2 or 4.
(Exercise 2.23 will likely be helpful.)
\bullet  If there are fifteen 2-Sylow subgroups, prove that there exists an element g \in  G
of order 2 contained in at least two of them. Prove that the centralizer of g
has
index 5.
Conclude that every simple group10 of order 60 contains a subgroup of index 5.
[4.22]
I have claimed that simple groups (in the sense of Definition 2.3) are the
`basic
constituents' of all finite groups. Among other things, the material in this
section
will (partially) justify this claim.

\subsection{The Jordan-H\"older theorem}
A series of subgroups Gi of a group G is a

decreasing sequence of subgroups starting from G:
G = G0 ? G1 ? G2 ? \dots  .
The length of a series is the number of strict inclusions.
A series is normal if Gi+1 is normal in Gi for all i. We will be interested in
the maximal length of a normal series in G; if finite, I will denote this
number11
by (G). The number (G) is a measure of how far G is from being simple. Indeed,
(G) = 0 if and only if G is trivial, and (G) = 1 if and only if G is simple: for
a
simple group, the only maximal normal series is
G ? {e}.
10 The reader will prove later (Exercise 4.22) that there is in fact only one
simple group of
order 60 up to isomorphism and that this group contains exactly five 2-Sylow
subgroups. The
result obtained here will be needed to establish this fact.
11There does not appear to be a standard notation for this concept.
Definition 3.1. A composition series for G is a normal series
G = G0 ? G1 ? G2 ? \dots  ? Gn = {e}
such that the successive quotients Gi /Gi+1 are simple.
?
It is clear (by induction on the order) that finite groups have composition
series,
while infinite groups do not necessarily have one (Exercise 3.3). It is also
clear that
if a normal series has maximal length (G), then it is a composition series. What
is not clear is that the converse holds: conceivably, there could exist
composition
series of different lengths (the longest ones having length (G)). For example,
why
can't there be a finite group G with (G) = 3 and two different composition
series
G ? G1 ? G2 ? {e}
and
G
 ?
 G\setminus  1
 ?
 {e}
(that is: a finite group G with (G) = 3 and a simple normal subgroup G\setminus
1 such
that G/G\setminus  1 is simple)?
Part of the content of the Jordan-H\"older theorem is that (luckily) this cannot
happen. In fact, the theorem is much more precise: not only do all composition
series have the same length, but they also have the same quotients (appearing,
however, in possibly different orders).
Theorem 3.2 (Jordan-H\"older). Let G be a group, and let
G = G0 ? G1 ? G2 ? \dots  ? Gn = {e},
G = G\setminus  0 ? G\setminus  1 ? G\setminus  2 ? \dots  ? Gm \setminus  = {e}
be two composition series for G. Then m = n, and the lists of quotient groups
Hi = Gi /Gi+1 , Hi \setminus  = G\setminus  i /G\setminus  i+1 agree (up to
isomorphism) after a permutation of
the indices.
Proof.(*)
Let
G = G0 ? G1 ? G2 ? \dots  ? Gn = {e}
be a composition series. Argue by induction on n: if n = 0, then G is trivial,
and
there is nothing to prove. Assume n > 0, and let
(**)
 G = G \setminus  0 ? G\setminus  1 ? G\setminus  2 ? \dots  ? G\setminus  m =
 {e}
be another composition series for G. If G1 = G\setminus  1 , then the result
follows from the
induction hypothesis, since G1 has a composition series of length n - 1 < n.
We may then assume G1 = G\setminus  1 . Note that G1G\setminus  1 = G: indeed,
G1 G\setminus  1 is normal
in G (Exercise 3.5), and G1 ? G1G\setminus  1 ; but there are no proper normal
subgroups
between G1 and G since G/G1 is simple.
Let K = G1 \cap  G\setminus  1 . The distinct subgroups Gi \cap  K determine a
composition
series
K ? K1 ? K2 ? \dots  ? Kr = {e}
of K: this is not difficult to see, and will be verified more formally in the
proof
of Proposition 3.4. By Proposition II.8.11 (the ``second isomorphism theorem''),
G 1
 G1
 G1G\setminus  1
 G
 G\setminus  1 G
=
 \sim 
 =
 = and
 \sim  =
K
 G1 \cap  G1
 \setminus 
 G1
 \setminus 
 G1
 \setminus 
 K
 G1
are simple. Therefore, we have two new composition series for G:
G ? G1 ? K ? K1 ? \dots  ? {e}
G ? G\setminus  1 ? K ? K1 ? \dots  ? {e}
which only differ at the first step. These two series trivially have the same
length
and the same quotients (the first two quotients get switched from one series to
the
other).
Now I claim that the first of these two series has the same length and quotients
as the series (*). Indeed,
G1 ? K ? K1 ? K2 ? \dots  ? Kr = {e}
is a composition series for G1 : by the induction hypothesis, it must have the
same
length and quotients as the composition series
G1 ? G2 ? \dots  ? Gn = {e};
verifying my claim (and note that, in particular, r = n - 2).
By the same token, applying the induction hypothesis to the series (of length
n - 1)
G\setminus  1 ? K ? K1 ? K2 ? \dots  ? Kn-2 = {e},
shows that the second series has the same length and quotients as (**), and the
statement follows.
 ?

\subsection{Composition factors; Schreier's theorem}
Two normal series are equiv-

alent if they have the same length and the same quotients (up to order). The
Jordan-H\"older theorem shows that any two maximal finite series of a group are
equivalent. That is, the (isomorphism classes of the) quotients of a composition
series depend only on the group, not on the chosen series. These are the compo-
sition factors of the group. They form a multiset12 of simple groups: the `basic
constituents' of our loose comment back in \S{}2.
It is clear that two isomorphic groups must have the same composition factors.
Unfortunately, it is not possible to reconstruct a group from its composition
factors
alone (Exercise 3.4). One has to take into account the way the simple groups are
`glued' together; we will come back to this point in \S{}5.2.
The intuition that the composition factors of a group are its basic constituents
is reinforced by the following fact: if G is a group with a composition series,
then
the composition factors of every normal subgroup N of G are composition factors
of G and the remaining ones are the composition factors of the quotient G/N .
12 See \S{}I.2.2 for a reminder on multisets: they are sets of elements counted
with multiplicity.
For example, the composition factors of Z/4Z form the multiset consisting of two
copies of Z/2Z.
Example 3.3. Let G = Z/6Z = {[0], [1], [2], [3], [4], [5]}. Then
{[0], [1], [2], [3], [4], [5]} ? {[0], [3]} ? {[0]}
is a composition series for G; the quotients are Z/3Z, Z/2Z, respectively. The
(nor-
mal) subgroup N = {[0], [2], [4]} `turns off' the second factor: indeed,
intersecting
the series with N gives
{[0], [2], [4]} ? {[0]} = {[0]},
a series with composition factor Z/3Z. On the other hand, `mod-ing out by N '
turns off the first factor: keeping in mind [3] + N = [1] + N , etc., we find
{[0] + N, [1] + N } = {[0] + N, [1] + N } ? {[0] + N },
a series with lone composition factor Z/2Z.
 ?
This phenomenon holds in complete generality:
Proposition 3.4. Let G be a group, and let N be a normal subgroup of G. Then
G has a composition series if and only if both N and G/N have composition
series.
Further, if this is the case, then
(G) = (N ) + (G/N ),
and the composition factors of G consist of the collection of composition
factors
of N and of G/N .
Proof. If G/N has a composition series, the subgroups appearing in it correspond
to subgroups of G containing N , with isomorphic quotients, by Proposition
II.8.10
(the ``third isomorphism theorem''). Thus, if both G/N and N have composition
series, juxtaposing them produces a composition series for G, with the stated
con-
sequence on composition factors.
The converse is a little trickier. Assume that G has a composition series
G = G0 ? G1 ? G2 ? \dots  ? Gn = {e}
and that N is a normal subgroup of G. Intersecting the series with N gives a
sequence of subgroups of the latter:
N = G \cap  N \supseteq  G1 \cap  N \supseteq  \dots  \supseteq  {e} \cap  N =
{e}
such that Gi+1 \cap  N is normal in Gi \cap  N , for all i. I claim that this
becomes a
composition series for N once repetitions are eliminated. Indeed, this follows
once
we establish that
Gi \cap  N
Gi+1 \cap  N
is either trivial (so that Gi+1 \cap  N = Gi \cap  N , and the corresponding
inclusion may
be omitted) or isomorphic to Gi /Gi+1 (hence simple, and one of the composition
factors of G). To see this, consider the homomorphism
Gi
Gi \cap  N \setminus \to  Gi \setminus 
 :
Gi+1
the kernel is clearly Gi+1 \cap N ; therefore (by the first isomorphism theorem)
we have
an injective homomorphism
Gi \cap  N
 Gi
\setminus \to 
Gi+1 \cap  N
 Gi+1
identifying (Gi \cap  N )/(Gi+1 \cap  N ) with a subgroup of Gi /Gi+1 . Now,
this subgroup
is normal (because N is normal in G) and Gi /Gi+1 is simple; our claim follows.
As for G/N , obtain a sequence of subgroups from a composition series for G:
G
 G1 N
 G2 N
 {eG}N
\supseteq 
 \supseteq 
 \supseteq  \cdot \cdot \cdot  \supseteq 
 = {eG/N },
N
 N
 N
 N
such that (Gi+1 N )/N is normal in (Gi N )/N . As above, we have to check that
(GiN )/N
(Gi+1 N )/N
is either trivial or isomorphic to Gi/Gi+1 . By the third isomorphism theorem,
this
quotient is isomorphic to (Gi N )/(Gi+1N ). This time, consider the homomorphism
Gi N
G i \setminus \to  GiN \setminus 
 :
Gi+1 N
this is surjective (check!), and the subgroup Gi+1 of the source is sent to the
identity
element in the target; hence (by Theorem II.7.12) there is an onto homomorphism
Gi
 Gi N
\setminus 
 .
Gi+1
 Gi+1N
Since Gi/Gi+1 is simple, it follows that (GiN )/(Gi+1 N ) is either trivial or
isomor-
phic to it (Exercise 2.4), as needed.
Summarizing, we have shown that if G has a composition series and N is normal
in G, then both N and G/N have composition series. The first part of the
argument
yields the statement on lengths and composition factors, concluding the proof. ?
One nice consequence of the Jordan-H\"older theorem is the following observa-
tion. A series is a refinement of another series if all terms of the first
appear in the
second.
Proposition 3.5. Any two normal series of a finite group ending with {e} admit
equivalent refinements.
Proof. Refine the series to a composition series; then apply the Jordan-H\"older
theorem.
 ?
In fact, Schreier's theorem asserts that this holds for all groups (while the
argument given here only works for groups admitting a composition series, e.g.,
finite groups). Proving this in general is reasonably straightforward, from
judicious
applications of the second isomorphism theorem (cf. Exercise 3.7).

\subsection{The commutator subgroup, derived series, and solvability}
It has

been a while since we have encountered a universal object; here is one. For any
group G, consider the category whose objects are group homomorphisms \alpha  : G
\to  A
from G to a commutative group and whose morphisms \alpha  \to  \beta  are (as
the reader
should expect) commutative diagrams
G 4
 4\alpha 
 4 \beta 
4 4?
 \phi 
 ?
A
 / B
where \phi  is a homomorphism.
Does this category have an initial object? That is, given a group G, does there
exist a commutative group which is universal with respect to the property of
being
a homomorphic image of G?
Yes.
Such a group may well be thought of as the closest `commutative approximation'
of the given group G. To verify that this universal object exists, we introduce
the
following important notion. (The diligent reader has begun exploring this
territory
already, in Exercise II.7.11.)
Definition 3.6. Let G be a group. The commutator subgroup of G is the subgroup
generated by all elements
-1 -1
ghg hwith g, h \in  G.
 ?
The element ghg -1 h-1 is often denoted [g, h] and is called the commutator of g
and h. Thus, g, h commute with each other if and only if [g, h] = e.
In the same notational style, the commutator subgroup of G should be denoted
[G, G]; this is a bit heavy, and the common shorthand for it is G\setminus  ,
which offers
the possibility of iterating the notation. Thus, G\setminus \setminus  may be
used to denote the
commutator subgroup of the commutator subgroup of G, and G(i) denotes the i-
th iterate. I will adopt this notation in this subsection for convenience, but
not
elsewhere in this book (as I want to be able to `prime' any letter I wish, for
any
reason).
First we record the following trivial, but useful, remark:
Lemma 3.7. Let \phi  : G1 \to  G2 be a group homomorphism. Then \forall g, h \in
G 1 we
have
\phi ([g, h]) = [\phi (g), \phi (h)]
and \phi (G1 \setminus 
 ) \subseteq  G2 \setminus 
 .
This simple observation makes the key properties of the commutator subgroup
essentially immediate (cf. Exercise II.7.11):
Proposition 3.8. Let G\setminus  be the commutator subgroup of G. Then
\bullet  G\setminus  is normal in G;
\bullet  the quotient G/G\setminus  is commutative;
\bullet  if \alpha  : G \to  A is a homomorphism of G to a commutative group, then G\setminus  \subseteq 
ker \alpha ;
\bullet  the natural projection G \to  G/G\setminus  is universal in the sense explained above.
Proof. These are all easy consequences of Lemma 3.7:
---By Lemma 3.7, the commutator subgroup is characteristic, hence normal
(cf. Exercise 2.2).
---By Lemma 3.7, the commutator of any two cosets gG\setminus , hG\setminus  is
the coset of the
commutator [g, h]; hence it is the identity in G/G\setminus . As noted above,
this implies
that G/G\setminus  is commutative.
---Let \alpha  : G \to  A be a homomorphism to a commutative group. By Lemma
3.7,
\alpha (G\setminus  ) \subseteq  A\setminus  = {e}: that is, G\setminus  \subseteq  ker \alpha .
---The universality follows from the previous point and from the universal prop-
erty of quotients (Theorem II.7.12).
 ?
Taking successive commutators of a group produces a descending sequence of
subgroups,
G \supseteq  G\setminus  \supseteq  G\setminus \setminus  \supseteq  G\setminus
\setminus \setminus  \supseteq  \dots  ,
which is `normal' in the sense indicated in \S{}3.1.
Definition 3.9. Let G be a group. The derived series of G is the sequence of
subgroups
G \supseteq  G\setminus  \supseteq  G\setminus \setminus  \supseteq  G\setminus
\setminus \setminus  \supseteq  \dots  .
 ?
The derived series may or may not end with the identity of G. For example,
if G is commutative, then the sequence gets there right away:
G \supseteq  G\setminus  = {e};
however, if G is simple and noncommutative, then it gets stuck at the first
step:
G = G \setminus  = G \setminus \setminus  = \dots 
(indeed, G\setminus  is normal and = {e} as G is noncommutative; but then G
\setminus  = G since
G is simple).
Definition 3.10. A group is solvable if its derived series terminates with the
iden-
tity.
 ?
For example, abelian groups are solvable.
The importance of this notion will be most apparent in the relatively distant
future because of a brilliant application of Galois theory (\S{}VII.7.4). But we
can
already appreciate it in the way it relates to the material we just covered. A
normal
series is abelian, resp., cyclic, if all quotients are abelian, resp., cyclic13
.
Proposition 3.11. For a finite group G, the following are equivalent:
(i) All composition factors of G are cyclic.
(ii) G admits a cyclic series ending in {e}.
13Thus, a composition series should be called `simple'; to our knowledge, it is
not.
(iii) G admits an abelian series ending in {e}.
(iv) G is solvable.
Proof. (i) \implies  (ii) \implies  (iii) are trivial. (iii) \implies  (i) is
obtained by refining an
abelian series to a composition series (keeping in mind that the simple abelian
groups are cyclic p-groups).
(iv) \implies  (iii) is also trivial, since the derived series is abelian (by
the second
point in Proposition 3.8).
Thus, we only have to prove (iii) \implies  (iv). For this, let
G = G0 ? G1 ? G2 ? \dots  ? Gn = {e}
be an abelian series. Then I claim that G(i) \subseteq  Gi for all i, where G(i)
denotes the
i-th `iterated' commutator subgroup.
This can be verified by induction. For i = 1, G/G 1 is commutative; thus
G\setminus  \subseteq  G1, by the third point in Proposition 3.8. Assuming we
know G(i) \subseteq  Gi , the
fact that Gi /G i+1 is abelian implies G\setminus  i \subseteq  G i+1, and hence
G(i+1) = (G(i) )\setminus  \subseteq  G\setminus  i \subseteq  Gi+1 ,
as claimed.
In particular we obtain that G(n) \subseteq  Gn = {e}: that is, the derived
series
terminates at {e}, as needed.
 ?
Example 3.12. All p-groups are solvable. Indeed, the composition factors of a
p-group are simple p-groups (what else could they be?), hence cyclic.
 ?
Corollary 3.13. Let N be a normal subgroup of a group G. Then G is solvable if
and only if both N and G/N are solvable.
Proof. This follows immediately from Proposition 3.4 and the formulation of
solv-
ability in terms of composition factors given in Proposition 3.11.
 ?
It is worth mentioning that any subgroup H of a solvable group is solvable:
indeed, the commutator H \setminus  of H is a subgroup of the commutator
G\setminus  of G, hence
H \setminus \setminus  \subseteq  G\setminus \setminus  , H \setminus \setminus
\setminus  \subseteq  G\setminus \setminus \setminus , and so on.
The Feit-Thompson theorem asserts that every finite group of odd order is
solvable. This result is many orders of magnitude beyond the scope of this book:
the original 1963 proof runs about 250 pages.

\subsection*{Exercises}

3.1. Prove that Z has normal series of arbitrary lengths. (Thus, (Z) is not
finite.)
3.2. Let G be a finite cyclic group. Compute (G) in terms of |G|. Generalize to
finite solvable groups.
3.3. ? Prove that every finite group has a composition series. Prove that Z does
not have a composition series. [\S{}3.1]
3.4. ? Find an example of two nonisomorphic groups with the same composition
factors. [\S{}3.2]
3.5. ? Show that if H, K are normal subgroups of a group G, then HK is a normal
subgroup of G. [\S{}3.1]
3.6. Prove that G1 \times  G2 has a composition series if and only if both G1
and G2
do, and explain how the corresponding composition factors are related.
3.7. ? Locate and understand a proof of (the general form of) Schreier's theorem
that does not use the Jordan-H\"older theorem. Then obtain an alternative proof
of
the Jordan-H\"older theorem, using Schreier's. [\S{}3.2]
3.8. ? Prove Lemma 3.7. [\S{}3.3]
3.9. Let G be a nontrivial p-group. Construct explicitly an abelian series for
G,
using the fact that the center of a nontrivial p-group is nontrivial (Corollary
1.9).
This gives an alternative proof of the fact that p-groups are solvable (Example
3.12).
3.10. \lnot  Let G be a group. Define inductively an increasing sequence Z0 =
{e} \subseteq
Z1 \subseteq  Z2 \subseteq  \dots  of subgroups of G as follows: for i \geq  1,
Zi is the subgroup of G
corresponding (as in Proposition II.8.9) to the center of G/Zi-1 .
\bullet  Prove that each Zi is normal in G, so that this definition makes sense.
A group is14 nilpotent if Zm = G for some m.
\bullet  Prove that G is nilpotent if and only if G/Z(G) is nilpotent.
\bullet  Prove that p-groups are nilpotent.
\bullet  Prove that nilpotent groups are solvable.
\bullet  Find a solvable group that is not nilpotent.
[3.11, 3.12, 5.1]
3.11. \lnot  Let H be a nontrivial normal subgroup of a nilpotent group G (cf.
Ex-
ercise 3.10). Prove that H intersects Z(G) nontrivially. (Hint: Let r \geq  1 be
the
smallest index such that \exists h = e, h \in  H \cap  Zr . Contemplate a
well-chosen commuta-
tor [g, h].) Since p-groups are nilpotent, this strengthens the result of
Exercise 1.9.
[3.14]
14 There are many alternative characterizations for this notion that are
equivalent to the one
given here but not too trivially so.
3.12. Let H be a proper subgroup of a finite nilpotent group G (cf. Exercise
3.10).
Prove that H ? NG (H). (Hint: Z(G) is nontrivial. First dispose of the case in
which H does not contain Z(G), and then use induction to deal with the case
in which H does contain Z(G).) Deduce that every Sylow subgroup of a finite
nilpotent group is normal15 . (Use Exercise 2.13.)
3.13. \lnot  For a group G, let G (i) denote the iterated commutator, as in
\S{}3.3. Prove
that each G(i) is characteristic (hence normal) in G. [3.14]
3.14. Let H be a nontrivial normal subgroup of a solvable group G.
\bullet  Prove that H contains a nontrivial commutative subgroup that is normal in G.
(Hint: Let r be the largest index such that K = H \cap  G(r) is nontrivial.
Prove
that K is commutative, and use Exercise 3.13 to show it is normal in G.)
\bullet  Find an example showing that H need not intersect the center of G nontrivially
(cf. Exercise 3.11).
3.15. Let p, q be prime integers, and let G be a group of order p2 q. Prove that
G
is solvable. (This is a particular case of Burnside's theorem: for p, q primes,
every
group of order paqb is solvable.)
3.16. ? Prove that every group of order < 120 and = 60 is solvable. [\S{}4.4,
\S{}VII.7.4]
3.17. Prove that the Feit-Thompson theorem is equivalent to the assertion that
every noncommutative finite simple group has even order.

\section{The symmetric group}


\subsection{Cycle notation. It is time to give a second look at symmetric groups}
Recall

that Sn denotes the group of permutations (i.e., automorphisms in Set) of the
set
{1, . . . , n}. In \S{}II.2 we denoted elements of Sn in a straightforward but
inconvenient
way:
1 2 3 4 5 6 7 8
\sigma  =
8 1 2 7 5 3 4 6
would stand for the element in S8 sending 1 to 8, 2 to 1, etc.
There is clearly ``too much'' information here (the first row should be
implicit),
and at the same time it seems hard to find out anything interesting about a per-
mutation from this notation. For example, can the reader say anything about the
conjugates of \sigma  in S8 ? For maximal enlightenment, try to do Exercise 4.1
now, and
then try again after absorbing the material in \S{}4.2. In short, we should be
able to
do better.
As often is the case, thinking in terms of actions helps. By its very
definition,
the group Sn acts on the set {1, . . . , n}; so does every subgroup of Sn .
Given a
permutation \sigma  \in  Sn , consider the cyclic group ?\sigma ? generated by
\sigma  and its action on
{1, . . . , n}. The orbits of this action form a partition of {1, . . . , n};
therefore, every
15This property characterizes finite nilpotent groups; cf. Exercise 5.1.
\sigma  \in  Sn determines a partition of {1, . . . , n}. For example, the element \sigma  \in  S8 given
above splits {1, . . . , 8} into three orbits:
{1, 2, 3, 6, 8},
 {4, 7},
 {5}.
The action of ?\sigma ? is transitive on each orbit. This means that one can get
from
any element of the orbit to any other element and then back to the original one
by
applying \sigma  enough times. In the example,
1 ?\to  8 ?\to  6 ?\to  3 ?\to  2 ?\to  1,
 4 ?\to  7 ?\to  4,
 5 ?\to  5.
Definition 4.1. A (nontrivial) cycle is an element of Sn with exactly one
nontrivial
orbit. For distinct a1 , . . . , ar in {1, . . . , n}, the notation
(a1 a2 . . . ar )
denotes the cycle in Sn with nontrivial orbit {a1, . . . , ar }, acting as
a1 ?\to  a 2 ?\to  \dots  ?\to  ar ?\to  a1.
In this case, r is the length of the cycle. A cycle of length r is called an
r-cycle. ?
The identity is considered a cycle of length 1 in a trivial way and is denoted
by
(1) (and could just as well be denoted by (i) for any i).
Note that (a1 a2 . . . ar ) = (a2 . . . ar a1 ) according to the notation
introduced
in Definition 4.1: the notation determines the cycle, but a nontrivial cycle
only
determines the notation `up to a cyclic permutation'.
Two cycles are disjoint if their nontrivial orbits are. The following
observation
deserves to be highlighted, but it does not seem to deserve a proof:
Lemma 4.2. Disjoint cycles commute.
The next one gives us the alternative notation we were looking for.
Lemma 4.3. Every \sigma  \in  Sn , \sigma  = e, can be written as a product of
disjoint nontrivial
cycles, in a unique way up to permutations of the factors.
Proof. As we have seen, every \sigma  \in  Sn determines a partition of {1, . .
. , n} into
orbits under the action of ?\sigma ?. If \sigma  = e, then ?\sigma ? has
nontrivial orbits. As \sigma  acts
as a cycle on each orbit, it follows that \sigma  may be written as a product of
cycles.
The proof of the uniqueness is left to the reader (Exercise 4.2).
 ?
The cycle notation for \sigma  \in  Sn is the (essentially) unique expression of
\sigma  as a
product of disjoint cycles found in Lemma 4.3 (or (1) for \sigma  = e). In our
running
example,
\sigma  = (18632)(47),
and keep in mind that this expression is unique cum grano salis: we could write
\sigma  = (63218)(47) = (74)(21863) = (32186)(74) = \dots  ,
and all of these are `the same' cycle notation for \sigma .

\subsection{Type and conjugacy classes in Sn}
The cycle notation has obviously an-

noying features---such as the not-too-unique uniqueness pointed out a moment
ago,
or the fact that
(123)
can be an element of S3 just as well as of S100000 , and only the context can
tell.
However, it is invaluable as it gives easy access to quite a bit of important
informa-
tion on a given permutation. In fact, much of this information is carried
already
by something even simpler than the cycle decomposition.
A partition of an integer n > 0 is a nonincreasing16 sequence of positive
integers
whose sum is n. It is easy to enumerate partitions for small values of n. For
example,
5 has 7 distinct partitions:
5=1+1+1+1+1
=2+1+1+1
=2+2+1
=3+1+1
=3+2
=4+1
= 5.
The partition \lambda 1 \geq  \lambda 2 \geq  \dots  \geq  \lambda r may be
denoted
[\lambda 1 , . . . , \lambda r ];
for example, the fourth partition listed above would be denoted [3, 1, 1]. A
nicer
`visual' representation is by means of the corresponding Young (or Ferrers)
diagram,
obtained by stacking \lambda 1 boxes on top of \lambda 2 boxes on top of \lambda
3 boxes on top of. . . .
For example, the diagrams corresponding to the seven partitions listed above are
Definition 4.4. The type of \sigma  \in  Sn is the partition of n given by the
sizes of the
orbits of the action of ?\sigma ? on {1, . . . , n}.
 ?
It is hopefully clear (from the argument proving Lemma 4.3) that the type
of \sigma  \in  Sn is simply given by the lengths of the cycles in the
decomposition of \sigma  as
the product of disjoint cycles, together with as many 1's as needed. In our
running
example,
\sigma  = (18632)(47) \in  S8
has type [5, 2, 1]:
16Of course this choice is arbitrary, and nondecreasing sequences would do just
as well.
The main reason why `types' are introduced is a consequence of the following
simple observation.
Lemma 4.5. Let \tau  \in  Sn, and let (a1 . . . ar ) be a cycle. Then
-1 -1 -1\tau  (a1 . . . ar )\tau  = (a1 \tau  . . . ar \tau  ).
-1 -1The funny notation a1 \tau  stands for the action of the permutation \tau
on
a1 \in  {1, . . . , n}; recall that we agreed in \S{}II.2.1 that we would let
our permutations
act on the right, for consistency with the usual notation for products in
groups.
Proof. This is verified by checking that both sides act in the same way on {1, .
. . , n}.
For example, for 1 \leq  i < r
-1 -1 -1 -1
(ai\tau  )(\tau  (a1 . . . ar )\tau  ) = ai (a1 . . . ar )\tau  = ai+1 \tau as
it should; the other cases are left to the reader.
 ?
-1By the usual trick of judiciously inserting identity factors \tau  \tau  ,
this formula
for computing conjugates extends immediately to any product of cycles:
\tau  (a1 . . . ar ) \dots  (b1 . . . bs )\tau  -1 = (a1\tau  -1 . . . ar \tau  -1 ) \dots  (b1 \tau  -1 . . . bs \tau  -1 ).
This holds whether the cycles are disjoint or not. However, since \tau  is a
bijection,
disjoint cycles remain disjoint after conjugation. This is essentially all there
is to
the following important observation:
Proposition 4.6. Two elements of Sn are conjugate in Sn if and only if they have
the same type.
Proof. The `only if' part of this statement follows immediately from the
preceding
considerations: conjugating a permutation yields a permutation of the same type.
As for the `if' part, suppose
\sigma 1 = (a1 . . . ar )(b1 . . . bs) \dots  (c1 . . . ct )
and
\sigma 2 = (a\setminus  1 . . . a\setminus  r )(b\setminus  1 . . . b\setminus  s) \dots  (c\setminus  1 . . . c\setminus  t )
are two permutations with the same type, written in cycle notation, with r \geq
s \geq
\dots  \geq  t (so the type is [r, s, . . . , t]). Let \tau  be any permutation such that ai = a\setminus  i \tau  ,
bj = b\setminus  j \tau  , . . . , ck = c\setminus  k \tau  for all i, j, . . .
, k. Then Lemma 4.5 implies \sigma 2 = \tau  \sigma 1 \tau  -1 ,
so \sigma 1 and \sigma 2 are conjugate, as needed.
 ?
Example 4.7. In S8 ,
(18632)(47) and (12345)(67)
must be conjugate, since they have the same type. The proof of Proposition 4.6
tells us that
-1\tau  (18632)(47)\tau  = (12345)(67)
for
1 2 3 4 5 6 7 8
\tau  =
 ,
1 8 6 3 2 4 7 5
and of course this may be checked by hand in a second. Running this check, and
especially staring at the second row in \tau  vis-\`{a}-vis the cycle notation
of the first
permutation, should clarify everything.
 ?
Summarizing, the type (or the corresponding Young diagram) tells us every-
thing about conjugation in Sn .
Corollary 4.8. The number of conjugacy classes in Sn equals the number of par-
titions of n.
For example, there are 7 conjugacy classes in S5 , indexed by the Tetris look-
alikes drawn above.
It is also reasonably straightforward to compute the number of elements in each
conjugacy class, in terms of the type. For example, in order to count the number
of permutations of type [2, 2, 1] in S5 , note that there are 5! = 120 ways to
fill the
corresponding Young diagram with the numbers 1, . . . , 5:
a1a3a5
a2
a4
that is, 120 ways to write a permutation as a product of two 2-cycles:
(a1a2 )(a3 a4 );
but switching a1 \leftrightarrow  a2 and a3 \leftrightarrow  a4 , as well as
switching the two cycles, gives the
same permutation. Therefore there are
= 15
2 \cdot  2 \cdot  2
permutations of type [2, 2, 1]. Performing this computation for all Young
diagrams
for S5 gives us the size of each conjugacy class, that is, the class formula
(cf. \S{}1.2)
for S5 :
120 = 1 + 10 + 15 + 20 + 20 + 30 + 24.
Example 4.9. There are no normal subgroups of size 30 in S5 .
Indeed, normal subgroups are unions of conjugacy classes (\S{}1.3); since the
identity is in every subgroup and 30 - 1 = 29 cannot be written as a sum of the
numbers appearing in the class formula for S5 , there is no such subgroup.
 ?
This observation will be dwarfed by much stronger results that we will prove
soon (such as Theorem 4.20, Corollary 4.21); but it is remarkable that such
precise
statements can be established with so little work.

\subsection{Transpositions, parity, and the alternating group}
For n \geq  1, consider

the polynomials
 ?
\Delta n =
 (xi - xj ) \in  Z[x1 , . . . , xn ],
1\leq i<j\leq n
that is,
\Delta 1 = 1,
\Delta 2 = x1 - x2 ,
\Delta 3 = (x 1 - x2 )(x1 - x3 )(x2 - x3 ),
\Delta 4 = (x 1 - x2 )(x1 - x3 )(x1 - x4 )(x2 - x3)(x2 - x4 )(x3 - x4 ),
\cdot \cdot \cdot 
We can act with any \sigma  \in  Sn on \Delta n , by permuting the indices
according to \sigma :
?
\Delta n \sigma  :=
 (xi\sigma  - xj\sigma  ).
1\leq i<j\leq n
For example,
\Delta 4 (1234) = (x2 - x3 )(x2 - x4 )(x2 - x1 )(x3 - x4 )(x3 - x1 )(x4 - x1 ) = -\Delta 4 .
In general, it is clear that \Delta n \sigma  is still the product of all
binomials (xi - xj ), where
the factors are permuted and some factors may change sign in the process. Hence,
\Delta n\sigma  = \pm \Delta n , where the factor \pm 1 depends on \sigma .
Definition 4.10. The sign of a permutation \sigma  \in  Sn , denoted (-1)\sigma
, is determined
by the action of \sigma  on \Delta n :
\Delta n \sigma  = (-1)\sigma  \Delta n .
We say that a permutation is even if its sign is +1 and odd if its sign is -1.
 ?
Note that \forall \sigma , \tau  \in  Sn we have
\Delta n (\sigma \tau  ) = (\Delta n \sigma )\tau  :
it follows that (-1)\sigma \tau  = (-1)\sigma  (-1)\tau  . Viewing {-1, +1} as a
group under multi-
plication17 , we see that the `sign' function
? : Sn \to  {-1, +1},
 ?(\sigma ) := (-1)\sigma 
is a homomorphism.
Here is a different viewpoint on this sign function. A transposition is a cycle
of
length 2. Every permutation is a product of transpositions:
Lemma 4.11. Transpositions generate Sn .
Proof. Indeed, by Lemma 4.3 it suffices to show that every cycle is a product of
transpositions, and indeed
(a1 . . . ar ) = (a1 a2 )(a1 a3 ) \dots  (a1 ar ),
as may be checked by applying18 both sides to every element of {1, . . . , n}.
 ?
17 This is the group of units in Z; of course it is isomorphic to C2
.
18Don't forget that permutations act on the right.
Of course a given permutation may be written as a product of transpositions
in many different ways. However, whether an even number of transpositions or an
odd number is needed only depends on \sigma . Indeed,
Lemma 4.12. Let \sigma  = \tau 1 \dots  \tau r be a product of transpositions.
Then \sigma  is even,
resp., odd, according to whether r is even, resp., odd.
Proof. This follows immediately from the facts that ? is a homomorphism and the
sign of a transposition is -1: indeed, (ij) acts on \Delta n by permuting its
factors and
changing the sign of an odd number of factors (for i < j, the factor (xi - xj )
and
the pairs of factors (xi - xk ), (xk - xj ) for all i < k < j).
 ?
Definition 4.13. The alternating group on {1, . . . , n}, denoted An , consists
of all
even permutations \sigma  \in  Sn.
 ?
The alternating group is a normal subgroup of Sn , and
[Sn : An ] = 2
for n \geq  2: indeed, An = ker ?, and ? is surjective for n \geq  2.
It is very easy to tell whether a permutation \sigma  belongs to An in terms of
the
type of \sigma . Indeed (by the argument proving Lemma 4.11) a cycle is even,
resp., odd,
if it has odd, resp., even, length19. Pictorially, a permutation \sigma  \in  Sn
is even if and
only if n and the number of rows in the Young diagram of \sigma  have the same
parity.
Take n = 5 for example:
The starred types correspond to even permutations. Adding up the sizes of the
corresponding conjugacy classes,
1 + 15 + 20 + 24 = 60,
confirms that A5 has index 2 in S5 .
However, do not read too much into this computation: I am not claiming that
these are the sizes of the conjugacy classes in A5 , only that these are the
conjugacy
classes in S5 making up the normal subgroup A5 . Conjugacy in An is rather
interesting and ties in nicely with the issue of the simplicity of An.

\subsection{Conjugacy in An ; simplicity of An and solvability of Sn}
Denote by

[\sigma ]Sn , resp., [\sigma ]An , the conjugacy class of an even permutation
\sigma  in Sn , resp., An .
Clearly [\sigma ]An \subseteq  [\sigma ]Sn ; we proceed to compare these two
sets.
Lemma 4.14. Let n \geq  2, and let \sigma  \in  A n. Then [\sigma ]An = [\sigma
]Sn or the size of [\sigma ]An
is half the size of [\sigma ]Sn , according to whether the centralizer ZSn
(\sigma ) is not or is
contained in An .
19Note the unfortunate terminology clash. For example, 3-cycles are even
permutations.
Proof. (Cf. Exercise 1.16.) Note that
ZAn (\sigma ) = An \cap  ZSn (\sigma ) :
this follows immediately from the definition of centralizer (Definition 1.6).
Now
recall that the centralizer of \sigma  is its stabilizer under conjugation, and
therefore the
size of the conjugacy class of \sigma  equals the index of its centralizer.
If ZSn (\sigma ) \subseteq  An , then ZAn (\sigma ) = ZSn (\sigma ), so that
[Sn : ZSn (\sigma )] = [Sn : ZAn (\sigma )] = [Sn : An ][An : ZAn (\sigma )] = 2
\cdot  [An : ZAn (\sigma )];
therefore, [\sigma ]An is half the size of [\sigma ]Sn in this case.
If ZSn (\sigma ) \subseteq  An , then note that An ZSn (\sigma ) = Sn : indeed,
An ZSn (\sigma ) is a sub-
group of Sn (because An is normal; cf. Proposition II.8.11), and it properly
con-
tains An, so it must equal Sn as An has index 2 in Sn. By index considerations
(cf. Exercise II.8.21)
[An : ZAn (\sigma )] = [An : An \cap  ZSn (\sigma )] = [A nZSn (\sigma ) : ZSn
(\sigma )] = [Sn : ZSn (\sigma )],
so the classes have the same size. Since [\sigma ]An \subseteq  [\sigma ]Sn in
any case, it follows that
[\sigma ]An = [\sigma ]Sn , completing the proof.
 ?
Therefore, conjugacy classes of even permutations either are preserved from Sn
to An or they split into two distinct, equal-sized classes. We are now in a
position
to give precise conditions determining which happens when.
Proposition 4.15. Let \sigma  \in  An , n \geq  2. Then the conjugacy class of
\sigma  in Sn splits
into two conjugacy classes in An precisely if the type of \sigma  consists of
distinct odd
numbers.
Proof. By Lemma 4.14, we have to verify that ZSn (\sigma ) is contained in An
precisely
when the stated condition is satisfied; that is, we have to show that
-1\sigma  = \tau  \sigma \tau  \implies  \tau  is even
precisely when the type of \sigma  consists of distinct odd numbers.
Write \sigma  in cycle notation (including cycles of length 1):
\sigma  = (a1 . . . a \lambda  )(b1 . . . b\mu  ) \dots  (c1 . . . c\nu  ),
and recall (Lemma 4.5) that
\tau  \sigma \tau  -1 = (a1 \tau  -1 . . . a\lambda  \tau  -1)(b1\tau  -1 . . . b\mu  \tau  -1 ) \dots  (c1 \tau  -1 . . . c\nu  \tau  -1 ).
-1Assume that \lambda , \mu  . . . , \nu  are odd and distinct. If \tau  \sigma
\tau  = \sigma , then conjugation
by \tau  must preserve each cycle in \sigma , as all cycle lengths are distinct:
-1\tau  (a1 . . . a\lambda  )\tau  = (a1 . . . a\lambda ),
 etc.,
that is,
-1 -1(a1\tau  . . . a\lambda  \tau  ) = (a1 . . . a\lambda ),
 etc.
This means that \tau  acts as a cyclic permutation on (e.g.) a1, . . . ,
a\lambda  and therefore
in the same way as a power of (a1 . . . a\lambda  ). It follows that
\tau  = (a1 . . . a\lambda  )r (b1 . . . b\mu  ) s \dots  (c1 . . . c\nu  )t
for suitable r, s, . . . , t. Since all cycles have odd lengths, each cycle is
an even
permutation; and \tau  must then be even as it is a product of even
permutations.
This proves that ZSn (\sigma ) \subseteq  An if the stated condition holds.
Conversely, assume that the stated condition does not hold: that is, either some
of the cycles in the cycle decomposition have even length or all have odd length
but
two of the cycles have the same length.
In the first case, let \tau  be an even-length cycle in the cycle decomposition
of \sigma .
-1Note that \tau  \sigma \tau  = \sigma : indeed, \tau  commutes with itself and
with all cycles in \sigma  other
than \tau  . Since \tau  has even length, then it is odd as a permutation: this
shows that
ZSn (\sigma ) \subseteq  An , as needed.
In the second case, without loss of generality assume \lambda  = \mu , and
consider the
odd permutation
\tau  = (a1 b1)(a2 b2 ) \dots  (a\lambda  b\lambda  ) :
-1conjugating by \tau  simply interchanges the first two cycles in \sigma ;
hence \tau  \sigma \tau  = \sigma .
As \tau  is odd, this again shows that Z S n (\sigma ) \subseteq  An , and we
are done.
 ?
Example 4.16. Looking again at A5 , we have noted in \S{}4.3 that the types of
the even permutations in S5 are [1, 1, 1, 1, 1], [2, 2, 1], [3, 1, 1], and [5].
By Proposi-
tion 4.15 the conjugacy classes corresponding to the first three types are
preserved
in A5, while the last one splits.
Therefore there are exactly 5 conjugacy classes in A5 , and the class formula
for A5 is
60 = 1 + 15 + 20 + 12 + 12.
 ?
Finally! We can now complete a circle of thought begun in the first section of
this chapter. Along the way, the reader has hopefully checked that every simple
group of order < 60 is commutative (Exercise 2.24); and we now see why 60 is
special:
Corollary 4.17. The alternating group A5 is a simple noncommutative group of
order 60.
Proof. A normal subgroup of A5 is necessarily the union of conjugacy classes,
contains the identity, and has order equal to a divisor of 60 (by Lagrange's
theorem).
The divisors of 60 other than 1 and 60 are
2 , 3 , 4 , 5 , 6 , 10 , 12 , 15 , 20 , 30;
counting the elements other than the identity would give one of
1 , 2 , 3 , 4 , 5 , 9 , 11 , 14 , 19 , 29
as a sum of numbers = 1 from the class formula for A5. But this simply does not
happen.
 ?
The reader will check that A6 is simple, by the same method (Exercise 4.21).
It is in fact the case that all groups An , n \geq  5, are simple (and
noncommutative),
implying that S n is not solvable for n \geq  5, and these facts are rather
important in
view of applications to Galois theory (cf., e.g., Corollary VII.7.16). Note that
A2
is trivial, A3 \sim 
 = Z/3Z is simple and abelian, and A4 is not simple (Exercise 2.24).
The alternating group A5 is also called the icosahedral (rotation) group:
indeed,
it is the group of symmetries of an icosahedron obtained through rigid motions.
(Can the reader verify20 this fact?)
The simplicity of An for n \geq  5 may be established by studying 3-cycles.
First
of all, it is natural to wonder whether every even permutation may be written as
a
product of 3-cycles, and this is indeed so:
Lemma 4.18. The alternating group An is generated by 3-cycles.
Proof. Since every even permutation is a product of an even number of 2-cycles,
it suffices to show that every product of two 2-cycles may be written as product
of
3-cycles. Therefore, consider a product
(ab)(cd)
with a = b, c = d. If (ab) = (cd), then this product is the identity, and there
is
nothing to prove. If {a, b}, {c, d} have exactly one element in common, then we
may assume c = a and observe
(ab)(ad) = (abd).
If {a, b}, {c, d} are disjoint, then
(ab)(cd) = (abc)(adc),
and we are done.
 ?
Now we can capitalize on our study of conjugacy in An:
Claim 4.19. Let n \geq  5. If a normal subgroup of An contains a 3-cycle, then
it
contains all 3-cycles.
Proof. Normal subgroups are unions of conjugacy classes, so we just need to
verify
that 3-cycles form a conjugacy class in An , for n \geq  5. But they do in Sn ,
and the
type of a 3-cycle is [3, 1, 1, . . . ] for n \geq  5; hence the conjugacy class
does not split
in An , by Proposition 4.15.
 ?
The general statement now follows easily by tying up loose ends:
Theorem 4.20. The alternating group An is simple for n \geq  5.
Proof. We have already checked this for n = 5, and the reader has checked it for
n = 6. For n > 6, let N be a nontrivial normal subgroup of An ; we will show
that
necessarily N = An , by proving that N contains 3-cycles.
Let \tau  \in  N , \tau  = (1), and let \sigma  \in  An be a 3-cycle. Since the
center of An is
trivial (Exercise 4.14) and 3-cycles generate An, we may assume that \tau  and
\sigma  do
not commute, that is, the commutator
[\tau , \sigma ] = \tau  (\sigma \tau  -1\sigma -1 ) = (\tau  \sigma \tau
-1)\sigma -1
20 It is good practice to start with smaller examples: for instance, the
tetrahedral rotation
group is isomorphic to A4 .
is not the identity. This element is in N (as is evident from the first
expression,
as N is normal) and is a product of two 3-cycles (as is evident from the second
expression, since the conjugate of a 3-cycle is a 3-cycle).
Therefore, replacing \tau  by [\tau , \sigma ] if necessary, we may assume that
\tau  \in  N is
a nonidentity permutation acting on \leq  6 elements: that is, on a subset of a
set
T \subseteq  {1, . . . , n} with |T | = 6. Now we may view A6 as a subgroup of
An , by letting
it act on T . The subgroup N \cap  A6 of A6 is then normal (because N is normal)
and
nontrivial (because \tau  \in  N \cap  A6 and \tau  = (1)). Since A6 is simple
(Exercise 4.21),
this implies N \cap  A6 = A6 . In particular, N contains 3-cycles.
By Claim 4.19, this implies that N contains all 3-cycles. By Lemma 4.18, it
follows that N = An , as needed.
 ?
Corollary 4.21. For n \geq  5, the group Sn is not solvable.
Proof. Since An is simple, the sequence
Sn ? An ? {(1)}
is a composition series for Sn. It follows that the composition factors of Sn
are
Z/2Z and An . By Proposition 3.11, Sn is not solvable.
 ?
In particular, S5 is a nonsolvable group of order 120. This is in fact the
smallest
order of a nonsimple, nonsolvable group; cf. Exercise 3.16.

\subsection*{Exercises}

4.1. ? Compute the number of elements in the conjugacy class of
1 2 3 4 5 6
 7 8
8 1 2 7 5 3
 4 6
in S8 . [\S{}4.1]
4.2. ? Suppose
(a1 . . . ar )(b1 . . . bs ) \dots  (c1 . . . ct ) = (d1 . . . du )(e1 . . . ev
) \dots  (f1 . . . fw )
are two products of disjoint cycles. Prove that the factors agree up to order.
(Hint: The two corresponding partitions of {1, . . . , n} must agree.) [\S{}4.1]
4.3. Assume \sigma  has type [\lambda 1 , . . . , \lambda r ] and that the
\lambda i 's are pairwise relatively prime.
What is |\sigma |? What can you say about |\sigma |, without the additional
hypothesis on
the numbers \lambda i ?
4.4. Make sense of the `Taylor series' of the infinite product
\cdot 
 \cdot 
 \cdot 
 \cdot 
 \cdot  \cdot \cdot \cdot  .
(1 - x) (1 - x ) (1 - x ) (1 - x ) (1 - x5 )
Prove that the coefficient of xn in this series is the number of partitions of
n.
4.5. Find the class formula for Sn, n \leq  6.
4.6. Let N be a normal subgroup of S4 . Prove that |N | = 1, 4, 12, or 24.
4.7. ? Prove that Sn is generated by (12) and (12 . . . n).
(Hint: It is enough to get all transpositions. What is the conjugate of (12) by
(12 . . . n)?) [4.9, \S{}VII.7.5]
4.8. \lnot  For n > 1, prove that the subgroup H of Sn consisting of
permutations
fixing 1 is isomorphic to Sn-1. Prove that there are no proper subgroups of Sn
properly containing H. [VII.7.17]
4.9. By Exercise 4.7, S4 is generated by (12) and (1234). Prove that (13) and
(1234) generate a copy of D8 in S4 . Prove that every subgroup of S4 of order 8
is
conjugate to ?(13), (1234)?. Prove there are exactly 3 such subgroups. For all n
\geq  3
prove that Sn contains a copy of the dihedral group D2n , and find generators
for it.
4.10. \lnot  \bullet  Prove that there are exactly (n - 1)! n-cycles in Sn.
\bullet  More generally, find a formula for the size of the conjugacy class of a permu-
tation of given type in Sn . [4.11]
4.11. Let p be a prime integer. Compute the number of p-Sylow subgroups of Sp .
(Use Exercise 4.10.) Use this result and Sylow's third theorem to prove again
the
`only if' implication in Wilson's theorem (cf. Exercise II.4.16.)
4.12. ? A subgroup G of Sn is transitive if the induced action of G on {1, . . .
, n}
is transitive.
\bullet  Prove that if G \subseteq  Sn is transitive, then |G| is a multiple of n.
\bullet  List the transitive subgroups of S3 .
\bullet  Prove that the following subgroups of S4 are all transitive:
-- ?(1234)? \sim 
 = C4 and its conjugates,
-- ?(12)(34), (13)(24)? \sim 
 = C2 \times  C2 ,
-- ?(12)(34), (1234)? \sim 
 = D8 and its conjugates,
-- A4 , and S4 .
With a bit of stamina, you can prove that these are the only transitive sub-
groups of S4 .
[\S{}VII.7.5]
4.13. (If you know about determinants.) Prove that the sign of a permutation
\sigma ,
as defined in Definition 4.10, equals the determinant of the matrix M\sigma
defined in
Exercise II.2.1.
4.14. ? Prove that the center of A n is trivial for n \geq  4. [\S{}4.4]
4.15. Justify the `pictorial' recipe given in \S{}4.3 to decide whether a
permutation
is even.
4.16. The number of conjugacy classes in An , n \geq  2, is (allegedly)
1, 3, 4, 5, 7, 9, 14, 18, 24, 31, 43, . . . .
Check the first several numbers in this list by finding the class formulas for
the
corresponding alternating groups.
4.17. ? \bullet  Find the class formula for A4 .
\bullet  Use it to prove that A4 has no subgroup of order 6. [\S{}II.8.5]
4.18. For n \geq  5, let H be a proper subgroup of An . Prove that [A n : H]
\geq  n.
Prove that An does have a subgroup of index n for all n \geq  3.
4.19. Prove that for n \geq  5 there are no nontrivial actions of An on any set
S with
|S| < n. Construct21 a nontrivial action of A4 on a set S, |S| = 3. Is there a
nontrivial action of A4 on a set S with |S| = 2?
4.20. \lnot  Find all fifteen elements of order 2 in A5, and prove that A5 has
exactly
five 2-Sylow subgroups. [4.22]
4.21. ? Prove that A6 is simple, by using its class formula (as is done for A5
in the
proof of Corollary 4.17). [\S{}4.4]
4.22. \lnot  Verify that A5 is the only simple group of order 60, up to
isomorphism.
(Hint: By Exercise 2.25, a simple group G of order 60 contains a subgroup of
index 5.
Use this fact to construct a homomorphism G \to  S5, and prove that the image of
this homomorphism must be A5.) Note that A5 has exactly five 2-Sylow subgroups;
cf. Exercise 4.20. Thus, the other possibility contemplated in Exercise 2.25
does
not occur. [2.25]

\section{Products of groups}

We already know that products exist in Grp (see \S{}II.3.4); here we analyze
this notion
further and explore variations on the same theme, with an eye towards the
question
of determining the information needed to reconstruct a group from its
composition
factors.

\subsection{The direct product}
Recall from \S{}II.3.4 that the (direct) product of two

groups H, K is the group supported on the set H \times  K, with operation
defined com-
ponentwise. We have checked (Proposition II.3.4) that the direct product
satisfies
the universal property defining products in the category Grp.
There are situations in which the direct product of two subgroups N , H of a
group G may be realized as a subgroup of G. Recall (Proposition II.8.11) that if
one of the subgroups is normal, then the subset N H of G is in fact a subgroup
of G. The relation between N H and N \times  H depends on how N and H intersect
in
G, so we take a look at this intersection.
The `commutator' [A, B] of two subsets A, B of G (see \S{}3.3) is the subgroup
generated by all commutators [a, b] with a \in  A, b \in  B.
Lemma 5.1. Let N , H be normal subgroups of a group G. Then
[N, H] \subseteq  N \cap  H.
21 You can think algebraically if you want; if you prefer geometry, visualize
pairs of opposite
sides on a tetrahedron.
Proof. It suffices to verify this on generators; that is, it suffices to check
that
[n, h] = n(hn-1 h-1) = (nhn-1 )h-1 \in  N \cap  H
for all n \in  N , h \in  H. But the first expression and the normality of N
show that
[n, h] \in  N ; the second expression and the normality of H show that [n, h]
\in  H. ?
Corollary 5.2. Let N , H be normal subgroups of a group G. Assume N \cap H =
{e}.
Then N , H commute with each other:
(\forall n \in  N ) (\forall h \in  H)
 nh = hn.
Proof. By Lemma 5.1, [N, H] = {e} if N \cap H = {e}; the result follows
immediately.
?
In fact, under the same hypothesis more is true:
Proposition 5.3. Let N , H be normal subgroups of a group G, such that N \cap  H
=
{e}. Then N H \sim 
 = N \times  H.
Proof. Consider the function
\phi  : N \times  H \to  NH
defined by \phi (n, h) = nh. Under the stated hypothesis, \phi  is a group
homomorphism:
indeed
\phi ((n1, h1 ) \cdot  (n2 , h2)) = \phi ((n1 n2 , h1 h2 ))
= n1 n2 h1 h2
= n1 h1n2 h2
since N , H commute by Corollary 5.2
= \phi ((n1 , h1 )) \cdot  \phi ((n2 , h2 )).
The homomorphism \phi  is surjective by definition of N H. To verify it is
injective,
consider its kernel:
ker \phi  = {(n, h) \in  N \times  H | nh = e}.
If nh = e, then n \in  N and n = h-1 \in  H; thus n = e since N \cap  H = {e}.
Using the
same token for h, we conclude h = e; hence (n, h) = the identity in N \times  H,
proving
that \phi  is injective.
Thus \phi  is an isomorphism, as needed.
 ?
Remark 5.4. This result gives an alternative argument for the proof of Claim
2.16:
if |G| = pq, with p < q prime integers, and G contains normal subgroups H, K of
order p, q, respectively (as is the case if q \equiv  1 mod p, by Sylow), then H
\cap  K = {e}
necessarily, and then Proposition 5.3 shows HK = \sim 
 H \times  K. As |HK| = |G| = pq,
this proves G \sim 
 = H \times  K \sim 
 = Z/pZ \times  Z/qZ. Finally, (1, 1) has order pq in this group,
so G is cyclic, with the same conclusion we obtained in Claim 2.16.
 ?

\subsection{Exact sequences of groups; extension problem}
Of course, the hypoth-

esis that both subgroups N , H are normal is necessary for the result of
Proposi-
tion 5.3: for example, the permutations (123) and (12) generate subgroups N , H
of S3 meeting only at {e}, and N is normal in S3 , but S3 = N H is not isomor-
phic to the direct product of N and H. It is natural to examine this more
general
situation.
Let N , H be subgroups of a group G, with N normal (but with no a priori
assumptions on H) and such that N \cap  H = {e}; assume G = N H. We are after a
description of the structure of G in terms of the structure of N and H.
It is notationally convenient to use the language of exact sequences, introduced
for modules in \S{}III.7.1. A (short) exact sequence of groups is a sequence of
groups
and group homomorphisms
 / N
 \phi 
 / G
 \psi 
 / H
 / 1
where \psi  is surjective and \phi  identifies N with ker \psi . In other words
(by the first
isomorphism theorem), use \phi  to identify N with a subgroup of G; then the
sequence
is exact if N is normal in G and \psi  induces an isomorphism G/N \to  H.
The reader should pause a moment and check that if G, N , H are abelian, then
this notion matches precisely the notion of short exact sequence of abelian
groups
(i.e., Z-modules) from \S{}III.7.1; a notational difference is that here the
trivial group
is denoted22 `1' rather than `0'.
Of course there always is an exact sequence
 / N
 / N \times  H
 / H
 / 1 :
map n \in  N to (n, eH ) and (n, h) \in  N \times  H to h. However, keep in mind
that this
is a very special case: to reiterate the example mentioned above, there also is
an
exact sequence
 / C3
 / S3
 / C2
 / 1 ,
yet S3 \sim 
 = C3 \times  C 2 .
Definition 5.5. Let N , H be groups. A group G is an extension of H by N if
there is an exact sequence of groups
 / N
 / G
 / H
 / 1 .
 ?
The extension problem aims to describe all extensions of two given groups,
up to isomorphism. For example, there are two extensions of C2 by C3 : namely
C6 \sim 
 = C3 \times  C2 and S3 ; we will soon be able to verify that, up to
 isomorphism, there
are no other extensions.
The extension problem is the `second half' of the classification problem: the
first half consists of determining all simple groups, and the second half
consists of
figuring out how these can be put together to construct any group23. For
example,
22 This is not unreasonable, since groups are more often written
`multiplicatively' rather than
additively, so the identity element is more likely to be denoted 1 rather than
0.
23 As mentioned earlier, the first half has been settled, although the
complexity of the work
leading to its solution justifies some skepticism concerning the absolute
correctness of the proof.
The status of the second half is, as far as I know, (even) murkier.
if
G = G0 ? G1 ? G2 ? G3 ? G4 = {e}
is a composition series, with (simple) quotients Hi = Gi /G i+1, then G is an
ex-
tension of H0 by an extension of H1 by an extension of H2 by H3 : knowing the
composition factors of G and the extension process, it should in principle be
possible
to reconstruct G.
We are going to `solve' the extension problem in the particular case in which
H is also a subgroup of G, intersecting N at {e}.
Definition 5.6. An exact sequence of groups
 / N
 / G
 / H
/ 1
(or the corresponding extension) is said to split if H may be identified with a
subgroup of G, so that N \cap  H = {e}.
We encountered this terminology in \S{}III.7.2 for modules, thus for abelian
groups.
Note that the notion examined there appears to be more restrictive than Defini-
tion 5.6, since it requires G to be isomorphic to a direct product N \times  H.
This
apparent mismatch evaporates because of Proposition 5.3: in the abelian case,
every split extension (according to Definition 5.6) is in fact a direct product.
Of course split extensions are anyway very special, since quotients of a group G
are usually not isomorphic to subgroups of G, even in the abelian case (cf.
Exer-
cise 5.4).
Lemma 5.7. Let N be a normal subgroup of a group G, and let H be a subgroup
of G such that G = N H and N \cap  H = {e}. Then G is a split extension of H by
N .
Proof. We have to construct an exact sequence
 / N
 / G
 / H
 / 1;
we let N \to  G be the inclusion map, and we prove that G/N = \sim 
 H. For this,
consider the composition
\alpha  : H \setminus \to  G \setminus  G/N.
Then \alpha  is surjective: indeed, since G = N H, \forall g \in  G we have g =
nh for some
n \in  N and h \in  H, and then
-1gN = nhN = h(h nh)N = hN = \alpha (h).
Further, ker \alpha  = {h \in  H | hN = N } = N \cap  H = {e}; therefore \alpha
is also injective,
as needed.
 ?
To recap, if in the situation of Lemma 5.7 we also require that H be normal
in G, then G is necessarily isomorphic to the `trivial' extension N \times  H:
this is what
we have proved in Proposition 5.3. We are seeking to describe the extension
`even
if' H is not normal in G.

\subsection{Internal/semidirect products}
The attentive reader should have noticed

that the key to Proposition 5.3 is really Corollary 5.2: if both N and H are
normal
and N \cap H = {e}, then N and H commute with each other. This is what
ultimately
causes the extension N H to be trivial. Now, recall that as soon as N is normal,
then every subgroup H of G acts on N by conjugation: in fact (cf. Exercise 1.21)
conjugation determines a homomorphism
\gamma  : H \to  AutGrp (N ),
 h ?\to  \gamma h .
(Explicitly, for h \in  H the automorphism \gamma h : N \to  N acts by \gamma h
(n) := hnh-1 .)
The subgroups H and N commute precisely when \gamma  is trivial. Corollary 5.2
shows
that if N and H are both normal and N \cap  H = {e}, then \gamma  is indeed
trivial.
This is the crucial remark. The next several considerations may be summarized
as follows: if N is normal in G, H is a subgroup of G, N \cap  H = {e} and G = N
H,
then the extension G of H by N may be reconstructed from the conjugation action
\gamma  : H \to  AutGrp (N ). The reader is advised to stare at the following triviality, which
is the motivating observation for the general discussion:
(*)
 (\forall n1, n2 \in  N ), (\forall h1 , h2 \in  H)
 n1h1 n2 h2 = (n1 (h1 n2 h-1
 1 )) (h1h2 ).
This says that if we know the conjugation action of H on N , then we can recover
the operation in G from this information and from the operations in N and H.
Here is the general discussion. It is natural to abstract the situation and
begin
with any two groups N , H and an arbitrary homomorphism24
\theta  : H \to  AutGrp (N ),
 h ?\to  \theta h .
Define an operation \bullet \theta  on the set N \times  H as follows: for n1 ,
n2 \in  N and h1 , h2 \in  H,
let
(n1 , h1 ) \bullet \theta  (n2 , h2 ) := (n1 \theta h1 (n2 ), h1 h2).
This will look more reasonable once it is compared with (*)!
Lemma 5.8. The resulting structure (N \times  H, \bullet \theta  ) is a group,
with identity element
(eN , eH ).
Proof. The reader should carefully verify this. For example, inverses exist
because
-1
 -1
 -1
 -1
 -1
(n1, h1 )\bullet \theta  (\theta h-1 (n1 ), h1 ) = (n1\theta h1 (\theta h-1 (n1
)), h1h1 ) = (n1 n1 , eH ) = (eN , eH )
and similarly in the reverse order.
 ?
Definition 5.9. The group (N \times  H, \bullet \theta  ) is a semidirect
product of N and H and is
denoted by N ?\theta  H.
 ?
For example, the ordinary direct product is a semidirect product and corre-
sponds to \theta  = the trivial map. If the reader feels a little uneasy about
giving
one name (semidirect product) for a whole host of different groups supported on
the Cartesian product, welcome to the club. In fact, it gets worse still: it is
not
24 The reason why I am not denoting the image of h by \theta  as \theta (h) is
that this is an automor-
phism of N and I dislike the notation \theta (h)(n) for the image of n \in  N
obtained by applying the
automorphism corresponding to h. The alternative \theta h (n) looks a little
easier to parse.
uncommon to omit `\theta ' from the notation and simply write N ? H for a
semidirect
product25.
In any case, the notation \bullet \theta  is too heavy to carry around, so we
generally revert
back to the usual simple juxtaposition of elements in order to denote
multiplication
in N ?\theta  H. The following proposition checks that semidirect products are
split
extensions:
Proposition 5.10. Let N , H be groups, and let \theta  : H \to  AutGrp (N ) be a
homo-
morphism; let G = N ?\theta  H be the corresponding semidirect product. Then
\bullet  G contains isomorphic copies of N and H;
\bullet  the natural projection G \to  H is a surjective homomorphism, with kernel N ;
thus N is normal in G, and the sequence
 / N
 / N ?\theta  H
 / H
 / 1
is (split) exact;
\bullet  N \cap  H = {eG };
\bullet  G = N H;
\bullet  the homomorphism \theta  is realized by conjugation in G: that is, for h \in  H and
n \in  N we have
-1
\theta h (n) = hnhin G.
Proof. The functions N \to  G, H \to  G defined for n \in  N , h \in  H by
n ?\to  (n, eH ),
 h ?\to  (eN , h)
are manifestly injective homomorphisms, allowing us to identify N , H with the
corresponding subgroups of G. It is clear that N \cap  H = {(eN , eH )} = {eG },
and
(n, eH ) \bullet \theta  (eN , h) = (n, h)
shows that G = N H.
The projection G \to  H defined by
(n, h) ?\to  h
is a surjective homomorphism, with kernel N ; therefore N is normal in G.
Finally,
(eN , h) \bullet \theta  (n, eH ) \bullet \theta  (eN , h)-1 = (\theta h (n), h)
\bullet \theta  (eN , h-1 ) = (\theta h(n), eH ),
as claimed in the last point.
 ?
Our original goal of `reconstructing' a given split extension of a group H by a
group N is a sort of converse to this proposition. More precisely,
25 This is actually OK, if N and H are given as subgroups of a common group G,
in which
case the implicit action is just conjugation.
Proposition 5.11. Let N , H be subgroups of a group G, with N normal in G.
Assume that N \cap  H = {e}, and G = N H. Let \gamma  : H \to  AutGrp (N ) be
defined by
conjugation: for h \in  H, n \in  N ,
\gamma h (n) = hnh-1 .
Then G \sim 
 = N ?\gamma  H.
Proof. Define a function
\phi  : N ?\gamma  H \to  G
by \phi (n, h) = nh; this is clearly a bijection. We need to verify that \phi
is a homomor-
phism, and indeed (\forall n1 , n2 \in  N ), (\forall h1, h2 \in  H):
\phi ((n1 , h1 ) \bullet \gamma  (n2 , h2)) = \phi ((n1 \gamma h1 (n2 ), h1 h2 ))
-1
= \phi ((n1 (h1 n2 h1 ), h1h2 ))
-1
= n 1 h1n2 (h1 h1)h2 = (n1h1 )(n2 h2 )
= \phi ((n1 , h1 ))\phi ((n2 , h2 ))
as needed.
 ?
When realized `within' a group, as in the previous proposition, the semidirect
product is sometime called an internal product.
Remark 5.12. If N and H commute, then the conjugation action of H on N is
trivial; therefore \gamma  is the trivial map, and the semidirect product N
?\gamma  H is the
direct product N \times  H. Thus, Proposition 5.11 recovers the result of
Proposition 5.3
in this case.
 ?
Example 5.13. The automorphism group of C3 is isomorphic to the cyclic group C2
:
if C3 = {e, y, y 2 }, then the two automorphisms of C3 are
\{
 \{
|
 |
 e ?\to  e,
 |
 |
 e ?\to  e,
id :
 y ?\to  y,
 \sigma  :
 y ?\to  y 2 ,
|
 \} 2
 2 |
 \} 2
y ?\to  y ,
 y ?\to  y.
Therefore, there are two homomorphisms C2 \to  AutGrp (C3 ): the trivial map,
and
the isomorphism sending the identity to id and the nonidentity element to \sigma
. The
semidirect product corresponding to the trivial map is the direct product C3
\times C2 \sim
 =
C6 ; the other semidirect product C3 ? C 2 is isomorphic to S3 . This can of
course
be checked by hand (and you should do it, for fun); but it also follows
immediately
from Proposition 5.11, since N = ?(123)?, H = ?(12)? \subseteq  S3 satisfy the
hypotheses
of this result.
 ?
The reader should contemplate carefully the slightly more general case of dihe-
dral groups (Exercise 5.11); this enriches (and in a sense explains) the
discussion
presented in Claim 2.17.
In fact, semidirect products shed light on all groups of order pq, for p < q
primes; the reader should be able to complete the classification of these groups
begun in \S{}2.5.2 and show that if q \equiv  1 mod p, then there is exactly one
such non-
commutative group up to isomorphism (Exercise 5.12).

\subsection*{Exercises}

The reader would in fact be well-advised to try to use semidirect products to
classify groups of small order: if a nontrivial normal subgroup N is found
(typically
by applying Sylow's theorems), with some luck the classification is reduced to
the
study of possible homomorphisms from known groups to AutGrp (N ) and can be
carried out. See Exercise 5.15 for a further illustration of this technique.
5.1. ? Let G be a finite group, and let P1 , . . . , Pr be its nontrivial Sylow
subgroups.
Assume all Pi are normal in G.
\bullet  Prove that G = \sim 
 P1 \times  \dots  \times  Pr . (Induction on r; use Proposition 5.3.)
\bullet  Prove that G is nilpotent. (Hint: Mod out by the center, and work by induction
on |G|. What is the center of a direct product of groups?)
Together with Exercise 3.10, this shows that a finite group is nilpotent if and
only
if each of its Sylow subgroups is normal. [3.12, \S{}6.1]
5.2. Let G be an extension of H by N . Prove that the composition factors of G
are the collection of the composition factors of H and those of N .
5.3. Let
G = G0 ? G1 ? \dots  ? Gr = {e}
be a normal series. Show how to `connect' {e} to G by means of r exact sequences
of groups using the groups Gi and the quotients Hi = G i /Gi+1 .
5.4. ? Prove that the sequence
 / Z
 \cdot 2
 / Z
 / Z/2Z
 / 0
is exact but does not split. [\S{}5.2]
5.5. In Proposition III.7.5 we have seen that if an exact sequence
 / M
 \phi 
 / N
 / N/(\phi (M ))
 / 0
of abelian groups splits, then \phi  has a left-inverse. Is this necessarily the
case for
split sequences of groups?
5.6. Prove Lemma 5.8.
5.7. Let N be a group, and let \alpha  : N \to  N be an automorphism of N .
Prove
that \alpha  may be realized as conjugation, in the sense that there exists a
group G
containing N as a normal subgroup and such that \alpha (n) = gng-1 for some g
\in  G.
5.8. Prove that any semidirect product of two solvable groups is solvable. Show
that semidirect products of nilpotent groups need not be nilpotent.
5.9. ? Prove that if G = N ? H is commutative, then G = \sim 
 N \times  H. [\S{}6.1]
5.10. Let N be a normal subgroup of a finite group G, and assume that |N |
and |G/N | are relatively prime. Assume there is a subgroup H in G such that
|H| = |G/N |. Prove that G is a semidirect product of N and H.
5.11. ? For all n > 0 express D2n as a semidirect product Cn ?\theta  C2 ,
finding \theta
explicitly. [\S{}5.3]
5.12. ? Classify groups G of order pq, with p < q prime: show that if |G| = pq,
then either G is cyclic or q \equiv  1 mod p and there is exactly one
isomorphism class
of noncommutative groups of order pq in this case. (You will likely have to use
the
fact that AutGrp (Cq ) = \sim 
 Cq-1 if q is prime; cf. Exercise II.4.15.) [\S{}2.5, \S{}5.3]
5.13. \lnot  Let G = N ?\theta  H be a semidirect product, and let K be the
subgroup of G
corresponding to ker \theta  \subseteq  H. Prove that K is the kernel of the
action of G on the
set G/H of left-cosets of H. [5.14]
5.14. Recall that S3 \sim 
 = AutGrp (C2 \times  C2 ) (Exercise II.4.13). Let \iota  be this isomor-
phism. Prove that (C2 \times  C2 ) ?\iota  S3 \sim 
 = S4 . (Hint: Exercise 5.13.)
5.15. ? Let G be a group of order 28.
\bullet  Prove that G contains a normal subgroup N of order 7.
\bullet  Recall (or prove again) that, up to isomorphism, the only groups of order 4 are
C4 and C2 \times  C2 . Prove that there are two homomorphisms C4 \to  AutGrp (N
)
and two homomorphisms C2 \times  C2 \to  AutGrp (N ) up to the choice of
generators
for the sources.
\bullet  Conclude that there are four groups of order 28 up to isomorphism: the two
direct products C4 \times  C7 , C2 \times  C2 \times  C7 , and two
noncommutative groups.
\bullet  Prove that D28 \sim 
 = C2 \times  D14. The other noncommutative group of order 28 is a
generalized quaternionic group.
[\S{}5.3]
5.16. Prove that the quaternionic group Q8 (cf. Exercise III.1.12) cannot be
written
as a semidirect product of two nontrivial subgroups.
5.17. Prove that the multiplicative group H\ast  of nonzero quaternions (cf.
Exer-
cise III.1.12) is isomorphic to a semidirect product SU(2) ? R+ . (Hint: Exer-
cise III.2.5.) Is this semidirect product in fact direct?

\section{Finite abelian groups}

I will end this chapter by treating in some detail the classification theorem
for finite
abelian groups mentioned in \S{}II.6.3.

\subsection{Classification of finite abelian groups}
Now that we have acquired more

familiarity with products, we are in a position to classify all finite abelian
groups26 .
26Of course fancier semidirect products will not be needed here; cf. Exercise
5.9.
In due time (Proposition VI.2.11, Exercise VI.2.19) we will in fact be able to
classify all finitely generated abelian groups: as mentioned in Example II.6.3,
all
such groups are products of cyclic groups27. In particular, this is the case for
finite
abelian groups: this is what we prove in this section.
Since in this section we exclusively deal with abelian groups, we revert to the
abelian style of notations: thus the operation will be denoted +; the identity
will
be 0; direct products will be called direct sums (and denoted \oplus ); and so
on.
First of all, I will congeal into an explicit statement a simple observation
that
has been with us in one form or another since at least as far back as 28
Exercise II.4.9.
Lemma 6.1. Let G be an abelian group, and let H, K be subgroups such that |H|,
|K| are relatively prime. Then H + K \sim 
 = H \oplus  K.
Proof. By Lagrange's theorem (Corollary II.8.14), H \cap  K = {0}. Since
subgroups
of abelian groups are automatically normal, the statement follows from Proposi-
tion 5.3.
 ?
Now let G be a finite abelian group. For each prime p, the p-Sylow subgroup
of G is unique, since it is automatically normal in G. Since the distinct
nontrivial
Sylow subgroups of G are p-groups for different primes p, Lemma 6.1 immediately
implies the following result.
Corollary 6.2. Every finite abelian group is the direct sum of its nontrivial
Sylow
subgroups.
(The diligent reader knew already that this had to be the case, since abelian
groups
are nilpotent; cf. Exercise 5.1.) Thus, we already know that every finite
abelian
group is a direct sum of p-groups, and our main task amounts to classifying
abelian
p-groups for a fixed prime p. This is somewhat technical; we will get there by a
seemingly roundabout path.
Lemma 6.3. Let G be an abelian p-group, and let g \in  G be an element of
maximal
order. Then the exact sequence
 / ?g?
 / G
 / G/?g?
 / 0
splits.
Put otherwise, there is a subgroup L of G such that L maps isomorphically to
G/?g? via the canonical projection, that is, such that ?g?\cap L = {0} and ?g?+L
= G.
Note that it will follow that G \sim 
 = ?g? \oplus  L, by Proposition 5.3.
The main technicality needed in order to prove this lemma is the following
particular case:
Lemma 6.4. Let p be a prime integer and r \geq  1. Let G be a noncyclic abelian
group of order pr+1 , and let g \in  G be an element of order pr . Then there
exists an
element h \in  G, h \in  ?g?, such that |h| = p.
27 The natural context to prove this more general result is that of modules over
Euclidean
rings or even principal ideal domains.
28 In fact, this observation will really find its most appropriate resting place
when we prove
the Chinese Remainder Theorem, Theorem V.6.1.
Lemma 6.4 is a special case of Lemma 6.3 in the sense that, with notation as in
the statement, necessarily ?h? \sim 
 = G/?g?, and in fact G = \sim 
 ?g? \oplus  ?h? (and the reader
is warmly encouraged to understand this before proceeding!). That is, we can
split
off the `large' cyclic subgroup ?g? as a direct summand of G, provided that G is
not cyclic and not much larger than ?g?. Lemma 6.3 claims that this can be done
whenever ?g? is a maximal cyclic subgroup of G. We will be able to prove this
more
general statement easily once the particular case is settled.
Proof of Lemma 6.4. Denote ?g? by K, and let h\setminus  be any element of G,
h\setminus  \in  K.
The subgroup K is normal in G since G is abelian; the quotient group G/K has
order p. Since h\setminus  \in  K, the coset h\setminus  + K has order p in G/K;
that is, ph\setminus  \in  K. Let
k = ph\setminus  .
Note that |k| divides pr ; hence it is a power of p. Also |k| = p r , otherwise
|h\setminus 
| = pr+1 and G would be cyclic, contrary to the hypothesis.
Therefore |k| = ps for some s < r; k generates a subgroup ?k? of the cyclic
group K, of order ps . By Proposition II.6.11, ?k? = ?pr-s g?. Since s < r, ?k?
\subseteq
?pg?; thus, k = mpg for some m \in  Z.
Then let h = h\setminus  - mg: h = 0 (since h\setminus  \in  K), and
ph = ph\setminus  - p(mg) = k - k = 0,
showing that |h| = p, as stated.
 ?
Proof of Lemma 6.3. Argue by induction on the order of G; the case |G| = p0 =
1 requires no proof. Thus we will assume that G is nontrivial and that the
statement
is true for every p-group smaller than G.
Let g \in  G be an element of maximal order, say pr , and denote by K the
subgroup ?g? generated by g; this subgroup is normal, as G is abelian. If G = K,
then the statement holds trivially. If not, G/K is a nontrivial p-group, and
hence it
contains an element of order p by Cauchy's theorem (Theorem 2.1). This element
generates a subgroup of order p in G/K, corresponding to a subgroup G\setminus
of G of
order pr+1 , containing K. This subgroup is not cyclic (otherwise the order of g
is
not maximal).
That is, we are in the situation of Lemma 6.4: hence we can conclude that there
is an element h \in  G\setminus  (and hence h \in  G) with h \in  K and |h| = p.
Let H = ?h? \subseteq  G
be the subgroup generated by h, and note that K \cap  H = {0}.
Now work modulo H. The quotient group G/H has smaller size than G, and
g + H generates a cyclic subgroup K \setminus  = (K + H)/H = \sim 
 K/(K \cap  H) = \sim 
 K of maximal
order in G/H. By the induction hypothesis, there is a subgroup L\setminus  of
G/H such
that K \setminus  + L\setminus  = G/H and K \setminus  \cap  L\setminus  = {0G/H
}. This subgroup L\setminus  corresponds to a
subgroup L of G containing H.
Now I claim that (i) K + L = G and (ii) K \cap  L = {0}. Indeed, we have the
following:
(i) For any a \in  G, there exist mg + H \in  K \setminus  , + H \in  L\setminus
such that a + H =
mg + + H (since K \setminus  + L\setminus  = G/H). This implies a - mg \in  L,
and hence
a \in  K + L as needed.
(ii) If a \in  K \cap  L, then a + H \in  K \setminus  \cap  L\setminus  = {0G/H
}, and hence a \in  H. In particular,
a \in  K \cap  H = {0}, forcing a = 0, as needed.
(i) and (ii) imply the lemma, as observed in the comments following the
statement.
?
Now we are ready to state the classification theorem; the proof is quite
straight-
forward after all this preparation work. We first give the statement in a
somewhat
coarse form, as a corollary of the previous considerations:
Corollary 6.5. Let G be a finite abelian group. Then G is a direct sum of cyclic
groups, which may be assumed to be cyclic p-groups.
Proof. As noted in Corollary 6.2, G is a direct sum of p-groups (as a
consequence
of the Sylow theorems). I claim that every abelian p-group P is a direct sum of
cyclic p-groups.
To establish this, argue by induction on |P |. There is nothing to prove if P is
trivial. If P is not trivial, let g be an element of P of maximal order. By
Lemma 6.3
P = ?g? \oplus  P \setminus 
\setminus  \setminus for some subgroup P of P ; by the induction hypothesis P is a direct sum of cyclic
p-groups, concluding the proof.
 ?

\subsection{Invariant factors and elementary divisors}
Here is a more precise version

of the classification theorem. It is common to state the result in two
equivalent
forms.
Theorem 6.6. Let G be a finite nontrivial abelian group. Then
\bullet  there exist prime integers p1 , . . . , pr and positive integers nij such that |G| =
?
 ni,j
i,j pi
 and
! Z
G \sim 
 =
 n
ij ;
i,j
 p i Z
\bullet  there exist positive integers 1 < d1 | \dots  | ds such that |G| = d1 \dots  ds and
G \sim 
 =
 Z
 \oplus  \cdot \cdot \cdot  \oplus 
 Z
 .
d1 Z
 ds Z
Further, these decompositions are uniquely determined by G.
The first form is nothing but a more explicit version of the statement of Corol-
lary 6.5, so it has already been proven. I will explain how to obtain the second
form from the first. The uniqueness statement29 is left to the reader (Exercise
6.1).
The prime powers appearing in the first form of Theorem 6.6 are called the
elementary divisors of G; the integers di appearing in the second form are
called
invariant factors. To go from elementary divisors to invariant factors, collect
the
29 Of course the `uniqueness' statement only holds up to trivial manipulation
such as a per-
mutation of the factors. The claim is that the factors themselves are determined
by G, in the sense
that two direct sums of either form given in the statement are isomorphic only
if their factors
match.
elementary divisors in a table, listing (for example) prime powers according to
in-
creasing primes in the horizontal direction and decreasing exponents in the
vertical
direction; then the invariant factors are obtained as products of the factors in
each
row:
dr =
 pn 1 11
 pn 2 21
 pn 3 31
 \cdot \cdot \cdot 
dr-1 =
 pn 1 12
 pn 2 22
 pn 3 32
 \cdot \cdot \cdot 
dr-2 =
 pn 1 13
 pn 2 23
 pn 3 33
 \cdot \cdot \cdot 
\cdot \cdot \cdot 
 \cdot \cdot \cdot 
 \cdot \cdot \cdot 
 \cdot \cdot \cdot 
 \cdot \cdot \cdot 
Conversely, given the invariant factors di, obtain the rows of this table by
factoring di into prime powers: the condition d1 | \dots  | dr guarantees that
these will
be decreasing.
Repeated applications of Lemma 6.1 show that if d = pn 1 1 \dots  pn r r for
distinct
primes pi and positive ni (as is the case in each row of the table), then
Z Z
 Z
\sim  = n1 \oplus  \dots  \oplus  n r ,
dZ
 p1 Z
 pr Z
proving that the two decompositions given in Theorem 6.6 are indeed equivalent.
This will likely be much clearer once the reader works through a few examples.
Example 6.7. Here are the two decompositions for a (random) group of order
29160 = 23 \cdot  36 \cdot  5:
\setminus 
 ?
 \setminus 
 ?
Z
 \oplus 
 Z
 \oplus 
 Z
 \oplus 
 Z
 \oplus 
 Z
 \oplus  Z
 \oplus  Z
 \oplus 
 Z
 \sim 
 =
 Z
 \oplus 
 Z
 \oplus 
 Z
 \oplus 
 Z
2Z
 2Z
 2Z
 3Z
 3Z
 32 Z
 32 Z
 5Z
 3Z
 6Z
 18Z
 90Z
and here is the corresponding table of invariant factors/elementary divisors:
90 =
 2 32
18 =
 2 32
6=
3=
Example 6.8. There are exactly 6 isomorphism classes of abelian groups of or-
der 360. Indeed, 360 = 23 \cdot  32 \cdot  5; the six possible tables of
elementary divisors
are shown below. In terms of invariant factors, the six distinct abelian groups
of
order 360 (up to isomorphism, by the uniqueness part of Theorem 6.6) are
therefore
Z
 Z
 Z
 Z
 Z
 Z
,
 \oplus 
 ,
 \oplus 
 \oplus 
 ,
360Z
 2Z 180Z
 2Z 2Z 90Z
Z
 Z
 Z
 Z
 Z
 Z
 Z
\oplus 
 ,
 \oplus 
 ,
 \oplus 
 \oplus 
 .
3Z 120Z
 6Z 60Z
 2Z 6Z 30Z
360=
 180=
 90=
2=
 2=
2=
120=
 3 5
 60=
 3 5
 30=
 2 3 5
3=
 6=
 6=
 2 3
2=

\subsection{Application: Finite subgroups of multiplicative groups of fields}
Any

classification theorem is useful in that it potentially reduces the proof of
general
facts to explicit verifications. Here is one example illustrating this strategy:
Lemma 6.9. Let G be a finite abelian group, and assume that for every integer
n > 0 the number of elements g \in  G such that ng = 0 is at most n. Then G is
cyclic.
The reader should try to prove this `by hand', to appreciate the fact that
it is not entirely trivial. It does become essentially immediate once we take
the
classification of finite abelian groups into account. Indeed, by Theorem 6.6
Z
 Z
G \sim 
 =
 \oplus  \cdot \cdot \cdot  \oplus 
d1 Z
 ds Z
for some positive integers 1 < d1 | \dots  | ds . But if s > 1, then |G| > ds
and d s g = 0
for all g \in  G (so that the order of g divides ds ), contradicting the
hypothesis.
Therefore s = 1; that is, G is cyclic.
Lemma 6.9 is the key to a particularly nice proof of the following important
fact, a weak form of which30 we ran across back in Example II.4.6. Recall that
the
\ast set F of nonzero elements of a field F is a commutative group under multiplication.
Also recall (Example III.4.7) that a polynomial f (x) \in  F [x] is divisible by
(x - a)
if and only if f (a) = 0; since a nonzero polynomial of degree n over a field
can
have at most n linear factors, this shows31 that if f (x) \in  F [x] has degree
n, then
f (a) = 0 for at most n distinct elements a \in  F .
Theorem 6.10. Let F be a field, and let G be a finite subgroup of the
multiplicative
group (F \ast  , \cdot ). Then G is cyclic.
Proof. By the considerations preceding the statement, for every n there are at
most n elements a \in  F such that an - 1 = 0, that is, at most n elements a \in
G
such that an = 1. Lemma 6.9 implies then that G is cyclic.
 ?
30 The diligent reader has proved that particular case in Exercise II.4.11. The
proof hinted
at in that exercise upgrades easily to the general case presented here. The
point is not that the
classification theorem is necessary in order to prove statements such as Theorem
6.10; the point
is that it makes such statements nearly evident.
31 Unique factorization in F [x] is secretly needed here. We will deal with this
issue more
formally later; cf. Lemma V.5.1.
As a (very) particular case, the multiplicative group ((Z/pZ)\ast  , \cdot ) is
cyclic: this
is the fact pointed to in Example II.4.6.
Preview of coming attractions: Finitely generated (as opposed to just finite)
abelian groups are also direct sums of cyclic groups. The only difference
between
the classification of finitely generated abelian groups and the classification
of finite
abelian groups explored here is the possible presence of a `free' factor Z\oplus
r in the
decomposition. The reader will first prove this fact in Exercise VI.2.19, as a
conse-
quence of `Gaussian elimination over integral domains', and then recover it
again
as a particular case of the classification theorem for finitely generated
modules
over PIDs, Theorem VI.5.6. Neither Gaussian elimination nor the very general
Theorem VI.5.6 are any harder to prove than the particular case of finite
abelian
groups laboriously worked out by hand in this section---a common benefit of
finding
the right general point of view is that, as a rule, proofs simplify. Technical
work
such as that performed in order to prove Lemma 6.3 is absorbed into the work
necessary to build up the more general apparatus; the need for such
technicalities
evaporates in the process.

\subsection*{Exercises}

6.1. ? Prove that the decomposition of a finite abelian group G as a direct sum
of cyclic p-groups is unique. (Hint: The prime factorization of |G| determines
the
primes, so it suffices to show that if
Z
 \oplus  \dots  \oplus  Z
 = \sim 
 Z
 \oplus  \dots  \oplus  Z
 ,
pr1 Z
 prm Z
 ps1 Z
 psn Z
with r1 \geq  \dots  \geq  rm and s1 \geq  \dots  \geq  sn , then m = n and ri =
si for all i. Do this by
induction, by considering the group pG obtained as the image of the homomorphism
G \to  G defined by g ?\to  pg.) [\S{}6.2, \S{}VI.5.3, VI.5.12]
6.2. Complete the classification of groups of order 8 (cf. Exercise 2.16).
6.3. Let G be a noncommutative group of order p3 , where p is a prime integer.
Prove that Z(G) \sim 
 = Z/pZ and G/Z(G) \sim 
 = Z/pZ \times  Z/pZ.
6.4. Classify abelian groups of order 400.
6.5. Let p be a prime integer. Prove that the number of distinct isomorphism
classes
of abelian groups of order pr equals the number of partitions of the integer r.
6.6. ? How many abelian groups of order 1024 are there, up to isomorphism?
[\S{}II.6.3]
6.7. \lnot  Let p > 0 be a prime integer, G a finite abelian group, and denote
by
\rho  : G \to  G the homomorphism defined by \rho (g) = pg.
\bullet  Let A be a finite abelian group such that pA = 0. Prove that A \sim 
 = Z/pZ \oplus  \dots  \oplus 
Z/pZ.
\bullet  Prove that p ker \rho  and p(coker \rho ) are both 0.
\bullet  Prove that ker \rho  \sim 
 = coker \rho .
\bullet  Prove that every subgroup of G of order p is contained in ker \rho  and that every
subgroup of G of index p contains im \rho .
\bullet  Prove that the number of subgroups of G of order p equals the number of
subgroups of G of index p.
[6.8]
6.8. \lnot  Let G be a finite abelian p-group, with elementary divisors p n1 , .
. . , pnr (n1 \geq
n2 \geq  \dots  ). Prove that G has a subgroup H with invariant divisors pm1 , .
. . , pms
(m1 \geq  m2 \geq  \dots  ) if and only if s \leq  r and mi \leq  ni for i = 1,
. . . , s. (Hint: One
direction is immediate. For the other, with notation as in Exercise 6.7, compare
ker \rho  for H and G to establish s \leq  r; this also proves the
 statement
 if all ni = 1.
"For
 the
 general
 case
 use
 induction,
 noting
 that
 if
 G
 \sim 
 =
 i
 Z/p
ni
 Z,
 then
 \rho (G) =
 \sim 
"
 ni -1
i Z/p
 Z.)
Prove that the same description holds for the homomorphic images of G. [6.9]
6.9. Let H be a subgroup of a finite abelian group G. Prove that G contains a
sub-
group isomorphic to G/H. (Reduce to the case of p-groups; then use Exercise
6.8.)
Show that both hypotheses `finite' and `abelian' are needed for this result.
(Hint:
Q8 has a unique subgroup of order 2.)
6.10. The dual of a finite group G is the abelian group G\vee  := HomGrp (G,
C\ast  ),
where C\ast  is the multiplicative group of C.
\bullet  Prove that the image of every \sigma  \in  G\vee  consists of roots of 1 in C, that is, roots
of polynomials xn - 1 for some n.
\bullet  Prove that if G is a finite abelian group, then G \sim 
 = G\vee  . (Hint: First prove
this for cyclic groups; then use the classification theorem to generalize to the
arbitrary case.)
In \S{}VIII.6.5 we will encounter another notion of `dual' of a group.
6.11. \bullet  Use the classification theorem for finite abelian groups (Theorem
6.6) to
classify all finite modules over the ring Z/nZ.
\bullet  Prove that if p is prime, all finite modules over Z/pZ are free32 .
6.12. Let G, H, K be finite abelian groups such that G \oplus  H \sim 
 = G \oplus  K. Prove that
H \sim 
 = K.
6.13. \lnot  Let G, H be finite abelian groups such that, for all positive
integers n, G
and H have the same number of elements of order n. Prove that G \sim 
 = H. (Note:
The `abelian' hypothesis is necessary! C4 \times  C4 and Q8 \times  C2 are
nonisomorphic
groups both with 1 element of order 1, 3 elements of order 2, and 12 elements of
order 4.) [\S{}II.4.3]
6.14. Let G be a finite abelian p-group, and assume G has only one subgroup of
or-
der p. Prove that G is cyclic. (This is in some sense a converse to Proposition
II.6.11.
You are welcome to try to prove it `by hand', but use of the classification
theorem
will simplify the argument considerably.)
32As we will see in Proposition VI.4.10, this property characterizes fields.
6.15. Let G be a finite abelian group, and let a \in  G be an element of maximal
order in G. Prove that the order of every b \in  G divides |a|. (This
essentially
reproduces the result of Exercise II.1.15.)
6.16. Let G be an abelian group of order n, and assume that G has at most one
subgroup of order d for all d | n. Prove that G is cyclic.
