\chapter{Irreducibility and factorization in integral domains}
\label{chap:irred}

We move our attention back to \emph{rings} and analyze several useful
classes of integral domains.  One guiding theme in this chapter is the
issue of \emph{factorization}: we will address the problem of
existence and uniqueness of factorizations of elements in a ring,
abstracting good factorization properties of rings such as $\Z$ or
$k[x]$ (for $k$ a field) to whole classes of integral domains.  The
reader may want to associate the following picture with the first part
of this chapter:


\begin{center}
  \begin{tikzpicture}[scale=0.85]
    \fill[gray!60] (0,0) ellipse (5.8cm and 3.4cm); \fill[gray!35]
    (0,-0.1) ellipse (4.8cm and 2.6cm); \fill[gray!55] (-0.7,-0.35)
    ellipse (2.9cm and 1.8cm); \fill[gray!55] (1.2,-0.35) ellipse
    (2.5cm and 1.75cm); \fill[gray!25] (0.1,-0.55) ellipse (1.45cm and
    1.0cm); \fill[gray!15] (0.1,-0.7) ellipse (0.95cm and 0.65cm);
    \fill[white] (0.1,-0.85) ellipse (0.58cm and 0.38cm);
    \node[font=\bfseries\itshape] at (0,3.05) {Integral domains};
    \node[font=\itshape] at (0.4,2.1) {Domains with factorization};
    \node[font=\itshape] at (-2.5,0.8) {UFDs}; \node[font=\itshape,
    rotate=-80] at (3.0,0.6) {Noetherian domains};
    \node[font=\itshape] at (0.1,-0.05) {PIDs}; \node[font=\itshape,
    scale=0.8] at (0.1,-0.58) {Euclidean Domains};
    \node[font=\itshape, scale=0.75] at (0.1,-0.85) {Fields};
  \end{tikzpicture}
\end{center}

\emph{Blanket assumption}: all rings considered in this chapter will
be \emph{commutative}\footnote{Also, recall that all our rings have 1;
  cf.\ \Cref{defn:rings:ring}.}.  In fact, most of the special classes
of rings we will consider will be \emph{integral domains}, that is,
commutative rings with 1 and with no nonzero zero-divisors (cf.\
\Cref{defn:rings:ring}0).

\section{Chain conditions and existence of factorizations}
\label{sec:irred:chain-conditions-and-existence-of-factorizations}


\subsection{Noetherian rings revisited}
\label{subsec:irred:noetherian-rings-revisited}

Let $R$ be a commutative ring. Recall that $R$ is said to be
\emph{Noetherian} if every ideal of $R$ is finitely generated
(\Cref{defn:rings:noetherian}). In fact, this is a special case of the
corresponding definition for modules: a module $M$ over a ring $R$ is
Noetherian if every submodule of $M$ is finitely generated
(\Cref{defn:rings:noetherian-module}). In~\S{}\ref{subsec:rings:submodule-generated-by-a-subset-noetherian-modules}
we have verified that this condition is preserved through exact
sequences: if $M$, $N$, $P$ are $R$-modules and
\[0 \longrightarrow N \longrightarrow M \longrightarrow P
  \longrightarrow 0\] is an exact sequence of $R$-modules, then $M$ is
Noetherian if and only if both $N$ and $P$ are Noetherian
(\Cref{prop:rings:noetherian-subquotient}). An easy and useful
consequence of this fact is that every finitely generated module over
a Noetherian ring is Noetherian
(\Cref{coro:rings:fg-module-over-noetherian-is-noetherian}).

The Noetherian condition may be expressed in alternative ways, and it
is useful to acquire some familiarity with them.
\begin{prop}{}{prop:irred:1.1}
  Let $R$ be a commutative ring, and let $M$ be an $R$-module. Then
  the following are equivalent:
  \begin{enumerate}[label=(\arabic*)]
  \item $M$ is Noetherian; that is, every submodule of $M$ is finitely
    generated.
  \item Every ascending chain of submodules of $M$ stabilizes; that
    is, if
    \[N_1 \subseteq N_2 \subseteq N_3 \subseteq \dots\] is a chain of
    submodules of $M$, then $\exists i$ such that
    $N_i = N_{i+1} = N_{i+2} = \dots$
  \item Every nonempty family of submodules of $M$ has a maximal
    element w.r.t.\ inclusion.
  \end{enumerate}
\end{prop}

The second condition listed here is called the \emph{ascending chain
  condition} (a.c.c.) for submodules. For $M = R$,
\cref{prop:irred:1.1} tells us (among other things) that a ring is
Noetherian if and only if the ascending chain condition holds for its
ideals.
\begin{proof}
  $(1) \implies (2)$: Assume that $M$ is Noetherian, and let
  $N_1 \subseteq N_2 \subseteq N_3 \subseteq \dots$ be a chain of
  submodules of $M$. Consider the union $N = \bigcup_i N_i$: the
  reader will verify that $N$ is a submodule of $M$. Since $M$ is
  Noetherian, $N$ is finitely generated, say
  $N = \langle n_1, \dots, n_r \rangle$. Now
  $n_k \in N \implies n_k \in N_i$ for some $i$; by picking the
  largest such $i$, we see that $\exists i$ such that all
  $n_1, \dots, n_r$ are contained in $N_i$. But then
  $N \subseteq N_i$, and since $N_i \subseteq N_{i+1} \subseteq \dots$
  are all contained in $N = N_i$, it follows that
  $N_i = N_{i+1} = N_{i+2} = \dots$ as needed.

  $(2) \implies (3)$: Arguing contrapositively, assume that $M$ admits
  a family $\mathscr{F}$ of submodules that does not have a maximal
  element.  Construct an infinite ascending chain as follows: let
  $N_1$ be any element of $\mathscr{F}$; since $N_1$ is not maximal in
  $\mathscr{F}$, there exists an element $N_2$ of $\mathscr{F}$ such
  that $N_1 \subsetneq N_2$; since $N_2$ is not maximal in
  $\mathscr{F}$, there exists an element $N_3$ of $\mathscr{F}$ such
  that $N_2 \subsetneq N_3$; etc.\ The chain
  $N_1 \subsetneq N_2 \subsetneq N_3 \subsetneq \dots$ does not
  stabilize, showing that $(2)$ does not hold.

  $(3) \implies (1)$: Assume $(3)$ holds, and let $N$ be a submodule
  of $M$.  Then the family $\mathscr{F}$ of finitely generated
  submodules of $N$ is nonempty (as $(0) \in \mathscr{F}$); hence it
  has a maximal element $N'$.  Say that
  $N' = \langle n_1, \dots, n_r \rangle$. Now I claim that $N' = N$:
  indeed, let $n \in N$; the submodule
  $\langle n_1, \dots, n_r, n \rangle$ is finitely generated, and
  therefore it is in $\mathscr{F}$; as it contains $N'$ and $N'$ is
  maximal, necessarily $\langle n_1, \dots, n_r, n \rangle = N'$; in
  particular $n \in N'$, as needed. This shows that $N = N'$ is
  finitely generated, and since $N \subseteq M$ was arbitrary, this
  implies that $M$ is Noetherian.
\end{proof}
Noetherian rings are a very useful and flexible class of
rings. In~\S{}\ref{subsec:rings:finitely-generated-vs-finite-type} I
mentioned the important fact that every \emph{finite-type algebra}
over a Noetherian ring is Noetherian. `Finite-type (commutative)
algebra' is just a fancy name for a quotient of a polynomial ring
(\S{}\ref{subsec:rings:finitely-generated-vs-finite-type}), so this is
what the fact states:

\begin{thm}{}{thm:irred:1.2}
  Let $R$ be a Noetherian ring, and let $J$ be an ideal of the
  polynomial ring $R[x_1, \dots, x_n]$. Then the ring
  $R[x_1, \dots, x_n]/J$ is Noetherian.
\end{thm}

Note that finite-type $R$-algebras are (in general) very far from
being finitely generated as modules over $R$ (cf.\ again
\S{}\ref{subsec:rings:finitely-generated-vs-finite-type}), so it would
be foolish to expect them to be Noetherian as $R$-modules. The fact
that they turn out to be Noetherian as rings (that is, as modules over
themselves) provides us with a huge class of examples of Noetherian
rings, among which are the rings of (classical) algebraic geometry and
number theory. Thus, entire fields of mathematics are a little more
manageable thanks to \cref{thm:irred:1.2}.

The proof of this deep fact is surprisingly easy. By
\Cref{exer:irred:quot-of-noeth-is-noeth}, it suffices to prove that
$R$ Noetherian $\implies$ $R[x_1, \dots, x_n]$ Noetherian; and an
immediate induction reduces the statement to the following particular
case, which carries a distinguished name:
\begin{lem}{Hilbert's basis theorem}{lem:irred:1.3}
  $R$ Noetherian $\implies$ $R[x]$ Noetherian.
\end{lem}

\begin{proof}
  Assume $R$ is Noetherian, and let $I$ be an ideal of $R[x]$. We have
  to prove that $I$ is finitely generated.

  Recall that if
  $f(x) = a_d x^d + a_{d-1} x^{d-1} + \dots + a_0 \in R[x]$ and
  $a_d \neq 0$, then $a_d$ is called the \emph{leading coefficient} of
  $f(x)$. Consider the following subset of $R$:
  \[A = \{0\} \cup \{a \in R \mid a \text{ is a leading coefficient of
      an element of } I\}.\] It is clear that $A$ is an ideal of $R$
  (\Cref{exer:irred:hilbert-basis-thm-set-a-is-ideal}); since $R$ is
  Noetherian, $A$ is finitely generated. Thus there exist elements
  $f_1(x), \dots, f_r(x) \in I$ whose leading coefficients
  $a_1, \dots, a_r$ generate $A$ as an ideal of $R$.

  Now let $d_i$ be the degree of $f_i(x)$, and let $d$ be the maximum
  among these degrees. Consider the sub-$R$-module
  \[M = \langle 1, x, x^2, \dots, x^{d-1} \rangle \subseteq R[x],\]
  that is, the $R$-module consisting of polynomials of degree $<
  d$. Since $M$ is finitely generated as a module over $R$, it is
  Noetherian as an $R$-module (by
  \Cref{coro:rings:fg-module-over-noetherian-is-noetherian}). Therefore,
  the submodule $M \cap I$ of $M$ is finitely generated over $R$, say
  by $g_1(x), \dots, g_s(x) \in I$.

\begin{fact}{Claim~1.4}{fact:irred:claim:1.4}
  $I = (f_1(x), \dots, f_r(x), g_1(x), \dots, g_s(x))$.
\end{fact}

This claim implies the statement of the theorem. To prove the claim,
we only need to prove the $\subseteq$ inclusion; to this end, let
$\alpha(x) \in I$ be an arbitrary polynomial in $I$. If
$\deg\alpha(x) \geq d$, let $a$ be the leading coefficient of
$\alpha(x)$. Then $a \in A$, so $\exists b_1, \dots, b_r \in R$ such
that $a = b_1 a_1 + \dots + b_r a_r$. Letting $e = \deg\alpha(x)$, so
that $e \geq d_i$ for all $i$, this says that
\[\alpha(x) - b_1 x^{e-d_1} f_1(x) - \dots - b_r x^{e-d_r} f_r(x)\]
has degree $< e$. Iterating this procedure, we obtain a finite list of
polynomials $\beta_1(x), \dots, \beta_r(x) \in R[x]$ such that
\[\alpha(x) - \beta_1(x) f_1(x) - \dots - \beta_r(x) f_r(x)\]
has degree $< d$. But this places this element in $M \cap I$;
therefore $\exists c_1, \dots, c_s \in R$ such that
\[\alpha(x) - \beta_1(x)f_1(x) - \dots - \beta_r(x)f_r(x)
  = c_1 g_1(x) + \dots + c_s g_s(x),\] and we are done, since this
verifies that
\begin{align*}
  \alpha(x) &= \beta_1(x)f_1(x) + \dots + \beta_r(x)f_r(x)
              + c_1 g_1(x) + \dots + c_s g_s(x) \\
            &\in (f_1(x), \dots, f_r(x), g_1(x), \dots, g_s(x)),
\end{align*}
completing the proof of \cref{fact:irred:claim:1.4}, hence of
\cref{lem:irred:1.3}, hence of \cref{thm:irred:1.2}.
\end{proof}

\subsection{Prime and irreducible elements}
\label{subsec:irred:prime-and-irreducible-elements}
Let $R$ be a (commutative) ring, and let $a, b \in R$. We say that $a$
\emph{divides} $b$, or that $a$ is a \emph{divisor} of $b$, or that
$b$ is a \emph{multiple} of $a$, if $b \in (a)$, that is,
\[(\exists c \in R),\quad b = ac.\] We use the notation $a \mid b$.

Two elements $a$, $b$ are \emph{associates} if $(a) = (b)$, that is,
if $a \mid b$ and $b \mid a$.
\begin{lem}{}{lem:irred:1.5}
  Let $a$, $b$ be nonzero elements of an integral domain $R$. Then $a$
  and $b$ are associates if and only if $a = ub$, for $u$ a unit in
  $R$.
\end{lem}

\begin{proof}
  Assume $a$ and $b$ are associates. Then $\exists c, d \in R$ such
  that $b = ac$, $a = bd$; therefore $a = bd = acd$, i.e.,
  $a(1 - cd) = 0$.  Since cancellation by nonzero elements holds in
  integral domains, this implies $cd = 1$. Thus $c$ is a unit, as
  needed.

  The converse is left to the reader.
\end{proof}
Incidentally, here the reader sees why it is convenient to restrict
our attention to integral domains. This argument really shows that if
$(a) = (b) \neq (0)$ in an integral domain, and $b = ca$, then $c$ is
necessarily a unit. Away from the comfortable environment of integral
domains, even such harmless-looking statements may fail: in $\Z/6\Z$
the classes $[2]_6$, $[4]_6$ of 2 and 4 are associates according to
our definition, and $[4]_6 = [2]_6 \cdot [2]_6$, yet $[2]_6$ is not a
unit.  However, $[4]_6 = [5]_6 \cdot [2]_6$ and $[5]_6$ is a unit, so
this is not a counterexample to \cref{lem:irred:1.5}. In fact,
\cref{lem:irred:1.5} may fail over rings with `non-harmless'
zero-divisors (yes, there is such a notion).

The notions reviewed above generalize directly the corresponding
notions in $\Z$. We are going to explore analogues of other common
notions in $\Z$, such as `primality' and `irreducibility', in more
general integral domains.
\begin{defn}{}{defn:irred:1.6}
  Let $R$ be an integral domain.
  \begin{itemize}
  \item An element $a \in R$ is \emph{prime} if the ideal $(a)$ is
    prime; that is, $a$ is not a unit and (cf.\
    \Cref{prop:rings:prime-maximal-equiv})
    \[a \mid bc \implies (a \mid b \quad\text{or}\quad a \mid c).\]
  \item An element $a \in R$ is \emph{irreducible} if $a$ is not a
    unit and
    \[a = bc \implies (b \text{ is a unit or } c \text{ is a unit}).\]
  \end{itemize}
\end{defn}

Note that $0$ is always reducible (integral domains are nonzero
rings!).  For nonzero elements, there are useful alternative ways to
think about the notion of `irreducible': a nonunit $a \neq 0$ is
irreducible if and only if
\begin{itemize}
\item $a = bc$ implies that $a$ is an associate of $b$ or of $c$;
\item $a = bc$ implies that $(a) = (b)$ or $(a) = (c)$
  (\cref{lem:irred:1.5});
\item $(a) \subseteq (b) \implies (b) = (a)$ or $(b) = (1)$
  (\Cref{exer:irred:irred-iff-a-maximal-principal});
\item $(a)$ is maximal among proper principal ideals (rephrasing the
  previous point!).
\end{itemize}

It is important to realize that primality and irreducibility are
\emph{not} equivalent, even for nonzero elements; this is somewhat
counterintuitive since they \emph{are} equivalent in $\Z$, as the
reader should verify\footnote{This fact will be fully explained by the
  general theory, so the reader should work this out right away.}
(\Cref{exer:irred:prime-iff-irred-in-z}). What is true in general is
that prime is stronger than irreducible:

\begin{lem}{}{lem:irred:1.7}
  Let $R$ be an integral domain, and let $a \in R$ be a nonzero prime
  element. Then $a$ is irreducible.
\end{lem}

\begin{proof}
  Since $(a)$ is prime, $(a) \neq (1)$; hence $a$ is not a unit. If
  $a = bc$, then $bc = a \in (a)$; therefore $b \in (a)$ or
  $c \in (a)$ since $(a)$ is prime. Assuming without loss of
  generality $b \in (a)$, we have $(b) \subseteq (a)$. On the other
  hand $a = bc$ implies $(a) \subseteq (b)$: hence $(a) = (b)$, that
  is, $a$ and $b$ are associates, as needed.
\end{proof}

We will soon see under what circumstances the converse statement
holds.

\subsection{Factorization into irreducibles; domains with
  factorizations}
\label{subsec:irred:factorization-into-irreducibles-domains-with-factorizations}

\begin{defn}{}{defn:irred:1.8}
  Let $R$ be an integral domain. An element $r \in R$ has a
  \emph{factorization} (or \emph{decomposition}) \emph{into
    irreducibles} if there exist irreducible elements
  $q_1, \dots, q_n$ such that $r = q_1 \cdots q_n$.

  This factorization is \emph{unique} if the elements $q_i$ are
  determined by $r$ up to order and associates, that is, if whenever
  $r = q'_1 \cdots q'_m$ is another factorization of $r$ into
  irreducibles, then $m = n$ and $q'_i$ is an associate of $q_i$ after
  (possibly) shuffling the factors.
\end{defn}

\begin{defn}{}{defn:irred:1.9}
  An integral domain $R$ is a \emph{domain with factorizations} (or
  `\emph{factorizations exist in} $R$') if every nonzero, nonunit
  element $r \in R$ has a factorization into irreducibles.
\end{defn}

\begin{defn}{}{defn:irred:1.10}
  An integral domain $R$ is \emph{factorial}, or a \emph{unique
    factorization domain} (abbreviated \emph{UFD}), if every nonzero,
  nonunit element $r \in R$ has a unique factorization into
  irreducibles.
\end{defn}

The terminology introduced in \cref{defn:irred:1.9} does not appear to
be too standard; by contrast, UFDs are famous.

We will study the unique factorization condition
in~\S{}\ref{sec:irred:ufds-pids-euclidean-domains}. For now, it seems
worth spending a little time contemplating the mere existence of
factorizations. Interestingly, this condition is implied by an
ascending chain condition, for a special class of ideals.

\begin{prop}{}{prop:irred:1.11}
  Let $R$ be an integral domain, and let $r$ be a nonzero, nonunit
  element of $R$. Assume that every ascending chain of principal
  ideals
  \[(r) \subseteq (r_1) \subseteq (r_2) \subseteq (r_3) \subseteq
    \dots\] stabilizes. Then $r$ has a factorization into
  irreducibles.
\end{prop}

\begin{proof}
  Assume that $r$ does not have a factorization into irreducible
  elements. In particular, $r$ is itself not irreducible; thus
  $\exists r_1, s_1 \in R$ such that $r = r_1 s_1$ and
  $(r) \subsetneq (r_1)$, $(r) \subsetneq (s_1)$. If both $r_1$, $s_1$
  have factorizations into irreducibles, the product of these
  factorizations gives a factorization of $r$; thus we may assume that
  (e.g.)  $r_1$ does not have a factorization into
  irreducibles. Therefore we have $(r) \subsetneq (r_1)$ and $r_1$
  does not have a factorization; iterating this argument constructs an
  infinitely increasing chain
  \[(r) \subsetneq (r_1) \subsetneq (r_2) \subsetneq (r_3) \subsetneq
    \dots,\] contradicting our hypothesis.
\end{proof}

Thus, factorizations exist in integral domains in which the ascending
chain condition holds for principal ideals. This has the following
immediate and convenient consequence:

\begin{coro}{}{coro:irred:1.12}
  Let $R$ be a Noetherian domain. Then factorizations exist in $R$.
\end{coro}

\begin{proof}
  By \cref{prop:irred:1.1}, Noetherian domains satisfy the ascending
  chain condition for all ideals.
\end{proof}

\cref{coro:irred:1.12} verifies part of the picture presented at the
beginning of the chapter: the class of Noetherian domains is contained
in the class of domains with factorization. This inclusion is proper:
for instance, the standard example of a non-Noetherian ring,
$\Z[x_1, x_2, x_3, \dots]$, (\Cref{exam:rings:infinite-poly-ring-fg})
does have factorizations. Indeed, every given polynomial
$f \in \Z[x_1, x_2, \dots]$ involves only finitely many
variables\footnote{Remember that polynomials are finite linear
  combinations of monomials;
  cf.~\S{}\ref{subsec:rings:polynomial-rings}.}, so it belongs to a
subring $\Z[x_1, \dots, x_n]$ isomorphic to an ordinary polynomial
ring over $\Z$; further, this subring contains every divisor of
$f$. It follows easily (cf.\
\Cref{exer:irred:acc-principal-ideals-infinite-vars}) that the
ascending chain condition for principal ideals holds in
$\Z[x_1, x_2, x_3, \dots]$, because it holds in $\Z[x_1, \dots, x_n]$
(since this ring is Noetherian, by Hilbert's basis theorem).

\subsection*{Exercises}

Remember that in this section all rings are taken to be
\emph{commutative}.
\begin{exer}{\exerused}{exer:irred:quot-of-noeth-is-noeth}
  Let $R$ be a Noetherian ring, and let $I$ be an ideal of $R$. Prove
  that $R/I$ is a Noetherian
  ring. [\S{}\ref{subsec:irred:noetherian-rings-revisited}]
\end{exer}

\begin{exer}{}{exer:irred:rx-noeth-implies-r-noeth}
  Prove that if $R[x]$ is Noetherian, so is $R$. (This is a `converse'
  to Hilbert's basis theorem.)
\end{exer}

\begin{exer}{}{exer:irred:subrings-of-kx-over-kt-noeth}
  Let $k$ be a field, and let $f \in k[x]$, $f \notin k$. For every
  subring $R$ of $k[x]$ containing $k$ and $f$, define a homomorphism
  $\phi : k[t] \to R$ by extending the identity on $k$ and mapping $t$
  to $f$. This makes every such $R$ a $k[t]$-algebra
  (\Cref{exam:rings:ring-hom-defines-module}).
  \begin{itemize}
  \item Prove that $k[x]$ is finitely generated as a $k[t]$-module.
  \item Prove that every subring $R$ as above is finitely generated as
    a $k[t]$-module.
  \item Prove that every subring of $k[x]$ containing $k$ is a
    Noetherian ring.
  \end{itemize}
\end{exer}

\begin{exer}{}{exer:irred:cont-fns-on-01-not-noeth}
  Let $R$ be the ring of real-valued continuous functions on the
  interval $[0, 1]$. Prove that $R$ is not Noetherian.
\end{exer}

\begin{exer}{}{exer:irred:power-set-ring-noeth-iff}
  Determine for which sets $S$ the power set ring $\mathscr{P}(S)$ is
  Noetherian. (Cf.\ \Cref{exer:rings:subsets-of-t-form-ideal}.)
\end{exer}
\begin{exer}{\exerused}{exer:irred:hilbert-basis-thm-set-a-is-ideal}
  Let $I$ be an ideal of $R[x]$, and let $A \subseteq R$ be the set
  defined in the proof of \cref{thm:irred:1.2}. Prove that $A$ is an
  ideal of $R$.  [\S{}\ref{subsec:irred:noetherian-rings-revisited}]
\end{exer}

\begin{exer}{}{exer:irred:power-series-ring-noeth}
  Prove that if $R$ is a Noetherian ring, then the ring of power
  series $R[[x]]$ (cf.~\S{}\ref{subsec:rings:polynomial-rings}) is
  also Noetherian. (Hint: The order of a power series
  $\sum_{i=0}^{\infty} a_i x^i$ is the smallest $i$ for which
  $a_i \neq 0$; the dominant coefficient is then $a_i$. Let
  $A_i \subseteq R$ be the set of dominant coefficients of series of
  order $i$ in $I$, together with $0$. Prove that $A_i$ is an ideal of
  $R$ and $A_0 \subseteq A_1 \subseteq A_2 \subseteq \dots$. This
  sequence stabilizes since $R$ is Noetherian, and each $A_i$ is
  finitely generated for the same reason. Now adapt the proof of
  \cref{lem:irred:1.3}.)
\end{exer}
\begin{exer}{}{exer:irred:ideal-contains-product-of-primes}
  Prove that every ideal in a Noetherian ring $R$ contains a finite
  product of prime ideals. (Hint: Let $\mathscr{F}$ be the family of
  ideals that do not contain finite products of prime ideals. If
  $\mathscr{F}$ is nonempty, it has a maximal element $M$ since $R$ is
  Noetherian. Since $M \in \mathscr{F}$, $M$ is not itself prime, so
  $\exists a, b \in R$ s.t.\ $a \notin M$, $b \notin M$, yet
  $ab \in M$. What's wrong with this?)
\end{exer}
\begin{exer}{$\lnot$}{exer:irred:noeth-implies-fin-minimal-primes}
  Let $R$ be a commutative ring, and let $I \subseteq R$ be a proper
  ideal.  The reader will prove in
  \Cref{exer:irred:minimal-primes-exist} that the set of prime ideals
  containing $I$ has minimal elements (the \emph{minimal primes} of
  $I$).  Prove that if $R$ is Noetherian, then the set of minimal
  primes of $I$ is finite. (Hint: Let $\mathscr{F}$ be the family of
  ideals that do not have finitely many minimal primes. If
  $\mathscr{F} \neq \emptyset$, note that $\mathscr{F}$ must have a
  maximal element $I$, and $I$ is not prime itself.  Find ideals
  $J_1$, $J_2$ strictly larger than $I$, such that
  $J_1 J_2 \subseteq I$, and deduce a contradiction.)
  [\hyperref[exer:linalg:intersection-of-assoc-primes-eq-nilradical]{VI.4.10}]
\end{exer}
\begin{exer}{$\lnot$}{exer:irred:artinian-basics}
  By \cref{prop:irred:1.1}, a ring $R$ is Noetherian if and only if it
  satisfies the a.c.c.\ for ideals. A ring is \emph{Artinian} if it
  satisfies the d.c.c.\ (descending chain condition) for ideals. Prove
  that if $R$ is Artinian and $I \subseteq R$ is an ideal, then $R/I$
  is Artinian.  Prove that if $R$ is an Artinian integral domain, then
  it is a field.  (Hint: Let $r \in R$, $r \neq 0$. The ideals $(r^n)$
  form a descending sequence; hence $(r^n) = (r^{n+1})$ for some
  $n$. Therefore\dots) Prove that Artinian rings have Krull dimension
  $0$ (that is, prime ideals are maximal in Artinian
  rings\footnote{One can prove that Artinian rings are necessarily
    Noetherian; in fact, a ring is Artinian if and only if it is
    Noetherian and has Krull dimension $0$. Thus, the d.c.c.\ implies
    the a.c.c., while the a.c.c.\ implies the d.c.c.\ if and only if
    all prime ideals are
    maximal.}). [\hyperref[exer:irred:pid-mod-ideal-is-artinian]{2.11}]
\end{exer}

\begin{exer}{}{exer:irred:associate-is-equiv-reln}
  Prove that the `associate' relation is an equivalence relation.
\end{exer}

\begin{exer}{\exerused}{exer:irred:irred-iff-a-maximal-principal}
  Let $R$ be an integral domain. Prove that a nonzero $a \in R$ is
  irreducible if and only if $(a)$ is maximal among proper principal
  ideals of
  $R$. [\S{}\ref{subsec:irred:prime-and-irreducible-elements},
  \S{}\ref{subsec:irred:pid-implies-ufd}]
\end{exer}

\begin{exer}{\exerused}{exer:irred:prime-iff-irred-in-z}
  Prove that, for nonzero elements, prime $\iff$ irreducible in $\Z$.
  [\S{}\ref{subsec:irred:prime-and-irreducible-elements},
  \S{}\ref{subsec:irred:pid-implies-ufd}]
\end{exer}

\begin{exer}{}{exer:irred:a-prime-mod-b-iff-b-prime-mod-a}
  For $a$, $b$ in a commutative ring $R$, prove that the class of $a$
  in $R/(b)$ is prime if and only if the class of $b$ in $R/(a)$ is
  prime.
\end{exer}

\begin{exer}{\exerused}{exer:irred:acc-principal-ideals-infinite-vars}
  Identify $S = \Z[x_1, \dots, x_n]$ in the natural way with a subring
  of the polynomial ring in countably infinitely many variables
  $R = \Z[x_1, x_2, x_3, \dots]$. Prove that if $f \in S$ and
  $(f) \subseteq (g)$ in $R$, then $g \in S$ as well. Conclude that
  the ascending chain condition for principal ideals holds in $R$, and
  hence $R$ is a domain with
  factorizations. [\S{}\ref{subsec:irred:factorization-into-irreducibles-domains-with-factorizations},
  \S{}\ref{subsec:irred:r-ufd-implies-r-x-ufd}]
\end{exer}

\begin{exer}{}{exer:irred:acc-principal-ideals-fails-example}
  Let
  \[R = \frac{\Z[x_1, x_2, x_3, \dots]}{(x_1 - x_2^2,\, x_2 - x_3^2,\,
      \dots)}.\] Does the ascending chain condition for principal
  ideals hold in $R$?
\end{exer}

\begin{exer}{\exerused}{exer:irred:z-sqrt-neg5-not-ufd}
  Consider the subring of $\C$:
  \[\Z[\sqrt{-5}] := \{a + bi\sqrt{5} \mid a, b \in \Z\}.\]
  \begin{itemize}
  \item Prove that this ring is isomorphic to $\Z[t]/(t^2 + 5)$.
  \item Prove that it is a Noetherian integral domain.
  \item Define a `norm' $N$ on $\Z[\sqrt{-5}]$ by setting
    $N(a + bi\sqrt{5}) = a^2 + 5b^2$. Prove that $N(zw) = N(z)N(w)$.
    (Cf.\ \Cref{exer:rings:z-sqrt-d-subring-of-c}.)
  \item Prove that the units in $\Z[\sqrt{-5}]$ are $\pm 1$. (Use the
    preceding point.)
  \item Prove that $2$, $3$, $1 + i\sqrt{5}$, $1 - i\sqrt{5}$ are all
    irreducible nonassociate elements of $\Z[\sqrt{-5}]$.
  \item Prove that no element listed in the preceding point is
    prime. (Prove that the rings obtained by mod-ing out the ideals
    generated by these elements are not integral domains.)
  \item Prove that $\Z[\sqrt{-5}]$ is not a UFD.
  \end{itemize} [\S{}\ref{subsec:irred:characterization-of-ufds},
  \ref{exer:irred:z-1-plus-sqrt-neg19-over-2-not-euclidean},
  \ref{exer:irred:z-sqrt-neg2-and-imaginary-quad-euclidean}]
\end{exer}

\section{UFDs, PIDs, Euclidean domains}
\label{sec:irred:ufds-pids-euclidean-domains}


\subsection{Irreducible factors and greatest common divisor}
\label{subsec:irred:irreducible-factors-and-greatest-common-divisor}
An integral domain $R$ is a UFD if factorizations exist in $R$ and are
unique in the sense of \cref{defn:irred:1.8}. Thus, in a UFD all
elements (other than $0$ and the units) determine a \emph{multiset} (a
set of elements `with multiplicity';
cf.~\S{}\ref{subsec:prelims:sets}) of irreducible factors, determined
up to the associate relation. We can also agree that units have no
factors; that is, the corresponding multiset is $\emptyset$.

The following trivial remark is at the root of most elementary facts
about UFDs, such as the characterization of \cref{thm:irred:2.5}:

\begin{lem}{}{lem:irred:2.1}
  Let $R$ be a UFD, and let $a$, $b$, $c$ be nonzero elements of
  $R$. Then
  \begin{itemize}
  \item $(a) \subseteq (b) \iff$ the multiset of irreducible factors
    of $b$ is contained in the multiset of irreducible factors of $a$;
  \item $a$ and $b$ are associates (that is, $(a) = (b)$) $\iff$ the
    two multisets coincide;
  \item the irreducible factors of a product $bc$ are the collection
    of all irreducible factors of $b$ and of $c$.
  \end{itemize}
\end{lem}

The proof is left to the reader
(\Cref{exer:irred:divisibility-via-irred-factors-in-ufd}). The
advantage of working in a UFD resides in the fact that ring-theoretic
statements about elements of the ring often reduce to straightforward
set-theoretic statements about multisets of irreducible elements, by
means of \cref{lem:irred:2.1}.  One important instance of this
mechanism is the existence of \emph{greatest common divisors}. I have
liberally used this notion (at least for integers) in previous
chapters; now we can appreciate it from a more technical perspective.

\begin{defn}{}{defn:irred:2.2}
  Let $R$ be an integral domain, and let $a, b \in R$. An element
  $d \in R$ is a \emph{greatest common divisor} (often abbreviated
  `gcd') of $a$ and $b$ if $(a, b) \subseteq (d)$ and $(d)$ is the
  smallest principal ideal in $R$ with this property.
\end{defn}

In other words, $d$ is a gcd of $a$ and $b$ if $d \mid a$, $d \mid b$,
and $c \mid a$, $c \mid b \implies c \mid d$. This definition is
immediately extended to any finite number of elements.

Note that greatest common divisors are not defined uniquely by this
prescription: if $d$ is a greatest common divisor of $a$ and $b$, so
is every associate of $d$. Thus, the notation `$\gcd(a, b)$' should
only be used for the associate class formed by all greatest common
divisors of $a$, $b$. Of course, language is often (harmlessly) abused
on this point.  For example, the fact that we can talk about
\emph{the} greatest common divisor of two integers is due to the fact
that in $\Z$ there is a convenient way to choose a distinguished
element in each class of associate integers (that is, the nonnegative
one).

Also note that greatest common divisors need not exist (cf.\
\Cref{exer:irred:no-gcd-example-in-subring-of-zt}); but they do exist
in UFDs:
\begin{lem}{}{lem:irred:2.3}
  Let $R$ be a UFD, and let $a$, $b$ be nonzero elements of $R$. Then
  $a$, $b$ have a greatest common divisor.
\end{lem}

\begin{proof}
  We can write
  \[a = u q_1^{\alpha_1} \cdots q_r^{\alpha_r}, \quad b = v
    q_1^{\beta_1} \cdots q_r^{\beta_r}\] where $u$ and $v$ are units,
  the elements $q_i$ are irreducible, $q_i$ is not an associate of
  $q_j$ for $i \neq j$, and $\alpha_i \geq 0$, $\beta_i \geq 0$ (so
  that the multisets of irreducible factors of $a$, resp., $b$,
  consist of those $q_i$ for which $\alpha_i > 0$, resp.,
  $\beta_i > 0$; the units $u$, $v$ are included since the irreducible
  factors are only defined up to the associate relation).

  I claim that
  $d = q_1^{\min(\alpha_1,\beta_1)} \cdots
  q_r^{\min(\alpha_r,\beta_r)}$ is a gcd of $a$ and $b$. Indeed, $d$
  is clearly a divisor of $a$ and $b$; and if $c$ also divides $a$ and
  $b$, then the multiset of factors of $c$ must be contained in both
  multisets of factors for $a$ and $b$ (by \cref{lem:irred:2.1}); that
  is, $c = w q_1^{\gamma_1} \cdots q_r^{\gamma_r}$ with $w$ a unit and
  $\gamma_i \leq \alpha_i$, $\gamma_i \leq \beta_i$. This implies
  $\gamma_i \leq \min(\alpha_i, \beta_i)$, and hence $c \mid d$ (again
  by \cref{lem:irred:2.1}), as needed.
\end{proof}
Of course the argument given in the proof generalizes one of the
standard ways to compute greatest common divisors in $\Z$: find
smallest exponents in prime factorizations. But note that this is not
the only way to compute the gcd in $\Z$; we will come back to this
point in a moment. In fact, greatest common divisors in $\Z$ have
properties that should not be expected in a more general domain: for
example, the result of \Cref{exer:groups:bezout} does not generalize
to arbitrary domains (and not even to arbitrary UFDs), as the reader
will check in \Cref{exer:irred:gcd-not-lin-comb-in-zxy}.

\subsection{Characterization of UFDs}
\label{subsec:irred:characterization-of-ufds}
It is easy to construct integral domains where unique factorization
fails.  The diligent reader has already analyzed one example
(\Cref{exer:irred:z-sqrt-neg5-not-ufd}); for another, in the
domain\footnote{In algebraic geometry, this is the ring of a `quadric
  cone in $\A^4$'. The vertex of this cone (at the origin) is a
  singular point, and this has to do with the fact that $R$ is not a
  UFD.}
\[R = \frac{\C[x,y,z,w]}{(xw - yz)}\] the (classes of the) elements
$x$, $y$, $z$, $w$ are irreducible and not associates of one another;
since $xw - yz = 0$ in $R$, the element $r = xw$ has two distinct
factorizations into irreducibles: $r = xw = yz$.

Note that this ring is Noetherian, by \cref{thm:irred:1.2} (and in
particular factorizations do exist in $R$). Thus, there are Noetherian
integral domains that are not UFDs.

Also note that this ring provides an example in which the converse to
\cref{lem:irred:1.7} does \emph{not} hold: indeed, (the class of) $x$
is irreducible, but the quotient
\[\frac{\C[x,y,z,w]}{(xw-yz)} \Big/ (x) \cong
  \frac{\C[x,y,z,w]}{(x, xw-yz)} = \frac{\C[x,y,z,w]} {(x,yz)}\] is
not an integral domain (because $y \neq 0$, $z \neq 0$, and yet
$yz = 0$ in this ring); that is, $x$ is not prime.

In fact, and maybe a little surprisingly, the issue of unique
factorization is inextricably linked with the relation between
primality and irreducibility. Indeed, the `converse' to
\cref{lem:irred:1.7} does hold in UFDs:
\begin{lem}{}{lem:irred:2.4}
  Let $R$ be a UFD, and let $a$ be an irreducible element of $R$. Then
  $a$ is prime.
\end{lem}

\begin{proof}
  The element $a$ is not a unit, by definition of irreducible. Assume
  $bc \in (a)$: thus $(bc) \subseteq (a)$, and by \cref{lem:irred:2.1}
  the irreducible factors of $a$, that is, $a$ itself, must be among
  the factors of $b$ or of $c$. We have $b \in (a)$ in the first case
  and $c \in (a)$ in the second. This shows that $(a)$ is a prime
  ideal, as needed.
\end{proof}

In fact, more is true. Provided that the ascending chain condition for
principal ideals holds, then UFDs are characterized by the equivalence
between irreducibility and primality.
\begin{thm}{}{thm:irred:2.5}
  An integral domain $R$ is a UFD if and only if
  \begin{itemize}
  \item the a.c.c.\ for principal ideals holds in $R$ and
  \item every irreducible element of $R$ is prime.
  \end{itemize}
\end{thm}

\begin{proof}
  $(\implies)$ Assume that $R$ is a UFD. \cref{lem:irred:2.4} shows
  that irreducible elements of $R$ are prime. To prove that the
  a.c.c.\ for principal ideals holds, consider an ascending chain
  $(r_1) \subseteq (r_2) \subseteq (r_3) \subseteq \dots$. By
  \cref{lem:irred:2.1}, this chain determines a corresponding
  \emph{descending} chain of multisets of irreducible factors. A
  descending chain of finite multisets clearly stabilizes, and it
  follows (by \cref{lem:irred:2.1} again) that
  $(r_i) = (r_{i+1}) = (r_{i+2}) = \dots$ for large enough $i$.

  $(\impliedby)$ Now assume that $R$ satisfies the a.c.c.\ for
  principal ideals and irreducibles are prime. \cref{prop:irred:1.11}
  implies that factorizations exist in $R$; we have to verify
  uniqueness. Let $q_1, \dots, q_m$ and $q'_1, \dots, q'_n$ be
  irreducible elements of $R$, and assume
  \[q_1 \cdots q_m = q'_1 \cdots q'_n.\] Then
  $q'_1 \cdots q'_n \in (q_1)$, and $(q_1)$ is a prime ideal (by
  hypothesis); thus $q'_i \in (q_1)$ for some $i$, which we may assume
  to be $1$ after changing the order of the factors. Therefore
  $q'_1 = uq_1$ for some $u \in R$. Since $q'_1$ is irreducible and
  $q_1$ is not a unit, necessarily $u$ is a unit. Thus $q_1$ and
  $q'_1$ are associates. Canceling $q_1$ and replacing $q'_2$ by
  $uq'_2$, we find
  \[q_2 \cdots q_m = q'_2 \cdots q'_n.\] Repeating this process
  matches all factors one by one. It is clear that $m = n$, because
  otherwise we will obtain $1 =$ a product of irreducibles,
  contradicting the fact that irreducibles are not units.
\end{proof}

\cref{thm:irred:2.5} is a pleasant statement, but it does not make it
particularly easy to check whether a given ring is in fact a UFD. The
situation is not unlike that with Noetherian rings: the a.c.c.\ is a
sharp equivalent formulation of the Noetherian condition, but in
practice one relies more often on other tools, such as Hilbert's basis
theorem, in order to establish that a given ring is Noetherian. Does
an analog of Hilbert's basis theorem hold for unique factorization
domains?

Answering this question requires some preparatory work, and we will
come back to it
in~\S{}\ref{sec:irred:unique-factorization-in-polynomial-rings}.

\subsection{\texorpdfstring{PID $\implies$ UFD}{PID implies UFD}}
\label{subsec:irred:pid-implies-ufd}
There are simple ways to produce examples of UFDs. Recall
(from~\S{}\ref{sec:rings:ideals-and-quotients-remarks-and-examples-prime-and})
that a \emph{principal ideal domain} (abbreviated PID) is an integral
domain in which every ideal is
principal. In~\S{}\ref{sec:rings:ideals-and-quotients-remarks-and-examples-prime-and}
we have observed that
\begin{itemize}
\item PIDs are Noetherian.
\item $\Z$ and $k[x]$ (where $k$ is a field) are PIDs.
\item If $R$ is a PID and $a, b \in R$, then $d$ is a greatest common
  divisor for $a$ and $b$ if and only if $(a, b) = (d)$. In
  particular, if $d$ is a greatest common divisor of $a$ and $b$ in a
  PID $R$, then $d$ is a linear combination of $a$ and $b$:
  $\exists r, s \in R$ such that $d = ra + sb$.
\item If $I$ is a nonzero ideal in a PID, then $I$ is prime if and
  only if it is maximal.
\end{itemize}
We now add one important point to this list:

\begin{prop}{}{prop:irred:2.6}
  If $R$ is a PID, then it is a UFD.
\end{prop}

\begin{proof}
  Let $R$ be a PID. The a.c.c.\ (for principal ideals, as all ideals
  in $R$ are principal!) holds in $R$ since PIDs are Noetherian. We
  verify that irreducible elements are prime in $R$, which implies
  that $R$ is a UFD by \cref{thm:irred:2.5}.

  Let $a \in R$ be an irreducible element. Ideals generated by
  irreducible elements are maximal among principal ideals
  (\Cref{exer:irred:irred-iff-a-maximal-principal}), hence $(a)$ is a
  maximal ideal in $R$ as all ideals in $R$ are principal. Since
  maximal ideals are prime it follows that $(a)$ is prime, as needed.
\end{proof}

In particular, $k[x]$ is a UFD if $k$ is a field, and $\Z$ is a
UFD---which hopefully will not come as a big surprise to our
readers\footnote{This fact is known as the fundamental theorem of
  arithmetic.}. Note that at this point we have recovered the fact
stated in \Cref{exer:irred:prime-iff-irred-in-z} in full glory.

\cref{prop:irred:2.6} justifies another feature of the picture given
at the beginning of this chapter: the class of PIDs is contained in
the class of UFDs. We will soon see that the inclusion is proper, that
is, that there are UFDs which are not PIDs; for example, $\Z[x]$ is
not a PID (\Cref{exer:irred:rx-pid-implies-r-field}), yet $\Z[x]$ is a
unique factorization domain as we will soon be able to prove
(\cref{thm:irred:4.14}). In fact, there are UFDs which are not
Noetherian, as represented in the picture; however, these examples are
best discussed after more material has been developed
(\cref{exam:irred:4.18}).

The reader can already get a feel for the gap between UFD and PID, by
contemplating the fact (recalled above and essentially tautological)
that, in a PID, greatest common divisors of $a$, $b$ are linear
combinations of $a$ and $b$. This is a very strong requirement: for
example, it characterizes PIDs among Noetherian domains, as the reader
will check (\Cref{exer:irred:noeth-gcd-lin-comb-implies-pid}); it does
not hold in general UFDs (\Cref{exer:irred:gcd-not-lin-comb-in-zxy}).

\subsection{\texorpdfstring{Euclidean domain $\implies$ PID}{Euclidean
    domain implies PID}}
\label{subsec:irred:euclidean-domain-implies-pid}
The excellent properties of $\Z$ and $k[x]$ (where $k$ is a field)
make these rings even more special than PIDs: they are \emph{Euclidean
  domains}. Informally, Euclidean domains are rings in which one may
perform a `division with remainder': this is the case for $\Z$ and for
$k[x]$, as observed
in~\S{}\ref{sec:rings:ideals-and-quotients-remarks-and-examples-prime-and}. The
point is that in both $\Z$ and $k[x]$ one can define a notion of
`size' of an element: $|n|$ for an integer $n$ and $\deg f(x)$ for a
polynomial $f(x)$. In both cases, one has control over the size of the
`remainder' in a division. The definition of Euclidean domain simply
abstracts this mechanism.

For the purpose of this discussion\footnote{Entire libraries have been
  written on the subject of valuations, studying a more precise notion
  than what is needed here.}, a \emph{valuation} on an integral domain
$R$ is any function $v : R \setminus \{0\} \to \Z^{\geq 0}$.

\begin{defn}{}{defn:irred:2.7}
  A \emph{Euclidean valuation} on an integral domain $R$ is a
  valuation satisfying the following property\footnote{It is not
    uncommon to also require that $v(ab) \geq v(b)$ for all nonzero
    $a, b \in R$; but this is not needed in the considerations that
    follow, and cf.\
    \Cref{exer:irred:euclidean-valuation-monotone-variant}.}: for all
  $a \in R$ and all nonzero $b \in R$ there exist $q, r \in R$ such
  that $a = qb + r$, with either $r = 0$ or $v(r) < v(b)$. An integral
  domain $R$ is a \emph{Euclidean domain} if it admits a Euclidean
  valuation.
\end{defn}

We say that $q$ is the \emph{quotient} of the division and $r$ is the
\emph{remainder}. Division with remainder in $\Z$ and in $k[x]$ (where
$k$ is a field) provide examples, so that $\Z$ and $k[x]$ are
Euclidean domains.
\begin{prop}{}{prop:irred:2.8}
  Let $R$ be a Euclidean domain. Then $R$ is a PID.
\end{prop}

The proof is modeled after the instances encountered for $\Z$
(\Cref{prop:rings:Z-PID}) and $k[x]$ (which the reader has hopefully
worked out in \Cref{exer:rings:field-poly-ring-pid}).

\begin{proof}
  Let $I$ be an ideal of $R$; we have to prove that $I$ is
  principal. If $I = \{0\}$, there is nothing to show; therefore,
  assume $I \neq \{0\}$.  The valuation maps the nonzero elements of
  $I$ to a subset of $\Z^{\geq 0}$; let $b \in I$ be an element with
  the smallest valuation. Then I claim that $I = (b)$; therefore $I$
  is principal, as needed. Since clearly $(b) \subseteq I$, we only
  need to verify that $I \subseteq (b)$.

  For this, let $a \in I$ and apply division with remainder: we have
  \[a = qb + r\] for some $q, r \in R$, with $r = 0$ or $v(r) <
  v(b)$. But $r = a - qb \in I$: by the minimality of $v(b)$ among
  nonzero elements of $I$, we cannot have $v(r) < v(b)$. Therefore
  $r = 0$, showing that $a = qb \in (b)$, as needed.
\end{proof}
\Cref{prop:irred:2.8} justifies one more feature of the picture at the
beginning of the chapter: the class of Euclidean domains is contained
in the class of principal ideal domains.

This inclusion is proper, as suggested in the picture. Producing an
explicit example of a PID which is not a Euclidean domain is not so
easy, but the gap between PIDs and Euclidean domains can in fact be
described very sharply: PIDs may be characterized as domains
satisfying a weaker requirement than `division with remainder'. More
precisely, a \emph{Dedekind--Hasse valuation} is a valuation $v$ such
that for all $a$, $b$, either $(a,b) = (b)$ (that is, $b$ divides $a$)
or there exists $r \in (a,b)$ such that $v(r) < v(b)$. This latter
condition amounts to requiring that there exist $q, s \in R$ such that
$as = bq + r$ with $v(r) < v(b)$; hence a Euclidean valuation (for
which we may in fact choose $s = 1$) is a Dedekind--Hasse
valuation. It is not hard to show that an integral domain is a PID if
and only if it admits a Dedekind--Hasse valuation
(\Cref{exer:irred:pid-iff-dedekind-hasse-valuation}).

For example, this can be used to show that the ring
$\Z[\frac{1+\sqrt{-19}}{2}]$ is a PID: the norm considered in
\Cref{exer:irred:z-1-plus-sqrt-neg19-over-2-not-euclidean} in order to
prove that this ring is not a Euclidean domain turns out to be a
Dedekind--Hasse valuation\footnote{This boils down to a case-by-case
  analysis, which I am happily leaving to my most patient
  readers.}. Thus, this ring gives an example of a PID that is not a
Euclidean domain.  One excellent feature of Euclidean domains, and the
one giving them their names, is the presence of an effective algorithm
computing greatest common divisors: the \emph{Euclidean algorithm}. As
Euclidean domains are PIDs, and hence UFDs, we know that they do have
greatest common divisors.  However, the `algorithm' obtained by
distilling the proof of \Cref{lem:irred:2.3} is highly impractical: if
we had to factor two integers $a$, $b$ in order to compute their gcd,
this would make it essentially impossible (with current technologies
and factorization algorithms) for integers of a few hundred
digits. The Euclidean algorithm bypasses the necessity of
factorization: greatest common divisors of thousand-digit integers may
be computed in a fraction of a second.

The key lemma on which the algorithm is based is the following trivial
general fact:
\begin{lem}{}{lem:irred:2.9}
  Let $a = bq + r$ in a ring $R$. Then $(a, b) = (b, r)$.
\end{lem}

\begin{proof}
  Indeed, $r = a - bq \in (a, b)$, proving $(b, r) \subseteq (a, b)$;
  and $a = bq + r \in (b, r)$, proving $(a, b) \subseteq (b, r)$.
\end{proof}
In particular, $(\forall c \in R)$,
\[(a, b) \subseteq (c) \iff (b, r) \subseteq (c);\] that is, the set
of common divisors of $a$, $b$ and the set of common divisors of $b$,
$r$ coincide. Therefore,

\begin{coro}{}{coro:irred:2.10}
  Assume $a = bq + r$. Then $a$, $b$ have a gcd if and only if $b$,
  $r$ have a gcd, and in this case $\gcd(a, b) = \gcd(b, r)$.
\end{coro}

Of course `$\gcd(a, b) = \gcd(b, r)$' means that the two classes of
associate elements coincide.  These considerations hold over any
integral domain; assume now that $R$ is a Euclidean domain. Then we
can use division with remainder to gain some control over the
remainders $r$. Given two elements $a$, $b$ in $R$, with $b \neq 0$,
we can apply division with remainder repeatedly:
\begin{align*}
  a &= bq_1 + r_1, \\
  b &= r_1 q_2 + r_2, \\
  r_1 &= r_2 q_3 + r_3, \\
    &\cdots
\end{align*}
as long as the remainder $r_i$ is nonzero.

\begin{fact}{Claim~2.11}{fact:irred:claim:2.11}
  This process terminates: that is, $r_N = 0$ for some $N$.
\end{fact}

\begin{proof}
  Each line in the table is a division with remainder. If no $r_i$
  were zero, we would have an infinite decreasing sequence
  $v(b) > v(r_1) > v(r_2) > v(r_3) > \dots$ of nonnegative integers,
  which is nonsense.
\end{proof}
Thus the table of divisions with remainders must be as follows:
letting $r_0 = b$,
\begin{align*}
  a &= r_0 q_1 + r_1, \\
  b &= r_1 q_2 + r_2, \\
  r_1 &= r_2 q_3 + r_3, \\
    &\cdots \\
  r_{N-3} &= r_{N-2} q_{N-1} + r_{N-1}, \\
  r_{N-2} &= r_{N-1} q_N
\end{align*}
with $r_{N-1} \neq 0$.

\begin{prop}{}{prop:irred:2.12}
  With notation as above, $r_{N-1}$ is a gcd of $a$, $b$.
\end{prop}

\begin{proof}
  By \Cref{coro:irred:2.10},
  \[\gcd(a, b) = \gcd(b, r_1) = \gcd(r_1, r_2) = \dots =
    \gcd(r_{N-2}, r_{N-1}).\] But $r_{N-2} = r_{N-1} q_N$ gives
  $r_{N-2} \in (r_{N-1})$; hence $(r_{N-2}, r_{N-1}) =
  (r_{N-1})$. Therefore $r_{N-1}$ is a gcd for $r_{N-2}$ and
  $r_{N-1}$, hence for $a$ and $b$, as needed.
\end{proof}
The ring of integers and the polynomial ring over a field are both
Euclidean domains. Fields are Euclidean domains (as represented in the
picture at the beginning of the chapter), but not for a very
interesting reason: the remainder of the division by a nonzero element
in a field is always zero, so every function qualifies as a `Euclidean
valuation' for trivial reasons. We will study another interesting
Euclidean domain later in this chapter
(\S{}\ref{subsec:irred:gaussian-integers}).

\subsection*{Exercises}

\begin{exer}{}{exer:irred:divisibility-via-irred-factors-in-ufd}
  \exerused{} Prove
  \Cref{lem:irred:2.1}. [\S{}\ref{subsec:irred:irreducible-factors-and-greatest-common-divisor}]
\end{exer}

\begin{exer}{}{exer:irred:euclids-lemma-in-ufd}
  Let $R$ be a UFD, and let $a$, $b$, $c$ be elements of $R$ such that
  $a \mid bc$ and $\gcd(a, b) = 1$. Prove that $a \mid c$.
\end{exer}

\begin{exer}{}{exer:irred:factors-of-n-vs-comp-factors-of-znz}
  Let $n$ be a positive integer. Prove that there is a one-to-one
  correspondence preserving multiplicities between the irreducible
  factors of $n$ (as an integer) and the composition factors of
  $\Z/n\Z$ (as a group). (In fact, the Jordan--H\"older theorem may be
  used to prove that $\Z$ is a UFD.)
\end{exer}
\begin{exer}{}{exer:irred:gcd-not-lin-comb-in-zxy}
  \exerused{} Consider the elements $x$, $y$ in $\Z[x, y]$. Prove that
  $1$ is a gcd of $x$ and $y$, and yet $1$ is not a linear combination
  of $x$ and $y$. (Cf.\ \Cref{exer:groups:bezout}.)
  [\S{}\ref{subsec:irred:irreducible-factors-and-greatest-common-divisor},
  \S{}\ref{subsec:irred:pid-implies-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:no-gcd-example-in-subring-of-zt}
  \exerused{} Let $R$ be the subring of $\Z[t]$ consisting of
  polynomials with no term of degree~$1$:
  $a_0 + a_2 t^2 + \dots + a_d t^d$.
  \begin{itemize}
  \item Prove that $R$ is indeed a subring of $\Z[t]$, and conclude
    that $R$ is an integral domain.
  \item List all common divisors of $t^5$ and $t^6$ in $R$.
  \item Prove that $t^5$ and $t^6$ have no gcd in $R$.
  \end{itemize}
  [\S{}\ref{subsec:irred:irreducible-factors-and-greatest-common-divisor}]
\end{exer}
\begin{exer}{}{exer:irred:principal-intersections-implies-gcd}
  Let $R$ be a domain with the property that the intersection of any
  family of principal ideals in $R$ is necessarily a principal ideal.
  \begin{itemize}
  \item Show that greatest common divisors exist in $R$.
  \item Show that UFDs satisfy this property.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:noeth-gcd-lin-comb-implies-pid}
  \exerused{} Let $R$ be a Noetherian domain, and assume that for all
  nonzero $a$, $b$ in $R$, the greatest common divisors of $a$ and $b$
  are linear combinations of $a$ and $b$. Prove that $R$ is a
  PID. [\S{}\ref{subsec:irred:pid-implies-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:ufd-desc-chain-principal-stabilizes}
  Let $R$ be a UFD, and let $I \neq (0)$ be an ideal of $R$. Prove
  that every descending chain of principal ideals containing $I$ must
  stabilize.
\end{exer}
\begin{exer}{}{exer:irred:ufd-height-1-prime-principal}
  $\lnot$ The \emph{height} of a prime ideal $P$ in a ring $R$ is (if
  finite) the maximum length $h$ of a chain of prime ideals
  $P_0 \subsetneq P_1 \subsetneq \dots \subsetneq P_h = P$ in
  $R$. (Thus, the Krull dimension of $R$, if finite, is the maximum
  height of a prime ideal in $R$.) Prove that if $R$ is a UFD, then
  every prime ideal of height~$1$ in $R$ is
  principal. [\ref{exer:irred:noeth-domain-ufd-iff-height-1-principal}]
\end{exer}
\begin{exer}{}{exer:irred:noeth-domain-ufd-iff-height-1-principal}
  $\lnot$ It is a consequence of a theorem known as Krull's
  Hauptidealsatz that every nonzero, nonunit element in a Noetherian
  domain is contained in a prime ideal of height~$1$. Assuming this,
  prove a converse to \Cref{exer:irred:ufd-height-1-prime-principal},
  and conclude that a Noetherian domain $R$ is a UFD if and only if
  every prime ideal of height~$1$ in $R$ is
  principal. [\ref{exer:irred:ufd-iff-localization-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:pid-mod-ideal-is-artinian}
  Let $R$ be a PID, and let $I$ be a nonzero ideal of $R$. Show that
  $R/I$ is an artinian ring (cf.\ \Cref{exer:irred:artinian-basics}),
  by proving explicitly that the d.c.c.\ holds in $R/I$.
\end{exer}
\begin{exer}{}{exer:irred:rx-pid-implies-r-field}
  \exerused{} Prove that if $R[x]$ is a PID, then $R$ is a field.
  [\S{}\ref{subsec:irred:pid-implies-ufd},
  \S{}\ref{subsec:linalg:linear-transformations-of-free-modules-actions-of-polynomial-rings}]
\end{exer}
\begin{exer}{}{exer:irred:power-minus-1-divisibility}
  \exerused{} For $a$, $b$, $c$ positive integers with $c > 1$, prove
  that $c^a - 1$ divides $c^b - 1$ if and only if $a \mid b$. Prove
  that $x^a - 1$ divides $x^b - 1$ in $\Z[x]$ if and only if
  $a \mid b$.  (Hint: For the interesting implications, write
  $b = ad + r$ with $0 \leq r < a$, and take `size' into account.)
  [\S{}\ref{subsec:fields:finite-fields},
  \ref{exer:fields:phi-n-a-divides-quotient}]
\end{exer}
\begin{exer}{}{exer:irred:formal-power-series-euclidean}
  \exerused{} Prove that if $k$ is a field, then $k[[x]]$ is a
  Euclidean domain. [\S{}\ref{subsec:irred:r-ufd-implies-r-x-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:euclidean-valuation-monotone-variant}
  \exerused{} Prove that if $R$ is a Euclidean domain, then $R$ admits
  a Euclidean valuation $v$ such that $v(ab) \geq v(b)$ for all
  nonzero $a, b \in R$. (Hint: Since $R$ is a Euclidean domain, it
  admits a valuation $v$ as in \Cref{defn:irred:2.7}. For $a \neq 0$,
  let $\bar{v}(a)$ be the minimum of all $v(ab)$ as $b \in R$,
  $b \neq 0$. To see that $R$ is a Euclidean domain with respect to
  $\bar{v}$ as well, let $a$, $b$ be nonzero in $R$, with $b \nmid a$;
  choose $q$, $r$ so that $a = bq + r$, with $\bar{v}(r)$ minimal;
  assume that $\bar{v}(r) \geq \bar{v}(b)$, and get a contradiction.)
  [\S{}\ref{subsec:irred:euclidean-domain-implies-pid},
  \ref{exer:irred:euclidean-valuation-associates-units}]
\end{exer}
\begin{exer}{}{exer:irred:euclidean-valuation-associates-units}
  Let $R$ be a Euclidean domain with Euclidean valuation $v$; assume
  that $v(ab) \geq v(b)$ for all nonzero $a, b \in R$ (cf.\
  \Cref{exer:irred:euclidean-valuation-monotone-variant}).  Prove that
  associate elements have the same valuation and that units have
  minimum valuation.
\end{exer}
\begin{exer}{}{exer:irred:euclidean-domain-remainder-unit-or-zero}
  $\lnot$ Let $R$ be a Euclidean domain that is not a field. Prove
  that there exists a nonzero, nonunit element $c$ in $R$ such that
  $\forall a \in R$, $\exists\, q, r \in R$ with $a = qc + r$ and
  either $r = 0$ or $r$ a
  unit. [\ref{exer:irred:z-1-plus-sqrt-neg19-over-2-not-euclidean}]
\end{exer}
\begin{exer}{}{exer:irred:z-1-plus-sqrt-neg19-over-2-not-euclidean}
  \exerused{} For an integer $d$, denote by $\Q(\sqrt{d})$ the
  smallest subfield of $\C$ containing $\Q$ and $\sqrt{d}$, with norm
  $N$ defined as in \Cref{exer:rings:z-sqrt-d-subring-of-c}. See
  \Cref{exer:irred:z-sqrt-neg5-not-ufd} for the case $d = -5$; in this
  problem, you will take $d = -19$.

  Let $\delta = \frac{1 + \sqrt{-19}}{2}$, and consider the following
  subring of $\Q(\sqrt{-19})$:
  \[\Z[\delta] := \left\{ a + b\delta \mid a, b \in \Z
    \right\}.\]
  \begin{itemize}
  \item Prove that the smallest values of $N(z)$ for
    $z = a + b\delta \in \Z[\delta]$ are $0, 1, 4, 5$. Prove that
    $N(a + b\delta) \geq 5$ if $b \neq 0$.
  \item Prove that the units in $\Z[\delta]$ are $\pm 1$.
  \item If $c \in \Z[\delta]$ satisfies the condition specified in
    \Cref{exer:irred:euclidean-domain-remainder-unit-or-zero}, prove
    that $c$ must divide $2$ or $3$ in $\Z[\delta]$, and conclude that
    $c = \pm 2$ or $c = \pm 3$.
  \item Now show that $\nexists\, q \in \Z[\delta]$ such that
    $\delta = qc + r$ with $c = \pm 2, \pm 3$ and $r = 0, \pm 1$.
  \end{itemize}
  Conclude that $\Z[\frac{1+\sqrt{-19}}{2}]$ is not a Euclidean
  domain. [\S{}\ref{subsec:irred:euclidean-domain-implies-pid},
  \ref{exer:irred:z-sqrt-neg2-and-imaginary-quad-euclidean}]
\end{exer}
\begin{exer}{}{exer:irred:discrete-valuation-ring-basics}
  $\lnot$ A \emph{discrete valuation} on a field $k$ is a surjective
  homomorphism of abelian groups $v : (k^*, \cdot) \to (\Z, +)$ such
  that $v(a + b) \geq \min(v(a), v(b))$ for all $a, b \in k^*$ such
  that $a + b \in k^*$.
  \begin{itemize}
  \item Prove that the set
    $R := \{a \in k^* \mid v(a) \geq 0\} \cup \{0\}$ is a subring of
    $k$.
  \item Prove that $R$ is a Euclidean domain.
  \end{itemize}
  Rings arising in this fashion are called \emph{discrete valuation
    rings}, abbreviated DVR. They arise naturally in number theory and
  algebraic geometry. Note that the Krull dimension of a DVR is~$1$
  (\Cref{exam:rings:speck-prime-ideals-maximal}); in algebraic
  geometry, DVRs correspond to particularly nice points on a `curve'.
  \begin{itemize}
  \item Prove that the ring of rational numbers $a/b$ with $b$ not
    divisible by a fixed prime integer $p$ is a DVR.
  \end{itemize} [\ref{exer:irred:dvr-is-pid-local-parameter},
  \ref{exer:linrep:p-adic-integers-basics}]
\end{exer}
\begin{exer}{}{exer:irred:dvr-is-pid-local-parameter}
  $\lnot$ As seen in \Cref{exer:irred:discrete-valuation-ring-basics},
  DVRs are Euclidean domains. In particular, they must be PIDs. Check
  this directly, as follows. Let $R$ be a DVR, and let $t \in R$ be an
  element such that $v(t) = 1$. Prove that if $I \subseteq R$ is any
  nonzero ideal, then $I = (t^k)$ for some $k \geq 1$. (The element
  $t$ is called a `local parameter' of $R$.)
  [\ref{exer:irred:dvr-is-local-pid-localization-is-dvr},
  \ref{exer:fields:two-curves-c1-c2-comparison}]
\end{exer}
\begin{exer}{}{exer:irred:pid-iff-dedekind-hasse-valuation}
  \exerused{} Prove that an integral domain is a PID if and only if it
  admits a Dedekind--Hasse valuation. (Hint: For the $\Leftarrow$
  implication, adapt the argument in \Cref{prop:irred:2.8}; for
  $\Rightarrow$, let $v(a)$ be the size of the multiset of irreducible
  factors of $a$.)
  [\S{}\ref{subsec:irred:euclidean-domain-implies-pid}]
\end{exer}
\begin{exer}{}{exer:irred:gcd-in-pid-remains-gcd-in-extension}
  $\lnot$ Suppose $R \subseteq S$ is an inclusion of integral domains,
  and assume that $R$ is a PID. Let $a, b \in R$, and let $d \in R$ be
  a gcd for $a$ and $b$ in $R$. Prove that $d$ is also a gcd for $a$
  and $b$ in~$S$. [\ref{exer:irred:irred-poly-no-multiple-roots}]
\end{exer}
\begin{exer}{}{exer:irred:compute-gcd-example}
  Compute $d = \gcd(5504227617645696, 2922476045110123)$. Further,
  find $a$, $b$ such that
  $d = 5504227617645696\, a + 2922476045110123\, b$.
\end{exer}

\begin{exer}{}{exer:irred:euclid-infinitude-of-primes}
  Prove that there are infinitely many prime integers. (Hint: Assume
  by contradiction that $p_1, \dots, p_N$ is a complete list of all
  positive prime integers. What can you say about
  $p_1 \cdots p_N + 1$? This argument was already known to Euclid,
  more than 2,000 years ago.)
  [\ref{exer:irred:euclid-variation-poly-values},
  \S{}\ref{subsec:irred:adding-roots-algebraically-closed-fields},
  \ref{exer:irred:finite-field-irreds-arb-high-degree}]
\end{exer}

\begin{exer}{}{exer:irred:euclid-variation-poly-values}
  $\lnot$ Variation on the theme of Euclid from
  \Cref{exer:irred:euclid-infinitude-of-primes}: Let $f(x) \in \Z[x]$
  be a nonconstant polynomial such that $f(0) = 1$.  Prove that
  infinitely many primes divide the numbers $f(n)$, as $n$ ranges in
  $\Z$. (If $p_1, \dots, p_N$ were a complete list of primes dividing
  the numbers $f(n)$, what could you say about $f(p_1 \cdots p_N x)$?)
  Once you are happy with this, show that the hypothesis $f(0) = 1$ is
  unnecessary. (If $f(0) = a \neq 0$, consider
  $f(p_1 \cdots p_N a x)$. Finally, note that there is nothing special
  about $0$.) [\ref{exer:fields:infinitely-many-primes-cong-1-mod-n}]
\end{exer}

\section{Intermezzo: Zorn's lemma}%
\label{sec:irred:intermezzo-zorn-s-lemma}

\subsection{Set theory, reprise}
\label{subsec:irred:set-theory-reprise}
We leave ring theory for a moment and take a little detour to
contemplate an issue from set theory. As remarked at the very outset,
only naive set theory is used in this book; all set-theoretic
operations we have used so far are nothing more than a formalization
of intuitive ideas regarding collections of objects. However, I will
occasionally need to refer to a less `intuitively obvious'
set-theoretic statement: for example, this statement is needed in
order to show that every proper ideal in a ring is contained in a
maximal ideal (\Cref{prop:irred:3.5}). This set-theoretic fact is
\emph{Zorn's lemma}. An \emph{order relation} on a set $Z$ is a
relation $\preceq$ which is reflexive, transitive, and antisymmetric:
the first two terms are familiar to the reader, and the third means
that
\[(\forall\, a, b \in Z),\quad a \preceq b \text{ and } b \preceq a
  \implies a = b.\] Typical prototypes are the $\leq$ relation on $\Z$
or the inclusion relation $\subseteq$ among subsets of a given set. We
use $a \prec b$ to denote $a \preceq b$ and $b \neq a$.  A pair
$(Z, \preceq)$, consisting of a set $Z$ and an order relation
$\preceq$ on $Z$, is called a \emph{poset}, for \emph{partially
  ordered set}. The qualifier `partially' is not necessary, but
convenient, as it reminds us that if $a, b \in Z$, then it is not
necessarily the case that $a \preceq b$ or $b \preceq a$: for example,
$\subseteq$ does not satisfy this additional requirement in general
(while $(\Z, \leq)$ does).  An order is \emph{total} if it does
satisfy this additional requirement. A totally ordered set is not
called a `toset', as would seem reasonable, but rather a \emph{chain}.

An element $m$ of a poset $Z$ is \emph{maximal} if nothing comes
properly `after' it in the order:
\[(\forall\, a \in Z),\quad m \preceq a \implies m = a.\] For example,
maximal ideals in a ring are maximal elements in the set of proper
ideals, that is, ideals $\neq (1)$ (by
\Cref{prop:rings:prime-maximal-equiv}). An \emph{upper bound} for a
subset $S$ of a poset $Z$ is an element $u \in Z$ coming after every
element of $S$:
\[(\forall\, a \in S),\quad a \preceq u.\] Notions of `smallest' (or,
rather, `least') or `largest' are defined in the evident way.

Posets may or may not have maximal elements, upper bounds, etc.: for
example, the family of finite subsets of an infinite set does not have
maximal elements. All this terminology comes together in the following
statement:
\begin{lem}{Zorn's lemma}{lem:irred:3.1}
  Let $Z$ be a nonempty poset. Assume that every chain in $Z$ has an
  upper bound in $Z$; then there exists a maximal element in $Z$.
\end{lem}

The status of Zorn's lemma is peculiar: on the one hand, it is complex
enough that no one I know of finds it `intuitively clear'; on the
other hand, it turns out to be logically
equivalent\footnote{Equivalent in the sense that each can be derived
  from the other together with the other axioms of
  `Zermelo--Fr\"ankel' set theory.} to the axiom of choice, which does
look reasonable to most people, and to the `well-ordering theorem',
which I find intuitively unreasonable. I will not prove all these
equivalences (the diligent reader will prove them in the exercises and
should in any case have no trouble locating more detailed proofs); but
I will attempt to describe the situation in general terms and explain
why the unreasonable statement implies the other two.  The \emph{axiom
  of choice} states that if $\mathscr{F}$ is a family of disjoint
nonempty subsets of a set $Z$, then we can form a new set by selecting
one element $x$ from each $X \in \mathscr{F}$. This may sound
reasonable, but it raises rather subtle points: for example, it can be
shown that this axiom is independent of the other axioms of
(Zermelo--Fr\"ankel) set theory. Also, it has disturbingly
counterintuitive consequences, such as the Banach--Tarski
paradox\footnote{One can subdivide a solid ball of radius~$1$ into
  finitely many pieces, then reassemble them after rotations and
  translations in $3$-space and obtain two balls of radius~$1$.}.

The subtlety of the axiom of choice boils down to the following
question: how does one choose the element $x$? This is not
controversial if $Z$ (and hence $\mathscr{F}$) is finite, but, not
surprisingly, it becomes an issue if $Z$ is infinite. A suitable order
relation on $Z$ would come in handy here: if $\preceq$ is an order
relation on $Z$ such that every nonempty subset of $Z$ has a least
element, then we could simply let $x$ be the least element of $X$ for
each $X \in \mathscr{F}$. We say that $Z$ is \emph{well-ordered} by
$\preceq$, or that $\preceq$ is a \emph{well-ordering} on $Z$, if this
is the case. (The abbreviation \emph{woset} is also used in the
literature, but not very often.) For example, the set $\Z_{>0}$ of
positive numbers is well-ordered by $\leq$; this fact is called the
\emph{well-ordering principle}.

Thus, the statement of the axiom of choice is really `clear' if the
set $Z$ is the set $\Z_{>0}$ of positive numbers. It may seem somewhat
less so for the set $\Z$, since $\Z$ is not well-ordered by $\leq$;
however, it takes a moment
(\Cref{exer:irred:explicit-well-ordering-on-z}) to construct a
different relation on $\Z$, making $\Z$ a well-ordered set. Thus, the
axiom of choice is also completely transparent for $Z = \Z$ or any
countable set (like $\Q$) for that matter.

The well-ordering principle is at the basis of proofs by induction: in
fact (cf.\ \Cref{exer:irred:induction-equiv-to-well-ordering-z-pos})
it is equivalent to the so-called principle of induction. More is
true, however. If $(Z, \preceq)$ is any woset, then we can consider
the following `principle of induction' on $Z$: Let $S \subseteq Z$ be
a subset such that
\[(\forall\, a \in Z),\quad \bigl((\forall\, b \in Z),\ b \prec a
  \implies b \in S\bigr) \implies a \in S;\] then $S = Z$. That is,
$Z$ is the only set $S$ with the property that $a \in S$ if all
$b \prec a$ are in $S$. (Note that the least element $a$ of $Z$ is
automatically in $S$, since the condition on all $b \prec a$ is
vacuously true in this case.) The reader will recognize that for
$Z = \Z_{>0}$ with the relation $\leq$, this is the ordinary principle
of induction.

\begin{fact}{Claim~3.2}{fact:irred:claim:3.2}
  Let $Z$ be any woset. Then the principle of induction holds for $Z$.
\end{fact}

\begin{proof}
  Let $S \subseteq Z$ be a subset with the property given above, and
  assume $S \subsetneq Z$. Then the complement $T$ of $S$ in $Z$ is
  nonempty; hence it has a least element $a$. Now if $b \prec a$, then
  necessarily $b \in S$ (otherwise $b \in T$, contradicting the fact
  that $a$ is the least element in $T$). But the property of $S$ then
  forces $a \in S$, contradiction. Therefore $T$ is empty; i.e.,
  $S = Z$.
\end{proof}

This is remarkable, since it extends induction to uncountable sets,
provided a well-ordering is available.  For example, if we had a
well-ordering on $\R$, then we could prove a statement about all reals
`by induction'. Many people find this somewhat counterintuitive, and
hence so seems the following amazing claim:

\begin{thm}{Well-ordering theorem}{thm:irred:3.3}
  Every set admits a well-ordering.
\end{thm}

As argued above, this implies that the axiom of choice holds on every
set: if you accept the well-ordering theorem, the statement of the
axiom of choice becomes as evident for every set as it is for the set
of positive integers. The catch is of course that the axiom of choice
is used in the proof of the well-ordering theorem; in fact, the
statements are equivalent.

In any case, the statement of the well-ordering theorem is easy to
absorb (even if perhaps less `intuitively clear' than the axiom of
choice). The well-ordering theorem is equivalent to Zorn's lemma,
since they are both equivalent to the axiom of choice. The good news
is that the derivation of Zorn's lemma from the well-ordering theorem
is reasonably straightforward:
\begin{proof}[Proof (Well-ordering theorem $\implies$ Zorn's lemma)]
  Let $(Z, \leq)$ be a nonempty poset such that every chain in $Z$ has
  an upper bound in $Z$. By the well-ordering theorem, there is a
  well-ordering\footnote{Watch out: we are considering two orderings
    on $Z$: $\leq$ (about which we want to say something) and
    $\preceq$ (about which we know something already).} $\preceq$ on
  $Z$. Define a function $f$ from $Z$ to the power set of $Z$ as
  follows\footnote{Even beyond; in the context of ordinals this
    induction principle is called \emph{transfinite induction}.}:
  \[f(a) = \begin{cases} \{a\} \cup \bigcup_{b \prec a} f(b) &
      \text{if } \bigl(\{a\} \cup {\textstyle\bigcup_{b \prec a}}
      f(b)\bigr)
                                                               \text{ is totally ordered by } \leq; \\[4pt]
    \emptyset & \text{otherwise.}
  \end{cases}\]
The fact that $Z$ is well-ordered by $\preceq$ implies that $f$ is defined
for every $a \in Z$. Indeed, let $T$ be the set of elements of $Z$ for
which $f$ is not defined; if $T \neq \emptyset$, $T$ has a least element
$a$ w.r.t.\ $\preceq$; but then $f(b)$ is defined for all $b \prec a$,
and the prescription given for $f(a)$ defines $f$ at $a$, a
contradiction.

Let $S = \bigcup_{a \in Z} f(a)$. It is clear that $S$ is totally
ordered by $\leq$: if $a, b \in S$, and (say) $b \prec a$, then both
$a$ and $b$ belong to $\{a\} \cup \bigcup_{b' \prec a} f(b')$, which
is totally ordered by $\leq$ by construction.

I claim that $S$ is a maximal totally ordered subset\footnote{The
  existence of maximal totally ordered subsets is known as the
  \emph{Hausdorff maximal principle}; it is also equivalent to the
  axiom of choice.} of $Z$:

\begin{fact}{Claim~3.4}{fact:irred:claim:3.4}
  If $S \subseteq S' \subseteq Z$ and $S'$ is totally ordered by
  $\leq$, then $S = S'$.
\end{fact}

Indeed, let $T$ be the complement of $S$ in $S'$. If $T$ is nonempty,
let $a \in T$ and observe that $\{a\} \cup \bigcup_{b \prec a} f(b)$
is totally ordered by $\leq$, since it is a subset of
$\{a\} \cup S \subseteq S'$. But then $f(a) = \{a\}$, that is,
$a \in S$, a contradiction.

Since $S$ is a chain, by the hypothesis of Zorn's lemma $S$ has an
upper bound $m$. It is now clear that $m$ must be maximal in $Z$
w.r.t.\ $\leq$, verifying the statement of Zorn's lemma. Indeed, if
$m' \geq m$, then $S \cup \{m'\}$ is totally ordered; hence
$S = S \cup \{m'\}$ by the claim. This means $m' \in S$, and hence
$m' = m$ since $m$ is an upper bound for $S$.
\end{proof}

Zorn's lemma is the key to several basic results in algebra, the first
of which we are about to encounter; the reader will sample a few more
results in the exercises. The reader will also encounter equally basic
applications of Zorn's lemma (or of other manifestations of the axiom
of choice) in other fields: for example Tychonoff's theorem in
topology and the Hahn--Banach theorem in functional analysis.

\subsection{Application: Existence of maximal ideals}
\label{subsec:irred:application-existence-of-maximal-ideals}
Recall (\S{}\ref{subsec:rings:prime-and-maximal-ideals}) that an ideal
$\mathfrak{m}$ of a ring $R$ is maximal if and only if
$R/\mathfrak{m}$ is a field, if and only if no other ideal stands
between $\mathfrak{m}$ and $R = (1)$, that is, if and only if
$\mathfrak{m}$ is maximal with respect to inclusion, in the family of
proper ideals of $R$. It is not obvious from this definition that
maximal ideals exist, but they do:

\begin{prop}{}{prop:irred:3.5}
  Let $I \neq (1)$ be a proper ideal of a commutative ring $R$. Then
  there exists a maximal ideal $\mathfrak{m}$ of $R$ containing $I$.
\end{prop}

This is immediate if $R$ is Noetherian, by applying condition~(3) from
\Cref{prop:irred:1.1} to the family of proper ideals of $R$ containing
$I$.  The argument for arbitrary rings is a classic application of
Zorn's lemma.  In fact, this is the first application given by M.~Zorn
in his 1935 article introducing a `maximum principle' (now known as
Zorn's lemma).  The result had earlier been proven by Krull, using the
well-ordering theorem.

\begin{proof}
  The set $\mathscr{I}$ of proper ideals of $R$ containing $I$ is
  ordered by inclusion. Then let $C$ be a chain of proper ideals, and
  consider
  \[U := \bigcup_{J \in C} J.\] I claim that $U$ is a proper ideal
  containing $I$; hence it is an upper bound for $C$ in
  $\mathscr{I}$. This proves that every chain in $\mathscr{I}$ has an
  upper bound, and it follows that $\mathscr{I}$ has maximal elements,
  by Zorn's lemma.

  To verify my claim, it is clear that $U$ contains $I$ and that it is
  an ideal (for example, if $a, b \in U$, then $\exists\, J \in C$
  such that $a, b \in J$; hence $a \pm b \in J$, and therefore
  $a \pm b \in U$). We have to check that $U$ is proper. But if
  $U = (1)$, then $1 \in J$ for some $J \in \mathscr{I}$,
  contradicting the fact that $\mathscr{I}$ consists of proper ideals.
\end{proof}

Note that this argument relies crucially on the fact that $R$ has a
multiplicative unit $1$, and indeed the stated fact is not true in
general for rings without~$1$
(\Cref{exer:irred:trivial-mult-rng-on-q-no-maximal-ideals}). Also, the
use of Zorn's lemma is not just a convenient trick; the statement is
known to be equivalent to the axiom of choice, by work of W.~Hodges.

\subsection*{Exercises}
\begin{exer}{}{exer:irred:woset-is-total-order}
  Prove that every well-ordering is total.
\end{exer}
\begin{exer}{}{exer:irred:woset-iff-desc-chains-stabilize}
  Prove that a totally ordered set $(Z, \preceq)$ is a woset if and
  only if every descending chain $z_1 \succ z_2 \succ z_3 \succ \dots$
  in $Z$ stabilizes.
\end{exer}
\begin{exer}{}{exer:irred:ac-iff-surjections-have-right-inverse}
  Prove that the axiom of choice is equivalent to the statement that a
  set-function is surjective if and only if it has a right-inverse
  (cf.\ \Cref{exer:prelim:inv-of-srjctv}).
\end{exer}
\begin{exer}{}{exer:irred:explicit-well-ordering-on-z}
  \exerused{} Construct explicitly a well-ordering on $\Z$. Explain
  why you know that $\Q$ can be well-ordered, even without performing
  an explicit
  construction. [\S{}\ref{subsec:irred:set-theory-reprise}]
\end{exer}
\begin{exer}{}{exer:irred:induction-equiv-to-well-ordering-z-pos}
  \exerused{} Prove that the (ordinary) principle of induction is
  equivalent to the statement that $\leq$ is a well-ordering on
  $\Z_{>0}$.  (To prove by induction that $(\Z_{>0}, \leq)$ is
  well-ordered, assume it is known that $1$ is the least element of
  $\Z_{>0}$ and that $\forall n \in \Z_{>0}$ there are no integers
  between $n$ and $n+1$.) [\S{}\ref{subsec:irred:set-theory-reprise}]
\end{exer}
\begin{exer}{}{exer:irred:zorn-implies-well-ordering-thm}
  In this exercise assume the truth of Zorn's lemma and the
  conventional set-theoretic constructions; you will be proving the
  well-ordering theorem. Let $Z$ be a nonempty set, and let
  $\mathscr{Z}$ be the set of pairs $(S, \leq)$ consisting of a subset
  $S$ of $Z$ and of a well-ordering $\leq$ on $S$. Note that
  $\mathscr{Z}$ is not empty (singletons can be well-ordered). Define
  a relation $\preceq$ on $\mathscr{Z}$ by prescribing
  $(S, \leq) \preceq (T, \leq')$ if and only if $S \subseteq T$,
  $\leq$ is the restriction of $\leq'$ to $S$, and every element of
  $S$ precedes every element of $T \setminus S$ w.r.t.\ $\leq'$.
  \begin{itemize}
  \item Prove that $\preceq$ is an order relation in $\mathscr{Z}$.
  \item Prove that every chain in $\mathscr{Z}$ has an upper bound in
    $\mathscr{Z}$.
  \item Use Zorn's lemma to obtain a maximal element $(M, \leq)$ in
    $\mathscr{Z}$. Prove that $M = Z$.
  \end{itemize}
  Thus every set admits a well-ordering, as stated in
  \Cref{thm:irred:3.3}.
\end{exer}
\begin{exer}{}{exer:irred:ac-implies-well-ordering-thm-direct}
  In this exercise assume the truth of the axiom of choice and the
  conventional set-theoretic constructions; you will be proving the
  well-ordering theorem\footnote{I learned this proof from notes of
    Dan~Grayson.}. Let $Z$ be a nonempty set. Use the axiom of choice
  to choose an element $\gamma(S) \in S$ for each proper subset
  $S \subsetneq Z$. Call a pair $(S, \leq)$ a \emph{$\gamma$-woset} if
  $S \subseteq Z$, $\leq$ is a well-ordering on $S$, and for every
  $a \in S$, $a = \gamma(\{b \in S \mid b < a\})$.
  \begin{itemize}
  \item Show how to begin constructing a $\gamma$-woset, and show that
    all $\gamma$-wosets must begin in the same way.
  \end{itemize}
  Define an ordering on $\gamma$-wosets by prescribing that
  $(U, \leq'') \preceq (T, \leq')$ if and only if $U \subseteq T$ and
  $\leq''$ is the restriction of $\leq'$.
  \begin{itemize}
  \item Prove that if $(U, \leq'') \prec (T, \leq')$, then
    $\gamma(U) \in T$.
  \item For two $\gamma$-wosets $(S, \leq)$ and $(T, \leq')$, prove
    that there is a maximal $\gamma$-woset $(U, \leq'')$ preceding
    both w.r.t.\ $\preceq$. (Note: There is no need to use Zorn's
    lemma!)
  \item Prove that the maximal $\gamma$-woset found in the previous
    point in fact equals $(S, \leq)$ or $(T, \leq')$. Thus, $\preceq$
    is a total ordering.
  \item Prove that there is a maximal $\gamma$-woset $(M, \leq)$
    w.r.t.\ $\preceq$. (Again, Zorn's lemma need not and should not be
    invoked.)
  \item Prove that $M = Z$.
  \end{itemize}
  Thus every set admits a well-ordering, as stated in
  \Cref{thm:irred:3.3}.
\end{exer}
\begin{exer}{}{exer:irred:fin-gen-grp-has-maximal-subgrp}
  Prove that every nontrivial finitely generated group has a maximal
  proper subgroup. Prove that $(\Q, +)$ has no maximal proper
  subgroup.
\end{exer}
\begin{exer}{}{exer:irred:trivial-mult-rng-on-q-no-maximal-ideals}
  \exerused{} Consider the rng (= ring without~$1$; cf.\
  \S{}\ref{subsec:rings:definition}) consisting of the abelian group
  $(\Q, +)$ endowed with the trivial multiplication $qr = 0$ for all
  $q, r \in \Q$. Prove that this rng has no maximal
  ideals. [\S{}\ref{subsec:irred:application-existence-of-maximal-ideals}]
\end{exer}
\begin{exer}{}{exer:irred:maximal-ideals-not-corresponding-to-points}
  $\lnot$ As shown in \Cref{exer:rings:continuous-functions-compact},
  every maximal ideal in the ring of continuous real-valued functions
  on a compact topological space $K$ consists of the functions
  vanishing at a point of $K$. Prove that there are maximal ideals in
  the ring of continuous real-valued functions on the real line that
  do not correspond to points of the real line in the same
  fashion. (Hint: Produce a proper ideal that is not contained in any
  maximal ideal corresponding to a point, and apply
  \Cref{prop:irred:3.5}.)
  [\ref{exer:rings:continuous-functions-compact}]
\end{exer}

\begin{exer}{}{exer:irred:ufd-is-pid-iff-primes-maximal}
  Prove that a UFD $R$ is a PID if and only if every nonzero prime
  ideal in $R$ is maximal. (Hint: One direction is
  \Cref{prop:rings:PID-prime-maximal}. For the other, assume that
  every nonzero prime ideal in a UFD $R$ is maximal, and prove that
  every maximal ideal in $R$ is principal; then use
  \Cref{prop:irred:3.5} to relate arbitrary ideals to maximal ideals,
  and prove that every ideal of $R$ is principal.)
\end{exer}

\begin{exer}{}{exer:irred:minimal-primes-exist}
  $\lnot$ Let $R$ be a commutative ring, and let $I \subseteq R$ be a
  proper ideal. Prove that the set of prime ideals containing $I$ has
  minimal elements. (These are the \emph{minimal primes} of $I$.)
  [\ref{exer:irred:noeth-implies-fin-minimal-primes}]
\end{exer}

\begin{exer}{}{exer:irred:nilradical-subset-intersection-of-primes}
  $\lnot$ Let $R$ be a commutative ring, and let $N$ be its nilradical
  (\Cref{exer:rings:nilpotents-form-ideal}). Let $r \in N$.
  \begin{itemize}
  \item Consider the family $\mathscr{F}$ of ideals of $R$ that do not
    contain any power $r^k$ of $r$ for $k > 0$. Prove that
    $\mathscr{F}$ has maximal elements.
  \item Let $I$ be a maximal element of $\mathscr{F}$. Prove that $I$
    is prime.
  \item Conclude: $r \in N \implies r$ is not in the intersection of
    all prime ideals of $R$.
  \end{itemize}
  Together with \Cref{exer:rings:nilradical-in-every-prime}, this
  shows that the nilradical of a commutative ring $R$ equals the
  intersection of all prime ideals of $R$.
  [\ref{exer:rings:nilradical-in-every-prime},
  \ref{exer:fields:radical-ideal-nilradical-correspondence}]
\end{exer}

\begin{exer}{}{exer:irred:jacobson-radical-characterization}
  $\lnot$ The \emph{Jacobson radical} of a commutative ring $R$ is the
  intersection of the maximal ideals in $R$. (Thus, the Jacobson
  radical contains the nilradical.) Prove that $r$ is in the Jacobson
  radical if and only if $1 + rs$ is invertible for every $s \in
  R$. [\ref{exer:linalg:nakayamas-lemma}]
\end{exer}

\begin{exer}{}{exer:irred:cohens-thm-primes-fin-gen-implies-noeth}
  $\lnot$ Recall that a (commutative) ring $R$ is Noetherian if every
  ideal of $R$ is finitely generated. Assume the seemingly weaker
  condition that every prime ideal of $R$ is finitely generated. Let
  $\mathscr{F}$ be the family of ideals that are not finitely
  generated in $R$. You will prove $\mathscr{F} = \emptyset$.
  \begin{itemize}
  \item If $\mathscr{F} \neq \emptyset$, prove that it has a maximal
    element $I$.
  \item Prove that $R/I$ is Noetherian.
  \item Prove that there are ideals $J_1$, $J_2$ containing $I$, such
    that $J_1 J_2 \subseteq I$.
  \item Give a structure of $R/I$-module to $I/J_1 J_2$ and
    $J_1/J_1 J_2$.
  \item Prove that $I/J_1 J_2$ is a finitely generated $R/I$-module.
  \item Prove that $I$ is finitely generated, thereby reaching a
    contradiction.
  \end{itemize}
  Thus, a ring is Noetherian if and only if its prime ideals are
  finitely generated.
\end{exer}

\section{Unique factorization in polynomial rings}
\label{sec:irred:unique-factorization-in-polynomial-rings}

We now return to regular programming and study unique factorization in
polynomial rings; we will finally establish the fact (already hinted
at) that $R[x]$ is a UFD if $R$ is a UFD. Among the necessary
preparatory work we also discuss the field of fractions of an integral
domain; this is a particular instance of the process of localization
of a ring (or a module) at a multiplicative subset, which the diligent
reader will explore a little more in the exercises.

\subsection{Primitivity and content; Gauss's lemma}
\label{subsec:irred:primitivity-and-content-gauss-s-lemma}
One issue that we have encountered already, and will encounter again,
is the description of the ideals of the polynomial ring $R[x]$, given
enough information about $R$. For example, we have proved that the
ideals of $k[x]$ are principal if $k$ is a field, and Hilbert's basis
theorem shows that all ideals of $R[x]$ are finitely generated if all
ideals of $R$ are (\Cref{exer:rings:field-poly-ring-pid} and
\Cref{lem:irred:1.3}).

An even more naive observation is simply that every ideal $I$ of $R$
generates an ideal of $R[x]$:
\[IR[x] := \{a_0 + a_1 x + \dots + a_d x^d \in R[x] \mid \forall\,
  i,\, a_i \in I\}.\]
\begin{lem}{}{lem:irred:4.1}
  Let $R$ be a ring, and let $I$ be an ideal of $R$. Then
  $\dfrac{R[x]}{IR[x]} \cong \dfrac{R}{I}[x]$.
\end{lem}

The proof of this lemma is a standard application of the first
isomorphism theorem and is left to the reader
(\Cref{exer:irred:poly-ring-mod-ideal-iso}).
\begin{coro}{}{coro:irred:4.2}
  If $I$ is a prime ideal of $R$, then $IR[x]$ is prime in $R[x]$.
\end{coro}

\begin{proof}
  If $I$ is prime in $R$, then $R/I$ is an integral domain; hence so
  is $R[x]/IR[x] \cong (R/I)[x]$, and therefore $IR[x]$ is prime in
  $R[x]$.
\end{proof}

We will use this fact in a moment.  The following definitions can be
studied for every commutative ring; our main application will be to
UFDs.

\begin{defn}{}{defn:irred:4.3}
  Let $R$ be a commutative ring, and let
  $f = a_0 + a_1 x + \dots + a_d x^d \in R[x]$ be a polynomial.
  \begin{itemize}
  \item $f$ is \emph{very primitive} if for all prime ideals
    $\mathfrak{p}$ of $R$, $f \notin \mathfrak{p}R[x]$.
  \item $f$ is \emph{primitive} if for all principal prime ideals
    $\mathfrak{p}$ of $R$, $f \notin \mathfrak{p}R[x]$.
  \end{itemize}
\end{defn}
The notion of `primitive' polynomial is standard, but it is not
usually presented in this way; the standard definition is the
equivalent formulation given in \Cref{lem:irred:4.5}. As for `very
primitive', this term is essentially a joke---my excuse for bringing
up this notion is that it is very natural and that unfortunately some
references blur the distinction between this notion and the notion of
`primitive'. I hope to ward off any possible confusion by being rather
explicit on this point. Very primitive polynomials are primitive, but
the converse does not hold in general, even for UFDs (cf.\
\Cref{exer:irred:primitive-iff-very-primitive-in-pid}).  Perhaps the
most important fact about `primitivity' is the following easy remark.

\begin{lem}{}{lem:irred:4.4}
  Let $R$ be a commutative ring. Then for $f, g \in R[x]$,
  \[fg \text{ is primitive} \iff \text{both } f \text{ and } g \text{
      are primitive.}\]
\end{lem}

\begin{proof}
  This is an easy consequence of \Cref{coro:irred:4.2}:
  \begin{align*}
    fg \text{ primitive}
    &\iff \forall\, \mathfrak{p} \text{ prime and principal in } R,\
      fg \notin \mathfrak{p}R[x] \\
    &\iff \forall\, \mathfrak{p} \text{ prime and principal in } R,\
      f \notin \mathfrak{p}R[x] \text{ and } g \notin \mathfrak{p}R[x] \\
    &\iff f \text{ is primitive and } g \text{ is primitive,}
  \end{align*}
  since $\mathfrak{p}R[x]$ is prime if $\mathfrak{p}$ is a prime
  ideal.
\end{proof}

The analogous equivalence holds for very primitive polynomials
(\Cref{exer:irred:product-very-primitive-iff-factors-are}).
\Cref{lem:irred:4.4} will be ultimately responsible for the fact that
$R[x]$ is a UFD if $R$ is a UFD, which is my main objective in this
section. If $R$ is a UFD, the notion of `primitive' has the following
interpretation; I accompany it with the `very primitive' case, for
comparison.
\begin{lem}{}{lem:irred:4.5}
  Let $R$ be a commutative ring and
  $f = a_0 + a_1 x + \dots + a_d x^d \in R[x]$ as above.
  \begin{itemize}
  \item $f$ is very primitive if and only if
    $(a_0, \dots, a_d) = (1)$.
  \item If $R$ is a UFD, then $f$ is primitive if and only if
    $\gcd(a_0, \dots, a_d) = 1$.
  \end{itemize}
\end{lem}
\begin{proof}
  If $(a_0, \dots, a_d) = (1)$, then no prime ideal can contain all
  coefficients $a_i$, and it follows that $f$ is very primitive.
  Conversely, if $f$ is very primitive, then the coefficients of $f$
  are not all contained in any one prime ideal, and in particular they
  are not all contained in any maximal ideal (since maximal ideals are
  prime).  Thus, $(a_0, \dots, a_d) = (1)$ in this case, since every
  proper ideal is contained in a maximal ideal by
  \Cref{prop:irred:3.5}. This proves the first point.

  For the second point note that, in a UFD,
  $\gcd(a_0, \dots, a_d) \neq 1$ if and only if there exists an
  irreducible element $q \in R$ such that
  $(a_0, \dots, a_d) \subseteq (q)$. As $(q)$ is then prime (by
  \Cref{lem:irred:2.4}), the second point follows as well.
\end{proof}
With this in mind, in order to capitalize on \Cref{lem:irred:4.4}, it
is convenient to give a name to the gcd of the coefficients of a
polynomial. I now assume that $R$ is a UFD, since this is the case of
greatest interest in the applications.

\begin{defn}{}{defn:irred:4.6}
  Let $R$ be a UFD. The \emph{content} of a nonzero polynomial
  $f \in R[x]$, denoted $\operatorname{cont} f$, is the gcd of its
  coefficients.
\end{defn}
Of course the content of a polynomial is only defined up to units. I
find this ambiguity distasteful. What is uniquely determined is the
principal ideal generated by the content of $f$, which I will denote
$(\operatorname{cont} f)$: thus $f$ is primitive precisely when
$(\operatorname{cont} f) = (1)$. I will take a somewhat stubborn
approach and deal with principal ideals throughout, rather than
individual (but only defined up to unit) elements. The reader should
keep in mind that ideals may be multiplied (cf.\
\S{}\ref{subsec:rings:basic-operations}), and if $(a)$, $(b)$ are
principal ideals, then their product $(a)(b)$ is the principal ideal
$(ab)$.

The proof of the following remarks is immediate from the definition;
hence it is left to the reader
(\Cref{exer:irred:content-primitive-part-factorization}):
\begin{lem}{}{lem:irred:4.7}
  Let $R$ be a UFD, and let $f \in R[x]$. Then
  \begin{itemize}
  \item $(f) = (\operatorname{cont} f)(\tilde{f})$, where $\tilde{f}$
    is primitive;
  \item if $(f) = (c)(g)$, with $c \in R$ and $g$ primitive, then
    $(c) = (\operatorname{cont} f)$.
  \end{itemize}
\end{lem}
In view of its consequences, the following fact is rather profound,
and it therefore deserves a name. Some references call the result of
\Cref{lem:irred:4.4}, or its main consequence (\Cref{thm:irred:4.14}),
Gauss's lemma. We prefer to use this name for the following statement.

\begin{prop}{Gauss's lemma}{prop:irred:4.8}
  Let $R$ be a UFD, and let $f, g \in R[x]$. Then
  $(\operatorname{cont}(fg)) = (\operatorname{cont}
  f)(\operatorname{cont} g)$.
\end{prop}

\begin{proof}
  This follows easily from our preparatory work. Write
  $(fg) = ((\operatorname{cont} f)(\tilde{f}))((\operatorname{cont}
  g)(\tilde{g})) = (\operatorname{cont} f)(\operatorname{cont}
  g)(\tilde{f}\tilde{g})$, with $\tilde{f}$, $\tilde{g}$ primitive. By
  \Cref{lem:irred:4.4}, $\tilde{f}\tilde{g}$ is primitive; by
  \Cref{lem:irred:4.7} it follows that
  $(\operatorname{cont} f)(\operatorname{cont} g)$ is the content of
  $fg$, as needed.
\end{proof}

Note the following immediate consequence:

\begin{coro}{}{coro:irred:4.9}
  Let $R$ be a UFD, and let $f, g \in R[x]$. Assume
  $(f) \subseteq (g)$.  Then
  $(\operatorname{cont} f) \subseteq (\operatorname{cont} g)$.
\end{coro}

\subsection{The field of fractions of an integral domain}
\label{subsec:irred:the-field-of-fractions-of-an-integral-domain}
Gauss's lemma is the key to the important observation that if $R$ is a
UFD, then so is $R[x]$. However, we need one more tool before we can
prove this fact; this is one instance of the important process of
localization.

In the case we need for our immediate goal, the process starts from
any integral domain and produces a field, in precisely the same way
the field $\Q$ of rational numbers may be obtained from the integral
domain $\Z$. This construction is known as the \emph{field of
  fractions} (or \emph{field of quotients}) of the integral domain
$R$. It satisfies the following beautiful universal property.  Given
an integral domain $R$, consider the category $\mathscr{R}$ whose
objects are pairs $(i, K)$, where $K$ is a field and $i : R \to K$ is
an injective ring homomorphism. Morphisms are defined in the style of
every analogous construction we have encountered: that is, a morphism
$(i, K) \to (j, L)$ is determined by a homomorphism of fields
$\alpha : K \to L$ making the following diagram commute:
\[\begin{tikzcd}
    K \arrow[rr, "\alpha"] & & L \\
    & R \arrow[ul, "i"] \arrow[ur, "j"']
  \end{tikzcd}\] Note that $\alpha$, as every homomorphism of fields,
is necessarily injective (\Cref{exer:rings:field-hom-injective}); this
is compatible with the requirement that $i : R \to K$ be injective to
begin with. The injectivity of $R \to K$ also forces $R$ to be an
integral domain, since subrings of fields are necessarily integral
domains.
\begin{defn}{}{defn:irred:4.10}
  The \emph{field of fractions} $K(R)$ of $R$ is an initial object of
  the category $\mathscr{R}$.
\end{defn}

Thus, $K(R)$ is the `smallest field containing $R$'.  The usual
caveats apply to any such definition: the initial object of
$\mathscr{R}$ carries not only the information of a field $K$ (the
field of fractions proper), but also of a specific realization of $R$
as a subring of $K$; this distinction is blurred in common use of the
language. Also, of course this prescription only defines the field of
fractions up to isomorphism (as is true of any universal object; cf.\
\Cref{prop:prelims:initial-final-unique}). Finally, just stating the
universal property does not guarantee that the sought-for initial
object exists. Thus, our next task is the construction of an initial
object for $\mathscr{R}$.  Fortunately, the construction is
essentially straightforward: we just need to formalize a notion of
`fraction' of elements of $R$.

Consider the set $R \times (R^*)$ of pairs $(a, r)$ of elements of
$R$, where $r \neq 0$. The pair $(a, r)$ will be associated with a
`fraction' $\frac{a}{r}$; the reader would be well-advised to put this
book away now and carry out the construction of $K(R)$ on his/her own,
profiting from this major hint.  Here is the construction. Denote by
$\frac{a}{r}$ the equivalence class of $(a, r) \in R \times (R^*)$
with respect to the following equivalence relation:
\[(a, r) \sim (b, s) \iff as - br = 0.\] The reader will verify that
this is indeed an equivalence relation. As a set, $K(R)$ is defined by
\[K(R) := \left\{ \frac{a}{r} \;\middle|\; a \in R,\, r \in R,\, r
    \neq 0 \right\}.\] It is clear that these `fractions' behave much
as ordinary fractions. For example, $\frac{a}{r} = \frac{as}{rs}$ if
$s \neq 0$: indeed, $(as)r = a(rs)$ by associativity and commutativity
in $R$, and this shows $(as, rs) \sim (a, r)$ as needed.  We define
operations on $K(R)$ as follows:
\[\frac{a}{r} + \frac{b}{s} = \frac{as + br}{rs}, \qquad
  \frac{a}{r} \cdot \frac{b}{s} = \frac{ab}{rs}.\] Of course one must
verify that these operations are well-defined; this is also left to
the reader. The fact that $R$ is an integral domain is used in these
definitions: it guarantees that $rs \neq 0$ if $r \neq 0$ and
$s \neq 0$.  Now I claim that $K(R)$ is made into a field by these
operations. This is a straightforward verification; for example,
distributivity amounts to the following computation:
\[\frac{a}{r}\left(\frac{b}{s} + \frac{c}{t}\right)
  = \frac{a}{r} \cdot \frac{bt + cs}{st} = \frac{a(bt+cs)}{r(st)} =
  \frac{a(bt)}{r(st)} + \frac{a(cs)}{r(st)} = \frac{ab}{rs} +
  \frac{ac}{rt} = \frac{a}{r}\cdot\frac{b}{s} +
  \frac{a}{r}\cdot\frac{c}{t}.\] The zero element is the fraction
$\frac{0}{1}$ (or in fact $\frac{0}{r}$ for any nonzero $r \in R$);
the multiplicative identity is the fraction $\frac{1}{1}$ (or in fact
$\frac{r}{r}$ for any nonzero $r \in R$).  Since
$\frac{r}{s} \cdot \frac{s}{r} = \frac{rs}{rs} = \frac{1}{1}$ for all
$rs \neq 0$ (that is, for all $r \neq 0$), every nonzero element in
$K(R)$ has an inverse, as promised.  An `inclusion map'
$i : R \to K(R)$ is defined by $a \mapsto \frac{a}{1}$; it is
immediately checked that this is an injective ring homomorphism.  It
is common to use this map to identify $R$ with its isomorphic copy
inside $K(R)$ and simply view $R$ as a subring of $K(R)$.
\begin{fact}{Claim~4.11}{fact:irred:claim:4.11}
  $(i, K(R))$ is initial in $\mathscr{R}$.
\end{fact}

\begin{proof}
  Let $j : R \to L$ be any injective ring homomorphism from $R$ to a
  field $L$. We need to define an induced homomorphism
  $\hat{\jmath} : K(R) \to L$ so that the diagram
  \[\begin{tikzcd}
      K(R) \arrow[rr, "\hat{\jmath}"] & & L \\
      & R \arrow[ul, "i"] \arrow[ur, "j"']
    \end{tikzcd}\] commutes, and we must show that $\hat{\jmath}$ is
  unique. Now, the definition of $\hat{\jmath}$ is in fact forced upon
  us: if $\hat{\jmath}$ exists as a homomorphism, then necessarily
  \[\hat{\jmath}\!\left(\frac{a}{r}\right)
    = \hat{\jmath}\!\left(\frac{a}{1}\right)
    \hat{\jmath}\!\left(\frac{r}{1}\right)^{-1} = (\hat{\jmath} \circ
    i(a))(\hat{\jmath} \circ i(r))^{-1} = j(a)j(r)^{-1}.\] Thus
  $\hat{\jmath}$ is indeed unique, if it exists. On the other hand,
  the prescription
  $\hat{\jmath}\!\left(\frac{a}{r}\right) := j(a)j(r)^{-1}$ does
  define a function $K(R) \to L$: indeed, if $(a, r) \sim (b, s)$,
  then $as = br$ in $R$, hence $j(a)j(s) = j(b)j(r)$ in $L$, and (note
  that $j(r), j(s)$ are nonzero in $L$ since $r, s$ are nonzero in $R$
  and $j$ is injective) $j(a)j(r)^{-1} = j(b)j(s)^{-1}$, showing that
  the proposed $\hat{\jmath}$ is well-defined. The reader will verify
  that it is a ring homomorphism, concluding the proof of the claim.
\end{proof}
With the notation introduced above, $K(\Z) = \Q$. The universal
property implies immediately that $F \to K(F)$ is an isomorphism if
$F$ is itself a field. Thus, the construction adds nothing to $\Q$,
$\R$, $\C$, $\Z/p\Z$, etc.

If $R$ is any integral domain, so that $R[x]$ is also an integral
domain, $K(R[x])$ is a famous field:

\begin{defn}{}{defn:irred:4.13}
  The \emph{field of rational functions} with coefficients in $R$ is
  the field of fractions of the ring $R[x]$. This field is denoted
  $R(x)$.
\end{defn}

Elements of $R(x)$ are fractions of polynomials $\frac{p(x)}{q(x)}$
with $p(x), q(x) \in R[x]$ and $q(x) \neq 0$. The term \emph{function}
is inaccurate, since the `function' $R \to R$ given by
$a \mapsto \frac{p(a)}{q(a)}$ is not defined for all $a \in R$ (not
for those for which $q(a) = 0$, that is); and the function itself does
not suffice to determine the element in $R(x)$ (cf.\
\Cref{exer:rings:z2-poly-function}).

\subsection{\texorpdfstring{$R$ UFD $\implies$ $R[x]$ UFD}{R UFD
    implies R[x] UFD}}
\label{subsec:irred:r-ufd-implies-r-x-ufd}
We are now in a position to prove the analogue of Hilbert's basis
theorem for unique factorization domains, that is,

\begin{thm}{}{thm:irred:4.14}
  Let $R$ be a UFD; then $R[x]$ is a UFD.
\end{thm}
For example, this result (and an immediate induction) shows that the
rings $\Z[x_1, \dots, x_n]$ and $k[x_1, \dots, x_n]$ (for $k$ a field)
are UFDs. \Cref{thm:irred:4.14} is also often called Gauss's lemma.

As a measure of how delicate the statement of \Cref{thm:irred:4.14}
is, note that the power series ring $R[[x]]$ is not necessarily a UFD
if $R$ is a UFD; examples of this phenomenon are however not easy to
construct. Of course $k[[x]]$ is a UFD if $k$ is a field, since
$k[[x]]$ is a Euclidean domain in this case
(\Cref{exer:irred:formal-power-series-euclidean}).  By
\Cref{thm:irred:2.5}, in order to prove \Cref{thm:irred:4.14}, we have
to verify that $R[x]$ satisfies the a.c.c.\ for principal ideals and
that every irreducible element in $R[x]$ is prime, provided that $R$
is itself a UFD. The general idea is to reduce these questions to
matters in $K[x]$, where $K = K(R)$ is the field of fractions of $R$:
as we know, $K[x]$ is a UFD (in fact it is a Euclidean domain, and
Euclidean domain $\implies$ PID $\implies$ UFD, as shown
in~\S{}\ref{sec:irred:ufds-pids-euclidean-domains}).

The following lemma captures the most crucial ingredient of the
interaction between $R[x]$ and $K[x]$:
\begin{lem}{}{lem:irred:4.15}
  Let $R$ be a UFD, and let $K = K(R)$ be its field of fractions. For
  nonzero $f, g \in R[x]$, denote by $(f)$, $(g)$ the principal ideals
  $fR[x]$, $gR[x]$ in $R[x]$, and denote by $(f)_K$, $(g)_K$ the
  principal ideals $fK[x]$, $gK[x]$ in $K[x]$. Assume
  \begin{itemize}
  \item $(\operatorname{cont} g) \subseteq (\operatorname{cont} f)$
    and
  \item $(g)_K \subseteq (f)_K$.
  \end{itemize}
  Then $(g) \subseteq (f)$.
\end{lem}
\begin{proof}
  Since $(g)_K \subseteq (f)_K$, we have $g = fh$, where $h \in K[x]$.
  Write $h = \frac{a}{b} \tilde{h}$, where $a, b \in R$ and
  $\tilde{h} \in R[x]$ is a primitive polynomial: this can be done by
  collecting common denominators in $h$, then applying the first point
  of \Cref{lem:irred:4.7}. We then have $bg = af\tilde{h}$ in
  $R[x]$. By Gauss's lemma and since $\tilde{h}$ is primitive,
  \[(a \operatorname{cont} f) = (b \operatorname{cont} g);\] and since
  $(\operatorname{cont} g) \subseteq (\operatorname{cont} f)$ by
  hypothesis, we obtain
  $(a \operatorname{cont} f) \subseteq (b \operatorname{cont} f)$.
  Since $R$ is an integral domain and
  $(\operatorname{cont} f) \neq (0)$, this implies $a = bc$ for some
  $c \in R$. But then
  $h = \frac{a}{b} \tilde{h} = c\tilde{h} \in R[x]$, and
  $g = fh \in (f)$; that is, $(g) \subseteq (f)$ in $R[x]$, as needed.
\end{proof}
The first application is the following description of the irreducible
elements of $R[x]$; this will also be used in the proof of
\Cref{thm:irred:4.14} and is independently interesting.

\begin{prop}{}{prop:irred:4.16}
  Let $R$ be a UFD, and let $K$ be its field of fractions. Let
  $f \in R[x]$ be a nonconstant, irreducible polynomial. Then $f$ is
  irreducible as an element of $K[x]$.
\end{prop}

\begin{proof}
  First note that $f$ is primitive: otherwise we could factor out its
  content, and $f$ would not be irreducible.

  Next, assume $f = gh$, with $g, h \in K[x]$; we have to prove that
  either $g$ or $h$ is a unit in $K[x]$. Let $c, d \in K$ such that
  $g = c\tilde{g}$, $h = d\tilde{h}$, and $\tilde{g}$, $\tilde{h}$ are
  primitive polynomials in $R[x]$. By \Cref{lem:irred:4.4},
  $\tilde{g}\tilde{h}$ is also primitive; thus
  $(\operatorname{cont}(\tilde{g}\tilde{h})) = (1) =
  (\operatorname{cont} f)$; further, $(f)_K = (\tilde{g}\tilde{h})_K$
  as $cd \neq 0$ is a unit in $K$. By \Cref{lem:irred:4.15} we obtain
  $(f) = (\tilde{g}\tilde{h})$ as ideals of $R[x]$; that is,
  $f = u\tilde{g}\tilde{h}$ with $u \in R[x]$ a unit.  As $f$ is
  irreducible in $R[x]$, this implies that either $\tilde{g}$ or
  $\tilde{h}$ is a unit in $R[x]$. But then $g$ or $h$ were units in
  $K[x]$, verifying that $f$ is irreducible in $K[x]$.
\end{proof}
I have always found this fact almost counterintuitive: $K$ is `larger'
than $R$, so one may expect that it should be `easier' to factor
polynomials over $K$ than over $R$. \Cref{prop:irred:4.16} tells us
that this is not the case: with due attention to special cases,
irreducibility in $R[x]$ is `the same as' irreducibility in $K[x]$. To
be precise,

\begin{coro}{}{coro:irred:4.17}
  Let $R$ be a UFD and $K$ the field of fractions of $R$. Let
  $f \in R[x]$ be a nonconstant polynomial. Then $f$ is irreducible in
  $R[x]$ if and only if it is irreducible in $K[x]$ and primitive.
\end{coro}

The proof amounts to tying up loose ends, and I leave it to the reader
(\Cref{exer:irred:establish-irred-poly-char-over-ufd}).  We can now
prove the main result of this section. We will make systematic use of
the characterization of UFDs found in \Cref{thm:irred:2.5}.

\begin{proof}[Proof of \Cref{thm:irred:4.14}]
  We begin by verifying the a.c.c.\ for principal ideals in
  $R[x]$. Let
  \[(f_1) \subseteq (f_2) \subseteq (f_3) \subseteq \dots\] be an
  ascending chain of principal ideals of $R[x]$. By
  \Cref{coro:irred:4.9}, this induces an ascending chain of principal
  ideals
  \[(\operatorname{cont} f_1) \subseteq (\operatorname{cont} f_2)
    \subseteq (\operatorname{cont} f_3) \subseteq \dots\] in $R$;
  since $R$ is a UFD, this chain stabilizes: that is,
  $(\operatorname{cont} f_i) = (\operatorname{cont} f_{i+1})$ for
  $i \gg 0$. On the other hand, with notation as in
  \Cref{lem:irred:4.15} we have
  \[(f_1)_K \subseteq (f_2)_K \subseteq (f_3)_K \subseteq \dots\] as a
  sequence of ideals in $K[x]$; since $K[x]$ is a UFD (because it is a
  PID; cf.\ \S{}\ref{subsec:irred:pid-implies-ufd}) this sequence
  stabilizes. Therefore, $(f_i)_K = (f_{i+1})_K$ for $i \gg 0$.

  By \Cref{lem:irred:4.15}, $(f_i) = (f_{i+1})$ for $i \gg 0$; that
  is, the given chain of principal ideals stabilizes, as needed.
  Next, we consider an irreducible element $f$ of $R[x]$. We verify
  that $(f)$ is a prime ideal; by \Cref{thm:irred:2.5} it then follows
  that $R[x]$ is a UFD, as stated. If $f$ is irreducible and constant,
  then $f$ is prime in $R$ as $R$ is a UFD, and it follows that $f$ is
  prime in $R[x]$ (by \Cref{coro:irred:4.2}). Thus we may assume that
  $f$ is nonconstant and irreducible (and in particular primitive) in
  $R[x]$.

  By \Cref{prop:irred:4.16}, $f$ is irreducible as an element of
  $K[x]$; since $K[x]$ is a PID, $(f)_K$ is prime in $K[x]$. Consider
  the composition
  \[\rho : R[x] \to K[x] \to \frac{K[x]}{(f)_K}.\]
  I claim that $\ker \rho = (f)$. Indeed, the inclusion $\supseteq$ is
  trivial; for the other inclusion, note that $\rho(g) = 0$ implies
  that $g$ is divisible by $f$ in $K[x]$: that is,
  $(g)_K \subseteq (f)_K$; and we have
  $(\operatorname{cont} g) \subseteq (\operatorname{cont} f)$ since
  $(\operatorname{cont} f) = (1)$ as $f$ is primitive. By
  \Cref{lem:irred:4.15}, we obtain that $(g) \subseteq (f)$, i.e., $g$
  is divisible by $f$ in $R[x]$, as needed. Since $\ker \rho = (f)$,
  we find that $\rho$ induces an injective homomorphism
  $\frac{R[x]}{(f)} \hookrightarrow \frac{K[x]}{(f)_K}$.  Since the
  ring on the right is an integral domain (as $(f)_K$ is prime in
  $K[x]$), so is the ring on the left. This proves that $(f)$ is prime
  in $R[x]$, and we are done.
\end{proof}
Summarizing, if $R$ is a UFD, then factorization in $R[x]$ is `the
same as' factorization in the polynomial ring $K[x]$ over the field of
quotients of $R$. If $f(x) \in R[x]$, then $f(x)$ has a prime
factorization in $K[x]$ for the simpler reason that $K[x]$ is a PID,
hence a UFD; but if $R$ itself is a UFD, then we know that each of the
factors may be assumed to be in $R[x]$ to begin with (cf.\
\Cref{exer:irred:gauss-lemma-clear-denominators}).

\begin{exam}{}{exam:irred:4.18}
  As mentioned already, \Cref{thm:irred:4.14} implies that several
  important rings, such as $\Z[x_1, \dots, x_n]$ or
  $\C[x_1, \dots, x_n]$, are UFDs; for example $\Z[x]$ is a UFD, as
  announced in~\S{}\ref{subsec:irred:pid-implies-ufd}.  Further,
  arguing as in \Cref{exer:irred:acc-principal-ideals-infinite-vars}
  to reduce to the case of finitely many indeterminates, it follows
  that $\Z[x_1, x_2, \dots]$ is a UFD: this is an example of a
  non-Noetherian UFD, promised a while back (and illustrating the last
  missing feature of the picture presented at the beginning of the
  chapter).
\end{exam}

\subsection*{Exercises}

\begin{exer}{}{exer:irred:poly-ring-mod-ideal-iso}
  \exerused{} Prove
  \Cref{lem:irred:4.1}. [\S{}\ref{subsec:irred:primitivity-and-content-gauss-s-lemma}]
\end{exer}
\begin{exer}{}{exer:irred:maximal-ideal-extension-to-poly-ring-q}
  Let $R$ be a ring, and let $I$ be an ideal of $R$. Prove or disprove
  that if $I$ is maximal in $R$, then $IR[x]$ is maximal in $R[x]$.
\end{exer}
\begin{exer}{}{exer:irred:primitive-iff-very-primitive-in-pid}
  \exerused{} Let $R$ be a PID, and let $f \in R[x]$. Prove that $f$
  is primitive if and only if it is very primitive. Prove that this is
  not necessarily the case in an arbitrary
  UFD. [\S{}\ref{subsec:irred:primitivity-and-content-gauss-s-lemma}]
\end{exer}
\begin{exer}{}{exer:irred:product-very-primitive-iff-factors-are}
  \exerused{} Let $R$ be a commutative ring, and let $f, g \in R[x]$.
  Prove that $fg$ is very primitive $\iff$ both $f$ and $g$ are very
  primitive. [\S{}\ref{subsec:irred:primitivity-and-content-gauss-s-lemma}]
\end{exer}
\begin{exer}{}{exer:irred:content-primitive-part-factorization}
  \exerused{} Prove
  \Cref{lem:irred:4.7}. [\S{}\ref{subsec:irred:primitivity-and-content-gauss-s-lemma}]
\end{exer}
\begin{exer}{}{exer:irred:partial-fractions-in-pid-frac-field}
  Let $R$ be a PID, and let $K$ be its field of fractions.
  \begin{itemize}
  \item Prove that every element $c \in K$ can be written as a finite
    sum $c = \sum_i \frac{a_i}{p_i^{r_i}}$, where the $p_i$ are
    nonassociate irreducible elements in $R$, $r_i \geq 0$, and $a_i$,
    $p_i^{r_i}$ are relatively prime.
  \item If
    $\sum_i \frac{a_i}{p_i^{r_i}} = \sum_j \frac{b_j}{q_j^{s_j}}$ are
    two such expressions, prove that (up to reshuffling) $p_i = q_i$,
    $r_i = s_i$, and $a_i \equiv b_i \pmod{p_i^{r_i}}$.
  \item Relate this to the process of integration by `partial
    fractions' you learned about when you took calculus.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:localization-construction}
  \exerused{} A subset $S$ of a commutative ring $R$ is a
  \emph{multiplicative subset} (or \emph{multiplicatively closed}) if
  (i)~$1 \in S$ and (ii)~$s, t \in S \implies st \in S$. Define a
  relation on the set of pairs $(a, s)$ with $a \in R$, $s \in S$ as
  follows:
  \[(a, s) \sim (a', s') \iff (\exists\, t \in S),\quad t(s'a - sa') =
    0.\] Note that if $R$ is an integral domain and
  $S = R \setminus \{0\}$, then $S$ is a multiplicative subset, and
  the relation agrees with the relation introduced
  in~\S{}\ref{subsec:irred:the-field-of-fractions-of-an-integral-domain}.
  \begin{itemize}
  \item Prove that the relation $\sim$ is an equivalence relation.
  \item Denote by $\frac{a}{s}$ the equivalence class of $(a,s)$, and
    define the same operations $+$, $\cdot$ on such `fractions' as the
    ones introduced in the special case
    of~\S{}\ref{subsec:irred:the-field-of-fractions-of-an-integral-domain}. Prove
    that these operations are well-defined.
  \item The set $S^{-1}R$ of fractions, endowed with the operations
    $+$, $\cdot$, is the \emph{localization}\footnote{The terminology
      is motivated by applications to algebraic geometry; see
      VII.2.17.} of $R$ at the multiplicative subset $S$. Prove that
    $S^{-1}R$ is a commutative ring and that the function
    $a \mapsto \frac{a}{1}$ defines a ring homomorphism
    $\phi : R \to S^{-1}R$.
  \item Prove that $\phi(s)$ is invertible for every $s \in S$.
  \item Prove that $R \to S^{-1}R$ is initial among ring homomorphisms
    $f : R \to R'$ such that $f(s)$ is invertible in $R'$ for every
    $s \in S$.
  \item Prove that $S^{-1}R$ is an integral domain if $R$ is an
    integral domain.
  \item Prove that $S^{-1}R$ is the zero-ring if and only if
    $0 \in S$.
  \end{itemize} [\ref{exer:irred:localization-of-module},
  \ref{exer:irred:ideal-extension-contraction-localization},
  \ref{exer:irred:localization-at-prime-is-local},
  \ref{exer:irred:localization-of-ufd-is-ufd},
  \ref{exer:fields:regular-fns-away-from-vg-is-localization},
  \ref{exer:linrep:localization-is-a-functor},
  \ref{exer:linrep:localization-is-tensor-with-localized-ring},
  \ref{exer:linrep:tensor-and-localization-commute},
  \ref{exer:linrep:localization-is-flat},
  \S{}\ref{subsec:homol:derived-categories}]
\end{exer}
\begin{exer}{}{exer:irred:localization-of-module}
  $\lnot$ Let $S$ be a multiplicative subset of a commutative ring
  $R$, as in \Cref{exer:irred:localization-construction}. For every
  $R$-module $M$, define a relation $\sim$ on the set of pairs
  $(m, s)$, where $m \in M$ and $s \in S$:
  \[(m, s) \sim (m', s') \iff (\exists\, t \in S),\quad t(s'm - sm') =
    0.\] Prove that this is an equivalence relation, and define an
  $S^{-1}R$-module structure on the set $S^{-1}M$ of equivalence
  classes, compatible with the $R$-module structure on $M$. The module
  $S^{-1}M$ is the \emph{localization} of $M$ at
  $S$. [\ref{exer:irred:ideal-extension-contraction-localization},
  \ref{exer:irred:localization-at-prime-is-local},
  \ref{exer:irred:localization-of-noeth-module-is-noeth},
  \ref{exer:linrep:localization-is-a-functor},
  \ref{exer:linrep:localization-is-tensor-with-localized-ring},
  \ref{exer:linrep:tensor-and-localization-commute}]
\end{exer}
\begin{exer}{}{exer:irred:ideal-extension-contraction-localization}
  $\lnot$ Let $S$ be a multiplicative subset of a commutative ring
  $R$, and consider the localization operation introduced in
  \Cref{exer:irred:localization-construction,exer:irred:localization-of-module}.
  \begin{itemize}
  \item Prove that if $I$ is an ideal of $R$ such that
    $I \cap S = \emptyset$, then\footnote{The superscript $e$ stands
      for `extension' (of the ideal from a smaller ring to a larger
      ring); the superscript $c$ in the next point stands for
      `contraction'.}  $I^e := S^{-1}I$ is a proper ideal of
    $S^{-1}R$.
  \item If $\phi : R \to S^{-1}R$ is the natural homomorphism, prove
    that if $J$ is a proper ideal of $S^{-1}R$, then
    $J^c := \phi^{-1}(J)$ is an ideal of $R$ such that
    $J^c \cap S = \emptyset$.
  \item Prove that $(J^c)^e = J$, while
    $(I^e)^c = \{a \in R \mid (\exists\, s \in S),\, sa \in I\}$.
  \item Find an example showing that $(I^e)^c$ need not equal $I$,
    even if $I \cap S = \emptyset$. (Hint: Let
    $S = \{1, x, x^2, \dots\}$ in $R = \C[x,y]$. What is $(I^e)^c$ for
    $I = (xy)$?)
  \end{itemize} [\ref{exer:irred:prime-ideal-bijection-localization},
  \ref{exer:irred:localization-of-noeth-module-is-noeth}]
\end{exer}
\begin{exer}{}{exer:irred:prime-ideal-bijection-localization}
  $\lnot$ With notation as in
  \Cref{exer:irred:ideal-extension-contraction-localization}, prove
  that the assignment $\mathfrak{p} \mapsto S^{-1}\mathfrak{p}$ gives
  an inclusion-preserving bijection between the set of prime ideals of
  $R$ disjoint from $S$ and the set of prime ideals of
  $S^{-1}R$. (Prove that $(\mathfrak{p}^e)^c = \mathfrak{p}$ if
  $\mathfrak{p}$ is a prime ideal disjoint from $S$.)
  [\ref{exer:irred:ufd-iff-localization-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:localization-at-prime-is-local}
  $\lnot$ (Notation as in
  \Cref{exer:irred:localization-construction,exer:irred:localization-of-module}.)
  A ring is said to be
  \emph{local} if it has a single maximal ideal.

  Let $R$ be a commutative ring, and let $\mathfrak{p}$ be a prime
  ideal of $R$. Prove that the set $S = R \setminus \mathfrak{p}$ is
  multiplicatively closed. The localizations $S^{-1}R$, $S^{-1}M$ are
  then denoted $R_{\mathfrak{p}}$, $M_{\mathfrak{p}}$.

  Prove that there is an inclusion-preserving bijection between the
  prime ideals of $R_{\mathfrak{p}}$ and the prime ideals of $R$
  contained in $\mathfrak{p}$. Deduce that $R_{\mathfrak{p}}$ is a
  local ring.  [\ref{exer:irred:module-zero-iff-zero-locally},
  \ref{exer:irred:dvr-is-local-pid-localization-is-dvr},
  \ref{exer:linalg:direct-summand-of-free-over-local-is-free},
  \ref{exer:fields:germs-identified-with-localization},
  \ref{exer:linrep:flatness-is-local-property}]
\end{exer}
\begin{exer}{}{exer:irred:module-zero-iff-zero-locally}
  $\lnot$ (Notation as in
  \Cref{exer:irred:localization-at-prime-is-local}.) Let $R$ be a
  commutative ring, and let $M$ be an $R$-module. Prove that the
  following are equivalent\footnote{The way to think of this type of
    results is that a module $M$ over $R$ is zero if and only if it is
    zero `at every point of $\operatorname{Spec} R$'. Working in the
    localization $M_{\mathfrak{p}}$ amounts to looking `near' the
    point $\mathfrak{p}$ of $\operatorname{Spec} R$; this is what is
    local about localization. As in this exercise, one can often
    detect `global' features by looking locally at every point.}:
  \begin{enumerate}[label=(\arabic*)]
  \item $M = 0$.
  \item $M_{\mathfrak{p}} = 0$ for every prime ideal $\mathfrak{p}$.
  \item $M_{\mathfrak{m}} = 0$ for every maximal ideal $\mathfrak{m}$.
  \end{enumerate}
  (\emph{Hint:} For the interesting implication, suppose that
  $m \neq 0$ in $M$; then the ideal $\{r \in R \mid rm = 0\}$ is
  proper. By \Cref{prop:irred:3.5}, it is contained in a maximal ideal
  $\mathfrak{m}$.  What can you say about $M_{\mathfrak{m}}$?)
  [\ref{exer:linrep:localization-is-faithfully-exact},
  \ref{exer:linrep:flatness-is-local-property}]
\end{exer}
\begin{exer}{}{exer:irred:dvr-is-local-pid-localization-is-dvr}
  $\lnot$ Let $k$ be a field, and let $v$ be a discrete valuation on
  $k$.  Let $R$ be the corresponding DVR, with local parameter $t$
  (see \Cref{exer:irred:dvr-is-pid-local-parameter}).
  \begin{enumerate}[label=(\arabic*)]
  \item Prove that $R$ is local
    (\Cref{exer:irred:localization-at-prime-is-local}), with maximal
    ideal $\mathfrak{m} = (t)$. (\emph{Hint:} Note that every element
    of $R \setminus \mathfrak{m}$ is invertible.)
  \item Prove that $k$ is the field of fractions of $R$.
  \item Now let $A$ be a PID, and let $\mathfrak{p}$ be a prime ideal
    in $A$. Prove that the localization $A_{\mathfrak{p}}$ (cf.\
    \Cref{exer:irred:localization-at-prime-is-local}) is a
    DVR. (\emph{Hint:} If $\mathfrak{p} = (p)$, define a valuation on
    the field of fractions of $A$ in terms of `divisibility by $p$'.)
  \end{enumerate} [\ref{exer:fields:two-curves-c1-c2-comparison}]
\end{exer}
\begin{exer}{}{exer:irred:localization-of-noeth-module-is-noeth}
  With notation as in \Cref{exer:irred:localization-of-module}, define
  operations $N \mapsto N^e$ and $\hat{N} \mapsto \hat{N}^c$ for
  submodules $N \subseteq M$, $\hat{N} \subseteq S^{-1}M$,
  respectively, analogously to the operations defined in
  \Cref{exer:irred:ideal-extension-contraction-localization}. Prove
  that $(\hat{N}^c)^e = \hat{N}$. Prove that every localization of a
  Noetherian module is Noetherian. In particular, all localizations
  $S^{-1}R$ of a Noetherian ring are Noetherian.
\end{exer}
\begin{exer}{}{exer:irred:localization-of-ufd-is-ufd}
  $\lnot$ Let $R$ be a UFD, and let $S$ be a multiplicatively closed
  subset of $R$ (cf.\ \Cref{exer:irred:localization-construction}).
  \begin{enumerate}[label=(\arabic*)]
  \item Prove that if $q$ is irreducible in $R$, then $q/1$ is either
    irreducible or a unit in $S^{-1}R$.
  \item Prove that if $a/s$ is irreducible in $S^{-1}R$, then $a/s$ is
    an associate of $q/1$ for some irreducible element $q$ of $R$.
  \item Prove that $S^{-1}R$ is also a UFD.
  \end{enumerate} [\ref{exer:irred:ufd-iff-localization-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:ufd-iff-localization-ufd}
  Let $R$ be a Noetherian integral domain, and let $s \in R$,
  $s \neq 0$, be a prime element. Consider the multiplicatively closed
  subset $S = \{1, s, s^2, \ldots\}$. Prove that $R$ is a UFD if and
  only if $S^{-1}R$ is a UFD. (\emph{Hint:} By
  \Cref{exer:irred:noeth-domain-ufd-iff-height-1-principal}, it
  suffices to show that every prime of height~1 is principal. Use
  \Cref{exer:irred:prime-ideal-bijection-localization} to relate prime
  ideals in $R$ to prime ideals in the localization.)

  On the basis of results such as this and of
  \Cref{exer:irred:localization-of-ufd-is-ufd}, one might suspect that
  being factorial is a local property, that is, that $R$ is a UFD if
  and only if $R_{\mathfrak{p}}$ is a UFD for all prime ideals
  $\mathfrak{p}$, if and only if $R_{\mathfrak{m}}$ is a UFD for all
  maximal ideals $\mathfrak{m}$. This is regrettably not the case. A
  ring $R$ is \emph{locally factorial} if $R_{\mathfrak{m}}$ is a UFD
  for all maximal ideals $\mathfrak{m}$; factorial implies locally
  factorial by \Cref{exer:irred:localization-of-ufd-is-ufd}, but
  locally factorial rings that are not factorial do exist.
\end{exer}
\begin{exer}{}{exer:irred:prime-subfield-char-0-or-p}
  \exerused Let $F$ be a field, and recall the notion of
  characteristic of a ring (\Cref{defn:rings:characteristic}); the
  characteristic of a field is either $0$ or a prime integer
  (\Cref{exer:rings:domain-characteristic-0-or-prime}).
  \begin{enumerate}[label=(\arabic*)]
  \item Show that $F$ has characteristic $0$ if and only if it
    contains a copy of $\Q$, and that $F$ has characteristic $p$ if
    and only if it contains a copy of the field $\Z/p\Z$.
  \item Show that (in both cases) this determines the smallest
    subfield of $F$; it is called the \emph{prime subfield} of $F$.
  \end{enumerate}
  [\S{}\ref{subsec:irred:adding-roots-algebraically-closed-fields},
  \S{}\ref{subsec:fields:basic-definitions}]
\end{exer}
\begin{exer}{}{exer:irred:units-of-domain-poly-ring}
  $\lnot$ Let $R$ be an integral domain. Prove that the invertible
  elements in $R[x]$ are the units of $R$, viewed as constant
  polynomials. [\ref{exer:irred:unit-in-poly-ring-general-ring}]
\end{exer}
\begin{exer}{}{exer:irred:nilpotent-implies-1-plus-a-unit}
  \exerused An element $a \in R$ in a ring is said to be
  \emph{nilpotent} if $a^n = 0$ for some $n \geq 0$. Prove that if $a$
  is nilpotent, then $1 + a$ is a unit in
  $R$. [\ref{exer:linalg:nilpotent-matrices-jordan-form-and-trace},
  \S{}\ref{subsec:fields:a-little-affine-algebraic-geometry}]
\end{exer}
\begin{exer}{}{exer:irred:unit-in-poly-ring-general-ring}
  Generalize the result of \Cref{exer:irred:units-of-domain-poly-ring}
  as follows: let $R$ be a commutative ring, and let
  $f = a_0 + a_1 x + \dots + a_d x^d \in R[x]$; prove that $f$ is a
  unit in $R[x]$ if and only if $a_0$ is a unit in $R$ and
  $a_1, \dots, a_d$ are nilpotent. (\emph{Hint:} If
  $b_0 + b_1 x + \dots + b_e x^e$ is the inverse of $f$, show by
  induction that $a_d^{i+1} b_{e-i} = 0$ for all $i \geq 0$, and
  deduce that $a_d$ is nilpotent.)
\end{exer}
\begin{exer}{}{exer:irred:establish-irred-poly-char-over-ufd}
  \exerused Establish the characterization of irreducible polynomials
  over a UFD given in
  \Cref{coro:irred:4.17}. [\S{}\ref{subsec:irred:r-ufd-implies-r-x-ufd}]
\end{exer}
\begin{exer}{}{exer:irred:common-factor-in-kxy-from-field-of-fractions}
  Let $k$ be a field, and let $f$, $g$ be two polynomials in
  $k[x, y] = k[x][y]$. Prove that if $f$ and $g$ have a nontrivial
  common factor in $k(x)[y]$, then they have a nontrivial common
  factor in $k[x, y]$.
\end{exer}
\begin{exer}{}{exer:irred:gauss-lemma-clear-denominators}
  \exerused Let $R$ be a UFD, $K$ its field of fractions,
  $f(x) \in R[x]$, and assume $f(x) = \alpha(x)\beta(x)$ with
  $\alpha(x)$, $\beta(x)$ in $K[x]$. Prove that there exists a
  $c \in K$ such that $c\alpha(x) \in R[x]$,
  $c^{-1}\beta(x) \in R[x]$, so that
  $f(x) = (c\alpha(x))(c^{-1}\beta(x))$ splits as a product of factors
  in $R[x]$.

  Deduce that if $\alpha(x)\beta(x) = f(x) \in R[x]$ is monic and
  $\alpha(x) \in K[x]$ is monic, then $\alpha(x)$, $\beta(x)$ are both
  in $R[x]$ and $\beta(x)$ is also monic.
  [\S{}\ref{subsec:irred:r-ufd-implies-r-x-ufd},
  \ref{exer:irred:coeff-products-lie-in-r},
  \S{}\ref{subsec:fields:cyclotomic-polynomials-and-fields}]
\end{exer}
\begin{exer}{}{exer:irred:coeff-products-lie-in-r}
  In the same situation as in
  \Cref{exer:irred:gauss-lemma-clear-denominators}, prove that the
  product of any coefficient of $\alpha$ with any coefficient of
  $\beta$ lies in $R$.
\end{exer}
\begin{exer}{}{exer:irred:flt-for-polynomials}
  Prove Fermat's last theorem for polynomials: the equation
  $f^n + g^n = h^n$ has no solutions in $\C[t]$ for $n > 2$ and $f$,
  $g$, $h$ not all constant. (\emph{Hint:} First, prove that $f$, $g$,
  $h$ may be assumed to be relatively prime. Next, the polynomial
  $1 - t^n$ factorizes in $\C[t]$ as $\prod_{i=1}^{n}(1 - \zeta^i t)$
  for $\zeta = e^{2\pi i/n}$; deduce that
  $f^n = \prod_{i=1}^{n}(h - \zeta^i g)$. Use unique factorization in
  $\C[t]$ to conclude that each of the factors $h - \zeta^i g$ is an
  $n$-th power. Now let $h - g = a^n$, $h - \zeta g = b^n$,
  $h - \zeta^2 g = c^n$ (this is where the $n > 2$ hypothesis enters).
  Use this to obtain a relation
  $(\lambda a)^n + (\mu b)^n = (\nu c)^n$, where $\lambda$, $\mu$,
  $\nu$ are suitable complex numbers. What's wrong with this?)

  The same pattern of proof would work in any environment where unique
  factorization is available; if adjoining to $\Z$ an $n$-th root of
  $1$ led to a unique factorization domain, the full-fledged Fermat's
  last theorem would be as easy to prove as indicated in this
  exercise.  This is not the case, a fact famously missed by G.\
  Lam\'e as he announced a `proof' of Fermat's last theorem to the
  Paris Academy on March~1, 1847.
\end{exer}

\section{Irreducibility of polynomials}
\label{sec:irred:irreducibility-of-polynomials}

Our work in \S{}\ref{subsec:irred:r-ufd-implies-r-x-ufd}, especially
\Cref{coro:irred:4.17}, links the irreducibility of polynomials over a
UFD with their irreducibility over the corresponding field of
fractions. For example, irreducibility of polynomials in $\Z[x]$ is
`essentially the same' as irreducibility in $\Q[x]$. This can be
useful in both directions, provided we have ways to effectively verify
irreducibility of polynomials over one kind of ring or the other. I
collect here several remarks aimed at establishing whether a given
polynomial is irreducible and briefly discuss the notion of
algebraically closed field.

\subsection{Roots and reducibility}
\label{subsec:irred:roots-and-reducibility}
Let $R$ be a ring and $f \in R[x]$. An element $a \in R$ is a
\emph{root} of $f$ if $f(a) = 0$. Recall
(\Cref{exam:rings:remainder-degree-1}) that a polynomial
$f(x) \in R[x]$ is divisible by $(x - a)$ if and only if $a$ is a root
of $f$. More generally, we say that $a$ is a root of $f$ with
\emph{multiplicity} $r$ if $(x - a)^r$ divides $f$ and $(x - a)^{r+1}$
does not divide $f$. The reader is probably familiar with this notion
from experience with calculus. For example, the graph of polynomials
with roots of multiplicities 1, 2, and 3 at 0 look (near the origin),
respectively, like
\begin{lem}{}{lem:irred:5.1}
  Let $R$ be an integral domain, and let $f \in R[x]$ be a polynomial
  of degree $n$. Then the number of roots of $f$, counted with
  multiplicity, is at most $n$.
\end{lem}
\begin{proof}
  The number of roots of $f$ in $R$ is less than or equal to the
  number of roots of $f$ viewed as a polynomial over the field of
  fractions $K$ of $R$; so we may replace $R$ by $K$.

  Now, $K[x]$ is a UFD, and the roots of $f$ correspond to the
  irreducible factors of $f$ of degree $1$. Since the product of all
  irreducible factors of $f$ has degree $n$, the number of factors of
  degree $1$ can be at most $n$, as claimed.
\end{proof}
It is worth remembering the trick of replacing an integral domain $R$
by its field of fractions $K$, used in this proof: one is often able
to use convenient properties due to the fact that $K$ is a field (the
fact that $K[x]$ is a UFD, in this case), even if $R$ is very far from
satisfying such properties ($R[x]$ may not be a UFD, since $R$ itself
need not be a UFD). Also note that the statement of
\Cref{lem:irred:5.1} may fail if $R$ is not an integral domain. For
example, the degree~2 polynomial $x^2 + x$ has four roots over
$\Z/6\Z$.

The innocent \Cref{lem:irred:5.1} has important applications: for
example, we used it (without much fanfare) in the proof of
\Cref{thm:groups2:finite-subgroup-of-field-mult-group-is-cyclic}. Also, recall that a polynomial $f \in R[x]$
determines an `evaluation function' (cf.\
\Cref{exam:rings:poly-universal-general}) $R \to R$, namely
$r \mapsto f(r)$. The reader checked in
\Cref{exer:rings:z2-poly-function} that in general the function does
not determine the polynomial; but polynomials are determined by the
corresponding functions over infinite integral domains:
\begin{coro}{}{coro:irred:5.2}
  Let $R$ be an infinite integral domain, and let $f, g \in R[x]$ be
  polynomials. Then $f = g$ if and only if the evaluation functions
  $r \mapsto f(r)$, $r \mapsto g(r)$ agree.
\end{coro}
\begin{proof}
  Indeed, the two functions agree if and only if every $a \in R$ is a
  root of $f - g$; but a nonzero polynomial over $R$ cannot have
  infinitely many roots, by \Cref{lem:irred:5.1}.
\end{proof}
Clearly, an irreducible polynomial of degree $\geq 2$ can have no
roots.  The converse holds for degree 2 and 3, over a field:

\begin{prop}{}{prop:irred:5.3}
  Let $k$ be a field. A polynomial $f \in k[x]$ of degree $2$ or $3$
  is irreducible if and only if it has no roots.
\end{prop}
\begin{proof}
  \Cref{exer:irred:deg-2-3-irred-iff-no-roots}.
\end{proof}
\begin{exam}{}{exam:irred:5.4}
  Let $\F_2$ be the field $\Z/2\Z$. The polynomial
  $f(t) = t^2 + t + 1 \in \F_2[t]$ is irreducible, since it has no
  roots: $f(0) = f(1) = 1$. Therefore the ideal $(t^2 + t + 1)$ is
  prime in $\F_2[t]$, hence maximal (because $\F_2[t]$ is a PID: don't
  forget \Cref{prop:rings:PID-prime-maximal}). This gives a
  one-sentence construction of a field with four elements:
  \[
    \frac{\F_2[t]}{(t^2 + t + 1)}.
  \]
  The reader (hopefully) constructed this field in
  \Cref{exer:rings:four-element-field}, by the tiresome process of
  searching by hand for a suitable multiplication table. The reader
  will now have no difficulty constructing much larger examples (cf.\
  \Cref{exer:irred:construct-fields-27-and-121-elements}).
\end{exam}
\Cref{prop:irred:5.3} is pleasant and can help to decide
irreducibility over more general rings: for example, a primitive
polynomial $f \in \Z[x]$ of degree $2$ or $3$ is irreducible if and
only if it has no roots in $\Q$. Indeed, $f$ is irreducible in $\Z[x]$
if and only if it is irreducible in $\Q[x]$ (by
\Cref{coro:irred:4.17}). However, note that e.g.\
$4x^2 - 1 = (2x - 1)(2x + 1)$ is primitive and reducible in $\Z[x]$,
although it has no integer roots. Also, keep in mind that the
statement of \Cref{prop:irred:5.3} may fail for polynomials of degree
$\geq 4$: for example, $x^4 + 2x^2 + 1 = (x^2 + 1)^2$ is reducible in
$\Q[x]$, but it has no rational roots.

Looking for rational roots of a polynomial in $\Z[x]$ is in principle
a finite endeavor, due to the following observation (which holds over
every UFD). This is often called the `rational root test'.
\begin{prop}{}{prop:irred:5.5}
  Let $R$ be a UFD, and let $K$ be its field of fractions. Let
  $f(x) = a_0 + a_1 x + \dots + a_n x^n \in R[x]$, and let
  $c = \frac{p}{q} \in K$ be a root of $f$, with $p, q \in R$,
  $\gcd(p, q) = 1$. Then $p \mid a_0$ and $q \mid a_n$ in $R$.
\end{prop}
\begin{proof}
  By hypothesis,
  \[
    a_0 + a_1 \frac{p}{q} + \dots + a_n \frac{p^n}{q^n} = 0;
  \]
  that is, $a_0 q^n + a_1 p q^{n-1} + \dots + a_n p^n = 0$. Therefore
  \[
    a_0 q^n = -p(a_1 q^{n-1} + \dots + a_n p^{n-1}),
  \]
  proving that $p \mid (a_0 q^n)$. Since $\gcd(p, q) = 1$, this
  implies that the multiset of factors of $p$ is contained in the
  multiset of irreducible factors of $a_0$, that is, $p \mid a_0$. An
  entirely similar argument proves that $q \mid a_n$.
\end{proof}
\begin{exam}{}{exam:irred:5.6}
  Looking for rational roots of the polynomial
  $3 - 2x + 3x^2 - 2x^3 + 3x^4 - 2x^5$ is therefore reduced to trying
  fractions $\frac{p}{q}$ with $q = \pm 1, \pm 2$, $p = \pm 1, \pm 3$.
  As it happens, $\frac{3}{2}$ is the only root found among these
  possibilities, and it follows that it is the only rational root of
  the polynomial.
\end{exam}

\subsection{Adding roots; algebraically closed fields}
\label{subsec:irred:adding-roots-algebraically-closed-fields}
A homomorphism of fields $i : k \to F$ is necessarily injective (cf.\
\Cref{exer:rings:field-hom-injective}): indeed, its kernel is a proper
ideal of $k$, and the only proper ideal of a field is $(0)$. In this
situation we say that $F$ (or more properly the homomorphism $i$) is
an \emph{extension} of $k$. Abusing language a little, we then think
of $k$ as contained in $F$: $k \subseteq F$. Keep in mind that this is
the case as soon as there is any ring homomorphism $k \to F$.

There is a standard procedure for constructing extensions in which a
given polynomial acquires a root; we have encountered an instance of
this process in \Cref{exam:rings:higher-degree-quotient}. In fact, the
resulting field is almost universal with respect to this requirement.
\begin{prop}{}{prop:irred:5.7}
  Let $k$ be a field, and let $f(t) \in k[t]$ be a nonzero irreducible
  polynomial. Then
  \[
    F := \frac{k[t]}{(f(t))}
  \]
  is a field, endowed with a natural homomorphism $i : k \to F$
  (obtained as the composition $k \to k[x] \to F$) realizing it as an
  extension of $k$. Further,
  \begin{enumerate}[label=(\arabic*)]
  \item $f(x) \in k[x] \subseteq F[x]$ has a root in $F$, namely the
    coset of $t$;
  \item if $k \subseteq K$ is any extension in which $f$ has a root,
    then there exists a homomorphism $j : F \to K$ such that the
    diagram
    \begin{equation*}
      \begin{tikzcd}
        k \ar[r, "i"] \ar[dr, hook] & F \ar[d, "j"] \\
        & K
      \end{tikzcd}
    \end{equation*}
    commutes.
  \end{enumerate}
\end{prop}
\begin{proof}
  Since $k$ is a field, $k[t]$ is a PID; hence $(f(t))$ is a maximal
  ideal of $k[t]$, by \Cref{prop:rings:PID-prime-maximal}. Therefore
  $F$ is indeed a field. Denoting cosets in $k[t]/(f(t)) = F$ by
  overlining, we have $f(\overline{t}) = \overline{f(t)} = 0$, as
  claimed.

  To verify the second part, suppose $k \subseteq K$ is an extension
  and $f(u) = 0$, with $u \in K$. This means that the evaluation
  homomorphism $\varphi : k[t] \to K$ defined by
  $\varphi(g(t)) := g(u)$ vanishes at $f(t)$; hence
  $(f(t)) \subseteq \ker(\varphi)$, and the universal property of
  quotients gives a unique homomorphism
  $j : F = \frac{k[t]}{(f(t))} \to K$ satisfying the stated
  requirement.
\end{proof}
I have stopped short of claiming that the extension constructed in
\Cref{prop:irred:5.7} is universal with respect to the requirement of
containing a root of $f$ because the homomorphism $j$ appearing in the
statement is not unique: in fact, the proof shows that there are as
many such homomorphisms as there are roots of $f$ in the larger field
$K$. We could say that $F$ is \emph{versal}, meaning `universal
without uni(queness)'. We would have full universality if we included
the information of the root of $f$; we will come back to this
construction in \Cref{chap:fields}, when we analyze field extensions
in greater depth.
\begin{exam}{}{exam:irred:5.8}
  For $k = \R$ and $f(x) = x^2 + 1$, the field constructed in
  \Cref{prop:irred:5.7} is (isomorphic to) $\C$: this was checked
  carefully in \Cref{exam:rings:higher-degree-quotient}. Similarly,
  $\Q[t]/(t^2 - 2)$ produces a field containing $\Q$ and in which
  there is a `square root of 2'. There are two embeddings of this
  field in $\R$, because $\R$ contains two distinct square roots of
  $2$: $\pm\sqrt{2}$.
\end{exam}
The fact that the irreducible polynomial $f \in k[x]$ acquires a root
in the extension $F$ constructed in \Cref{prop:irred:5.7} implies that
$f$ has a linear (i.e., degree~1) factor over $F$; in particular, if
$\deg(f) > 1$, then $f$ is no longer irreducible over $F$. Given any
polynomial $f \in k[x]$, it is easy to construct an extension of $k$
in which $f$ factors completely as a product of linear factors
(\Cref{exer:irred:extension-where-f-splits-linear}). The case in which
this already happens in $k$ itself for every nonzero $f$ is very
important, and hence it is given a name.
\begin{defn}{}{defn:irred:5.9}
  A field $k$ is \emph{algebraically closed} if all irreducible
  polynomials in $k[x]$ have degree $1$.
\end{defn}
We have encountered this notion in passing, back in
\Cref{exam:rings:speck-prime-ideals-maximal} (and
\Cref{exer:rings:maximal-ideal-linear-poly}). The following lemma is
left to the reader:

\begin{lem}{}{lem:irred:5.10}
  A field $k$ is algebraically closed if and only if every nonconstant
  polynomial $f \in k[x]$ factors completely as a product of linear
  factors, if and only if every nonconstant polynomial $f \in k[x]$
  has a root in $k$.
\end{lem}

In other words, a field $k$ is algebraically closed if the only
polynomial equations with coefficients in $k$ and no solutions are
equations `of degree 0', such as $1 = 0$. The result known as
Nullstellensatz (also due to David Hilbert; I have mentioned this
result in \Cref{chap:rings}) is a vast generalization of this
observation and one of the pillars of (old-fashioned) algebraic
geometry. We will come back to all of this in
\S{}\ref{sec:fields:algebraic-closure-nullstellensatz-and-a-little-algebraic}.
Is any of the finite fields we have encountered (such as $\Z/p\Z$, for
$p$ prime) algebraically closed? No. Adapting Euclid's argument
proving the infinitude of prime integers
(\Cref{exer:irred:euclid-infinitude-of-primes}) reveals that
algebraically closed fields are infinite.

\begin{prop}{}{prop:irred:5.11}
  Let $k$ be an algebraically closed field. Then $k$ is infinite.
\end{prop}
\begin{proof}
  By contradiction, assume that $k$ is algebraically closed and
  finite; let the elements of $k$ be $c_1, \dots, c_N$. Then there are
  exactly $N$ irreducible monic polynomials in $k[x]$, namely
  $(x - c_1), \dots, (x - c_N)$. Consider the polynomial
  \[
    f(x) = (x - c_1) \cdots (x - c_N) + 1:
  \]
  for all $c \in k$ we have
  $f(c) = (c - c_1) \cdots (c - c_N) + 1 = 1 \neq 0$, since $c$ equals
  one of the $c_i$'s. Thus $f(x)$ is a nonconstant polynomial with no
  roots, contradicting \Cref{lem:irred:5.10}.
\end{proof}
Since finite fields are not algebraically closed, the reader may
suspect that algebraically closed fields necessarily have
characteristic $0$ (cf.\
\Cref{exer:irred:prime-subfield-char-0-or-p}). This is not so: there
are algebraically closed fields of any characteristic. In fact, as we
will see in due time, every field $F$ may be embedded into an
algebraically closed field; the smallest such extension is called the
\emph{algebraic closure} of $F$ and is denoted $\overline{F}$. Thus,
for example, $\overline{\Z/2\Z}$ is an algebraically closed field of
characteristic $2$. The algebraic closure $\overline{\Q}$ is a
countable subfield of $\C$.  The extension
$\Q \subseteq \overline{\Q}$, and especially its `Galois group', cf.\
\S{}\ref{sec:fields:a-little-galois-theory}, is considered one of the
most important objects of study in mathematics (cf.\ the discussion
following \Cref{coro:fields:every-finite-group-is-a-galois-group}). We will construct the algebraic
closure of a field rather explicitly, in
\S{}\ref{subsec:fields:algebraic-closure}.

\subsection{Irreducibility in C[x], R[x], Q[x]}
\label{subsec:irred:irreducibility-in-c-x-r-x-q-x}
Every polynomial $f \in \C[x]$ factors completely over $\C$. Indeed,

\begin{thm}{}{thm:irred:5.12}
  $\C$ is algebraically closed.
\end{thm}

Gauss is credited with providing the first proof\footnote{Actually,
  Gauss's first proof apparently had a gap; the first rigorous proof
  is due to Argand. Later, Gauss fixed the gap in his proof and
  provided several other proofs.} of this fundamental theorem (which
is indeed known as the fundamental theorem of algebra). `Algebraic'
proofs of the fundamental theorem of algebra require more than we know
at this point (we will encounter one in
\S{}\ref{subsec:fields:fundamental-theorem-of-algebra}, after we have
seen a little Galois theory); curiously, a little complex analysis
makes the statement nearly trivial.
\begin{proof}
  Here is a sketch. Let $f \in \C[x]$ be a nonconstant polynomial; the
  task (cf.\ \Cref{lem:irred:5.10}) consists of proving that $f$ has a
  root in $\C$. Whatever $f(0)$ is, we can find an $r \in \R$ large
  enough that $|f(z)| > |f(0)|$ for all $z$ on the circle $|z| = r$
  (since $f$ is nonconstant, $\lim_{z \to \infty} |f(z)| =
  +\infty$). The disk $|z| \leq r$ is compact, so the continuous
  function $|f(z)|$ has a minimum on it; by the choice of $r$, it must
  be somewhere in the interior of the disk, say at $z = a$. The
  minimum modulus principle (cf.\
  \Cref{exer:irred:minimum-modulus-principle}) then implies that
  $f(a) = 0$.
\end{proof}
By \Cref{thm:irred:5.12}, irreducibility of polynomials in $\C[x]$ is
as simple as it can be: a nonconstant polynomial $f \in \C[x]$ is
irreducible if and only if it has degree $1$. Every $f \in \C[x]$ of
degree $\geq 2$ is reducible. Over $\R$, the situation is almost as
simple:

\begin{prop}{}{prop:irred:5.13}
  Every polynomial $f \in \R[x]$ of degree $\geq 3$ is reducible.  The
  nonconstant irreducible polynomials in $\R[x]$ are precisely the
  polynomials of degree $1$ and the quadratic polynomials
  $f = ax^2 + bx + c$ with $b^2 - 4ac < 0$.
\end{prop}
\begin{proof}
  Let $f \in \R[x]$ be a nonconstant polynomial:
  $f = a_0 + a_1 x + \dots + a_n x^n$, with all $a_i \in \R$.  By
  \Cref{thm:irred:5.12}, $f$ has a complex root $z$:
  $a_0 + a_1 z + \dots + a_n z^n = 0$. Applying complex conjugation
  $z \mapsto \overline{z}$ and noting that $\overline{a_i} = a_i$
  since $a_i \in \R$,
  \[
    \overline{a_0 + a_1 z + \dots + a_n z^n} = a_0 + a_1 \overline{z}
    + \dots + a_n \overline{z}^n = 0:
  \]
  this says that $\overline{z}$ is also a root of $f$. There are two
  possibilities:
  \begin{enumerate}[label=(\arabic*)]
  \item Either $z = \overline{z}$, that is, $z = r$ was real to begin
    with, and hence $(x - r)$ is a factor of $f$ in $\R[x]$; or
  \item $z \neq \overline{z}$, and then $(x - z)$ and
    $(x - \overline{z})$ are nonassociate irreducible factors of $f$
    in $\C[x]$. In this case (since $\C[x]$ is a UFD!)
    \[
      x^2 - (z + \overline{z})x + z\overline{z} = (x - z)(x -
      \overline{z})
    \]
    divides $f$.
  \end{enumerate}
  Since $z + \overline{z}$ and $z\overline{z}$ are both real numbers,
  this analysis shows that every nonconstant $f \in \R[x]$ has an
  irreducible factor of degree $\leq 2$, proving the first
  statement. The second statement then follows immediately from
  \Cref{prop:irred:5.3} and the fact that the quadratic polynomials in
  $\R[x]$ with no real roots are precisely the polynomials
  $ax^2 + bx + c$ with $b^2 - 4ac < 0$.
\end{proof}
The following remark is an immediate consequence of
\Cref{prop:irred:5.13} and is otherwise evident from elementary
calculus (\Cref{exer:irred:odd-degree-real-poly-has-real-root}):

\begin{coro}{}{coro:irred:5.14}
  Every polynomial $f \in \R[x]$ of odd degree has a real root.
\end{coro}
Summarizing, the issue of irreducibility of polynomials in $\C[x]$ or
$\R[x]$ is very straightforward. By contrast, irreducibility in
$\Q[x]$ is a complicated business: as we have seen
(\Cref{prop:irred:4.16}), it is just as subtle as irreducibility in
$\Z[x]$. There is no sharp characterization of irreducibility in
$\Z[x]$; but the following simple-minded remarks (and Eisenstein's
criterion, discussed in \S{}\ref{subsec:irred:eisenstein-s-criterion})
suffice in many interesting cases.

One simple-minded remark is that if $\varphi : R \to S$ is a
homomorphism, and $a$ is reducible in $R$, then $\varphi(a)$ is likely
to be reducible in $S$. This is just because if $a = bc$, then
$\varphi(a) = \varphi(b)\varphi(c)$. Of course this is not quite
right: for example because $\varphi(a)$ and/or $\varphi(b)$ and/or
$\varphi(c)$ may be units in $S$. But with due care for special cases
this observation may be used to great effect, especially in the
contrapositive form: if $\varphi(a)$ is irreducible in $S$, then $a$
is (likely\dots) irreducible in $R$.

Here is a simple statement formalizing these remarks, for $R = \Z[x]$.
The natural projection $\Z \to \Z/p\Z$ induces a surjective
homomorphism $\pi : \Z[x] \to \Z/p\Z[x]$: quite simply, $\pi(f)$ is
obtained from $f$ by reading all coefficients modulo $p$; I will write
`$f$ mod $p$' for the result of this operation.
\begin{prop}{}{prop:irred:5.15}
  Let $f \in \Z[x]$ be a primitive polynomial, and let $p$ be a prime
  integer. Assume $f \bmod p$ has the same degree as $f$ and is
  irreducible in $(\Z/p\Z)[x]$. Then $f$ is irreducible in $\Z[x]$.
\end{prop}
\begin{proof}
  Argue contrapositively: if $f$ is primitive and reducible in $\Z[x]$
  and $\deg f = n$, then $f = gh$ with $\deg g = d$, $\deg h = e$,
  $d + e = n$, and both $d$, $e$ positive. But then the same can be
  said of $f \bmod p$, so $f \bmod p$ is also reducible.
\end{proof}
The hypothesis on degrees is necessary, precisely to take care of the
`special cases' mentioned in the discussion preceding the
statement. For example, $(2x^3 + 3x^2 + 3x + 1)$ equals
$(x^2 + x + 1) \bmod 2$, and the latter is irreducible in
$(\Z/2\Z)[x]$; yet $(2x + 1)$ is a factor of the former. The point is
of course that $2x + 1$ is a unit mod $2$.

\textbf{Warning:} As we will see in a fairly distant future
(\Cref{exam:fields:exam:5.3}), there are irreducible polynomials in
$\Z[x]$ which are reducible modulo all primes! So
\Cref{prop:irred:5.15} cannot be turned into a characterization of
irreducibility in $\Z[x]$.

It is however rather useful in producing examples. For instance,

\begin{coro}{}{coro:irred:5.16}
  There are irreducible polynomials in $\Z[x]$ and $\Q[x]$ of
  arbitrarily large degree.
\end{coro}
\begin{proof}
  By \Cref{prop:irred:4.16}, the statement for $\Z[x]$ implies the one
  for $\Q[x]$. By \Cref{prop:irred:5.15}, it suffices to verify that
  there are irreducible polynomials in $(\Z/p\Z)[x]$ of arbitrarily
  large degree, for any prime integer $p$. This is a particular case
  of \Cref{exer:irred:finite-field-irreds-arb-high-degree}.
\end{proof}

\subsection{Eisenstein's criterion}
\label{subsec:irred:eisenstein-s-criterion}
I am giving this result in a separate subsection only on account of
the fact that it is very famous; it is also very simple-minded, like
the considerations leading up to \Cref{prop:irred:5.15}, and further
it is not really a `criterion', in the sense that it also does not
give a characterization of irreducibility. The result is usually
stated for $\Z$, but it holds for every ring $R$.
\begin{prop}{Eisenstein's criterion}{prop:irred:5.17}
  Let $R$ be a (commutative) ring, and let $\mathfrak{p}$ be a prime
  ideal of $R$. Let $f = a_0 + a_1 x + \dots + a_n x^n \in R[x]$ be a
  polynomial, and assume that
  \begin{enumerate}[label=(\arabic*)]
  \item $a_n \notin \mathfrak{p}$;
  \item $a_i \in \mathfrak{p}$ for $i = 0, \ldots, n - 1$;
  \item $a_0 \notin \mathfrak{p}^2$.
  \end{enumerate}
  Then $f$ is not the product of polynomials of degree $< n$ in
  $R[x]$.
\end{prop}
\begin{proof}
  Argue by contradiction. Assume $f = gh$ in $R[x]$, with both
  $d = \deg g$ and $e = \deg h$ less than $n = \deg f$; write
  \[g = b_0 + b_1 x + \cdots + b_d x^d, \quad h = c_0 + c_1 x + \cdots
    + c_e x^e,\] and note that necessarily $d > 0$ and $e >
  0$. Consider $f$ modulo $\mathfrak{p}$: thus
  \[\underline{f} = \underline{g}\,\underline{h}
    \quad \text{in } (R/\mathfrak{p})[x],\] where $\underline{f}$
  denotes $f$ modulo $\mathfrak{p}$, etc.

  By hypothesis, $\underline{f} = \underline{a_n} x^n$ modulo
  $\mathfrak{p}$, where $\underline{a_n} \neq 0$ in
  $R/\mathfrak{p}$. Since $R/\mathfrak{p}$ is an integral domain,
  factors of $\underline{f}$ must also be monomials: that is,
  necessarily
  \[\underline{g} = \underline{b_d} x^d,
    \quad \underline{h} = \underline{c_e} x^e.\] Since $d > 0$,
  $e > 0$, this implies $b_0 \in \mathfrak{p}$,
  $c_0 \in \mathfrak{p}$.

  But then $a_0 = b_0 c_0 \in \mathfrak{p}^2$, contradicting the
  hypothesis.
\end{proof}

For example, $x^4 + 2x^2 + 2$ must be irreducible in $\Z[x]$ (and
hence in $\Q[x]$): apply Eisenstein's criterion with $R = \Z$,
$\mathfrak{p} = (2)$. More exciting applications involve whole classes
of polynomials:

\begin{exam}{}{exam:irred:5.18}
  For all $n$ and all primes $p$, the polynomial $x^n - p$ is
  irreducible in $\Z[x]$. This follows immediately from Eisenstein's
  criterion and gives an alternative proof of \cref{coro:irred:5.16}.
\end{exam}

\begin{exam}{}{exam:irred:5.19}
  This is probably the most famous application of Eisenstein's
  criterion.  Let $p$ be a prime integer, and let
  \[f(x) = 1 + x + x^2 + \cdots + x^{p-1} \in \Z[x].\] These
  polynomials are called \emph{cyclotomic}; we will encounter them
  again in~\S{}\ref{subsec:fields:cyclotomic-polynomials-and-fields}.

  It may not look like it, but Eisenstein's criterion may be used to
  prove that $f(x)$ is irreducible. The trick (which the reader should
  endeavor to remember) is to apply the shift $x \to x + 1$. It is
  hopefully clear that $f(x)$ is irreducible if and only if $f(x + 1)$
  is; thus we are reduced to showing that
  \[f(x + 1) = 1 + (x + 1) + (x + 1)^2 + \cdots + (x + 1)^{p-1}\] is
  irreducible. This is better than it looks: since
  $f(x) = (x^p - 1)/(x - 1)$,
  \[f(x + 1) = \frac{(x + 1)^p - 1}{(x + 1) - 1} = x^{p-1} +
    \binom{p}{p-1} x^{p-2} + \cdots + \binom{p}{2} x + \binom{p}{1};\]
  Eisenstein's criterion proves that this is irreducible, since
  $\binom{p}{1} = p$ and

\begin{fact}{Claim~5.20}{fact:irred:claim:5.20}
  For $p$ prime and $k = 1, \ldots, p - 1$, $p$ divides
  $\binom{p}{k}$.
\end{fact}

There is nothing to this, because
$\binom{p}{k} = \frac{p!}{k!(p - k)!}$ and $p$ divides the numerator
and does not divide the denominator; the claim follows as $p$ is
prime. It may be worthwhile noting that this fact does \emph{not} hold
if $p$ is not prime: for example, $4$ does not divide~$\binom{4}{2}$.
\end{exam}

\subsection*{Exercises}
\begin{exer}{}{exer:irred:root-multiplicity-via-derivative}
  $\lnot$ Let $f(x) \in \C[x]$. Prove that $a \in \C$ is a root of $f$
  with multiplicity $r$ if and only if
  $f(a) = f'(a) = \cdots = f^{(r-1)}(a) = 0$ and $f^{(r)}(a) \neq 0$,
  where $f^{(k)}(a)$ denotes the value of the $k$-th derivative of $f$
  at $a$. Deduce that $f(x) \in \C[x]$ has multiple roots if and only
  if $\gcd(f(x), f'(x)) \neq
  1$. [\ref{exer:irred:irred-poly-no-multiple-roots}]
\end{exer}
\begin{exer}{}{exer:irred:irred-poly-no-multiple-roots}
  Let $F$ be a subfield of $\C$, and let $f(x)$ be an irreducible
  polynomial in $F[x]$. Prove that $f(x)$ has no multiple roots in
  $\C$. (Use
  \Cref{exer:irred:gcd-in-pid-remains-gcd-in-extension,exer:irred:root-multiplicity-via-derivative}.)
\end{exer}
\begin{exer}{}{exer:irred:even-poly-factor-closed-under-negation}
  Let $R$ be a ring, and let
  $f(x) = a_{2n} x^{2n} + a_{2n-2} x^{2n-2} + \cdots + a_2 x^2 + a_0
  \in R[x]$ be a polynomial only involving even powers of $x$. Prove
  that if $g(x)$ is a factor of $f(x)$, so is $g(-x)$.
\end{exer}
\begin{exer}{}{exer:irred:x4-plus-x2-plus-1-reducible}
  Show that $x^4 + x^2 + 1$ is reducible in $\Z[x]$. Prove that it has
  no rational roots, without finding its (complex) roots.
\end{exer}
\begin{exer}{}{exer:irred:deg-2-3-irred-iff-no-roots}
  \exerused{} Prove
  \Cref{prop:irred:5.3}. [\S{}\ref{subsec:irred:roots-and-reducibility}]
\end{exer}
\begin{exer}{}{exer:irred:construct-fields-27-and-121-elements}
  \exerused{} Construct fields with 27 elements and with 121 elements.
  [\S{}\ref{subsec:irred:roots-and-reducibility}]
\end{exer}
\begin{exer}{}{exer:irred:poly-determined-by-d-plus-1-values}
  Let $R$ be an integral domain, and let $f(x) \in R[x]$ be a
  polynomial of degree $d$. Prove that $f(x)$ is determined by its
  value at any $d + 1$ distinct elements of $R$.
\end{exer}
\begin{exer}{}{exer:irred:lagrange-interpolation-construction}
  $\lnot$ Let $K$ be a field, and let $a_0, \ldots, a_d$ be distinct
  elements of $K$. Given any elements $b_0, \ldots, b_d$ in $K$,
  construct explicitly a polynomial $f(x) \in K[x]$ of degree at most
  $d$ such that $f(a_0) = b_0, \ldots, f(a_d) = b_d$, and show that
  this polynomial is unique. (Hint: First solve the problem assuming
  that only one $b_i$ is not equal to zero.) This process is called
  Lagrange
  interpolation. [\ref{exer:irred:factor-polys-via-lagrange-interpolation}]
\end{exer}
\begin{exer}{}{exer:irred:factor-polys-via-lagrange-interpolation}
  $\lnot$ Pretend you can factor integers, and then use Lagrange
  interpolation (cf.\
  \Cref{exer:irred:lagrange-interpolation-construction}) to give a
  finite algorithm to factor polynomials with integer coefficients
  over $\Q[x]$. Use your algorithm to factor
  $(x - 1)(x - 2)(x - 3)(x - 4) +
  1$. [\ref{exer:irred:product-linear-minus-1-irreducible}]
\end{exer}
\begin{exer}{}{exer:irred:product-linear-minus-1-irreducible}
  Prove that the polynomial $(x - 1)(x - 2) \cdots (x - n) - 1$ is
  irreducible in $\Q[x]$ for all $n \geq 1$. (Hint: Think along the
  lines of \Cref{exer:irred:factor-polys-via-lagrange-interpolation}.)
\end{exer}
\begin{exer}{}{exer:irred:finite-field-irreds-arb-high-degree}
  \exerused{} Let $F$ be a finite field. Prove that there are
  irreducible polynomials in $F[x]$ of arbitrarily high degree. (Hint:
  \Cref{exer:irred:euclid-infinitude-of-primes}.)
  [\S{}\ref{subsec:irred:irreducibility-in-c-x-r-x-q-x}]
\end{exer}
\begin{exer}{}{exer:irred:construction-on-linear-poly-gives-iso-field}
  Prove that applying the construction in \Cref{prop:irred:5.7} to an
  irreducible linear polynomial in $k[x]$ produces a field isomorphic
  to $k$.
\end{exer}
\begin{exer}{}{exer:irred:extension-where-f-splits-linear}
  \exerused{} Let $k$ be a field, and let $f \in k[x]$ be any
  polynomial.  Prove that there is an extension $k \subseteq F$ in
  which $f$ factors completely as a product of linear
  terms. [\S{}\ref{subsec:irred:adding-roots-algebraically-closed-fields},
  \S{}\ref{subsec:linalg:jordan-canonical-form}]
\end{exer}
\begin{exer}{}{exer:irred:embeddings-of-cubic-ext-in-r-and-c}
  How many different embeddings of the field $\Q[t]/(t^3 - 2)$ are
  there in $\R$? How many in $\C$?
\end{exer}
\begin{exer}{}{exer:irred:alg-closed-iff-poly-splits-iff-has-root}
  Prove \Cref{lem:irred:5.10}.
\end{exer}
\begin{exer}{}{exer:irred:minimum-modulus-principle}
  \exerused{} If you know about the `maximum modulus principle' in
  complex analysis: formulate and prove the `minimum modulus
  principle' used in the sketch of the proof of the fundamental
  theorem of
  algebra. [\S{}\ref{subsec:irred:irreducibility-in-c-x-r-x-q-x}]
\end{exer}
\begin{exer}{}{exer:irred:odd-degree-real-poly-has-real-root}
  \exerused{} Let $f \in \R[x]$ be a polynomial of odd degree. Use the
  intermediate value theorem to give an `algebra-free' proof of the
  fact that $f$ has real
  roots. [\S{}\ref{subsec:irred:irreducibility-in-c-x-r-x-q-x},
  \S{}\ref{subsec:fields:fundamental-theorem-of-algebra}]
\end{exer}
\begin{exer}{}{exer:irred:cubic-odd-values-irreducible}
  Let $f \in \Z[x]$ be a cubic polynomial such that $f(0)$ and $f(1)$
  are odd and with odd leading coefficient. Prove that $f$ is
  irreducible in $\Q[x]$.
\end{exer}
\begin{exer}{}{exer:irred:sqrt2-irrational-via-eisenstein}
  Give a proof of the fact that $\sqrt{2}$ is not rational by using
  Eisenstein's criterion.
\end{exer}
\begin{exer}{}{exer:irred:x6-plus-4x3-plus-1-irreducible}
  Prove that $x^6 + 4x^3 + 1$ is irreducible by using Eisenstein's
  criterion.
\end{exer}
\begin{exer}{}{exer:irred:sum-of-powers-reducible-if-n-not-prime}
  Prove that $1 + x + x^2 + \cdots + x^{n-1}$ is reducible over $\Z$
  if $n$ is not prime.
\end{exer}
\begin{exer}{}{exer:irred:xn-minus-a-irreducible-squarefree}
  Let $R$ be a UFD, and let $a \in R$ be an element that is not
  divisible by the square of some irreducible element in its
  factorization. Prove that $x^n - a$ is irreducible for every integer
  $n \geq 1$.
\end{exer}
\begin{exer}{}{exer:irred:bivariate-poly-reducibility-example}
  Decide whether $y^5 + x^2 y^3 + x^3 y^2 + x$ is reducible or
  irreducible in $\C[x, y]$.
\end{exer}

\begin{exer}{}{exer:irred:quadric-cone-coord-ring-is-domain}
  Prove that $\C[x, y, z, w]/(xw - yz)$ is an integral domain, by
  using Eisenstein's criterion. (I used this ring as an example
  in~\S{}\ref{subsec:irred:characterization-of-ufds}, but I did not
  have the patience to prove that it was a domain back then!)
\end{exer}

\section{Further remarks and examples}
\label{sec:irred:further-remarks-and-examples}


\subsection{Chinese remainder theorem}
\label{subsec:irred:chinese-remainder-theorem}
Suppose all you know about an integer is its class modulo several
numbers; can you reconstruct the integer? Also, if you are given
arbitrary classes in $\Z/n\Z$ for several integers $n$, can you find
an $N \in \Z$ satisfying all these congruences simultaneously?

The answer is no in both cases, for trivial reasons. If an integer $N$
satisfies given congruences modulo $n_1, \ldots, n_k$, then adding any
multiple of $n_1 \cdots n_k$ to $N$ produces an integer satisfying the
same congruences (so $N$ cannot be entirely reconstructed from the
given data); and there is no integer $N$ such that
$N \equiv 1 \bmod 2$ and $N \equiv 2 \bmod 4$, so there are plenty of
congruences which cannot be simultaneously satisfied.

The \emph{Chinese remainder theorem} (CRT) sharpens these questions so
that they can in fact be answered affirmatively, and (in its modern
form) it generalizes them to a much broader setting. Its statement is
most pleasant for PIDs; it is very impressive\footnote{Allegedly, a
  large number of `Putnam' problems can be solved by clever
  applications of the CRT.} even in the limited context of
$\Z$. Protoversions of the theorem have already appeared here and
there in this book, e.g., \Cref{lem:groups2:coprime-order-subgroups-direct-sum}.

I will first give the more general version, which is in some way
simpler.  Let $R$ be any commutative ring.
\begin{thm}{}{thm:irred:6.1}
  Let $I_1, \ldots, I_k$ be ideals of $R$ such that $I_i + I_j = (1)$
  for all $i \neq j$. Then the natural homomorphism
  \[\varphi : R \longrightarrow \frac{R}{I_1} \times \cdots \times
    \frac{R}{I_k}\] is surjective and induces an isomorphism
  \[\tilde{\varphi} : \frac{R}{I_1 \cdots I_k}
    \xrightarrow{\;\sim\;} \frac{R}{I_1} \times \cdots \times
    \frac{R}{I_k}.\]
\end{thm}

The `natural' homomorphism $\varphi$ is determined by the canonical
projections $R \to R/I_j$ and the universal property of products; the
homomorphism $\tilde\varphi$ is induced by virtue of the universal
property of quotients, since $I_1 \cdots I_k \subseteq I_j$ for all
$j$, hence $I_1 \cdots I_k \subseteq
\ker\varphi$. \cref{thm:irred:6.1} is proven by an induction relying
on the following lemma.
\begin{lem}{}{lem:irred:6.2}
  Let $I_1, \ldots, I_k$ be ideals of $R$ such that $I_i + I_k = (1)$
  for all $i = 1, \ldots, k-1$. Then
  $(I_1 \cdots I_{k-1}) + I_k = (1)$.
\end{lem}

\begin{proof}
  By hypothesis, for $i = 1, \ldots, k-1$ there exists $a_i \in I_k$
  such that $1 - a_i \in I_i$. Then
  \[(1 - a_1) \cdots (1 - a_{k-1}) \in I_1 \cdots I_{k-1},\] and
  \[1 - (1 - a_1) \cdots (1 - a_{k-1}) \in I_k,\] because it is a
  combination of $a_1, \ldots, a_{k-1} \in I_k$.
\end{proof}
Since clearly $\ker\varphi = I_1 \cap \cdots \cap I_k$, the second
part of the statement of \cref{thm:irred:6.1} follows immediately from
the first part, the `first isomorphism theorem', and the following
(independently interesting) observation:
\begin{lem}{}{lem:irred:6.3}
  Let $I_1, \ldots, I_k$ be ideals of $R$ such that $I_i + I_j = (1)$
  for all $i \neq j$. Then
  $I_1 \cdots I_k = I_1 \cap \cdots \cap I_k$.
\end{lem}

\begin{proof}
  By \cref{lem:irred:6.2}, under the stated hypotheses we have that
  $I_1 \cdots I_{k-1} + I_k = (1)$ for $k \geq 3$. Thus, the general
  statement is reduced by induction to the case $k = 2$. (By the way,
  this case is \Cref{exer:rings:coprime-ideals-product-intersection}!)
  Assume $I$ and $J$ are ideals of $R$ such that $I + J = (1)$. The
  inclusion $IJ \subseteq I \cap J$ holds for all ideals $I$, $J$, so
  the task amounts to proving $I \cap J \subseteq IJ$ when
  $I + J = (1)$.  If $I + J = (1)$, then there exist elements
  $a \in I$, $b \in J$ such that $a + b = 1$. But if $r \in I \cap J$,
  then
  \[r = r \cdot 1 = r(a + b) = ra + rb \in IJ:\] because $ra \in IJ$
  as $r \in J$ and $a \in I$, while $rb \in IJ$ as $r \in I$ and
  $b \in J$. The statement follows.
\end{proof}
Thus, we only need to prove the first part of \cref{thm:irred:6.1}.

\begin{proof}[Proof of \cref{thm:irred:6.1}]
  Argue by induction on $k$. For $k = 1$, there is nothing to
  show. For $k > 1$, assume the statement is known for fewer
  ideals. Thus, we may assume that the natural projection induces an
  isomorphism
  \[\frac{R}{I_1 \cdots I_{k-1}} \cong
    \frac{R}{I_1} \times \cdots \times \frac{R}{I_{k-1}};\] and all
  that we have left to prove is that the natural homomorphism
  \[R \to \frac{R}{I_1 \cdots I_{k-1}} \times \frac{R}{I_k}\] is
  surjective. By \cref{lem:irred:6.2},
  $(I_1 \cdots I_{k-1}) + I_k = (1)$; thus we are reduced to the case
  of two ideals.  Let then $I$, $J$ be ideals of a commutative ring
  $R$, such that $I + J = (1)$, and let $r_I, r_J \in R$; we have to
  verify that $\exists r \in R$ such that $r \equiv r_I \bmod I$ and
  $r \equiv r_J \bmod J$.  Since $I + J = (1)$, there are $a \in I$,
  $b \in J$ such that $a + b = 1$.  Let $r = ar_J + br_I$: then
  \[r = ar_J + (1 - a)r_I = r_I + a(r_J - r_I) \equiv r_I \bmod I\] as
  $a \in I$, and
  \[r = (1 - b)r_J + br_I = r_J + b(r_I - r_J) \equiv r_J \bmod J\] as
  $b \in J$, as needed, and completing the proof.
\end{proof}
In a PID, the CRT takes the following form:

\begin{coro}{}{coro:irred:6.4}
  Let $R$ be a PID, and let $a_1, \dots, a_k \in R$ be elements such
  that $\gcd(a_i, a_j) = 1$ for all $i \neq j$. Let
  $a = a_1 \cdots a_k$. Then the function
  \[\phi : \frac{R}{(a)} \to
    \frac{R}{(a_1)} \times \cdots \times \frac{R}{(a_k)}\] defined by
  $r + (a) \mapsto (r + (a_1), \dots, r + (a_k))$ is an isomorphism.
\end{coro}
This is an immediate consequence of \cref{thm:irred:6.1}, since (in a
PID!)  $\gcd(a, b) = 1$ if and only if $(a, b) = (1)$ as ideals. This
is not the case for arbitrary UFDs, and indeed the natural map
\[\Z[x] \to \frac{\Z[x]}{(2)} \times
  \frac{\Z[x]}{(x)}\] is not surjective (check this!), even though
$\gcd(2, x) = 1$. But the kernel of this map is $(2x)$; and this is
what can be expected in general from the CRT over UFDs
(\Cref{exer:irred:ufd-coprime-intersection-and-crt-injective}).  As
$\Z$ is a PID, \cref{coro:irred:6.4} holds for $\Z$ and gives the
promised answer to a revised version of the questions posed at the
beginning of this subsection. In fact, tracing the proof in this case
gives an effective procedure to solve simultaneous congruences over
$\Z$ (or in fact over any Euclidean domain, as will be apparent from
the argument).  To see this more explicitly, let $n_1, \dots, n_k$ be
pairwise relatively prime integers, and let $n = n_1 \cdots n_k$; for
each $i$, let $m_i = n/n_i$. Then $n_i$ and $m_i$ are relatively
prime,\footnote{This is a particular case of \cref{lem:irred:6.2} and
  is clear anyway from gcd considerations.}  and we can use the
Euclidean algorithm to explicitly find integers $a_i$, $b_i$ such that
\[a_i n_i + b_i m_i = 1.\] The numbers $q_i = b_i m_i$ have the
property that
\[q_i \equiv 1 \bmod n_i, \quad q_i \equiv 0 \bmod n_j \quad \forall j
  \neq i\] (why?). These integers may be used to solve any given
system of congruences modulo $n_1, \dots, n_k$: indeed, if
$r_1, \dots, r_k \in \Z$ are given, then
\[N := r_1 q_1 + \dots + r_k q_k\] satisfies
\[N \equiv r_1 \cdot 0 + \dots + r_i \cdot 1 + \dots + r_k \cdot 0
  \equiv r_i \bmod n_i\] for all $i$.

The same procedure can be applied over any Euclidean domain $R$; see
\Cref{exer:irred:crt-construction-example-qx} for an example.

\subsection{Gaussian integers}
\label{subsec:irred:gaussian-integers}
The rings $\Z$ and $k[x]$ (where $k$ is a field) may be the only
examples of Euclidean domains known to our reader. The next most
famous example is the ring of \emph{Gaussian integers}; this is a very
pretty ring, and it has elementary but striking applications in number
theory (one instance of which we will see in
\S{}\ref{subsec:irred:fermat-s-theorem-on-sums-of-squares}).
Abstractly, we may define this ring as
\[\Z[i] := \frac{\Z[x]}{(x^2 + 1)};\]
and we are going to verify that $\Z[i]$ is a Euclidean domain.  The
notation $\Z[i]$ is justified by the discussion in
\S{}\ref{subsec:irred:adding-roots-algebraically-closed-fields} (cf.\
especially \Cref{prop:irred:5.7}): $\Z[i]$ may be viewed as the
`smallest' ring containing $\Z$ and a root of $x^2 + 1$, that is, a
square root $i$ of $-1$. The natural embedding
\[\frac{\Z[x]}{(x^2 + 1)} \subseteq \frac{\R[x]}{(x^2 +
    1)}\] and the identification of the rightmost ring with $\C$
(\Cref{exam:rings:higher-degree-quotient}) realize $\Z[i]$ as a
subring of $\C$; tracing this embedding shows that we could
equivalently define
\[\Z[i] = \{a + bi \in \C \mid a, b \in \Z\}.\]
Thus, the reader should think of $\Z[i]$ as consisting of the complex
numbers whose real and imaginary parts are integers; this may be
referred to as the `integer lattice' in $\C$.

This picture is particularly compelling, because it allows us to
`visualize' principal ideals in $\Z[i]$: simple properties of complex
multiplication show that the multiples of a fixed $w \in \Z[i]$ form a
regular lattice superimposed on $\Z[i]$.  For example, the fattened
dots represent the ideal $(-2 - i)$ in $\Z[i]$: an enlarged, tilted
lattice superimposed on the integer lattice in $\C$. The reader should
stop reading now and make sure to understand why this works out so
neatly.  A Gaussian integer is a complex number, and as such it has a
\emph{norm}
\[N(a + bi) := (a + bi)(a - bi) = a^2 + b^2.\] Geometrically, this is
the square of the distance from the origin to $a + bi$.  The norm of a
Gaussian integer is a nonnegative integer; thus $N$ is a function
$\Z[i] \to \Z^{\geq 0}$.

\begin{lem}{}{lem:irred:6.5}
  The function $N$ is a Euclidean valuation on $\Z[i]$; further, $N$
  is multiplicative in the sense that $\forall z, w \in \Z[i]$
  \[N(zw) = N(z)N(w).\]
\end{lem}
\begin{proof}
  The multiplicativity is an immediate consequence of the elementary
  properties of complex conjugation:
  \[N(zw) = (zw)\overline{(zw)} = (z\bar{z})(w\bar{w}) = N(z)N(w).\]
  To see that $N$ is a Euclidean valuation, we have to show how to
  perform a `division with remainder' in $\Z[i]$. It is not hard, but
  it is a bit messy, to do this algebraically; I will attempt to
  convince the reader that this is in fact visually evident (and then
  the reader can have fun producing the needed algebraic
  computations).  Let $z, w \in \Z[i]$, and assume $w \neq 0$. The
  ideal $(w)$ is a lattice superimposed on $\Z[i]$. The given $z$ is
  either one of the vertices of this lattice (in which case $z$ is a
  multiple of $w$, so that the division $z/w$ can be performed in
  $\Z[i]$, with remainder $0$) or it sits inside one of the `boxes' of
  the lattice. In the latter case, pick any of the vertices of that
  box, that is, a multiple $qw$ of $w$, and let $r = z - qw$. Then we
  have obtained
  \[z = qw + r,\] and the norm of $r$ is the square of the length of
  the segment between $qw$ and $z$. Since this segment is contained in
  a box, and the square of the size of the box is $N(w)$, we have
  achieved
  \[N(r) < N(w),\] completing the proof.
\end{proof}
As we know, all sorts of amazing properties hold for the ring of
Gaussian integers as a consequence of \cref{lem:irred:6.5}: $\Z[i]$ is
a PID and a UFD; irreducible elements are prime in $\Z[i]$; greatest
common divisors exist and may be found by applying the Euclidean
algorithm, etc. Any one of these facts would seem fairly challenging
in itself, but they are all immediate consequences of the existence of
a Euclidean valuation and of the general considerations in
\S{}\ref{sec:irred:ufds-pids-euclidean-domains}.

The fact that the norm is multiplicative simplifies further the
analysis of the ring $\Z[i]$.
\begin{lem}{}{lem:irred:6.6}
  The units of $\Z[i]$ are $\pm 1, \pm i$.
\end{lem}

\begin{proof}
  If $u$ is a unit in $\Z[i]$, then there exists $v \in \Z[i]$ such
  that $uv = 1$. But then $N(u)N(v) = N(uv) = N(1) = 1$ by
  multiplicativity, so $N(u)$ is a unit in $\Z$. This implies
  $N(u) = 1$, and the only elements in $\Z[i]$ with norm $1$ are
  $\pm 1, \pm i$. The statement follows.
\end{proof}
\begin{lem}{}{lem:irred:6.7}
  Let $q \in \Z[i]$ be a prime element. Then there is a prime integer
  $p \in \Z$ such that $N(q) = p$ or $N(q) = p^2$.
\end{lem}

\begin{proof}
  Since $q$ is not a unit, $N(q) \neq 1$ (by
  \cref{lem:irred:6.6}). Thus $N(q)$ is a nontrivial product of
  (integer) primes, and since $q$ is prime in $\Z[i] \supseteq \Z$,
  $q$ must divide one of the prime integer factors of $N(q)$; let $p$
  be this integer prime. But then $q \mid p$ in $\Z[i]$, and it
  follows (by multiplicativity of the norm) that
  $N(q) \mid N(p) = p^2$. Since $N(q) \neq 1$, the only possibilities
  are $N(q) = p$ and $N(q) = p^2$, as claimed.
\end{proof}
Both possibilities presented in \cref{lem:irred:6.7} occur, and
studying this dichotomy further will be key to the result in
\S{}\ref{subsec:irred:fermat-s-theorem-on-sums-of-squares}. But we can
already work out several cases `by hand':
\begin{exam}{}{exam:irred:6.8}
  The prime integer $3$ is a prime element of $\Z[i]$; this can be
  verified by proving that $3$ is irreducible in $\Z[i]$ (since
  $\Z[i]$ is a UFD). For this purpose, note that since $N(3) = 9$, the
  norm of a factor of $3$ would have to be a divisor of $9$, that is,
  $1$, $3$, or $9$. Gaussian integers with norm $1$ are units, and
  those with norm $9$ are associates of $3$
  (\Cref{exer:irred:gaussian-int-associates-same-norm}); thus a
  nontrivial factor of $3$ would necessarily have norm equal to
  $3$. But there are no such elements in $\Z[i]$; hence $3$ is indeed
  irreducible.  The prime integer $5$ is \emph{not} a prime element of
  $\Z[i]$: running through the same argument as we just did for $3$,
  we find that a factor of $5$ should have norm $5$, and $2 + i$
  does. In fact, $5 = (2 + i)(2 - i)$ is a prime factorization of $5$
  in $\Z[i]$.

  Visually, what happens is that the lattice generated by $3$ in
  $\Z[i]$ cannot be refined further into a tighter lattice, while the
  one generated by $5$ does admit a refinement (the circles represent
  complex numbers with norm $3$, $5$, respectively).
\end{exam}

The reader is encouraged to work out several other examples and to try
to figure out what property of an integer prime makes it split in
$\Z[i]$ (like $5$ and unlike $3$). But this should be done before
reading on!

\subsection{Fermat's theorem on sums of squares}
\label{subsec:irred:fermat-s-theorem-on-sums-of-squares}
Once armed with our knowledge about rings, playing with $\Z[i]$ a
little further teaches us something new about $\Z$---this is a
beautiful and famous example of the success achieved by judicious use
of generalization over brute force.

First, let us complete the circle of thought begun with
\cref{lem:irred:6.7}.  Say that a prime integer $p$ \emph{splits} in
$\Z[i]$ if it is not a prime element of $\Z[i]$; we have seen in
\cref{exam:irred:6.8} that $5$ splits in $\Z[i]$, while $3$ does not.
\begin{lem}{}{lem:irred:6.9}
  A positive integer prime $p \in \Z$ splits in $\Z[i]$ if and only if
  it is the sum of two squares in $\Z$.
\end{lem}
\begin{proof}
  First assume that $p = a^2 + b^2$, with $a, b \in \Z$. Then
  \[p = (a + bi)(a - bi)\] in $\Z[i]$, and
  $N(a \pm bi) = a^2 + b^2 = p \neq 1$, so neither of the two factors
  is a unit in $\Z[i]$ (\cref{lem:irred:6.6}).  Thus $p$ is not
  irreducible, hence not prime, in $\Z[i]$.  Conversely, assume that
  $p$ is not irreducible in $\Z[i]$: then it has an irreducible factor
  $q \in \Z[i]$, which is not an associate of $p$. Since $q \mid p$,
  by multiplicativity of the norm we have $N(q) \mid N(p) = p^2$, and
  hence $N(q) = p$ since $q$ and $p$ are not associates (thus
  $N(q) \neq p^2$) and $q$ is not a unit (thus $N(q) \neq 1$). If
  $q = a + bi$, we find
  \[p = N(q) = a^2 + b^2,\] verifying that $p$ is the sum of two
  squares and completing the proof.
\end{proof}
Thus, $2$ splits (as $2 = 1^2 + 1^2$), and so do
\[5 = 1^2 + 2^2, \quad 13 = 2^2 + 3^2, \quad 17 = 1^2 + 4^2, \quad
  \dots\] while $3, 7, 11, 19, 23, \dots$ remain prime if viewed as
elements of $\Z[i]$.

The next puzzle is as follows: what else distinguishes the primes in
the first list from the primes in the second list? Again, the reader
who does not know the answer already should pause and try to come up
with a conjecture and then prove that conjecture! Do not read the next
statement before trying this on your own!
\begin{lem}{}{lem:irred:6.10}
  A positive odd prime integer $p$ splits in $\Z[i]$ if and only if it
  is congruent to $1$ modulo $4$.
\end{lem}

\begin{proof}
  The question is whether $p$ is prime as an element of $\Z[i]$, that
  is, whether $\Z[i]/(p)$ is an integral domain. But we have
  isomorphisms
  \[\frac{\Z[i]}{(p)} \cong \frac{\Z[x]/(x^2+1)}{(p)} \cong
    \frac{\Z[x]}{(p, x^2+1)} \cong \frac{\Z[x]/(p)}{(x^2+1)} \cong
    \frac{\Z/p\Z[x]}{(x^2+1)}\] by virtue of the usual isomorphism
  theorems (including an appearance of
  \Cref{lem:irred:4.1})\footnote{The point I am trying to make is that
    these are not difficult facts, and a conscientious reader really
    is in the position of discovering and proving them on his or her
    own. This is extremely remarkable: the theorem we are heading
    towards was stated by Fermat, without proof, in 1640, and had to
    wait about one hundred years before being proven rigorously (by
    Euler, using `infinite descent'). Proofs using Gaussian integers
    are due to Dedekind and had to wait another hundred years. The
    moral is, of course, that the modern algebraic apparatus acts as a
    tremendous amplifier of our skills.} and taking some liberties
  with the language (for example, $(p)$ means three different things,
  hopefully self-clarifying through the context). Therefore,
  \begin{align*}
    p \text{ splits in } \Z[i]
    &\iff \Z[i]/(p) \text{ is not an integral domain} \\
    &\iff \Z/p\Z[x]/(x^2+1) \text{ is not an integral domain}
    \\
    &\iff x^2 + 1 \text{ is not irreducible in } \Z/p\Z[x] \\
    &\iff x^2 + 1 \text{ has a root in } \Z/p\Z \\
    &\iff \exists\, n \in \Z \text{ such that } n^2 \equiv -1 \bmod p,
  \end{align*}
  and we are reduced to verifying that this last condition is
  equivalent to $p \equiv 1 \bmod 4$, provided that $p$ is an odd
  prime.  For this purpose, recall (\Cref{thm:groups2:finite-subgroup-of-field-mult-group-is-cyclic}) that the
  multiplicative group $G$ of $\Z/p\Z$ is cyclic; let $g \in G$ be a
  generator of $G$. Since $p$ is assumed to be odd, $p - 1$ is an even
  number, say $2\ell$: thus $g$ has order $|G| = 2\ell$. Also, denote
  the class of an integer $n \bmod (p)$ by $\underline{n}$. Since $g$
  is a generator of $G$, for every integer $n \notin (p)$ there is an
  integer $m$ such that $\underline{n} = g^m$.  The class
  $\underline{-1}$ generates the unique subgroup of order $2$ of $G$
  (unique by the classification of subgroups of a cyclic group,
  \Cref{prop:groups:subgroups-of-znz}); since $g^\ell$ does the same,
  we have
  \[g^\ell = \underline{-1}.\] Therefore, with $\underline{n} = g^m$,
  we see that $n^2 \equiv -1 \bmod p$ if and only if
  $g^{2m} = g^\ell$, that is, if and only if
  \[2m \equiv \ell \bmod 2\ell.\] Summarizing, $\exists n \in \Z$ such
  that $n^2 \equiv -1 \bmod p$ if and only if $\exists m \in \Z$ such
  that $2m \equiv \ell \bmod 2\ell$. Now this is clearly the case if
  and only if $\ell$ is even, that is, if and only if
  $(p - 1) = 2\ell$ is a multiple of $4$, that is, if and only if
  $p \equiv 1 \bmod 4$; and we are done.
\end{proof}
Putting together \cref{lem:irred:6.9} and \cref{lem:irred:6.10} gives
the following beautiful number-theoretic statement:

\begin{thm}{Fermat}{thm:irred:6.11}
  A positive odd prime $p \in \Z$ is a sum of two squares if and only
  if $p \equiv 1 \bmod 4$.
\end{thm}
\begin{rem}{}{irred:6.12}
  Lagrange proved (in 1770) that every positive integer is the sum of
  four squares. One way to prove this is not unlike the proof given
  for \cref{thm:irred:6.11}: it boils down to analyzing the splitting
  of (integer) primes in a ring; the role of $\Z[i]$ is taken here by
  a ring of `integral quaternions'.
\end{rem}
The reader should pause and note that there is no mention of UFDs,
Euclidean domains, complex numbers, cyclic groups, etc., in the
statement of \cref{thm:irred:6.11}, although a large selection of
these tools was used in its proof. This is of course the dream of the
generalizer: to set up an abstract machinery making interesting facts
(nearly) evident---facts that would be extremely mysterious or that
would require Fermat-grade cleverness to be understood without that
machinery.

\subsection*{Exercises}

\begin{exer}{}{exer:irred:crt-exact-sequence-two-ideals}
  Generalize the CRT for two ideals, as follows. Let $I$, $J$ be
  ideals in a commutative ring $R$; prove that there is an exact
  sequence of $R$-modules
  \[0 \longrightarrow I \cap J \longrightarrow R
    \xrightarrow{\;\varphi\;} \frac{R}{I} \times \frac{R}{J}
    \longrightarrow \frac{R}{I + J} \longrightarrow 0\] where
  $\varphi$ is the natural map. (Also, explain why this implies the
  first part of \cref{thm:irred:6.1}, for $k = 2$.)
\end{exer}
\begin{exer}{}{exer:irred:idempotent-ring-decomposition}
  Let $R$ be a commutative ring, and let $a \in R$ be an element such
  that $a^2 = a$. Prove that $R \cong R/(a) \times R/(1 - a)$.

  Show that the multiplication in $R$ endows the ideal $(a)$ with a
  ring structure, with $a$ as the identity.\footnote{This is an
    extremely unusual situation. Note that this ring $(a)$ is
    \emph{not} a subring of $R$ if $a \neq 1$ according to
    \Cref{defn:rings:subring}, since the identities in $(a)$ and $R$
    differ.}  Prove that $(a) \cong R/(1 - a)$ as rings. Prove that
  $R \cong (a) \times (1 - a)$ as rings.
\end{exer}
\begin{exer}{}{exer:irred:finite-boolean-ring-is-product-of-z2}
  Recall (\Cref{exer:rings:boolean-power-set}) that a ring $R$ is
  called \emph{Boolean} if $a^2 = a$ for all $a \in R$. Let $R$ be a
  finite Boolean ring; prove that
  $R \cong \Z/2\Z \times \dots \times \Z/2\Z$.
\end{exer}
\begin{exer}{}{exer:irred:bound-on-pairwise-comaximal-ideals}
  Let $R$ be a finite commutative ring, and let $p$ be the smallest
  prime dividing $|R|$. Let $I_1, \dots, I_k$ be proper ideals such
  that $I_i + I_j = (1)$ for $i \neq j$. Prove that
  $k \leq \log_p |R|$. (Hint: Prove
  $|R|^{k-1} \leq |I_1| \cdots |I_k| \leq (|R|/p)^k$.)
\end{exer}
\begin{exer}{}{exer:irred:crt-map-not-surjective-example}
  Show that the map $\Z[x] \to \Z[x]/(2) \times \Z[x]/(x)$ is not
  surjective.
\end{exer}
\begin{exer}{}{exer:irred:ufd-coprime-intersection-and-crt-injective}
  \exerused{}Let $R$ be a UFD.
  \begin{itemize}
  \item Let $a, b \in R$ such that $\gcd(a, b) = 1$. Prove that
    $(a) \cap (b) = (ab)$.
  \item Under the hypotheses of \cref{coro:irred:6.4} (but only
    assuming that $R$ is a UFD) prove that the function $\varphi$ is
    injective.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:crt-construction-example-qx}
  \exerused{}Find a polynomial $f \in \Q[x]$ such that
  $f \equiv 1 \bmod (x^2 + 1)$ and $f \equiv x \bmod x^{100}$.
\end{exer}
\begin{exer}{}{exer:irred:crt-znz-decomp-and-totient-formula}
  $\lnot$ Let $n \in \Z$ be a positive integer and
  $n = p_1^{a_1} \cdots p_r^{a_r}$ its prime factorization. By the
  classification theorem for finite abelian groups (or, in fact,
  simpler considerations; cf.\ \Cref{exer:groups:cmn-iso-cm-times-cn})
  \[\frac{\Z}{(n)} \cong \frac{\Z}{(p_1^{a_1})} \times
    \dots \times \frac{\Z}{(p_r^{a_r})}\] as abelian groups.
  \begin{itemize}
  \item Use the CRT to prove that this is in fact a ring isomorphism.
  \item Prove that
    \[\left(\frac{\Z}{(n)}\right)^* \cong
      \left(\frac{\Z}{(p_1^{a_1})}\right)^* \times \cdots \times
      \left(\frac{\Z}{(p_r^{a_r})}\right)^*\] (recall that
    $(\Z/n\Z)^*$ denotes the group of units of $\Z/n\Z$).
  \item Recall (\Cref{exer:groups:euler-totient-sum-formula}) that
    Euler's $\varphi$-function $\varphi(n)$ denotes the number of
    positive integers $\leq n$ that are relatively prime to $n$. Prove
    that
    \[\varphi(n) = p_1^{a_1 - 1}(p_1 - 1) \cdots p_r^{a_r - 1}(p_r -
      1).\]
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:gaussian-int-quotient-finite}
  Let $I$ be a nonzero ideal of $\Z[i]$. Prove that $\Z[i]/I$ is
  finite.
\end{exer}
\begin{exer}{}{exer:irred:gaussian-int-associates-same-norm}
  \exerused{}Let $z, w \in \Z[i]$. Show that if $z$ and $w$ are
  associates, then $N(z) = N(w)$. Show that if $w \in (z)$ and
  $N(z) = N(w)$, then $z$ and $w$ are associates.
\end{exer}
\begin{exer}{}{exer:irred:gaussian-int-irreducibles-classification}
  Prove that the irreducible elements in $\Z[i]$ are, up to
  associates: $1 + i$; the integer primes congruent to $3 \bmod 4$;
  and the elements $a \pm bi$ with $a^2 + b^2$ an integer prime
  congruent to $1 \bmod 4$.
\end{exer}
\begin{exer}{}{exer:irred:gaussian-int-euclidean-algebraic-proof}
  $\lnot$ Prove \cref{lem:irred:6.5} without any `visual' aid. (Hint:
  Let $z = a + bi$, $w = c + di$ be Gaussian integers, with
  $w \neq 0$. Then
  $z/w = \frac{ac+bd}{c^2+d^2} + \frac{bc-ad}{c^2+d^2} i$. Find
  integers $e$, $f$ such that
  $|e - \frac{ac+bd}{c^2+d^2}| \leq \frac{1}{2}$ and
  $|f - \frac{bc-ad}{c^2+d^2}| \leq \frac{1}{2}$, and set
  $q = e + if$. Prove that $|\frac{z}{w} - q| < 1$. Why does this do
  the job?)
\end{exer}
\begin{exer}{}{exer:irred:z-sqrt2-euclidean-domain}
  $\lnot$ Consider the set
  $\Z[\sqrt{2}] = \{a + b\sqrt{2} \mid a, b \in \Z\} \subseteq \C$.
  \begin{itemize}
  \item Prove that $\Z[\sqrt{2}]$ is a ring, isomorphic to
    $\Z[t]/(t^2 - 2)$.
  \item Prove that the function $N : \Z[\sqrt{2}] \to \Z$ defined by
    $N(a + b\sqrt{2}) = a^2 - 2b^2$ is multiplicative:
    $N(zw) = N(z)N(w)$. (Cf.\
    \Cref{exer:rings:z-sqrt-d-subring-of-c}.)
  \item Prove that $\Z[\sqrt{2}]$ has infinitely many units.
  \item Prove that $\Z[\sqrt{2}]$ is a Euclidean domain, by using the
    absolute value of $N$ as valuation. (Hint: Follow the same steps
    as in \Cref{exer:irred:gaussian-int-euclidean-algebraic-proof}.)
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:z-sqrt-neg2-and-imaginary-quad-euclidean}
  Working as in \Cref{exer:irred:z-sqrt2-euclidean-domain}, prove that
  $\Z[\sqrt{-2}]$ is a Euclidean domain. (Use the norm
  $N(a + b\sqrt{-2}) = a^2 + 2b^2$.)

  If you are particularly adventurous, prove that
  $\Z[(1 + \sqrt{d})/2]$ is also a Euclidean domain\footnote{You are
    probably wondering why we switched from $\Z[\sqrt{d}]$ to
    $\Z[(1 + \sqrt{d})/2]$. These rings are the `rings of integers' in
    $\Q(\sqrt{d})$; the form they take depends on the class of $d$
    modulo $4$. Their study is a cornerstone of algebraic number
    theory.} for $d = -3, -7, -11$. (You can still use the norm
  defined by $N(a + b\sqrt{d}) = a^2 - db^2$; note that this is still
  an integer on $\Z[(1 + \sqrt{d})/2]$, if $d \equiv 1 \bmod 4$.)

  The five values $d = -1, -2$, resp., $-3, -7, -11$, are the only
  ones for which $\Z[\sqrt{d}]$, resp., $\Z[(1 + \sqrt{d})/2]$, is
  Euclidean. For the values $d = -19, -43, -67, -163$, the ring
  $\Z[(1 + \sqrt{d})/2]$ is still a PID (cf.\
  \S{}\ref{subsec:irred:euclidean-domain-implies-pid} and
  \Cref{exer:irred:z-1-plus-sqrt-neg19-over-2-not-euclidean} for
  $d = -19$); the fact that there are no other negative values for
  which the ring of integers in $\Q(\sqrt{d})$ is a PID was
  conjectured by Gauss and only proven by Alan Baker and Harold Stark
  around 1966. Also, keep in mind that $\Z[\sqrt{-5}]$ is not even a
  UFD, as you have proved all by yourself in
  \Cref{exer:irred:z-sqrt-neg5-not-ufd}.
\end{exer}
\begin{exer}{}{exer:irred:n-cong-3-mod-4-not-sum-of-two-squares}
  Give an elementary proof (using modular arithmetic) of the fact that
  if an integer $n$ is congruent to $3$ modulo $4$, then it is not the
  sum of two squares.
\end{exer}
\begin{exer}{}{exer:irred:product-of-sums-of-two-squares}
  Prove that if $m$ and $n$ are two integers both of which can be
  written as sums of two squares, then $mn$ can also be written as the
  sum of two squares.
\end{exer}
\begin{exer}{}{exer:irred:sum-of-two-squares-characterization}
  Let $n$ be a positive integer.
  \begin{itemize}
  \item Prove that $n$ is a sum of two squares if and only if it is
    the norm of a Gaussian integer $a + bi$.
  \item By factoring $a^2 + b^2$ in $\Z$ and $a + bi$ in $\Z[i]$,
    prove that $n$ is a sum of two squares if and only if each integer
    prime factor $p$ of $n$ such that $p \equiv 3 \bmod 4$ appears
    with an even power in $n$.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:four-squares-lemma-np-eq-1-plus-a2-plus-b2}
  $\lnot$ One ingredient in the proof of Lagrange's theorem on four
  squares is the following result, which can be proven by completely
  elementary means.  Let $p > 0$ be an odd prime integer. Then there
  exists an integer $n$, $0 < n < p$, such that $np$ may be written as
  $1 + a^2 + b^2$ for two integers $a$, $b$. Prove this result, as
  follows:
  \begin{itemize}
  \item Prove that the numbers $a^2$, $0 \leq a \leq (p-1)/2$,
    represent $(p+1)/2$ distinct congruence classes $\bmod p$.
  \item Prove the same for numbers of the form $-1 - b^2$,
    $0 \leq b \leq (p-1)/2$.
  \item Now conclude, using the pigeon-hole principle.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:integral-quaternions-ring-basics}
  $\lnot$ Let $\field{I} \subseteq \field{H}$ be the set of
  quaternions (cf.\ \Cref{exer:rings:quaternions}) of the form
  $\frac{a}{2}(1 + i + j + k) + bi + cj + dk$ with
  $a, b, c, d \in \Z$.
  \begin{itemize}
  \item Prove that $\field{I}$ is a (noncommutative) subring of the
    ring of quaternions.
  \item Prove that the norm $N(w)$ (\Cref{exer:rings:quaternion-norm})
    of an integral quaternion $w \in \field{I}$ is an integer and
    $N(w_1 w_2) = N(w_1)N(w_2)$.
  \item Prove $\field{I}$ has exactly $24$ units in $\field{I}$:
    $\pm 1, \pm i, \pm j, \pm k$, and
    $\frac{1}{2}(\pm 1 \pm i \pm j \pm k)$.
  \item Prove that every $w \in \field{I}$ is an associate of an
    element $a + bi + cj + dk \in \field{I}$ with $a, b, c, d \in \Z$.
  \end{itemize}
  The ring $\field{I}$ is called the ring of \emph{integral
    quaternions}.
\end{exer}
\begin{exer}{}{exer:irred:integral-quaternions-euclidean-like}
  $\lnot$ Let $\field{I}$ be as in
  \Cref{exer:irred:integral-quaternions-ring-basics}. Prove that
  $\field{I}$ shares most good properties of a Euclidean domain,
  notwithstanding the fact that it is noncommutative.
  \begin{itemize}
  \item Let $z, w \in \field{I}$, with $w \neq 0$. Prove that
    $\exists q, r \in \field{I}$ such that $z = qw + r$, with
    $N(r) < N(w)$. (This is a little tricky; don't feel too bad if you
    have to cheat and look it up somewhere.)
  \item Prove that every left-ideal in $\field{I}$ is of the form
    $\field{I}w$ for some $w \in \field{I}$.
  \item Prove that every $z, w \in \field{I}$, not both zero, have a
    `greatest common right-divisor' $d$ in $\field{I}$, of the form
    $\alpha z + \beta w$ for $\alpha, \beta \in \field{I}$.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:irred:lagranges-four-squares-theorem}
  Prove Lagrange's theorem on four squares. Use notation as in
  \Cref{exer:irred:integral-quaternions-ring-basics,exer:irred:integral-quaternions-euclidean-like}.
  \begin{itemize}
  \item Let $z \in \field{I}$ and $n \in \Z$. Prove that the greatest
    common right-divisor of $z$ and $n$ in $\field{I}$ is $1$ if and
    only if $(N(z), n) = 1$ in $\Z$. (If $\alpha z + \beta n = 1$,
    then
    $N(\alpha)N(z) = N(1 - \beta n) = (1 - \beta n)(1 -
    \overline{\beta n})$, where $\overline{\beta}$ is obtained by
    changing the signs of the coefficients of $i, j, k$. Expand, and
    deduce that $(N(z), n) \mid 1$.)
  \item For an odd prime integer $p$, use
    \Cref{exer:irred:four-squares-lemma-np-eq-1-plus-a2-plus-b2} to
    obtain an integral quaternion $z = 1 + ai + bj$ such that
    $p \mid N(z)$. Prove that $z$ and $p$ have a common right-divisor
    that is not a unit and not an associate of $p$.
  \item Say that $w \in \field{I}$ is irreducible if $w = \alpha\beta$
    implies that either $\alpha$ or $\beta$ is a unit. Prove that
    integer primes are not irreducible in $\field{I}$. Deduce that
    every positive prime integer is the norm of some integral
    quaternion.
  \item Prove that every positive integer is the norm of some integral
    quaternion.
  \item Finally, use the last point of
    \Cref{exer:irred:integral-quaternions-ring-basics} to deduce that
    every positive integer may be written as the sum of four perfect
    squares.
  \end{itemize}
\end{exer}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-engine: luatex
%%% TeX-master: "../master"
%%% End:
