\chapter{Linear algebra}
\label{chap:linalg}

In several branches of science, `algebra' means `linear algebra': the
study of vector spaces and linear maps, that is, of the category of
modules over a ring $R$ in the very special case in which $R$ is a
\emph{field} (and often one restricts attention to very special
fields, such as $\R$ or $\C$).

This will be one of the main themes in this chapter. However, I will
stress that much of what can be done over a field can in fact be done
over less special rings. In fact, I will argue that working out the
general theory over \emph{integral domains} produces invaluable tools
when we are working over \emph{fields}: the paramount example being
canonical forms for matrices with entries in a field, which will be
obtained as a corollary of the classification of finitely generated
modules over a PID.

Throughout the main body of this chapter (but not necessarily in the
exercises) $R$ will denote an integral domain; most of the theory can
be extended to arbitrary commutative\footnote{The hypothesis of
  commutativity is very convenient, as it allows us to identify the
  notion of \emph{left}- and \emph{right}-modules, and the fact that
  integral domains have fields of fractions simplifies many
  arguments. The reader should keep in mind that many results in this
  chapter can be extended to more general rings.} rings without major
difficulty.

\section{Free modules revisited}
\label{sec:linalg:free-modules-revisited}


\subsection{R-Mod}
\label{subsec:linalg:r-mod}
For generalities on modules, see
\S{}\ref{sec:rings:modules-over-a-ring}. A \emph{module} over $R$ is
an abelian group $M$, endowed with an action of $R$. The action of
$r \in R$ on $m \in M$ is denoted $rm$: there is a notational bias
towards \emph{left}-modules (but the distinction between left- and
right-modules will be immaterial here as $R$ is commutative). The
defining axioms of a module tell us that for all $r_1, r_2, r \in R$
and $m, m_1, m_2 \in M$,
\begin{itemize}
\item $(r_1 + r_2)m = r_1 m + r_2 m$,
\item $1m = m$ and $(r_1 r_2)m = r_1(r_2 m)$,
\item $r(m_1 + m_2) = rm_1 + rm_2$.
\end{itemize}
Modules over a ring $R$ form the category $R$-$\categ{Mod}$, which we
encountered in \Cref{chap:rings}. This category reflects subtle and
important properties of $R$, and we are going to attempt to uncover
some of these features. As a first approximation, we look at the `full
subcategory' (cf.\ \Cref{exer:prelim:subcat}) of $R$-$\categ{Mod}$
whose objects are \emph{free} modules and in which morphisms are
ordinary morphisms in $R$-$\categ{Mod}$, that is, $R$-linear group
homomorphisms.

The goal of the first few sections of this chapter is to give a very
explicit description of this subcategory: in the case of finitely
generated free modules, this can be done by means of \emph{matrices}
with entries in $R$. Later in the chapter, we will see that matrices
may also be used to describe important classes of nonfree modules.

\subsection{Linear independence and bases}%
\label{subsec:linalg:linear-independence-and-bases}
The reader is invited to review the definition of free $R$-modules
given in \S{}\ref{subsec:rings:free-modules-and-free-algebras}:
$F^R(S)$ denotes an $R$-module containing a given set $S$ and
universal with respect to the existence of a set-map from $S$. We
proved (\Cref{prop:rings:free-module}) that the module $R^{\oplus S}$
with `one component for each element of $S$' gives an explicit
realization of $F^R(S)$.

The main point of this subsection will be that, for reasonable rings
$R$, the set $S$ can be recovered\footnote{This is not necessarily the
  case if $R$ is not commutative.} `abstractly' from the free module
$F^R(S)$. For example, it will follow easily that $R^m \cong R^n$ if
and only if $m = n$; this is the first indication that the category of
(finitely generated) free modules does indeed admit a simple
description.  Our main tool will be the famous concepts of
\emph{linearly independent subsets} and \emph{bases}. It is easy to be
imprecise in defining these notions. In order to avoid obvious traps,
I will give the definitions for \emph{indexed sets} (cf.\
\S{}\ref{subsec:prelims:examples-multisets-indexed-sets}), that is,
for functions $i : I \to M$ from a (nonempty) indexing set $I$ to a
given module $M$. The reader should think of $i$ as a selection of
elements of $M$, allowing for the possibility that the elements
$m_\alpha \in M$ corresponding to $\alpha \in I$ may not all be
distinct.

Recall that for all sets $I$ there is a canonical injection
$j : I \to F^R(I)$ and any function $i : I \to M$ determines a unique
$R$-module homomorphism $\varphi : F^R(I) \to M$ making the diagram
\[\begin{tikzcd}
    F^R(I) \arrow[r, "\varphi"] & M \\
    I \arrow[u, "j"] \arrow[ur, "i"']
  \end{tikzcd}\] commute: this is precisely the universal property
satisfied by $F^R(I)$.
\begin{defn}{}{defn:linalg:linear-independence-and-generation}
  We say that the indexed set $i : I \to M$ is \emph{linearly
    independent} if $\varphi$ is injective; $i$ is \emph{linearly
    dependent} otherwise. We say that $i$ \emph{generates} $M$ if
  $\varphi$ is surjective.
\end{defn}
Put in a slightly messier, but perhaps more common, way, an indexed
set $S = \{m_\alpha\}_{\alpha \in I}$ of elements of $M$ is linearly
independent if the only vanishing linear combination
\[\sum_{\alpha \in I} r_\alpha m_\alpha = 0\]
is obtained by choosing $r_\alpha = 0$ $\forall \alpha \in I$; $S$ is
linearly dependent otherwise. The indexed set generates $M$ if every
element of $M$ may be written as
$\sum_{\alpha \in I} r_\alpha m_\alpha$ for some choice of
$r_\alpha$. (As a notational warning/reminder, keep in mind that only
finite sums are defined in a module; therefore, in this context the
notation $\sum$ stands for a finite sum. When writing
$\sum_{\alpha \in I} m_\alpha$ in an ordinary module, one is
implicitly assuming that $m_\alpha = 0$ for all but finitely many
$\alpha \in I$.)  Using indexed sets in \cref{defn:linalg:linear-independence-and-generation} takes care
of obvious particular cases such as $m$ and $m$ being linearly
dependent since $1 \cdot m + (-1) \cdot m = 0$: if $i : I \to M$ is
not itself injective, then $\varphi$ is surely not injective (by the
commutativity of the diagram and the injectivity of $j$), so that $i$
is linearly dependent in this case. Because of this fact, the datum of
a linearly independent $i : I \to M$ amounts to a (special) choice of
\emph{distinct} elements of $M$; the temptation to identify the
elements of $I$ with the corresponding elements of $M$ is historically
irresistible and essentially harmless. Thus, it is common to speak
about linearly dependent/independent \emph{subsets} of $M$.  I will
conform to this common practice; the reader should parse the
statements carefully and correct obvious imprecisions that may arise.
A simple application of Zorn's lemma shows that every module has
maximal linearly independent subsets. In fact, it gives the following
conveniently stronger statement:

\begin{lem}{}{lem:linalg:maximal-linearly-independent-subset-exists}
  Let $M$ be an $R$-module, and let $S \subseteq M$ be a linearly
  independent subset. Then there exists a maximal linearly independent
  subset of $M$ containing $S$.
\end{lem}
\begin{proof}
  Consider the family $\mathscr{S}$ of linearly independent subsets of
  $M$ containing $S$, ordered by inclusion. Since $S$ is linearly
  independent, $\mathscr{S} \neq \emptyset$. By Zorn's lemma, it
  suffices to verify that every chain in $\mathscr{S}$ has an upper
  bound. Indeed, the union of a chain of linearly independent subsets
  containing $S$ is also linearly independent: because any relation of
  linear dependence only involves finitely many elements and these
  elements would all belong to one subset in the chain.
\end{proof}
\begin{rem}{}{rem:linalg:maximal-lin-indep-subset-equiv-to-choice}
  This statement is in fact known to be equivalent to the axiom of
  choice; therefore, the use of Zorn's lemma in one form or another
  cannot be by-passed.
\end{rem}

Note that the singleton $\{2\} \subseteq \Z$ is a `maximal linearly
independent subset' of $\Z$, but it does not generate $\Z$. In general
this is an additional requirement, leading to the definition of a
basis.

\begin{defn}{}{defn:linalg:basis}
  An indexed set $B \to M$ is a \emph{basis} if it generates $M$ and
  is linearly independent.
\end{defn}
Again, one often talks of bases as `subsets' of the module $M$; since
the images of all $b \in B$ are necessarily distinct elements, this is
rather harmless. When $B$ is finite (or at any rate countable), the
extra information carried by the indexed set can be encoded by
\emph{ordering} the elements of $B$; to emphasize this, one talks
about \emph{ordered bases}.

Bases are necessarily maximal linearly independent subsets and minimal
generating subsets; this holds over every ring. What will make modules
over a \emph{field}, i.e., vector spaces, so special is that the
converse will also hold.

In any case, only very special modules admit bases:
\begin{lem}{}{lem:linalg:free-iff-has-basis}
  An $R$-module $M$ is free if and only if it admits a basis. In fact,
  $B \subseteq M$ is a basis if and only if the natural homomorphism
  $R^{\oplus B} \to M$ is an isomorphism.
\end{lem}

\begin{proof}
  This is immediate from \cref{defn:linalg:linear-independence-and-generation}: if $B \subseteq M$ is
  linearly independent and generates $M$, then the corresponding
  homomorphism $R^{\oplus B} \to M$ is injective and
  surjective. Conversely, if $\varphi : R^{\oplus B} \to M$ is an
  isomorphism, then $B$ is identified with a subset of $M$ which
  generates it (because $\varphi$ is surjective) and is linearly
  independent (because $\varphi$ is injective).
\end{proof}
By \cref{lem:linalg:free-iff-has-basis}, the choice of a basis $B$ of a free module $M$
amounts to the choice of an isomorphism $R^{\oplus B} \cong M$; this
will be an important observation in due time (e.g., in
\S{}\ref{subsec:linalg:change-of-basis}).

Once a basis $B$ has been chosen for a free module $M$, then every
element $m \in M$ can be written uniquely as a linear combination
\[m = \sum_{b \in B} r_b b\] with $r_b \in R$. As always, remember
that all but finitely many of the coefficients $r_b$ are $0$ in any
such expression.

\subsection{Vector spaces}
\label{subsec:linalg:vector-spaces}
\cref{lem:linalg:free-iff-has-basis} is all that is needed to prove the fundamental
observation that modules over a field are necessarily free. Recall
that modules over a field $k$ are called \emph{$k$-vector spaces}
(\Cref{exam:rings:k-vect}). Elements of a vector space are called
(surprise, surprise) \emph{vectors}, while elements of the field are
called\footnote{This terminology is also often used for free modules
  over any ring.}  \emph{scalars}.

By \cref{lem:linalg:free-iff-has-basis}, proving that vector spaces are free modules
amounts to proving that they admit bases; \cref{lem:linalg:maximal-linearly-independent-subset-exists} reduces
the matter to the following:

\begin{lem}{}{lem:linalg:field-maximal-lin-indep-is-basis}
  Let $R = k$ be a field, and let $V$ be a $k$-vector space. Let $B$
  be a maximal linearly independent subset of $V$; then $B$ is a basis
  of $V$.
\end{lem}

Again, this should be contrasted with the situation over rings:
$\{2\}$ is a maximal linearly independent subset of $\Z$, but it is
not a basis.
\begin{proof}
  Let $v \in V$, $v \notin B$. Then $B \cup \{v\}$ is not linearly
  independent, by the maximality of $B$; therefore, there exist
  $c_0, \dots, c_t \in k$ and (distinct) $b_1, \dots, b_t \in B$ such
  that
  \[c_0 v + c_1 b_1 + \dots + c_t b_t = 0,\] with not all
  $c_0, \dots, c_t$ equal to $0$. Now, $c_0 \neq 0$: otherwise we
  would get a linear dependence relation among elements of $B$. Since
  $k$ is a field, $c_0$ is a unit; but then
  \[v = (-c_0^{-1} c_1) b_1 + \dots + (-c_0^{-1} c_t) b_t,\] proving
  that $v$ is in the span of $B$. It follows that $B$ generates $V$,
  as needed.
\end{proof}
Summarizing,

\begin{prop}{}{prop:linalg:lin-indep-extends-to-basis}
  Let $R = k$ be a field, and let $V$ be a $k$-vector space. Let $S$
  be a linearly independent set of vectors of $V$. Then there exists a
  basis $B$ of $V$ containing $S$.

  In particular, $V$ is free as a $k$-module.
\end{prop}

\begin{proof}
  Put \cref{lem:linalg:maximal-linearly-independent-subset-exists}, \cref{lem:linalg:free-iff-has-basis}, and \cref{lem:linalg:field-maximal-lin-indep-is-basis}
  together.
\end{proof}
We could also contemplate this situation from the `mirror' point of
view of generating sets:

\begin{lem}{}{lem:linalg:minimal-generating-set-is-basis}
  Let $R = k$ be a field, and let $V$ be a $k$-vector space. Let $B$
  be a minimal generating set for $V$; then $B$ is a basis of $V$.

  Every set generating $V$ contains a basis of $V$.
\end{lem}

\begin{proof}
  \Cref{exer:linalg:min-generating-set-is-basis}.
\end{proof}

\cref{lem:linalg:minimal-generating-set-is-basis} also fails on more general rings
(\Cref{exer:linalg:complete-to-basis-over-domain-q}). To reiterate,
over fields (but not over general rings) a subset $B$ of a vector
space is a basis $\iff$ it is a maximal linearly independent subset
$\iff$ it is a minimal generating set.

\subsection{\texorpdfstring{Recovering $B$ from $F^R(B)$}{Recovering B
    from F\^{}R(B)}}
\label{subsec:linalg:recovering-b-from-f-r-b}
We are ready for the `reconstruction' of a set $B$ (up to a
bijection!)  from the corresponding free module $F^R(B)$. This is the
result justifying the notion of \emph{dimension} of a vector space,
or, more generally, the \emph{rank} of a free module. Again, we prove
a somewhat stronger statement.

\begin{prop}{}{prop:linalg:basis-cardinality-comparison}
  Let $R$ be an integral domain, and let $M$ be a free $R$-module. Let
  $B$ be a maximal linearly independent subset of $M$, and let $S$ be
  a linearly independent subset. Then\footnote{Here, $|A|$ denotes the
    \emph{cardinality} of the set $A$, a notion with which the reader
    is hopefully familiar. The reader will not lose much by only
    considering the case in which $B$, $S$ are finite sets; but the
    fact is true for `infinite-dimensional spaces' as well, as the
    argument shows.}  $|S| \leq |B|$.

  In particular, any two maximal linearly independent subsets of a
  free module over an integral domain have the same cardinality.
\end{prop}
\begin{proof}
  By taking fields of fractions, the general case over an integral
  domain is easily reduced to the case of vector spaces over a field;
  see
  \Cref{exer:linalg:lin-indep-in-free-module-iff-in-frac-field}. We
  may then assume that $R = k$ is a field and $M = V$ is a $k$-vector
  space.

  We have to prove that there is an injective map
  $j : S \hookrightarrow B$, and this can be done by an inductive
  process, replacing elements of $B$ by elements of $S$
  `one-by-one'. For this, let $\leq$ be a well-ordering on $S$, let
  $v \in S$, and assume we have defined $j$ for all $w \in S$ with
  $w < v$. Let $B'$ be the set obtained from $B$ by replacing all
  $j(w)$ by $w$, for $w < v$, and assume (inductively) that $B'$ is
  still a maximal linearly independent subset of $V$. Then I claim
  that $j(v) \in B'$ may be defined so that
  \begin{itemize}
  \item $j(v) \neq j(w)$ for all $w < v$;
  \item the set $B''$ obtained from $B'$ by replacing $j(v)$ by $v$ is
    still a maximal linearly independent subset.
  \end{itemize}
  (Transfinite) induction (\Cref{fact:irred:induction-holds-for-wosets}) then shows
  that $j$ is defined and injective on $S$, as needed.

  To verify my claim, since $B'$ is a maximal linearly independent
  set, $B' \cup \{v\}$ is linearly dependent (as an indexed
  set\footnote{This allows for the possibility that $v \in B'$
    `already'. In this case, the reader can check that the process I
    am about to describe gives $j(v) = v$.}), so that there exists a
  linear dependence relation
  \[
    \tag{$*$} c_0v + c_1b_1 + \dots + c_tb_t = 0
  \]
  with not all $c_i$ equal to zero and the $b_i$ distinct in $B'$.
  Necessarily $c_0 \neq 0$ (because $B'$ is linearly independent);
  also, necessarily not all the $b_i$ with $c_i \neq 0$ are elements
  of $S$ (because $S$ is linearly independent). Without loss of
  generality we may then assume that $c_1 \neq 0$ and
  $b_1 \in B' \setminus S$. This guarantees that $b_1 \neq j(w)$ for
  all $w < v$; I set $j(v) = b_1$.

  All that is left now is the verification that the set $B''$ obtained
  by replacing $b_1$ by $v$ in $B'$ is a maximal linearly independent
  subset. But by using $(*)$ to write
  \[
    v = -c_0^{-1}c_1b_1 - \dots - c_0^{-1}c_tb_t,
  \]
  this is an easy consequence of the fact that $B'$ is a maximal
  linearly independent subset. Further details are left to the reader.
\end{proof}

\begin{exam}{}{exam:linalg:uncountable-subset-of-poly-ring-dependent}
  An \emph{uncountable} subset of $\C[x]$ is necessarily linearly
  dependent. Indeed, $\C[x]$ has a countable basis over $\C$: for
  example, $\{1, x, x^2, x^3, \dots\}$.
\end{exam}

\begin{coro}{}{coro:linalg:free-module-iso-iff-bijection-of-sets}
  Let $R$ be an integral domain, and let $A$, $B$ be sets. Then
  \[
    F^R(A) \cong F^R(B) \iff \text{there is a bijection } A \cong B.
  \]
\end{coro}
\begin{proof}
  \Cref{exer:linalg:deduce-free-module-iso-iff-bijection}.
\end{proof}

\begin{rem}{}{rem:linalg:completing-linearly-independent-subsets}
  We have learned in \cref{lem:linalg:maximal-linearly-independent-subset-exists} that we can `complete' every
  linearly independent subset $S$ to a maximal one. The argument used
  in the proof of \cref{prop:linalg:basis-cardinality-comparison} shows that we can in fact do this
  by borrowing elements of a given maximal linearly independent
  subset.
\end{rem}

\begin{rem}{}{rem:linalg:invariant-basis-number-property}
  As a particular case of \cref{coro:linalg:free-module-iso-iff-bijection-of-sets}, we see that if $R$ is an
  integral domain, then $R^m \cong R^n$ if and only if $m = n$. This
  says that integral domains satisfy the `IBN (Invariant Basis Number)
  property'.

  Strange as it may seem, this obvious-looking fact does \emph{not}
  hold over arbitrary rings: for example, the ring of endomorphisms of
  an infinite-dimensional vector space does not satisfy the IBN
  property.  On the other hand, integral domains are a bit of an
  overshoot: all commutative rings satisfy the IBN property
  (\Cref{exer:linalg:comm-rings-satisfy-ibn}).
\end{rem}

One way to think about this is that the category of finitely generated
free modules over (say) an integral domain is `classified' by
$\Z^{\geq 0}$: up to isomorphisms, there is exactly one finitely
generated free module for any nonnegative integer. The task of
describing this category then amounts to describing the homomorphisms
between objects corresponding to two given nonnegative integers; this
will be done in \S{}\ref{subsec:linalg:matrices}.

As a byproduct of the result of \cref{prop:linalg:basis-cardinality-comparison}, we can now give the
following important definition.

\begin{defn}{}{defn:linalg:rank-and-dimension}
  Let $R$ be an integral domain. The \emph{rank} of a free $R$-module
  $M$, denoted $\operatorname{rk}_R M$, is the cardinality of a
  maximal linearly independent subset of $M$. The rank of a vector
  space is called the \emph{dimension}, denoted $\dim_k V$.
\end{defn}

This definition will in fact be adopted for more general finitely
generated modules when the time comes, in
\S{}\ref{subsec:linalg:the-classification-theorem}.

Finite-dimensional vector spaces over a fixed field form a category.
Since vector spaces are free modules (\cref{prop:linalg:lin-indep-extends-to-basis}),
\cref{coro:linalg:free-module-iso-iff-bijection-of-sets} implies that two finite-dimensional vector spaces
are isomorphic if and only if they have the same dimension.

The subscripts $R$, $k$ are often omitted, if the context permits. But
note that, for example, viewing the complex numbers as a real vector
space, we have $\dim_{\R} \C = 2$, while $\dim_{\C} \C = 1$. So some
care is warranted.

\cref{prop:linalg:basis-cardinality-comparison} tells us that every linearly independent subset $S$
of a free $R$-module $M$ must have cardinality lower than or equal to
$\operatorname{rk}_R M$. Similarly, every generating set must have
cardinality higher than or equal to the rank. Indeed,

\begin{prop}{}{prop:linalg:generating-set-contains-max-lin-indep-subset}
  Let $R$ be an integral domain, and let $M$ be a free $R$-module;
  assume that $M$ is generated by $S$: $M = \langle S \rangle$. Then
  $S$ contains a maximal linearly independent subset of $M$.
\end{prop}
\begin{proof}
  By \Cref{exer:linalg:lin-indep-in-free-module-iff-in-frac-field} we
  may assume that $R$ is a field and $M = V$ is a vector space. Use
  Zorn's lemma to obtain a linearly independent subset $B \subseteq S$
  which is maximal among subsets of $S$. Arguing as in the proof of
  \cref{lem:linalg:field-maximal-lin-indep-is-basis} shows that $S$ is in the span of $B$, and it
  follows that $B$ generates $V$. Thus $B$ is a basis, and hence a
  maximal linearly independent subset of $V$, as needed.
\end{proof}

\begin{rem}{}{rem:linalg:fraction-field-trick-caveat}
  I have used again the trick of switching from an integral domain to
  its field of fractions. The second part of the argument would not
  work over an arbitrary integral domain, since maximal linearly
  independent subsets over an integral domain are not generating sets
  in general.

  Another standard method to reduce questions about modules over
  arbitrary commutative rings to vector spaces is to mod out by a
  maximal ideal; cf.\
  \Cref{exer:linalg:module-killed-by-maximal-ideal-is-vector-space}
  and following.
\end{rem}

\subsection*{Exercises}

\begin{exer}{$\lnot$}{exer:linalg:r-and-c-iso-as-q-vector-spaces}
  Prove that $\R$ and $\C$ are isomorphic \emph{as $\Q$-vector
    spaces}. (In particular, $(\R, +)$ and $(\C, +)$ are isomorphic as
  groups.) [\ref{exer:groups:z-q-r-not-isomorphic}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:lie-algebra-sets-are-r-vector-spaces}
  Prove that the sets listed in \Cref{exer:rings:matrix-ring} are all
  $\R$-vector spaces, and compute their
  dimensions. [\ref{exer:linalg:su2-iso-so3-as-vector-spaces}]
\end{exer}
\begin{exer}{}{exer:linalg:su2-iso-so3-as-vector-spaces}
  Prove that $\mathfrak{su}(2) \cong \mathfrak{so}_3(\R)$ as
  $\R$-vector spaces. (This is immediate, and not particularly
  interesting, from the dimension computation of
  \Cref{exer:linalg:lie-algebra-sets-are-r-vector-spaces}. However,
  these two spaces may be viewed as the tangent spaces to
  $\mathrm{SU}(2)$, resp., $\mathrm{SO}_3(\R)$, at $I$; the surjective
  homomorphism $\mathrm{SU}(2) \to \mathrm{SO}_3(\R)$ you constructed
  in \Cref{exer:groups:so3-su2-mod-plusminus-i} induces a more
  `meaningful' isomorphism $\mathfrak{su}(2) \to
  \mathfrak{so}_3(\R)$. Can you find this isomorphism?)
\end{exer}
\begin{exer}{}{exer:linalg:lie-bracket-axioms-and-category}
  Let $V$ be a vector space over a field $k$. A \emph{Lie bracket} on
  $V$ is an operation $[\cdot,\cdot] : V \times V \to V$ such that
  \begin{itemize}
  \item $(\forall u,v,w \in V), (\forall a,b \in k)$,
    \[ [au+bv,w] = a[u,w]+b[v,w], \qquad [w,au+bv] = a[w,u]+b[w,v],
    \]
  \item $(\forall v \in V)$, $[v,v] = 0$,
  \item and $(\forall u,v,w \in V)$,
    $[[u,v],w] + [[v,w],u] + [[w,u],v] = 0$.
  \end{itemize}
  (This axiom is called the \emph{Jacobi identity}.) A vector space
  endowed with a Lie bracket is called a \emph{Lie algebra}. Define a
  category of Lie algebras over a given field.

  Prove the following:
  \begin{itemize}
  \item In a Lie algebra $V$, $[u,v] = -[v,u]$ for all $u,v \in V$.
  \item If $V$ is a $k$-algebra (\Cref{defn:rings:R-algebra}), then
    $[v,w] := vw - wv$ defines a Lie bracket on $V$, so that $V$ is a
    Lie algebra in a natural way.
  \item This makes $\mathfrak{gl}_n(\R)$, $\mathfrak{gl}_n(\C)$ into
    Lie algebras. The sets listed in \Cref{exer:rings:matrix-ring} are
    all Lie algebras, with respect to a Lie bracket induced from
    $\mathfrak{gl}$.
  \item $\mathfrak{su}(2)$ and $\mathfrak{so}_3(\R)$ are isomorphic as
    Lie algebras over $\R$.
  \end{itemize}
\end{exer}
\begin{exer}{\exerused}{exer:linalg:complete-to-basis-over-domain-q}
  Let $R$ be an integral domain. Prove or disprove the following:
  \begin{itemize}
  \item Every linearly independent subset of a free $R$-module may be
    completed to a basis.
  \item Every generating subset of a free $R$-module contains a basis.
  \end{itemize} [\S{}\ref{subsec:linalg:vector-spaces}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:min-generating-set-is-basis}
  Prove \cref{lem:linalg:minimal-generating-set-is-basis}. [\S{}\ref{subsec:linalg:vector-spaces}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:lin-indep-in-free-module-iff-in-frac-field}
  Let $R$ be an integral domain, and let $M = R^{\oplus A}$ be a free
  $R$-module. Let $K$ be the field of fractions of $R$, and view $M$
  as a subset of $V = K^{\oplus A}$ in the evident way. Prove that a
  subset $S \subseteq M$ is linearly independent in $M$ (over $R$) if
  and only if it is linearly independent in $V$ (over $K$). Conclude
  that the rank of $M$ (as an $R$-module) equals the dimension of $V$
  (as a $K$-vector space). Prove that if $S$ generates $M$ over $R$,
  then it generates $V$ over $K$. Is the converse true?
  [\S{}\ref{subsec:linalg:recovering-b-from-f-r-b}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:deduce-free-module-iso-iff-bijection}
  Deduce \cref{coro:linalg:free-module-iso-iff-bijection-of-sets} from
  \cref{prop:linalg:basis-cardinality-comparison}. [\S{}\ref{subsec:linalg:recovering-b-from-f-r-b}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:module-killed-by-maximal-ideal-is-vector-space}
  Let $R$ be a commutative ring, and let $M$ be an $R$-module. Let
  $\mathfrak{m}$ be a maximal ideal in $R$, such that
  $\mathfrak{m}M = 0$ (that is, $rm = 0$ for all $r \in \mathfrak{m}$,
  $m \in M$). Define in a natural way a vector space structure over
  $R/\mathfrak{m}$ on $M$.
  [\S{}\ref{subsec:linalg:recovering-b-from-f-r-b}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:quotient-of-free-module-by-max-ideal}
  Let $R$ be a commutative ring, and let $F = R^{\oplus B}$ be a free
  module over $R$. Let $\mathfrak{m}$ be a maximal ideal of $R$, and
  let $k = R/\mathfrak{m}$ be the quotient field. Prove that
  $F/\mathfrak{m}F \cong k^{\oplus B}$ as $k$-vector
  spaces. [\ref{exer:linalg:comm-rings-satisfy-ibn}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:comm-rings-satisfy-ibn}
  Prove that commutative rings satisfy the IBN property. (Use
  \Cref{prop:irred:proper-ideal-contained-in-maximal} and
  \Cref{exer:linalg:quotient-of-free-module-by-max-ideal}.)
  [\S{}\ref{subsec:linalg:recovering-b-from-f-r-b}]
\end{exer}
\begin{exer}{}{exer:linalg:ibn-fails-for-endomorphism-ring}
  Let $V$ be a vector space over a field $k$, and let
  $R = \End_{k\text{-}\categ{Vect}}(V)$ be its ring of endomorphisms
  (cf.\ \Cref{exer:rings:endomorphism-algebra}). (Note that $R$ is
  \emph{not} commutative in general.)
  \begin{itemize}
  \item Prove that
    $\End_{k\text{-}\categ{Vect}}(V \oplus V) \cong R^4$ as an
    $R$-module.
  \item Prove that $R$ does not satisfy the IBN property if
    $V = k^{\oplus \N}$. (Note that $V \cong V \oplus V$ if
    $V = k^{\oplus \N}$.)
  \end{itemize}
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:end-a-field-char-0-implies-a-iso-q}
  Let $A$ be an abelian group such that $\End_{\categ{Ab}}(A)$ is a
  field of characteristic $0$. Prove that $A \cong \Q$.  (Hint: Prove
  that $A$ carries a $\Q$-vector space structure; what must its
  dimension be?)
  [\ref{exer:homol:fin-dim-vector-spaces-not-fully-embeddable-in-ab}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:stable-ker-image-fin-dim-endomorphism}
  Let $V$ be a finite-dimensional vector space, and let
  $\varphi : V \to V$ be a homomorphism of vector spaces. Prove that
  there is an integer $n$ such that
  $\ker\varphi^{n+1} = \ker\varphi^n$ and
  $\image\varphi^{n+1} = \image\varphi^n$.

  Show that both claims may fail if $V$ has infinite
  dimension. [\ref{exer:linalg:stable-ker-image-free-module-over-domain}]
\end{exer}
\begin{exer}{}{exer:linalg:stable-ker-image-free-module-over-domain}
  Consider the question of
  \Cref{exer:linalg:stable-ker-image-fin-dim-endomorphism} for free
  $R$-modules $F$ of finite rank, where $R$ is an integral domain that
  is not a field. Let $\varphi : F \to F$ be an $R$-module
  homomorphism.

  What property of $R$ immediately guarantees that
  $\ker\varphi^{n+1} = \ker\varphi^n$ for $n \gg 0$?

  Show that there is an $R$-module homomorphism $\varphi : F \to F$
  such that $\image\varphi^{n+1} \subsetneq \image\varphi^n$ for all
  $n \geq 0$.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:jordan-holder-for-modules}
  Let $M$ be a module over a ring $R$. A \emph{finite composition
    series} for $M$ (if it exists) is a decreasing sequence of
  submodules
  \[
    M = M_0 \supsetneq M_1 \supsetneq \dots \supsetneq M_m = \langle 0
    \rangle
  \]
  in which all quotients $M_i/M_{i+1}$ are \emph{simple} $R$-modules
  (cf.\ \Cref{exer:rings:schur}). The \emph{length} of a series is the
  number of strict inclusions. The \emph{composition factors} are the
  quotients $M_i/M_{i+1}$.

  Prove a Jordan-H\"older theorem for modules: any two finite
  composition series of a module have the same length and the same
  (multiset of) composition factors. (Adapt the proof of
  \Cref{thm:groups2:jordan-holder}.)

  We say that $M$ has \emph{length} $m$ if $M$ admits a finite
  composition series of length $m$. This notion is well-defined as a
  consequence of the result you just
  proved. [\ref{exer:linalg:length-of-vector-space-equals-dim},
  \ref{exer:linalg:submodule-of-fin-length-module-dcc},
  \ref{exer:linalg:grothendieck-grp-universal-property-fin-length},
  \ref{exer:linalg:linear-transformation-preserves-complete-flag}]
\end{exer}
\begin{exer}{}{exer:linalg:length-of-vector-space-equals-dim}
  Prove that a $k$-vector space $V$ has finite length as a module over
  $k$ (cf.\ \Cref{exer:linalg:jordan-holder-for-modules}) if and only
  if it is finite-dimensional and that in this case its length equals
  its dimension.
\end{exer}
\begin{exer}{}{exer:linalg:submodule-of-fin-length-module-dcc}
  Let $M$ be an $R$-module of finite length $m$ (cf.\
  \Cref{exer:linalg:jordan-holder-for-modules}).
  \begin{itemize}
  \item Prove that every submodule $N$ of $M$ has finite length
    $n \leq m$. (Adapt the proof of
    \Cref{prop:groups2:comp-series-normal}.)
  \item Prove that the `descending chain condition' (d.c.c.) for
    submodules holds in $M$. (Use induction on the length.)
  \item Prove that if $R$ is an integral domain that is not a field
    and $F$ is a free $R$-module, then $F$ has finite length if and
    only if it is the $0$-module.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:linalg:poly-multiple-with-prime-exponents}
  Let $k$ be a field, and let $f(x) \in k[x]$ be any polynomial. Prove
  that there exists a multiple of $f(x)$ in which all exponents of
  nonzero monomials are prime integers. (Example: for
  $f(x) = 1+x^5+x^6$,
  \[
    (1+x^5+x^6)(2x^2-x^3+x^5-x^8+x^9-x^{10}+x^{11}) =
    2x^2-x^3+x^5+2x^7+2x^{11}-x^{13}+x^{17}.
  \]
  )

  (Hint: $k[x]/(f(x))$ is a finite-dimensional $k$-vector space.)
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:free-groups-iso-iff-bijection-of-sets}
  Let $A$, $B$ be sets. Prove that the free groups $F(A)$, $F(B)$
  (\S{}\ref{sec:groups:free-groups}) are isomorphic if and only if
  there is a bijection $A \cong B$. (For the interesting direction:
  remember that $F(A) \cong F(B) \implies F^{ab}(A) \cong F^{ab}(B)$,
  by \Cref{exer:groups:free-mod-commutator-is-fab}). This extends the
  result of
  \Cref{exer:groups:free-groups-isomorphic-implies-sets-isomorphic} to
  possibly infinite sets $A$,
  $B$. [\ref{exer:groups:fab-quotient-by-2g}]
\end{exer}

\section{Homomorphisms of free modules, I}
\label{sec:linalg:homomorphisms-of-free-modules-i}


\subsection{Matrices}
\label{subsec:linalg:matrices}
As pointed out in \S{}\ref{sec:linalg:free-modules-revisited},
\Cref{coro:linalg:free-module-iso-iff-bijection-of-sets} amounts to a classification of free modules over an
integral domain: if $F$ is a free module, then there is a set $A$
(determined up to a bijection) such that $F \cong R^{\oplus A}$. The
choice of such an isomorphism is precisely the same thing as the
choice of a basis of $F$ (\Cref{lem:linalg:free-iff-has-basis}).

This is simultaneously good and bad news. The good news is that if
$F_1, F_2$ are free, then it must be possible to
`understand'\footnote{For notational simplicity I will denote
  $\Hom_{R\text{-}\categ{Mod}}(M,N)$ by $\Hom_R(M,N)$; this is very
  common, and no confusion is likely.}
\[
  \Hom_R(F_1, F_2)
\]
entirely in terms of the corresponding sets $A_1, A_2$ such that
$F_1 \cong R^{\oplus A_1}$, $F_2 \cong R^{\oplus A_2}$. That is, this
set of morphisms in $R$-$\categ{Mod}$ may be identified with
\[
  \Hom_R(R^{\oplus A_1}, R^{\oplus A_2}).
\]
The bad news is that this identification
$\Hom_R(F_1, F_2) \cong \Hom_R(R^{\oplus A_1}, R^{\oplus A_2})$ is not
`canonical', because it depends on the chosen isomorphisms
$F_1 \cong R^{\oplus A_1}$, $F_2 \cong R^{\oplus A_2}$, that is, on
the choice of bases. We will therefore have to do some work to deal
with this ambiguity (in \S{}\ref{subsec:linalg:change-of-basis}).

The point of this subsection is to deal with the good news, that is,
describe $\Hom_R(R^{\oplus A_1}, R^{\oplus A_2})$. This can be done in
a particularly convenient way when the free modules are finitely
generated, the case we are going to analyze more carefully.  Also
recall (from \S{}\ref{subsec:rings:the-category-r-mod}) that one of
the good features of the category $R$-$\categ{Mod}$ is that the set of
morphisms $\Hom_R(M,N)$ between two $R$-modules is itself an
$R$-module, in a natural way. The task is then to describe as
explicitly as possible\footnote{I am now writing $R^n$ for what was
  denoted $R^{\oplus n}$ in previous chapters.  This is a slight abuse
  of language, but it makes for easier typesetting.}
\[
  \Hom_R(R^n, R^m)
\]
as an $R$-module, for every choice of $m, n \in \Z^{\geq 0}$.

This will be done by means of \emph{matrices} with entries in $R$. I
trust that the reader is familiar with the general notion of an
$m \times n$ matrix; I have occasionally used matrices in examples
given in previous chapters, and they have showed up in several
exercises. An $m \times n$ matrix with entries in $R$ is simply a
choice of $mn$ elements of $R$. It is common to arrange these elements
as an array consisting of $m$ rows and $n$ columns:
\[
  (r_{ij})_{\substack{i=1,\dots,m \\ j=1,\dots,n}} =
  \begin{pmatrix}
    r_{11} & r_{12} & \dots & r_{1n} \\
    r_{21} & r_{22} & \dots & r_{2n} \\
    \vdots & \vdots & \ddots & \vdots \\
    r_{m1} & r_{m2} & \dots & r_{mn}
  \end{pmatrix}
\]
with $r_{ij} \in R$. For any given $m, n$ the set
$\mathcal{M}_{m,n}(R)$ of $m \times n$ matrices with entries in $R$ is
an abelian group under entrywise \emph{addition}\footnote{Note that I
  will be dropping the extra subscripts giving the range of the
  indices and will write $(r_{ij})$ rather than
  $(r_{ij})_{\substack{i=1,\dots,m \\ j=1,\dots,n}}$: these subscripts
  are an eyesore, and the context usually makes the information
  redundant.}
\[
  (a_{ij}) + (b_{ij}) := (a_{ij} + b_{ij})
\]
(cf.\ \Cref{exam:groups:gln-matrices}); this is in fact an $R$-module,
under the action
\[
  r(a_{ij}) := (r a_{ij})
\]
for $r \in R$. From this point of view, the set of $m \times n$
matrices is simply a copy of the $R$-module $R^{mn}$.  However, there
are other interesting operations on these sets. If $A = (a_{ik})$ is
an $m \times p$ matrix and $B = (b_{kj})$ is a $p \times n$ matrix,
then one may define the \emph{product} of $A$ and $B$ as
\[
  A \cdot B = (a_{ik}) \cdot (b_{kj}) := \left( \sum_{k=1}^p a_{ik}
    b_{kj} \right);
\]
this operation is (clearly) distributive with respect to addition and
compatible with the $R$-module structure. It is just a tiny bit
messier to check that it is associative, in the sense that if
$A, B, C$ are matrices of size, respectively, $m \times p$,
$p \times q$, and $q \times n$, then
\[
  (A \cdot B) \cdot C = A \cdot (B \cdot C).
\]
The reader should have no trouble reconstructing the proof of this
fact (\Cref{exer:linalg:matrix-mult-is-associative}).

In particular, we have a binary operation on the abelian group
$\mathcal{M}_n(R)$ of square $n \times n$ matrices, and this operation
is associative, distributive w.r.t.\ $+$, and admits the identity
element
\[
  \begin{pmatrix}
    1 & 0 & \dots & 0 \\
    0 & 1 & \dots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \dots & 1
  \end{pmatrix}
\]
(the \emph{identity matrix}, denoted $I_n$). That is,
$\mathcal{M}_n(R)$ is a ring
(\Cref{exam:rings:matrix-ring-noncommutative}) and in fact an
$R$-algebra (\Cref{exer:linalg:mn-r-iso-hom-as-r-algebras}).  Except
in very special cases (such as $n = 1$) this ring is not commutative.

Matrices of type $n \times 1$ are called \emph{column $n$-vectors};
matrices of type $1 \times m$ are called \emph{row $m$-vectors}, for
evident reasons. An element of a free $R$-module $R^n$ is nothing but
the choice of $n$ elements of $R$, and we can arrange these elements
into row or column vectors if we please; the standard choice is to
represent them as \emph{column} vectors:
\[
  \mathbf{v} = \begin{pmatrix} v_1 \\ v_2 \\ \vdots \\
    v_n \end{pmatrix} \in R^n.
\]
I will denote by $\mathbf{e}_i$ the elements of the `standard basis'
of $R^n$:
\[
  \mathbf{e}_1 = \begin{pmatrix} 1 \\ 0 \\ \vdots \\ 0 \end{pmatrix},
  \quad \mathbf{e}_2 = \begin{pmatrix} 0 \\ 1 \\ \vdots \\
    0 \end{pmatrix}, \quad \dots, \quad \mathbf{e}_n = \begin{pmatrix}
      0 \\ 0 \\ \vdots \\ 1 \end{pmatrix},
  \]
  so that
  \[
    \mathbf{v} = \begin{pmatrix} v_1 \\ \vdots \\ v_n \end{pmatrix} =
    \sum_{j=1}^n v_j \mathbf{e}_j.
  \]
  The elements $v_j \in R$ are the `components' of $\mathbf{v}$.

  Interpreting elements of $R^n$ as column vectors, we can act on
  $R^n$ with an $m \times n$ matrix, by left-multiplication: if
  $A = (a_{ij})$ is an $m \times n$ matrix and $\mathbf{v} \in R^n$ is
  a column vector, the product is a column vector in $R^m$:
  \[
    A \cdot \mathbf{v} =
    \begin{pmatrix}
      a_{11} & a_{12} & \dots & a_{1n} \\
      a_{21} & a_{22} & \dots & a_{2n} \\
      \vdots & \vdots & \ddots & \vdots \\
      a_{m1} & a_{m2} & \dots & a_{mn}
    \end{pmatrix}
    \cdot
    \begin{pmatrix} v_1 \\ v_2 \\ \vdots \\ v_n \end{pmatrix} =
    \begin{pmatrix}
      a_{11}v_1 + a_{12}v_2 + \dots + a_{1n}v_n \\
      a_{21}v_1 + a_{22}v_2 + \dots + a_{2n}v_n \\
      \vdots \\
      a_{m1}v_1 + a_{m2}v_2 + \dots + a_{mn}v_n
    \end{pmatrix}
    \in R^m.
  \]

\begin{lem}{}{lem:linalg:matrices-correspond-to-homomorphisms}
  For all $m \times n$ matrices $A$ with entries in $R$:
  \begin{itemize}
  \item The function $\varphi : R^n \to R^m$ defined by
    $\varphi(\mathbf{v}) = A \cdot \mathbf{v}$ is a homomorphism of
    $R$-modules.
  \item Every $R$-module homomorphism $R^n \to R^m$ is determined in
    this way by a unique $m \times n$ matrix.
  \end{itemize}
\end{lem}
\begin{proof}
  The first point follows immediately from the elementary properties
  of matrix multiplication recalled above: $\forall r, s \in R$,
  $\forall \mathbf{v}, \mathbf{w} \in R^n$
  \[
    \varphi(r\mathbf{v} + s\mathbf{w}) = A \cdot (r\mathbf{v} +
    s\mathbf{w}) = rA \cdot \mathbf{v} + sA \cdot \mathbf{w} =
    r\varphi(\mathbf{v}) + s\varphi(\mathbf{w})
  \]
  as needed.

  For the second point, let $\varphi : R^n \to R^m$ be a homomorphism
  of $R$-modules; let $a_{ij}$ be the $i$-th component of
  $\varphi(\mathbf{e}_j)$, so that
  \[
    \varphi(\mathbf{e}_j) = \begin{pmatrix} a_{1j} \\ \vdots \\ a_{mj}
    \end{pmatrix}.
  \]
  Then $A = (a_{ij})$ is an $m \times n$ matrix, and
  $\forall \mathbf{v} \in R^n$ with components $v_j$,
  \[
    A \cdot \mathbf{v} =
    \begin{pmatrix}
      a_{11}v_1 + a_{12}v_2 + \dots + a_{1n}v_n \\
      a_{21}v_1 + a_{22}v_2 + \dots + a_{2n}v_n \\
      \vdots \\
      a_{m1}v_1 + a_{m2}v_2 + \dots + a_{mn}v_n
    \end{pmatrix}
    = \sum_{j=1}^n \begin{pmatrix} a_{1j}v_j \\ a_{2j}v_j \\ \vdots \\
      a_{mj}v_j \end{pmatrix} = \sum_{j=1}^n v_j \varphi(\mathbf{e}_j)
  = \varphi\left( \sum_{j=1}^n v_j \mathbf{e}_j \right) =
  \varphi(\mathbf{v}),
\]
as needed. The homomorphism induced by a nonzero matrix is manifestly
nontrivial; this implies that the matrix $A$ associated to a
homomorphism $\varphi$ is uniquely determined by $\varphi$.
\end{proof}

The reader should attempt to remember the recipe associating a matrix
$A$ to a homomorphism $\varphi$ as in the proof of \Cref{lem:linalg:matrices-correspond-to-homomorphisms}:
the $j$-th column of $A$ is simply the column vector
$\varphi(\mathbf{e}_j)$. The collection of these vectors determines
$\varphi$---since a homomorphism is determined by its action on a set
of generators.

\Cref{lem:linalg:matrices-correspond-to-homomorphisms} yields the promised explicit description of
$\Hom_R(R^n, R^m)$:
\begin{coro}{}{coro:linalg:matrix-ring-iso-to-hom}
  The correspondence introduced in \Cref{lem:linalg:matrices-correspond-to-homomorphisms} gives an
  isomorphism of $R$-modules
  \[
    \mathcal{M}_{m,n}(R) \cong \Hom_R(R^n, R^m).
  \]
\end{coro}
\begin{proof}
  The reader will check that the correspondence is a bijective
  homomorphism of $R$-modules; this is enough, by
  \Cref{exer:rings:bijective-iso}.
\end{proof}

This is very good news, and it gets even better. Don't forget that
$R$-$\categ{Mod}$ is a category; that is, we can compose morphisms.
Therefore, there is a function\footnote{Here I am reversing the order
  of the Hom sets on the left w.r.t.\ the convention used in
  \Cref{defn:prelim:category}.}
\[
  \Hom_R(R^p, R^m) \times \Hom_R(R^n, R^p) \to \Hom_R(R^n, R^m),
\]
mapping $(\varphi, \psi)$ to $\varphi \circ \psi$; on the other hand,
the matrix product gives a function
\[
  \mathcal{M}_{m,p}(R) \times \mathcal{M}_{p,n}(R) \to
  \mathcal{M}_{m,n}(R),
\]
mapping $(A, B)$ to $A \cdot B$. That is, we have a diagram
\[
  \begin{tikzcd}
    \mathcal{M}_{m,p}(R) \times \mathcal{M}_{p,n}(R) \arrow[r]
    \arrow[d, "\sim"'] &
    \mathcal{M}_{m,n}(R) \arrow[d, "\sim"] \\
    \Hom_R(R^p, R^m) \times \Hom_R(R^n, R^p) \arrow[r] & \Hom_R(R^n,
    R^m)
  \end{tikzcd}
\]
where the vertical maps are (induced by) the isomorphisms obtained in
\Cref{coro:linalg:matrix-ring-iso-to-hom}.

\begin{lem}{}{lem:linalg:matrix-of-composition-is-product}
  This diagram commutes. That is, the matrix corresponding to a
  composition $\varphi \circ \psi$ is the product of the matrices
  corresponding to $\varphi$ and $\psi$.
\end{lem}
\begin{proof}
  This follows immediately from the associativity of matrix
  multiplication: for $\mathbf{v} \in R^n$ and
  $A \in \mathcal{M}_{m,p}(R)$, $B \in \mathcal{M}_{p,n}(R)$,
  \[
    A \cdot (B \cdot \mathbf{v}) = (A \cdot B) \cdot \mathbf{v};
  \]
  that is, the successive action of $B$ and $A$ is equivalent to the
  action of $A \cdot B$.
\end{proof}

Here is a summary of the situation. Given an integral domain $R$, we
can construct a `toy category' as follows: we let $\Z^{\geq 0}$ be the
set of objects, and we define the set of morphisms $\Hom(n,m)$ to be
the set of matrices $\mathcal{M}_{m,n}(R)$, with composition given by
the matrix product (cf.\ \Cref{exer:prelim:categ-matrix}). Then we
have verified that this toy category is `essentially the same as' the
category of finitely generated free $R$-modules and $R$-module
homomorphisms\footnote{The reader should not take this statement too
  seriously: we have identified together all free modules isomorphic
  to a given one, which is a rather drastic operation. We will take a
  more formal look at this situation in \Cref{exam:linrep:matrix-category-equivalent-to-vector-spaces}.}.

For example, by virtue of \Cref{prop:linalg:lin-indep-extends-to-basis}, if $R = k$ is a field,
then $R$-$\categ{Mod}$ = $k$-$\categ{Vect}$ consists exclusively of
free $k$-modules. The toy category I just described is a faithful
snapshot of the category of finite-dimensional $k$-vector spaces.

\subsection{Change of basis}
\label{subsec:linalg:change-of-basis}
It is time to deal with the bad news.

Free modules are only defined \emph{up to isomorphism}, as is any
structure satisfying a universal property. Writing a free module as a
direct sum $R^{\oplus A}$ amounts to making the choice of one specific
realization. As I pointed out at the end of
\S{}\ref{subsec:linalg:linear-independence-and-bases}, by
\Cref{lem:linalg:free-iff-has-basis} this choice is equivalent to the choice of a basis
of the module.

The price to pay for representing homomorphisms of free modules by
matrices is that this correspondence relies on the choice of a basis:
we say that it is \emph{not canonical}. It is essential to be able to
keep track of this choice. For finitely generated free modules, this
also boils down to the action of a matrix, as we proceed to see.

Let $F$ be a free module, and choose two bases $A$, $B$ for $F$;
assume that $F$ is finitely generated, so that $A$, $B$ are finite
sets, and further $|A| = |B|$ ($= \operatorname{rk} F$) by
\cref{prop:linalg:basis-cardinality-comparison}. The two bases correspond to two isomorphisms
\[
  R^{\oplus A} \xrightarrow{\varphi} F, \qquad R^{\oplus B}
  \xrightarrow{\psi} F.
\]
Then
\[
  R^{\oplus A} \xrightarrow{\psi^{-1} \circ \varphi} R^{\oplus B}
\]
is an isomorphism, which corresponds to a matrix as seen in
\Cref{lem:linalg:matrices-correspond-to-homomorphisms}; we may call this matrix $N_A^B$. Explicitly, the
$j$-th column of $N_A^B$ is the image of the $j$-th element of $A$,
viewed as a column vector in $R^{\oplus B}$. That is,
letting\footnote{Ordering the elements of a basis is convenient, for
  example because it allows us to talk about the `$j$-th element' of
  the basis. Hence we have the `tuple' notation.}
\[
  A = (\mathbf{a}_1, \dots, \mathbf{a}_r), \qquad B = (\mathbf{b}_1,
  \dots, \mathbf{b}_r),
\]
we have $N_A^B = (n_{ij})_{\substack{i=1,\dots,r \\ j=1,\dots,r}}$,
with
\[
  \psi^{-1} \circ \varphi(\mathbf{a}_j) = \sum_{i=1}^r n_{ij}
  \mathbf{b}_i.
\]
This equality is written \emph{in $R^{\oplus B}$}; in practice one
views the basis vectors $\mathbf{a}_j$, $\mathbf{b}_i$ as vectors
\emph{of $F$} and would therefore simply write
\[
  \mathbf{a}_j = \sum_{i=1}^r n_{ij} \mathbf{b}_i,
\]
that is, the corresponding equality in $F$.

\begin{defn}{}{defn:linalg:change-of-basis-matrix}
  The matrix $N_A^B$ is called the \emph{matrix of the change of
    basis}.
\end{defn}

Since it represents an isomorphism, the matrix of the change of basis
is necessarily an invertible matrix.

To my knowledge there is no established convention on the notation for
this matrix, and in any case I would find it futile to try to remember
any such convention. Any choice of notation will lead to a pleasant
`calculus': for example, the definition given above implies
immediately $N_B^A = (N_A^B)^{-1}$ and $N_B^C N_A^B = N_A^C$. In my
experience, any such manipulation is immediately clarified by writing
isomorphisms out explicitly, so no convention is necessary.

For example, let's work out the action of a change of basis on the
matrix representation of a homomorphism $\alpha : F \to G$ of two free
modules. The diagram taking care of the needed manipulations is
\[
  \begin{tikzcd}
    R^{\oplus A} \arrow[dr, "\varphi"] \arrow[dd, bend right=60,
    "\nu_A^B"'] &
    & R^{\oplus C} \arrow[dl, "\rho"'] \arrow[dd, bend left=60, "\mu_C^D"] \\
    & F \arrow[r, "\alpha"] & G & \\
    R^{\oplus B} \arrow[ur, "\psi"'] & & R^{\oplus D} \arrow[ul,
    "\sigma"]
  \end{tikzcd}
\]
Let $N_A^B$ be the matrix of $\nu_A^B = \psi^{-1} \circ \varphi$, as
above, and let $M_C^D$ be the matrix for
$\mu_C^D = \sigma^{-1} \circ \rho$.  Choose the basis $A$ for $F$ and
$C$ for $G$. Then the matrix representing $\alpha$ will be the matrix
$P$ corresponding to
\[
  \rho^{-1} \circ \alpha \circ \varphi : R^{\oplus A} \to R^{\oplus
    C}.
\]
If on the other hand we choose the basis $B$ for $F$ and $D$ for $G$,
then we represent $\alpha$ by the matrix $Q$ corresponding to
\[
  \sigma^{-1} \circ \alpha \circ \psi : R^{\oplus B} \to R^{\oplus D}.
\]
Now,
\[
  \sigma^{-1} \circ \alpha \circ \psi = (\mu_C^D \circ \rho^{-1})
  \circ \alpha \circ (\varphi \circ (\nu_A^B)^{-1}) = \mu_C^D \circ
  (\rho^{-1} \circ \alpha \circ \varphi) \circ (\nu_A^B)^{-1},
\]
and by \Cref{lem:linalg:matrix-of-composition-is-product} this shows
\[
  Q = M_C^D \cdot P \cdot (N_A^B)^{-1} = M_C^D \cdot P \cdot N_B^A.
\]
This is hardly surprising, of course: starting from a vector expressed
as a combination of $\mathbf{b}$'s, $N_B^A$ converts it into a
combination of $\mathbf{a}$'s; $P$ acts on it by giving its image
under $\alpha$ as a combination of $\mathbf{c}$'s; and $M_C^D$
converts it into $\mathbf{d}$'s. That accomplishes $Q$'s job, as it
should.

Summarizing this discussion,
\begin{prop}{}{prop:linalg:matrices-representing-same-hom-differ-by-units}
  Let $\alpha : F \to G$ be a homomorphism of finitely generated free
  modules, and let $P$ be a matrix representing it with respect to any
  choice of bases for $F$ and $G$. Then the matrices representing
  $\alpha$ with respect to any other choice of bases are all and only
  the matrices of the form
  \[
    M \cdot P \cdot N,
  \]
  where $M$ and $N$ are invertible matrices.
\end{prop}
\begin{defn}{}{defn:linalg:equivalent-matrices}
  Two matrices $P, Q \in \mathcal{M}_{m,n}(R)$ are \emph{equivalent}
  if they represent the same homomorphism of free modules
  $R^n \to R^m$ up to a choice of basis.
\end{defn}
This is manifestly an equivalence relation. Properly speaking, the
`abstract' homomorphism $\alpha : F \to G$ is not represented by one
matrix as much as by the whole equivalence class with respect to this
relation. \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units} gives a computational interpretation of
equivalence of matrices: $P$ and $Q$ are equivalent if and only if
there are invertible $M$ and $N$ such that $Q = MPN$.

\subsection{Elementary operations and Gaussian elimination}
\label{subsec:linalg:elementary-operations-and-gaussian-elimination}
The idea is now to capitalize on \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units}, as follows: given
a homomorphism $\alpha : F \to G$ between two free modules, find
`special' bases in $F$ and $G$ so that the matrix of $\alpha$ takes a
particularly convenient form. That is, look for a particularly
convenient matrix in each equivalence class with respect to the
relation introduced in \Cref{defn:linalg:equivalent-matrices}.

For this, we can start with random bases in $F$ and $G$, representing
$\alpha$ by a matrix $P$ and then (by \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units}) multiply $P$
on the right and left by invertible matrices, in order to bring $P$
into whatever form is best suited for our needs.

There is an even more concrete way to deal with equivalence
computationally. Consider the following three `elementary (row/column)
operations' that can be performed on a matrix $P$:
\begin{itemize}
\item switch two rows (or two columns) of $P$;
\item add to one row (resp., column) a multiple of another row (resp.,
  column);
\item multiply all entries in one row (or column) of $P$ by a unit of
  $R$.
\end{itemize}
\begin{prop}{}{prop:linalg:equivalent-via-elementary-operations}
  Two matrices $P, Q \in \mathcal{M}_{m,n}(R)$ are equivalent if $Q$
  may be obtained from $P$ by a sequence of elementary operations.
\end{prop}
\begin{proof}
  To see that elementary operations produce equivalent matrices, it
  suffices (by \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units}) to express them as multiplications
  on the left or right\footnote{Multiplying on the \emph{left} acts on
    the rows of the matrix; multiplying on the \emph{right} acts on
    its columns.} by invertible matrices. Indeed, these operations may
  be performed by suitably multiplying by the matrices obtained from
  the identity matrix by performing the same operation. For example,
  multiplying on the left by
  \[
    \begin{pmatrix}
      1 & 0 & 0 & 0 \\
      0 & 0 & 0 & 1 \\
      0 & 0 & 1 & 0 \\
      0 & 1 & 0 & 0
    \end{pmatrix}
  \]
  interchanges the second and fourth row of a $4 \times n$ matrix;
  multiplying on the right by
  \[
    \begin{pmatrix}
      1 & 0 & c \\
      0 & 1 & 0 \\
      0 & 0 & 1
    \end{pmatrix}
  \]
  adds to the third column of a $m \times 3$ matrix the $c$-multiple
  of the first column. The diligent reader will formalize this
  discussion and prove that all these matrices are indeed invertible
  (\Cref{exer:linalg:elementary-ops-give-equivalent-matrices}).
\end{proof}

The matrices corresponding to the elementary operations are called
\emph{elementary matrices}. Linear algebra over arbitrary rings would
be a great deal simpler if \Cref{prop:linalg:equivalent-via-elementary-operations} were an `if and only if'
statement. This boils down to the question of whether every invertible
matrix may be written as a product of elementary matrices, that is,
whether elementary matrices generate the `general linear group'.

\begin{defn}{}{defn:linalg:general-linear-group}
  The $n$-th \emph{general linear group} over the ring $R$, denoted
  $\mathrm{GL}_n(R)$, is the group of units in $\mathcal{M}_n(R)$,
  that is, the group of invertible $n \times n$ matrices with entries
  in $R$.
\end{defn}
Brief mentions of this notion have occurred already; cf.\
\Cref{exam:groups:gln-matrices} and several exercises in previous
chapters.

The elementary matrices are elements of $\mathrm{GL}_n(R)$; in fact,
the inverse of an elementary matrix is (of course) again an elementary
matrix. The following observation is surely known to the reader, in
one form or another; in view of the foregoing considerations, it says
that the relation introduced in \Cref{defn:linalg:equivalent-matrices} is under good
control over fields.

\begin{prop}{}{prop:linalg:gln-generated-by-elementary-matrices}
  Let $R = k$ be a field, and let $n \geq 0$ be an integer. Then
  $\mathrm{GL}_n(k)$ is generated by elementary matrices.

  Thus, two matrices are equivalent over a field if and only if they
  are linked by a sequence of elementary operations.
\end{prop}
\begin{proof}
  Let $A = (a_{ij})$ be an $n \times n$ invertible matrix. In
  particular, some entry in the first column of $A$ is nonzero; by
  performing a row switch if necessary, we may assume that $a_{11}$ is
  nonzero. Multiplying the first row by $a_{11}^{-1}$, we may assume
  that $a_{11} = 1$:
  \[
    \begin{pmatrix}
      1 & a_{12} & \dots & a_{1n} \\
      a_{21} & a_{22} & \dots & a_{1n} \\
      \vdots & \vdots & \ddots & \vdots \\
      a_{n1} & a_{n2} & \dots & a_{nn}
    \end{pmatrix}.
  \]
  Adding to the second row the $(-a_{21})$-multiple of the first row
  clears the $(2,1)$ entry. After performing the analogous operation
  on all rows, we may assume that the only nonzero entry in the first
  column is the $(1,1)$ entry:
  \[
    \begin{pmatrix}
      1 & a_{12} & \dots & a_{1n} \\
      0 & a'_{22} & \dots & a'_{1n} \\
      \vdots & \vdots & \ddots & \vdots \\
      0 & a'_{n2} & \dots & a'_{nn}
    \end{pmatrix}.
  \]
  Similarly, adding to the second column the $(-a_{12})$-multiple of
  the first column clears the $(1,2)$ entry. Performing this operation
  on all columns reduces the matrix to the form
  \[
    \begin{pmatrix}
      1 & 0 & \dots & 0 \\
      0 & a'_{22} & \dots & a'_{1n} \\
      \vdots & \vdots & \ddots & \vdots \\
      0 & a'_{n2} & \dots & a'_{nn}
    \end{pmatrix}
    = \left(
      \begin{array}{c|c}
        1 & 0 \\
        \hline
        0 & A'
      \end{array}
    \right),
  \]
  where $A'$ denotes a (clearly invertible) $(n-1) \times (n-1)$
  matrix\footnote{The vertical and horizontal lines alert the reader
    to the fact that the sectors of the matrix are themselves
    matrices; this notation is called a \emph{block matrix}.}.

  Repeating the process on $A'$ and on subsequent smaller matrices
  reduces $A$ to the identity matrix $I_n$. In other words, $I_n$ may
  be obtained from $A$ by a sequence of elementary operations:
  \[
    I_n = M \cdot A \cdot N,
  \]
  where $M$ and $N$ are products of elementary matrices. But then
  \[
    A = M^{-1} \cdot N^{-1}
  \]
  is itself a product of elementary matrices, yielding the statement.
\end{proof}

Note that, with notation as in the preceding proof,
$A^{-1} = N \cdot M$; thus, the process explained in the proof may be
used to compute the inverse of a matrix (also cf.\
\Cref{exer:linalg:row-reduction-computes-matrix-inverse}).

When applied only to the rows of a matrix, the simplification of a
matrix by means of elementary operations is called \emph{Gaussian
  elimination}. This corresponds to multiplying the given matrix on
the left by a product of elementary matrices, and it suffices in order
to reduce any square invertible matrix to the identity
(\Cref{exer:linalg:invertible-iff-row-equivalent-to-identity}).

I will sloppily call `Gaussian elimination' the more drastic process
including column operations as well as row operations; this is in line
with our focus on equivalence rather than `row-equivalence'. Applied
to any rectangular matrix, this process yields the following:
\begin{prop}{}{prop:linalg:matrix-equivalent-to-rank-normal-form}
  Over a field, every $m \times n$ matrix is equivalent to a matrix of
  the form
  \[
    \left(
      \begin{array}{c|c}
        I_r & 0 \\
        \hline
        0 & 0
      \end{array}
    \right)
  \]
  (where $r \leq \min(m,n)$ and `0' stands for null matrices of
  appropriate sizes).
\end{prop}
Different matrices of the type displayed in \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form} are
inequivalent (for example by rank considerations; cf.\
\S{}\ref{subsec:linalg:rank-and-nullity}). Thus, \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form}
describes all equivalence classes of matrices over a field and shows
that for any given $m, n$ there are in fact only \emph{finitely many}
such classes (over a field!).

\subsection{Gaussian elimination over Euclidean domains}
\label{subsec:linalg:gaussian-elimination-over-euclidean-domains}
With due care, Gaussian elimination may be performed over every
\emph{Euclidean domain}. A $2 \times 2$ example should suffice to
illustrate the general case: let
\[
  \begin{pmatrix} a & b \\ c & d \end{pmatrix} \in \mathcal{M}_2(R),
\]
for a Euclidean domain $R$, with Euclidean valuation $N$. After
switching rows and/or columns if necessary, we may assume that $N(a)$
is the minimum of the valuations of all entries in the
matrices. Division with remainder gives
\[
  b = aq + r
\]
with $r = 0$ or $N(r) < N(a)$. Adding to the second column the
$(-q)$-multiple of the first produces the matrix
\[
  \begin{pmatrix} a & r \\ c & d - qc \end{pmatrix}.
\]
If $r \neq 0$, so that $N(r) < N(a)$, we begin again and shuffle rows
and columns so that the $(1,1)$ entry has minimum valuation. This
process may be repeated, but after a finite number of steps the
$(1,2)$ entry will have to vanish: because valuations are nonnegative
integers and at each iteration the valuation of the $(1,1)$ entry
decreases.

Trivial variations of the same procedure will clear the $(2,1)$ entry
as well, producing a matrix
\[
  \begin{pmatrix} e & 0 \\ 0 & f \end{pmatrix}.
\]
Now (this is the cleverest part) I claim that we may assume that $e$
divides $f$ in $R$, with no remainder. Indeed, otherwise we can add
the second row to the first,
\[
  \begin{pmatrix} e & f \\ 0 & f \end{pmatrix},
\]
and start all over with this new matrix. Again, the effect of all the
operations will be to decrease the valuation of the $(1,1)$ entry, so
after a final number of steps we must reach the condition $e \mid f$.

The reader will have some fun describing this process in the general
case. The end result is the following useful remark:
\begin{prop}{}{prop:linalg:euclidean-domain-normal-form}
  Let $R$ be a Euclidean domain, and let $P \in \mathcal{M}_{m,n}(R)$.
  Then $P$ is equivalent to a matrix of the form\footnote{Here
    $r \leq \min(m,n)$, the bottom-right $0$ stands for a null
    $(m-r) \times (n-r)$ matrix, etc.}
  \[
    \left(
      \begin{array}{ccc|c}
        d_1 & \dots & 0 & 0 \\
        \vdots & \ddots & \vdots & \vdots \\
        0 & \dots & d_r & 0 \\
        \hline
        0 & \dots & 0 & 0
      \end{array}
    \right)
  \]
  with $d_1 \mid \dots \mid d_r$.
\end{prop}
This is called the \emph{Smith normal form} of the matrix.

\Cref{prop:linalg:euclidean-domain-normal-form} suffices to prove a weak form of one of our main
goals (a classification theorem for finitely generated modules over
PIDs) over Euclidean domains; this will hopefully be clear in a little
while, but the reader can gain the needed insight right away by
contemplating \Cref{prop:linalg:euclidean-domain-normal-form} vis-\`{a}-vis the classification of
finite abelian groups
(\Cref{exer:linalg:fin-gen-abel-grp-direct-sum-cyclic-via-matrices}).

\begin{rem}{}{rem:linalg:elementary-matrices-generate-gln-euclidean-domain}
  As a consequence of the preceding considerations and arguing as in
  the proof of \Cref{prop:linalg:gln-generated-by-elementary-matrices}, we see that $\mathrm{GL}_n(R)$ is
  generated by elementary matrices if $R$ is a Euclidean domain. The
  reader may think that some cleverness in handling Gaussian
  elimination may extend this to more general rings, but this will not
  go too far: there are examples of PIDs $R$ for which
  $\mathrm{GL}_n(R)$ is \emph{not} generated by elementary matrices.

  On the other hand, some cleverness \emph{does} manage to produce a
  Smith normal form for any matrix with entries in a PID: the presence
  of a good gcd suffices to adapt the procedure sketched above (but
  one may need more than elementary row and column operations). We
  will take a direct approach to this question in
  \S{}\ref{sec:linalg:classification-of-finitely-generated-modules-over-pids}.
\end{rem}

\subsection*{Exercises}

\begin{exer}{}{exer:linalg:unipotent-matrices-iso-to-r-plus}
  Prove that the subset of $\mathcal{M}_2(R)$ consisting of matrices
  of the form
  \[
    \begin{pmatrix} 1 & 0 \\ r & 1 \end{pmatrix}
  \]
  is a group under matrix multiplication and is isomorphic to $(R,+)$.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:matrix-mult-is-associative}
  Prove that matrix multiplication is
  associative. [\S{}\ref{subsec:linalg:matrices}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:mn-r-iso-hom-as-r-algebras}
  Prove that both $\mathcal{M}_n(R)$ and $\Hom_R(R^n, R^n)$ are
  $R$-algebras in a natural way and the bijection
  $\Hom_R(R^n, R^n) \cong \mathcal{M}_n(R)$ of \cref{coro:linalg:matrix-ring-iso-to-hom} is an
  isomorphism of $R$-algebras. (Cf.\
  \Cref{exer:rings:endomorphism-algebra}.) In particular, if the
  matrix $M$ corresponds to the homomorphism $\varphi : R^n \to R^n$,
  then $M$ is invertible in $\mathcal{M}_n(R)$ if and only if
  $\varphi$ is an isomorphism. (Note that $R$ is commutative by
  default.) [\S{}\ref{subsec:linalg:matrices},
  \S{}\ref{subsec:linalg:the-determinant},
  \S{}\ref{subsec:linalg:endomorphisms-and-similarity}]
\end{exer}
\begin{exer}{}{exer:linalg:hom-of-free-modules-iso-matrices}
  Prove \cref{coro:linalg:matrix-ring-iso-to-hom}.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:elementary-ops-give-equivalent-matrices}
  Give a formal argument proving
  \cref{prop:linalg:equivalent-via-elementary-operations}. [\S{}\ref{subsec:linalg:elementary-operations-and-gaussian-elimination}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:gaussian-elimination-to-reduced-echelon}
  A matrix with entries in a field is in \emph{row echelon form} if
  \begin{itemize}
  \item its nonzero rows are all above the zero rows and
  \item the leftmost nonzero entry of each row is $1$, and it is
    strictly to the right of the leftmost nonzero entry of the row
    above it.
  \end{itemize}
  The matrix is further in \emph{reduced} row echelon form if
  \begin{itemize}
  \item the leftmost nonzero entry of each row is the only nonzero
    entry in its column.
  \end{itemize}
  The leftmost nonzero entries in a matrix in row echelon form are
  called \emph{pivots}.

  Prove that any matrix with entries in a field can be brought into
  reduced echelon form by a sequence of elementary operations on rows.
  (This is what is more properly called Gaussian elimination.)
  [\ref{exer:linalg:pivots-determine-row-combination-coeffs},
  \ref{exer:linalg:unique-reduced-echelon-form-per-class}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:pivots-determine-row-combination-coeffs}
  Let $M$ be a matrix with entries in a field and in reduced row
  echelon form
  (\Cref{exer:linalg:gaussian-elimination-to-reduced-echelon}). Prove
  that if a row vector $\mathbf{r}$ is a linear combination
  $\sum a_i \mathbf{r}_i$ of the nonzero rows of $M$, then $a_i$
  equals the component of $\mathbf{r}$ at the position corresponding
  to the pivot on the $i$-th row of $M$. Deduce that the nonzero rows
  of $M$ are linearly
  independent. [\ref{exer:linalg:unique-reduced-echelon-form-per-class}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:row-equivalence-is-equivalence-relation}
  Two matrices $M$, $N$ are \emph{row-equivalent} if $M = PN$ for an
  invertible matrix $P$. Prove that this is indeed an equivalence
  relation, and that two matrices with entries in a field are
  row-equivalent if and only if one may be obtained from the other by
  a sequence of elementary operations on
  rows. [\ref{exer:linalg:unique-reduced-echelon-form-per-class},
  \ref{exer:linalg:bijection-row-equiv-classes-and-subspaces}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:unique-reduced-echelon-form-per-class}
  Let $k$ be a field, and consider row-equivalence
  (\Cref{exer:linalg:row-equivalence-is-equivalence-relation}) on the
  set of $m \times n$ matrices $\mathcal{M}_{m,n}(k)$. Prove that each
  equivalence class contains exactly one matrix in reduced row echelon
  form
  (\Cref{exer:linalg:gaussian-elimination-to-reduced-echelon}). (Hint:
  To prove uniqueness, argue by contradiction. Let $M$, $N$ be
  different row-equivalent reduced row echelon matrices; assume that
  they have the minimum number of columns with this property. If the
  leftmost column at which $M$ and $N$ differ is the $k$-th column,
  use the minimality to prove that $M$, $N$ may be assumed to be of
  the form
  \[
    \left(\begin{array}{c|c} I_{k-1} & * \\ \hline 0 &
        * \end{array}\right) \quad \text{or} \quad
    \left(\begin{array}{c|c} I_{k-1} & * \end{array}\right).
  \]
  Use \Cref{exer:linalg:pivots-determine-row-combination-coeffs} to
  obtain a contradiction.)

  The unique matrix in reduced row echelon form that is row-equivalent
  to a given matrix $M$ is called the \emph{reduced echelon form} of
  $M$.  [\ref{exer:linalg:dim-row-space-equals-nonzero-echelon-rows}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:row-equivalent-same-row-space}
  The \emph{row space} of a matrix $M$ is the span of its rows; the
  \emph{column space} of $M$ is the span of its column. Prove that
  row-equivalent matrices have the same row space and isomorphic
  column
  spaces. [\ref{exer:linalg:bijection-row-equiv-classes-and-subspaces},
  \S{} ef{subsec:linalg:rank-and-nullity}]
\end{exer}
\begin{exer}{}{exer:linalg:dim-row-space-equals-nonzero-echelon-rows}
  Let $k$ be a field and $M \in \mathcal{M}_{m,n}(k)$. Prove that the
  dimension of the space spanned by the rows of $M$ equals the number
  of nonzero rows in the reduced echelon form of $M$ (cf.\
  \Cref{exer:linalg:unique-reduced-echelon-form-per-class}).
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:bijection-row-equiv-classes-and-subspaces}
  Let $k$ be a field, and consider row-equivalence on
  $\mathcal{M}_{m,n}(k)$
  (\Cref{exer:linalg:row-equivalence-is-equivalence-relation}). By
  \Cref{exer:linalg:row-equivalent-same-row-space}, row-equivalent
  matrices have the same row space. Prove that, conversely, there is
  exactly one row-equivalence class in $\mathcal{M}_{m,n}(k)$ for each
  subspace of $k^n$ of dimension $\leq
  m$. [\ref{exer:linalg:grassmannian-cell-decomposition},
  \ref{exer:linalg:gr-2-4-union-of-six-schubert-cells}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:grassmannian-cell-decomposition}
  The set of subspaces of given dimension in a fixed vector space is
  called a \emph{Grassmannian}. In
  \Cref{exer:linalg:bijection-row-equiv-classes-and-subspaces} you
  have constructed a bijection between the Grassmannian of
  $r$-dimensional subspaces of $k^n$ and the set of reduced row
  echelon matrices with $n$ columns and $r$ nonzero rows.

  For $r = 1$, the Grassmannian is called the \emph{projective space}.
  For a vector space $V$, the corresponding projective space
  $\field{P}V$ is the set of `lines' (1-dimensional subspaces) in
  $V$. For $V = k^n$, $\field{P}V$ may be denoted $\P_k^{n-1}$, and
  the field $k$ may be omitted if it is clear from the context. Show
  that $\P_k^{n-1}$ may be written as a union
  $k^{n-1} \cup k^{n-2} \cup \dots \cup k^1 \cup k^0$, and describe
  each of these subsets `geometrically'.

  Thus, $\P^{n-1}$ is the union of $n$ `cells'\footnote{Here, a `cell'
    is simply a subset endowed with a natural bijection with $k^\ell$
    for some $\ell$.}, the largest one having dimension $n-1$
  (accounting for the choice of notation). Similarly, all
  Grassmannians may be written as unions of cells. These are called
  \emph{Schubert cells}.

  Prove that the Grassmannian of $(n-1)$-dimensional subspaces of
  $k^n$ admits a cell decomposition entirely analogous to that of
  $\P_k^{n-1}$. (This phenomenon will be explained in
  \Cref{exer:linrep:grassmannian-bijection-with-dual-grassmannian}.)
  [\ref{exer:fields:affine-to-projective-space-covering},
  \ref{exer:linrep:plucker-embedding-well-defined},
  \ref{exer:linrep:grassmannian-bijection-with-dual-grassmannian}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:gr-2-4-union-of-six-schubert-cells}
  Show that the Grassmannian $\operatorname{Gr}_k(2,4)$ of
  2-dimensional subspaces of $k^4$ is the union of 6 Schubert cells:
  $k^4 \cup k^3 \cup k^2 \cup k^2 \cup k^1 \cup k^0$. (Use
  \Cref{exer:linalg:bijection-row-equiv-classes-and-subspaces}; list
  all the possible reduced echelon forms.)
  [\ref{exer:linrep:grassmannian-2-4-as-projective-quadric}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:invertible-iff-row-equivalent-to-identity}
  Prove that a square matrix with entries in a field is invertible if
  and only if it is equivalent to the identity, if and only if it is
  row-equivalent to the identity, if and only if its reduced echelon
  form is the
  identity. [\S{}\ref{subsec:linalg:elementary-operations-and-gaussian-elimination},
  \ref{exer:linalg:row-reduction-computes-matrix-inverse}]
\end{exer}
\begin{exer}{}{exer:linalg:matrix-equiv-to-block-form-over-field}
  Prove \cref{prop:linalg:matrix-equivalent-to-rank-normal-form}.
\end{exer}
\begin{exer}{}{exer:linalg:matrix-equiv-to-diagonal-form-euclidean-domain}
  Prove \cref{prop:linalg:euclidean-domain-normal-form}.
\end{exer}
\begin{exer}{}{exer:linalg:diagonal-form-example-z3-to-z2}
  Suppose $\alpha : \Z^3 \to \Z^2$ is represented by the matrix
  \[
    \begin{pmatrix} -6 & 12 & 18 \\ -15 & 36 & 54 \end{pmatrix}
  \]
  with respect to the standard bases. Find bases of $\Z^3$, $\Z^2$
  with respect to which $\alpha$ is given by a matrix of the form
  obtained in \cref{prop:linalg:euclidean-domain-normal-form}.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:fin-gen-abel-grp-direct-sum-cyclic-via-matrices}
  Prove \cref{coro:groups2:finite-abelian-is-sum-of-cyclic-p-groups} again as a corollary of
  \cref{prop:linalg:euclidean-domain-normal-form}. In fact, prove the more general fact that every
  finitely generated abelian group is a direct sum of cyclic groups.
  [\S{}\ref{subsec:groups:example-subgroup-generated-by-a-subset},
  \S{}\ref{subsec:groups2:the-direct-product},
  \S{}\ref{subsec:groups2:internal-semidirect-products},
  \S{}\ref{subsec:linalg:gaussian-elimination-over-euclidean-domains},
  \S{}\ref{subsec:linalg:torsion},
  \S{}\ref{subsec:linalg:reading-a-presentation}]
\end{exer}

\section{Homomorphisms of free modules, II}
\label{sec:linalg:homomorphisms-of-free-modules-ii}


The work in \S{}\ref{sec:linalg:homomorphisms-of-free-modules-i}
accomplishes the goal of describing $\Hom_R(F,G)$, where $F$ and $G$
are free $R$-modules of finite rank; as we have seen, this can be done
most explicitly if, for example, $R$ is a field or a Euclidean
domain. We are not quite done, though: even over fields, it is
important to `understand the answer'---that is, examine the
classification of homomorphisms into finitely many classes,
highlighted at the end of
\S{}\ref{subsec:linalg:elementary-operations-and-gaussian-elimination}. Also,
the frequent appearance of invertible matrices makes it necessary to
develop criteria to tell whether a given matrix is or is not in the
general linear group. I begin by pointing out a straightforward
application of the classification of homomorphisms.

\subsection{Solving systems of linear equations}
\label{subsec:linalg:solving-systems-of-linear-equations}
A moment's thought reveals that Gaussian elimination, through the
auspices of statements such as \Cref{prop:linalg:gln-generated-by-elementary-matrices}, gives us a tool to
solve systems of linear equations over a field or a Euclidean
domain\footnote{Of course one can deal with systems over arbitrary
  integral domains $R$ by reducing to the field case, by embedding $R$
  in its field of fractions.}. This is another point which does not
seem to warrant a careful treatment or memory gymnastic: writing
things out carefully should take care of producing tools as need
be. To describe a typical situation, suppose
\[
  \begin{cases}
    a_{11}x_1 + \dots + a_{1n}x_n = b_1 \\
    \dots \\
    a_{m1}x_1 + \dots + a_{mn}x_n = b_m
  \end{cases}
\]
is a system of $m$ equations in $n$ unknowns, with $a_{ij}$, $b_i$ in
a Euclidean domain $R$. We want to find all solutions
$x_1, \dots, x_n$ in $R$.

With
\[
  A = \begin{pmatrix} a_{11} & \dots & a_{1n} \\ \vdots & \ddots &
    \vdots \\ a_{m1} & \dots & a_{mn} \end{pmatrix}, \quad \mathbf{b}
  = \begin{pmatrix} b_1 \\ \vdots \\ b_m \end{pmatrix}, \quad
  \text{and} \quad \mathbf{x} = \begin{pmatrix} x_1 \\ \vdots \\
    x_n \end{pmatrix},
\]
this amounts to `solving for $\mathbf{x}$' in the matrix equation
\[
  A \cdot \mathbf{x} = \mathbf{b}.
\]
Of course, if $m = n$ and the matrix $A$ is square and invertible,
then the solutions are obtained simply as
\[
  \mathbf{x} = A^{-1} \cdot \mathbf{b}.
\]
Here $A^{-1}$ may be obtained through Gaussian elimination (cf.\
\Cref{prop:linalg:gln-generated-by-elementary-matrices}) or through determinants
(\S{}\ref{subsec:linalg:the-determinant}).

Even without requiring anything special of the `matrix of
coefficients' $A$, row and column operations will take it to the
standard form presented in \Cref{prop:linalg:euclidean-domain-normal-form}; that is, it will yield
invertible matrices $M$, $N$ such that
\[
  M \cdot A \cdot N = \left(
    \begin{array}{ccc|c}
      d_1 & \dots & 0 & 0 \\
      \vdots & \ddots & \vdots & \vdots \\
      0 & \dots & d_r & 0
    \end{array}
  \right)
\]
with notation as in \Cref{prop:linalg:euclidean-domain-normal-form}. Gaussian elimination is a
constructive procedure: watching yourself as you switch/combine rows
or columns will produce $M$ and $N$ explicitly. Now, letting
$\mathbf{y} = (y_j)$ and $\mathbf{c} = M\mathbf{b}$, the system
\[
  (M \cdot A \cdot N) \cdot \mathbf{y} = \mathbf{c}
\]
solves itself:
\[
  \begin{cases}
    d_1 y_1 = c_1 \\
    \dots \\
    d_r y_r = c_r \\
    0 = c_{r+1} \\
    \dots
  \end{cases}
\]
has solutions if and only if $d_j \mid c_j$ for all $j$ and $c_j = 0$
for $j > r$; moreover, in this case $y_j = d_j^{-1} c_j$ for
$j = 1, \dots, r$, and $y_j$ is arbitrary for $j > r$. This yields
$\mathbf{y}$, and the reader will check that
$\mathbf{x} = N\mathbf{y}$ gives all solutions to the original system.

Such arguments can be packaged into convenient explicit procedures to
solve systems of linear equations. Again, it seems futile to list (or
try to remember) any such procedure; it seems more important to know
what is behind such techniques, so as to be able to come up with one
when needed.

In any case, the reader should be able to justify such recipes
rigorously. The most famous one is possibly Cramer's rule
(\Cref{prop:linalg:cramers-rule}) which relies on determinants and is, incidentally,
essentially useless in practice for any decent-size problem.

\subsection{The determinant}
\label{subsec:linalg:the-determinant}
Let $\alpha : F \to G$ be a homomorphism of free $R$-modules of the
same rank, and let $A$ be the matrix representing $\alpha$ with
respect to a choice of bases for $F$ and $G$. As a consequence of
\Cref{lem:linalg:matrix-of-composition-is-product} (cf.\
\Cref{exer:linalg:mn-r-iso-hom-as-r-algebras}), $\alpha$ is an
isomorphism if and only if $A$ is a unit in $\mathcal{M}_n(R)$, that
is, if and only if it is invertible as a matrix with entries in
$R$. This may be detected by computing the \emph{determinant} of $A$.

\begin{defn}{}{defn:linalg:determinant}
  Let $A = (a_{ij}) \in \mathcal{M}_n(R)$ be a square matrix. Then the
  \emph{determinant} of $A$ is the element
  \[
    \det(A) = \sum_{\sigma \in S_n} (-1)^\sigma \prod_{i=1}^n
    a_{i\sigma(i)} \in R.
  \]
\end{defn}
Here $S_n$ denotes the symmetric group on $\{1, \dots, n\}$, and I
write\footnote{Consistency with previous encounters with the symmetric
  group (e.g., \S{}\ref{sec:groups2:the-symmetric-group}) would demand
  that I write $i\sigma$; but then I would end up with things like
  `$a_{i\sigma i}$', which are very hard to parse.} $\sigma(i)$ for
the action of $\sigma \in S_n$ on $i \in \{1, \dots, n\}$;
$(-1)^\sigma$ is the \emph{sign} of a permutation
(\Cref{defn:groups2:sign}, \Cref{lem:groups2:parity-well-defined}).

The reader is surely familiar with determinants, at least over fields.
They satisfy a number of remarkable properties; here is a selection:

---The determinant of a matrix $A$ equals the determinant of its
\emph{transpose} $A^t = (a_{ij}^t)$, defined by setting
\[
  a_{ij}^t = a_{ji}
\]
for all $i$ and $j$ (that is, the rows of $A^t$ are the columns of
$A$). Indeed,
\[
  \det(A^t) = \sum_{\sigma \in S_n} (-1)^\sigma \prod_{i=1}^n
  a^t_{i\sigma(i)} = \sum_{\sigma \in S_n} (-1)^\sigma \prod_{i=1}^n
  a_{\sigma(i)i} = \sum_{\sigma \in S_n} (-1)^\sigma \prod_{i=1}^n
  a_{i\sigma^{-1}(i)}
\]
by the commutativity of the product in $R$; and $\sigma^{-1}$ ranges
over all permutations of $\{1, \dots, n\}$ as $\sigma$ does the same,
and with the same sign, so the right-most term equals $\det(A)$.

---If two rows or columns of a square matrix $A$ agree, then
$\det(A) = 0$. Indeed, it is enough to check this for matching
columns; the case for rows follows by applying the previous
observation. If columns $j$ and $j'$ of $A$ are equal, the
contribution to $\det(A)$ due to a $\sigma \in S_n$ is equal and
opposite in sign to the contribution due to the product of $\sigma$
and the transposition $(jj')$, so $\det(A) = 0$.

---Suppose $A = (a_{ij})$ and $B = (b_{ij})$ agree on all but at most
one row: $a_{ij} = b_{ij}$ if $i \neq k$, for all $j$ and some fixed
$k$. Let $c_{ij} := a_{ij} = b_{ij}$ for $i \neq k$,
$c_{kj} := a_{kj} + b_{kj}$, and let $C := (c_{ij})$. Then
\[
  \det(C) = \det(A) + \det(B).
\]
This follows immediately from \Cref{defn:linalg:determinant} and distributivity.
Applying this observation to the transpose matrices gives an analogous
statement for matrices differing at most along a column.

I will record more officially the effect of elementary operations on
determinants:
\begin{lem}{}{lem:linalg:determinant-row-column-operations}
  Let $A$ be a square matrix with entries in an integral domain $R$.
  \begin{itemize}
  \item Let $A'$ be obtained from $A$ by switching two rows or two
    columns. Then $\det(A') = -\det(A)$.
  \item Let $A'$ be obtained from $A$ by adding to a row (column) a
    multiple of another row (column). Then $\det(A') = \det(A)$.
  \item Let $A'$ be obtained from $A$ by multiplying a row (column) by
    an element\footnote{For an elementary operation, $c$ should be a
      unit; this restriction is not necessary here.} $c \in R$. Then
    $\det(A') = c\det(A)$.
  \end{itemize}
  In other words, the effect of an elementary operation on $\det(A)$
  is the same as multiplying $\det(A)$ by the determinant of the
  corresponding elementary matrix.
\end{lem}
\begin{proof}
  These are all essentially immediate from \Cref{defn:linalg:determinant}. For
  example, switching two columns amounts to correcting each $\sigma$
  in the definition by a fixed transposition, changing the sign of all
  contributions to the $\sum$ in the definition. The third point is
  immediate from distributivity. Combining the third operation and the
  two remarks preceding the statement yields the second point. Details
  are left to the reader
  (\Cref{exer:linalg:det-row-col-operations-details}).
\end{proof}

This observation simplifies the theory of determinants drastically. If
$R = k$ is a field and $P \in \mathcal{M}_n(k)$, Gaussian elimination
(\Cref{prop:linalg:matrix-equivalent-to-rank-normal-form}) shows
\[
  A = E_1 \dots E_a \cdot \left(
    \begin{array}{c|c}
      I_r & 0 \\
      \hline
      0 & 0
    \end{array}
  \right) \cdot E_1' \dots E_b',
\]
where $r \leq n$ and $E_i$, $E_j'$ are elementary matrices. Then
\Cref{lem:linalg:determinant-row-column-operations} gives that
\[
  \det(A) = \prod_i \det(E_i) \prod_j \det(E_j') \det \left(
    \begin{array}{c|c}
      I_r & 0 \\
      \hline
      0 & 0
    \end{array}
  \right).
\]
(In particular, $\det A \neq 0$ only if $r = n$.) Useful facts about
the determinant follow from this remark. For example,

\begin{prop}{}{prop:linalg:determinant-invertible-iff-unit-homomorphism}
  Let $R$ be a commutative ring.
  \begin{itemize}
  \item A square matrix $A \in \mathcal{M}_n(R)$ is invertible if and
    only if $\det(A)$ is a unit in $R$.
  \item The determinant is a homomorphism\footnote{Recall that
      $(R^*, \cdot)$ denotes the groups of units of $R$.}
    $\mathrm{GL}_n(R) \to (R^*, \cdot)$. More generally, for
    $A, B \in \mathcal{M}_n(R)$,
    \[
      \det(A \cdot B) = \det(A) \det(B).
    \]
  \end{itemize}
\end{prop}
\begin{proof}[Proof for $R = $ a field]
  If $R = k$ is a field, we can use the considerations immediately
  preceding the statement. The first point is reduced to the case of a
  block matrix
  \[
    \left(
      \begin{array}{c|c}
        I_r & 0 \\
        \hline
        0 & 0
      \end{array}
    \right),
  \]
  for which it is immediate. In fact, this shows that $\det(A) = 0$ if
  and only if the linear map $k^n \to k^n$ corresponding to $A$ is not
  an isomorphism. In particular, for all $A, B$ we have $\det(AB) = 0$
  if and only if $AB$ is not an isomorphism, if and only if $A$ or $B$
  is not an isomorphism, if and only if $\det(A) = 0$ or
  $\det(B) = 0$. So we only need to check the homomorphism property
  for invertible matrices.  These are products of elementary matrices
  `on the nose', and the homomorphism property then follows from
  \Cref{lem:linalg:determinant-row-column-operations}.
\end{proof}

Before giving the (easy) extension to the case of arbitrary
commutative rings, it is helpful to note the following explicit
formulas. A `submatrix' obtained from a given matrix $A$ by removing a
number of rows and columns is called a \emph{minor} of $A$; more
properly, this term refers to the determinants of square submatrices
obtained in these ways. If $A \in \mathcal{M}_n(R)$, the
\emph{cofactors} of $A$ are the $(n-1) \times (n-1)$ minors of $A$,
corrected by a sign. More precisely, for $A = (a_{ij})$ I will let
\[
  A^{(ij)} := (-1)^{i+j} \det
  \begin{pmatrix}
    a_{11} & \cdots & a_{1,j-1} & a_{1,j+1} & \cdots & a_{1n} \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots \\
    a_{i-1,1} & \cdots & a_{i-1,j-1} & a_{i-1,j+1} & \cdots & a_{i-1,n} \\
    a_{i+1,1} & \cdots & a_{i+1,j-1} & a_{i+1,j+1} & \cdots & a_{i+1,n} \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots \\
    a_{n1} & \cdots & a_{n,j-1} & a_{n,j+1} & \cdots & a_{nn}
  \end{pmatrix}.
\]

\begin{lem}{}{lem:linalg:cofactor-expansion}
  With notation as above,
  \begin{itemize}
  \item for all $i = 1, \dots, n$,
    $\det(A) = \sum_{j=1}^n a_{ij} A^{(ij)}$,
  \item for all $j = 1, \dots, n$,
    $\det(A) = \sum_{i=1}^n a_{ij} A^{(ij)}$.
  \end{itemize}
\end{lem}
\begin{proof}
  This is a simple (if slightly messy) induction on $n$, which I leave
  to the diligent reader.
\end{proof}

Of course \Cref{lem:linalg:cofactor-expansion} is simply stating the famous strategy
computing a determinant by expanding it with respect to your favorite
row or column. This works wonderfully for very small matrices and is
totally useless for large ones, since the number of computations
needed to apply it grows as the factorial of the size of the
matrix. From a computational point of view, it makes much better sense
to apply Gaussian elimination and use the considerations preceding
\Cref{prop:linalg:determinant-invertible-iff-unit-homomorphism}.

But \Cref{lem:linalg:cofactor-expansion} has the following important implication:
\begin{coro}{}{coro:linalg:adjugate-matrix-formula}
  Let $R$ be a commutative ring and $A \in \mathcal{M}_n(R)$. Then
  \[
    A \cdot
    \begin{pmatrix}
      A^{(11)} & \cdots & A^{(n1)} \\
      \vdots & \ddots & \vdots \\
      A^{(1n)} & \cdots & A^{(nn)}
    \end{pmatrix}
    =
    \begin{pmatrix}
      A^{(11)} & \cdots & A^{(n1)} \\
      \vdots & \ddots & \vdots \\
      A^{(1n)} & \cdots & A^{(nn)}
    \end{pmatrix}
    \cdot A = \det(A) I_n.
  \]
  Note the switch in the role of $i$ and $j$ in the matrix of
  cofactors.  This matrix is called the \emph{adjoint matrix} of $A$.
\end{coro}
\begin{proof}
  Along the diagonal of the right-hand side, this is a restatement of
  \Cref{lem:linalg:cofactor-expansion}. Off the diagonal, one is evaluating (for example)
  \[
    \sum_{j=1}^n a_{i'j} A^{(ij)}
  \]
  for $i' \neq i$. By \Cref{lem:linalg:cofactor-expansion} this is the same as the
  determinant of the matrix obtained by replacing the $i$-th row with
  the $i'$-th row; the resulting matrix has two equal rows, so its
  determinant is $0$, as needed.
\end{proof}

In particular, \Cref{coro:linalg:adjugate-matrix-formula} proves that we can invert a matrix if
we can invert its determinant:
\[
  A^{-1} = \det(A)^{-1}
  \begin{pmatrix}
    A^{(11)} & \dots & A^{(n1)} \\
    \vdots & \ddots & \vdots \\
    A^{(1n)} & \dots & A^{(nn)}
  \end{pmatrix};
\]
this holds over any commutative ring, as soon as $\det(A)$ is a unit.
In practice the computation of cofactors is `expensive', so in any
given concrete case Gaussian elimination is likely a better
alternative (at least over fields); cf.\
\Cref{exer:linalg:row-reduction-computes-matrix-inverse}.

But this formula for the inverse has good theoretical
significance. For example, we are now in a position to complete the
proof of \Cref{prop:linalg:determinant-invertible-iff-unit-homomorphism}.
\begin{proof}[Proof of \Cref{prop:linalg:determinant-invertible-iff-unit-homomorphism} for commutative rings]
  The first point follows from the second and from what we have just
  seen. Indeed, we have checked that $A \in \mathcal{M}_n(R)$ admits
  an inverse $A^{-1} \in \mathcal{M}_n(R)$ if $\det(A)$ is a unit in
  $R$; conversely, if $A$ admits an inverse
  $A^{-1} \in \mathcal{M}_n(R)$, then
  \[
    \det(A) \det(A^{-1}) = \det(AA^{-1}) = \det(I_n) = 1
  \]
  by the second statement, so that $\det(A^{-1})$ is the inverse of
  $\det(A)$ in $R$.

  Thus, we just need to verify the second point, that is, the
  homomorphism property of determinants. In order to verify this over
  every (commutative) ring, it suffices to verify the `universal'
  identity obtained by writing out the claimed equality, for matrices
  with indeterminate entries. For example, for $n = 2$ the statement
  is
  \[
    \det \begin{pmatrix} x_1 & x_2 \\ x_3 & x_4 \end{pmatrix}
    \det \begin{pmatrix} y_1 & y_2 \\ y_3 & y_4 \end{pmatrix} = \det
    \begin{pmatrix}
      x_1y_1 + x_2y_3 & x_1y_2 + x_2y_4 \\
      x_3y_1 + x_4y_3 & x_3y_2 + x_4y_4
    \end{pmatrix},
  \]
  which translates into the identity
  \[
    (x_1x_4 - x_2x_3)(y_1y_4 - y_2y_3) = (x_1y_1 + x_2y_3)(x_3y_2 +
    x_4y_4) - (x_1y_2 + x_2y_4)(x_3y_1 + x_4y_3);
  \]
  since this identity holds in $\Z[x_1, \dots, y_4]$, it must hold in
  any commutative ring, for any choice of $x_1, \dots, y_4$: indeed,
  $\Z$ is initial in $\categ{Ring}$.

  Now, we have verified that the homomorphism property holds over
  fields; in particular it holds over the field of fractions of
  $\Z[x_{11}, \dots, x_{nn}, y_{11}, \dots, y_{nn}]$. It follows that
  it does hold in $\Z[x_{11}, \dots, x_{nn}, y_{11}, \dots, y_{nn}]$,
  and we are done.
\end{proof}

The `universal identity' argument extending the result from fields to
arbitrary commutative rings is a useful device, and the reader is
invited to contemplate it carefully.

As an application of determinants (and especially cofactors) we can
now go back to the special case of a system of $n$ equations in $n$
unknowns
\[
  A \cdot \mathbf{x} = \mathbf{b},
\]
cf.\ \S{}\ref{subsec:linalg:solving-systems-of-linear-equations}, in
the case in which $\det(A)$ is a unit.

\begin{prop}{Cramer's rule}{prop:linalg:cramers-rule}
  Assume $\det(A)$ is a unit, and let $A^{(j)}$ be the matrix obtained
  by replacing the $j$-th column of $A$ by the column vector
  $\mathbf{b}$.  Then
  \[
    x_j = \det(A)^{-1} \det(A^{(j)}).
  \]
\end{prop}
\begin{proof}
  Using \Cref{lem:linalg:cofactor-expansion}, expand $\det(A^{(j)})$ with respect to the
  $j$-th column:
  \[
    \det(A^{(j)}) = \sum_{i=1}^n A^{(ij)} b_i.
  \]
  Therefore
  \[
    \begin{pmatrix} x_1 \\ \vdots \\ x_n \end{pmatrix} =
    A^{-1}\mathbf{b} = \det(A)^{-1}
    \begin{pmatrix}
      A^{(11)} & \cdots & A^{(n1)} \\
      \vdots & \ddots & \vdots \\
      A^{(1n)} & \cdots & A^{(nn)}
    \end{pmatrix}
    \cdot
    \begin{pmatrix} b_1 \\ \vdots \\ b_n \end{pmatrix} = \det(A)^{-1}
    \begin{pmatrix}
      \sum_{i=1}^n A^{(i1)} b_i \\ \vdots \\ \sum_{i=1}^n A^{(in)} b_i
    \end{pmatrix}
    =
    \begin{pmatrix}
      \det(A)^{-1} \det(A^{(1)}) \\ \vdots \\ \det(A)^{-1} \det(A^{(n)})
    \end{pmatrix},
  \]
  which gives the statement.
\end{proof}

\subsection{Rank and nullity}
\label{subsec:linalg:rank-and-nullity}
According to \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form}, each equivalence class of matrices
over a field has a representative of the type
\[
  \left(
    \begin{array}{c|c}
      I_r & 0 \\
      \hline
      0 & 0
    \end{array}
  \right).
\]
The reason why two different $m \times n$ matrices of this type are
surely inequivalent is that if $\alpha : V \to W$ (with $V$ and $W$
free modules, vector spaces in this case) is represented by a matrix
of this form, then $r$ is the dimension of the image of $\alpha$.
Therefore, matrices with different $r$ cannot represent the same
$\alpha$.

This integer $r$ is called the \emph{rank} of the matrix and deserves
some attention. I will discuss it for matrices with entries in a
field, leaving generalizations to more general rings to the reader
(for now at least).

The \emph{column (row) space} of a matrix $P$ over a field $k$ is the
span of the columns (rows) of $P$. The \emph{column (row) rank} of $P$
is the dimension of the column (row) space of $P$.

\begin{prop}{}{prop:linalg:row-rank-equals-column-rank}
  The row rank of a matrix over a field $k$ equals its column rank.
\end{prop}
\begin{proof}
  Equivalent matrices have the same ranks. Indeed, let
  $P \in \mathcal{M}_{m,n}(k)$; the row space of $P$ consists of all
  row vectors
  \[
    \begin{pmatrix} a_1 & \dots & a_m \end{pmatrix} =
    \begin{pmatrix} v_1 & \dots & v_m \end{pmatrix} \cdot P
  \]
  obtained as each $v_i$ ranges in $k$. If $Q = MPN$, with $M$ and $N$
  invertible, let $\begin{pmatrix} w_1 & \dots & w_m \end{pmatrix} =
  \begin{pmatrix} v_1 & \dots & v_m \end{pmatrix} \cdot M^{-1}$; then
  \[
    \begin{pmatrix} w_1 & \dots & w_m \end{pmatrix} \cdot Q =
    \begin{pmatrix} v_1 & \dots & v_m \end{pmatrix} M^{-1}(MPN) =
    \begin{pmatrix} a_1 & \dots & a_m \end{pmatrix} \cdot N.
  \]
  This shows that multiplication on the right by $N$ maps the row
  space of $P$ (isomorphically, since $N$ is invertible) to the row
  space of $Q$; thus the two spaces have the same dimension, as
  claimed. Minimal variations on the same argument show that the
  column ranks of $P$ and $Q$ agree. (Cf.\
  \Cref{exer:linalg:row-equivalent-same-row-space}.)

  Applying this observation and \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form} reduces the
  question to matrices
  \[
    \left(
      \begin{array}{c|c}
        I_r & 0 \\
        \hline
        0 & 0
      \end{array}
    \right),
  \]
  for which row rank $= r =$ column rank, proving the statement.
\end{proof}

The result upgrades easily to matrices over an arbitrary integral
domain $R$ (applying the usual trick of embedding $R$ in its field of
fractions).

In view of \Cref{prop:linalg:row-rank-equals-column-rank}, we can simply talk about the rank of a
matrix:
\begin{defn}{}{defn:linalg:rank-of-a-matrix}
  Let $M \in \mathcal{M}_{m,n}(k)$ be a matrix over a field $k$. The
  \emph{rank} of $M$ is the dimension of its column (or, equivalently,
  row) space.
\end{defn}
One way to rephrase \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form} is that matrices over a field
are classified up to equivalence by their rank.

The foregoing considerations translate nicely in more abstract terms
for a linear map $\alpha : V \to W$ between finite-dimensional vector
spaces over a field $k$. Using the convenient language introduced in
\S{}\ref{subsec:rings:complexes-and-exact-sequences}, note that each
$\alpha$ determines an exact sequence of vector spaces
\[
  0 \to \ker\alpha \to V \to \image\alpha \to 0.
\]
\begin{defn}{}{defn:linalg:rank-and-nullity-of-transformation}
  The \emph{rank} of $\alpha$, denoted $\operatorname{rk}\alpha$, is
  the dimension of $\image\alpha$. The \emph{nullity} of $\alpha$ is
  $\dim(\ker\alpha)$.
\end{defn}
\begin{prop}{}{prop:linalg:rank-nullity-theorem}
  Let $\alpha : V \to W$ be a linear map of finite-dimensional vector
  spaces. Then
  \[
    (\text{rank of } \alpha) + (\text{nullity of } \alpha) = \dim V.
  \]
\end{prop}
\begin{proof}
  Let $n = \dim V$ and $m = \dim W$. By \Cref{prop:linalg:matrix-equivalent-to-rank-normal-form} we can
  represent $\alpha$ by an $m \times n$ matrix of the form
  \[
    \left(
      \begin{array}{c|c}
        I_r & 0 \\
        \hline
        0 & 0
      \end{array}
    \right).
  \]
  From this representation it is immediate that
  $\operatorname{rk}\alpha = r$ and the nullity of $\alpha$ is
  $n - r$, with the stated consequence.
\end{proof}

Summarizing, $\operatorname{rk}\alpha$ equals the (column) rank of any
matrix $P$ representing $\alpha$; similarly, the nullity of $\alpha$
equals `$\dim V$ minus the (row) rank' of $P$. \Cref{prop:linalg:rank-nullity-theorem} is
the abstract version of the equality of row rank and column rank.

\subsection{Euler characteristic and the Grothendieck group}
\label{subsec:linalg:euler-characteristic-and-the-grothendieck-group}
Against my best efforts, I cannot resist extending these simple
observations to more general complexes. \Cref{prop:linalg:rank-nullity-theorem} may be
reformulated as follows:
\begin{prop}{}{prop:linalg:dimension-additive-in-exact-sequence}
  Let
  \[
    0 \to U \to V \to W \to 0
  \]
  be a short exact sequence of finite-dimensional vector spaces. Then
  \[
    \dim(V) = \dim(U) + \dim(W).
  \]
\end{prop}
Equivalently, this amounts to the relation
$\dim(V/U) = \dim(V) - \dim(U)$.

Consider then a complex of finite-dimensional vector spaces and linear
maps:
\[
  V_\bullet : \quad 0 \to V_N \xrightarrow{\alpha_N} V_{N-1}
  \xrightarrow{\alpha_{N-1}} \dots \xrightarrow{\alpha_2} V_1
  \xrightarrow{\alpha_1} V_0 \to 0
\]
(cf.\ \S{}\ref{subsec:rings:complexes-and-exact-sequences}). Thus,
$\alpha_{i-1} \circ \alpha_i = 0$ for all $i$.  This condition is
equivalent to the requirement that
$\image(\alpha_{i+1}) \subseteq \ker(\alpha_i)$; recall that the
\emph{homology} of this complex is defined as the collection of spaces
\[
  H_i(V_\bullet) = \frac{\ker(\alpha_i)}{\image(\alpha_{i+1})}.
\]
The complex is \emph{exact} if $\image(\alpha_{i+1}) = \ker(\alpha_i)$
for all $i$, that is, if $H_i(V_\bullet) = 0$ for all $i$.

\begin{defn}{}{defn:linalg:euler-characteristic}
  The \emph{Euler characteristic} of $V_\bullet$ is the integer
  \[
    \chi(V_\bullet) := \sum_i (-1)^i \dim(V_i).
  \]
\end{defn}
The original motivation for the introduction of this number is
topological: with suitable positions, this Euler characteristic equals
the Euler characteristic obtained by triangulating a manifold and then
computing the number of vertices of the triangulation, minus the
number of edges, plus the number of faces, etc.

The following simple result is then a straightforward (and very
useful) generalization of \Cref{prop:linalg:dimension-additive-in-exact-sequence}:
\begin{prop}{}{prop:linalg:euler-characteristic-via-homology}
  With notation as above,
  \[
    \chi(V_\bullet) = \sum_{i=0}^N (-1)^i \dim(H_i(V_\bullet)).
  \]
  In particular, if $V_\bullet$ is exact, then $\chi(V_\bullet) = 0$.
\end{prop}
\begin{proof}
  There is nothing to show for $N = 0$, and the result follows
  directly from \Cref{prop:linalg:dimension-additive-in-exact-sequence} if $N = 1$
  (\Cref{exer:linalg:euler-char-case-n-eq-1}). Arguing by induction,
  given a complex
  \[
    V_\bullet : \quad 0 \to V_N \xrightarrow{\alpha_N} V_{N-1}
    \xrightarrow{\alpha_{N-1}} \dots \xrightarrow{\alpha_2} V_1
    \xrightarrow{\alpha_1} V_0 \to 0,
  \]
  we may assume that the result is known for `shorter' complexes.
  Consider then the truncation
  \[
    V_\bullet' : \quad 0 \to V_{N-1} \xrightarrow{\alpha_{N-1}} \dots
    \xrightarrow{\alpha_2} V_1 \xrightarrow{\alpha_1} V_0 \to 0.
  \]
  Then
  \[
    \chi(V_\bullet) = \chi(V_\bullet') + (-1)^N \dim(V_N),
  \]
  and
  \[
    H_i(V_\bullet) = H_i(V_\bullet') \quad \text{for } 0 \leq i \leq
    N-2,
  \]
  while
  \[
    H_{N-1}(V_\bullet') = \ker(\alpha_{N-1}), \quad H_{N-1}(V_\bullet)
    = \frac{\ker(\alpha_{N-1})}{\image(\alpha_N)}, \quad
    H_N(V_\bullet) = \ker(\alpha_N).
  \]
  By \Cref{prop:linalg:dimension-additive-in-exact-sequence} (cf.\ \Cref{prop:linalg:rank-nullity-theorem}),
  \[
    \dim(V_N) = \dim(\image(\alpha_N)) + \dim(\ker(\alpha_N))
  \]
  and
  \[
    \dim(H_{N-1}(V_\bullet)) = \dim(\ker(\alpha_{N-1})) -
    \dim(\image(\alpha_N));
  \]
  therefore
  \[
    \dim(H_{N-1}(V_\bullet')) - \dim(V_N) = \dim(H_{N-1}(V_\bullet)) -
    \dim(H_N(V_\bullet)).
  \]
  Putting all of this together with the induction hypothesis,
  \[
    \chi(V_\bullet') = \sum_{i=0}^{N-1} (-1)^i \dim(H_i(V_\bullet'))
  \]
  gives
  \[
    \chi(V_\bullet) = \chi(V_\bullet') + (-1)^N \dim(V_N)
  \]
  \[
    = \sum_{i=0}^{N-1} (-1)^i \dim(H_i(V_\bullet')) + (-1)^N \dim(V_N)
  \]
  \[
    = \sum_{i=0}^{N-2} (-1)^i \dim(H_i(V_\bullet')) + (-1)^{N-1}
    (\dim(H_{N-1}(V_\bullet')) - \dim(V_N))
  \]
  \[
    = \sum_{i=0}^{N-2} (-1)^i \dim(H_i(V_\bullet)) + (-1)^{N-1}
    (\dim(H_{N-1}(V_\bullet)) - \dim(H_N(V_\bullet)))
  \]
  \[
    = \sum_{i=0}^N (-1)^i \dim(H_i(V_\bullet))
  \]
  as needed.
\end{proof}

In terms of the topological motivation recalled above,
\Cref{prop:linalg:euler-characteristic-via-homology} tells us that the Euler characteristic of a
manifold may be computed as the alternating sum of the ranks of its
homology, that is, of its \emph{Betti numbers}.

Having come this far, I cannot refrain from mentioning the next,
equally simple-minded, generalization. The reader has surely noticed
that the only tool used in the proof of \Cref{prop:linalg:euler-characteristic-via-homology} was the
`additivity' property of dimension, established in \Cref{prop:linalg:dimension-additive-in-exact-sequence}:
if
\[
  0 \to U \to V \to W \to 0
\]
is exact, then
\[
  \dim(V) = \dim(U) + \dim(W).
\]
\Cref{prop:linalg:euler-characteristic-via-homology} is a formal consequence of this one property of
$\dim$.

With this in mind, we can reinterpret what we have just done in the
following curious way. Consider the category
$k$-$\categ{Vect}$$^f$ of finite-dimensional
$k$-vector spaces. Each object $V$ of
$k$-$\categ{Vect}$$^f$ determines an isomorphism class $[V]$. Let
$F(k\text{-}\categ{Vect}^f)$ be the free abelian group on the set of
these isomorphism classes; further, let $E$ be the subgroup generated
by the elements
\[ [V] - [U] - [W]
\]
for all short exact sequences
\[
  0 \to U \to V \to W \to 0
\]
in $k$-$\categ{Vect}$$^f$. The quotient group
\[
  K(k\text{-}\categ{Vect}^f) := \frac{F(k\text{-}\categ{Vect}^f)}{E}
\]
is called the \emph{Grothendieck group} of the category
$k-\categ{Vect}^f$. The element determined by
$V$ in the Grothendieck group is still denoted $[V]$.

More generally, a Grothendieck group may be defined for any category
admitting a notion of exact sequence.

Every complex $V_\bullet$ determines an element in
$K(k\text{-}\categ{Vect}^f)$, namely
\[
  \chi_K(V_\bullet) \df \sum_i (-1)^i [V_i] \in
  K(k\text{-}\categ{Vect}^f).
\]

\begin{prop}{}{prop:linalg:euler-characteristic-properties}
  With notation as above, we have the following:
  \begin{itemize}
  \item $\chi_K$ `is an Euler characteristic', in the sense that it
    satisfies the formula given in \Cref{prop:linalg:euler-characteristic-via-homology}:
    \[
      \chi_K(V_\bullet) = \sum_i (-1)^i [H_i(V_\bullet)].
    \]
  \item $\chi_K$ is a `universal Euler characteristic', in the
    following sense. Let $G$ be an abelian group, and let $\delta$ be
    a function associating an element of $G$ to each
    finite-dimensional vector space, such that
    $\delta(V) = \delta(V')$ if $V \cong V'$ and
    $\delta(V/U) = \delta(V) - \delta(U)$. For $V_\bullet$ a complex,
    define
    \[
      \chi_G(V_\bullet) = \sum_i (-1)^i \delta(V_i).
    \]
    Then $\delta$ induces a (unique) group homomorphism
    \[
      K(k\text{-}\categ{Vect}^f) \to G
    \]
    mapping $\chi_K(V_\bullet)$ to $\chi_G(V_\bullet)$.
  \item In particular, $\delta = \dim$ induces a group homomorphism
    \[
      K(k- \categ{Vect}^f) \to \Z
    \]
    such that $\chi_K(V_\bullet) \mapsto \chi(V_\bullet)$.
  \item This is in fact an isomorphism.
  \end{itemize}
\end{prop}
The best way to convince the reader that this impressive claim is
completely trivial is to leave its proof to the reader
(\Cref{exer:linalg:euler-char-properties}). The first point is proved
by adapting the proof of \Cref{prop:linalg:euler-characteristic-via-homology}; the second point is a
harmless mixture of universal properties; the third point follows from
the second; and the last point follows from the fact that
$\dim(k) = 1$ and the IBN property.

The last point is, in fact, rather anticlimactic: if the impressively
abstract Grothendieck group turns out to just be a copy of the
integers, why bother defining it? The answer is, of course, that this
only stresses how special the category
$k$-$\categ{Vect}$$^f$ is. The definition of the Grothendieck group
can be given in any context in which complexes and a notion of
exactness are available (for example, in the category of finitely
generated\footnote{Note that some finiteness condition is likely to be
  necessary. We cannot define a `Grothendieck group of
  $k$-$\categ{Vect}$' because the isomorphism classes of objects in
  $k$-$\categ{Vect}$ do not form a set: there is one for each cardinal
  number, and cardinal numbers do not form a set.}  modules over any
ring). The formal arguments proving \Cref{prop:linalg:euler-characteristic-properties} will go through
in any such context and provide us with a useful notion of `universal
Euler characteristic'.

We will come back to complexes and homology in \Cref{chap:homol}.

\subsection*{Exercises}

\begin{exer}{}{exer:linalg:gaussian-elimination-integer-solutions-example}
  Use Gaussian elimination to find all integer solutions of the system
  of equations
  \[
    \begin{cases}
      7x - 36y + 12z = 1, \\
      -8x + 42y - 14z = 2.
    \end{cases}
  \]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:det-row-col-operations-details}
  Provide details for the proof of
  \Cref{lem:linalg:determinant-row-column-operations}. [\S{}\ref{subsec:linalg:the-determinant}]
\end{exer}
\begin{exer}{}{exer:linalg:redo-sln-normal-in-gln-with-det}
  Redo \Cref{exer:groups:sln-normal-in-gln}.
\end{exer}
\begin{exer}{}{exer:linalg:universal-identities-formalization}
  Formalize the discussion of `universal identities': by what cocktail
  of universal properties is it true that if an identity holds in
  $\Z[x_1, \dots, x_r]$, then it holds over every commutative ring
  $R$, for every choice of $x_i \in R$? (Is the commutativity of $R$
  necessary?)
\end{exer}
\begin{exer}{\exerused}{exer:linalg:row-reduction-computes-matrix-inverse}
  Let $A$ be an $n \times n$ square invertible matrix with entries in
  a field, and consider the $n \times (2n)$ matrix $B = (A | I_n)$
  obtained by placing the identity matrix to the side of $A$. Perform
  elementary row operations on $B$ so as to reduce $A$ to $I_n$ (cf.\
  \Cref{exer:linalg:invertible-iff-row-equivalent-to-identity}). Prove
  that this transforms $B$ into $(I_n | A^{-1})$.

  (This is a much more efficient way to compute the inverse of a
  matrix than by using determinants as in
  \S{}\ref{subsec:linalg:the-determinant}.)
  [\S{}\ref{subsec:linalg:elementary-operations-and-gaussian-elimination},
  \S{}\ref{subsec:linalg:the-determinant}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:det-a-annihilates-module-determinant-trick}
  Let $R$ be a commutative ring and
  $M = \langle m_1, \dots, m_r \rangle$ a finitely generated
  $R$-module. Let $A \in \mathcal{M}_r(R)$ be a matrix such that
  \[
    A \cdot \begin{pmatrix} m_1 \\ \vdots \\ m_r \end{pmatrix} =
    \begin{pmatrix} 0 \\ \vdots \\ 0 \end{pmatrix}.
  \]
  Prove that $\det(A) m = 0$ for all $m \in M$. (Hint: Multiply by the
  adjoint.) [\ref{exer:linalg:jm-eq-m-implies-1-plus-b-kills-m}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:jm-eq-m-implies-1-plus-b-kills-m}
  Let $R$ be a commutative ring, $M$ a finitely generated $R$-module,
  and let $J$ be an ideal of $R$. Assume $JM = M$. Prove that there
  exists an element $b \in J$ such that $(1+b)M = 0$. (Let
  $m_1, \dots, m_r$ be generators for $M$. Find an $r \times r$ matrix
  $B$ with entries in $J$ such that
  \[
    \begin{pmatrix} m_1 \\ \vdots \\ m_r \end{pmatrix} = B \cdot
    \begin{pmatrix} m_1 \\ \vdots \\ m_r \end{pmatrix}.
  \]
  Then use
  \Cref{exer:linalg:det-a-annihilates-module-determinant-trick}.)
  [\ref{exer:linalg:nakayamas-lemma},
  \ref{exer:linrep:intersection-of-ideal-powers-krull}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:nakayamas-lemma}
  Let $R$ be a commutative ring, $M$ be a finitely generated
  $R$-module, and let $J$ be an ideal of $R$ contained in the Jacobson
  radical of $R$
  (\Cref{exer:irred:jacobson-radical-characterization}). Prove that
  $M = 0 \iff JM = M$. (Use
  \Cref{exer:linalg:jm-eq-m-implies-1-plus-b-kills-m}.  This is
  \emph{Nakayama's lemma}, a result with important applications in
  commutative algebra and algebraic geometry. A particular case was
  given as \Cref{exer:rings:nakayama-nilpotent}.)
  [\ref{exer:rings:nakayama-nilpotent},
  \ref{exer:linalg:local-ring-m-eq-mm-plus-n-implies-m-eq-n},
  \ref{exer:linalg:direct-summand-of-free-over-local-is-free}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:local-ring-m-eq-mm-plus-n-implies-m-eq-n}
  Let $R$ be a commutative local ring, that is, a ring with a single
  maximal ideal $\mathfrak{m}$, and let $M$, $N$ be finitely generated
  $R$-modules. Prove that if $M = \mathfrak{m}M + N$, then $M = N$.
  (Apply Nakayama's lemma, that is,
  \Cref{exer:linalg:nakayamas-lemma}, to $M/N$. Note that the Jacobson
  radical of $R$ is $\mathfrak{m}$.)
  [\ref{exer:linalg:nakayama-gives-generators-from-quotient-basis}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:nakayama-gives-generators-from-quotient-basis}
  Let $R$ be a commutative local ring, and let $M$ be a finitely
  generated $R$-module. Note that $M/\mathfrak{m}M$ is a
  finite-dimensional vector space over the field $R/\mathfrak{m}$; let
  $m_1, \dots, m_r \in M$ be elements whose cosets mod $\mathfrak{m}M$
  form a basis of $M/\mathfrak{m}M$. Prove that $m_1, \dots, m_r$
  generate $M$.

  (Show that $\langle m_1, \dots, m_r \rangle + \mathfrak{m}M = M$;
  then apply Nakayama's lemma in the form of
  \Cref{exer:linalg:local-ring-m-eq-mm-plus-n-implies-m-eq-n}.)
  [\ref{exer:linalg:direct-summand-of-free-over-local-is-free},
  \ref{exer:linrep:fin-gen-flat-over-noeth-local-is-free}]
\end{exer}
\begin{exer}{}{exer:linalg:gaussian-elim-finds-row-col-space-bases}
  Explain how to use Gaussian elimination to find bases for the row
  space and the column space of a matrix over a field.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:more-columns-than-rows-linearly-dependent}
  Let $R$ be an integral domain, and let $M \in \mathcal{M}_{m,n}(R)$,
  with $m < n$. Prove that the columns of $M$ are linearly dependent
  over $R$. [\ref{exer:linalg:rank-bounded-by-num-generators}]
\end{exer}
\begin{exer}{}{exer:linalg:rank-leq-r-iff-pq-factorization}
  Let $k$ be a field. Prove that a matrix $M \in \mathcal{M}_{m,n}(k)$
  has rank $\leq r$ if and only if there exist matrices
  $P \in \mathcal{M}_{m,r}(k)$, $Q \in \mathcal{M}_{r,n}(k)$ such that
  $M = PQ$. (Thus the rank of $M$ is the smallest such integer.)
\end{exer}
\begin{exer}{}{exer:linalg:generalize-exact-seq-dim-formula-to-domains}
  Generalize \Cref{prop:linalg:dimension-additive-in-exact-sequence} to the case of finitely generated free
  modules over any integral domain. (Embed the integral domain in its
  field of fractions.)
\end{exer}
\begin{exer}{\exerused}{exer:linalg:euler-char-case-n-eq-1}
  Prove \Cref{prop:linalg:euler-characteristic-via-homology} for the case $N =
  1$. [\S{}\ref{subsec:linalg:euler-characteristic-and-the-grothendieck-group}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:euler-char-properties}
  Prove
  \cref{prop:linalg:euler-characteristic-properties}. [\S{}\ref{subsec:linalg:euler-characteristic-and-the-grothendieck-group}]
\end{exer}
\begin{exer}{}{exer:linalg:grothendieck-grp-countable-dim-trivial}
  Extend the definition of Grothendieck group of vector spaces given
  in
  \S{}\ref{subsec:linalg:euler-characteristic-and-the-grothendieck-group}
  to the category of vector spaces of countable (possibly infinite)
  dimension, and prove that it is the trivial group.
\end{exer}
\begin{exer}{}{exer:linalg:grothendieck-grp-fin-gen-abelian-iso-z}
  Let $\categ{Ab}^{fg}$ be the category of finitely generated abelian
  groups. Define a Grothendieck group of this category in the style of
  the construction of $K(k\text{-}\categ{Vect}^f)$, and prove that
  $K(\categ{Ab}^{fg}) \cong \Z$.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:order-hom-from-grothendieck-grp-finite-abelian}
  Let $\categ{Ab}^f$ be the category of finite abelian groups. Prove
  that assigning to every finite abelian group its order extends to a
  homomorphism from the Grothendieck group $K(\categ{Ab}^f)$ to the
  multiplicative group $(\Q^*,
  \cdot)$. [\ref{exer:linalg:grothendieck-grp-universal-property-fin-length}]
\end{exer}
\begin{exer}{}{exer:linalg:grothendieck-grp-universal-property-fin-length}
  Let $R-\categ{Mod}^f$ be the category of modules of finite length
  (cf.\ \Cref{exer:linalg:jordan-holder-for-modules}) over a ring
  $R$. Let $G$ be an abelian group, and let $\delta$ be a function
  assigning an element of $G$ to every simple $R$-module. Prove that
  $\delta$ extends to a homomorphism from the Grothendieck group of
  $R-\categ{Mod}^f$ to $G$.

  Explain why
  \Cref{exer:linalg:order-hom-from-grothendieck-grp-finite-abelian} is
  a particular case of this observation.

  (For another example, letting $\delta(M) = 1 \in \Z$ for every
  simple module $M$ shows that length itself extends to a homomorphism
  from the Grothendieck group of $R-\categ{Mod}^f$ to $\Z$.)
\end{exer}

\section{Presentations and resolutions}
\label{sec:linalg:presentations-and-resolutions}

After this excursion into the idyllic world of free modules, we can
come back to earth and see if we have learned something that may be
useful for more general situations. Modules over a field are
necessarily free (\Cref{prop:linalg:lin-indep-extends-to-basis}), not so for modules over more
general rings. In fact, this property will turn out to be a
characterization of fields (\Cref{prop:linalg:field-iff-fg-modules-free}).

It is important that we develop some understanding of nonfree modules.
In this section we will see that homomorphisms of free modules carry
enough information to allow us to deal with many nonfree modules.

\subsection{Torsion}
\label{subsec:linalg:torsion}
There are several ways in which a module $M$ may fail to be free: the
most spectacular one is that $M$ may have \emph{torsion}.

\begin{defn}{}{defn:linalg:torsion-element-and-module}
  Let $M$ be an $R$-module. An element $m \in M$ is a \emph{torsion
    element} if $\{m\}$ is linearly dependent, that is, if
  $\exists r \in R$, $r \neq 0$, such that $rm = 0$. The subset of
  torsion elements of $M$ is denoted $\Tor_R(M)$. A module $M$ is
  \emph{torsion-free} if $\Tor_R(M) = \{0\}$. A \emph{torsion module}
  is a module $M$ in which every element is a torsion element.
\end{defn}
The subscript $R$ is usually omitted if there is no uncertainty about
the base ring.

A commutative ring is torsion-free as a module over itself if and only
if it is an integral domain; this is a good reason to limit the
discussion to integral domains in this chapter. Also, the reader will
check (\Cref{exer:linalg:torsion-submodule-needs-domain}) that if $R$
is an integral domain, then $\Tor(M)$ is a \emph{submodule} of
$M$. Equally easy is the following observation:
\begin{lem}{}{lem:linalg:torsion-free-closure-properties}
  Submodules and direct sums of torsion-free modules are torsion-free.
  Free modules over an integral domain are torsion-free.
\end{lem}
\begin{proof}
  The first statement is immediate; the second follows from the first,
  since an integral domain is torsion-free as a module over itself.
\end{proof}

\Cref{lem:linalg:torsion-free-closure-properties} gives a good source of torsion-free modules: for
example, ideals in an integral domain $R$ are torsion-free (because
they are submodules of the free module $R^1$). In fact, ideals provide
us with examples of another mechanism in which a module may fail to be
free.

\begin{exam}{}{exam:linalg:nonprincipal-ideal-torsion-free-not-free}
  Let $R = \Z[x]$, and let $I = (2,x)$. Then $I$ is not a free
  $R$-module. More generally, let $I$ be any nonprincipal ideal of an
  integral domain $R$; then $I$ is a torsion-free module which is not
  free.
\end{exam}
Indeed, if $I$ were free, then its rank would have to be $1$ at most,
by \Cref{prop:linalg:basis-cardinality-comparison} (a basis for $I$ would be a linearly independent
subset of $R$, and $R$ has rank $1$ over itself); thus one element
would suffice to generate $I$, and $I$ would be principal.

What is behind this example is a characterization of PIDs in terms of
`torsion-free submodules of a rank-1 free module'
(\Cref{exer:linalg:domain-is-pid-iff-submodules-of-r-free}). This is a
facet of the main result towards which we are heading, that is, the
classification of finitely generated modules over PIDs
(\Cref{thm:linalg:structure-theorem-fg-modules-over-pid}). The gist of this classification is that finitely
generated modules over a PID can be decomposed into cyclic modules. We
have essentially proved this fact already for modules over Euclidean
domains (it follows from \Cref{prop:linalg:euclidean-domain-normal-form}; see
\Cref{exer:linalg:fin-gen-abel-grp-direct-sum-cyclic-via-matrices}),
and we have looked in great detail at the particular case of
$\Z$-modules, a.k.a.\ abelian groups
(\S{}\ref{sec:groups2:finite-abelian-groups}); we are almost ready to
deal with the general case of PIDs.

\begin{defn}{}{defn:linalg:cyclic-module}
  An $R$-module $M$ is \emph{cyclic} if it is generated by a
  singleton, that is, if $M \cong R/I$ for some ideal $I$ of $R$.
\end{defn}
The equivalence in the definition is hopefully clear to our reader, as
an immediate consequence of the first isomorphism theorem for modules
(\Cref{coro:rings:module-first-iso}). If not, go back and (re)do
\Cref{exer:rings:cyclic-module}.

Cyclic modules are witness to the difference between fields and more
general rings: over a field $k$, a cyclic module is just a
1-dimensional vector space, that is a `copy of $k$'; over more general
rings, cyclic modules may be very interesting (think of the many hours
spent contemplating cyclic groups). In fact, we can tell that a ring
is a field by just looking at its cyclic modules:
\begin{lem}{}{lem:linalg:cyclic-modules-torsion-free-implies-field}
  Let $R$ be an integral domain. Assume that every cyclic $R$-module
  is torsion-free. Then $R$ is a field.
\end{lem}
\begin{proof}
  Let $c \in R$, $c \neq 0$; then $M = R/(c)$ is a cyclic module. Note
  that $\Tor(M) = M$: indeed, the class of $1$ generates $R/(c)$ and
  belongs to $\Tor(M)$ since $c \cdot 1$ is $0$ mod $(c)$ and
  $c \neq 0$. However, by hypothesis $M$ is torsion-free; that is,
  $\Tor(M) = \{0\}$. Therefore $M = \Tor(M)$ is the zero module.

  This shows $R/(c)$ is the zero $R$-module; that is, $(c) = (1)$.
  Therefore, $c$ is a unit. Thus every nonzero element of $R$ is a
  unit, proving that $R$ is a field.
\end{proof}

\Cref{lem:linalg:cyclic-modules-torsion-free-implies-field} is a simple-minded illustration of the fact that we
can study a ring $R$ by studying the module structure over $R$, that
is, the category $R$-$\categ{Mod}$, and that we may not even need to
look at the whole of $R$-$\categ{Mod}$ to be able to draw strong
conclusions about $R$.

\subsection{Finitely presented modules and free resolutions}
\label{subsec:linalg:finitely-presented-modules-and-free-resolutions}
The `right' way to think of a cyclic $R$-module $M$ is as a module
which admits an epimorphism from $R$, viewing the latter as the free
rank-1 $R$-module\footnote{In context the exactness of a sequence of
  $R$-modules will be understood, so the displayed sequence is a way
  to denote the fact that there exists a surjective homomorphism of
  $R$-modules from $R$ to $M$; cf.\
  \Cref{exam:rings:exact-at-n-epimorphism}. Also note the convention
  of denoting $R$ by $R^1$ when it is viewed as a module over
  itself.}:
\[
  R^1 \to M \to 0.
\]
The fact that $M$ is surjected upon by a free $R$-module is nothing
special. In fact, every module $M$ admits such an epimorphism:
\[
  R^{\oplus A} \to M \to 0
\]
provided that we are willing to take $A$ large enough; if we are
desperate, $A = M$ will surely do. This is immediate from the
universal property of free modules; if the reader does not agree, it
is time to go back and review
\S{}\ref{subsec:rings:free-modules-and-free-algebras}. What makes
cyclic modules special is that $A$ can be chosen to be a singleton.

We are now going to focus on a case which is also special, but not
quite as special as cyclic modules: \emph{finitely generated} modules
are modules for which we can choose $A$ to be a finite set (cf.\
\S{}\ref{subsec:rings:submodule-generated-by-a-subset-noetherian-modules}). Thus,
we will assume that $M$ admits an epimorphism from a finite-rank free
module:
\[
  R^m \xrightarrow{\pi} M \to 0
\]
for some integer $m$. The image by $\pi$ of the $m$ vectors in a basis
of $R^m$ is a set of generators for $M$.

Finitely generated modules are much easier to handle than arbitrary
modules. For example, an ideal of $R$ can tell us whether a finitely
generated module is torsion.
\begin{defn}{}{defn:linalg:annihilator}
  The \emph{annihilator} of an $R$-module $M$ is
  \[
    \operatorname{Ann}_R(M) := \{r \in R \mid \forall m \in M, rm =
    0\}.
  \]
\end{defn}
The subscript is usually omitted. The reader will check
(\Cref{exer:linalg:annihilator-is-ideal-torsion-criterion}) that
$\operatorname{Ann}(M)$ is an ideal of $R$ and that if $M$ is a
finitely generated module and $R$ is an integral domain, then $M$ is
torsion if and only if $\operatorname{Ann}(M) \neq 0$.

We would like to develop tools to deal with finitely generated
modules.  It turns out that matrices allow us to describe a
comfortably large collection of such modules.
\begin{defn}{}{defn:linalg:finitely-presented-module}
  An $R$-module $M$ is \emph{finitely presented} if for some positive
  integers $m, n$ there is an exact sequence
  \[
    R^n \xrightarrow{\varphi} R^m \to M \to 0.
  \]
  Such a sequence is called a \emph{presentation} of $M$.
\end{defn}
In other words, finitely presented modules are cokernels (cf.\
III.6.2) of homomorphisms between finitely generated free modules.
Everything about $M$ must be encoded in the homomorphism $\varphi$;
therefore, we should be able to describe the module $M$ by studying
the matrix corresponding to $\varphi$.

There is a gap between finitely presented modules and finitely
generated modules, but on reasonable rings the two notions coincide:
\begin{lem}{}{lem:linalg:noetherian-fg-implies-finitely-presented}
  If $R$ is a Noetherian ring, then every finitely generated
  $R$-module is finitely presented.
\end{lem}
\begin{proof}
  If $M$ is a finitely generated module, there is an exact sequence
  \[
    R^m \xrightarrow{\pi} M \to 0
  \]
  for some $m$. Since $R$ is Noetherian, $R^m$ is Noetherian as an
  $R$-module
  (\Cref{coro:rings:fg-module-over-noetherian-is-noetherian}). Thus
  $\ker\pi$ is finitely generated; that is, there is an exact sequence
  \[
    R^n \to \ker\pi \to 0
  \]
  for some $n$. Putting together the two sequences gives a
  presentation of $M$.
\end{proof}

Once we have gone one step to obtain generators and two steps to get a
presentation, we should hit upon the idea to keep going:
\begin{defn}{}{defn:linalg:free-resolution}
  A \emph{resolution} of an $R$-module $M$ by finitely generated free
  modules is an exact complex
  \[
    \dots \to R^{m_3} \to R^{m_2} \to R^{m_1} \to R^{m_0} \to M \to 0.
  \]
\end{defn}
Iterating the argument proving \Cref{lem:linalg:noetherian-fg-implies-finitely-presented} shows that if $R$ is
Noetherian, then every finitely generated module has a resolution as
in \Cref{defn:linalg:free-resolution}.

It is an important conceptual step to realize that $M$ may be studied
by studying an exact complex of free modules
\[
  \dots \to R^{m_3} \to R^{m_2} \to R^{m_1} \to R^{m_0}
\]
resolving $M$, that is, such that $M$ is the cokernel of the last map.
The $R^{m_0}$ piece keeps track of the generators of $M$; $R^{m_1}$
accounts for the relations among these generators; $R^{m_2}$ records
relations among the relations; and so on.

Developing this idea in full generality would take us too far for now:
for example, we would have to deal with the fact that every module
admits many different resolutions (for example, we can bump up every
$m_i$ by one by direct-summing each term in the complex with a copy of
$R^1$, sent to itself by the maps in the complex). We will do this
very carefully later on, in \Cref{chap:homol}.

However, we can already learn something by considering coarse
questions, such as `how long' a resolution can be. \emph{A priori},
there is no reason to expect a free resolutions to be `finite', that
is, such that $m_i = 0$ for $i \gg 0$. Such finiteness conditions tell
us something special about the base ring $R$.

The first natural question of this type is, for which rings $R$ is it
the case that every finitely generated $R$-module $M$ has a free
resolution `of length 0', that is, stopping at $m_0$? That would mean
that there is an exact sequence
\[
  0 \to R^{m_0} \to M \to 0.
\]
Therefore, $M$ itself must be free. What does this say about $R$?

\begin{prop}{}{prop:linalg:field-iff-fg-modules-free}
  Let $R$ be an integral domain. Then $R$ is a field if and only if
  every finitely generated $R$-module is free.
\end{prop}
\begin{proof}
  If $R$ is a field, then every $R$-module is free, by
  \Cref{prop:linalg:lin-indep-extends-to-basis}.  For the converse, assume that every finitely
  generated $R$-module is free; in particular, every cyclic module is
  free; in particular, every cyclic module is torsion-free. But then
  $R$ is a field, by \Cref{lem:linalg:cyclic-modules-torsion-free-implies-field}.
\end{proof}

The next natural question concerns rings for which finitely generated
modules admit free resolutions of length $1$. It is convenient to
phrase the question in stronger terms, that is, to require that for
every finitely generated $R$-module $M$ and every beginning of a free
resolution
\[
  R^{m_0} \xrightarrow{\pi} M \to 0,
\]
the resolution can be completed to a length $1$ free resolution. This
would amount to demanding that there exist an integer $m_1$ and an
$R$-module homomorphism $R^{m_1} \to R^{m_0}$ such that the sequence
\[
  0 \to R^{m_1} \to R^{m_0} \xrightarrow{\pi} M \to 0
\]
is exact. Equivalently, this condition requires that the module
$\ker\pi$ of relations among the $m_0$ generators necessarily be free.

\begin{prop}{}{prop:linalg:submodule-freeness-property-implies-pid}
  Let $R$ be an integral domain satisfying this property. Then $R$ is
  a PID.
\end{prop}
\begin{proof}
  Let $I$ be an ideal of $R$, and apply the condition to $M =
  R/I$. Since we have an epimorphism
  \[
    R^1 \xrightarrow{\pi} R/I \to 0,
  \]
  the condition says that $\ker\pi$ is free; that is, $I$ is
  free. Since $I$ is a free submodule of $R$, which is free of rank
  $1$, $I$ must be free of rank $\leq 1$ by
  \Cref{prop:linalg:basis-cardinality-comparison}. Therefore $I$ is generated by one element, as
  needed.
\end{proof}

The classification result for finitely generated modules over PIDs
(\Cref{thm:linalg:structure-theorem-fg-modules-over-pid}), which I keep bringing up, will essentially be a
converse to \Cref{prop:linalg:submodule-freeness-property-implies-pid}: the mysterious condition requiring
free resolutions of finitely generated modules to have length at most
$1$ turns out to be a characterization of PIDs, just as the length $0$
condition is a characterization of fields (as proved in
\Cref{prop:linalg:field-iff-fg-modules-free}). We will work this out in
\S{}\ref{subsec:linalg:pids-and-resolutions}.

\subsection{Reading a presentation}
\label{subsec:linalg:reading-a-presentation}
Let us return to the brilliant idea of studying a finitely presented
module $M$ by studying a homomorphism of free modules
\[
  \tag{*} \varphi : R^n \to R^m
\]
such that $M = \coker\varphi$. As we know, we can describe $\varphi$
completely by considering a matrix $A$ representing it, and therefore
we can describe any finitely presented module by giving a matrix
corresponding to (a homomorphism corresponding to) it.

In many cases, judicious use of the material developed in
\S{}\ref{sec:linalg:homomorphisms-of-free-modules-i} allows us to
determine the module $M$ explicitly. For example, take
\[
  \begin{pmatrix} 1 & 3 \\ 2 & 3 \\ 5 & 9 \end{pmatrix};
\]
this matrix corresponds to a homomorphism $\Z^2 \to \Z^3$, hence to a
$\Z$-module, that is, a finitely generated abelian group $G$. The
reader should figure out what $G$ is more explicitly (in terms of the
classification of \S{}\ref{sec:groups2:finite-abelian-groups}; cf.\
\Cref{exer:linalg:fin-gen-abel-grp-direct-sum-cyclic-via-matrices})
before reading on. In the rest of this section I will simply tie up
loose ends into a more concrete recipe to perform these operations.

Incidentally, a number of software packages can perform sophisticated
operations on modules (say, over polynomial rings); a personal
favorite is \texttt{Macaulay2}. These packages rely precisely on the
correspondence between modules and matrices: with due care, every
operation on modules (such as direct sums, tensors, quotients, etc.)
can be executed on the corresponding matrices. For example,
\begin{lem}{}{lem:linalg:block-matrix-for-direct-sum}
  Let $A$, $B$ be matrices with entries in an integral domain $R$, and
  let $M$, $N$ denote the corresponding $R$-modules. Then $M \oplus N$
  corresponds to the block matrix
  \[
    \left(
      \begin{array}{c|c}
        A & 0 \\
        \hline
        0 & B
      \end{array}
    \right).
  \]
\end{lem}
\begin{proof}
  This follows immediately from
  \Cref{exer:linalg:coker-of-direct-sum-of-homomorphisms}.
\end{proof}

Coming back to $(*)$, note that the module $M$ cannot know which bases
we have chosen for $R^n$ or $R^m$; that is, $M = \coker\varphi$ really
depends on the homomorphism $\varphi$, not on the specific matrix
representation we have chosen for $\varphi$. This is an issue that we
have already encountered, and treated rather thoroughly, in
\S{}\ref{subsec:linalg:change-of-basis} and following: `equivalent'
matrices represent the same homomorphism and hence the same module. In
the context we are exploring now, \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units} tells us that two
matrices $A$, $B$ represent the same module $M$ if there exist
invertible matrices $P$, $Q$ such that $B = PAQ$.

But this is not the whole story. Two different homomorphisms
$\varphi_1, \varphi_2$ may have isomorphic cokernels, even if they act
between different modules: the extreme case being any isomorphism
\[
  R^m \to R^m,
\]
whose cokernel is $0$ (regardless of the isomorphism and no matter
what $m$ is). Therefore, if a matrix $A'$ corresponds to a module $M$,
then (by \Cref{lem:linalg:block-matrix-for-direct-sum}) so does the block matrix
\[
  A = \left(
    \begin{array}{c|c}
      I_r & 0 \\
      \hline
      0 & A'
    \end{array}
  \right),
\]
where $I_r$ is the $r \times r$ identity matrix (and $r$ is any
nonnegative integer); in fact, $I_r$ could be replaced here by any
invertible matrix.

The following proposition attempts to formalize these observations.
\begin{prop}{}{prop:linalg:presentation-matrix-operations-preserve-module}
  Let $A$ be a matrix with entries in an integral domain $R$, and let
  $B$ be obtained from $A$ by any sequence of the following
  operations:
  \begin{itemize}
  \item switch two rows or two columns;
  \item add to one row (resp., column) a multiple of another row
    (resp., column);
  \item multiply all entries in one row (or column) by a unit of $R$;
  \item if a unit is the only nonzero entry in a row (or column),
    remove the row and column containing that entry.
  \end{itemize}
  Then $B$ represents the same $R$-module as $A$, up to isomorphism.
\end{prop}
\begin{proof}
  The first three operations are the `elementary operations' of
  \S{}\ref{subsec:linalg:elementary-operations-and-gaussian-elimination},
  and they transform a matrix into an equivalent one (by
  \Cref{prop:linalg:equivalent-via-elementary-operations}); as observed above, this does not affect the
  corresponding module, up to isomorphism.

  As for the fourth operation, if $u$ is a unit and the only nonzero
  entry in (say) a row, then by applications of the second elementary
  operation we may assume that $u$ is also the only nonzero entry in
  its column; without loss of generality we may assume that $u$ is in
  fact the $(1,1)$ entry of the matrix; that is, the matrix is in
  block form:
  \[
    A = \left(
      \begin{array}{c|c}
        u & 0 \\
        \hline
        0 & A'
      \end{array}
    \right).
  \]
  But then $A$ and $A'$ represent the same module, as needed.
\end{proof}

\begin{exam}{}{exam:linalg:abelian-group-from-presentation-matrix}
  The matrix with integer entries
  \[
    \begin{pmatrix} 1 & 3 \\ 2 & 3 \\ 5 & 9 \end{pmatrix}
  \]
  determines an abelian group $G$. Subtract three times the first
  column from the second column, obtaining
  \[
    \begin{pmatrix} 1 & 0 \\ 2 & -3 \\ 5 & -6 \end{pmatrix};
  \]
  the $(1,1)$ entry is a unit and the only nonzero entry in the first
  row, so we can remove the first row and column:
  \[
    \begin{pmatrix} -3 \\ -6 \end{pmatrix};
  \]
  now change the sign, and subtract twice the first row from the
  second, leaving
  \[
    \begin{pmatrix} 3 \\ 0 \end{pmatrix}.
  \]
  Therefore $G$ is isomorphic to the cokernel of the homomorphism
  \[
    \varphi : \Z \to \Z \oplus \Z
  \]
  mapping $1$ to $(3,0)$. This homomorphism is injective and
  identifies $\Z$ with the subgroup $3\Z \oplus 0$ of the target.
  Therefore
  \[
    G \cong \coker\varphi \cong \frac{\Z \oplus \Z}{3\Z \oplus 0}
    \cong \frac{\Z}{3\Z} \oplus \Z.
  \]
\end{exam}

\subsection*{Exercises}

By virtue of Gaussian elimination, the `algorithm' implicitly
described in \Cref{prop:linalg:presentation-matrix-operations-preserve-module} will work without fail over Euclidean
domains (e.g., over the polynomial ring in one variable over a field),
in the sense that it will identify the finitely generated module
corresponding to a matrix with an explicit direct sum of cyclic
modules, as in \Cref{exam:linalg:abelian-group-from-presentation-matrix}. This is too much to expect over
more general rings, since in general elementary transformations do not
generate $\mathrm{GL}$; cf.\ \Cref{rem:linalg:elementary-matrices-generate-gln-euclidean-domain}.
\begin{exer}{\exerused}{exer:linalg:torsion-submodule-needs-domain}
  Prove that if $R$ is an integral domain and $M$ is an $R$-module,
  then $\Tor(M)$ is a submodule of $M$. Give an example showing that
  the hypothesis that $R$ is an integral domain is necessary.
  [\S{}\ref{subsec:linalg:torsion}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:hom-torsion-free-if-target-torsion-free}
  Let $M$ be a module over an integral domain $R$, and let $N$ be a
  torsion-free module. Prove that $\Hom_R(M,N)$ is torsion-free. In
  particular, $\Hom_R(M,R)$ is torsion-free. (We will run into this
  fact again; see \Cref{prop:linrep:dual-module-torsion-free}.)
  [\S{}\ref{subsec:linrep:duals-and-matrices-biduality}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:domain-is-pid-iff-submodules-of-r-free}
  Prove that an integral domain $R$ is a PID if and only if every
  submodule of $R$ itself is free. [\S{}\ref{subsec:linalg:torsion},
  \ref{exer:linalg:torsion-free-iff-free-characterizes-pids}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:annihilator-is-ideal-torsion-criterion}
  Let $R$ be a commutative ring and $M$ an $R$-module.
  \begin{itemize}
  \item Prove that $\operatorname{Ann}(M)$ is an ideal of $R$.
  \item If $R$ is an integral domain and $M$ is finitely generated,
    prove that $M$ is torsion if and only if
    $\operatorname{Ann}(M) \neq 0$.
  \item Give an example of a torsion module $M$ over an integral
    domain, such that $\operatorname{Ann}(M) = 0$. (Of course this
    example cannot be finitely generated!)
  \end{itemize}
  [\S{}\ref{subsec:linalg:finitely-presented-modules-and-free-resolutions},
  \S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:associated-primes-basics}
  Let $M$ be a module over a commutative ring $R$. Prove that an ideal
  $I$ of $R$ is the annihilator of an element of $M$ if and only if
  $M$ contains an isomorphic copy of $R/I$ (viewed as an $R$-module).

  The \emph{associated primes} of $M$ are the prime ideals among the
  ideals $\operatorname{Ann}(m)$, for $m \in M$. The set of the
  associated primes of a module $M$ is denoted
  $\operatorname{Ass}_R(M)$. Note that every prime in
  $\operatorname{Ass}_R(M)$ contains
  $\operatorname{Ann}_R(M)$. [\ref{exer:linalg:maximal-annihilators-are-prime},
  \ref{exer:linalg:prime-filtration-of-fin-gen-module},
  \ref{exer:linalg:primes-in-elem-divisors-are-assoc-primes}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:maximal-annihilators-are-prime}
  Let $M$ be a module over a commutative ring $R$, and consider the
  family of ideals $\operatorname{Ann}(m)$, as $m$ ranges over the
  nonzero elements of $M$. Prove that the maximal elements in this
  family are prime ideals of $R$. Conclude that if $R$ is Noetherian,
  then $\operatorname{Ass}_R(M) \neq \emptyset$ (cf.\
  \Cref{exer:linalg:associated-primes-basics}).
  [\ref{exer:linalg:prime-filtration-of-fin-gen-module},
  \ref{exer:linalg:union-of-annihilators-eq-union-of-assoc-primes}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:prime-filtration-of-fin-gen-module}
  Let $R$ be a commutative Noetherian ring, and let $M$ be a finitely
  generated module over $R$. Prove that $M$ admits a finite series
  \[
    M = M_0 \supsetneq M_1 \supsetneq \dots \supsetneq M_m = \langle 0
    \rangle
  \]
  in which all quotients $M_i/M_{i+1}$ are of the form
  $R/\mathfrak{p}$ for some prime ideal $\mathfrak{p}$ of $R$.

  (Hint: Use
  \Cref{exer:linalg:associated-primes-basics,exer:linalg:maximal-annihilators-are-prime}
  to show that $M$ contains an
  isomorphic copy $M'$ of $R/\mathfrak{p}_1$ for some prime
  $\mathfrak{p}_1$. Then do the same with $M/M'$, producing an
  $M'' \supseteq M'$ such that $M''/M' \cong R/\mathfrak{p}_2$ for
  some prime $\mathfrak{p}_2$. Why must this process stop after
  finitely many steps?)
  [\ref{exer:linalg:assoc-primes-appear-in-prime-filtration}]
\end{exer}
\begin{exer}{}{exer:linalg:assoc-primes-appear-in-prime-filtration}
  Let $R$ be a commutative Noetherian ring, and let $M$ be a finitely
  generated module over $R$. Prove that every prime in
  $\operatorname{Ass}_R(M)$ appears in the list of primes produced by
  the procedure presented in
  \Cref{exer:linalg:prime-filtration-of-fin-gen-module}. (If
  $\mathfrak{p}$ is an associated prime, then $M$ contains an
  isomorphic copy $N$ of $R/\mathfrak{p}$. With notation as in the
  hint in \Cref{exer:linalg:prime-filtration-of-fin-gen-module}, prove
  that either $\mathfrak{p}_1 = \mathfrak{p}$ or $N \cap M' = 0$. In
  the latter case, $N$ maps isomorphically to a copy of
  $R/\mathfrak{p}$ in $M/M'$; iterate the reasoning.)

  In particular, if $M$ is a finitely generated module over a
  Noetherian ring, then $\operatorname{Ass}(M)$ is finite.
\end{exer}
\begin{exer}{}{exer:linalg:union-of-annihilators-eq-union-of-assoc-primes}
  Let $M$ be a module over a commutative Noetherian ring $R$. Prove
  that the union of all annihilators of nonzero elements of $M$ equals
  the union of all associated primes of $M$. (Use
  \Cref{exer:linalg:maximal-annihilators-are-prime}.)

  Deduce that the union of the associated primes of a Noetherian ring
  $R$ (viewed as a module over itself) equals the set of zero-divisors
  of $R$.
\end{exer}
\begin{exer}{}{exer:linalg:intersection-of-assoc-primes-eq-nilradical}
  Let $R$ be a commutative Noetherian ring. One can prove that the
  minimal primes of $\operatorname{Ann}(M)$ (cf.\
  \Cref{exer:irred:noeth-implies-fin-minimal-primes}) are in
  $\operatorname{Ass}(M)$. Assuming this, prove that the intersection
  of the associated primes of a Noetherian ring $R$ (viewed as a
  module over itself) equals the nilradical of $R$.
\end{exer}
\begin{exer}{}{exer:linalg:group-vs-module-presentations}
  Review the notion of presentation of a group,
  (\S{}\ref{subsec:groups:presentations}), and relate it to the notion
  of presentation introduced in
  \S{}\ref{subsec:linalg:finitely-presented-modules-and-free-resolutions}.
\end{exer}
\begin{exer}{}{exer:linalg:fin-gen-module-over-poly-quotient-fin-presented}
  Let $\mathfrak{p}$ be a prime ideal of a polynomial ring
  $k[x_1, \dots, x_n]$ over a field $k$, and let
  $R = k[x_1, \dots, x_n]/\mathfrak{p}$. Prove that every finitely
  generated module over $R$ has a finite presentation.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:koszul-complex-length-2}
  Let $R$ be a commutative ring. A tuple $(a_1, a_2, \dots, a_n)$ of
  elements of $R$ is a \emph{regular sequence} if $a_1$ is a
  non-zero-divisor in $R$, $a_2$ is a non-zero-divisor
  modulo\footnote{That is, the class of $a_2$ in $R/(a_1)$ is a
    non-zero-divisor in $R/(a_1)$.} $(a_1)$, $a_3$ is a
  non-zero-divisor modulo $(a_1, a_2)$, and so on.

  For $a, b$ in $R$, consider the following complex of $R$-modules:
  \[
    \tag{*} 0 \to R \xrightarrow{d_2} R \oplus R \xrightarrow{d_1} R
    \xrightarrow{\pi} \frac{R}{(a,b)} \to 0
  \]
  where $\pi$ is the canonical projection, $d_1(r,s) = ra + sb$, and
  $d_2(t) = (bt, -at)$. Put otherwise, $d_1$ and $d_2$ correspond,
  respectively, to the matrices
  \[
    \begin{pmatrix} a & b \end{pmatrix}, \quad
    \begin{pmatrix} b \\ -a \end{pmatrix}.
  \]
  \begin{itemize}
  \item Prove that this is indeed a complex, for every $a$ and $b$.
  \item Prove that if $(a,b)$ is a regular sequence, this complex is
    exact.
  \end{itemize}
  The complex $(*)$ is called the \emph{Koszul complex} of
  $(a,b)$. Thus, when $(a,b)$ is a regular sequence, the Koszul
  complex provides us with a free resolution of the module
  $R/(a,b)$. [\ref{exer:linalg:koszul-complex-length-3},
  \ref{exer:linalg:krull-dim-geq-2-example},
  \ref{exer:linrep:koszul-complex-definition}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:koszul-complex-length-3}
  A Koszul complex may be defined for any sequence $a_1, \dots, a_n$
  of elements of a commutative ring $R$. The case $n = 2$ seen in
  \Cref{exer:linalg:koszul-complex-length-2} and the case $n = 3$
  reviewed here will hopefully suffice to get a gist of the general
  construction; the general case will be given in VIII.4.22.

  Let $a, b, c \in R$. Consider the following complex:
  \[
    0 \to R \xrightarrow{d_3} R \oplus R \oplus R \xrightarrow{d_2} R
    \oplus R \oplus R \xrightarrow{d_1} R \xrightarrow{\pi}
    \frac{R}{(a,b,c)} \to 0
  \]
  where $\pi$ is the canonical projection and the matrices for
  $d_1, d_2, d_3$ are, respectively,
  \[
    \begin{pmatrix} a & b & c \end{pmatrix}, \quad
    \begin{pmatrix} 0 & -c & -b \\ -c & 0 & a \\ b & a &
      0 \end{pmatrix}, \quad
    \begin{pmatrix} a \\ -b \\ c \end{pmatrix}.
  \]
  \begin{itemize}
  \item Prove that this is indeed a complex, for every $a, b, c$.
  \item Prove that if $(a,b,c)$ is a regular sequence, this complex is
    exact.
  \end{itemize}
  Koszul complexes are very important in commutative algebra and
  algebraic geometry. [\ref{exer:linrep:koszul-complex-definition}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:free-resolution-of-z-over-zxy}
  View $\Z$ as a module over the ring $R = \Z[x,y]$, where $x$ and $y$
  act by $0$. Find a free resolution of $\Z$ over
  $R$. [\ref{exer:linrep:rees-algebra-example-iso-to-quotient}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:coker-of-direct-sum-of-homomorphisms}
  Let $\varphi : R^n \to R^m$ and $\psi : R^p \to R^q$ be two
  $R$-module homomorphisms, and let
  \[
    \varphi \oplus \psi : R^n \oplus R^p \to R^m \oplus R^q
  \]
  be the morphism induced on direct sums. Prove that
  \[
    \coker(\varphi \oplus \psi) = \coker\varphi \oplus \coker\psi.
  \]
\end{exer}
\begin{exer}{}{exer:linalg:determine-module-from-matrix-example}
  Determine (as a better known entity) the module represented by the
  matrix
  \[
    \begin{pmatrix}
      1+3x & 2x & 3x \\
      1+2x & 1+2x-x^2 & 2x \\
      x & x^2 & x
    \end{pmatrix}
  \]
  over the polynomial ring $k[x]$ over a field.
\end{exer}

\section{Classification of finitely generated modules over PIDs}
\label{sec:linalg:classification-of-finitely-generated-modules-over-pids}


It is finally time to prove the classification theorem for finitely
generated modules over arbitrary PIDs. We have already proved this
statement in the special case of finite $\Z$-modules
(\S{}\ref{sec:groups2:finite-abelian-groups}), and the diligent reader
has worked out a proof in the less special case of finitely generated
modules over Euclidean domains, in
\Cref{exer:linalg:fin-gen-abel-grp-direct-sum-cyclic-via-matrices}. Now
we go for the real thing.

\subsection{Submodules of free modules}
\label{subsec:linalg:submodules-of-free-modules}
Recall (\Cref{lem:linalg:torsion-free-closure-properties}, \Cref{exam:linalg:nonprincipal-ideal-torsion-free-not-free}) that a submodule of a
free module over an arbitrary integral domain $R$ is necessarily
torsion-free but need not be free. For example, the ideal $I = (x,y)$
of $R = k[x,y]$ (with $k$ a field, for example) is torsion-free as an
$R$-module, but not free: for this to become really, really evident,
it is helpful to rename $x = a$, $y = b$ when these are viewed as
elements of $I$ and observe that $a$ and $b$ are not linearly
independent over $k[x,y]$, since $ya - xb = 0$.

On the other hand, submodules of a free module over a field are
automatically free: simply because every module over a field is free
(\Cref{prop:linalg:lin-indep-extends-to-basis}). It is reasonable to expect that `some property in
between' being a field and being a friendly UFD such as $k[x,y]$ will
guarantee that a submodule of a free module is free. We will now prove
that this property is precisely that of being a \emph{principal ideal
  domain}.
\begin{prop}{}{prop:linalg:submodule-of-free-over-pid-is-free}
  Let $R$ be a PID, let $F$ be a finitely generated free module over
  $R$, and let $M \subseteq F$ be a submodule. Then $M$ is free.
\end{prop}
We will actually prove a more precise result, in view of the full
statement of the classification theorem: we will show that there is a
basis $(x_1, \dots, x_n)$ of $F$ and elements $a_1, \dots, a_m$ of $R$
(with $m \leq n$) such that
\[
  y_1 = a_1 x_1, \dots, \quad y_m = a_m x_m
\]
form a basis of $M$. That is, not only do we prove that $M$ is free,
but we also show that there are `compatible' bases of $F$ and $M$.

In order to do this, we may of course assume that $M \neq 0$:
otherwise there is nothing to prove. Most of the work will then go
into showing that if $M \neq 0$, we can split one direct summand off
$M$; iterating this process will prove the proposition\footnote{In a
  loose sense, this is the same strategy we employed in the proof of
  the classification result for finite abelian groups in
  \S{}\ref{sec:groups2:finite-abelian-groups}.}. This is where the PID
condition is used, so I will single out the main technical point into
the following statement.
\begin{lem}{}{lem:linalg:pid-submodule-structure-step}
  Let $R$ be a PID, let $F$ be a finitely generated free module over
  $R$, and let $M \subseteq F$ be a nonzero submodule. Then there
  exist $a \in R$, $x \in F$, $y \in M$, and submodules
  $F' \subseteq F$ and $M' \subseteq M$, such that $y = ax \neq 0$,
  $M' = F' \cap M$, and
  \[
    F = \langle x \rangle \oplus F', \quad M = \langle y \rangle
    \oplus M'.
  \]
\end{lem}
The reader would find it instructive to pause a moment and try to
imagine how the PID hypothesis may enter into the proof of this lemma.
The PID hypothesis is a hypothesis on $R$, so the question is, where
is the special copy of $R$ to which we can apply it? Don't read on
until you have spent a moment thinking about this.

It is tempting to look for this copy of $R$ among the `factors' of
$F = R^n$: for example, we could map $F$ to $R$ by projecting onto the
first component:
\[
  \pi(r_1, \dots, r_n) = r_1.
\]
But there is nothing special about the first component of $R^n$; in
fact there is nothing special about the particular basis chosen for
$F$, that is, the particular representation of $F$ as
$R^n$. Therefore, this does not look too promising. The way out of
this bind is to democratically map $F$ to $R$ in every possible way:
we consider \emph{all} homomorphisms $\varphi : F \to R$. This is a
set that does not depend on any extra choice, so it is more likely to
carry the information we need.

For each $\varphi$, $\varphi(M)$ is a submodule of $R$, that is, an
ideal of $R$; so we have a chance to use the PID hypothesis with
profit. In fact, the much weaker fact that $R$ is Noetherian
guarantees already that there must be some homomorphism $\alpha$ for
which $\alpha(M)$ is maximal among all ideals $\varphi(M)$. This copy
of $R$, that is, the target of such a homomorphism $\alpha$, is
special for a good reason, independent of inessential choices. The
fact that $R$ is a PID tells us that $\alpha(M)$ is principal, and
then we are clearly in business.

Here is the formal argument:
\begin{proof}
  For all $\varphi \in \Hom_R(F,R)$, $\varphi(M)$ is a submodule of
  $R$, that is, an ideal. The family of all these ideals is nonempty,
  and PIDs are Noetherian; therefore (by \Cref{prop:irred:noetherian-module-tfae}) there
  exists a maximal element in the family, say $\alpha(M)$, for a
  homomorphism $\alpha : M \to R$. The fact that $M \neq 0$ implies
  immediately that some $\varphi(M) \neq 0$ (for example, take for
  $\varphi$ the projection to a suitable factor of $R^n$); hence
  $\alpha(M) \neq 0$.

  Since $R$ is a PID, $\alpha(M)$ is principal: $\alpha(M) = (a)$ for
  some $a \in R$, $a \neq 0$. Since $a \in \alpha(M)$, there exists an
  element $y \in M$, $y \neq 0$, such that $\alpha(y) = a$. These are
  the elements $a, y$ mentioned in the statement.

  I claim that $a$ divides $\varphi(y)$ for all
  $\varphi \in \Hom_R(F,R)$.  Indeed, let $b$ be a generator of
  $(a, \varphi(y))$ (which exists since $R$ is a PID; of course $b$ is
  simply a gcd of $a$ and $\varphi(y)$), and let $r, s \in R$ such
  that $b = ra + s\varphi(y)$; consider the homomorphism
  $\psi := r\alpha + s\varphi$. Since $a \in (b)$, we have
  $\alpha(M) \subseteq (b)$. On the other hand
  \[
    b = ra + s\varphi(y) = (r\alpha + s\varphi)(y) = \psi(y) \in
    \psi(M);
  \]
  therefore $(b) \subseteq \psi(M)$. It follows that
  $\alpha(M) \subseteq \psi(M)$, and by maximality
  $\alpha(M) = \psi(M)$; hence $(a) = (b)$, and in particular $a$
  divides $\varphi(y)$, as claimed.

  Let $y = (s_1, \dots, s_n)$ as an element of $F = R^n$. Each $s_i$
  is the image of $y$ by a homomorphism $F \to R$ (that is, the $i$-th
  projection), so $a$ divides all of them by what we just proved.
  Therefore $\exists r_1, \dots, r_n \in R$ such that $s_i = ar_i$;
  let
  \[
    x = (r_1, \dots, r_n) \in F.
  \]
  This is the element $x$ mentioned in the statement. By construction,
  $y = ax$. Further, $a = \alpha(y) = \alpha(ax) = a\alpha(x)$; since
  $R$ is an integral domain and $a \neq 0$, this implies
  $\alpha(x) = 1$.

  Finally, we let $F' = \ker\alpha$ and $M' = F' \cap M$, and we can
  proceed to verify the stated direct sums.

  First, every $z \in F$ may be written as
  \[
    z = \alpha(z)x + (z - \alpha(z)x);
  \]
  by linearity
  \[
    \alpha(z - \alpha(z)x) = \alpha(z) - \alpha(z)\alpha(x) =
    \alpha(z) - \alpha(z) = 0,
  \]
  that is, $z - \alpha(z)x \in \ker\alpha$. This implies that
  $F = \langle x \rangle + F'$. On the other hand,
  $rx \in F' \implies \alpha(rx) = 0 \implies r\alpha(x) = 0 \implies
  r = 0$: that is, $\langle x \rangle \cap F' = 0$. Therefore
  \[
    F = \langle x \rangle \oplus F',
  \]
  as claimed (cf.\
  \Cref{exer:linalg:internal-direct-sum-for-modules}).

  Second, if $z \in M$, then $a$ divides $\alpha(z)$: indeed,
  $\alpha(z) \in \alpha(M) = (a)$. Writing $\alpha(z) = ca$, we have
  $\alpha(z)x = cax = cy$; splitting $z$ as above, we note
  \[
    z - \alpha(z)x = z - cy \in M \cap F' = M',
  \]
  and this leads as before to
  \[
    M = \langle y \rangle \oplus M',
  \]
  concluding the proof.
\end{proof}

Once \Cref{lem:linalg:pid-submodule-structure-step} is established, the proof of \Cref{prop:linalg:submodule-of-free-over-pid-is-free}
is mere busywork:
\begin{proof}[Proof of \Cref{prop:linalg:submodule-of-free-over-pid-is-free}]
  If $M = 0$, we are done. If not, applying \Cref{lem:linalg:pid-submodule-structure-step} to
  $M \subseteq F$ produces an element $y_1 \in M$ and a submodule
  $M^{(1)} \subseteq M$ such that
  \[
    M = \langle y_1 \rangle \oplus M^{(1)}.
  \]
  If $M^{(1)} = 0$, we are done; otherwise, apply \Cref{lem:linalg:pid-submodule-structure-step}
  again (to $M^{(1)} \subseteq F$) to obtain $y_2 \in M^{(1)}$ and
  $M^{(2)} \subseteq M^{(1)}$ so that
  \[
    M = \langle y_1 \rangle \oplus (\langle y_2 \rangle \oplus
    M^{(2)}).
  \]
  This process may be continued, producing elements
  $y_1, \dots, y_m \in M$ such that
  \[
    M = \langle y_1 \rangle \oplus \dots \oplus \langle y_m \rangle
    \oplus M^{(m)},
  \]
  so long as the module $M^{(m)}$ is nonzero. However, by
  \Cref{prop:linalg:basis-cardinality-comparison} we know $m \leq n$, since $y_1, \dots, y_m$ are
  linearly independent in $F$. It follows that the process must stop;
  that is, $M^{(m)} = 0$ for some $m \leq n$. That is,
  \[
    M = \langle y_1 \rangle \oplus \dots \oplus \langle y_m \rangle
  \]
  is free, as needed.
\end{proof}

Note that the proof has not used part of the result of
\Cref{lem:linalg:pid-submodule-structure-step}, that is, the fact that the `factor'
$\langle y \rangle$ of $M$ is a submodule of a corresponding factor
$\langle x \rangle$ of $F$. This is needed in order to upgrade
\Cref{prop:linalg:submodule-of-free-over-pid-is-free} along the lines mentioned after its statement. Here
is that stronger statement:
\begin{coro}{}{coro:linalg:submodule-basis-with-divisibility-chain}
  Let $R$ be a PID, let $F$ be a finitely generated free module over
  $R$, and let $M \subseteq F$ be a submodule. Then there exist a
  basis $(x_1, \dots, x_n)$ of $F$ and nonzero elements
  $a_1, \dots, a_m$ of $R$ ($m \leq n$) such that
  $(a_1 x_1, \dots, a_m x_m)$ is a basis of $M$. Further, we may
  assume $a_1 \mid a_2 \mid \dots \mid a_m$.
\end{coro}
This statement should be compared with \Cref{prop:linalg:euclidean-domain-normal-form}: it amounts
to a `Smith normal form over PIDs'; cf.\ \Cref{rem:linalg:elementary-matrices-generate-gln-euclidean-domain}.
\begin{proof}
  Now that we know that submodules of a free module are free, we see
  that the submodule $F' \subseteq F$ produced in \Cref{lem:linalg:pid-submodule-structure-step} is
  free. The first part of the statement then follows from
  \Cref{lem:linalg:pid-submodule-structure-step}, by an inductive argument analogous to the proof
  of \Cref{prop:linalg:submodule-of-free-over-pid-is-free}, and is left to the reader.

  The most delicate part of the statement is the divisibility
  condition.  By induction it suffices to prove $a_1 \mid a_2$, and
  for this we refer back to the proof of \Cref{lem:linalg:pid-submodule-structure-step}: $(a_1)$ is
  maximal among the ideals $\varphi(M)$ of $R$, as $\varphi$ ranges
  over all homomorphisms $F \to R$. Consider then any\footnote{In
    order to define a homomorphism on a free module $F$, one may
    define it on a basis of $F$ and then extend it by linearity; there
    is no restriction on the possible choices for the images of basis
    elements. The reader who is not sure about this should review the
    universal property of free modules!} homomorphism $\varphi$ such
  that $\varphi(x_1) = \varphi(x_2) = 1$.

  Since $\varphi(y_1) = \varphi(a_1x_1) = a_1$, we have
  $(a_1) \subseteq \varphi(M)$; by maximality $(a_1) = \varphi(M)$.
  Therefore
  $a_2 = \varphi(a_2x_2) = \varphi(y_2) \in \varphi(M) = (a_1)$,
  proving $a_1 \mid a_2$ as needed.
\end{proof}

The logic of the argument given in this section is somewhat tricky: it
takes \Cref{lem:linalg:pid-submodule-structure-step} to prove \Cref{prop:linalg:submodule-of-free-over-pid-is-free}; but then it takes
\Cref{prop:linalg:submodule-of-free-over-pid-is-free} (proving that $F'$ is free) to revisit
\Cref{lem:linalg:pid-submodule-structure-step} and squeeze from it the stronger \Cref{coro:linalg:submodule-basis-with-divisibility-chain}.

\subsection{PIDs and resolutions}
\label{subsec:linalg:pids-and-resolutions}
\Cref{prop:linalg:submodule-of-free-over-pid-is-free} allows us to complete the circle of ideas begun in
\S{}\ref{subsec:linalg:finitely-presented-modules-and-free-resolutions}.
\begin{prop}{}{prop:linalg:pid-characterization-via-presentations}
  Let $R$ be an integral domain. Then $R$ is a PID if and only if for
  every finitely generated $R$-module $M$ and every epimorphism
  \[
    R^{m_0} \xrightarrow{\pi_0} M \to 0,
  \]
  there exist a free $R$-module $R^{m_1}$ and a homomorphism
  $\pi_1 : R^{m_1} \to R^{m_0}$ such that the sequence
  \[
    0 \to R^{m_1} \xrightarrow{\pi_1} R^{m_0} \xrightarrow{\pi_0} M
    \to 0
  \]
  is exact.
\end{prop}
\begin{proof}
  The fact that the stated condition implies that $R$ is a PID was
  proved in \Cref{prop:linalg:submodule-freeness-property-implies-pid}. For the converse, let
  $\pi_0 : R^{m_0} \to M$ be an epimorphism; then $\ker\pi_0$ is free
  by \Cref{prop:linalg:submodule-of-free-over-pid-is-free}; the result follows by choosing any isomorphism
  $\pi_1 : R^{m_1} \to \ker(\pi_0)$.
\end{proof}

The content of \Cref{prop:linalg:pid-characterization-via-presentations} is possibly simpler than its awkward
formulation: \Cref{prop:linalg:pid-characterization-via-presentations} is simply a characterization for PIDs
analogous to the characterization of fields worked out in
\Cref{prop:linalg:field-iff-fg-modules-free}. This is another example of the fact that we can
study a ring $R$ by looking at how $R$-$\categ{Mod}$ is put together.

The reader can well imagine the `length-$n$' version of the condition
appearing in these results: we may ask for the class of integral
domains $R$ with the property that for all finitely generated modules
$M$ and all partial free resolutions of $M$,
\[
  R^{m_{n-1}} \xrightarrow{\pi_{n-1}} \dots \xrightarrow{\pi_1}
  R^{m_0} \xrightarrow{\pi_0} M \to 0,
\]
there exists a homomorphism $\pi_n : R^{m_n} \to R^{m_{n-1}}$ such
that
\[
  0 \to R^{m_n} \xrightarrow{\pi_n} R^{m_{n-1}}
  \xrightarrow{\pi_{n-1}} \dots \xrightarrow{\pi_1} R^{m_0}
  \xrightarrow{\pi_0} M \to 0
\]
is exact. We have proved that this condition characterizes fields for
$n = 0$ and PIDs for $n = 1$.

The alert reader should now have a \emph{d\'ej\`{a} vu}, as we have
already encountered a notion that is $0$ precisely for fields and $1$
for PIDs: the Krull dimension of a ring; cf.\
\Cref{exam:rings:speck-prime-ideals-maximal}. Of course this is not a
coincidence, but the full relation between the Krull dimension and the
bound on the length of free resolutions is beyond the scope of this
book\footnote{The most persistent of readers will take a more
  sophisticated look at these bounds in the exercises to
  \S{}\ref{sec:homol:double-complexes}.}. `Most' rings (even rings
with finite Krull dimension) do not satisfy any such bound for all
modules. The local rings satisfying the finiteness condition described
above for some $n$ correspond in the language of algebraic geometry to
`smooth' points on varieties of dimension $\leq n$.

\subsection{The classification theorem}
\label{subsec:linalg:the-classification-theorem}
The notion of the \emph{rank} of a free module (\Cref{defn:linalg:rank-and-dimension})
extends naturally to every finitely generated module $M$ over an
integral domain $R$.
\begin{defn}{}{defn:linalg:rank-of-fg-module}
  Let $R$ be an integral domain. The \emph{rank} $\operatorname{rk} M$
  of a finitely generated $R$-module $M$ is the maximum number of
  linearly independent elements in $M$.
\end{defn}
It should be clear that this number is finite
(\Cref{exer:linalg:rank-bounded-by-num-generators}).

\begin{thm}{}{thm:linalg:structure-theorem-fg-modules-over-pid}
  Let $R$ be a PID, and let $M$ be a finitely generated
  $R$-module. Then the following hold:
  \begin{itemize}
  \item There exist distinct prime ideals
    $(q_1), \dots, (q_n) \subseteq R$, positive integers $r_{ij}$, and
    an isomorphism
    \[
      M \cong R^{\operatorname{rk} M} \oplus \left( \bigoplus_{i,j}
        \frac{R}{(q_i^{r_{ij}})} \right).
    \]
  \item There exist nonzero, nonunit ideals $(a_1), \dots, (a_m)$ of
    $R$, such that
    $(a_1) \supseteq (a_2) \supseteq \dots \supseteq (a_m)$, and an
    isomorphism
    \[
      M \cong R^{\operatorname{rk} M} \oplus \frac{R}{(a_1)} \oplus
      \dots \oplus \frac{R}{(a_m)}.
    \]
  \end{itemize}
  These decompositions are unique (in the evident sense).
\end{thm}
After all our preparatory work, this statement practically proves
itself. I will describe the arguments in general terms, leaving the
details to the enterprising reader.

The two forms taken by the theorem go under the name of
\emph{invariant factors} and \emph{elementary divisors}, as in the
case of abelian groups. As in the case of abelian groups, the
equivalence between the two formulations amounts to careful
bookkeeping, and we will not prove it formally; see
\S{}\ref{subsec:groups2:invariant-factors-and-elementary-divisors} for
a reminder of how to go from one to the other. The role played by
\Cref{lem:groups2:coprime-order-subgroups-direct-sum} in that discussion is taken over by the Chinese
remainder theorem, \Cref{thm:irred:chinese-remainder-theorem}, in this more general setting.

As the two formulations are equivalent, it suffices to prove the
existence and uniqueness of the decompositions given in
\Cref{thm:linalg:structure-theorem-fg-modules-over-pid} for any one of the two.

The existence is a direct consequence of the results in
\S{}\ref{subsec:linalg:submodules-of-free-modules}.  Indeed, let $M$
be a finitely generated module; thus there is an epimorphism
\[
  R^n \xrightarrow{\pi} M \to 0
\]
where $n$ is the number of generators of $M$. Apply \Cref{coro:linalg:submodule-basis-with-divisibility-chain}
to the submodule $\ker\pi \subseteq R^n$: there exist a basis
$(x_1, \dots, x_n)$ of $R^n$ and nonzero elements $a_1, \dots, a_m$ of
$R$ such that $(a_1 x_1, \dots, a_m x_m)$ is a basis of $\ker\pi$ and
further $a_1 \mid \dots \mid a_m$. That is, $M$ is presented by the
$n \times m$ matrix
\[
  \begin{pmatrix}
    a_1 & \cdots & 0 \\
    \vdots & \ddots & \vdots \\
    0 & \cdots & a_m \\
    0 & \dots & 0 \\
    \vdots & \ddots & \vdots \\
    0 & \dots & 0
  \end{pmatrix}.
\]
By \Cref{prop:linalg:presentation-matrix-operations-preserve-module}, we may assume that $a_1, \dots, a_m$ are not
units: if any one of them is, the corresponding row and column may be
omitted from the matrix (and $n$ corrected accordingly). It follows
that
\[
  M \cong \frac{R}{(a_1)} \oplus \dots \oplus \frac{R}{(a_m)} \oplus
  R^{(n-m)}
\]
with $a_1 \mid \dots \mid a_m$ nonzero nonunits, as prescribed by
\Cref{thm:linalg:structure-theorem-fg-modules-over-pid}, and the existence is proved.

As for the uniqueness of the representations, if
\[
  M \cong R^r \oplus T,
\]
with $T$ a torsion module\footnote{Recall that this means that every
  element of $T$ is a torsion element; cf.\
  \S{}\ref{subsec:linalg:torsion}.}, then $r = \operatorname{rk} M$
and $T \cong \Tor_R(M)$ $\cong \Tor_R(M)$
(\Cref{exer:linalg:direct-proof-rank-and-torsion-submodule}). Thus,
the `free part' and the `torsion part' in the decompositions given in
\Cref{thm:linalg:structure-theorem-fg-modules-over-pid} are determined uniquely.

This reduces the uniqueness question to the structure of torsion
modules: assuming that
\[
  \tag{*} \bigoplus_{i,j} \frac{R}{(p_i^{r_{ij}})} \cong
  \bigoplus_{k,\ell} \frac{R}{(q_k^{s_{k\ell}})},
\]
with $p_i, q_k$ irreducible elements of $R$, the task is to show that
the range of the indices is the same and, up to reordering,
$(p_i) = (q_i)$ and $r_{ij} = s_{ij}$ for all $i, j$.

Since a finitely generated module is torsion if and only if its
annihilator is nonzero
(\Cref{exer:linalg:annihilator-is-ideal-torsion-criterion}), it is
reasonable that $\operatorname{Ann}(M)$ will be of some use here.
\begin{lem}{}{lem:linalg:torsion-module-annihilator-and-primes}
  Let $M$ be a torsion module, expressed as in \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} (with
  $\operatorname{rk} M = 0$). Then $\operatorname{Ann}(M) = (a_m)$.
  Further, the prime ideals $(q_i)$ are precisely the prime ideals of
  $R$ containing $\operatorname{Ann}(M)$.
\end{lem}
\begin{proof}
  By hypothesis
  \[
    M \cong \frac{R}{(a_1)} \oplus \dots \oplus \frac{R}{(a_m)},
  \]
  with $a_1 \mid \dots \mid a_m$. If $r \in \operatorname{Ann}(M)$,
  then
  \[
    0 = r(1, \dots, 1) = (r, \dots, r).
  \]
  In particular $r \equiv 0$ modulo $(a_m)$; that is, $r \in
  (a_m)$. Thus $\operatorname{Ann}(M) \subseteq (a_m)$.

  For the reverse inclusion, assume $r \in (a_m)$ and $y \in M$.
  Identifying $M$ with its decomposition, write
  $y = (y_1, \dots, y_m)$, with $y_i \in R/(a_i)$. Since
  $r \in (a_m) \subseteq (a_i)$, we have $ry_i = 0$ for all $i$;
  therefore $ry = 0$, and $r \in \operatorname{Ann}(M)$ as needed.

  For the second part of the statement, tracing the equivalence
  between the two decompositions in \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} shows that the
  $q_i$'s are precisely the irreducible factors of $a_m$, so the
  assertion follows from the first part.
\end{proof}

By \Cref{lem:linalg:torsion-module-annihilator-and-primes}, the sets $\{p_i\}$, $\{q_k\}$ of irreducibles
appearing in $(*)$ must coincide (up to inessential units): the ideals
they generate are precisely the prime ideals containing
$\operatorname{Ann}(M)$. The fact that the sets coincide also follows
from the observation that the isomorphism in $(*)$ must match like
primes, in the sense that an element of a factor $R/(p_i^{r_{ij}})$ in
the source must land in a combination of quotients in the target by
powers of the same prime $p_i$; this follows by comparing annihilators
(cf.\
\Cref{exer:linalg:annihilator-inclusion-under-homomorphism}). Therefore,
uniqueness is reduced to the case of a single irreducible $q \in R$:
it suffices to show that if
\[
  \frac{R}{(q^{r_1})} \oplus \dots \oplus \frac{R}{(q^{r_m})} \cong
  \frac{R}{(q^{s_1})} \oplus \dots \oplus \frac{R}{(q^{s_n})},
\]
with $r_1 \geq \dots \geq r_m$ and $s_1 \geq \dots \geq s_n$, then
$m = n$ and $r_i = s_i$ for all $i$. This situation reproduces
precisely the key step the reader used in the proof of uniqueness for
the particular case of abelian groups, in
\Cref{exer:groups2:uniqueness-of-p-grp-decomp}. Of course I will not
spoil the reader's fun by giving any more details this time
(\Cref{exer:linalg:complete-uniqueness-proof-classification-thm}).

\begin{rem}{}{rem:linalg:torsion-module-invariant-factors-summary}
  To summarize, the information carried by a torsion module over a PID
  is equivalent to the choice of a selection of nonzero, nonunit
  ideals $(a_1), \dots, (a_m)$ such that
  \[
    (a_1) \supseteq (a_2) \supseteq \dots \supseteq (a_m).
  \]
  We have seen that $(a_m)$ is in fact the annihilator ideal of the
  module; it would be nice if we had a similarly explicit description
  of all invariants $(a_1), \dots, (a_m)$. As a preview of coming
  attractions, we could call the ideal
  \[
    (a_1 \dots a_m)
  \]
  the \emph{characteristic ideal} of the module\footnote{Warning: This
    does not seem to be standard terminology.}. In situations in which
  we can compute both the annihilator and the characteristic ideal,
  comparing them may lead to strong conclusions about the module. For
  example, a torsion module is cyclic if and only if its annihilator
  and characteristic ideals coincide.

  Further, the prime ideals appearing in \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} have been
  characterized as the prime ideals containing the annihilator ideal.
  They may just as well be characterized as those prime ideals
  containing the characteristic ideal, as the reader will check
  (\Cref{exer:linalg:primes-in-elem-divisors-contain-char-ideal}).

  Finally, note that from this point of view it is immediate that the
  characteristic ideal is contained in the annihilator ideal. We will
  run into this fact again, in an important application
  (\Cref{thm:linalg:cayley-hamilton-theorem}).
\end{rem}

\subsection*{Exercises}

\begin{exer}{\exerused}{exer:linalg:internal-direct-sum-for-modules}
  Let $N$, $P$ be submodules of a module $M$, such that
  $N \cap P = \{0\}$ and $M = N + P$. Prove that $M \cong N \oplus
  P$. (This is a word-for-word repetition of \Cref{prop:groups2:disjoint-normal-subgroups-product-is-direct}
  for modules.) [\S{}\ref{subsec:linalg:submodules-of-free-modules}]
\end{exer}
\begin{exer}{}{exer:linalg:torsion-iff-rank-zero}
  Let $R$ be an integral domain, and let $M$ be a finitely generated
  $R$-module. Prove that $M$ is torsion if and only if
  $\operatorname{rk} M = 0$.
\end{exer}
\begin{exer}{}{exer:linalg:complete-proof-submodule-of-free-module-structure}
  Complete the proof of \Cref{coro:linalg:submodule-basis-with-divisibility-chain}.
\end{exer}
\begin{exer}{}{exer:linalg:krull-dim-geq-2-example}
  Let $R$ be an integral domain, and assume that $a, b \in R$ are such
  that $a \neq 0$, $b \notin (a)$, and $R/(a)$, $R/(a,b)$ are both
  integral domains.
  \begin{itemize}
  \item Prove that the Krull dimension of $R$ is at least $2$.
  \item Prove that if $R$ satisfies the finiteness condition discussed
    in \S{}\ref{subsec:linalg:pids-and-resolutions} for some $n$, then
    $n \geq 2$.
  \end{itemize}
  You can prove this second point by appealing to
  \Cref{prop:linalg:pid-characterization-via-presentations}. For a more concrete argument, you should look for
  an $R$-module admitting a free resolution of length $2$ which cannot
  be shortened.
  \begin{itemize}
  \item Prove that $(a,b)$ is a regular sequence in $R$
    (\Cref{exer:linalg:koszul-complex-length-2}).
  \item Prove that the $R$-module $R/(a,b)$ has a free resolution of
    length exactly $2$.
  \end{itemize}
  Can you see how to construct analogous situations with $n \geq 3$
  elements $a_1, \dots, a_n$?
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:direct-summand-of-free-over-local-is-free}
  Recall (\Cref{exer:irred:localization-at-prime-is-local}) that a
  commutative ring is \emph{local} if it has a single maximal ideal
  $\mathfrak{m}$. Let $R$ be a local ring, and let $M$ be a direct
  summand of a finitely generated free $R$-module: that is, there
  exists an $R$-module $N$ such that $M \oplus N$ is a free
  $R$-module.
  \begin{itemize}
  \item Choose elements $m_1, \dots, m_r \in M$ whose cosets mod
    $\mathfrak{m}M$ are a basis of $M/\mathfrak{m}M$ as a vector space
    over the field $R/\mathfrak{m}$. By Nakayama's lemma,
    $M = \langle m_1, \dots, m_r \rangle$
    (\Cref{exer:linalg:nakayama-gives-generators-from-quotient-basis}).
  \item Obtain a surjective homomorphism
    $\pi : F = R^{\oplus r} \to M$.
  \item Show that $\pi$ splits, giving an isomorphism
    $F \cong M \oplus \ker\pi$. (Apply
    \Cref{exer:rings:free-projective} to the surjective homomorphism
    $\pi$ and the free module $M \oplus N$ to obtain a splitting
    $M \to F$; then use \Cref{prop:rings:split-sequence}.)
  \item Show $\ker\pi/\mathfrak{m}\ker\pi = 0$. Use Nakayama's lemma
    (\Cref{exer:linalg:nakayamas-lemma}) to deduce that $\ker\pi = 0$.
  \item Conclude that $M \cong F$ is in fact
    free. [\ref{exer:linrep:fin-gen-flat-over-noeth-local-is-free},
    \ref{exer:linrep:combine-results-reprove-rank-thm},
    \ref{exer:linrep:fin-gen-over-local-ring-projective-iff-free}]
  \end{itemize}
  Summarizing, over a local ring, every direct summand of a finitely
  generated\footnote{The finite rank hypothesis is actually
    unnecessary, but the proof is harder without this condition.} free
  $R$-module is free. Using the terminology we will introduce in
  \Cref{chap:linrep}, we would say that `projective modules over local
  rings are free'. This result has strong implications in algebraic
  geometry, since it underlies the notion of vector bundle.

  Contrast this fact with \Cref{prop:linalg:submodule-of-free-over-pid-is-free}, which shows that, over a
  PID, every submodule of a finitely generated free module is free.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:rank-bounded-by-num-generators}
  Let $R$ be an integral domain, and let
  $M = \langle m_1, \dots, m_r \rangle$ be a finitely generated
  module. Prove that $\operatorname{rk} M \leq r$. (Use
  \Cref{exer:linalg:more-columns-than-rows-linearly-dependent}.)
  [\S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{}{exer:linalg:rank-of-m-equals-rank-of-m-mod-torsion}
  Let $R$ be an integral domain, and let $M$ be a finitely generated
  module over $R$. Prove that
  $\operatorname{rk} M = \operatorname{rk}(M/\Tor(M))$.
\end{exer}
\begin{exer}{}{exer:linalg:rank-r-iff-free-submodule-torsion-quotient}
  Let $R$ be an integral domain, and let $M$ be a finitely generated
  module over $R$. Prove that $\operatorname{rk} M = r$ if and only if
  $M$ has a free submodule $N \cong R^r$, such that $M/N$ is torsion.

  If $R$ is a PID, then $N$ may be chosen so that
  $0 \to N \to M \to M/N \to 0$ splits.
\end{exer}
\begin{exer}{}{exer:linalg:rank-additive-in-exact-sequences}
  Let $R$ be an integral domain, and let
  \[
    0 \to M_1 \to M_2 \to M_3 \to 0
  \]
  be an exact sequence of finitely generated $R$-modules. Prove that
  $\operatorname{rk} M_2 = \operatorname{rk} M_1 + \operatorname{rk}
  M_3$.

  Deduce that `rank' defines a homomorphism from the Grothendieck
  group of the category of finitely generated $R$-modules to $\Z$
  (cf.\
  \S{}\ref{subsec:linalg:euler-characteristic-and-the-grothendieck-group}).
\end{exer}
\begin{exer}{\exerused}{exer:linalg:direct-proof-rank-and-torsion-submodule}
  Let $R$ be an integral domain, $M$ an $R$-module, and assume
  $M \cong R^r \oplus T$, with $T$ a torsion module. Prove directly
  (that is, without using \Cref{thm:linalg:structure-theorem-fg-modules-over-pid}) that
  $r = \operatorname{rk} M$ and $T \cong
  \Tor_R(M)$. [\S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:annihilator-inclusion-under-homomorphism}
  Let $R$ be an integral domain, let $M$, $N$ be $R$-modules, and let
  $\varphi : M \to N$ be a homomorphism. For $m \in M$, show that
  $\operatorname{Ann}(\langle m \rangle) \subseteq
  \operatorname{Ann}(\langle \varphi(m)
  \rangle)$. [\S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:complete-uniqueness-proof-classification-thm}
  Complete the proof of uniqueness in \Cref{thm:linalg:structure-theorem-fg-modules-over-pid}. (The hint in
  \Cref{exer:groups2:uniqueness-of-p-grp-decomp} may be helpful.)
  [\S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{}{exer:linalg:torsion-free-iff-free-characterizes-pids}
  Let $M$ be a finitely generated module over an integral domain $R$.
  Prove that if $R$ is a PID, then $M$ is torsion-free if and only if
  it is free. Prove that this property characterizes PIDs. (Cf.\
  \Cref{exer:linalg:domain-is-pid-iff-submodules-of-r-free}.)
\end{exer}
\begin{exer}{}{exer:linalg:example-not-direct-sum-of-cyclics}
  Give an example of a finitely generated module over an integral
  domain which is not isomorphic to a direct sum of cyclic modules.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:primes-in-elem-divisors-contain-char-ideal}
  Prove that the prime ideals appearing in the elementary divisor
  version of the classification theorem for a torsion module $M$ over
  a PID are the prime ideals containing the characteristic ideal of
  $M$, as defined in
  \Cref{rem:linalg:torsion-module-invariant-factors-summary}. [\S{}\ref{subsec:linalg:the-classification-theorem}]
\end{exer}
\begin{exer}{}{exer:linalg:primes-in-elem-divisors-are-assoc-primes}
  Prove that the prime ideals appearing in the elementary divisor
  version of the classification theorem for a module $M$ over a PID
  are the associated primes of $M$, as defined in
  \Cref{exer:linalg:associated-primes-basics}.
\end{exer}
\begin{exer}{}{exer:linalg:grothendieck-grp-fin-gen-modules-over-pid}
  Let $R$ be a PID. Prove that the Grothendieck group (cf.\
  \S{}\ref{subsec:linalg:euler-characteristic-and-the-grothendieck-group})
  of the category of finitely generated $R$-modules is isomorphic to
  $\Z$.
\end{exer}

\section{Linear transformations of a free module}
\label{sec:linalg:linear-transformations-of-a-free-module}


One beautiful application of the classification theorem for finitely
generated modules over PIDs is the determination of `special' forms
for matrices of linear maps of a vector space to itself.

Several fundamental concepts are associated with the general notion of
a linear map from a vector space to itself, and we will review these
concepts in this section. Not surprisingly, much of the discussion may
be carried out for free modules over any integral domain $R$, and we
will stay at this level of generality in (most of) the section.
\Cref{thm:linalg:structure-theorem-fg-modules-over-pid} will be used with great profit when $R$ is a field,
in the next section.

\subsection{Endomorphisms and similarity}
\label{subsec:linalg:endomorphisms-and-similarity}
Let $R$ be an integral domain. We have considered in some detail the
module $\Hom_R(F,G)$ of $R$-module homomorphisms $F \to G$ between two
free modules. For example, I have argued that describing such
homomorphisms for finitely generated free modules $F$, $G$ amounts to
describing matrices with entries in $R$, up to `equivalence': the
equivalence relation introduced in
\S{}\ref{subsec:linalg:change-of-basis} accounts for the (arbitrary)
choice of bases for $F$ and $G$.

We now shift the focus a little and consider the special case in which
$F = G$, that is, the $R$-module $\End_R(F)$ of endomorphisms $\alpha$
of a fixed free $R$-module $F$:
\[
  F \xrightarrow{\alpha} F.
\]
Note that $\End_R(F)$ is in fact an $R$-algebra: the operation of
composition makes it a ring, compatibly with the $R$-module structure
(cf.\ \Cref{exer:linalg:mn-r-iso-hom-as-r-algebras}).

From the point of view championed in
\S{}\ref{sec:linalg:homomorphisms-of-free-modules-i}, the two copies
of $F$ appearing in $\End_R(F) = \Hom_R(F,F)$ are unrelated; they are
just (any) isomorphic representatives of a free module of a given
rank. There are circumstances, however, in which we need to really
choose one representative $F$ and stick with it: that is, view
$\alpha$ as acting from a selected free module $F$ to \emph{itself},
not just to an isomorphic copy of itself. In this situation we also
say that $\alpha$ is a \emph{linear transformation} of $F$, or an
\emph{operator} on $F$.

From this different point of view it makes sense to compare elements
of $F$ `before and after' we apply $\alpha$; for example, we could ask
whether for some $\mathbf{v} \in F$ we may have
$\alpha(\mathbf{v}) = \lambda\mathbf{v}$ for some $\lambda \in R$, or
more generally whether a submodule $M$ of $F$ may be sent to itself by
$\alpha$. In other words, we can compare the action of $\alpha$ with
the identity\footnote{The existence of the identity is one of the
  axioms of a category; every now and then, it is handy to be able to
  invoke it.} $I : F \to F$.

In terms of matrix representations (in the finite rank case) the
description of $\alpha$ can be carried out as we did in
\S{}\ref{sec:linalg:homomorphisms-of-free-modules-i}, but with one
interesting twist. In \S{}\ref{subsec:linalg:change-of-basis} we dealt
with how the matrix representation of a homomorphism changes when we
change bases in the source \emph{and} in the target. As we are now
\emph{identifying} source and target, we must choose \emph{the same}
basis for both source and target. This leads to a different notion of
equivalence of matrices:
\begin{defn}{}{defn:linalg:similar-matrices}
  Two square matrices $A, B \in \mathcal{M}_n(R)$ are \emph{similar}
  if they represent the same homomorphism $F \to F$ of a free rank-$n$
  module $F$ to itself, up to the choice of a basis for $F$.
\end{defn}
This is clearly an equivalence relation. The analog of
\Cref{prop:linalg:matrices-representing-same-hom-differ-by-units} for this new notion is readily obtained:
\begin{prop}{}{prop:linalg:similarity-conjugation-characterization}
  Two matrices $A, B \in \mathcal{M}_n(R)$ are similar if and only if
  there exists an invertible matrix $P$ such that
  \[
    B = PAP^{-1}.
  \]
\end{prop}
The reader who has really understood \Cref{prop:linalg:matrices-representing-same-hom-differ-by-units} will not need
any detailed proof of this statement: it should be apparent from
staring at the butterfly diagram
\[
  \begin{tikzcd}
    R^n \arrow[dr, "\varphi"] \arrow[dd, bend right=60, "\pi"'] & &
    R^n \arrow[dl, "\varphi"'] \arrow[dd, bend left=60, "\pi"] \\
    & F \arrow[r, "\alpha"] & F & \\
    R^n \arrow[ur, "\psi"'] & & R^n \arrow[ul, "\psi"]
  \end{tikzcd}
\]
which I am essentially copying from
\S{}\ref{subsec:linalg:change-of-basis}. The difference here is that
we are choosing the same basis for source and target; hence the two
triangles keeping track of the change of basis are the same.  Suppose
$A$ (resp., $B$) represents $\alpha$ with respect to the choice of
basis dictated by $\varphi$ (resp., $\psi$); that is, $A$ (resp., $B$)
is the matrix of the `top' (resp., bottom) composition
\[
  \varphi^{-1} \circ \alpha \circ \varphi : R^n \to R^n
\]
(resp., $\psi^{-1} \circ \alpha \circ \psi$). If $P$ is the matrix
representing the change of basis $\pi$, then $B = PAP^{-1}$ simply
because the diagram commutes:
\[
  \psi^{-1} \circ \alpha \circ \psi = (\pi \circ \varphi^{-1}) \circ
  \alpha \circ (\varphi \circ \pi^{-1}) = \pi \circ (\varphi^{-1}
  \circ \alpha \circ \varphi) \circ \pi^{-1}.
\]

\Cref{prop:linalg:similarity-conjugation-characterization} suggests a useful equivalence relation among
endomorphisms:
\begin{defn}{}{defn:linalg:similar-transformations}
  Two $R$-module homomorphisms of a free module $F$ to itself,
  \[
    \alpha, \beta : F \to F,
  \]
  are similar if there exists an automorphism $\pi : F \to F$ such
  that
  \[
    \beta = \pi \circ \alpha \circ \pi^{-1}.
  \]
\end{defn}
In the finite rank case, similar endomorphisms are represented by
similar matrices, and two endomorphisms $\alpha, \beta$ are similar if
and only if they may be represented by the same matrix by choosing
appropriate (possibly different) bases on $F$. The interesting
invariants determined by similar endomorphisms will turn out (not
surprisingly) to be the same.

From a group-theoretic viewpoint, similarity is an eminently natural
notion to study: the group $\mathrm{GL}(F) = \Aut_R(F)$ of
automorphisms of a free module acts on $\End_R(F)$ by conjugation, and
two endomorphisms $\alpha, \beta$ are similar if and only if they are
in the same orbit under this action. If the reader prefers matrices,
the group $\mathrm{GL}_n(R)$ of invertible $n \times n$ matrices with
entries in $R$ acts by conjugation on the module $\mathcal{M}_n(R)$ of
square $n \times n$ matrices, and two matrices are similar if and only
if they are in the same orbit.

Natural questions arise in this context. For example, we should look
for ways to distinguish different orbits: given two endomorphisms
$\alpha, \beta$, can we effectively decide whether $\alpha$ and
$\beta$ are similar? One way to approach this question is to determine
invariants which can distinguish different orbits. Better still, we
can look for `special' representatives within each orbit, that is, of
a given similarity class. In other words, given an endomorphism
$\alpha : F \to F$, find a basis of $F$ with respect to which the
matrix description of $\alpha$ has a particular, predictable shape.
Then $\alpha, \beta$ are similar if these distinguished
representatives coincide.

This is what we are eventually going to squeeze out of
\Cref{thm:linalg:structure-theorem-fg-modules-over-pid}, in the friendly case of vector spaces over a fixed
field.

\subsection{The characteristic and minimal polynomials of an
  endomorphism}
\label{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}
Let $\alpha \in \End_R(F)$ be an endomorphism of a free $R$-module;
henceforth I am often tacitly going to assume that $F$ is finitely
generated (this is necessary anyway as we are aiming to translate what
we do into the language of matrices). We want to identify
\emph{invariants} of the similarity class of $\alpha$, that is,
quantities that will not change if we replace $\alpha$ with a similar
linear map $\beta \in
\End_R(F)$.

For example, the determinant is such a quantity:
\begin{defn}{}{defn:linalg:determinant-of-transformation}
  Let $\alpha \in \End_R(F)$. The \emph{determinant} of $\alpha$ is
  \[
    \det(\alpha) := \det(A),
  \]
  where $A$ is the matrix representing $\alpha$ with respect to any
  choice of basis of $F$.
\end{defn}
Of course there is something to check here, that is, that $\det(A)$
does not depend on the choice of basis. But we know
(\Cref{prop:linalg:similarity-conjugation-characterization}) that $A, B$ represent the same endomorphism
$\alpha$ if and only if there exists an invertible matrix $P$ such
that $B = PAP^{-1}$; then
\[
  \det(B) = \det(PAP^{-1}) = \det(P)\det(A)\det(P^{-1}) = \det(A)
\]
by \Cref{prop:linalg:determinant-invertible-iff-unit-homomorphism}. Thus the determinant is indeed independent of
the choice of basis. Essentially the same argument shows that if
$\alpha$ and $\beta$ are similar linear transformations, then
$\det(\alpha) = \det(\beta)$
(\Cref{exer:linalg:similar-matrices-same-det-and-trace}).

By \Cref{prop:linalg:determinant-invertible-iff-unit-homomorphism}, a linear transformation $\alpha$ is invertible
if and only if $\det(\alpha)$ is a unit in $R$. If $R$ is a field,
this of course means simply $\det(\alpha) \neq 0$. Even if $R$ is not
a field, $\det(\alpha) \neq 0$ says something interesting:
\begin{prop}{}{prop:linalg:determinant-nonzero-iff-injective}
  Let $\alpha$ be a linear transformation of a free $R$-module
  $F \cong R^n$. Then $\det(\alpha) \neq 0$ if and only if $\alpha$ is
  injective.
\end{prop}
\begin{proof}
  Embed $R$ in its field of fractions $K$, and view $\alpha$ as a
  linear transformation of $K^n$; note that the determinant of
  $\alpha$ is the same whether it is computed over $R$ or over
  $K$. Then $\alpha$ is injective as a linear transformation
  $R^n \to R^n$ if and only if it is injective as a linear
  transformation $K^n \to K^n$, if and only if it is invertible as a
  linear transformation $K^n \to K^n$, if and only if
  $\det(\alpha) \neq 0$.
\end{proof}

Of course over integral domains other than fields $\alpha$ may fail to
be surjective even if $\det(\alpha) \neq 0$; and care is required even
over fields, if we abandon the hypothesis that the free modules are
finitely generated (cf.\
\Cref{exer:linalg:injective-not-surjective-iff-det-not-unit,exer:linalg:injective-not-surjective-example-on-kt}).

Another quantity that is invariant under similarity is the
\emph{trace}. The trace of a square matrix
$A = (a_{ij}) \in \mathcal{M}_n(R)$ is
\[
  \operatorname{tr}(A) := \sum_{i=1}^n a_{ii},
\]
that is, the sum of its diagonal entries.
\begin{defn}{}{defn:linalg:trace-of-transformation}
  Let $\alpha \in \End_R(F)$. The \emph{trace} of $\alpha$ is defined
  to be $\operatorname{tr}(\alpha) := \operatorname{tr}(A)$, where $A$
  is the matrix representing $\alpha$ with respect to any choice of
  basis of $F$.
\end{defn}
Again, we have to check that this is independent of the choice of
basis; the key is the following computation.
\begin{lem}{}{lem:linalg:trace-of-product-is-commutative}
  Let $A, B \in \mathcal{M}_n(R)$. Then
  $\operatorname{tr}(AB) = \operatorname{tr}(BA)$.
\end{lem}
\begin{proof}
  Let $A = (a_{ij})$, $B = (b_{ij})$. Then
  $AB = (\sum_{k=1}^n a_{ik}b_{kj})$; hence
  \[
    \operatorname{tr}(AB) = \sum_{i=1}^n \sum_{k=1}^n a_{ik}b_{ki}.
  \]
  This expression is symmetric in $A, B$, so it must equal
  $\operatorname{tr}(BA)$.
\end{proof}

With this understood, if $B = PAP^{-1}$, then
\[
  \operatorname{tr}(B) = \operatorname{tr}((PA)P^{-1}) =
  \operatorname{tr}(P^{-1}(PA)) = \operatorname{tr}((P^{-1}P)A) =
  \operatorname{tr}(A),
\]
showing (by \Cref{prop:linalg:similarity-conjugation-characterization}) that similar matrices have the same
trace, as needed.

Again, the reader will check that the same token shows that similar
linear transformations have the same trace.

The trace and determinant of an endomorphism $\alpha$ of a rank-$n$
free module are in fact just two of a sequence of $n$ invariants,
which can be nicely collected together into the \emph{characteristic
  polynomial} of $\alpha$.
\begin{defn}{}{defn:linalg:characteristic-polynomial}
  Let $F$ be a free $R$-module, and let $\alpha \in \End_R(F)$. Denote
  by $I$ the identity map $F \to F$. The \emph{characteristic
    polynomial} of $\alpha$ is the polynomial
  \[
    P_\alpha(t) := \det(tI - \alpha) \in R[t].
  \]
\end{defn}
\begin{prop}{}{prop:linalg:characteristic-polynomial-coefficients}
  Let $F$ be a free $R$-module of rank $n$, and let
  $\alpha \in \End_R(F)$.
  \begin{itemize}
  \item The characteristic polynomial $P_\alpha(t)$ is a monic
    polynomial of degree $n$.
  \item The coefficient of $t^{n-1}$ in $P_\alpha(t)$ equals
    $-\operatorname{tr}(\alpha)$.
  \item The constant term of $P_\alpha(t)$ equals
    $(-1)^n \det(\alpha)$.
  \item If $\alpha$ and $\beta$ are similar, then
    $P_\alpha(t) = P_\beta(t)$.
  \end{itemize}
\end{prop}
\begin{proof}
  The first point is immediate, and the third is checked by setting
  $t = 0$. To verify the second assertion, let $A = (a_{ij})$ be a
  matrix representing $\alpha$ with respect to any basis for $F$, so
  that
  \[
    P_\alpha(t) = \det
    \begin{pmatrix}
      t-a_{11} & -a_{12} & \dots & -a_{1n} \\
      -a_{21} & t-a_{22} & \dots & -a_{2n} \\
      \vdots & \vdots & \ddots & \vdots \\
      -a_{n1} & -a_{n2} & \dots & t-a_{nn}
    \end{pmatrix}.
  \]
  Expanding the determinant according to \Cref{defn:linalg:determinant} (or in any
  other way), we see that the only contributions to the coefficient of
  $t^{n-1}$ come from the diagonal entries, in the form
  \[
    \sum_{i=1}^n t \dots t \cdot (-a_{ii}) \cdot t \dots t,
  \]
  and the statement follows.

  Finally, assume that $\alpha$ and $\beta$ are similar. Then there
  exists an invertible $\pi$ such that
  $\beta = \pi \circ \alpha \circ \pi^{-1}$, and hence
  \[
    tI - \beta = \pi \circ (tI - \alpha) \circ \pi^{-1}
  \]
  are similar (as endomorphisms of $R[t]^n$). By
  \Cref{exer:linalg:similar-matrices-same-det-and-trace} these two
  transformations must have the same determinant, and this proves the
  fourth point.
\end{proof}

By \Cref{prop:linalg:characteristic-polynomial-coefficients}, all coefficients in the characteristic
polynomial
\[
  t^n - \operatorname{tr}(\alpha)t^{n-1} + \dots + (-1)^n \det(\alpha)
\]
are invariant under similarity; as far as I know, trace and
determinant are the only ones that have special names.

Determinants, traces, and more generally the characteristic polynomial
can show very quickly that two linear transformations are \emph{not}
similar; but do they tell us unfailingly when two transformations
\emph{are} similar? In general, the answer is no, even over
fields. For example, the two matrices
\[
  I = \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}, \quad A
  = \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}
\]
both have characteristic polynomial $(t-1)^2$, but they are not
similar. Indeed, $I$ is the identity and clearly only the identity is
similar to the identity: $PIP^{-1} = I$ for all invertible matrices
$P$.

Therefore, the equivalence relation defined by prescribing that two
transformations have the same characteristic polynomial is coarser
than similarity. This hints that there must be other interesting
quantities associated with a linear transformation and invariant under
similarity.

We are going to understand this much more thoroughly in a short while
(at least over fields), but we can already gain some insight by
contemplating another kind of `polynomial' information, which also
turns out to be invariant under similarity. As $\End_R(F)$ is an
$R$-algebra, we can evaluate every polynomial
\[
  f(t) = r_m t^m + r_{m-1} t^{m-1} + \dots + r_0 \in R[t]
\]
at any $\alpha \in \End_R(F)$:
\[
  f(\alpha) = r_m \alpha^m + r_{m-1} \alpha^{m-1} + \dots + r_0 \in
  \End_R(F).
\]
In other words, we can perform these operations in the ring
$\End_R(F)$; multiplication by $r \in R$ amounts to composition with
$rI \in \End_R(F)$, and $\alpha^k$ stands for the $k$-fold composition
$\alpha \circ \dots \circ \alpha$ of $\alpha$ with itself. The set of
polynomials such that $f(\alpha) = 0$ is an ideal of $R[t]$
(\Cref{exer:linalg:annihilator-ideal-of-linear-transformation}), which
I will denote $\mathscr{I}_\alpha$ and call the \emph{annihilator
  ideal} of $\alpha$.
\begin{lem}{}{lem:linalg:similar-transformations-same-annihilator-ideal}
  If $\alpha$ and $\beta$ are similar, then
  $\mathscr{I}_\alpha = \mathscr{I}_\beta$.
\end{lem}
\begin{proof}
  By hypothesis there exists an invertible $\pi$ such that
  $\beta = \pi \circ \alpha \circ \pi^{-1}$. As
  \[
    \beta^k = (\pi \circ \alpha \circ \pi^{-1})^k = (\pi \circ \alpha
    \circ \pi^{-1}) \circ (\pi \circ \alpha \circ \pi^{-1}) \circ
    \dots \circ (\pi \circ \alpha \circ \pi^{-1}) = \pi \circ \alpha^k
    \circ \pi^{-1},
  \]
  we see that for all $f(t) \in R[t]$ we have
  \[
    f(\beta) = \pi \circ f(\alpha) \circ \pi^{-1}.
  \]
  It follows immediately that $f(\alpha) = 0 \iff f(\beta) = 0$, which
  is the statement.
\end{proof}

Going back to the simple example shown above, the polynomial $t - 1$
is in the annihilator ideal of the identity, while it is not in the
annihilator ideal of the matrix
\[
  A = \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}.
\]
An optimistic reader might now guess that two linear transformations
are similar if and only if both their characteristic polynomials and
annihilator ideals coincide. This is unfortunatley not the case in
general, but I promise that the situation will be considerably
clarified in short order (cf.\
\Cref{exer:linalg:similar-iff-same-char-min-poly-low-dim}).

In any case, even the simple example given above allows me to point
out a remarkable fact. Note that the (common) characteristic
polynomial $(t-1)^2$ of $I$ and $A$ annihilates both:
\[
  (A - I)^2 = \begin{pmatrix} 0 & 1 \\ 0 & 0 \end{pmatrix}^2 =
  \begin{pmatrix} 0 & 0 \\ 0 & 0 \end{pmatrix}.
\]
This is not a coincidence: $P_\alpha(t) \in \mathscr{I}_\alpha$ for
all linear transformations $\alpha$. That is,
\begin{thm}{Cayley-Hamilton}{thm:linalg:cayley-hamilton-theorem}
  Let $P_\alpha(t)$ be the characteristic polynomial of the linear
  transformation $\alpha \in \End_R(F)$. Then
  \[
    P_\alpha(\alpha) = 0.
  \]
\end{thm}
This beautiful observation can be proved directly by judicious use of
Cramer's rule\footnote{Here is an even more direct `proof':
  $P_\alpha(\alpha) = \det(\alpha I - \alpha) = \det(0) = 0$.
  Unfortunately this does not work: in the definition
  $\det(tI - \alpha)$ of the characteristic polynomial, $t$ is assumed
  to act as a scalar and cannot be replaced by $\alpha$. Applying
  Cramer's rule circumvents this obstacle.}, in the form of
\Cref{coro:linalg:adjugate-matrix-formula}; cf.\ \Cref{exer:linalg:cayley-hamilton-theorem}. In
any case, the Cayley-Hamilton theorem will become essentially evident
once we connect these linear algebra considerations with the
classification theorem for finitely generated modules over a PID; the
adventurous reader can already look back at \Cref{rem:linalg:torsion-module-invariant-factors-summary} and
figure out why the Cayley-Hamilton theorem is obvious.

If $R$ is an arbitrary integral domain, we cannot expect too much of
$R[t]$, and it seems hard to say something \emph{a priori} concerning
$\mathscr{I}_\alpha$. However, consider the field of fractions $K$ of
$R$
(\S{}\ref{subsec:irred:the-field-of-fractions-of-an-integral-domain});
viewing $\alpha$ as an element of $\End_K(K^n)$ (that is, viewing the
entries of a matrix representation of $\alpha$ as elements of $K$,
rather than $R$), $\alpha$ will have an annihilator ideal
$\mathscr{I}_\alpha^{(K)}$ `over $K$', and it is clear that
\[
  \mathscr{I}_\alpha = \mathscr{I}_\alpha^{(K)} \cap R[t].
\]
The advantage of considering $\mathscr{I}_\alpha^{(K)} \subseteq K[t]$
is that $K[t]$ is a PID, and it follows that
$\mathscr{I}_\alpha^{(K)}$ has a (unique) monic generator.
\begin{defn}{}{defn:linalg:minimal-polynomial}
  Let $F$ be a free $R$-module, and let $\alpha \in \End_R(F)$. Let
  $K$ be the field of fractions of $R$. The \emph{minimal polynomial}
  of $\alpha$ is the monic generator $m_\alpha(t) \in K[t]$ of
  $\mathscr{I}_\alpha^{(K)}$.
\end{defn}
With this terminology, the Cayley-Hamilton theorem amounts to the
assertion that the minimal polynomial divides the characteristic
polynomial: $m_\alpha(t) \mid P_\alpha(t)$.

Of course the situation is simplified, at least from an expository
point of view, if $R$ is itself a field: then $K = R$,
$m_\alpha(t) \in R[t]$, and $\mathscr{I}_\alpha = (m_\alpha(t))$. This
is one reason why we will eventually assume that $R$ is a field.

\subsection{Eigenvalues, eigenvectors, eigenspaces}
\label{subsec:linalg:eigenvalues-eigenvectors-eigenspaces}

\begin{defn}{}{defn:linalg:eigenvalue}
  Let $F$ be a free $R$-module, and let $\alpha \in \End_R(F)$ be a
  linear transformation of $F$.  A scalar $\lambda \in R$ is an
  \emph{eigenvalue} for $\alpha$ if there exists $\mathbf{v} \in F$,
  $\mathbf{v} \neq 0$, such that
  \[
    \alpha(\mathbf{v}) = \lambda\mathbf{v}.
  \]
\end{defn}
For example, $0$ is an eigenvalue for $\alpha$ precisely when $\alpha$
has a nontrivial kernel. The notion of eigenvalue is one of the most
important in linear algebra, if not in algebra, if not in mathematics,
if not in the whole of science. The set of eigenvalues of a linear
transformation is called its \emph{spectrum}. Spectra of operators
show up everywhere, from number theory to differential equations to
quantum mechanics. The spectrum of a ring, encountered briefly in
\S{}\ref{subsec:rings:prime-and-maximal-ideals}, was so named because
it may be interpreted (in important motivating contexts) as a spectrum
in the sense of \Cref{defn:linalg:eigenvalue}. The `spectrum of the hydrogen
atom' is also a spectrum in this sense.

It is hopefully immediate that similar transformations have the same
spectrum: for if $\beta = \pi \circ \alpha \circ \pi^{-1}$ and
$\alpha(\mathbf{v}) = \lambda\mathbf{v}$, then
\[
  \beta(\pi(\mathbf{v})) = \pi \circ \alpha \circ
  \pi^{-1}(\pi(\mathbf{v})) = \pi \circ (\alpha(\mathbf{v})) =
  \pi(\lambda\mathbf{v}) = \lambda\pi(\mathbf{v});
\]
as $\pi(\mathbf{v}) \neq 0$ if $\mathbf{v} \neq 0$, this shows that
every eigenvalue of $\alpha$ is an eigenvalue of $\beta$.

If $F$ is finitely generated, then we have the following useful
translation of the notion of eigenvalue:
\begin{lem}{}{lem:linalg:eigenvalues-are-roots-of-characteristic-poly}
  Let $F$ be a finitely generated $R$-module, and let
  $\alpha \in \End_R(F)$. Then the set of eigenvalues of $\alpha$ is
  precisely the set of roots in $R$ of the characteristic polynomial
  $P_\alpha(t)$.
\end{lem}
\begin{proof}
  This is a straightforward consequence of \Cref{prop:linalg:determinant-nonzero-iff-injective}:
  \[
    \begin{aligned}
      \lambda \text{ is an eigenvalue for } \alpha
      &\iff \exists \mathbf{v} \neq 0 \text{ such that }
        \alpha(\mathbf{v}) = \lambda I(\mathbf{v}) \\
      &\iff \exists \mathbf{v} \neq 0 \text{ such that }
        (\lambda I - \alpha)(\mathbf{v}) = 0 \\
      &\iff \lambda I - \alpha \text{ is not injective} \\
      &\iff \det(\lambda I - \alpha) = 0 \\
      &\iff P_\alpha(\lambda) = 0
    \end{aligned}
  \]
  as claimed.
\end{proof}

It is an evident consequence of \Cref{lem:linalg:eigenvalues-are-roots-of-characteristic-poly} that eigenvalue
considerations depend very much on the base ring $R$.
\begin{exam}{}{exam:linalg:rotation-matrix-no-real-eigenvalues}
  The matrix
  \[
    \begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}
  \]
  has no eigenvalues over $\R$, while it has eigenvalues over $\C$:
  indeed, the characteristic polynomial $t^2 + 1$ has no real roots
  and two complex roots. The reader should observe that, as a linear
  transformation of the real plane $\R^2$, this matrix corresponds to
  a $90^\circ$ counterclockwise rotation; the reason why this
  transformation has no (real) eigenvalues is that no direction in the
  plane is preserved through a $90^\circ$ rotation.
\end{exam}
\begin{exam}{}{exam:linalg:eigenvalues-over-r-not-over-q}
  An example with a different flavor is given by the matrix
  \[
    \begin{pmatrix} 0 & 2 \\ 1 & 0 \end{pmatrix}.
  \]
  This has characteristic polynomial $t^2 - 2$; hence it has
  eigenvalues over $\R$, but not over $\Q$. Geometrically, the
  corresponding transformation flips the plane about a line; but that
  line has irrational slope, so it contains no nonzero vectors with
  rational components.
\end{exam}
As another benefit of \Cref{lem:linalg:eigenvalues-are-roots-of-characteristic-poly}, we may now introduce the
following notion:
\begin{defn}{}{defn:linalg:algebraic-multiplicity}
  The \emph{algebraic multiplicity} of an eigenvalue of a linear
  transformation $\alpha$ of a finitely generated free module is its
  multiplicity as a root of the characteristic polynomial of $\alpha$.
\end{defn}
For example, the identity on $F$ has the single eigenvalue $1$, with
(algebraic) multiplicity equal to the rank of $F$.

The sum of the algebraic multiplicities is bounded by the degree of
the characteristic polynomial, that is, the dimension of the space. In
particular,
\begin{coro}{}{coro:linalg:eigenvalue-count-bound}
  The number of eigenvalues of a linear transformation of $R^n$ is at
  most $n$. If the base ring $R$ is an algebraically closed field,
  then every linear transformation has exactly $n$ eigenvalues
  (counted with algebraic multiplicity).
\end{coro}
\begin{proof}
  Immediate from \Cref{lem:linalg:eigenvalues-are-roots-of-characteristic-poly,lem:irred:polynomial-root-count-bound,lem:irred:algebraically-closed-tfae}.
\end{proof}

There is a different notion of multiplicity of an eigenvalue, related
to how big the corresponding\footnote{If we were serious about
  pursuing the theory over arbitrary integral domains, I would feel
  compelled to call this object the \emph{eigenmodule} of $\alpha$.
  However, we are eventually going to restrict our attention to the
  case in which the base ring is a field, so I will not bother.}
eigenspace may be.
\begin{defn}{}{defn:linalg:eigenvector-and-eigenspace}
  Let $\lambda$ be an eigenvalue of a linear transformation $\alpha$
  of a free $R$-module $F$. Then a nonzero $\mathbf{v} \in F$ is an
  \emph{eigenvector} for $\alpha$, corresponding to the eigenvalue
  $\lambda$, if $\alpha(\mathbf{v}) = \lambda\mathbf{v}$, that is, if
  $\mathbf{v} \in \ker(\lambda I - \alpha)$. The submodule
  $\ker(\lambda I - \alpha)$ is the \emph{eigenspace} corresponding to
  $\lambda$.
\end{defn}
\begin{defn}{}{defn:linalg:geometric-multiplicity}
  The \emph{geometric multiplicity} of an eigenvalue $\lambda$ is the
  rank of its eigenspace.
\end{defn}
This is clearly invariant under similarity
(\Cref{exer:linalg:geometric-multiplicity-invariant-under-similarity}). Geometric
and algebraic multiplicities do not necessarily coincide: for example,
the matrix
\[
  \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}
\]
has characteristic polynomial $(t-1)^2$, and hence the single
eigenvalue $1$, with algebraic multiplicity $2$. However, for a vector
$\mathbf{v} = \begin{pmatrix} v_1 \\ v_2 \end{pmatrix}$,
\[
  \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}
  \begin{pmatrix} v_1 \\ v_2 \end{pmatrix} =
  \begin{pmatrix} v_1 + v_2 \\ v_2 \end{pmatrix}
\]
equals $1\mathbf{v}$ if and only if $v_2 = 0$: that is, the eigenspace
corresponding to the single eigenvalue $1$ consists of the span of the
vector $\begin{pmatrix} 1 \\ 0 \end{pmatrix}$ and has dimension $1$.
Thus, the geometric multiplicity of the eigenvalue is $1$ in this
example.

Contemplating this example will convince the reader that the geometric
multiplicity of an eigenvalue is always \emph{less than} its algebraic
multiplicity. In the neatest possible situation, an operator $\alpha$
on a free module $F$ of rank $n$ may have all its $n$ eigenvalues in
the base ring $R$, and each algebraic multiplicity may agree with the
corresponding geometric multiplicity. If this is the case, $F$ may
then be expressed as a direct sum of the eigenspaces, producing the
so-called \emph{spectral decomposition} of $F$ determined by $\alpha$
(cf.\ \Cref{exer:linalg:eigenvectors-distinct-eigenvalues-lin-indep}
for a concrete instance of this situation). The action of $\alpha$ on
$F$ is then completely transparent, as it amounts to simply applying a
(possibly) different scaling factor on each piece of the spectral
decomposition.

If $A$ is a matrix representing $\alpha \in \End_R(V)$ with respect to
any basis, then $\alpha$ admits a spectral decomposition if and only
if $A$ is similar to a diagonal matrix:
\[
  A = P
  \begin{pmatrix}
    \lambda_1 & 0 & \dots & 0 \\
    0 & \lambda_2 & \dots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \dots & \lambda_n
  \end{pmatrix}
  P^{-1},
\]
where $\lambda_1, \dots, \lambda_n$ are the eigenvalues of
$\alpha$. In this case we say that $A$ (or $\alpha$) is
\emph{diagonalizable}. A moment's thought reveals that the columns of
$P$ are then a basis of $V$ consisting of eigenvectors of $\alpha$.

Once more, I promise that the situation will become (even) clearer
once we bring \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} into the picture.

\subsection*{Exercises}

In these exercises, $R$ denotes an integral domain.
\begin{exer}{}{exer:linalg:finite-equiv-classes-infinite-similarity-classes}
  Let $k$ be an infinite field, and let $n$ be any positive integer.
  \begin{itemize}
  \item Prove that there are finitely many equivalence classes of
    matrices in $\mathcal{M}_n(k)$.
  \item Prove that there are infinitely many similarity classes of
    matrices in $\mathcal{M}_n(k)$.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:linalg:det-multiplicative-invertible-iff-det-invertible}
  Let $F$ be a free $R$-module of rank $n$, and let
  $\alpha, \beta \in \End_R(F)$. Prove that
  $\det(\alpha \circ \beta) = \det(\alpha)\det(\beta)$. Prove that
  $\alpha$ is invertible if and only if $\det(\alpha)$ is invertible.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:similar-matrices-same-det-and-trace}
  Prove that if $\alpha$ and $\beta$ are similar in the sense of
  \Cref{defn:linalg:similar-transformations}, then $\det(\alpha) = \det(\beta)$ and
  $\operatorname{tr}(\alpha) = \operatorname{tr}(\beta)$.  (Do this
  without using \Cref{prop:linalg:characteristic-polynomial-coefficients}!)
  [\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:injective-not-surjective-iff-det-not-unit}
  Let $F$ be a finitely generated free $R$-module, and let $\alpha$ be
  a linear transformation of $F$. Give an example of an injective
  $\alpha$ which is not surjective; in fact, prove that $\alpha$ is
  not surjective precisely when $\det(\alpha)$ is not a
  unit. [\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:injective-not-surjective-example-on-kt}
  Let $k$ be a field, and view $k[t]$ as a vector space over $k$ in
  the evident way. Give an example of a $k$-linear transformation
  $k[t] \to k[t]$ which is injective but not surjective; give an
  example of a linear transformation which is surjective but not
  injective.
  [\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism},
  \S{}\ref{subsec:fields:splitting-fields-and-normal-extensions}]
\end{exer}
\begin{exer}{}{exer:linalg:same-char-poly-2x2-iff-trace-det}
  Prove that two $2 \times 2$ matrices have the same characteristic
  polynomial if and only if they have the same trace and determinant.
  Find two $3 \times 3$ matrices with the same trace and determinant
  but different characteristic polynomials.
\end{exer}
\begin{exer}{\exerused}{exer:linalg:annihilator-ideal-of-linear-transformation}
  Let $\alpha \in \End_R(F)$ be a linear transformation on a free
  $R$-module $F$. Prove that the set of polynomials $f(t) \in R[t]$
  such that $f(\alpha) = 0$ is an ideal of $R[t]$.
  [\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:matrix-transpose-same-char-poly-annihilator}
  Let $A \in \mathcal{M}_n(R)$ be a square matrix, and let $A^t$ be
  its transpose. Prove that $A$ and $A^t$ have the same characteristic
  polynomial and the same annihilator
  ideals. [\S{}\ref{subsec:linalg:k-t-modules-and-the-rational-canonical-form}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:cayley-hamilton-theorem}
  Prove the Cayley-Hamilton theorem, as follows. Recall that every
  square matrix $M$ has an adjoint matrix, which I will denote
  $\operatorname{adj}(M)$, and that we proved (\Cref{coro:linalg:adjugate-matrix-formula}) that
  $\operatorname{adj}(M) \cdot M = \det(M) \cdot I$. Applying this to
  $M = tI - A$ (with $A$ a matrix realization of
  $\alpha \in \End_R(F)$) gives
  \[
    \tag{*} \operatorname{adj}(tI - A) \cdot (tI - A) = P_\alpha(t)
    \cdot I.
  \]
  Prove that there exist matrices $B_k \in \mathcal{M}_n(R)$ such that
  $\operatorname{adj}(tI - A) = \sum_{k=0}^{n-1} B_k t^k$; then use
  $(*)$ to obtain $P_\alpha(A) = 0$, proving the Cayley-Hamilton
  theorem.
  [\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:char-poly-multiplicative-direct-sums}
  Let $F_1, F_2$ be free $R$-modules of finite rank, and let
  $\alpha_1$, resp., $\alpha_2$, be linear transformations of $F_1$,
  resp., $F_2$. Let $F = F_1 \oplus F_2$, and let
  $\alpha = \alpha_1 \oplus \alpha_2$ be the linear transformation of
  $F$ restricting to $\alpha_1$ on $F_1$ and $\alpha_2$ on $F_2$.
  \begin{itemize}
  \item Prove that $P_\alpha(t) =
    P_{\alpha_1}(t)P_{\alpha_2}(t)$. That is, the characteristic
    polynomial is multiplicative under direct sums.
  \item Find an example showing that the minimal polynomial is
    \emph{not} multiplicative under direct sums.
  \end{itemize}
  [\ref{exer:linalg:char-poly-multiplicative-invariant-subspace}, \S{}
  ef{subsec:linalg:k-t-modules-and-the-rational-canonical-form}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:char-poly-multiplicative-invariant-subspace}
  Let $\alpha$ be a linear transformation of a finite-dimensional
  vector space $V$, and let $V_1$ be an invariant subspace, that is,
  such that $\alpha(V_1) \subseteq V_1$. Let $\alpha_1$ be the
  restriction of $\alpha$ to $V_1$, and let $V_2 = V/V_1$. Prove that
  $\alpha$ induces a linear transformation $\alpha_2$ on $V_2$, and
  (in the same vein as
  \Cref{exer:linalg:char-poly-multiplicative-direct-sums}) show that
  $P_\alpha(t) = P_{\alpha_1}(t)P_{\alpha_2}(t)$.  Also, prove that
  $\operatorname{tr}(\alpha^r) = \operatorname{tr}(\alpha_1^r) +
  \operatorname{tr}(\alpha_2^r)$, for all $r \geq
  0$. [\ref{exer:linalg:zeta-function-identity-linear-transformation}]
\end{exer}
\begin{exer}{}{exer:linalg:zeta-function-identity-linear-transformation}
  Let $\alpha$ be a linear transformation of a finite-dimensional
  $\C$-vector space $V$. Prove the identity of formal power series
  with coefficients in $\C$:
  \[
    \frac{1}{\det(1-\alpha t)} = \exp \left( \sum_{r=1}^\infty
      \operatorname{tr}(\alpha^r) \frac{t^r}{r} \right).
  \]
  (Hint: The left-hand side is essentially the inverse of the
  characteristic polynomial of $\alpha$. Use
  \Cref{exer:linalg:char-poly-multiplicative-invariant-subspace} to
  show that both sides are multiplicative with respect to exact
  sequences $0 \to V_1 \to V \to V_2 \to 0$, where $V_1$ is an
  invariant subspace. Use this and the fact that $\alpha$ admits
  nontrivial invariant subspaces since $\C$ is algebraically closed
  (why?) to reduce to the case $\dim V = 1$, for which the identity is
  an elementary calculus exercise.)

  With due care, the identity can be stated and proved (in the same
  way) over arbitrary fields of characteristic $0$. It is an
  ingredient in the cohomological interpretation of the `Weil
  conjectures'.
\end{exer}
\begin{exer}{}{exer:linalg:rational-eigenvalue-of-integer-matrix-is-integer}
  Let $A$ be a square matrix with integer entries. Prove that if
  $\lambda$ is a rational eigenvalue of $A$, then in fact
  $\lambda \in \Z$. (Hint: \Cref{prop:irred:rational-root-theorem}.)
\end{exer}
\begin{exer}{\exerused}{exer:linalg:geometric-multiplicity-invariant-under-similarity}
  Let $\lambda$ be an eigenvalue of two similar transformations
  $\alpha, \beta$. Prove that the geometric multiplicities of
  $\lambda$ with respect to $\alpha$ and $\beta$
  coincide. [\S{}\ref{subsec:linalg:eigenvalues-eigenvectors-eigenspaces}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:eigenvectors-distinct-eigenvalues-lin-indep}
  Let $\alpha$ be a linear transformation on a free $R$-module $F$,
  and let $\mathbf{v}_1, \dots, \mathbf{v}_n$ be eigenvectors
  corresponding to pairwise distinct eigenvalues
  $\lambda_1, \dots, \lambda_n$. Prove that
  $\mathbf{v}_1, \dots, \mathbf{v}_n$ are linearly independent.
  (Hint: If not, there is a shortest linear combination
  $r_1\mathbf{v}_{i_1} + \dots + r_m\mathbf{v}_{i_m} = 0$ with all
  $r_j \in R$, $r_j \neq 0$. Compare the action of $\alpha$ on this
  linear combination with the product by $\lambda_{i_1}$.)

  It follows that if $F$ has rank $n$ and $\alpha$ has $n$ distinct
  eigenvalues, then $\alpha$ induces a spectral decomposition of $F$.
  [\S{}\ref{subsec:linalg:eigenvalues-eigenvectors-eigenspaces},
  \ref{exer:fields:distinct-automorphisms-linearly-independent}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:orthogonal-unitary-preserve-inner-product}
  The \emph{standard inner product} on $V = \R^n$ is the map
  $V \times V \to \R$ defined by
  \[
    (\mathbf{v}, \mathbf{w}) := \mathbf{v}^t \cdot \mathbf{w}
  \]
  (viewing elements $\mathbf{v} \in V$ as column vectors). The
  \emph{standard hermitian product} on $W = \C^n$ is the map
  $W \times W \to \C$ defined by
  \[
    (\mathbf{v}, \mathbf{w}) := \mathbf{v}^\dagger \cdot \mathbf{w},
  \]
  where for any matrix $M$, $M^\dagger$ stands for the matrix obtained
  by taking the complex conjugates of the entries of the transpose
  $M^t$.

  These products satisfy evident linearity properties: for example,
  for $\lambda \in \C$ and $\mathbf{v}, \mathbf{w} \in W$
  \[
    (\lambda\mathbf{v}, \mathbf{w}) = \lambda(\mathbf{v}, \mathbf{w}).
  \]
  Prove\footnote{See \Cref{exer:groups:matrix-subgroups-inclusions}
    for the definitions of $\mathrm{O}_n(\R)$ and $\mathrm{U}(n)$. I
    trust that basic facts on inner products are not new to the
    reader. Among these, recall that for
    $\mathbf{v}, \mathbf{w} \in \R^n$, $(\mathbf{v}, \mathbf{v})$
    equals the square of the length of $\mathbf{v}$, and
    $(\mathbf{v}, \mathbf{w})$ equals the product of the lengths of
    $\mathbf{v}$ and $\mathbf{w}$ and the cosine of the angle formed
    by $\mathbf{v}$ and $\mathbf{w}$. Thus, the reader is essentially
    asked to verify that $M \in \mathrm{O}_n(\R)$ if and only if $M$
    preserves lengths and angles, and to check an analogous statement
    in the complex environment.} that a matrix
  $M \in \mathcal{M}_n(\R)$ belongs to $\mathrm{O}_n(\R)$ if and only
  if it preserves the standard inner product on $\R^n$:
  \[
    (\forall \mathbf{v}, \mathbf{w} \in \R^n) \quad (M\mathbf{v},
    M\mathbf{w}) = (\mathbf{v}, \mathbf{w}).
  \]
  Likewise, prove that a matrix $M \in \mathcal{M}_n(\C)$ belongs to
  $\mathrm{U}(n)$ if and only if it preserves the standard hermitian
  product on
  $\C^n$. [\ref{exer:linalg:on-r-iff-orthonormal-columns-rows}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:orthogonal-complement-is-subspace-dim-n-minus-1}
  We say that two vectors $\mathbf{v}, \mathbf{w}$ of $\R^n$ or $\C^n$
  are \emph{orthogonal} if $(\mathbf{v}, \mathbf{w}) = 0$.  The
  \emph{orthogonal complement} $\mathbf{v}^\perp$ of $\mathbf{v}$ is
  the set of vectors $\mathbf{w}$ that are orthogonal to $\mathbf{v}$.
  Prove that if $\mathbf{v} \neq 0$ in $V = \R^n$ or $\C^n$, then
  $\mathbf{v}^\perp$ is a subspace of $V$ of dimension
  $n-1$. [\ref{exer:linalg:schur-form},
  \ref{exer:linrep:annihilator-of-subspace-basics}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:on-r-iff-orthonormal-columns-rows}
  Let $V = \R^n$, endowed with the standard inner product defined in
  \Cref{exer:linalg:orthogonal-unitary-preserve-inner-product}. A set
  of distinct vectors $\mathbf{v}_1, \dots, \mathbf{v}_r$ is
  \emph{orthonormal} if $(\mathbf{v}_i, \mathbf{v}_j) = 0$ for
  $i \neq j$ and $1$ for $i = j$.  Geometrically, this means that each
  vector has length $1$, and different vectors are orthogonal. The
  same terminology may be used in $\C^n$, w.r.t.\ the standard
  hermitian product.

  Prove that $M \in \mathrm{O}_n(\R)$ if and only if the columns of
  $M$ are orthonormal, if and only if the rows of $M$ are
  orthonormal\footnote{The group $\mathrm{O}_n(\R)$ is called the
    \emph{orthogonal group}; orthonormal group would probably be a
    more appropriate terminology.}.

  Formulate and prove an analogous statement for $\mathrm{U}(n)$. (The
  group $\mathrm{U}(n)$ is called the \emph{unitary group}. Note that,
  for $n = 1$, it consists of the complex numbers of norm $1$.)
  [\ref{exer:linalg:orthonormal-vectors-coefficient-projection-formulas},
  \ref{exer:linalg:schur-form},
  \ref{exer:linrep:nondegenerate-nonsingular-bilinear-forms}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:orthonormal-vectors-coefficient-projection-formulas}
  Let $\mathbf{v}_1, \dots, \mathbf{v}_r$ form an orthonormal set of
  vectors in $\R^n$.
  \begin{itemize}
  \item Prove that $\mathbf{v}_1, \dots, \mathbf{v}_r$ are linearly
    independent; so they form an orthonormal basis of the space $V$
    they span.
  \item Let $\mathbf{w} = a_1\mathbf{v}_1 + \dots + a_r\mathbf{v}_r$
    be a vector of $V$. Prove that $a_i = (\mathbf{v}_i, \mathbf{w})$.
  \item More generally, prove that if $\mathbf{w} \in \R^n$, then
    $(\mathbf{v}_i, \mathbf{w})$ is the component of $\mathbf{v}_i$ in
    the orthogonal projection $\mathbf{w}_V$ of $\mathbf{w}$ onto $V$.
    (That is, prove that $\mathbf{w} - \mathbf{w}_V$ is orthogonal to
    all vectors of $V$.)
  \end{itemize}
  \begin{center}
    \begin{tikzpicture}
      \draw[fill=gray!10] (-1.5,-0.5) -- (2.5,-0.5) -- (3.2,0.5) --
      (-0.8,0.5) -- cycle; \node at (2.9,0) {$V$}; \draw[->, thick]
      (0,0) -- (2,2) node[above] {$\mathbf{w}$}; \draw[->, thick]
      (0,0) -- (1,0) node[below] {$\mathbf{w}_V$}; \draw[dashed] (2,2)
      -- (1,0);
    \end{tikzpicture}
  \end{center}
  For reasons such as these, it is convenient to work with orthonormal
  bases. Note that, by
  \Cref{exer:linalg:on-r-iff-orthonormal-columns-rows}, a matrix is in
  $\mathrm{O}_n(\R)$ if and only if its columns form an orthonormal
  basis of $\R^n$. Again, the reader should formulate parallel
  statements for hermitian products in
  $\C^n$. [\ref{exer:linalg:gram-schmidt-process}]
\end{exer}
\begin{exer}{}{exer:linalg:gram-schmidt-process}
  The \emph{Gram-Schmidt process} takes as input a choice of linearly
  independent vectors $\mathbf{v}_1, \dots, \mathbf{v}_r$ of $\R^n$
  and returns vectors $\mathbf{w}_1, \dots, \mathbf{w}_r$ spanning the
  same space $V$ spanned by $\mathbf{v}_1, \dots, \mathbf{v}_r$ and
  such that $(\mathbf{w}_i, \mathbf{w}_j) = 0$ for $i \neq j$. Scaling
  each $\mathbf{w}_i$ by its length, this yields an orthonormal basis
  for $V$.

  Here is how the Gram-Schmidt process works:
  \begin{itemize}
  \item[---] $\mathbf{w}_1 := \mathbf{v}_1$.
  \item[---] For $k \geq 1$, $\mathbf{w}_k := \mathbf{v}_k -$
    orthogonal projection of $\mathbf{v}_k$ onto
    $\langle \mathbf{w}_1, \dots, \mathbf{w}_{k-1} \rangle$.
  \end{itemize}
  (Cf.\
  \Cref{exer:linalg:orthonormal-vectors-coefficient-projection-formulas}.)
  Prove that this process accomplishes the stated goal.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:symmetric-hermitian-iff-self-adjoint}
  A matrix $M \in \mathcal{M}_n(\R)$ is \emph{symmetric} if $M^t =
  M$. Prove that $M$ is symmetric if and only if
  $(\forall \mathbf{v}, \mathbf{w} \in \C^n)$,
  $(M\mathbf{v}, \mathbf{w}) = (\mathbf{v}, M\mathbf{w})$.

  A matrix $M \in \mathcal{M}_n(\C)$ is \emph{hermitian} if
  $M^\dagger = M$. Prove that $M$ is hermitian if and only if
  $(\forall \mathbf{v}, \mathbf{w} \in \C^n)$,
  $(M\mathbf{v}, \mathbf{w}) = (\mathbf{v}, M\mathbf{w})$.

  In both cases, one may say that $M$ is \emph{self-adjoint}; this
  means that shuttling it from one side of the product to the other
  does not change the result of the operation.

  A hermitian matrix with real entries is symmetric. It is in fact
  useful to think of real symmetric matrices as particular cases of
  hermitian
  matrices. [\ref{exer:linalg:hermitian-eigenvalues-real-orthogonal-eigenvectors}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:hermitian-eigenvalues-real-orthogonal-eigenvectors}
  Prove that the eigenvalues of a hermitian matrix
  (\Cref{exer:linalg:symmetric-hermitian-iff-self-adjoint}) are
  real. Also, prove that if $\mathbf{v}, \mathbf{w}$ are eigenvectors
  of a hermitian matrix, corresponding to different eigenvalues, then
  $(\mathbf{v}, \mathbf{w}) = 0$. (Thus, eigenvectors with distinct
  eigenvalues for a real symmetric matrix are orthogonal.)
  [\ref{exer:linalg:real-symmetric-diagonalizable-orthonormal-eigenbasis}]
\end{exer}

\section{Canonical forms}
\label{sec:linalg:canonical-forms}

\subsection{Linear transformations of free modules; actions of
  polynomial rings}
\label{subsec:linalg:linear-transformations-of-free-modules-actions-of-polynomial-rings}
I have promised to use \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} to better understand the
similarity relation and to obtain special forms for square matrices
with entries in a field. I am now ready to make good on my promise.

The questions the reader should be anxiously asking are, where is the
PID? Where is the finitely generated module? Why would a
classification theorem for these things have anything to tell us about
matrices? Once these questions are answered, everything else will
follow easily. In a sense, this is the \emph{one} issue that I keep in
my mind concerning all these considerations: once I remember how to
get an interesting module out of a linear transformation of vector
spaces, the rest is essentially an immediate consequence of standard
notions.

Once more, while the main application will be to vector spaces and
matrices over a \emph{field}, we do not need to preoccupy ourselves
with specializing the situation before we get to the key
point. Therefore, let's keep working for a while longer on our given
integral domain\footnote{In fact, the requirements that $R$ be an
  integral domain and that $F$ be free are immaterial here; cf.\
  \Cref{exer:rings:Rx-module-structures}.} $R$.

\begin{prop}{}{prop:linalg:transformation-as-rt-module-structure}
  Giving a linear transformation on a free $R$-module $F$ is the same
  as giving an $R[t]$-module structure on $F$, compatible with its
  $R$-module structure.
\end{prop}

Here $R[t]$ is the polynomial ring in one indeterminate $t$ over the
ring $R$. The claim is much less impressive than it sounds; in fact,
it is completely tautological.

Giving an $R[t]$-module structure on $F$ compatible with the
$R$-module structure is the same as giving a homomorphism of
$R$-algebras
\[
  \varphi : R[t] \to \End_R(F)
\]
(this is an insignificant upgrade of the very definition of module
from \S{}\ref{subsec:rings:definition-of-left-r-module}). By the
universal property satisfied by polynomial rings
(\S{}\ref{subsec:rings:universal-property-of-polynomial-rings},
especially \Cref{exam:rings:poly-universal-general}), giving $\varphi$
as an extension of the basic $R$-algebra structure of $\End_R(F)$ is
just the same as specifying the image $\varphi(t)$, that is, choosing
an element of $\End_R(F)$. This is precisely what the claim says.

My propensity for the use of a certain language may obfuscate the
matter, which is very simple. Given a linear transformation $\alpha$
of a free module $F$, we can define the action of a polynomial
\[
  f(t) = r_m t^m + r_{m-1} t^{m-1} + \dots + r_0 \in R[t]
\]
on $F$ as follows: for every $\mathbf{v} \in F$, set
\[
  f(t)(\mathbf{v}) := r_m \alpha^m(\mathbf{v}) + r_{m-1}
  \alpha^{m-1}(\mathbf{v}) + \dots + r_0 \mathbf{v},
\]
where $\alpha^k$ denotes the $k$-fold composition of $\alpha$ with
itself; we have already run into this in
\S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}. Conversely,
an action of $R[t]$ on $F$ determines in particular a map
$\alpha : F \to F$, that is, `multiplication by $t$':
\[
  \mathbf{v} \mapsto t\mathbf{v};
\]
the compatibility with the $R$-module structure on $F$ tells us
precisely that $\alpha$ is a linear transformation of $F$. Again, this
is the content of Claim~\ref{prop:linalg:transformation-as-rt-module-structure}.

Tautological as it is, Claim~\ref{prop:linalg:transformation-as-rt-module-structure} packs a good amount of
information. Indeed, the $R[t]$-module knows everything about
similarity:

\begin{lem}{}{lem:linalg:rt-module-iso-iff-similar}
  Let $\alpha$, $\beta$ be linear transformations of a free $R$-module
  $F$. Then the corresponding $R[t]$-module structures on $F$ are
  isomorphic if and only if $\alpha$ and $\beta$ are similar.
\end{lem}

\begin{proof}
  Denote by $F_\alpha$, $F_\beta$ the two $R[t]$-modules defined on
  $F$ by $\alpha$, $\beta$ as per Claim~\ref{prop:linalg:transformation-as-rt-module-structure}.

  Assume first that $\alpha$ and $\beta$ are similar. Then there
  exists an invertible $R$-linear transformation $\pi : F \to F$ such
  that
  \[
    \beta = \pi \circ \alpha \circ \pi^{-1};
  \]
  that is, $\pi \circ \alpha = \beta \circ \pi$. We can view $\pi$ as
  an $R$-linear map
  \[
    F_\alpha \to F_\beta.
  \]
  I claim that it is $R[t]$-linear: indeed, multiplication by $t$ is
  $\alpha$ in $F_\alpha$ and $\beta$ in $F_\beta$, so
  \[
    \pi(t\mathbf{v}) = \pi \circ \alpha(\mathbf{v}) = \beta \circ
    \pi(\mathbf{v}) = t\pi(\mathbf{v}).
  \]
  Thus $\pi$ is an invertible $R[t]$-linear map
  $F_\alpha \to F_\beta$, proving that $F_\alpha$ and $F_\beta$ are
  isomorphic as $R[t]$-modules.

  The converse implication is obtained essentially by running this
  argument in reverse and is left to the reader
  (\Cref{exer:linalg:complete-proof-rt-module-iso-iff-similar}).
\end{proof}

\begin{coro}{}{coro:linalg:similarity-classes-correspond-to-rt-module-classes}
  There is a one-to-one correspondence between the similarity classes
  of $R$-linear transformations of a free $R$-module $F$ and the
  isomorphism classes of $R[t]$-module structures on $F$.
\end{coro}

Of course the same statement holds for square matrices with entries in
$R$, in the finite rank case.

Corollary~\ref{coro:linalg:similarity-classes-correspond-to-rt-module-classes} is the tool we were looking for, bridging
between classifying similarity classes of linear transformations or
matrices and classifying modules. The task now becomes that of
translating the notions we have encountered in
\S{}\ref{sec:linalg:linear-transformations-of-a-free-module} into the
module language and seeing if this teaches us anything new.

In any case, since we have classified finitely generated modules over
PIDs, it is clear that we will be able to classify similarity classes
of transformations of finite-rank free modules---provided that $R[t]$
is a PID. This is where the additional hypothesis that $R$ be a field
enters the discussion (cf.\ \Cref{exer:irred:rx-pid-implies-r-field}).

\subsection{\texorpdfstring{$k[t]$}{k[t]}-modules and the rational
  canonical form}
\label{subsec:linalg:k-t-modules-and-the-rational-canonical-form}
It is finally time to specialize to the case in which $R = k$ is a
field and $F$ has finite rank; so $F = V$ is simply a
finite-dimensional vector space. Let $n = \dim V$.

By the preceding discussion, choosing a linear transformation of $V$
is the same as giving $V$ a $k[t]$-module structure (compatible with
its vector space structure); similar linear transformations correspond
to isomorphic $k[t]$-modules. Then $V$ is a finitely generated module
over $k[t]$, and $k[t]$ is a PID since $k$ is a field
(\Cref{exer:rings:field-poly-ring-pid} and
\S{}\ref{sec:irred:ufds-pids-euclidean-domains}); therefore, we are
precisely in the situation covered by the classification theorem.

Even before spelling out the result of applying \Cref{thm:linalg:structure-theorem-fg-modules-over-pid}, we
can make use of its slogan version: every finitely generated module
over a PID is a direct sum of cyclic modules. This tells us that we
can understand all linear transformations of finite-dimensional vector
spaces, up to similarity, if we understand the linear transformation
given by multiplication by $t$ on a cyclic $k[t]$-module
\[
  V = \frac{k[t]}{(f(t))},
\]
where $f(t)$ is a nonconstant monic polynomial:
\[
  f(t) = t^n + r_{n-1} t^{n-1} + \dots + r_0.
\]
It is worthwhile obtaining a good matrix representation of this poster
case. We choose the basis
\[
  1, \quad t, \quad \dots, \quad t^{n-1}
\]
of $V$ (cf.\ \Cref{prop:rings:poly-group-iso}). Recall that the
columns of the matrix corresponding to a transformation consist of the
images of the chosen basis (cf.\ the comments preceding
\Cref{coro:linalg:matrix-ring-iso-to-hom}). Since multiplication by $t$ on $V$ acts as
\[
  \begin{cases}
    1 \mapsto t, \\
    t \mapsto t^2, \\
    \dots \\
    t^{n-1} \mapsto t^n = -r_{n-1} t^{n-1} - \dots - r_0,
  \end{cases}
\]
the matrix corresponding to this linear transformation is
\[
  \begin{pmatrix}
    0 & 0 & 0 & \dots & 0 & -r_0 \\
    1 & 0 & 0 & \dots & 0 & -r_1 \\
    0 & 1 & 0 & \dots & 0 & -r_2 \\
    \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
    0 & 0 & 0 & \dots & 0 & -r_{n-2} \\
    0 & 0 & 0 & \dots & 1 & -r_{n-1}
  \end{pmatrix}.
\]

\begin{defn}{}{defn:linalg:companion-matrix}
  This matrix is called the \emph{companion matrix} of the polynomial
  $f(t)$, denoted $C_{f(t)}$.
\end{defn}

\Cref{thm:linalg:structure-theorem-fg-modules-over-pid} tells us (among other things) that every linear
transformation admits a matrix representation into blocks, each of
which is the companion matrix of a polynomial. Here is the statement
of \Cref{thm:linalg:structure-theorem-fg-modules-over-pid} in this context:

\begin{thm}{}{thm:linalg:rational-canonical-form-existence}
  Let $k$ be a field, and let $V$ be a finite-dimensional vector
  space.  Let $\alpha$ be a linear transformation on $V$, and endow
  $V$ with the corresponding $k[t]$-module structure, as in
  Claim~\ref{prop:linalg:transformation-as-rt-module-structure}.  Then the following hold:
  \begin{itemize}
  \item There exist distinct monic irreducible polynomials
    $p_1(t), \dots, p_s(t) \in k[t]$ and positive integers $r_{ij}$
    such that
    \[
      V \cong \bigoplus_{i,j} \frac{k[t]}{(p_i(t)^{r_{ij}})}
    \]
    as $k[t]$-modules.
  \item There exist monic nonconstant polynomials
    $f_1(t), \dots, f_m(t) \in k[t]$ such that
    $f_1(t) \mid \dots \mid f_m(t)$ and
    \[
      V \cong \frac{k[t]}{(f_1(t))} \oplus \dots \oplus
      \frac{k[t]}{(f_m(t))}
    \]
    as $k[t]$-modules.
  \end{itemize}
  Via these isomorphisms, the action of $\alpha$ on $V$ corresponds to
  multiplication by $t$.

  Further, two linear transformations $\alpha$, $\beta$ are similar if
  and only if they have the same collections of invariants
  $p_i(t)^{r_{ij}}$ (`elementary divisors'), $f_i(t)$ (`invariant
  factors').
\end{thm}

\begin{proof}
  Since $\dim V$ is finite, $V$ is finitely generated as a $k$-module
  and a fortiori as a $k[t]$-module. The two isomorphisms are then
  obtained by applying \Cref{thm:linalg:structure-theorem-fg-modules-over-pid}. All the relevant polynomials
  may be chosen to be monic since every polynomial over a field is the
  associate of a (unique) monic polynomial
  (\Cref{exer:rings:field-poly-ideal-monic-generator}). The fact that
  the action of $\alpha$ corresponds to multiplication by $t$ is
  precisely what defines the corresponding $k[t]$-module structure on
  $V$. The statement about similar transformations follows from
  Corollary~\ref{coro:linalg:similarity-classes-correspond-to-rt-module-classes}.
\end{proof}

Theorem~\ref{thm:linalg:rational-canonical-form-existence} answers (over fields) the question raised in
\S{}\ref{subsec:linalg:endomorphisms-and-similarity}: we now have a
list of invariants which describe completely the similarity class of a
linear transformation. Further, these invariants provide us with a
special matrix representation of a given linear transformation.

\begin{defn}{}{defn:linalg:rational-canonical-form}
  The \emph{rational canonical form} of a linear transformation
  $\alpha$ of a vector space $V$ is the block matrix
  \[
    \begin{pmatrix}
      C_{f_1(t)} & & \\
                 & \ddots & \\
                 & & C_{f_m(t)}
    \end{pmatrix},
  \]
  where $f_1(t), \dots, f_m(t)$ are the invariant factors of $\alpha$.
\end{defn}

The rational canonical form of a square matrix is (of course) the
rational canonical form of the corresponding linear
transformation. The following statement is an immediate consequence of
\Cref{thm:linalg:rational-canonical-form-existence}.

\begin{coro}{}{coro:linalg:rational-canonical-form-exists-and-unique}
  Every linear transformation admits a rational canonical form. Two
  linear transformations have the same rational canonical form if and
  only if they are similar.
\end{coro}

\begin{rem}{}{rem:linalg:rational-canonical-form-naming}
  The `rational' in rational canonical form has nothing to do with
  $\Q$; it is meant to remind the reader that this form can be found
  without leaving the base field. The other canonical form we will
  encounter will have entries in a possibly larger field, where the
  characteristic polynomial of the transformation factors completely.
\end{rem}

Corollary~\ref{coro:linalg:rational-canonical-form-exists-and-unique} fulfills explicitly another wish expressed
at the end of \S{}\ref{subsec:linalg:endomorphisms-and-similarity},
that is, to find one distinguished representative in each similarity
class of matrices/linear transformations. The rational canonical form
is such a representative.

The next obvious question is how the invariants we just found relate
to our more naive attempts to produce invariants of similar
transformations in
\S{}\ref{sec:linalg:linear-transformations-of-a-free-module}.

\begin{prop}{}{prop:linalg:minimal-and-characteristic-poly-from-invariant-factors}
  Let $f_1(t) \mid \dots \mid f_m(t)$ be the invariant factors of a
  linear transformation $\alpha$ on a vector space $V$. Then the
  minimal polynomial $m_\alpha(t)$ equals $f_m(t)$, and the
  characteristic polynomial $P_\alpha(t)$ equals the product
  $f_1(t) \cdots f_m(t)$.
\end{prop}

\begin{proof}
  Tracing definitions, $(m_\alpha(t))$ is the annihilator ideal of $V$
  when this is viewed as a $k[t]$-module via $\alpha$ (as in
  Claim~\ref{prop:linalg:transformation-as-rt-module-structure}). Therefore the equality of $m_\alpha(t)$ and
  $f_m(t)$ is a restatement of \Cref{lem:linalg:torsion-module-annihilator-and-primes}.

  Concerning the characteristic polynomial, by
  \Cref{exer:linalg:char-poly-multiplicative-direct-sums} it suffices
  to prove the statement in the cyclic case: that is, it suffices to
  prove that if $f(t)$ is a monic polynomial, then $f(t)$ equals the
  characteristic polynomial of the companion matrix $C_{f(t)}$.
  Explicitly, let
  \[
    f(t) = t^n + r_{n-1} t^{n-1} + \dots + r_0
  \]
  be a monic polynomial; then the task amounts to showing that
  \[
    \det \begin{pmatrix}
      t & 0 & 0 & \dots & 0 & r_0 \\
      -1 & t & 0 & \dots & 0 & r_1 \\
      0 & -1 & t & \dots & 0 & r_2 \\
      \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
      0 & 0 & 0 & \dots & t & r_{n-2} \\
      0 & 0 & 0 & \dots & -1 & t + r_{n-1}
    \end{pmatrix} = f(t).
  \]
  This can be done by an easy induction and is left to the reader
  (\Cref{exer:linalg:char-poly-of-companion-matrix}).
\end{proof}

\begin{coro}{Cayley-Hamilton}{coro:linalg:minimal-poly-divides-characteristic-poly}
  The minimal polynomial of a linear transformation divides its
  characteristic polynomial.
\end{coro}

\begin{proof}
  This has now become evident, as promised in
  \S{}\ref{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}.
\end{proof}

Finding the rational canonical form of a given matrix $A$ amounts to
working out the classification theorem for the specific $k[t]$-module
corresponding to $A$ as in Claim~\ref{prop:linalg:transformation-as-rt-module-structure}. As $k[t]$ is in
fact a Euclidean domain, we know that this can be done by Gaussian
elimination (cf.\
\S{}\ref{subsec:linalg:gaussian-elimination-over-euclidean-domains}). In
practice, the process consists of compiling the information obtained
while diagonalizing $tI - A$ over $k[t]$ by elementary operations. The
reader is invited to either produce an original algorithm or at least
look it up in a standard reference to see what is involved.

Major shortcuts may come to our aid. For example, if the minimal and
characteristic polynomials coincide, then one knows a priori that the
corresponding module is cyclic (cf.\ \Cref{rem:linalg:torsion-module-invariant-factors-summary}), and it
follows that the rational canonical form is simply the companion
matrix of the characteristic polynomial.

In any case, the strength of a canonical form rests in the fact that
it allows us to reduce general facts to specific, standard cases. For
example, the following statement is somewhat mysterious if all one
knows are the definitions, but it becomes essentially trivial with a
pinch of rational canonical forms:

\begin{prop}{}{prop:linalg:matrix-similar-to-transpose}
  Let $A \in \mathcal{M}_n(k)$ be a square matrix. Then $A$ is similar
  to its transpose.
\end{prop}

\begin{proof}
  If $B$ is similar to $A$ and we can prove that $B$ is similar to its
  transpose $B^t$, then $A$ is similar to its transpose $A^t$: because
  $B = PAP^{-1}$, $B^t = QBQ^{-1}$ give
  \[
    A^t = (P^t Q P) A (P^t Q P)^{-1}.
  \]
  Therefore, it suffices to prove the statement for matrices in
  rational canonical form.

  Further, to prove the statement for a block matrix, it clearly
  suffices to prove it for each block; so we may assume that $A$ is
  the companion matrix $C$ of a polynomial $f(t)$. Since the
  characteristic and minimal polynomials of the transpose $C^t$
  coincide with those of $C$
  (\Cref{exer:linalg:matrix-transpose-same-char-poly-annihilator}),
  they are both equal to $f(t)$. It follows that the rational
  canonical form of $C^t$ is again the companion matrix to $f(t)$;
  therefore $C^t$ and $C$ are similar, as needed.
\end{proof}

\subsection{Jordan canonical form}
\label{subsec:linalg:jordan-canonical-form}
% We have derived the rational canonical form from

% the invariant factors of the module corresponding to a linear
% transformation. A useful alternative can be obtained from the
% elementary divisors, at least in the case in which the
% characteristic polynomial factors completely over the field k. If
% this is not the case, one can enlarge k so as to include all the
% roots of the characteristic polynomial: the reader proved this in
% Exercise V.5.13. The price to pay will be that the Jordan canonical
% form of a linear transformation \alpha \in Endk (V ) may be a matrix
% with entries in a field larger than k. In any case, whether two
% transformations are
similar or not is independent of the base field
(\Cref{exer:linalg:similarity-over-k-iff-over-extension}), so this
does not affect the issue at hand.

Given $\alpha \in \End_k(V)$, obtain the elementary divisor
decomposition of the corresponding $k[t]$-module, as in
Theorem~\ref{thm:linalg:rational-canonical-form-existence}:
\[
  V \cong \bigoplus_{i,j} \frac{k[t]}{(p_i(t)^{r_{ij}})}.
\]
It is an immediate consequence of Proposition~\ref{prop:linalg:minimal-and-characteristic-poly-from-invariant-factors} and the
equivalence between the elementary divisor and invariant factor
formulations that the characteristic polynomial $P_\alpha(t)$ equals
the product
\[
  \prod_{i,j} p_i(t)^{r_{ij}}.
\]

\begin{lem}{}{lem:linalg:characteristic-poly-factors-completely-setup}
  Assume that the characteristic polynomial $P_\alpha(t)$ factors
  completely; that is,
  \[
    P_\alpha(t) = \prod_{i=1}^s (t - \lambda_i)^{m_i}
  \]
  where $\lambda_i$, $i = 1, \dots, s$, are the distinct eigenvalues
  of $\alpha$ (and $m_i$ are their algebraic multiplicities; cf.\
  \S{}\ref{subsec:linalg:eigenvalues-eigenvectors-eigenspaces}).  Then
  $p_i(t) = (t - \lambda_i)$, and $m_i = \sum_j r_{ij}$.

  In this situation, the minimal polynomial of $\alpha$ equals
  \[
    m_\alpha(t) = \prod_{i=1}^s (t - \lambda_i)^{\max_j \{r_{ij}\}}.
  \]
\end{lem}

\begin{proof}
  The first statement follows from uniqueness of factorizations. The
  statement about the minimal polynomial is immediate from
  Proposition~\ref{prop:linalg:minimal-and-characteristic-poly-from-invariant-factors} and the bookkeeping giving the
  equivalence of the two formulations in Theorem~\ref{thm:linalg:rational-canonical-form-existence}.
\end{proof}

The elementary divisor decomposition splits $V$ into a different
collection of cyclic modules than the decomposition in invariant
factors: the basic cyclic bricks are now of the form $k[t]/(p(t)^r)$
for a monic prime $p(t)$; by Lemma~\ref{lem:linalg:characteristic-poly-factors-completely-setup}, assuming that
the characteristic polynomial factors completely over $k$, they are in
fact of the form
\[
  \frac{k[t]}{((t - \lambda)^r)}
\]
for some $\lambda \in k$ (which equals an eigenvalue of $\alpha$) and
$r > 0$. Just as we did for `companion matrices', we now look for a
basis with respect to which $\alpha$ (= multiplication by $t$; keep in
mind the fine print in Theorem~\ref{thm:linalg:rational-canonical-form-existence}) has a particularly
simple matrix representation. This time we choose the basis
\[
  (t - \lambda)^{r-1}, \quad (t - \lambda)^{r-2}, \quad \dots, \quad
  (t - \lambda)^0 = 1.
\]
Multiplication by $t$ on $V$ acts (in the first line, use the fact
that $(t - \lambda)^r = 0$ in $V$) as
\[
  \begin{cases}
    (t - \lambda)^{r-1} \mapsto t(t - \lambda)^{r-1} = \lambda(t -
    \lambda)^{r-1} + (t - \lambda)^r = \lambda(t - \lambda)^{r-1}, \\
    (t - \lambda)^{r-2} \mapsto t(t - \lambda)^{r-2} = (t - \lambda)^{r-1} +
    \lambda(t - \lambda)^{r-2}, \\
    \dots \\
    1 \mapsto t = (t - \lambda) + \lambda.
  \end{cases}
\]
Therefore, with respect to this basis the linear transformation has
matrix
\[
  \begin{pmatrix}
    \lambda & 1 & 0 & \dots & 0 & 0 \\
    0 & \lambda & 1 & \dots & 0 & 0 \\
    0 & 0 & \lambda & \dots & 0 & 0 \\
    \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
    0 & 0 & 0 & \dots & \lambda & 1 \\
    0 & 0 & 0 & \dots & 0 & \lambda
  \end{pmatrix}.
\]

\begin{defn}{}{defn:linalg:jordan-block}
  This matrix is the \emph{Jordan block} of size $r$ corresponding to
  $\lambda$, denoted $J_{\lambda, r}$.
\end{defn}

We can put several blocks together for a given $\lambda$: for
$(r_j) = (r_1, \dots, r_\ell)$, let
\[
  J_{\lambda, (r_j)} := \begin{pmatrix}
    J_{\lambda, r_1} & & \\
                     & \ddots & \\
                     & & J_{\lambda, r_\ell}
  \end{pmatrix}.
\]
With this notation in hand, we get new distinguished representatives
of the similarity class of a given linear transformation:

\begin{defn}{}{defn:linalg:jordan-canonical-form}
  The \emph{Jordan canonical form} of a linear transformation $\alpha$
  of a vector space $V$ is the block matrix
  \[
    \begin{pmatrix}
      J_{\lambda_1, (r_{1j})} & & \\
                              & \ddots & \\
                              & & J_{\lambda_s, (r_{sj})}
    \end{pmatrix},
  \]
  where $(t - \lambda_i)^{r_{ij}}$ are the elementary divisors of
  $\alpha$ (cf.\ Lemma~\ref{lem:linalg:characteristic-poly-factors-completely-setup}).
\end{defn}

This canonical form is `a little less canonical' than the rational
canonical form, in the sense that while the rational canonical form is
\emph{really} unique, some ambiguity is left in the Jordan canonical
form: there is no special way to choose the order of the different
blocks any more than there is a way to order\footnote{Of course if the
  ground field is $\Q$ or $\R$, then we can order the $\lambda_i$'s
  in, for example, increasing order. However, this is not an option on
  fields like $\C$.} the eigenvalues $\lambda_1, \dots, \lambda_s$.

Therefore, the analog of Corollary~\ref{coro:linalg:rational-canonical-form-exists-and-unique} should state that
two linear transformations are similar if and only if they have the
same Jordan canonical form up to a reordering of the blocks. As we
have seen, every linear transformation $\alpha \in \End_k(V)$ admits a
Jordan canonical form if $P_\alpha(t)$ factors completely over $k$;
for example, we can always find a Jordan canonical form with entries
in the algebraic closure of $k$.

\begin{exam}{}{exam:linalg:jordan-form-similarity-classes-example}
  One use of the Jordan canonical form is the enumeration of all
  possible similarity classes of transformations with given
  eigenvalues. For example, there are $5$ similarity classes of linear
  transformations with a single eigenvalue $\lambda$ with algebraic
  multiplicity $4$, over a $4$-dimensional vector space: indeed, there
  are $5$ different ways to stack together Jordan blocks corresponding
  to the same eigenvalue, within a $4 \times 4$ square matrix:
  \[
    \begin{pmatrix}
      \lambda & 0 & 0 & 0 \\
      0 & \lambda & 0 & 0 \\
      0 & 0 & \lambda & 0 \\
      0 & 0 & 0 & \lambda
    \end{pmatrix}, \quad
    \begin{pmatrix}
      \lambda & 1 & 0 & 0 \\
      0 & \lambda & 0 & 0 \\
      0 & 0 & \lambda & 0 \\
      0 & 0 & 0 & \lambda
    \end{pmatrix}, \quad
    \begin{pmatrix}
      \lambda & 1 & 0 & 0 \\
      0 & \lambda & 0 & 0 \\
      0 & 0 & \lambda & 1 \\
      0 & 0 & 0 & \lambda
    \end{pmatrix},
  \]
  \[
    \begin{pmatrix}
      \lambda & 1 & 0 & 0 \\
      0 & \lambda & 1 & 0 \\
      0 & 0 & \lambda & 0 \\
      0 & 0 & 0 & \lambda
    \end{pmatrix}, \quad
    \begin{pmatrix}
      \lambda & 1 & 0 & 0 \\
      0 & \lambda & 1 & 0 \\
      0 & 0 & \lambda & 1 \\
      0 & 0 & 0 & \lambda
    \end{pmatrix}.
  \]
\end{exam}

The Jordan canonical form clarifies the difference between
\emph{algebraic} and \emph{geometric} multiplicities of an eigenvalue.
The algebraic multiplicity of $\lambda$ as an eigenvalue of a linear
transformation $\alpha$ is (of course) the number of times $\lambda$
appears in the Jordan canonical form of $\alpha$. Recall that the
geometric multiplicity of $\lambda$ equals the dimension of the
eigenspace of $\lambda$.

\begin{prop}{}{prop:linalg:geometric-multiplicity-equals-jordan-block-count}
  The geometric multiplicity of $\lambda$ as an eigenvalue of $\alpha$
  equals the number of Jordan blocks corresponding to $\lambda$ in the
  Jordan canonical form of $\alpha$.
\end{prop}

\begin{proof}
  As the geometric multiplicity is clearly additive in direct sums, it
  suffices to show that the geometric multiplicity of $\lambda$ for
  the transformation corresponding to a single Jordan block
  \[
    J = \begin{pmatrix}
      \lambda & 1 & \dots & 0 & 0 \\
      0 & \lambda & \dots & 0 & 0 \\
      \vdots & \vdots & \ddots & \vdots & \vdots \\
      0 & 0 & \dots & \lambda & 1 \\
      0 & 0 & \dots & 0 & \lambda
    \end{pmatrix}
  \]
  is $1$.

  Let
  $\mathbf{v} = \begin{pmatrix} v_1 \\ \vdots \\ v_r \end{pmatrix}$ be
  an eigenvector corresponding to $\lambda$. Then
  \[
    \lambda \begin{pmatrix} v_1 \\ v_2 \\ \vdots \\ v_r \end{pmatrix}
    = J \begin{pmatrix} v_1 \\ v_2 \\ \vdots \\ v_r \end{pmatrix} =
    \begin{pmatrix} \lambda v_1 + v_2 \\ \lambda v_2 + v_3 \\ \vdots \\
      \lambda v_r \end{pmatrix},
  \]
  yielding
  \[
    v_2 = \dots = v_r = 0.
  \]
  That is, the eigenspace of $\lambda$ is generated by
  \[
    \mathbf{v} = \begin{pmatrix} 1 \\ 0 \\ \vdots \\ 0 \end{pmatrix}
  \]
  and has dimension $1$, as needed.
\end{proof}

\subsection{Diagonalizability}
\label{subsec:linalg:diagonalizability}
A linear transformation of a vector space $V$ is \emph{diagonalizable}
if it can be represented by a diagonal matrix.

The question of whether a given linear transformation is or is not
diagonalizable is interesting and important: as I pointed out already
in \S{}\ref{subsec:linalg:eigenvalues-eigenvectors-eigenspaces}, when
$\alpha \in \End_k(V)$ is diagonalizable, then the vector space $V$
admits a corresponding spectral decomposition: $\alpha$ is
diagonalizable if and only if $V$ admits a basis of eigenvectors of
$\alpha$. Much more could be said on this important theme, but we are
already in the position of giving useful criteria for
diagonalizability.

For example, the diagonal matrix representing $\alpha$ will
necessarily be a Jordan canonical form for $\alpha$, consisting of
blocks of size $1$. Proposition~\ref{prop:linalg:geometric-multiplicity-equals-jordan-block-count} tells us that this
can be detected by the difference between algebraic and geometric
multiplicity, so we get the following characterization of
diagonalizable transformations:

\begin{coro}{}{coro:linalg:diagonalizable-iff-multiplicities-equal}
  Assume the characteristic polynomial of $\alpha \in
  \End_k(V)$ factors completely over $k$. Then $\alpha$ is
  diagonalizable if and only if the geometric and algebraic
  multiplicities of all its eigenvalues coincide.
\end{coro}

Another view of the same result is the following:

\begin{prop}{}{prop:linalg:diagonalizable-iff-minimal-poly-squarefree}
  Assume the characteristic polynomial of $\alpha \in
  \End_k(V)$ factors completely over $k$. Then $\alpha$ is
  diagonalizable if and only if the minimal polynomial of $\alpha$ has
  no multiple roots.
\end{prop}

\begin{proof}
  Again, diagonalizability is equivalent to having all Jordan blocks
  of size $1$ in the Jordan canonical form of $\alpha$. Therefore, if
  the characteristic polynomial of $\alpha$ factors completely, then
  $\alpha$ is diagonalizable if and only if all exponents $r_{ij}$
  appearing in Theorem~\ref{thm:linalg:rational-canonical-form-existence} equal $1$. By the second part
  of Lemma~\ref{lem:linalg:characteristic-poly-factors-completely-setup}, this is equivalent to the requirement
  that the minimal polynomial of $\alpha$ have no multiple roots.
\end{proof}

The condition of factorizability in these statements is necessary in
order to guarantee that a Jordan canonical form exists. Not
surprisingly, whether a matrix is diagonalizable or not depends on the
ground field. For example,
\[
  \begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}
\]
is not diagonalizable over $\R$ (cf.\ \Cref{exam:linalg:rotation-matrix-no-real-eigenvalues}), while it is
diagonalizable over $\C$.

Endowing the vector space $V$ with additional structure (such as an
inner product) singles out important classes of operators, for which
one can obtain precise and delicate results on the existence of
spectral decompositions. For example, this can be used to prove that
real symmetric matrices are diagonalizable
(\Cref{exer:linalg:real-symmetric-diagonalizable-orthonormal-eigenbasis}.)
The reader will get a taste of these `spectral theorems' in the
exercises.

\subsection*{Exercises}

As above, $k$ denotes a \emph{field}.

\begin{exer}{\exerused}{exer:linalg:complete-proof-rt-module-iso-iff-similar}
  Complete the proof of Lemma~\ref{lem:linalg:rt-module-iso-iff-similar}. [\S{}
  ef{subsec:linalg:linear-transformations-of-free-modules-actions-of-polynomial-rings}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:char-poly-of-companion-matrix}
  Prove that the characteristic polynomial of the companion matrix of
  a monic polynomial $f(t)$ equals $f(t)$. [\S{}
  ef{subsec:linalg:k-t-modules-and-the-rational-canonical-form}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:similar-iff-same-char-min-poly-low-dim}
  Prove that two linear transformations of a vector space of dimension
  $\leq 3$ are similar if and only if they have the same
  characteristic and minimal polynomials. Is this true in dimension
  $4$? [\S{}
  ef{subsec:linalg:the-characteristic-and-minimal-polynomials-of-an-endomorphism}]
\end{exer}
\begin{exer}{\exerused}{exer:linalg:similarity-over-k-iff-over-extension}
  Let $k$ be a field, and let $K$ be a field containing $k$. Two
  square matrices $A, B \in \mathcal{M}_n(k)$ may be viewed as
  matrices with entries in the larger field $K$. Prove that $A$ and
  $B$ are similar over $k$ if and only if they are similar over
  $K$. [\S{} ef{subsec:linalg:jordan-canonical-form}]
\end{exer}
\begin{exer}{}{exer:linalg:rational-canonical-form-of-diagonal-matrix}
  Find the rational canonical form of a diagonal matrix
  \[
    \begin{pmatrix}
      \lambda_1 & 0 & \dots & 0 \\
      0 & \lambda_2 & \dots & 0 \\
      \vdots & \vdots & \ddots & \vdots \\
      0 & 0 & \dots & \lambda_r
    \end{pmatrix},
  \]
  assuming that the $\lambda_i$ are all distinct.
\end{exer}
\begin{exer}{}{exer:linalg:compute-char-poly-min-poly-rcf-example}
  Let
  \[
    A = \begin{pmatrix} 6 & -10 & -10 \\ 3 & -5 & -6 \\ -1 & 2 & 3
    \end{pmatrix}.
  \]
  \begin{itemize}
  \item Compute the characteristic polynomial of $A$.
  \item Find the minimal polynomial of $A$ (use the Cayley-Hamilton
    theorem!).
  \item Find the invariant factors of $A$.
  \item Find the rational canonical form of $A$.
  \end{itemize}
\end{exer}
\begin{exer}{}{exer:linalg:min-eq-char-poly-iff-cyclic-vector-exists}
  Let $V$ be a $k$-vector space of dimension $n$, and let $\alpha \in
  \End_k(V)$. Prove that the minimal and characteristic polynomials of
  $\alpha$ coincide if and only if there is a vector
  $\mathbf{v} \in V$ such that
  \[
    \mathbf{v}, \quad \alpha(\mathbf{v}), \quad \dots, \quad
    \alpha^{n-1}(\mathbf{v})
  \]
  is a basis of $V$.
\end{exer}
\begin{exer}{}{exer:linalg:char-poly-divides-power-of-min-poly}
  Let $V$ be a $k$-vector space of dimension $n$, and let $\alpha \in
  \End_k(V)$. Prove that the characteristic polynomial $P_\alpha(t)$
  divides a power of the minimal polynomial $m_\alpha(t)$.
\end{exer}
\begin{exer}{}{exer:linalg:num-similarity-classes-single-eigenvalue}
  What is the number of distinct similarity classes of linear
  transformations on an $n$-dimensional vector space with one fixed
  eigenvalue $\lambda$ with algebraic multiplicity $n$?
\end{exer}
\begin{exer}{}{exer:linalg:classify-idempotent-matrices-up-to-similarity}
  Classify all square matrices $A \in \mathcal{M}_n(k)$ such that
  $A^2 = A$, up to similarity. Describe the action of such matrices
  `geometrically'.
\end{exer}
\begin{exer}{}{exer:linalg:nilpotent-matrices-jordan-form-and-trace}
  A square matrix $A \in \mathcal{M}_n(k)$ is \emph{nilpotent} (cf.\
  \Cref{exer:irred:nilpotent-implies-1-plus-a-unit}) if $A^k = 0$ for
  some integer $k$.
  \begin{itemize}
  \item Characterize nilpotent matrices in terms of their Jordan
    canonical form.
  \item Prove that if $A^k = 0$ for some integer $k$, then $A^k = 0$
    for some integer $k$ no larger than $n$ (= the size of the
    matrix).
  \item Prove that the trace of a nilpotent matrix is $0$.
  \end{itemize}
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:diagonalizable-restricted-to-invariant-subspace}
  Let $V$ be a finite-dimensional $k$-vector space, and let
  $\alpha \in
  \End_k(V)$ be a diagonalizable linear transformation.  Assume that
  $W \subseteq V$ is an invariant subspace, so that $\alpha$ induces a
  linear transformation $\alpha|_W \in \End_k(W)$.  Prove that
  $\alpha|_W$ is also diagonalizable. (Use
  Proposition~\ref{prop:linalg:diagonalizable-iff-minimal-poly-squarefree}.)
  [\ref{exer:linalg:commuting-diagonalizable-simultaneously-diagonalizable}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:commuting-matrix-diagonal-in-eigenbasis}
  Let $R$ be an integral domain. Assume that $A \in \mathcal{M}_n(R)$
  is diagonalizable, with distinct eigenvalues. Let
  $B \in \mathcal{M}_n(R)$ be such that $AB = BA$. Prove that $B$ is
  also diagonalizable, and in fact it is diagonal w.r.t.\ a basis of
  eigenvectors of $A$. (If $P$ is such that $P^{-1}AP$ is diagonal,
  note that $P^{-1}AP$ and $P^{-1}BP$ also commute.)
  [\ref{exer:linalg:commuting-diagonalizable-simultaneously-diagonalizable}]
\end{exer}
\begin{exer}{}{exer:linalg:commuting-diagonalizable-simultaneously-diagonalizable}
  Prove that `commuting transformations may be simultaneously
  diagonalized', in the following sense. Let $V$ be a
  finite-dimensional vector space, and let
  $\alpha, \beta \in \End_k(V)$ be diagonalizable
  transformations. Assume that $\alpha\beta = \beta\alpha$.  Prove
  that $V$ has a basis consisting of eigenvectors of both $\alpha$ and
  $\beta$. (Argue as in
  Exercise~\ref{exer:linalg:commuting-matrix-diagonal-in-eigenbasis}
  to reduce to the case in which $V$ is an eigenspace for $\alpha$;
  then use
  Exercise~\ref{exer:linalg:diagonalizable-restricted-to-invariant-subspace}.)
\end{exer}
\begin{exer}{}{exer:linalg:linear-transformation-preserves-complete-flag}
  A \emph{complete flag} of subspaces of a vector space $V$ of
  dimension $n$ is a sequence of nested subspaces
  \[
    0 = V_0 \subsetneq V_1 \subsetneq \dots \subsetneq V_{n-1}
    \subsetneq V_n = V
  \]
  with $\dim V_i = i$. In other words, a complete flag is a
  composition series in the sense of
  \Cref{exer:linalg:jordan-holder-for-modules}.

  Let $V$ be a finite-dimensional vector space over an algebraically
  closed field. Prove that every linear transformation $\alpha$ of $V$
  preserves a complete flag: that is, there is a complete flag as
  above and such that $\alpha(V_i) \subseteq V_i$.

  Find a linear transformation of $\R^2$ that does not preserve a
  complete flag.
\end{exer}
\begin{exer}{}{exer:linalg:schur-form}
  (Schur form) Let $A \in \mathcal{M}_n(\C)$. Prove that there exists
  a unitary matrix $P \in \mathrm{U}(n)$ such that
  \[
    A = P \begin{pmatrix}
      \lambda_1 & \ast & \dots & \ast \\
      0 & \lambda_2 & \dots & \ast \\
      \vdots & \vdots & \ddots & \vdots \\
      0 & 0 & \dots & \lambda_n
    \end{pmatrix} P^{-1}.
  \]
  So not only is $A$ similar to an upper triangular matrix (this is
  already guaranteed by the Jordan canonical form), but a matrix of
  change of basis $P$ `triangularizing' the matrix $A$ can in fact be
  chosen to be in $\mathrm{U}(n)$. (Argue inductively on $n$. Since
  $\C$ is algebraically closed, $A$ has at least one eigenvector
  $\mathbf{v}_1$, and we may assume
  $(\mathbf{v}_1, \mathbf{v}_1) = 1$.  For the induction step,
  consider the complement $\mathbf{v}_1^\perp$; cf.\
  \Cref{exer:linalg:orthogonal-complement-is-subspace-dim-n-minus-1},
  and keep in mind
  \Cref{exer:linalg:on-r-iff-orthonormal-columns-rows}.)

  The upper triangular matrix is a \emph{Schur form} for $A$. It is
  (of course) not unique.
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:triangular-normal-matrices-are-diagonal}
  A matrix $M \in \mathcal{M}_n(\C)$ is \emph{normal} if
  $MM^\dagger = M^\dagger M$. Note that unitary matrices and hermitian
  matrices are both normal.

  Prove that triangular normal matrices are
  diagonal. [\ref{exer:linalg:spectral-theorem-for-normal-operators}]
\end{exer}
\begin{exer}{$\lnot$}{exer:linalg:spectral-theorem-for-normal-operators}
  Prove the \emph{spectral theorem for normal operators}: if $M$ is a
  normal matrix, then there exists an orthonormal basis of
  eigenvectors of $M$. Therefore, normal operators are diagonalizable;
  they determine a spectral decomposition of the vector space on which
  they act.  (Consider the Schur form of $M$; cf.\
  Exercise~\ref{exer:linalg:schur-form}.  Use
  Exercise~\ref{exer:linalg:triangular-normal-matrices-are-diagonal}.)
  [\ref{exer:linalg:normal-iff-orthonormal-eigenbasis-exists},
  \ref{exer:linalg:real-symmetric-diagonalizable-orthonormal-eigenbasis}]
\end{exer}
\begin{exer}{}{exer:linalg:normal-iff-orthonormal-eigenbasis-exists}
  Prove that a matrix $M \in \mathcal{M}_n(\C)$ is normal if and only
  if it admits an orthonormal basis of eigenvectors.
  (Exercise~\ref{exer:linalg:spectral-theorem-for-normal-operators}
  gives one direction; prove the converse.)
\end{exer}
\begin{exer}{\exerused}{exer:linalg:real-symmetric-diagonalizable-orthonormal-eigenbasis}
  Prove that real symmetric matrices are diagonalizable and admit an
  orthonormal basis of eigenvectors. (This is an immediate consequence
  of Exercise~\ref{exer:linalg:spectral-theorem-for-normal-operators}
  and
  \Cref{exer:linalg:hermitian-eigenvalues-real-orthogonal-eigenvectors}.)
  [\S{} ef{subsec:linalg:diagonalizability}]
\end{exer}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-engine: luatex
%%% TeX-master: "../master"
%%% End:
